### 2019-04-20 paper.tex
paper

%
% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf]{acmart}
\settopmatter{printacmref=false}
\usepackage[T1]{fontenc}
%\usepackage{lmodern}
\usepackage{mathtools}
%
% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

% Rights management information. 
% This information is sent to you when you complete the rights form.
% These commands have SAMPLE values in them; it is your responsibility as an author to replace
% the commands and values with those provided to you when you complete the rights form.
%
% These commands are for a PROCEEDINGS abstract or paper.
\copyrightyear{2018}
\acmYear{2018}
\setcopyright{acmlicensed}
\acmConference[RecSys '19]{Recomme '18: ACM Symposium on Neural Gaze Detection}{September 16--20, 2019}{Copenhagen, Denmark}
\acmBooktitle{RecSys '19: ACM Conference on Neural Gaze Detection, June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmDOI{10.1145/1122445.1122456}
\acmISBN{978-1-4503-9999-9/18/06}

%
% These commands are for a JOURNAL article.
%\setcopyright{acmcopyright}
%\acmJournal{TOG}
%\acmYear{2018}\acmVolume{37}\acmNumber{4}\acmArticle{111}\acmMonth{8}
%\acmDOI{10.1145/1122445.1122456}

%
% Submission ID. 
% Use this when submitting an article to a sponsored event. You'll receive a unique submission ID from the organizers
% of the event, and this ID should be used as the parameter to this command.
%\acmSubmissionID{123-A56-BU3}

%
% The majority of ACM publications use numbered citations and references. If you are preparing content for an event
% sponsored by ACM SIGGRAPH, you must use the "author year" style of citations and references. Uncommenting
% the next command will enable that style.
%\citestyle{acmauthoryear}

%
% end of the preamble, start of the body of the document source.
\begin{document}

%
% The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
\title{The Ex-Ante View of RS Design}
%
% By default, the full list of authors will be used in the page headers. Often, this list is too long, and will overlap
% other information printed in the page headers. This command allows the author to define a more concise list
% of authors' names for this purpose.

%
% The abstract is a short summary of the work to be presented in the article.
\begin{abstract}
Recommender systems (RS) are traditionally deployed in environments where users are uncertain about their preferences and thus face a problem of choice under uncertainty, but most popular design approaches ignore this fact. We argue that predicting and modeling consumer choice in these contexts can improve the usefulness of RS and reframe the RS problem as providing useful information to help reduce user uncertainty as opposed to simply predicting user preferences. We show how this insight can be utilized to design RS that mitigate the negative consequences of RS such as filter bubble and user homogenization effects as well as better understand the role that RS play in contributing to these phenomena.
\end{abstract}

%
% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
% Please copy and paste the code instead of the example below.
%

%
% Keywords. The author(s) should pick words that accurately describe the work being
% presented. Separate the keywords with commas.
\keywords{Recommender Systems, Beliefs, Decision Theory, Filter Bubbles}

%
% This command processes the author and affiliation and title information and builds
% the first part of the formatted document.
\maketitle

\section{Introduction}

Recommender Systems (RS) have been widely deployed in online platforms such as Netflix, Amazon, and YouTube. On these platforms the combination of the fact that most goods are experience goods and that the number of choices available to users is incredibly large results in users not knowing their true preference ordering over all of the items. For instance, on Netflix there are thousands of movies, on Amazon there are millions of products, and on YouTube there are billions of videos. As a result, RS have been useful in helping users make decisions on these platforms because they provide users with information to help them make their choices under uncertainty. 
\par
However, while acknowledging that this is why RS are useful, most of the literature on RS ignores this user uncertainty and takes what we call an \textit{ex-post} viewpoint to preferences, focusing on the reported utility of users after they consume the item or inferring them via behavioral data \cite{adomavicius2005toward, zhao2018interpreting}. The RS problem traditionally is cast as solving the following prediction problem:
\begin{align*}
\arg\max\limits_{x \in A} u(x)
\end{align*}
\noindent where $A$ is the set of possible items and $u(x)$ represents the utility a user gets to consume good $x$. Recently there has been a move away from metrics focused on maximizing the accuracy of this prediction problem and towards optimizing for metrics such as serendipity or discoverability \cite{mcnee2006being, vargas2011rank} due to  the realization that accurate recommendations may not be the most useful recommendations for users. Furthermore, deployed RS have come under scrutiny for leading to ``filter bubbles" \cite{pariser2011filter} where users consume a less diverse set of content over time as well as increased homogenization amongst users \cite{chaney2018algorithmic, hosanagar2013will}.
\par
In this paper we want to put forward the argument that one approach to overcoming these problems is to make understanding how users choose which goods to consume a first-order consideration when designing RS. In particular, we differentiate between the previous \textit{ex-post} interpretation of the RS problem and what we define as the \textit{ex-ante} viewpoint. Unlike the ex-post viewpoint taken in the traditional prediction problem, the ex-ante viewpoint simply states that users face uncertainty and so maximize their expected utility as is defined in economic decision theory under uncertainty \citep[see e.g.][]{mas1995microeconomic}:
\begin{align*}
\arg\max\limits_{x \in A} \mathbb{E}[u(x)]
\end{align*}
Fundamentally, the move away from accuracy towards alternative metrics as well as tensions that have arisen from increased user homogenization and filter bubbles come from the \textit{choices} that users make. This shift from thinking about true preferences after all uncertainty is resolved (the ex-post view) to user decisions under uncertainty (the ex-ante view) can serve as a theoretical tool to better understand what recommendations are useful for users as well as understanding what of the documented perverse effects of RS come from recommendation per-se and what effects come from inherent uncertainty in user decision-making.
\section{The Value of Understanding User Beliefs in RS}

In this section we illustrate the differences between the ex-post and ex-ante view and show how it is useful for thinking about the design of RS. Our first point is trivial to state - since users do not know the true consumption values of the items, without any further information they may make sub-optimal consumption decisions. Consider the following stylized example. The choice set of the individual is $\Omega= \{x_1, x_2, x_3 \}$ and the ex-post utility values are given by
$u(x_1) = 1$, $u(x_2) = 2$, $u(x_3) = 3$. 
\par
However, when users are making decisions they have unobservable \textit{beliefs} about the ex-post utility value of the items and make decisions given these beliefs. If users had more information then their beliefs would converge to the ex-post utility values but, generally, users do not have enough information and so have beliefs over the space of possible ex-post utility values. In this example we suppose these beliefs are simple so that users view each good as simple lotteries: for $n=1,2,3$, $u(x_n) = n + \varepsilon_n$, $\varepsilon_1 \sim \mathcal N (2, 1)$, $\varepsilon_2 \sim \mathcal N (0, 1)$, $\varepsilon_3 \sim \mathcal N (-5, 1)$.
\par
In this example, after uncertainty is unresolved, the user ought to consume $x_3$, but an expected utility maximizing consumer (without further information) would consume $x_1$.\footnote{Note that here we ignore an important component of decision-making under uncertainty from decision theory, risk preferences, and suppose users are risk-neutral. A \textit{risk-averse} user may choose to consume a good with lower expected payoff but lower ex-ante variance compared to another good. While risk-aversion is an important component of decision-making and important in the context of RS when considering choices under uncertainty, it is not our main focus here.}
\par
This observation, while seemingly trivial, has important implications. The fact that the user would choose $C$ over $A$ reveals that $C$ prefers $A$ and allows us to get some information about a user's beliefs. Why are user beliefs important objects for designers of RS to understand and utilize?
\par
The first is that they are useful for understanding how users interact and get value from RS. In particular, users use RS to get \textit{information} about the set of items and update their beliefs about the value of the items. For instance, in the example, a RS could recommend to a user $x_3$ over $x_2$ and $x_1$ and that user will use this to update her beliefs about $x_3$. It is an empirical question that we leave for future work precisely how and to what extent users update their beliefs from recommendation and what factors of RS deisgn influence this.\footnote{There has been some work in this direction in understanding when recommendations are persuasive \cite{cremonesi2012investigating, gretzel2006persuasion} as well as the effect of their delivery on effectiveness \cite{murphy2014recommendation}.}
\par
Better understanding user beliefs is important for improving the performance of RS that optimize for metrics that move away from accuracy such as serendipity \cite{kotkov2016survey}. Serendipity-based RS attempt to provide recommendations that ``have the quality of being both unexpected and useful" \cite{maksai2015predicting} but it is hard to know what items are both unexpected and useful \cite{kotkov2016survey}. Understanding \textit{why} a user has not consumed a good depends on the beliefs that the user has and this is important for designing serendipitious recommendations. For instance, in the example, the user does not consume $x_3$ because she believes that the good has the potential to be catastrophically bad but, given more information, she would prefer to consume it over $x_1$ and $x_2$. It is possible that the ex-post and ex-ante utility of $x_3$ could be at best as good as the other two. The user may have an alternative with a known value -- say going for a walk around the neighborhood -- which she could always prefer to watching a given movie that has uncertain value but that the consumer thinks is always lower than the value of the walk.
\footnote{An additional reason this could be useful is if one considers a risk-averse individual, then it may be that a user has not consumed an item because she has a high perceived variance for the item but high expected quality.}
As a result, understanding the beliefs of users helps us understand why users have not consumed certain items and, in particular, which items a user may find useful given more information about it. This poses an important problem: designing systems that elicit not only ex-post valuations but also ex-ante beliefs about valuations, as only by knowing these can RS be effective in steering users' choices.
\par
Understanding the choices that individuals will make as well as what they will end up liking are complementary problems but require different viewpoints. \cite{jiang2014choice, saavedra2016choice} argue that choice-based approaches alone can help in designing better RS, though they argue that for these approaches because they do a better job at providing more accurate recommendations. We argue that the two problems should be considered separately -- though they may interact in interesting and useful ways.\footnote{In particular, some of the results from \cite{jiang2014choice, saavedra2016choice} may be driven by the fact that user choice embeds ratings information that may not have been observed by the recommender but is observed by the user. For instance, on Netflix I may have seen every episode of Friends when it was on TV which influences the consumption choices that I make. Alternatively, my friends can be a source of information that affect beliefs and choices. Netflix cannot observe neither of these data.} 
\par
RS are designed to give users \textit{information} that is useful to help them make decisions. On the one hand, predicting what are a user's ex-post preferences is useful to know what items the user would like to consume. On the other hand, predicting what items a user will actually consume (without recommendation) can help reason about what information may actually be useful to give her. Furthermore, in designing RS to avoid filter bubble and homogenization effects, having accurate choice predictions allows mitigating such adverse effects to become a first-order component of design. In particular, by predicting both choice and ratings, RS can provide information to users that leads them to useful goods but prevents them from falling into a filter bubble. 
\par
Solving this choice prediction problem requires data on the choices that individuals make and the sets that they choose them from as opposed to the traditional ratings or behavioral data that is collected. In the example given previously the fact that the consumer chooses $x_1$ from the set $\{x_1, x_2, x_3 \}$ would be the data that is useful for the choice prediction problem.

\section{Consumer Learning and Filter Bubbles}\label{sec:consumer_learning}
In this section we further show that differentiating between the ex-ante and ex-post view allows us to not only design better RS but also better understand their consequences. To do so, we study a simple model where users have beliefs about how good each item is and can learn their valuation over the items by consuming them. However, this learning process has spill-overs to their beliefs about other similar items which influences their future consumption choices. For instance, in our model, if a user watches \textit{The Matrix} and finds out that it is a good movie then they will update their beliefs positively about a similar movie such as \textit{The Matrix Reloaded} (the sequel to \textit{The Matrix}) but not update them much for a very different movie such as \textit{La La Land}.

Viewing the goal of the RS as providing users with additional information they can utilize to update their beliefs, we first show that, without recommendation, users may get stuck in a ``filter bubble" whereby they consume increasingly similar content and that, in this case, welfare will be negatively correlated with diversity. We show how providing information to users via recommendation can lead users out of this filter bubble. However, if recommendation does not properly take into account user beliefs, recommendations can lead to increasing homogenization among users and lower welfare and consumption diversity. Finally we then show that when the recommender does not know the user's beliefs about her valuation, the user may choose not to follow the recommendation.
\par
\noindent \textbf{Setup}. Suppose that there is a group $I$ of individuals and that each individual $i \in I$ can choose from the same finite set of experience goods $N=\left\{0,1,...,|N|-1\right\}$. Each of these goods $n$ is located on the unit circle and is equidistant to its neighbors, where $x_n \in \mathbb R^2$ denotes good $n$'s type or position. For simplicity, we assume that individuals only derive pleasure from a good $n \in N$ the first time they consume it. We denote by $u_{in}$ individual $i$'s utility from consuming good $n$.  We define a distance between the goods by $d:N\times N \to [0,1]$, where $d(n,m)=\min\{n-m \mod N\,,\,m-n \mod N \}$.
\par
In particular, we consider that the utility derived from a given good can be decomposed in the following manner: $u_{in}=\beta v_{in} + (1-\beta) v_n$, where $v_{in}$ denotes an idiosyncratic component -- i.e. consumer $i$'s idiosyncratic taste for good $n$ --  and $v_{n}$, a common-value component. 
One can interpret $v_n$ as a measure of how much good $n$ is valued in society in general. In a sense, $v_{in}$ denotes how $i$ diverges from this overall ranking. 
The scalar $\beta \in [0,1]$ denotes the degree to which utilities are idiosyncratic to each individual or common across individuals. If $\beta=1$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta=0$, every individual has the same preferences.
Stacking utilities in vector-form, we get ${\left(u_{in}\right)}_{n \in N}=:U_i=\beta V_i+(1 - \beta) V $, the vector of utilities associated with each good, where $V_i ={\left(v_{in}\right)}_{n \in N}$ and $V={\left(v_{n}\right)}_{n \in N}$.
\par
We assume that user $i$ starts with some beliefs about $U_i$, namely that the idiosyncratic and common-value parts of the utilities are independent -- $V_i \perp \!\!\! \perp V$ -- and that each is multivariate normal $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ and $V \sim \mathcal N(\overline V, \Sigma)$, with $\overline V =0$.
Furthermore, we assume that learning about the utility of good $n$ reveals more about the utility associated to goods that are closer to it, i.e. $n\pm1$ (mod $N$), than about those farther away. This captures the idea that trying a product provides more information about similar products than about dissimilar ones.
To this effect, we consider that the entry of $n$-th row and the ($m$)-th column of $\Sigma_i$ is given by $\sigma_i^2 \rho^{d(n,m)}$, and that of $\Sigma$ is given by $\sigma^2 \rho^{d(n,m)}$. The scalar $\rho \in [0,1]$ the covariance structure: a higher $\rho$ implies that learning the utility of $n$ is more informative about products nearby. Informativeness, for any $\rho \in (0,1)$, is decreasing in distance.
\par
Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{in}\right)_{n \in N}$ are drawn independently from a jointly normal distribution, where $\overline v_{in} \sim \mathcal N (0, \overline \sigma^2)$ are i.i.d. These $\overline v_{in}$ denote the prior belief that $i$ holds about the her valuation over good $n$. As people are exposed to different backgrounds, their beliefs about what is good for them also varies and $\overline v_{in}$ denotes this idiosyncrasy at the level of prior beliefs. 
\par
We assume the user makes $T$ choices and therefore can only consume up to $T$ goods, where $T$ is but a small fraction of $N$. This captures the idea that users are faced with an immense choice set but that ultimately they end up experiencing (and learning) about just a small fraction of it. Each period $t=1,...,T$, the consumer chooses the best product that she has not yet tried ($n_i^t$) given the information that past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and that that the RS offers ($R$), i.e. $n_i^t \in \arg \max_{ n \in N \backslash C_i^{t-1}} \mathbb E \left[u_{in} \mid C_i^{t-1}, R\right]$, with ties broken uniformly at random. 
\par
We are going to analyze three outcomes: how diverse consumer choices are, how similar are the choices that different individuals make and the resulting welfare. The first will be measured by $D_i =\frac{1}{T (T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$, the average pairwise distance between the consumed products, a natural measure of diversity in this environment [CITATION]. The second will be measured by a consumer homogeneity index $H:=1-\frac{1}{|I|(|I|-1)}\sum_{i,j \in I: i \ne j}d_J(C_i^T,C_j^T) $, where $d_J$ denotes the Jaccard distance on the space of subsets of $(N,d)$ and $H \in [0,1]$ [CITATION]. We will simulate a population of users and compare their consumption choices under different recommendation system regimes, where a higher $H$ denotes a higher user homogeneity. Consumer $i$'s welfare will just be the average of realized utilities, to control for the effect of $T$, $W_i=\frac{1}{T}\sum_{n \in C_i^T} u_{in}$.
\par
Finally, we are going to study three different cases: (i) no recommendation, (ii) omniscient recommender and (iii) partially informed recommender observes utilities accrued, but does not know consumer $i$'s starting beliefs $\overline U_i$. In the first case, users choose their items myopically, that is, using a greedy algorithm, and pick the item that has maximum expected utility. Upon consuming that item and learning its value, the users update their beliefs about their valuation of other goods and iterate this process $T$ times. In the second case, the recommender knows exactly $V_i$ and $V$ and can therefore recommend the best item for each consumer. In the third case, the recommender knows $V$ but does not know neither $V_i$ nor, crucially, users' beliefs $\overline V_i$. The recommendation will be that the consumer chooses $r^{t} \in \arg \max_{n \in N \backslash C_i^T} u_n$, but we assume the recommender provides full information about $V$. For instance, the recommender could display the whole distribution of utilities reported by other users or even its average, which is a good proxy for the common value component.\footnote{
The best item that could recommended with such information if the item with highest common value component. However, in that case, recommending only a single item generates costly updating to the consumer and provides little guidance as to what she should indeed pick as the common value might be of little importance when compared to the idiosyncratic component. Therefore, we assume that the RS reports the whole $V$, which results in higher expected welfare in choices and is not costly to implement.
} Note that it is not necessarily optimal for the user to follow the recommendation, that is, to pick the item with the highest common-value component $v_n$. Consumer's beliefs about her valuation of each item become crucial in this case: knowing $V$ may change the original ranking, but given this new information the consumer may find it best to pick an item other than the one recommended.
\par
\noindent \textbf{Simulation Details}. We study our model via numerical simulation, varying $\sigma, \rho, \beta$ and for $N = 1000, T = 25$. We report results for 50 populations and 50 individuals and track consumer welfare, the effect of recommendation, diversity, and homogeneity.\footnote{Due to space constraints we omit simulations for different values of $T$ but, unless otherwise noted, our results are robust to changes to $T$ as long as $T$ is not too large or too small.}
\par
\noindent \textbf{Results}. Table \ref{table:agg_results} shows aggregate results for welfare, diversity, and homogeneity. The omniscient recommendation case leads to the optimal consumption path for a consumer since she consumes the $T$ items with the highest utility. Compared to this case, both partial recommendation and no recommendation not only lead to lower welfare but also have lower diversity. In fact, the no recommendation case leads to the lowest levels of consumption diversity since users do not explore the space of products sufficiently and do excessive amounts of local search [can I look at how many jumps of less than 10 there were]. The partial recommendation case leads to higher consumption diversity and higher welfare, but has a significantly higher level of homogeneity.

\begin{table}[ht]
\centering
\begin{tabular}{rlll}
  \hline
Rec Policy & Welfare & Diversity & Homogeneity \\ 
  \hline
No Rec & 0.42 $\pm$ 0.002 & 191.04 $\pm$ 0.27 & 0.001 $\pm$ 0 \\ 
Partial Rec & 1.25 $\pm$ 0.005 & 221.83 $\pm$ 0.20 & 0.174 $\pm$ 0.004 \\ 
Omniscient Rec & 2.24 $\pm$ 0.006 & 237.59 $\pm$ 0.12 & 0.035 $\pm$ 0.002 \\ 
   \hline
\end{tabular}
\caption{Mean Welfare, Diversity, and Homogeneity for $N = 1000, T = 25, I = 50, S = 50$. Reported intervals are 95\% confidence intervals.}
\label{table:agg_results}
\end{table}

In the no recommendation case, users are not homogeneous since they primarily do local search based only on their priors in different areas of the product space. In the partial recommendation case, the information on the common-value component leads users to have more information on where to explore in the product space. However, the fact that this information is about the common-value component leads them to explore the same sections of the product space and leads to substantially higher homogeneity among users.

In our model, filter bubble effects occur simply because of the correlation structure between the utility and beliefs of products. Figure \ref{fig:local_consumption} shows how ``local" consumption choices are by looking at what fraction of consecutive consumption choices are close to each other with the definition of close being less than distance of 10. It highlights the fact that the correlation structure induces users to consume increasingly narrow content when they get no guidance from recommendation whereas the ``optimal" consumption path given by the omniscient policy does not exhibit this. The results of our model are consistent with the empirical results found in \cite{nguyen2014exploring} who show that filter bubble effects can arise even among users who do not follow recommendations.
\begin{figure}
\includegraphics[scale=0.08]{figures/local_moves_25}
\caption{Fraction of users who engage in ``local" consumption - moving less than 10 units}
\label{fig:local_consumption}
\end{figure}

\begin{figure}
\includegraphics[scale=0.05]{figures/beta_homogeneity_t_25}
\includegraphics[scale=0.05]{figures/beta_welfare_t_25}
\caption{Welfare and Homogeneity varying $\beta$}
\label{fig:beta_vary}
\end{figure}
\par
Furthermore, \cite{nguyen2014exploring} also find that recommendation can lead to an increase in consumption diversity. As Table \ref{table:agg_results} shows, our model is consistent with this effect but it depends on the nature of recommendation. If recommendation only provides information on the common-value component then welfare and consumption diversity increase but, correspondingly, homogenization increases dramatically. If recommendation takes into account user beliefs on the idiosyncratic  component of utility then welfare and consumption diversity can further increase and homogeneity will decrease.
\par
Low consumption diversity by itself is not necessarily bad. Figure \ref{fig:diversity_welfare_no_rec} shows that in the no recommendation case welfare and consumption diversity are negatively correlated. This is primarily because content diversity in the no recommendation case can arise from the fact that the user consumes a bad product and is forced to jump around the product space looking for ``good " products. However, if a user finds a high utility product right away then staying in this neighborhood may yield higher utility due to the correlation of utilities. The problem is that, even when the user finds a ``good" product, this may only be ``locally" good and the user will then excessively consume goods around it.
\par
As users learn more about the idiosyncratic component of their own preferences (i.e. as $t$ increases), the value of recommendation decreases. Figure \ref{fig:rec_obey} shows that recommendation compliance decreases as $t$ increases. Additionally, Figure \ref{fig:fig:beta_vary} shows how homogeneity and welfare change as $\beta$ increases  Most interestingly, when $\beta$ is high recommendation can be harmful to users - user homogeneity remains similar as $\beta$ increases but welfare drops to be even lower than under no recommendation.

\begin{figure}
   \begin{minipage}{0.48\textwidth}
     \centering
     \includegraphics[width=.5\linewidth]{figures/rec_obedience_25}
     \caption{Recommendation Effectiveness}\label{fig:rec_obey}
   \end{minipage}\hfill
   \begin{minipage}{0.48\textwidth}
     \centering
     \includegraphics[width=.5\linewidth]{"figures/Diversity vs Welfare - No Recommendation"}
     \caption{Diversity vs Welfare under No Recommendation}\label{fig:diversity_welfare_no_rec}
   \end{minipage}
\end{figure}

\section{Conclusion}

We have argued that incorporating user choice under uncertainty should be a first-order component of RS design. By collecting appropriate data about user choices and user beliefs, RS can be built to better understand what choices users are likely to make and thus what information would be useful to give them as opposed to simply predicting what items a user will like. This approach can not only aid in designing more useful RS, but can also be utilized to better understand recently documented adverse effects of RS such as filter bubbles and homogenization. We presented a model of user choice under uncertainty with correlated utilities that naturally generates filter bubble effects without any recommendation. We showed how, by not properly incorporating user beliefs into recommendation, RS can help alleviate filter bubble effects and increase utility but also substantially increase homogenization amongst users. Incorporating user beliefs can help increase utility even further but decrease homogenization effects. We leave for future work approaches that incorporate the insights from our work into traditional recommender system design methods.

\bibliographystyle{plain}
\bibliography{refs}
\end{document}

### 2019-04-20 create_graphs.R

@@ -57,7 +57,7 @@ graph_stats_diversity <- function(df, var, use_hrbrthemes){
   N <- df$N[1]
   
   g <- ggplot(df, aes_string(x=var, y="diversity_mean")) +
-      geom_line(aes(colour=regime)) +
+      geom_line(aes(colour=formatted_regime)) +
       geom_errorbar(aes(ymin=lower_ci_diversity, ymax=upper_ci_diveristy), width=.02,
                     position=position_dodge(.9)) + 
       labs(x=var, y="diversity",
@@ -89,7 +89,7 @@ graph_stats_welfare <- function(d, var, use_hrbrthemes){
   T_val <- d$T[1]
   N <- d$N[1]
   g <- ggplot(d, aes_string(x=var, y="welfare_mean")) +
-    geom_line(aes(colour=regime)) +
+    geom_line(aes(colour=formatted_regime)) +
     geom_errorbar(aes(ymin=lower_ci_welfare, ymax=upper_ci_welfare), width=.02, position=position_dodge(.9)) + 
     labs(x=var, y="welfare",
          title=paste("N =", N, "T =", T_val, "Welfare",sep=" ")) 
@@ -133,31 +133,15 @@ get_stats_welfare <- function(rec_data, var){
 ## HOMOGENEITY DATA
 
 get_stats_homogeneity <- function(stats_df, var){
-  if (var == "rho") { # there must be a better way
-    df <- stats_df %>%
-      group_by(regime, rho) %>% 
-      mutate(jaccard_mean = mean(jaccard),
-             jaccard_sd = sd(jaccard),
-             jaccard_n = n()) %>%
-      mutate(jaccard_se =  jaccard_sd/ sqrt(jaccard_n),
-             lower_ci_jaccard = jaccard_mean - qt(1 - (0.05 / 2), jaccard_n - 1) * jaccard_se,
-             upper_ci_jaccard = jaccard_mean + qt(1 - (0.05 / 2), jaccard_n - 1) * jaccard_se) %>% 
-      slice(1)
-    return(df)
+  df <- stats_df
+  if (var == "rho") {
+    df <- df %>% group_by(regime, rho) 
   } else if (var == "beta") {
-    df <- stats_df %>%
-      group_by(regime, beta) %>% 
-      mutate(jaccard_mean = mean(jaccard),
-             jaccard_sd = sd(jaccard),
-             jaccard_n = n()) %>%
[…]

### 2019-04-20 create_graphs.R
have two graphs in one ggplot

@@ -3,6 +3,8 @@ library(dplyr)
 library(hrbrthemes)
 library(tidyverse)
 
+library(gridExtra)
+
 
 get_stats_diversity <- function(rec_data, var){
   df <- rec_data
@@ -92,7 +94,7 @@ graph_stats_welfare <- function(d, var, use_hrbrthemes){
     geom_line(aes(colour=formatted_regime)) +
     geom_errorbar(aes(ymin=lower_ci_welfare, ymax=upper_ci_welfare), width=.02, position=position_dodge(.9)) + 
     labs(x=var, y="welfare",
-         title=paste("N =", N, "T =", T_val, "Welfare",sep=" ")) 
+         title=paste( "Welfare",sep=" ")) 
   if(use_hrbrthemes){
     g <- g + theme_ipsum_rc()
   }
@@ -171,7 +173,7 @@ graph_stats_homo <- function(df, var){
       geom_errorbar(aes(ymin=lower_ci_jaccard, ymax=upper_ci_jaccard), width=.02,
                     position=position_dodge(.9)) + 
       labs(x="beta", y="homogeneity",
-           title=paste("N =", N, "T =", T_val, "Homogeneity",sep=" ")) + theme_ipsum_rc()
+           title=paste("Homogeneity",sep=" ")) + theme_ipsum_rc()
     return(g)
   } else if (var == "sigma") {
     g <- ggplot(df, aes(x=sigma, y=jaccard_mean)) +
@@ -184,8 +186,8 @@ graph_stats_homo <- function(df, var){
   }
 }
 
-WORKING_DIR <- "/Users/guyaridor/Desktop/recommender_systems/rec_sys_conf_paper/"
-#WORKING_DIR <- "/Users/ssikdar/Downloads/recommender_systems/rec_sys_conf_paper"
+#WORKING_DIR <- "/Users/guyaridor/Desktop/recommender_systems/rec_sys_conf_paper/"
+WORKING_DIR <- "/Users/ssikdar/Downloads/econ/bubbles/"
 
 rec_data <- read.csv(paste(WORKING_DIR, "rec_data_t_25.csv", sep=""))
 rec_data <- rec_data %>% mutate(formatted_regime = ifelse(regime == "rec", "Omniscient", ifelse(regime == "no_rec", "No Rec", "Partial")))
@@ -321,6 +323,34 @@ if(use_hrbrthemes){
 ggsave(
[…]

### 2019-04-20 create_graphs.R
fix hardcoded path

@@ -295,7 +295,7 @@ df <- merge(df, homo, by="regime")
 library(xtable)
 xtable(df)
 
-time_dat <- read.csv("/Users/guyaridor/Desktop/recommender_systems/rec_sys_conf_paper/time_path_25.csv")
+time_dat <- read.csv(paste(WORKING_DIR, "time_path_25.csv", sep=""))
 time_dat <- time_dat %>% mutate(formatted_regime = ifelse(regime == "rec", "Omniscient", ifelse(regime == "no_rec", "No Rec", "Partial")))
 time_dat <- time_dat %>% mutate(local_move = as.numeric(consumption_dist < 10))
 

### 2019-04-20 paper.tex
updated apper

@@
 \usepackage{mathtools}
+\usepackage[font=small]{caption}
+\usepackage{subcaption}
+\usepackage{float}
 %
@@
 % The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
-\title{The Ex-Ante View of RS Design}
+\title{The Ex-Ante View of Recommender System Design}
 %
@@
 Recommender Systems (RS) have been widely deployed in online platforms such as Netflix, Amazon, and YouTube.
-On these platforms the combination of the fact that most goods are experience goods and that the number of choices available to users is incredibly large results in users not knowing their true preference ordering over all of the items.
-For instance, on Netflix there are thousands of movies, on Amazon there are millions of products, and on YouTube there are billions of videos.
+On these platforms the combination of the fact that most items are experience goods (i.e.
+valuation is only learned after consumption) and that the number of choices available to users is incredibly large, results in users not knowing their true preference ordering over all of the items.
+For instance, on Netflix there are thousands of movies, on Amazon millions of products, and YouTube billions of videos.
 As a result, RS have been useful in helping users make decisions on these platforms because they provide users with information to help them make their choices under uncertainty.
@@
 \par
-In this paper we want to put forward the argument that one approach to overcoming these problems is to make understanding how users choose which goods to consume a first-order consideration when designing RS.
+In this paper we want to put forward the argument that one approach to overcoming these problems is to make understanding how users choose which items to consume a first-order consideration when designing RS.
 In particular, we differentiate between the previous \textit{ex-post} interpretation of the RS problem and what we define as the \textit{ex-ante} viewpoint.
@@
 This shift from thinking about true preferences after all uncertainty is resolved (the ex-post view) to user decisions under uncertainty (the ex-ante view) can serve as a theoretical tool to better understand what recommendations are useful for users as well as understanding what of the documented perverse effects of RS come from recommendation per-se and what effects come from inherent uncertainty in user decision-making.
-\section{The Value of Understanding User Beliefs in RS}
+\section{The Value of the Ex-Ante View}
 
@@
 This observation, while seemingly trivial, has important implications.
-The fact that the user would choose $C$ over $A$ reveals that $C$ prefers $A$ and allows us to get some information about a user's beliefs.
+The fact that the user would choose $C$ over $A$ reveals that $C$ prefers $A$ given her current beliefs\footnote{This is a statement of \textit{revealed preference} which has been studied heavily in economic decision theory (see e.g.
+\cite[]{mas1995microeconomic}).} and allows us to get some information about a user's beliefs.
 Why are user beliefs important objects for designers of RS to understand and utilize?
@@
 It is possible that the ex-post and ex-ante utility of $x_3$ could be at best as good as the other two.
-The user may have an alternative with a known value -- say going for a walk around the neighborhood -- which she could always prefer to watching a given movie that has uncertain value but that the consumer thinks is always lower than the value of the walk.
-\footnote{An additional reason this could be useful is if one considers a risk-averse individual, then it may be that a user has not consumed an item because she has a high perceived variance for the item but high expected quality.}
+The user may have an alternative with a known value -- say going for a walk around the neighborhood -- which she could always prefer to watching a given movie that has uncertain value but that the consumer thinks is always lower than the value of the walk.\footnote{An additional reason this could be useful is if one considers a risk-averse individual, then it may be that a user has not consumed an item because she has a high perceived variance for the item but high expected quality.}
 As a result, understanding the beliefs of users helps us understand why users have not consumed certain items and, in particular, which items a user may find useful given more information about it.
-This poses an important problem: designing systems that elicit not only ex-post valuations but also ex-ante beliefs about valuations, as only by knowing these can RS be effective in steering users' choices.
+This poses an important problem: designing systems that elicit not only ex-post valuations but also ex-ante beliefs about valuations, as only by knowing these can RS be effective in steering users' choices.\footnote{\cite{schotter2014belief} provides a survey of belief elicitation methods that are utilized in economics.}
 \par
 Understanding the choices that individuals will make as well as what they will end up liking are complementary problems but require different viewpoints.
-\cite{jiang2014choice, saavedra2016choice} argue that choice-based approaches alone can help in designing better RS, though they argue that for these approaches because they do a better job at providing more accurate recommendations.
+\cite{jiang2014choice, saavedra2016choice} argue that choice-based approaches alone can help in designing better RS, though they argue for these approaches because they do a better job at providing more accurate recommendations.
 We argue that the two problems should be considered separately -- though they may interact in interesting and useful ways.\footnote{In particular, some of the results from \cite{jiang2014choice, saavedra2016choice} may be driven by the fact that user choice embeds ratings information that may not have been observed by the recommender but is observed by the user.
@@
 RS are designed to give users \textit{information} that is useful to help them make decisions.
-On the one hand, predicting what are a user's ex-post preferences is useful to know what items the user would like to consume.
+On the one hand, predicting a user's ex-post preferences is useful to know what items the user would like to consume.
 On the other hand, predicting what items a user will actually consume (without recommendation) can help reason about what information may actually be useful to give her.
 Furthermore, in designing RS to avoid filter bubble and homogenization effects, having accurate choice predictions allows mitigating such adverse effects to become a first-order component of design.
-In particular, by predicting both choice and ratings, RS can provide information to users that leads them to useful goods but prevents them from falling into a filter bubble.
+In particular, by predicting both choice and ratings, RS can provide information to users that leads them to useful items but prevents them from falling into a filter bubble.
 \par
-Solving this choice prediction problem requires data on the choices that individuals make and the sets that they choose them from as opposed to the traditional ratings or behavioral data that is collected.
+Solving this choice prediction problem requires data on the choices that individuals make and the sets that they choose them from in addition to the traditional ratings or behavioral data that is collected.
 In the example given previously the fact that the consumer chooses $x_1$ from the set $\{x_1, x_2, x_3 \}$ would be the data that is useful for the choice prediction problem.
-
-\section{Consumer Learning and Filter Bubbles}\label{sec:consumer_learning}
+\section{Filter Bubbles and Homogenization}\label{sec:consumer_learning}
 In this section we further show that differentiating between the ex-ante and ex-post view allows us to not only design better RS but also better understand their consequences.
 To do so, we study a simple model where users have beliefs about how good each item is and can learn their valuation over the items by consuming them.
-However, this learning process has spill-overs to their beliefs about other similar items which influences their future consumption choices.
+However, drawing from the main idea behind content-based RS, similar items have similar valuations and consumers' beliefs about neighboring items update after consuming an item which influences future consumption decisions.
 For instance, in our model, if a user watches \textit{The Matrix} and finds out that it is a good movie then they will update their beliefs positively about a similar movie such as \textit{The Matrix Reloaded} (the sequel to \textit{The Matrix}) but not update them much for a very different movie such as \textit{La La Land}.
@@
 However, if recommendation does not properly take into account user beliefs, recommendations can lead to increasing homogenization among users and lower welfare and consumption diversity.
-Finally we then show that when the recommender does not know the user's beliefs about her valuation, the user may choose not to follow the recommendation.
+Finally we then show that when the recommender does not know the user's beliefs about her valuation, recommendation becomes decreasingly useful as the user learns more about her own preferences.
 \par
 \noindent \textbf{Setup}.
-Suppose that there is a group $I$ of individuals and that each individual $i \in I$ can choose from the same finite set of experience goods $N=\left\{0,1,...,|N|-1\right\}$.
-Each of these goods $n$ is located on the unit circle and is equidistant to its neighbors, where $x_n \in \mathbb R^2$ denotes good $n$'s type or position.
-For simplicity, we assume that individuals only derive pleasure from a good $n \in N$ the first time they consume it.
-We denote by $u_{in}$ individual $i$'s utility from consuming good $n$.
-We define a distance between the goods by $d:N\times N \to [0,1]$, where $d(n,m)=\min\{n-m \mod N\,,\,m-n \mod N \}$.
+Suppose that there is a group $I$ of individuals and that each individual $i \in I$ can choose from the same finite set of $N$ items $\mathcal{I} = \left\{0,1,...,N-1\right\}$.
+For simplicity, we assume that individuals only derive pleasure from a item $n \in \mathcal{I}$ the first time they consume it.
+We denote by $u_{in}$ individual $i$'s realized utility from consuming item $n$.
+We define the distance between two items as $d(n,m)=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}$ where $m$ and $n$ are the index of the items in $\mathcal{I}$ so that, conceptually, the items are located in a ring of size $N$.
 \par
-In particular, we consider that the utility derived from a given good can be decomposed in the following manner: $u_{in}=\beta v_{in} + (1-\beta) v_n$, where $v_{in}$ denotes an idiosyncratic component -- i.e.
+In particular, we consider that the utility derived from a given item can be decomposed in the following manner: $u_{in}=\beta v_{in} + (1-\beta) v_n$, where $v_{in}$ denotes an idiosyncratic component -- i.e.
 consumer $i$'s idiosyncratic taste for good $n$ --  and $v_{n}$, a common-value component.
-One can interpret $v_n$ as a measure of how much good $n$ is valued in society in general. In a sense, $v_{in}$ denotes how $i$ diverges from this overall ranking. 
+One can interpret $v_n$ as a measure of how much good $n$ is valued in society in general and, in a sense, $v_{in}$ denotes how $i$ diverges from this overall ranking. 
 The scalar $\beta \in [0,1]$ denotes the degree to which utilities are idiosyncratic to each individual or common across individuals.
 If $\beta=1$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta=0$, every individual has the same preferences.
-Stacking utilities in vector-form, we get ${\left(u_{in}\right)}_{n \in N}=:U_i=\beta V_i+(1 - \beta) V $, the vector of utilities associated with each good, where $V_i ={\left(v_{in}\right)}_{n \in N}$ and $V={\left(v_{n}\right)}_{n \in N}$.
+Stacking utilities in vector-form, we get ${\left(u_{in}\right)}_{n \in \mathcal{I}}=:U_i=\beta V_i+(1 - \beta) V $, the vector of utilities associated with each good, where $V_i ={\left(v_{in}\right)}_{n \in \mathcal{I}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{I}}$.
 \par
@@
 \perp V$ -- and that each is multivariate normal $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ and $V \sim \mathcal N(\overline V, \Sigma)$, with $\overline V =0$.
-Furthermore, we assume that learning about the utility of good $n$ reveals more about the utility associated to goods that are closer to it, i.e.
+Furthermore, we assume that learning about the utility of good $n$ reveals more about the utility associated to items that are closer to it, i.e.
 $n\pm1$ (mod $N$), than about those farther away.
@@
 \par
-Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{in}\right)_{n \in N}$ are drawn independently from a jointly normal distribution, where $\overline v_{in} \sim \mathcal N (0, \overline \sigma^2)$ are i.i.d.
+Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{in}\right)_{n \in \mathcal{I}}$ are drawn independently from a jointly normal distribution, where $\overline v_{in} \sim \mathcal N (0, \overline \sigma^2)$ are i.i.d.
 These $\overline v_{in}$ denote the prior belief that $i$ holds about the her valuation over good $n$.
@@
 \par
-We assume the user makes $T$ choices and therefore can only consume up to $T$ goods, where $T$ is but a small fraction of $N$.
+We assume the user makes $T$ choices and therefore can only consume up to $T$ items, where $T$ is but a small fraction of $N$.
 This captures the idea that users are faced with an immense choice set but that ultimately they end up experiencing (and learning) about just a small fraction of it.
 Each period $t=1,...,T$, the consumer chooses the best product that she has not yet tried ($n_i^t$) given the information that past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and that that the RS offers ($R$), i.e.
-$n_i^t \in \arg \max_{ n \in N \backslash C_i^{t-1}} \mathbb E \left[u_{in} \mid C_i^{t-1}, R\right]$, with ties broken uniformly at random.
+$n_i^t \in \arg \max_{ n \in \mathcal{I} \backslash C_i^{t-1}} \mathbb E \left[u_{in} \mid C_i^{t-1}, R\right]$, with ties broken uniformly at random.
 \par
 We are going to analyze three outcomes: how diverse consumer choices are, how similar are the choices that different individuals make and the resulting welfare.
-The first will be measured by $D_i =\frac{1}{T (T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$, the average pairwise distance between the consumed products, a natural measure of diversity in this environment [CITATION].
-The second will be measured by a consumer homogeneity index $H:=1-\frac{1}{|I|(|I|-1)}\sum_{i,j \in I: i \ne j}d_J(C_i^T,C_j^T) $, where $d_J$ denotes the Jaccard distance on the space of subsets of $(N,d)$ and $H \in [0,1]$ [CITATION].
-We will simulate a population of users and compare their consumption choices under different recommendation system regimes, where a higher $H$ denotes a higher user homogeneity.
+The first will be measured by $D_i =\frac{1}{T (T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$, the average pairwise distance between the consumed products, a natural measure of diversity in this environment that is utilized in the literature \cite{ziegler2005improving}.
+The second will be measured by a consumer homogeneity index $H:=1-\frac{1}{|I|(|I|-1)}\sum_{i,j \in I: i \ne j}d_J(C_i^T,C_j^T) $, where $d_J$ denotes the Jaccard distance and $H \in [0,1]$ which is similar to the approach to measuring homogeneity utilized in \cite{chaney2018algorithmic}.
+Note that a higher $H$ indicates more homogeneity.
 Consumer $i$'s welfare will just be the average of realized utilities, to control for the effect of $T$, $W_i=\frac{1}{T}\sum_{n \in C_i^T} u_{in}$.
 \par
-Finally, we are going to study three different cases: (i) no recommendation, (ii) omniscient recommender and (iii) partially informed recommender observes utilities accrued, but does not know consumer $i$'s starting beliefs $\overline U_i$.
-In the first case, users choose their items myopically, that is, using a greedy algorithm, and pick the item that has maximum expected utility.
-Upon consuming that item and learning its value, the users update their beliefs about their valuation of other goods and iterate this process $T$ times.
+Finally, we are going to study three different cases: (i) no recommendation, (ii) omniscient recommender and (iii) partially informed recommender where the recommender observes utilities accrued, but does not know consumer $i$'s starting beliefs $\overline U_i$.
+In the first case, users choose their items myopically, that is, using a greedy algorithm, and pick the item that has maximum expected utility given beliefs at time $t$.
+Upon consuming that item and learning its value, the users update their beliefs about their valuation of other items and iterate this process $T$ times.
 In the second case, the recommender knows exactly $V_i$ and $V$ and can therefore recommend the best item for each consumer.
 In the third case, the recommender knows $V$ but does not know neither $V_i$ nor, crucially, users' beliefs $\overline V_i$.
-The recommendation will be that the consumer chooses $r^{t} \in \arg \max_{n \in N \backslash C_i^T} u_n$, but we assume the recommender provides full information about $V$.
+The recommendation will be that the consumer chooses $r^{t} \in \arg \max_{n \in \mathcal{I} \backslash C_i^T} u_n$, but we assume the recommender provides full information about $V$.
 For instance, the recommender could display the whole distribution of utilities reported by other users or even its average, which is a good proxy for the common value component.\footnote{
@@
 \noindent \textbf{Simulation Details}.
-We study our model via numerical simulation, varying $\sigma, \rho, \beta$ and for $N = 1000, T = 25$.
-We report results for 50 populations and 50 individuals and track consumer welfare, the effect of recommendation, diversity, and homogeneity.\footnote{Due to space constraints we omit simulations for different values of $T$ but, unless otherwise noted, our results are robust to changes to $T$ as long as $T$ is not too large or too small.}
+We will simulate $P$ populations of $I$ users and compare their consumption choices under different recommendation system regimes.
+We simulate over a grid of parameter values for $\sigma, \sigma_i, \rho, \rho_i, \beta$ and report results for $P=50, I = 50, N = 1000, T = 25$.\footnote{Due to space constraints we omit simulations for different values of $T$ but, unless otherwise noted, our results are robust to changes to $T$ as long as $T$ is not too large (close to $N$) or too small (close to $0$).}
 \par
-\noindent \textbf{Results}.
-Table \ref{table:agg_results} shows aggregate results for welfare, diversity, and homogeneity.
-The omniscient recommendation case leads to the optimal consumption path for a consumer since she consumes the $T$ items with the highest utility.
-Compared to this case, both partial recommendation and no recommendation not only lead to lower welfare but also have lower diversity.
-In fact, the no recommendation case leads to the lowest levels of consumption diversity since users do not explore the space of products sufficiently and do excessive amounts of local search [can I look at how many jumps of less than 10 there were].
-The partial recommendation case leads to higher consumption diversity and higher welfare, but has a significantly higher level of homogeneity.
 
-\begin{table}[ht]
+\begin{table}[bt]
 \centering
@@
 
-In the no recommendation case, users are not homogeneous since they primarily do local search based only on their priors in different areas of the product space.
-In the partial recommendation case, the information on the common-value component leads users to have more information on where to explore in the product space.
-However, the fact that this information is about the common-value component leads them to explore the same sections of the product space and leads to substantially higher homogeneity among users.
+\begin{figure*}[!htbp]
+    \centering
+    \begin{minipage}[b]{0.5\textwidth}
+        \centering
+        \includegraphics[width=\linewidth, height=1.75in]{figures/local_moves_25}
+        \caption{Fraction of Local Search}
+        \label{fig:local_consumption}
+    \end{minipage}%
+    ~ 
+    \begin{minipage}[b]{0.5\textwidth}
+        \centering
+        \includegraphics[width=\linewidth, height=1.75in]{figures/welfare_homo_combo}
+        \caption{Welfare and Homogeneity varying $\beta$}
+        \label{fig:beta_vary}
+    \end{minipage}
+    ~ \\
+     \begin{minipage}{0.48\textwidth}
+     \centering
+     \includegraphics[width=\linewidth, height=1.75in]{figures/rec_obedience_25}
+     \caption{Recommendation Effectiveness}\label{fig:rec_obey}
+   \end{minipage}\hfill
+   ~
+   \begin{minipage}{0.48\textwidth}
+     \centering
+     \includegraphics[width=\linewidth, height=1.75in]{"figures/Diversity vs Welfare - No Recommendation"}
+     \caption{ Diversity vs Welfare under No Recommendation}\label{fig:diversity_welfare_no_rec}
+   \end{minipage}
+\end{figure*}
+\small
+\noindent \textbf{Results}.
+Table \ref{table:agg_results} shows aggregate results for welfare, diversity, and homogeneity.
+The omniscient recommendation case leads to the optimal consumption path for a consumer since she consumes the $T$ items with the highest utility.
+Compared to this case, both partial recommendation and no recommendation not only lead to lower welfare but also have lower diversity.
+In fact, the no recommendation case leads to the lowest levels of consumption diversity since users do not explore the space of products sufficiently and do excessive amounts of local search as can be seen in Figure \ref{fig:local_consumption}.
+Figure \ref{fig:local_consumption} shows how ``local" consumption choices are by looking at what fraction of consecutive consumption choices are close to each other with the definition of close being less than distance of 10.
+The partial recommendation case leads to higher consumption diversity and higher welfare, but has a significantly higher level of homogeneity.
 
 In our model, filter bubble effects occur simply because of the correlation structure between the utility and beliefs of products.
-Figure \ref{fig:local_consumption} shows how ``local" consumption choices are by looking at what fraction of consecutive consumption choices are close to each other with the definition of close being less than distance of 10.
 It highlights the fact that the correlation structure induces users to consume increasingly narrow content when they get no guidance from recommendation whereas the ``optimal" consumption path given by the omniscient policy does not exhibit this.
-The results of our model are consistent with the empirical results found in \cite{nguyen2014exploring} who show that filter bubble effects can arise even among users who do not follow recommendations.
-\begin{figure}
-\includegraphics[scale=0.08]{figures/local_moves_25}
-\caption{Fraction of users who engage in ``local" consumption - moving less than 10 units}
-\label{fig:local_consumption}
-\end{figure}
-
-\begin{figure}
-\includegraphics[scale=0.05]{figures/beta_homogeneity_t_25}
-\includegraphics[scale=0.05]{figures/beta_welfare_t_25}
-\caption{Welfare and Homogeneity varying $\beta$}
-\label{fig:beta_vary}
-\end{figure}
+The results of our model are consistent with the empirical results found in \cite{nguyen2014exploring} who show that filter bubble effects can arise even amongst users who do not utilize recommendations.
 \par
@@
 If recommendation only provides information on the common-value component then welfare and consumption diversity increase but, correspondingly, homogenization increases dramatically.
+In this case the information on the common-value component leads users to have more information on where to explore in the product space.
+However, the fact that this information is about the common-value component and the same across all individuals leads them to explore the same sections of the product space and leads to substantially higher homogeneity among users.
 If recommendation takes into account user beliefs on the idiosyncratic  component of utility then welfare and consumption diversity can further increase and homogeneity will decrease.
@@
 Figure \ref{fig:diversity_welfare_no_rec} shows that in the no recommendation case welfare and consumption diversity are negatively correlated.
-This is primarily because content diversity in the no recommendation case can arise from the fact that the user consumes a bad product and is forced to jump around the product space looking for ``good " products.
+This is primarily because content diversity in the no recommendation case can arise from the fact that the user consumes a bad product and is forced to jump around the product space looking for ``good" products.
 However, if a user finds a high utility product right away then staying in this neighborhood may yield higher utility due to the correlation of utilities.
-The problem is that, even when the user finds a ``good" product, this may only be ``locally" good and the user will then excessively consume goods around it.
+The problem is that, even when the user finds a ``good" product, this may only be ``locally" good and the user will then excessively consume items around it.
 \par
-As users learn more about the idiosyncratic component of their own preferences (i.e.
-as $t$ increases), the value of recommendation decreases.
+Finally, the value of recommendation in the partial recommendation case decreases over time as users learn more about the idiosyncratic component of their own preferences (i.e.
+as $t$ increases).
 Figure \ref{fig:rec_obey} shows that recommendation compliance decreases as $t$ increases.
-Additionally, Figure \ref{fig:fig:beta_vary} shows how homogeneity and welfare change as $\beta$ increases  Most interestingly, when $\beta$ is high recommendation can be harmful to users - user homogeneity remains similar as $\beta$ increases but welfare drops to be even lower than under no recommendation.
-
-\begin{figure}
-   \begin{minipage}{0.48\textwidth}
-     \centering
-     \includegraphics[width=.5\linewidth]{figures/rec_obedience_25}
-     \caption{Recommendation Effectiveness}\label{fig:rec_obey}
-   \end{minipage}\hfill
-   \begin{minipage}{0.48\textwidth}
-     \centering
-     \includegraphics[width=.5\linewidth]{"figures/Diversity vs Welfare - No Recommendation"}
-     \caption{Diversity vs Welfare under No Recommendation}\label{fig:diversity_welfare_no_rec}
-   \end{minipage}
-\end{figure}
+Additionally, Figure \ref{fig:beta_vary} shows how homogeneity and welfare change as $\beta$ increases.
+Most interestingly, when $\beta$ is high so that the common-value component is small, recommendation can be harmful to users.
+The information from recommendation still leads to large user homogeneity even though the optimum would be for there to be almost none and leads users to have lower welfare than under no recommendation.
 
@@
 Incorporating user beliefs can help increase utility even further but decrease homogenization effects.
-We leave for future work approaches that incorporate the insights from our work into traditional recommender system design methods.
+We leave for future work approaches that incorporate our insights from into traditional recommender system design methods.
 

### 2019-04-20 create_graphs.R

@@ -186,8 +186,8 @@ graph_stats_homo <- function(df, var){
   }
 }
 
-#WORKING_DIR <- "/Users/guyaridor/Desktop/recommender_systems/rec_sys_conf_paper/"
-WORKING_DIR <- "/Users/ssikdar/Downloads/econ/bubbles/"
+WORKING_DIR <- "/Users/guyaridor/Desktop/recommender_systems/rec_sys_conf_paper/"
+#WORKING_DIR <- "/Users/ssikdar/Downloads/econ/bubbles/"
 
 rec_data <- read.csv(paste(WORKING_DIR, "rec_data_t_25.csv", sep=""))
 rec_data <- rec_data %>% mutate(formatted_regime = ifelse(regime == "rec", "Omniscient", ifelse(regime == "no_rec", "No Rec", "Partial")))
@@ -297,9 +297,10 @@ xtable(df)
 
 time_dat <- read.csv(paste(WORKING_DIR, "time_path_25.csv", sep=""))
 time_dat <- time_dat %>% mutate(formatted_regime = ifelse(regime == "rec", "Omniscient", ifelse(regime == "no_rec", "No Rec", "Partial")))
-time_dat <- time_dat %>% mutate(local_move = as.numeric(consumption_dist < 10))
+time_dat <- time_dat %>% mutate(local_move = as.numeric(consumption_dist < 5))
+t <- time_dat %>% filter(t > 0) # local move isn't defined at the first time step so drop it to properly have smoothed plots
 
-g <- ggplot(time_dat, aes(x=t, y=local_move)) +
+g <- ggplot(t, aes(x=t, y=local_move)) +
     geom_smooth(aes(colour=formatted_regime)) +
     labs(x="t", y="Fraction of Local Search"
     )

### 2019-04-20 paper.tex
some changes

--- 2019-04-20 paper.tex
@@
 \begin{abstract}
-Recommender systems (RS) are traditionally deployed in environments where users are uncertain about their preferences and thus face a problem of choice under uncertainty, but most popular design approaches ignore this fact.
-We argue that predicting and modeling consumer choice in these contexts can improve the usefulness of RS and reframe the RS problem as providing useful information to help reduce user uncertainty as opposed to simply predicting user preferences.
-We show how this insight can be utilized to design RS that mitigate the negative consequences of RS such as filter bubble and user homogenization effects as well as better understand the role that RS play in contributing to these phenomena.
+Recommender Systems (RS) are traditionally deployed in environments where users are uncertain about their preferences and thus face a problem of choice under uncertainty, but most popular design approaches ignore this fact.
+We argue that predicting and modeling consumer choice in these contexts can improve the usefulness of RS and reframe the problem as providing useful information to help reduce user uncertainty as opposed to simply predicting user preferences.
+We show how this insight can be utilized to design RS that mitigate the negative consequences such as filter bubble and user homogenization effects, as well as better understand the role that RS play in contributing to these phenomena.
 \end{abstract}
@@
 Recommender Systems (RS) have been widely deployed in online platforms such as Netflix, Amazon, and YouTube.
-On these platforms the combination of the fact that most goods are experience goods and that the number of choices available to users is incredibly large results in users not knowing their true preference ordering over all of the items.
-For instance, on Netflix there are thousands of movies, on Amazon there are millions of products, and on YouTube there are billions of videos.
+On these platforms the combination of most goods being experience goods and the number of choices available to users being incredibly large, results in users not knowing their true preference ordering over all of the items.
+For instance, on Netflix there are thousands of movies, Amazon millions of products, and YouTube billions of videos.
 As a result, RS have been useful in helping users make decisions on these platforms because they provide users with information to help them make their choices under uncertainty.
@@
 \par
-However, when users are making decisions they have unobservable \textit{beliefs} about the ex-post utility value of the items and make decisions given these beliefs.
+However when users are making decisions, they have unobservable \textit{beliefs} about the ex-post utility value of the items, and make decisions given these beliefs.
 If users had more information then their beliefs would converge to the ex-post utility values but, generally, users do not have enough information and so have beliefs over the space of possible ex-post utility values.
@@
 For instance, in the example, a RS could recommend to a user $x_3$ over $x_2$ and $x_1$ and that user will use this to update her beliefs about $x_3$.
-It is an empirical question that we leave for future work precisely how and to what extent users update their beliefs from recommendation and what factors of RS deisgn influence this.\footnote{There has been some work in this direction in understanding when recommendations are persuasive \cite{cremonesi2012investigating, gretzel2006persuasion} as well as the effect of their delivery on effectiveness \cite{murphy2014recommendation}.}
+It is an empirical question that we leave for future work, precisely how and to what extent users update their beliefs from recommendation and what factors of RS deisgn influence this.\footnote{There has been some work in this direction in understanding when recommendations are persuasive \cite{cremonesi2012investigating, gretzel2006persuasion} as well as the effect of their delivery on effectiveness \cite{murphy2014recommendation}.}
 \par
@@
 In this section we further show that differentiating between the ex-ante and ex-post view allows us to not only design better RS but also better understand their consequences.
-To do so, we study a simple model where users have beliefs about how good each item is and can learn their valuation over the items by consuming them.
+We study a simple model where users have beliefs about how good each item is and can learn their valuation over the items by consuming them.
 However, this learning process has spill-overs to their beliefs about other similar items which influences their future consumption choices.
@@
 \begin{figure}
-\includegraphics[scale=0.05]{figures/beta_homogeneity_t_25}
-\includegraphics[scale=0.05]{figures/beta_welfare_t_25}
+\includegraphics[scale=0.05]{figures/welfare_homo_combo}
 \caption{Welfare and Homogeneity varying $\beta$}

### 2019-04-20 paper.tex

--- 2019-04-20 paper.tex
@@
 Recommender Systems (RS) have been widely deployed in online platforms such as Netflix, Amazon, and YouTube.
-On these platforms the combination of the fact that most items are experience goods (i.e.
-valuation is only learned after consumption) and that the number of choices available to users is incredibly large, results in users not knowing their true preference ordering over all of the items.
-For instance, on Netflix there are thousands of movies, on Amazon millions of products, and YouTube billions of videos.
+On these platforms the combination of most goods being experience goods and the number of choices available to users being incredibly large, results in users not knowing their true preference ordering over all of the items.
+For instance, on Netflix there are thousands of movies, Amazon millions of products, and YouTube billions of videos.
 As a result, RS have been useful in helping users make decisions on these platforms because they provide users with information to help them make their choices under uncertainty.

### 2019-04-20 create_graphs.R
fix local graph

@@ -186,8 +186,8 @@ graph_stats_homo <- function(df, var){
   }
 }
 
-WORKING_DIR <- "/Users/guyaridor/Desktop/recommender_systems/rec_sys_conf_paper/"
-#WORKING_DIR <- "/Users/ssikdar/Downloads/econ/bubbles/"
+#WORKING_DIR <- "/Users/guyaridor/Desktop/recommender_systems/rec_sys_conf_paper/"
+WORKING_DIR <- "/Users/ssikdar/Downloads/econ/bubbles/"
 
 rec_data <- read.csv(paste(WORKING_DIR, "rec_data_t_25.csv", sep=""))
 rec_data <- rec_data %>% mutate(formatted_regime = ifelse(regime == "rec", "Omniscient", ifelse(regime == "no_rec", "No Rec", "Partial")))
@@ -300,6 +300,7 @@ time_dat <- time_dat %>% mutate(formatted_regime = ifelse(regime == "rec", "Omni
 time_dat <- time_dat %>% mutate(local_move = as.numeric(consumption_dist < 5))
 t <- time_dat %>% filter(t > 0) # local move isn't defined at the first time step so drop it to properly have smoothed plots
 
+# https://stackoverflow.com/questions/21066077/remove-fill-around-legend-key-in-ggplot
 g <- ggplot(t, aes(x=t, y=local_move)) +
     geom_smooth(aes(colour=formatted_regime)) +
     labs(x="t", y="Fraction of Local Search"
@@ -307,6 +308,7 @@ g <- ggplot(t, aes(x=t, y=local_move)) +
 if(use_hrbrthemes){
   g <- g + theme_ipsum_rc()
 }
+g <- g + theme(legend.position="bottom", legend.title = element_blank())  + guides(color=guide_legend(override.aes=list(fill=NA)))
 ggsave(
   filename=paste(WORKING_DIR, "figures/local_moves_25.jpeg", sep=""), 
   plot=g

### 2019-08-25 create_graphs.R
fix name of _rec graphs

@@ -247,36 +247,7 @@ graph_stats_homo <- function(df, var){
 }
 
 
-#' Calculate consumption diversity w/ Confidence Intervals derived from time_path
-#'
-#' @param stats_df Dataframe from time_path
-#' @param var  one of rho | beta | sigma | alpha
-#' @return Dataframe w/ mean and confidence intervals
-#' @examples
-#' get_stats_consumption_diversity(stats_df, "rho")
-get_stats_consumption_diversity_time_path <- function(stats_df, var){
 
-  df <- stats_df
-  
-  if (var == "rho") {
-    df <- df %>% group_by(regime, rho) 
-  } else if (var == "beta") {
-    df <- df %>% group_by(regime, beta) 
-  } else if (var == "sigma") {
-    df <- df %>% group_by(regime, sigma) 
-  } else if (var == "alpha") {
-    df <- df %>% group_by(regime, alpha)
-  }
-  df <- df %>%
-    mutate(local_move_mean = mean(local_move),
-           local_move_sd = sd(local_move),
-           local_move_n = n()) %>%
-    mutate(local_move_se =  local_move_mean/ sqrt(local_move_sd),
-           lower_ci_local_move = local_move_mean - qt(1 - (0.05 / 2), local_move_n - 1) * local_move_se,
-           upper_ci_local_move = local_move_mean + qt(1 - (0.05 / 2), local_move_n - 1) * local_move_se) %>% 
-    slice(1)
-  return(df)
-}
 
 scatter <- function(df, var_x, var_y, use_hrbrthemes, title){
   g <- ggplot(df, aes_string(var_x, var_y)) +
@@ -322,7 +293,7 @@ process_rec_homo_data <- function(N, T, use_hrbrthemes){
         g <- graph_stats_welfare(get_stats_welfare(rec_data, variable), variable,  use_hrbrthemes)
         ggsave(filename=file_name, plot=g)
         
-        rec_file_name <- paste(WORKING_DIR, "figures/", variable, "_rec.jpeg", sep="")
+        rec_file_name <- paste(WORKING_DIR, "figures/", variable,"_", metric,"_N_", N, "_T_", T, "_rec.jpeg", sep="")
         print(rec_file_name)
         g <- graph_stats_rec(get_stats_rec(filter(rec_data, regime == "partial"), variable), variable,  use_hrbrthemes)
[…]

### 2019-08-25 create_graphs.R
comment back in

@@ -503,7 +503,7 @@ N_s <- list(2000, 200)
 
 for(N in N_s){
   process_rec_homo_data(N, 20, use_hrbrthemes)
-  #process_time_path(N, 20, use_hrbrthemes)
+  process_time_path(N, 20, use_hrbrthemes)
 }
 
 

### 2019-08-28 paper/ex_ante_rec_sys.tex
add draft of paper to the repo

--- 2019-04-20 paper.tex
@@
-%
-% The first command in your LaTeX source must be the \documentclass command.
 \documentclass[sigconf]{acmart}
@@
 %\usepackage{lmodern}
-\usepackage{mathtools}
 \usepackage[font=small]{caption}
-\usepackage{subcaption}
 \usepackage{float}
@@
 \begin{document}
+% end of the preamble, start of the body of the document source.
 
@@
 \begin{abstract}
-Recommender systems (RS) are traditionally deployed in environments where users are uncertain about their preferences and thus face a problem of choice under uncertainty, but most popular design approaches ignore this fact.
-We argue that predicting and modeling consumer choice in these contexts can improve the usefulness of RS and reframe the RS problem as providing useful information to help reduce user uncertainty as opposed to simply predicting user preferences.
-We show how this insight can be utilized to design RS that mitigate the negative consequences of RS such as filter bubble and user homogenization effects as well as better understand the role that RS play in contributing to these phenomena.
 \end{abstract}
@@
 % the first part of the formatted document.
+
 \maketitle
@@
 
-Recommender Systems (RS) have been widely deployed in online platforms such as Netflix, Amazon, and YouTube.
-On these platforms the combination of the fact that most items are experience goods (i.e.
-valuation is only learned after consumption) and that the number of choices available to users is incredibly large, results in users not knowing their true preference ordering over all of the items.
-For instance, on Netflix there are thousands of movies, on Amazon millions of products, and YouTube billions of videos.
-As a result, RS have been useful in helping users make decisions on these platforms because they provide users with information to help them make their choices under uncertainty.
-\par
-However, while acknowledging that this is why RS are useful, most of the literature on RS ignores this user uncertainty and takes what we call an \textit{ex-post} viewpoint to preferences, focusing on the reported utility of users after they consume the item or inferring them via behavioral data \cite{adomavicius2005toward, zhao2018interpreting}.
-The RS problem traditionally is cast as solving the following prediction problem:
-\begin{align*}
-\arg\max\limits_{x \in A} u(x)
-\end{align*}
-\noindent where $A$ is the set of possible items and $u(x)$ represents the utility a user gets to consume good $x$.
-Recently there has been a move away from metrics focused on maximizing the accuracy of this prediction problem and towards optimizing for metrics such as serendipity or discoverability \cite{mcnee2006being, vargas2011rank} due to  the realization that accurate recommendations may not be the most useful recommendations for users.
-Furthermore, deployed RS have come under scrutiny for leading to ``filter bubbles" \cite{pariser2011filter} where users consume a less diverse set of content over time as well as increased homogenization amongst users \cite{chaney2018algorithmic, hosanagar2013will}.
-\par
-In this paper we want to put forward the argument that one approach to overcoming these problems is to make understanding how users choose which items to consume a first-order consideration when designing RS.
-In particular, we differentiate between the previous \textit{ex-post} interpretation of the RS problem and what we define as the \textit{ex-ante} viewpoint.
-Unlike the ex-post viewpoint taken in the traditional prediction problem, the ex-ante viewpoint simply states that users face uncertainty and so maximize their expected utility as is defined in economic decision theory under uncertainty \citep[see e.g.][]{mas1995microeconomic}:
-\begin{align*}
-\arg\max\limits_{x \in A} \mathbb{E}[u(x)]
-\end{align*}
-Fundamentally, the move away from accuracy towards alternative metrics as well as tensions that have arisen from increased user homogenization and filter bubbles come from the \textit{choices} that users make.
-This shift from thinking about true preferences after all uncertainty is resolved (the ex-post view) to user decisions under uncertainty (the ex-ante view) can serve as a theoretical tool to better understand what recommendations are useful for users as well as understanding what of the documented perverse effects of RS come from recommendation per-se and what effects come from inherent uncertainty in user decision-making.
-\section{The Value of the Ex-Ante View}
+Recommender Systems (RS) have become critical for assisting consumers in navigating the large choice sets that they face in many online markets.
+For instance, consumers have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
+Consumers in many cases are not aware of most items, let alone have full information about their preferences over them and, to make matters worse, the goods in these markets are usually experience goods whose true utility can only be learned after consumption.
+Furthermore, consumers interact with these systems routinely and watch more than one movie on Netflix, buy more than one product on Amazon, and listen to more than one video on YouTube.
 
-In this section we illustrate the differences between the ex-post and ex-ante view and show how it is useful for thinking about the design of RS.
-Our first point is trivial to state - since users do not know the true consumption values of the items, without any further information they may make sub-optimal consumption decisions.
-Consider the following stylized example.
-The choice set of the individual is $\Omega= \{x_1, x_2, x_3 \}$ and the ex-post utility values are given by
-$u(x_1) = 1$, $u(x_2) = 2$, $u(x_3) = 3$. 
-\par
-However, when users are making decisions they have unobservable \textit{beliefs} about the ex-post utility value of the items and make decisions given these beliefs.
-If users had more information then their beliefs would converge to the ex-post utility values but, generally, users do not have enough information and so have beliefs over the space of possible ex-post utility values.
-In this example we suppose these beliefs are simple so that users view each good as simple lotteries: for $n=1,2,3$, $u(x_n) = n + \varepsilon_n$, $\varepsilon_1 \sim \mathcal N (2, 1)$, $\varepsilon_2 \sim \mathcal N (0, 1)$, $\varepsilon_3 \sim \mathcal N (-5, 1)$.
-\par
-In this example, after uncertainty is unresolved, the user ought to consume $x_3$, but an expected utility maximizing consumer (without further information) would consume $x_1$.\footnote{Note that here we ignore an important component of decision-making under uncertainty from decision theory, risk preferences, and suppose users are risk-neutral.
-A \textit{risk-averse} user may choose to consume a good with lower expected payoff but lower ex-ante variance compared to another good.
-While risk-aversion is an important component of decision-making and important in the context of RS when considering choices under uncertainty, it is not our main focus here.}
-\par
-This observation, while seemingly trivial, has important implications.
-The fact that the user would choose $C$ over $A$ reveals that $C$ prefers $A$ given her current beliefs\footnote{This is a statement of \textit{revealed preference} which has been studied heavily in economic decision theory (see e.g.
-\cite[]{mas1995microeconomic}).} and allows us to get some information about a user's beliefs.
-Why are user beliefs important objects for designers of RS to understand and utilize?
-\par
-The first is that they are useful for understanding how users interact and get value from RS.
-In particular, users use RS to get \textit{information} about the set of items and update their beliefs about the value of the items.
-For instance, in the example, a RS could recommend to a user $x_3$ over $x_2$ and $x_1$ and that user will use this to update her beliefs about $x_3$.
-It is an empirical question that we leave for future work precisely how and to what extent users update their beliefs from recommendation and what factors of RS deisgn influence this.\footnote{There has been some work in this direction in understanding when recommendations are persuasive \cite{cremonesi2012investigating, gretzel2006persuasion} as well as the effect of their delivery on effectiveness \cite{murphy2014recommendation}.}
-\par
-Better understanding user beliefs is important for improving the performance of RS that optimize for metrics that move away from accuracy such as serendipity \cite{kotkov2016survey}.
-Serendipity-based RS attempt to provide recommendations that ``have the quality of being both unexpected and useful" \cite{maksai2015predicting} but it is hard to know what items are both unexpected and useful \cite{kotkov2016survey}.
-Understanding \textit{why} a user has not consumed a good depends on the beliefs that the user has and this is important for designing serendipitious recommendations.
-For instance, in the example, the user does not consume $x_3$ because she believes that the good has the potential to be catastrophically bad but, given more information, she would prefer to consume it over $x_1$ and $x_2$.
-It is possible that the ex-post and ex-ante utility of $x_3$ could be at best as good as the other two.
-The user may have an alternative with a known value -- say going for a walk around the neighborhood -- which she could always prefer to watching a given movie that has uncertain value but that the consumer thinks is always lower than the value of the walk.\footnote{An additional reason this could be useful is if one considers a risk-averse individual, then it may be that a user has not consumed an item because she has a high perceived variance for the item but high expected quality.}
-As a result, understanding the beliefs of users helps us understand why users have not consumed certain items and, in particular, which items a user may find useful given more information about it.
-This poses an important problem: designing systems that elicit not only ex-post valuations but also ex-ante beliefs about valuations, as only by knowing these can RS be effective in steering users' choices.\footnote{\cite{schotter2014belief} provides a survey of belief elicitation methods that are utilized in economics.}
-\par
-Understanding the choices that individuals will make as well as what they will end up liking are complementary problems but require different viewpoints.
-\cite{jiang2014choice, saavedra2016choice} argue that choice-based approaches alone can help in designing better RS, though they argue for these approaches because they do a better job at providing more accurate recommendations.
-We argue that the two problems should be considered separately -- though they may interact in interesting and useful ways.\footnote{In particular, some of the results from \cite{jiang2014choice, saavedra2016choice} may be driven by the fact that user choice embeds ratings information that may not have been observed by the recommender but is observed by the user.
-For instance, on Netflix I may have seen every episode of Friends when it was on TV which influences the consumption choices that I make.
-Alternatively, my friends can be a source of information that affect beliefs and choices.
-Netflix cannot observe neither of these data.}
-\par
-RS are designed to give users \textit{information} that is useful to help them make decisions.
-On the one hand, predicting a user's ex-post preferences is useful to know what items the user would like to consume.
-On the other hand, predicting what items a user will actually consume (without recommendation) can help reason about what information may actually be useful to give her.
-Furthermore, in designing RS to avoid filter bubble and homogenization effects, having accurate choice predictions allows mitigating such adverse effects to become a first-order component of design.
-In particular, by predicting both choice and ratings, RS can provide information to users that leads them to useful items but prevents them from falling into a filter bubble.
-\par
-Solving this choice prediction problem requires data on the choices that individuals make and the sets that they choose them from in addition to the traditional ratings or behavioral data that is collected.
-In the example given previously the fact that the consumer chooses $x_1$ from the set $\{x_1, x_2, x_3 \}$ would be the data that is useful for the choice prediction problem.
+Recommender systems have been influential in shaping consumer choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.
+While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems.
+There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals\footnote{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350/}, that personalized recommender systems lead consumers into \textit{filter bubbles} where they get effectively isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead consumers to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
+
+\textbf{Our Model and Contributions} In this paper, drawing from recent work in psychology and decision theory in economics, we provide a theoretical model for how personalized recommender systems can shape the behavior of long-lived consumers and how the insights from our model can be used to improve recommender system design as well as allow us to better understand the consequences of these systems.
+In our model, consumers have uncertainty about their true utility of the goods in their choice set and consume only a small fraction of the large choice set they face.
+
+First, we focus on understanding the degree to which recommender systems can lead consumers into \textit{filter bubbles}.
+Consistent with recent empirical work in the case of movie consumption \cite{nguyen2014exploring}, our model shows that individual consumption behavior over time naturally exhibits patterns consistent with a filter bubble effect \textit{without recommendation}.
+The key component of our model, drawn from the results of \cite{schulz2019structured}, is that the utilities of similar goods are correlated.
+This means that when consumers consume a good and learn its true utility this gives them information about similar goods.
+Crucially, this not only impacts the underlying belief about the average utility of the good, but also the amount of uncertainty.
+This learning spill-over induces consumers to consume goods ``similar" to those they consumed before leading to an increasing narrowing of consumption patterns.
+This effect is further exemplified when consumers are \textit{risk-averse}, a concept from economic decision theory where, all else being equal, consumers have a preference for goods with lower uncertainty to those with higher uncertainty.
+
+We then consider a stylized model of user recommendation and explore how recommendation can affect consumption patterns.
+We suppose each good is a sum of a common value component and an idiosyncratic component, where the idiosyncratic component is inherently unpredictable given other individual's preferences.
+We model existing recommendation systems as providing consumers with information on the common-value component and denote as this as partial recommendation.
+We consider how consumer behavior varies as we move from no recommendation to partial recommendation to omniscient recommendation, which is the optimal consumption path for consumers if they knew the true utility of all the goods in their choice set.
+
+Consistent with \cite{nguyen2014exploring}, we find that partial recommendation increases consumption diversity.
+However, we also find that partial recommendation increases homogeneity amongst consumers.
+
+\textbf{Related Work.}
+Recommender Systems
+
 \section{Filter Bubbles and Homogenization}\label{sec:consumer_learning}
+
 In this section we further show that differentiating between the ex-ante and ex-post view allows us to not only design better RS but also better understand their consequences.
@@
 \par
-\noindent \textbf{Setup}.
-Suppose that there is a group $I$ of individuals and that each individual $i \in I$ can choose from the same finite set of $N$ items $\mathcal{I} = \left\{0,1,...,N-1\right\}$.
-For simplicity, we assume that individuals only derive pleasure from a item $n \in \mathcal{I}$ the first time they consume it.
+\section{Our Model and Preliminaries}
+\noindent \textbf{Expected Utility Theory} First, we review basic economic decision theory. [Duarte can do this]
+
+\noindent \textbf{Consumers} We consider a set of consumers $I$ where each individual $i \in I$ faces the same finite set of $N$ items $\mathcal{N} = \left\{0,1,...,N-1\right\}$.
+For simplicity, we assume that individuals only derive pleasure from item $n \in \mathcal{N}$ the first time they consume it.
+
 We denote by $u_{in}$ individual $i$'s realized utility from consuming item $n$.
-We define the distance between two items as $d(n,m)=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}$ where $m$ and $n$ are the index of the items in $\mathcal{I}$ so that, conceptually, the items are located in a ring of size $N$.
-\par
-In particular, we consider that the utility derived from a given item can be decomposed in the following manner: $u_{in}=\beta v_{in} + (1-\beta) v_n$, where $v_{in}$ denotes an idiosyncratic component -- i.e.
+By realized utility we In particular, we consider that the realized utility derived from a given item can be decomposed in the following manner: $u_{in}= v_{in} + \beta v_n$, where $v_{in}$ denotes an idiosyncratic component -- i.e.
 consumer $i$'s idiosyncratic taste for good $n$ --  and $v_{n}$, a common-value component.
@@
 The scalar $\beta \in [0,1]$ denotes the degree to which utilities are idiosyncratic to each individual or common across individuals.
-If $\beta=1$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta=0$, every individual has the same preferences.
+If $\beta=0$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta \rightarrow \infty$, every individual has the same preferences.
+
 Stacking utilities in vector-form, we get ${\left(u_{in}\right)}_{n \in \mathcal{I}}=:U_i=\beta V_i+(1 - \beta) V $, the vector of utilities associated with each good, where $V_i ={\left(v_{in}\right)}_{n \in \mathcal{I}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{I}}$.
-\par
-We assume that user $i$ starts with some beliefs about $U_i$, namely that the idiosyncratic and common-value parts of the utilities are independent -- $V_i \perp \!\!\!
+
+\noindent \textbf{Consumer Beliefs} We assume that consumers do not know the realized utilities of the goods before consuming an item.
+We make this assumption as our model pertains to markets where the goods are experience goods and modeling consumers as being uncertain over their preferences in these environments  is common in the economics literature [add citations].
+Even though consumers have access to online reviews for goods in many of the markets where recommender systems are deployed, it is prohibitively costly for consumers to acquire information on all of the goods in their choice set.
+Even if they did so, this information would be helpful for reducing uncertainty on the common-value component of their utility and not the idiosyncratic component.
+While we do not explicitly incorporate this into our model, we use it to motivate that consumers make consumption decisions under some degree of uncertainty.
+
+Formally, consumer $i$ starts with some beliefs about $U_i$, namely that the idiosyncratic and common-value parts of the utilities are independent -- $V_i \perp \!\!\!
 \perp V$ -- and that each is multivariate normal $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ and $V \sim \mathcal N(\overline V, \Sigma)$, with $\overline V =0$.
-Furthermore, we assume that learning about the utility of good $n$ reveals more about the utility associated to items that are closer to it, i.e.
-$n\pm1$ (mod $N$), than about those farther away.
-This captures the idea that trying a product provides more information about similar products than about dissimilar ones.
+We impose the normality assumption for two reasons.
+The first is that consumers update their beliefs using Bayesian updating and recall that the normal distribution forms a conjugate family, which allows for simple posterior updates.
+The second is that it allows us to incorporate an easily interpretable correlation structure between the items.
+
+Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{in}\right)_{n \in \mathcal{I}}$ are drawn independently from a jointly normal distribution, where $\overline v_{in} \sim \mathcal N (0, \overline \sigma^2)$ are i.i.d.
+These $\overline v_{in}$ denote the prior belief that $i$ holds about the her valuation over good $n$.
+As people are exposed to different backgrounds, their beliefs about what is good for them also varies and $\overline v_{in}$ denotes this idiosyncrasy at the level of prior beliefs.
+
+\textbf{Consumer Learning}
+When a consumer consumes a good $n$ they learn the realized utility for that good.
+In our model we incorporate the idea that learning the utility of good $n$ gives a consumer information about similar items.
+This is drawn from recent empirical evidence in \cite{schulz2019structured} that consumers learn how to navigate large choice sets using similarity-based generalization.
+We assume that learning about the utility of good $n$ reveals more about the utility associated to items that are closer to it, which captures the idea that trying a product provides more information about similar products than about dissimilar ones.
+
+In order to have a well-defined notion of similarity we need to define a distance function between goods, which we define as $d(n,m):=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}$ where $m$ and $n$ are the index of the items in $\mathcal{N}$.
+This distance function has several nice properties.
+
 To this effect, we consider that the entry of $n$-th row and the ($m$)-th column of $\Sigma_i$ is given by $\sigma_i^2 \rho^{d(n,m)}$, and that of $\Sigma$ is given by $\sigma^2 \rho^{d(n,m)}$.
@@
 \par
-Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{in}\right)_{n \in \mathcal{I}}$ are drawn independently from a jointly normal distribution, where $\overline v_{in} \sim \mathcal N (0, \overline \sigma^2)$ are i.i.d.
-These $\overline v_{in}$ denote the prior belief that $i$ holds about the her valuation over good $n$.
-As people are exposed to different backgrounds, their beliefs about what is good for them also varies and $\overline v_{in}$ denotes this idiosyncrasy at the level of prior beliefs.
+
 \par
@@
 We are going to analyze three outcomes: how diverse consumer choices are, how similar are the choices that different individuals make and the resulting welfare.
-The first will be measured by $D_i =\frac{1}{T (T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$, the average pairwise distance between the consumed products, a natural measure of diversity in this environment that is utilized in the literature \cite{ziegler2005improving}.
-The second will be measured by a consumer homogeneity index $H:=1-\frac{1}{|I|(|I|-1)}\sum_{i,j \in I: i \ne j}d_J(C_i^T,C_j^T) $, where $d_J$ denotes the Jaccard distance and $H \in [0,1]$ which is similar to the approach to measuring homogeneity utilized in \cite{chaney2018algorithmic}.
+The first will be measured by $D_i =\frac{1}{N} \frac{1}{T (T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$, the average normalized pairwise distance between the consumed products, a natural measure of diversity in this environment that is utilized in the literature \cite{ziegler2005improving}.
+The second will be measured by a consumer homogeneity index $H:=\frac{1}{|I|(|I|-1)}\sum_{i,j \in I: i \ne j}d_J(C_i^T,C_j^T) $, where $d_J$ denotes the Jaccard index and $H \in [0,1]$ which is similar to the approach to measuring homogeneity utilized in \cite{chaney2018algorithmic}.
 Note that a higher $H$ indicates more homogeneity.
@@
 In the third case, the recommender knows $V$ but does not know neither $V_i$ nor, crucially, users' beliefs $\overline V_i$.
-The recommendation will be that the consumer chooses $r^{t} \in \arg \max_{n \in \mathcal{I} \backslash C_i^T} u_n$, but we assume the recommender provides full information about $V$.
+The recommendation will be that the consumer $i$ chooses $r_{it} \in \arg \max_{n \in \mathcal{I} \backslash C_i^T} u_n$, but we assume the recommender provides full information about $V$.
 For instance, the recommender could display the whole distribution of utilities reported by other users or even its average, which is a good proxy for the common value component.\footnote{
@@
 We simulate over a grid of parameter values for $\sigma, \sigma_i, \rho, \rho_i, \beta$ and report results for $P=50, I = 50, N = 1000, T = 25$.\footnote{Due to space constraints we omit simulations for different values of $T$ but, unless otherwise noted, our results are robust to changes to $T$ as long as $T$ is not too large (close to $N$) or too small (close to $0$).}
-\par
-
 \begin{table}[bt]
@@
   \hline
-No Rec & 0.42 $\pm$ 0.002 & 191.04 $\pm$ 0.27 & 0.001 $\pm$ 0 \\ 
-Partial Rec & 1.25 $\pm$ 0.005 & 221.83 $\pm$ 0.20 & 0.174 $\pm$ 0.004 \\ 
-Omniscient Rec & 2.24 $\pm$ 0.006 & 237.59 $\pm$ 0.12 & 0.035 $\pm$ 0.002 \\ 
+No Rec & 0.42 $\pm$ 0.002 & 0.19 $\pm$ 0.0003 & 0.001 $\pm$ 0 \\ 
+Partial Rec & 1.25 $\pm$ 0.005 & 0.22 $\pm$ 0.0002 & 0.174 $\pm$ 0.004 \\ 
+Omniscient Rec & 2.24 $\pm$ 0.006 & 0.24 $\pm$ 0.0001 & 0.035 $\pm$ 0.002 \\ 
    \hline
@@
 \end{table}
-
 \begin{figure*}[!htbp]
@@
         \centering
-        \includegraphics[width=\linewidth, height=1.75in]{figures/local_moves_25}
-        \caption{Fraction of Local Search}
+        \includegraphics[width=0.9\linewidth, height=1.5in]{local_moves_25}
+        \caption{Extent of ``Local"  Consumption}
         \label{fig:local_consumption}
@@
         \centering
-        \includegraphics[width=\linewidth, height=1.75in]{figures/welfare_homo_combo}
-        \caption{Welfare and Homogeneity varying $\beta$}
+        \includegraphics[width=0.9\linewidth, height=1.5in]{welfare_homo_combo}
+        \caption{Welfare and Homogeneity}
         \label{fig:beta_vary}
@@
      \centering
-     \includegraphics[width=\linewidth, height=1.75in]{figures/rec_obedience_25}
+     \includegraphics[width=0.9\linewidth, height=1.5in]{rec_obedience_25}
      \caption{Recommendation Effectiveness}\label{fig:rec_obey}
@@
      \centering
-     \includegraphics[width=\linewidth, height=1.75in]{"figures/Diversity vs Welfare - No Recommendation"}
-     \caption{ Diversity vs Welfare under No Recommendation}\label{fig:diversity_welfare_no_rec}
+     \includegraphics[width=0.9\linewidth, height=1.5in]{Diversity_vs_Welfare_No_Recommendation}
+     \caption{Diversity vs Welfare under No Recommendation}\label{fig:diversity_welfare_no_rec}
    \end{minipage}
 \end{figure*}
-\small
+\par
 \noindent \textbf{Results}.
@@
 Figure \ref{fig:local_consumption} shows how ``local" consumption choices are by looking at what fraction of consecutive consumption choices are close to each other with the definition of close being less than distance of 10.
-The partial recommendation case leads to higher consumption diversity and higher welfare, but has a significantly higher level of homogeneity.
+The partial recommendation case leads to higher consumption diversity and higher welfare, but also significantly higher homogeneity.
 
@@
 As Table \ref{table:agg_results} shows, our model is consistent with this effect but it depends on the nature of recommendation.
-If recommendation only provides information on the common-value component then welfare and consumption diversity increase but, correspondingly, homogenization increases dramatically.
-In this case the information on the common-value component leads users to have more information on where to explore in the product space.
+If recommendations only provides information on the common-value component (partially informed recommender) then welfare and consumption diversity increase when contrasted to the no recommendation case, but homogenization increases dramatically.
+Knowing the common-value component leads users to have more information on where to explore in the product space.
 However, the fact that this information is about the common-value component and the same across all individuals leads them to explore the same sections of the product space and leads to substantially higher homogeneity among users.
-If recommendation takes into account user beliefs on the idiosyncratic  component of utility then welfare and consumption diversity can further increase and homogeneity will decrease.
+If recommendation takes into account user beliefs on the idiosyncratic component of utility (omniscient recommender), then welfare and consumption diversity can further increase and homogeneity will decrease.
 \par
-Low consumption diversity by itself is not necessarily bad.
-Figure \ref{fig:diversity_welfare_no_rec} shows that in the no recommendation case welfare and consumption diversity are negatively correlated.
+Low consumption diversity is not necessarily bad by itself.
+Figure \ref{fig:diversity_welfare_no_rec} shows that, in the no recommendation case, welfare and consumption diversity are negatively correlated.
 This is primarily because content diversity in the no recommendation case can arise from the fact that the user consumes a bad product and is forced to jump around the product space looking for ``good" products.
-However, if a user finds a high utility product right away then staying in this neighborhood may yield higher utility due to the correlation of utilities.
-The problem is that, even when the user finds a ``good" product, this may only be ``locally" good and the user will then excessively consume items around it.
+However, if a user finds a high utility product right away, then staying in this neighborhood may yield higher utility due to the correlation of utilities.
+The problem is that, even when the user finds a ``good" product, this may only be ``locally" good and the user will then excessively consume items around it due to having more information about these than more ``different'' ones, i.e.
+farther away.
 \par
@@
 Most interestingly, when $\beta$ is high so that the common-value component is small, recommendation can be harmful to users.
-The information from recommendation still leads to large user homogeneity even though the optimum would be for there to be almost none and leads users to have lower welfare than under no recommendation.
+The information from recommendation still leads to large user homogeneity even though the optimum would be for there to be almost none and leads to lower welfare than under no recommendation.
+
+\section{The Value of the Ex-Ante View}
+
+In this section we illustrate the differences between the ex-post and ex-ante view and show how it is useful for thinking about the design of RS.
+Our first point is trivial to state - since users do not know the true consumption values of the items, without any further information they may make \textit{ex-post} sub-optimal consumption decisions, but \textit{ex-ante} optimal given the information available to them at the moment of choice.
+Consider the following stylized example.
+The choice set of the individual is $\Omega= \{x_1, x_2, x_3 \}$ and the ex-post utility values are given by
+$u(x_1) = 1$, $u(x_2) = 2$, $u(x_3) = 3$. 
+\par
+However, when users are making decisions they have unobservable \textit{beliefs} about the ex-post utility value of the items and make decisions given these beliefs.
+As users obtain more information, their beliefs converge to the ex-post utility values but, generally, users do not have enough information and so have beliefs over the space of possible ex-post utility values.
+In this example we suppose these beliefs are simple so that users view each good as simple lotteries: for $n=1,2,3$, $u(x_n) = n + \varepsilon_n$, $\varepsilon_1 \sim \mathcal N (2, 1)$, $\varepsilon_2 \sim \mathcal N (0, 1)$, $\varepsilon_3 \sim \mathcal N (-5, 2)$.
+\par
+After uncertainty is unresolved, the user would consume $x_3$, but an expected utility maximizing consumer (without further information) would choose $x_1$.\footnote{We ignore an important component of decision-making under uncertainty, risk attitudes, by focusing on risk-neutral consumers.
+A \textit{risk-averse} user may choose to consume a good with lower expected \textit{ex-post} payoff but lower ex-ante variance compared to another good.
+While risk-aversion is important for decision-making under uncertainty, especially in the context of RS, it is not our main focus here.}
+\par
+This observation, while seemingly trivial, has important implications.
+The fact that the user would choose $C$ over $A$ reveals that $C$ prefers $A$ given her current beliefs\footnote{This is a statement of \textit{revealed preference} which has been studied heavily in economic decision theory (see e.g.
+\cite[]{mas1995microeconomic}).} and allows the system designer to get some information about a user's beliefs.
+Why are user beliefs important objects for designers of RS to understand and utilize?
+\par
+Beliefs are useful for understanding how users interact and get value from RS.
+In particular, users use RS to get \textit{information} about the set of items and update their beliefs about the value of the items.
+For instance, in the example, a RS could recommend to a user $x_3$ over $x_2$ and $x_1$ and that user will use this to update her beliefs about $x_3$.
+It is an empirical question that we leave for future work precisely how and to what extent users update their beliefs from recommendation and what factors of RS deisgn influence this.\footnote{There has been some work in this direction in understanding when recommendations are persuasive \cite{cremonesi2012investigating, gretzel2006persuasion} as well as the effect of their delivery on effectiveness \cite{murphy2014recommendation}.}
+\par
+Better understanding user beliefs is important for improving the performance of RS that optimize for metrics that move away from accuracy such as serendipity \cite{kotkov2016survey}.
+Serendipity-based RS attempt to provide recommendations that ``have the quality of being both unexpected and useful" \cite{maksai2015predicting} but it is hard to know what items are both unexpected and useful \cite{kotkov2016survey}.
+Understanding \textit{why} a user has not consumed a good depends on the beliefs that the user has and this is important for designing serendipitious recommendations.
+For instance, in the example, the user does not consume $x_3$ because she expects the good to turn out worse than the other two.
+However, if given more information, she would prefer to consume it over $x_1$ and $x_2$.
+Moreover, when the user can use her known valuation of an item to make inferences on how good another item is, consumption is path-dependent, as beliefs guide consumption and consumption affects beliefs about other goods.
+For instance, learning the value of $x_1$ says something about how good $x_2$ is.
+%When considering risk-averse individuals, when the user has alternatives with a known value,The consumer may have also consumed Another reason why the user may not .
+The user may have an alternative with a known value -- say going for a walk around the neighborhood -- which she could always prefer to watching a given movie that has uncertain value but that the consumer thinks is always lower than the value of the walk.\footnote{An additional reason this could be useful is if one considers a risk-averse individual, then it may be that a user has not consumed an item because she has a high perceived variance for the item's quality.}
+As a result, understanding the beliefs of users helps us understand why users have not consumed certain items and, in particular, which items a user may find useful given more information about it.
+This poses an important problem: designing systems that elicit not only ex-post valuations but also ex-ante beliefs about valuations, as only by knowing these can RS be effective in steering users' choices.
+In particular, choice has to be perceived as guided by ex-ante utility, whereas ratings by the ex-post realized utility.
+\par
+Understanding the choices that individuals will make in addition to what they will end up liking are complementary problems but require different viewpoints.
+\cite{jiang2014choice, saavedra2016choice} argue that choice-based approaches alone can help in designing better RS, though they argue for these approaches because they do a better job at providing more accurate recommendations.
+We argue that the two problems should be considered separately -- though they may interact in interesting and useful ways.\footnote{In particular, some results in \cite{jiang2014choice, saavedra2016choice} may be driven by the fact that user choice embeds ratings information that may not have been observed by the recommender but is observed by the user.
+Ratings, reviews, friends, there are many sources of information affecting a person's beliefs and choices that are unobservable by RS.
+Other consumption choices, e.g.
+movies seen in the cinema, change beliefs about for instance how good a director is, which affect beliefs about other movies and guide choices on a streaming platform that is unable to observe these data.}
+\par
+RS are designed to give users \textit{information} that is useful to help them make decisions.
+On the one hand, predicting a user's ex-post preferences is useful to know what items the user would like to consume.
+On the other hand, predicting what items a user will actually consume (without recommendation) can help reason about what information may actually be useful to give her.
+Furthermore, in designing RS to avoid filter bubble and homogenization effects, having accurate choice predictions allows mitigating such adverse effects to become a first-order component of design.
+In particular, by predicting both choice and ratings, RS can provide information to users that leads them to useful items but prevents them from falling into a filter bubble.
+\par
+Solving this choice prediction problem requires data on the choices that individuals make and the sets that they choose them from in addition to the traditional ratings or behavioral data that is collected.
+In the example given previously the fact that the consumer chooses $x_1$ from the set $\{x_1, x_2, x_3 \}$ would be the data that is useful for the choice prediction problem, as would the consumer's ex-ante beliefs about her valuation of the goods.
 
@@
 By collecting appropriate data about user choices and user beliefs, RS can be built to better understand what choices users are likely to make and thus what information would be useful to give them as opposed to simply predicting what items a user will like.
-This approach can not only aid in designing more useful RS, but can also be utilized to better understand recently documented adverse effects of RS such as filter bubbles and homogenization.
-We presented a model of user choice under uncertainty with correlated utilities that naturally generates filter bubble effects without any recommendation.
-We showed how, by not properly incorporating user beliefs into recommendation, RS can help alleviate filter bubble effects and increase utility but also substantially increase homogenization amongst users.
-Incorporating user beliefs can help increase utility even further but decrease homogenization effects.
-We leave for future work approaches that incorporate our insights from into traditional recommender system design methods.
+This approach can not only aid in designing more useful RS, but can also be utilized to better understand and prevent recently documented adverse effects of RS such as filter bubbles and homogenization.
 

### 2019-08-28 paper/dump_to_dataset.py

@@ -1,0 +1,140 @@
+import pickle
+import csv
+from copy import copy
+
+from scipy.spatial.distance import jaccard
+
+OMNI = 'rec'
+PARTIAL = 'partial'
+NO_REC = 'no_rec'
+rec_policy_keys = [OMNI, PARTIAL, NO_REC]
+WORKING_DIR = ''
+with open(WORKING_DIR + 'sim_results_t_25.p', 'rb') as fp:
+    df = pickle.load(fp)
+
+def d(i,j, N):
+    return min(abs(i-j), abs(j-i), abs(j-i-N), abs(i-j-N))
+
+def div_fun(CiT, T, N):
+    div_score = 0.0
+    for i in range(len(CiT)):
+        for j in range(len(CiT)):
+            if i == j: continue
+            div_score = div_score + d(CiT[i],CiT[j], N)
+    return div_score*(1.0/(T*(T-1)))
+
+def homogeneity(C1, C2, type_sim="jaccard"):
+    if type_sim == "jaccard":
+        return 1 - jaccard(C1, C2)
+    return None
+
+def follow_rec(Ci, Rec, T, N):
+    s = 0.0
+    for i in range(len(Ci)):
+        if Ci[i] == Rec[i]:
+            s += 1
+    return s / len(Ci)
+
+
+INDIVIDUAL_FIELD_NAMES =['pop_idx', 'indiv_idx', 'regime', 'welfare', 'diversity_score', 'rho', 'beta', 'follow_recommendation', 'N', 'T', 'sigma']
+with open(WORKING_DIR + 'rec_data_t_25.csv', 'w') as rec_csv:
+    data_writer = csv.DictWriter(rec_csv, fieldnames=INDIVIDUAL_FIELD_NAMES)
+    data_writer.writeheader()
+    for key, value in df.items():
+        for pop_idx in range(len(value)):
+            N = key[0]
+            T = key[1]
+            dat = {
+                'pop_idx': pop_idx,
+                'N': N,
+                'T': T,
+                'rho': key[2],
+                'beta': key[3],
+                'sigma': key[4]
+            }
+            cur = value[pop_idx]
+            welfare = cur['Welfare']
+            consumption = cur['Consumption']
+            for policy in rec_policy_keys:
+            	dat['regime'] = policy
+            	for indiv_idx in range(len(welfare[policy])):
+                    dat['indiv_idx'] = indiv_idx
+                    dat['welfare'] = welfare[policy][indiv_idx]
[…]

### 2019-08-28 paper/plots.R

@@ -1,0 +1,357 @@
+library(ggplot2)
+library(dplyr)
+library(hrbrthemes)
+library(tidyverse)
+
+library(gridExtra)
+
+
+get_stats_diversity <- function(rec_data, var){
+  df <- rec_data
+  if (var == "rho") { 
+    df <- df %>% group_by(regime, rho) 
+  } else if (var == "beta") {
+    df <- df %>% group_by(regime, beta) 
+  } else if (var == "sigma") {
+    df <- rec_data %>% group_by(regime, sigma) 
+  }
+  
+  df <- df %>% 
+    mutate(diversity_mean = mean(diversity_score),
+           diversity_sd = sd(diversity_score),
+           diversity_n = n()) %>%
+    mutate(diversity_se =  diversity_sd/ sqrt(diversity_n),
+           lower_ci_diversity = diversity_mean - qt(1 - (0.05 / 2), diversity_n - 1) * diversity_se,
+           upper_ci_diveristy = diversity_mean + qt(1 - (0.05 / 2), diversity_n - 1) * diversity_se) %>% 
+    slice(1)
+  return(df)
+  
+}
+
+get_stats_rec <- function(rec_data, var){
+  rec_data$follow_recommendation <- as.numeric(levels(rec_data$follow_recommendation)[rec_data$follow_recommendation])
+  
+  df <- rec_data
+  
+  if (var == "rho") { # there must be a better way
+    df <- df %>% group_by(rho) 
+  } else if (var == "beta") {
+    df <- df %>% group_by(beta) 
+  } else if (var == "sigma") {
+    df <- df %>% group_by(sigma) 
+  }
+  
+  df <- df %>% 
+    mutate(rec_mean = mean(follow_recommendation),
+           rec_sd = sd(follow_recommendation),
+           rec_n = n()) %>%
+    mutate(rec_se =  rec_sd/ sqrt(rec_n),
+           lower_ci_rec = rec_mean - qt(1 - (0.05 / 2), rec_n - 1) * rec_se,
+           upper_ci_rec = rec_mean + qt(1 - (0.05 / 2), rec_n - 1) * rec_se) %>% 
+    slice(1)
+  return(df)
+}
+
+# https://stackoverflow.com/questions/22309285/how-to-use-a-variable-to-specify-column-name-in-ggplot
+# graph w/ themes
+graph_stats_diversity <- function(df, var, use_hrbrthemes){
+  T_val <- df$T[1]
[…]

### 2019-08-28 paper/simulations.py

@@ -1,0 +1,212 @@
+import pandas as pd
+import numpy as np
+import pickle
+import cProfile
+from copy import copy
+
+from joblib import Parallel, delayed
+
+import itertools
+from scipy import stats
+from scipy.stats import beta
+
+
+import pandas as pd
+import numpy as np
+import pickle
+import cProfile
+import numba
+
+from joblib import Parallel, delayed
+
+import itertools
+from scipy import stats
+from scipy.stats import beta
+
+WRITE_FILE = 'sim_results_t_25.p'
+OMNI = 'rec'
+PARTIAL = 'partial'
+NO_REC = 'no_rec'
+
+
+###SETUP FUNCTIONS
+def d(i,j, N):
+    return min(abs(i-j), abs(j-i), abs(j-i-N), abs(i-j-N))
+
+def cov_mat_fun(sigm, rh, N):
+    cov_mat = np.ones((N,N))
+    for i in range(0,N):
+        for j in range(0,N):
+            cov_mat[i,j] = rh**d(i,j, N)
+    cov_mat = sigm * cov_mat
+    return cov_mat
+
+
+
+### Welfare Functions - Statistic Calculation Functions
+
+
+def w_fun(CiT,Ui):
+    w_score = 0.0
+    for i in range(len(CiT)):
+        w_score = w_score + Ui[CiT[i]]
+    return w_score*(T**(-1))
+
+
+## Bayesian updating
+@numba.jit
+def inv_nla_jit(A):
+  return np.linalg.inv(A)
+
+def update_Ui(Cit,Ui,mu_Ui, Sigma_Ui, Nset):
+    x1 = Cit
+    x2 = [n for n in Nset if n not in Cit]
+    Nit = [n for n in Nset if n not in Cit]
+    mu1 = np.array([mu_Ui[n] for n in x1]).reshape((1,len(x1)))
+    mu2 = np.array([mu_Ui[n] for n in x2]).reshape((1,len(x2)))
+    Sigma11 = np.ones((len(x1),len(x1)))
+    Sigma21 = np.ones((len(x2),len(x1)))
+    for i in range(len(Cit)):
+        for j in range(len(Cit)):
+            Sigma11[i,j] = Sigma_Ui[Cit[i],Cit[j]]
+    for i in range(len(Cit)):
[…]

### 2019-08-30 paper/ex_ante_rec_sys.tex
updated paper

@@
 
-\textbf{Our Model and Contributions} In this paper, drawing from recent work in psychology and decision theory in economics, we provide a theoretical model for how personalized recommender systems can shape the behavior of long-lived consumers and how the insights from our model can be used to improve recommender system design as well as allow us to better understand the consequences of these systems.
-In our model, consumers have uncertainty about their true utility of the goods in their choice set and consume only a small fraction of the large choice set they face.
+\textbf{Our Model and Contributions} In this paper, drawing from recent work in psychology and decision theory in economics, we first develop a model of consumer decision-making in the context of large choice sets faced by consumers in online markets.
+We then ask how a stylized model of recommendation can shape the behavior of long-lived consumers and finally ask how the insights from our model can be used to improve recommender system design and to better understand the consequences of these systems.
 
-First, we focus on understanding the degree to which recommender systems can lead consumers into \textit{filter bubbles}.
+We focus on understanding the degree to which recommender systems can lead consumers into \textit{filter bubbles}.
 Consistent with recent empirical work in the case of movie consumption \cite{nguyen2014exploring}, our model shows that individual consumption behavior over time naturally exhibits patterns consistent with a filter bubble effect \textit{without recommendation}.
@@
 
-\section{Filter Bubbles and Homogenization}\label{sec:consumer_learning}
+\section{Our Model and Preliminaries}
 
-In this section we further show that differentiating between the ex-ante and ex-post view allows us to not only design better RS but also better understand their consequences.
-To do so, we study a simple model where users have beliefs about how good each item is and can learn their valuation over the items by consuming them.
-However, drawing from the main idea behind content-based RS, similar items have similar valuations and consumers' beliefs about neighboring items update after consuming an item which influences future consumption decisions.
-For instance, in our model, if a user watches \textit{The Matrix} and finds out that it is a good movie then they will update their beliefs positively about a similar movie such as \textit{The Matrix Reloaded} (the sequel to \textit{The Matrix}) but not update them much for a very different movie such as \textit{La La Land}.
+\subsection{Preliminaries and Illustrative Example}
+[Duarte can do this] \\
+\noindent \textbf{Expected Utility Theory} First, we review basic economic decision theory.
 
-Viewing the goal of the RS as providing users with additional information they can utilize to update their beliefs, we first show that, without recommendation, users may get stuck in a ``filter bubble" whereby they consume increasingly similar content and that, in this case, welfare will be negatively correlated with diversity.
-We show how providing information to users via recommendation can lead users out of this filter bubble.
-However, if recommendation does not properly take into account user beliefs, recommendations can lead to increasing homogenization among users and lower welfare and consumption diversity.
-Finally we then show that when the recommender does not know the user's beliefs about her valuation, recommendation becomes decreasingly useful as the user learns more about her own preferences.
+\noindent \textbf{Illustrative Example}
+
+\subsection{Model}
 \par
-\section{Our Model and Preliminaries}
-\noindent \textbf{Expected Utility Theory} First, we review basic economic decision theory. [Duarte can do this]
-
 \noindent \textbf{Consumers} We consider a set of consumers $I$ where each individual $i \in I$ faces the same finite set of $N$ items $\mathcal{N} = \left\{0,1,...,N-1\right\}$.
@@
 
-Stacking utilities in vector-form, we get ${\left(u_{in}\right)}_{n \in \mathcal{I}}=:U_i=\beta V_i+(1 - \beta) V $, the vector of utilities associated with each good, where $V_i ={\left(v_{in}\right)}_{n \in \mathcal{I}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{I}}$.
-
+Stacking utilities in vector-form, we get ${\left(u_{in}\right)}_{n \in \mathcal{N}}=:U_i=V_i+ \beta V $, the vector of utilities associated with each good, where $V_i ={\left(v_{in}\right)}_{n \in \mathcal{N}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{N}}$.
+\par
 \noindent \textbf{Consumer Beliefs} We assume that consumers do not know the realized utilities of the goods before consuming an item.
@@
 Formally, consumer $i$ starts with some beliefs about $U_i$, namely that the idiosyncratic and common-value parts of the utilities are independent -- $V_i \perp \!\!\!
-\perp V$ -- and that each is multivariate normal $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ and $V \sim \mathcal N(\overline V, \Sigma)$, with $\overline V =0$.
+\perp V$ -- and that each is multivariate normal
+\begin{definition} The idiosyncratic utility and common-value utility are distributed as follows:
+\begin{center}
+\begin{enumerate}
+\item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ 
+\item $V \sim \mathcal N(\overline V, \Sigma)$ with $\overline V =0$
+\end{enumerate}
+\end{center}
+\end{definition}
+
 We impose the normality assumption for two reasons.
@@
 
-Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{in}\right)_{n \in \mathcal{I}}$ are drawn independently from a jointly normal distribution, where $\overline v_{in} \sim \mathcal N (0, \overline \sigma^2)$ are i.i.d.
+Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{in}\right)_{n \in \mathcal{N}}$ are drawn independently from a jointly normal distribution, where $\overline v_{in} \sim \mathcal N (0, \overline \sigma^2)$ are i.i.d.
 These $\overline v_{in}$ denote the prior belief that $i$ holds about the her valuation over good $n$.
 As people are exposed to different backgrounds, their beliefs about what is good for them also varies and $\overline v_{in}$ denotes this idiosyncrasy at the level of prior beliefs.
-
-\textbf{Consumer Learning}
+\par
+\noindent \textbf{Consumer Learning}
 When a consumer consumes a good $n$ they learn the realized utility for that good.
@@
 In order to have a well-defined notion of similarity we need to define a distance function between goods, which we define as $d(n,m):=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}$ where $m$ and $n$ are the index of the items in $\mathcal{N}$.
-This distance function has several nice properties.
-
-To this effect, we consider that the entry of $n$-th row and the ($m$)-th column of $\Sigma_i$ is given by $\sigma_i^2 \rho^{d(n,m)}$, and that of $\Sigma$ is given by $\sigma^2 \rho^{d(n,m)}$.
-The scalar $\rho \in [0,1]$ the covariance structure: a higher $\rho$ implies that learning the utility of $n$ is more informative about products nearby.
+We consider that the entry of $n$-th row and the ($m$)-th column of $\Sigma_i$ is given by $\sigma_i^2 \rho^{d(n,m)}$, and that of $\Sigma$ is given by $\sigma^2 \rho^{d(n,m)}$.
+The scalar $\rho \in [0,1]$ therefore impacts the covariance structure: a higher $\rho$ implies that learning the utility of $n$ is more informative about products nearby.
 Informativeness, for any $\rho \in (0,1)$, is decreasing in distance.
+The particular distance function that we utilize leads to this covariance structure being simple, where the $(n,n+1)$-th entry in the covariance matrix is $\rho$ , $(n,n+2)$-th entry is $\rho^2$, etc.
+This spill-over process is represented in Figure [insert fig].
 \par
-
-\par
+\textbf{Consumer Decision-Making}
 We assume the user makes $T$ choices and therefore can only consume up to $T$ items, where $T$ is but a small fraction of $N$.
 This captures the idea that users are faced with an immense choice set but that ultimately they end up experiencing (and learning) about just a small fraction of it.
-Each period $t=1,...,T$, the consumer chooses the best product that she has not yet tried ($n_i^t$) given the information that past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and that that the RS offers ($R$), i.e.
-$n_i^t \in \arg \max_{ n \in \mathcal{I} \backslash C_i^{t-1}} \mathbb E \left[u_{in} \mid C_i^{t-1}, R\right]$, with ties broken uniformly at random.
+Since this is a sequential decision-making problem under uncertainty, in principle consumers face an exploration-exploitation trade-off.
+However, for tractability, we impose that consumers are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
+Any ties are broken uniformly at random.
+This assumption is critical not only for tractability but also in order to have easily defined benchmarks.
 \par
-We are going to analyze three outcomes: how diverse consumer choices are, how similar are the choices that different individuals make and the resulting welfare.
-The first will be measured by $D_i =\frac{1}{N} \frac{1}{T (T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$, the average normalized pairwise distance between the consumed products, a natural measure of diversity in this environment that is utilized in the literature \cite{ziegler2005improving}.
-The second will be measured by a consumer homogeneity index $H:=\frac{1}{|I|(|I|-1)}\sum_{i,j \in I: i \ne j}d_J(C_i^T,C_j^T) $, where $d_J$ denotes the Jaccard index and $H \in [0,1]$ which is similar to the approach to measuring homogeneity utilized in \cite{chaney2018algorithmic}.
-Note that a higher $H$ indicates more homogeneity.
-Consumer $i$'s welfare will just be the average of realized utilities, to control for the effect of $T$, $W_i=\frac{1}{T}\sum_{n \in C_i^T} u_{in}$.
-\par
-Finally, we are going to study three different cases: (i) no recommendation, (ii) omniscient recommender and (iii) partially informed recommender where the recommender observes utilities accrued, but does not know consumer $i$'s starting beliefs $\overline U_i$.
-In the first case, users choose their items myopically, that is, using a greedy algorithm, and pick the item that has maximum expected utility given beliefs at time $t$.
-Upon consuming that item and learning its value, the users update their beliefs about their valuation of other items and iterate this process $T$ times.
-In the second case, the recommender knows exactly $V_i$ and $V$ and can therefore recommend the best item for each consumer.
-In the third case, the recommender knows $V$ but does not know neither $V_i$ nor, crucially, users' beliefs $\overline V_i$.
-The recommendation will be that the consumer $i$ chooses $r_{it} \in \arg \max_{n \in \mathcal{I} \backslash C_i^T} u_n$, but we assume the recommender provides full information about $V$.
+\noindent \textbf{Recommendation}
+Our model of recommendation is stylized in order to provide qualitative insights into how recommendation shapes behavior, instead of looking at realistic implementations of recommender systems.
+We model recommendation as giving consumers information about the utility of the goods.
+
+We will consider three cases.
+The case of primary interest is \textit{partial recommendation} where the recommender observes utilities accrued, but does not know consumer $i$'s starting beliefs $\overline U_i$.
+In this case, the recommender knows $V$ but does not know neither $V_i$ nor, crucially, users' beliefs $\overline V_i$.
+The recommendation will be that the consumer $i$ chooses $r_{it} \in \arg \max_{n \in \mathcal{N} \backslash C_i^T} u_n$, but we assume the recommender provides full information about $V$.
 For instance, the recommender could display the whole distribution of utilities reported by other users or even its average, which is a good proxy for the common value component.\footnote{
 The best item that could recommended with such information if the item with highest common value component.
-However, in that case, recommending only a single item generates costly updating to the consumer and provides little guidance as to what she should indeed pick as the common value might be of little importance when compared to the idiosyncratic component.
+However, in that case, recommending only a single item generates costly Bayesian updating to the consumer and provides little guidance as to what she should indeed pick as the common value might be of little importance when compared to the idiosyncratic component.
 Therefore, we assume that the RS reports the whole $V$, which results in higher expected welfare in choices and is not costly to implement.
-} Note that it is not necessarily optimal for the user to follow the recommendation, that is, to pick the item with the highest common-value component $v_n$.
+}
+Note that with partial recommendation it is not necessarily optimal for the user to follow the recommendation, that is, to pick the item with the highest common-value component $v_n$.
 Consumer's beliefs about her valuation of each item become crucial in this case: knowing $V$ may change the original ranking, but given this new information the consumer may find it best to pick an item other than the one recommended.
+
+We further consider two cases that serve primarily as benchmarks.
+The first is \textit{no recommendation}, where consumers get no additional information about utilities and make consumption choices based on their beliefs and consumption history.
+This gives us a benchmark as to how consumers in our model would behave \textit{without} recommendation so that we can analyze what changes with the introduction of recommendation.
+The second is \textit{omniscient recommendation} where the recommender knows the true utility of each good for each consumer and can therefore recommend the best remaining good in every period.
+This gives us a full information benchmark, which is the optimal consumption path for a consumer if all uncertainty about their preferences was resolved.
+We compare this benchmark to the partial recommendation case and it allows us to understand how much this form of recommendation affects behavior relative to the ideal.
 \par
 \noindent \textbf{Simulation Details}.
-We will simulate $P$ populations of $I$ users and compare their consumption choices under different recommendation system regimes.
-We simulate over a grid of parameter values for $\sigma, \sigma_i, \rho, \rho_i, \beta$ and report results for $P=50, I = 50, N = 1000, T = 25$.\footnote{Due to space constraints we omit simulations for different values of $T$ but, unless otherwise noted, our results are robust to changes to $T$ as long as $T$ is not too large (close to $N$) or too small (close to $0$).}
+We analyze our model using numerical simulation since the sequential decision-making component of our model paired with the rich covariance structure between the items make it difficult to characterize optimal consumer behavior analytically.
+We explore how consumption patterns differ as we consider different recommendation regimes and report representative results from our simulations.
+$P$ populations of $I$ users and compare their consumption choices under different recommendation system regimes.
+We simulate over $100$ populations with $100$ consumers.
+\section{Results}
+
 \begin{table}[bt]
@@
 \end{figure*}
+
+We are going to analyze three outcomes: how diverse consumer choices are, how similar are the choices that different individuals make and the resulting welfare.
+The first will be measured by $D_i =\frac{1}{N} \frac{1}{T (T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$, the average normalized pairwise distance between the consumed products, a natural measure of diversity in this environment that is utilized in the literature \cite{ziegler2005improving}.
+The second will be measured by a consumer homogeneity index $H:=\frac{1}{|I|(|I|-1)}\sum_{i,j \in I: i \ne j}d_J(C_i^T,C_j^T) $, where $d_J$ denotes the Jaccard index and $H \in [0,1]$ which is similar to the approach to measuring homogeneity utilized in \cite{chaney2018algorithmic}.
+Note that a higher $H$ indicates more homogeneity.
+Consumer $i$'s welfare will just be the average of realized utilities, to control for the effect of $T$, $W_i=\frac{1}{T}\sum_{n \in C_i^T} u_{in}$.
+\par
 \par

### 2019-08-30 create_graphs.R
Data Analysis with the datasets (#10)
* just a random push so i dont loose what im doing

* Add support for multiple N's

* add local moves graph into the loop

* think things work but damn gave my laptop a workout

* fix name of _rec graphs

* comment back in

* Code to create rings

@@ -2,8 +2,9 @@ library(ggplot2)
 library(dplyr)
 library(hrbrthemes)
 library(tidyverse)
-
+library(xtable)
 library(gridExtra)
+
 
 #' Calculate diversity mean w/ Confidence Intervals 
 #'
@@ -245,174 +246,272 @@ graph_stats_homo <- function(df, var){
   }
 }
 
-#WORKING_DIR <- "/Users/guyaridor/Desktop/recommender_systems/rec_sys_conf_paper/"
-#WORKING_DIR <- "/Users/ssikdar/Downloads/econ/bubbles/"
-WORKING_DIR <- "/Users/guyaridor/Desktop/ExAnteFilterBubble/"
-rec_data <- read.csv(paste(WORKING_DIR, "data/rec_data.csv", sep=""))
-rec_data <- rec_data %>% mutate(formatted_regime = ifelse(regime == "omni", "Omniscient", ifelse(regime == "no_rec", "No Rec", "Partial")))
-
-homogeneity <- read.csv(paste(WORKING_DIR, "data/homogeneity_data.csv", sep=""))
-homogeneity <- homogeneity%>% mutate(formatted_regime = ifelse(regime == "omni", "Omniscient", ifelse(regime == "no_rec", "No Rec", "Partial")))
 
 
-# CREATE GRAPHS
-use_hrbrthemes <- TRUE
-variables=list("rho", "beta", "sigma", "alpha")
-metrics=list("diversity", "welfare", "homogeneity")
-
-for(metric in metrics){
-  
-  print(metric)
-  for(variable in variables){
-    print(variable)
-    
-    file_name <- paste(WORKING_DIR,"figures/",variable,"_", metric, ".jpeg", sep="")
-    print(file_name)
-    
-    if (metric == "diversity"){
-      g <- graph_stats_diversity(get_stats_diversity(rec_data, variable), variable, use_hrbrthemes)
-      ggsave(filename=file_name, plot=g)
-    }
-    else if (metric == "welfare"){
-      #welfare
-      g <- graph_stats_welfare(get_stats_welfare(rec_data, variable), variable,  use_hrbrthemes)
-      ggsave(filename=file_name, plot=g)
-      
-      rec_file_name <- paste(WORKING_DIR, "figures/", variable, "_rec.jpeg", sep="")
-      print(rec_file_name)
-      g <- graph_stats_rec(get_stats_rec(filter(rec_data, regime == "partial"), variable), variable,  use_hrbrthemes)
[…]

### 2019-08-30 create_ring_graphs.R

@@ -1,0 +1,37 @@
+library(viridis)
+library(tidygraph)
+library(ggraph)
+library(igraph)
+
+N = 10
+ring <- create_ring(N)
+
+hop_distance <- function(i,j,N){min(abs(i-j), abs(j-i), abs(j-i-N), abs(i-j-N))}
+
+viridis_pallete <- viridis::viridis(N)
+viridis_pallete_rev <- rev(viridis_pallete)
+
+white_pallete <- c()
+for (i in 1:10){white_pallete <- c(white_pallete, "white")}
+
+png("circle1.png")
+plot(ring, 
+     vertex.size=10, vertex.label=NA, layout=layout_in_circle, vertex.color = white_pallete)
+dev.off()
+
+white_pallete[N] <- viridis_pallete[N]
+
+png("circle2.png")
+plot(ring, 
+     vertex.size=10, vertex.label=NA, layout=layout_in_circle, vertex.color = white_pallete)
+dev.off()
+
+new_pal <- c()
+for (i in 1:10){new_pal <- c(new_pal, viridis_pallete_rev[hop_distance(1,i,10)])}
+new_pal[1] <- viridis_pallete[N]
+png("circle3.png")
+plot(ring, vertex.size=10, vertex.label=NA, layout=layout_in_circle, vertex.color = new_pal)
+dev.off()
+
+
+

### 2019-08-30 simulations.jl
updated parameter values

@@ -381,31 +381,31 @@ end
 end
 
 #
-nr_pop = 50
+nr_pop = 100
 #
-nr_ind = 25
+nr_ind = 100
 #
 sigma_ibar = .1
 #
 rho_ibar = 0.0
 
-N_vals = [2000]
+N_vals = [200]
 
 T_vals = [20]
 
 # Covariance structure
-rho_vals = [0.1, 0.5, 0.9]
+rho_vals = [0.1, 0.3, 0.5, 0.7, 0.9]
 
 # utility idiosyncratic degree 
-beta_vals = [0, 1, 5, 10]
+beta_vals = [0, 0.4, 0.8, 1, 2, 5]
 
 # absolute risk aversion
-alpha_vals = [0, 1, 5, 10]
+alpha_vals = [0, 0.3, 0.6, 1, 5]
 
 # action of the time for random exploration
 epsilon_vals = [0.0]
 
-sigma_vals = [0.25, 1.0, 4.0]
+sigma_vals = [0.25, 0.5, 1.0, 2.0, 4.0]
 
 params = Iterators.product(N_vals, T_vals, rho_vals, beta_vals, sigma_vals, alpha_vals, epsilon_vals)
 

### 2019-08-30 simulations.jl
add rho = 0

@@ -394,7 +394,7 @@ N_vals = [200]
 T_vals = [20]
 
 # Covariance structure
-rho_vals = [0.1, 0.3, 0.5, 0.7, 0.9]
+rho_vals = [0, 0.1, 0.3, 0.5, 0.7, 0.9]
 
 # utility idiosyncratic degree 
 beta_vals = [0, 0.4, 0.8, 1, 2, 5]

### 2019-08-30 simulations.jl
fix type issue

@@ -75,7 +75,7 @@ CARA
 """
 
 @everywhere function certainty_equivalent(
-        alpha::Int64, 
+        alpha::Float64, 
         mu::Array{Float64,1},
         sig::Array{Float64,2}
     )
@@ -208,7 +208,7 @@ end
     T::Int64,
     N::Int64,
     Nset::Array{Int64,1},
-    alpha::Int64,
+    alpha::Float64,
     epsilon::Float64,
     beta::Int64
 )
@@ -259,7 +259,7 @@ end
 			T::Int64, 
 			N::Int64, 
 			Nset::Array{Int64,1}, 
-			alpha::Int64, 
+			alpha::Float64, 
 			epsilon::Float64)
 
     C_iT::Array{Int64,1} = []
@@ -299,7 +299,7 @@ end
         Sigma_V_i::Array{Float64,2},
         Sigma_V::Array{Float64,2},
         Sigma_V_ibar::Array{Float64,2},
-        alpha::Int64,
+        alpha::Float64,
         epsilon::Float64,
         seed::Float64
     )
@@ -318,7 +318,7 @@ end
     Sigma_V_i::Array{Float64,2},
     Sigma_V::Array{Float64,2},
     Sigma_V_ibar::Array{Float64,2},
-    alpha::Int64,
+    alpha::Float64,
     epsilon::Float64,
     seed::Int64
     )
@@ -394,13 +394,13 @@ N_vals = [200]
 T_vals = [20]
 
 # Covariance structure
-rho_vals = [0, 0.1, 0.3, 0.5, 0.7, 0.9]
+rho_vals = [0., 0.1, 0.3, 0.5, 0.7, 0.9]
 
 # utility idiosyncratic degree 
-beta_vals = [0, 0.4, 0.8, 1, 2, 5]
+beta_vals = [0., 0.4, 0.8, 1., 2., 5.]
 
 # absolute risk aversion
-alpha_vals = [0, 0.3, 0.6, 1, 5]
+alpha_vals = [0., 0.3, 0.6, 1., 5.]
 
 # action of the time for random exploration
 epsilon_vals = [0.0]

### 2019-08-30 simulations.jl
fix types

@@ -210,7 +210,7 @@ end
     Nset::Array{Int64,1},
     alpha::Float64,
     epsilon::Float64,
-    beta::Int64
+    beta::Float64
 )
     C_iT::Array{Int64,1} = []
     R_iT::Array{Int64,1} = []
@@ -294,7 +294,7 @@ end
         sigma::Float64,
         sigma_i::Float64, 
         sigma_ibar::Float64,
-        beta::Int64, 
+        beta::Float64, 
         nr_ind::Int64,
         Sigma_V_i::Array{Float64,2},
         Sigma_V::Array{Float64,2},
@@ -313,7 +313,7 @@ end
     sigma::Float64,
     sigma_i::Float64, 
     sigma_ibar::Float64,
-    beta::Int64, 
+    beta::Float64, 
     nr_ind::Int64,
     Sigma_V_i::Array{Float64,2},
     Sigma_V::Array{Float64,2},

### 2019-08-30 simulations.jl
store entire welfare array

@@ -330,7 +330,7 @@ end
     Nset = [ n for n=1:N]   # set of N items I = {1, ..., N}
 
     C_pop = Dict( "no_rec"  => zeros(Int64, nr_ind,T), "omni"  => zeros(Int64, nr_ind,T), "partial" => zeros(Int64, nr_ind,T))
-    W_pop = Dict( "no_rec"  => zeros(nr_ind), "omni"  => zeros(nr_ind), "partial" => zeros(nr_ind))
+    W_pop = Dict( "no_rec"  => zeros(nr_ind,T), "omni"  => zeros(nr_ind,T), "partial" => zeros(nr_ind,T))
     R_pop = Dict( "no_rec"  => zeros(nr_ind,T), "omni"  => zeros(nr_ind,T), "partial" => zeros(nr_ind,T))
 
     # V = (v_n) n in I aka: common value component v_n in vector form
@@ -358,26 +358,23 @@ end
         Sigma_U_i = Sigma_V_i + beta^2 * (Sigma_V)
         C_iT_no_rec = choice_ind(copy(U_i), copy(mu_U_i), copy(Sigma_U_i),T,N, Nset, alpha, epsilon)
         C_pop["no_rec"][it_ind,:] = C_iT_no_rec
-        w_val = w_fun(C_iT_no_rec, U_i, T)
-        W_pop["no_rec"][it_ind] = w_val
+        W_pop["no_rec"][it_ind,:] = copy(U_i[C_iT_no_rec])
 
         
         ## OMNISCIENT CASE
         C_iT = choice_omni(copy(U_i),T,N, Nset)
-        C_pop["omni"][it_ind,:] = copy(C_iT)
-        w_val = w_fun(C_iT, U_i, T)
-        W_pop["omni"][it_ind] = w_val
+        C_pop["omni"][it_ind,:] = C_iT
+        W_pop["omni"][it_ind,:] = copy(U_i[C_iT])
 
  
         ## PARTIAL REC Case
         C_iT_partial, R_iT = choice_part(copy(V_i), copy(mu_V_i), copy(Sigma_V_i), copy(V), T, N, Nset, alpha, epsilon, beta)
         C_pop["partial"][it_ind,:] = C_iT_partial
-        w_val = w_fun(C_iT_partial, U_i, T)
-        W_pop["partial"][it_ind] = w_val
+        W_pop["partial"][it_ind,:] = copy(U_i[C_iT_partial])
[…]

### 2019-08-30 simulations.jl
fix typo

@@ -374,7 +374,7 @@ end
         R_pop["partial"][it_ind,:] = R_iT
     end
 
-    return Dict( "Consumption" => C_pop, "Welfare" => W_pop, "Welfare" "Rec" => R_pop )
+    return Dict( "Consumption" => C_pop, "Welfare" => W_pop, "Rec" => R_pop )
 end
 
 #

### 2019-08-30 dump_to_dataset.py
update welfare calculation cc @ssikdar

@@ -26,7 +26,7 @@ def div_fun(CiT, T, N):
         for j in range(len(CiT)):
             if i == j: continue
             div_score = div_score + d(CiT[i],CiT[j], N)#euclidean( iota(CiT[i], N), iota(CiT[j], N) )
-    return div_score*(1.0/(T*(T-1)))
+    return div_score*(1.0/(T*(T-1))) / N
 
 def homogeneity(C1, C2, type_sim="jaccard"):
     if type_sim == "jaccard":
@@ -54,6 +54,50 @@ def parse_pickle_key(key):
     }
     return dat
 
+
+INDIVIDUAL_FIELD_NAMES =['pop_idx', 'regime', 'rho', 'beta', 'epsilon', 'alpha', 'N', 'T', 'sigma', 't', 'follow_recommendation', 'consumption_dist', 'cur_utility', 'cumulative_utility', 'utility_difference']
+with open(WORKING_DIR + 'time_path.csv', 'w') as rec_csv:
+    data_writer = csv.DictWriter(rec_csv, fieldnames=INDIVIDUAL_FIELD_NAMES)
+    data_writer.writeheader()
+    for key, value in df.items():
+        for pop_idx in range(len(value)):
+            dat = parse_pickle_key(key)
+            N = dat['N']
+            dat['pop_idx'] = pop_idx
+            cur = value[pop_idx]
+            consumption = cur['Consumption']
+            welfare = cur['Welfare']
+            for policy in rec_policy_keys:
+                consumption_arr = np.array(consumption[policy])
+                welfare_arr = np.array(welfare[policy])
+                dat['regime'] = policy
+                if policy == PARTIAL:
+                    follow_rec_arr = np.array(cur['Rec'][policy])
+                for indiv_idx in range(len(consumption_arr[0,:])):
+                    c1 = consumption_arr[:,indiv_idx]
+                    welf = welfare_arr[:,indiv_idx]
+                    prev_consumption = None
+                    prev_welfare = None
+                    rec = None
+                    if policy == PARTIAL:
+                        rec = follow_rec_arr[:,indiv_idx]
+                    cumulative_welfare = 0.0
+                    for t in range(len(c1)):
+                        dat['t'] = t
+                        dat['cur_utility'] = welf[t]
+                        cumulative_welfare += welf[t]
+                        dat['cumulative_utility'] = cumulative_welfare
+                        if prev_consumption != None:
[…]

### 2019-09-10 dump_to_dataset.py
first push

@@ -66,80 +66,81 @@ def parse_pickle_key(key):
     return dat
 
 
-INDIVIDUAL_FIELD_NAMES =['pop_idx', 'indiv_idx', 'regime', 'welfare', 'diversity_score', 'rho', 'beta', 'epsilon', 'follow_recommendation', 'N', 'T', 'sigma', 'alpha', 'nr_pop', 'nr_ind']
-with open(WORKING_DIR + 'rec_data.csv', 'w') as rec_csv:
-    data_writer = csv.DictWriter(rec_csv, fieldnames=INDIVIDUAL_FIELD_NAMES)
-    data_writer.writeheader()
-    for file in data_files:
-        print(file)
-        print("@ {}".format(datetime.datetime.now()))
-        df = None
-        with open(file, 'r') as fp:
-            df = json.load(fp)
-
-        for key, value in df.items():
-            for pop_idx in range(len(value)):
-                dat = parse_pickle_key(key)
-                T = dat['T']
-                N = dat['N']
-                dat['pop_idx'] = pop_idx
-                cur = value[pop_idx]
-                welfare = cur['Welfare']
-                consumption = cur['Consumption']
-                for policy in rec_policy_keys:
-                    dat['regime'] = policy
-                    for indiv_idx in range(len(welfare[policy])):
-                        dat['indiv_idx'] = indiv_idx
-                        welfare_arr = np.array(welfare[policy])
-                        dat['welfare'] = np.sum(welfare_arr[:,indiv_idx]) / T
-                        consumption_arr = np.array(consumption[policy])
-                        dat['diversity_score'] = div_fun(consumption_arr[:,indiv_idx], T, N)
-                        dat['follow_recommendation'] = False
-                        if policy == PARTIAL:
-                            follow_rec_arr = np.array(cur['Rec'][policy])
-                            dat['follow_recommendation'] = follow_rec(consumption_arr[:,indiv_idx], follow_rec_arr[:,indiv_idx], T, N)
-                        data_writer.writerow(copy(dat))
-
-print("STARTING HOMOGENEITY")
-INDIVIDUAL_FIELD_NAMES =['pop_idx', 'regime', 'rho', 'beta', 'N', 'T', 'sigma', 'jaccard', 'beta', 'alpha', 'epsilon', 'nr_pop', 'nr_ind']
-with open(WORKING_DIR + 'homogeneity_data.csv', 'w') as rec_csv:
[…]

### 2019-09-10 dump_to_dataset.py
consumption dist average

@@ -166,9 +166,17 @@ with open(WORKING_DIR + 'time_path.csv', 'w') as rec_csv:
                         dat['t'] = t
 
                         # consumption dist average    
+                        c1 =  np.mean(consumption_arr[t, :])
+                        if t != 0:
+                            c2 = np.mean(consumption_arr[t-1, :]) 
+                            dat['consumption_dist'] = d(c2, c1, N)
+                        else:
+                            dat['consumption_dist'] = 0
                         
                         # cumulative welfare avg
                         dat['average_cumulative_utility'] = np.mean(welfare_arr[t, :])
+
+                        # instantaneous (at time t) welfare avg
 
                         # local move avg 
 

### 2019-09-10 dump_to_dataset.py
local move

@@ -136,7 +136,7 @@ def parse_pickle_key(key):
 #                    data_writer.writerow(cur_dat)
 
 print("STARTING TIME PATH")
-INDIVIDUAL_FIELD_NAMES =['pop_idx', 'regime', 'rho', 'beta', 'epsilon', 'alpha', 'N', 'T', 'sigma', 't', 'follow_recommendation', 'consumption_dist', 'cur_utility', 'average_cumulative_utility', 'utility_difference']
+INDIVIDUAL_FIELD_NAMES =['pop_idx', 'regime', 'rho', 'beta', 'epsilon', 'alpha', 'N', 'T', 'sigma', 't', 'follow_recommendation', 'consumption_dist', 'cur_utility', 'average_cumulative_utility', 'utility_difference', 'local_move']
 with open(WORKING_DIR + 'time_path.csv', 'w') as rec_csv:
     data_writer = csv.DictWriter(rec_csv, fieldnames=INDIVIDUAL_FIELD_NAMES)
     data_writer.writeheader()
@@ -179,30 +179,7 @@ with open(WORKING_DIR + 'time_path.csv', 'w') as rec_csv:
                         # instantaneous (at time t) welfare avg
 
                         # local move avg 
-
+                        dat['local_move'] =  int(dat['consumption_dist'] < (dat['N'] * 0.05))
+                        
                         cur_dat = copy(dat)
                         data_writer.writerow(cur_dat)
-
-                    #for indiv_idx in range(len(consumption_arr[0,:])): 
-                    #    c1 = consumption_arr[:,indiv_idx]
-                    #    prev_consumption = None
-                    #    prev_welfare = None
-                    #    rec = None
-                    #    if policy == PARTIAL:
-                    #        rec = follow_rec_arr[:,indiv_idx]
-                    #    cumulative_welfare = 0.0
-                    #    for t in range(len(c1)):
-                    #        dat['t'] = t
-                    #        dat['cur_utility'] = welf[t]
-                    #        cumulative_welfare += welf[t]
-                    #        dat['cumulative_utility'] = cumulative_welfare
-                    #        if prev_consumption != None:
-                    #            dat['consumption_dist'] = d(prev_consumption, c1[t], N)
-                    #            dat['utility_difference'] = welf[t] - prev_welfare
-                    #        prev_consumption = c1[t]
-                    #        prev_welfare = welf[t]
-                    #        dat['follow_recommendation'] = False
-                    #        if policy == PARTIAL:
[…]

### 2019-09-10 dump_to_dataset.py
remove comments

@@ -66,74 +66,74 @@ def parse_pickle_key(key):
     return dat
 
 
-#INDIVIDUAL_FIELD_NAMES =['pop_idx', 'indiv_idx', 'regime', 'welfare', 'diversity_score', 'rho', 'beta', 'epsilon', 'follow_recommendation', 'N', 'T', 'sigma', 'alpha', 'nr_pop', 'nr_ind']
-#with open(WORKING_DIR + 'rec_data.csv', 'w') as rec_csv:
-#    data_writer = csv.DictWriter(rec_csv, fieldnames=INDIVIDUAL_FIELD_NAMES)
-#    data_writer.writeheader()
-#    for file in data_files:
-#        print(file)
-#        print("@ {}".format(datetime.datetime.now()))
-#        df = None
-#        with open(file, 'r') as fp:
-#            df = json.load(fp)
-#
-#        for key, value in df.items():
-#            for pop_idx in range(len(value)):
-#                dat = parse_pickle_key(key)
-#                T = dat['T']
-#                N = dat['N']
-#                dat['pop_idx'] = pop_idx
-#                cur = value[pop_idx]
-#                welfare = cur['Welfare']
-#                consumption = cur['Consumption']
-#                for policy in rec_policy_keys:
-#                    dat['regime'] = policy
-#                    for indiv_idx in range(len(welfare[policy])):
-#                        dat['indiv_idx'] = indiv_idx
-#                        welfare_arr = np.array(welfare[policy])
-#                        dat['welfare'] = np.sum(welfare_arr[:,indiv_idx]) / T
-#                        consumption_arr = np.array(consumption[policy])
-#                        dat['diversity_score'] = div_fun(consumption_arr[:,indiv_idx], T, N)
-#                        dat['follow_recommendation'] = False
-#                        if policy == PARTIAL:
-#                            follow_rec_arr = np.array(cur['Rec'][policy])
-#                            dat['follow_recommendation'] = follow_rec(consumption_arr[:,indiv_idx], follow_rec_arr[:,indiv_idx], T, N)
-#                        data_writer.writerow(copy(dat))
-#
-#print("STARTING HOMOGENEITY")
[…]

### 2019-09-10 dump_to_dataset.py
instantaneous_welfare_average

@@ -136,7 +136,7 @@ with open(WORKING_DIR + 'homogeneity_data.csv', 'w') as rec_csv:
                     data_writer.writerow(cur_dat)
 
 print("STARTING TIME PATH")
-INDIVIDUAL_FIELD_NAMES =['pop_idx', 'regime', 'rho', 'beta', 'epsilon', 'alpha', 'N', 'T', 'sigma', 't', 'follow_recommendation', 'consumption_dist', 'cur_utility', 'average_cumulative_utility', 'utility_difference', 'local_move']
+INDIVIDUAL_FIELD_NAMES =['pop_idx', 'regime', 'rho', 'beta', 'epsilon', 'alpha', 'N', 'T', 'sigma', 't', 'follow_recommendation', 'consumption_dist', 'cur_utility', 'average_cumulative_utility', 'utility_difference', 'local_move', 'instantaneous_welfare_average']
 with open(WORKING_DIR + 'time_path.csv', 'w') as rec_csv:
     data_writer = csv.DictWriter(rec_csv, fieldnames=INDIVIDUAL_FIELD_NAMES)
     data_writer.writeheader()
@@ -162,6 +162,7 @@ with open(WORKING_DIR + 'time_path.csv', 'w') as rec_csv:
                     if policy == PARTIAL:
                         follow_rec_arr = np.array(cur['Rec'][policy])
                     
+                    cum_welfare = 0
                     for t in range(int(dat['T'])) :
                         dat['t'] = t
 
@@ -177,6 +178,8 @@ with open(WORKING_DIR + 'time_path.csv', 'w') as rec_csv:
                         dat['average_cumulative_utility'] = np.mean(welfare_arr[t, :])
 
                         # instantaneous (at time t) welfare avg
+                        cum_welfare += np.mean(welfare_arr[t, :])
+                        dat['instantaneous_welfare_average'] = float(cum_welfare)/dat['T']
 
                         # local move avg 
                         dat['local_move'] =  int(dat['consumption_dist'] < (dat['N'] * 0.05))

### 2019-09-11 dump_to_dataset.py
fixups

@@ -65,7 +65,7 @@ def parse_pickle_key(key):
     }
     return dat
 
-
+'''
 INDIVIDUAL_FIELD_NAMES =['pop_idx', 'indiv_idx', 'regime', 'welfare', 'diversity_score', 'rho', 'beta', 'epsilon', 'follow_recommendation', 'N', 'T', 'sigma', 'alpha', 'nr_pop', 'nr_ind']
 with open(WORKING_DIR + 'rec_data.csv', 'w') as rec_csv:
     data_writer = csv.DictWriter(rec_csv, fieldnames=INDIVIDUAL_FIELD_NAMES)
@@ -134,13 +134,12 @@ with open(WORKING_DIR + 'homogeneity_data.csv', 'w') as rec_csv:
                     dat['jaccard'] = tot
                     cur_dat = copy(dat)
                     data_writer.writerow(cur_dat)
-
+'''
 print("STARTING TIME PATH")
-INDIVIDUAL_FIELD_NAMES =['pop_idx', 'regime', 'rho', 'beta', 'epsilon', 'alpha', 'N', 'T', 'sigma', 't', 'follow_recommendation', 'consumption_dist', 'cur_utility', 'average_cumulative_utility', 'utility_difference', 'local_move', 'instantaneous_welfare_average']
+INDIVIDUAL_FIELD_NAMES =['pop_idx', 'regime', 'rho', 'beta', 'epsilon', 'alpha', 'N', 'T', 'sigma', 't', 'follow_recommendation', 'consumption_dist', 'cur_utility', 'average_cumulative_utility', 'utility_difference', 'local_move_05', 'local_move_025', 'local_move_10', 'instantaneous_welfare_average']
 with open(WORKING_DIR + 'time_path.csv', 'w') as rec_csv:
     data_writer = csv.DictWriter(rec_csv, fieldnames=INDIVIDUAL_FIELD_NAMES)
     data_writer.writeheader()
-    data_files = ["/media/IntentMedia/rec_data/sim_results/new_sim_34.json"]
     for file in data_files:
         print(file)
         print("@ {}".format(datetime.datetime.now()))
@@ -166,23 +165,25 @@ with open(WORKING_DIR + 'time_path.csv', 'w') as rec_csv:
                     for t in range(int(dat['T'])) :
                         dat['t'] = t
 
-                        # consumption dist average    
-                        c1 =  np.mean(consumption_arr[t, :])
+                        # consumption dist average
+                        l = len(consumption_arr[t, :])
[…]

### 2019-09-11 paper/filter_bubbles.tex

--- 2019-08-30 paper/ex_ante_rec_sys.tex
@@
 We simulate over $100$ populations with $100$ consumers.
-\section{Results}
+We average over the consumers in a population to get a representative consumer and then report results where a single data point is over the set of representative consumers.
+
+\section{User Behavior}
+
+\subsection{Local Search}
+
+We explore the degree to which users engage in \textit{local} consumption and how this varies with recommendation.
+We isolate the two main mechanisms in our model that can generate local search without recommendation.
+The first is by exploiting the correlation structure
+
+\subsection{User Welfare}
+
+\subsection{Diversity}
+
+\subsection{Homogenization}
+
+\subsection{Effectiveness of Recommendation}
 
@@
 \end{table}
-\begin{figure*}[!htbp]
-    \centering
-    \begin{minipage}[b]{0.5\textwidth}
-        \centering
-        \includegraphics[width=0.9\linewidth, height=1.5in]{local_moves_25}
-        \caption{Extent of ``Local"  Consumption}
-        \label{fig:local_consumption}
-    \end{minipage}%
-    ~ 
-    \begin{minipage}[b]{0.5\textwidth}
-        \centering
-        \includegraphics[width=0.9\linewidth, height=1.5in]{welfare_homo_combo}
-        \caption{Welfare and Homogeneity}
-        \label{fig:beta_vary}
-    \end{minipage}
-    ~ \\
-     \begin{minipage}{0.48\textwidth}
-     \centering
-     \includegraphics[width=0.9\linewidth, height=1.5in]{rec_obedience_25}
-     \caption{Recommendation Effectiveness}\label{fig:rec_obey}
-   \end{minipage}\hfill
-   ~
-   \begin{minipage}{0.48\textwidth}
-     \centering
-     \includegraphics[width=0.9\linewidth, height=1.5in]{Diversity_vs_Welfare_No_Recommendation}
-     \caption{Diversity vs Welfare under No Recommendation}\label{fig:diversity_welfare_no_rec}
-   \end{minipage}
-\end{figure*}
 

### 2019-09-12 create_graphs.R
updates

@@ -115,13 +115,13 @@ get_stats_homogeneity <- function(df, var){
     df <- df %>% group_by(formatted_regime, alpha)
   }
   df <- df %>%
-      mutate(jaccard_mean = mean(pop_jaccard_avg),
-             jaccard_sd = sd(pop_jaccard_avg),
-             jaccard_n = n()) %>%
-      mutate(jaccard_se =  jaccard_sd/ sqrt(jaccard_n),
-             lower_ci_jaccard = jaccard_mean - qt(1 - (0.05 / 2), jaccard_n - 1) * jaccard_se,
-             upper_ci_jaccard = jaccard_mean + qt(1 - (0.05 / 2), jaccard_n - 1) * jaccard_se) %>% 
-      slice(1)
+    mutate(jaccard_mean = mean(pop_jaccard_avg),
+           jaccard_sd = sd(pop_jaccard_avg),
+           jaccard_n = n()) %>%
+    mutate(jaccard_se =  jaccard_sd/ sqrt(jaccard_n),
+           lower_ci_jaccard = jaccard_mean - qt(1 - (0.05 / 2), jaccard_n - 1) * jaccard_se,
+           upper_ci_jaccard = jaccard_mean + qt(1 - (0.05 / 2), jaccard_n - 1) * jaccard_se) %>% 
+    slice(1)
   return(df)
 }
 
@@ -136,15 +136,15 @@ graph_stats_diversity <- function(df, var, N, T_val, use_hrbrthemes){
   # graph w/ themes
   
   g <- ggplot(df, aes_string(x=var, y="diversity_mean")) +
-      geom_line(aes(colour=formatted_regime)) +
-      geom_errorbar(aes(ymin=lower_ci_diversity, ymax=upper_ci_diveristy), width=.02,
-                    position=position_dodge(.9)) + 
-      labs(x=var, y="diversity",
-           title=paste("N =", N, "T =", T_val, "Diversity",sep=" "))
+    geom_line(aes(colour=formatted_regime)) +
+    geom_errorbar(aes(ymin=lower_ci_diversity, ymax=upper_ci_diveristy), width=.02,
+                  position=position_dodge(.9)) + 
+    labs(x=var, y="diversity",
+         title=paste("N =", N, "T =", T_val, "Diversity",sep=" "))
   if(use_hrbrthemes){
     g <- g + theme_ipsum_rc()
   }
-    return(g)
+  return(g)
 }
 
 #' Graph rec mean as function of `var`
[…]

### 2019-09-12 simulations.jl

@@ -426,9 +426,6 @@ for (N, T, rho, beta, sigma, alpha, epsilon) in params
     Sigma_V_i = cov_mat_fun(sigma, rho, N)
     Sigma_V = cov_mat_fun(sigma,rho,N)
 
-    if beta != 0
-        Sigma_V = Sigma_V / beta^2
-    end
     Sigma_V_ibar = cov_mat_fun(sigma_ibar,rho_ibar,N)
     global sim_results[(N, T, rho, beta, sigma, alpha, epsilon, nr_pop, nr_ind)] = @sync @distributed vcat for i= 1:nr_pop
         simulate(N, T,sigma, sigma_i, sigma_ibar, beta, nr_ind, Sigma_V_i,  Sigma_V,  Sigma_V_ibar,  alpha, epsilon, i)

### 2019-09-12 dump_to_dataset.py
add back otehr stuff

@@ -52,6 +52,7 @@ def follow_rec(Ci, Rec, N, T):
     return s / T
 
 def parse_pickle_key(key):
+    print(key)
     key = eval(key)
     dat = {
         'N': float(key[0]),
@@ -64,8 +65,59 @@ def parse_pickle_key(key):
         #'pop_idx': pop_idx
     }
     return dat
+             
+print("STARTING TIME PATH")
+INDIVIDUAL_FIELD_NAMES =['pop_idx', 'regime', 'rho', 'beta', 'epsilon', 'alpha', 'N', 'T', 'sigma', 't', 'follow_recommendation', 'consumption_dist', 'cur_utility', 'average_cumulative_utility', 'utility_difference', 'local_move_05', 'local_move_025', 'local_move_10', 'instantaneous_welfare_average']
+with open(WORKING_DIR + 'time_path.csv', 'w') as rec_csv:
+    data_writer = csv.DictWriter(rec_csv, fieldnames=INDIVIDUAL_FIELD_NAMES)
+    data_writer.writeheader()
+    for file in data_files:
+        print(file)
+        print("@ {}".format(datetime.datetime.now()))
+        df = None
+        with open(file, 'r') as fp:
+            df = json.load(fp)
+        for key, value in df.items():
+            for pop_idx in range(len(value)):
+                dat = parse_pickle_key(key)
+                N = dat['N']
+                dat['pop_idx'] = pop_idx
+                cur = value[pop_idx]
+                consumption = cur['Consumption']
+                welfare = cur['Welfare']
+                for policy in rec_policy_keys:
+                    consumption_arr = np.array(consumption[policy])
+                    welfare_arr = np.array(welfare[policy])
+                    dat['regime'] = policy
+                    if policy == PARTIAL:
+                        follow_rec_arr = np.array(cur['Rec'][policy])
+                    
+                    cum_welfare = 0
+                    for t in range(int(dat['T'])) :
+                        dat['t'] = t
 
-'''
+                        # consumption dist average
+                        l = len(consumption_arr[t, :])
+                        if t != 0:
+                            distance = [d(consumption_arr[t, cur], consumption_arr[t-1, cur], N) for cur in range(l)]
+                            dat['mean_consumption_dist'] = np.mean(distance)
+                            dat['median_consumption_dist'] = np.median(distance)
+                            dat['sd_consumption_dist'] = np.std(distance)
[…]

### 2019-09-12 dump_to_dataset.py
add back other stuff

@@ -67,7 +67,7 @@ def parse_pickle_key(key):
     return dat
              
 print("STARTING TIME PATH")
-INDIVIDUAL_FIELD_NAMES =['pop_idx', 'regime', 'rho', 'beta', 'epsilon', 'alpha', 'N', 'T', 'sigma', 't', 'follow_recommendation', 'consumption_dist', 'cur_utility', 'average_cumulative_utility', 'utility_difference', 'local_move_05', 'local_move_025', 'local_move_10', 'instantaneous_welfare_average']
+INDIVIDUAL_FIELD_NAMES =['pop_idx', 'regime', 'rho', 'beta', 'epsilon', 'alpha', 'N', 'T', 'sigma', 't', 'follow_recommendation', 'mean_consumption_dist', 'median_consumption_dist', 'sd_consumption_dist', 'cur_utility', 'average_cumulative_utility', 'utility_difference', 'local_move_05', 'local_move_025', 'local_move_10', 'instantaneous_welfare_average']
 with open(WORKING_DIR + 'time_path.csv', 'w') as rec_csv:
     data_writer = csv.DictWriter(rec_csv, fieldnames=INDIVIDUAL_FIELD_NAMES)
     data_writer.writeheader()

### 2019-09-12 dump_to_dataset.py
remove log line

@@ -52,7 +52,6 @@ def follow_rec(Ci, Rec, N, T):
     return s / T
 
 def parse_pickle_key(key):
-    print(key)
     key = eval(key)
     dat = {
         'N': float(key[0]),

### 2019-09-12 paper/dump_to_dataset.py
clean up repo

@@ -1,140 +1,0 @@
-import pickle
-import csv
-from copy import copy
-
-from scipy.spatial.distance import jaccard
-
-OMNI = 'rec'
-PARTIAL = 'partial'
-NO_REC = 'no_rec'
-rec_policy_keys = [OMNI, PARTIAL, NO_REC]
-WORKING_DIR = ''
-with open(WORKING_DIR + 'sim_results_t_25.p', 'rb') as fp:
-    df = pickle.load(fp)
-
-def d(i,j, N):
-    return min(abs(i-j), abs(j-i), abs(j-i-N), abs(i-j-N))
-
-def div_fun(CiT, T, N):
-    div_score = 0.0
-    for i in range(len(CiT)):
-        for j in range(len(CiT)):
-            if i == j: continue
-            div_score = div_score + d(CiT[i],CiT[j], N)
-    return div_score*(1.0/(T*(T-1)))
-
-def homogeneity(C1, C2, type_sim="jaccard"):
-    if type_sim == "jaccard":
-        return 1 - jaccard(C1, C2)
-    return None
-
-def follow_rec(Ci, Rec, T, N):
-    s = 0.0
-    for i in range(len(Ci)):
-        if Ci[i] == Rec[i]:
-            s += 1
-    return s / len(Ci)
-
-
-INDIVIDUAL_FIELD_NAMES =['pop_idx', 'indiv_idx', 'regime', 'welfare', 'diversity_score', 'rho', 'beta', 'follow_recommendation', 'N', 'T', 'sigma']
-with open(WORKING_DIR + 'rec_data_t_25.csv', 'w') as rec_csv:
-    data_writer = csv.DictWriter(rec_csv, fieldnames=INDIVIDUAL_FIELD_NAMES)
-    data_writer.writeheader()
-    for key, value in df.items():
-        for pop_idx in range(len(value)):
-            N = key[0]
-            T = key[1]
-            dat = {
-                'pop_idx': pop_idx,
-                'N': N,
-                'T': T,
-                'rho': key[2],
-                'beta': key[3],
-                'sigma': key[4]
-            }
-            cur = value[pop_idx]
-            welfare = cur['Welfare']
-            consumption = cur['Consumption']
-            for policy in rec_policy_keys:
-            	dat['regime'] = policy
-            	for indiv_idx in range(len(welfare[policy])):
-                    dat['indiv_idx'] = indiv_idx
-                    dat['welfare'] = welfare[policy][indiv_idx]
[…]

### 2019-09-12 paper/plots.R

@@ -1,357 +1,0 @@
-library(ggplot2)
-library(dplyr)
-library(hrbrthemes)
-library(tidyverse)
-
-library(gridExtra)
-
-
-get_stats_diversity <- function(rec_data, var){
-  df <- rec_data
-  if (var == "rho") { 
-    df <- df %>% group_by(regime, rho) 
-  } else if (var == "beta") {
-    df <- df %>% group_by(regime, beta) 
-  } else if (var == "sigma") {
-    df <- rec_data %>% group_by(regime, sigma) 
-  }
-  
-  df <- df %>% 
-    mutate(diversity_mean = mean(diversity_score),
-           diversity_sd = sd(diversity_score),
-           diversity_n = n()) %>%
-    mutate(diversity_se =  diversity_sd/ sqrt(diversity_n),
-           lower_ci_diversity = diversity_mean - qt(1 - (0.05 / 2), diversity_n - 1) * diversity_se,
-           upper_ci_diveristy = diversity_mean + qt(1 - (0.05 / 2), diversity_n - 1) * diversity_se) %>% 
-    slice(1)
-  return(df)
-  
-}
-
-get_stats_rec <- function(rec_data, var){
-  rec_data$follow_recommendation <- as.numeric(levels(rec_data$follow_recommendation)[rec_data$follow_recommendation])
-  
-  df <- rec_data
-  
-  if (var == "rho") { # there must be a better way
-    df <- df %>% group_by(rho) 
-  } else if (var == "beta") {
-    df <- df %>% group_by(beta) 
-  } else if (var == "sigma") {
-    df <- df %>% group_by(sigma) 
-  }
-  
-  df <- df %>% 
-    mutate(rec_mean = mean(follow_recommendation),
-           rec_sd = sd(follow_recommendation),
-           rec_n = n()) %>%
-    mutate(rec_se =  rec_sd/ sqrt(rec_n),
-           lower_ci_rec = rec_mean - qt(1 - (0.05 / 2), rec_n - 1) * rec_se,
-           upper_ci_rec = rec_mean + qt(1 - (0.05 / 2), rec_n - 1) * rec_se) %>% 
-    slice(1)
-  return(df)
-}
-
-# https://stackoverflow.com/questions/22309285/how-to-use-a-variable-to-specify-column-name-in-ggplot
-# graph w/ themes
-graph_stats_diversity <- function(df, var, use_hrbrthemes){
-  T_val <- df$T[1]
[…]

### 2019-09-12 paper/simulations.py

@@ -1,212 +1,0 @@
-import pandas as pd
-import numpy as np
-import pickle
-import cProfile
-from copy import copy
-
-from joblib import Parallel, delayed
-
-import itertools
-from scipy import stats
-from scipy.stats import beta
-
-
-import pandas as pd
-import numpy as np
-import pickle
-import cProfile
-import numba
-
-from joblib import Parallel, delayed
-
-import itertools
-from scipy import stats
-from scipy.stats import beta
-
-WRITE_FILE = 'sim_results_t_25.p'
-OMNI = 'rec'
-PARTIAL = 'partial'
-NO_REC = 'no_rec'
-
-
-###SETUP FUNCTIONS
-def d(i,j, N):
-    return min(abs(i-j), abs(j-i), abs(j-i-N), abs(i-j-N))
-
-def cov_mat_fun(sigm, rh, N):
-    cov_mat = np.ones((N,N))
-    for i in range(0,N):
-        for j in range(0,N):
-            cov_mat[i,j] = rh**d(i,j, N)
-    cov_mat = sigm * cov_mat
-    return cov_mat
-
-
-
-### Welfare Functions - Statistic Calculation Functions
-
-
-def w_fun(CiT,Ui):
-    w_score = 0.0
-    for i in range(len(CiT)):
-        w_score = w_score + Ui[CiT[i]]
-    return w_score*(T**(-1))
-
-
-## Bayesian updating
-@numba.jit
-def inv_nla_jit(A):
-  return np.linalg.inv(A)
-
-def update_Ui(Cit,Ui,mu_Ui, Sigma_Ui, Nset):
-    x1 = Cit
-    x2 = [n for n in Nset if n not in Cit]
-    Nit = [n for n in Nset if n not in Cit]
-    mu1 = np.array([mu_Ui[n] for n in x1]).reshape((1,len(x1)))
-    mu2 = np.array([mu_Ui[n] for n in x2]).reshape((1,len(x2)))
-    Sigma11 = np.ones((len(x1),len(x1)))
-    Sigma21 = np.ones((len(x2),len(x1)))
-    for i in range(len(Cit)):
-        for j in range(len(Cit)):
-            Sigma11[i,j] = Sigma_Ui[Cit[i],Cit[j]]
-    for i in range(len(Cit)):
[…]

### 2019-09-13 analysis/create_graphs.R
add latest figures

@@ -20,8 +20,8 @@ get_stats_diversity <- function(df, var){
     df <- df %>% group_by(formatted_regime, beta) 
   } else if (var == "sigma") {
     df <- df %>% group_by(formatted_regime, sigma) 
-  } else if (var == "alpha") {
-    df <- df %>% group_by(formatted_regime, alpha)
+  } else if (var == "gamma") {
+    df <- df %>% group_by(formatted_regime, gamma)
   }
   
   df <- df %>% 
@@ -50,8 +50,8 @@ get_stats_rec <- function(df, var){
     df <- df %>% group_by(formatted_regime, beta) 
   } else if (var == "sigma") {
     df <- df %>% group_by(formatted_regime, sigma) 
-  } else if (var == "alpha") {
-    df <- df %>% group_by(formatted_regime, alpha)
+  } else if (var == "gamma") {
+    df <- df %>% group_by(formatted_regime, gamma)
   }
   
   df <- df %>% 
@@ -80,8 +80,8 @@ get_stats_welfare <- function(df, var){
     df <- df %>% group_by(formatted_regime, beta) 
   } else if (var == "sigma") {
     df <- df %>% group_by(formatted_regime, sigma) 
-  } else if (var == "alpha") {
-    df <- df %>% group_by(formatted_regime, alpha)
+  } else if (var == "gamma") {
+    df <- df %>% group_by(formatted_regime, gamma)
   }
   
   df <- df %>%
@@ -111,8 +111,8 @@ get_stats_homogeneity <- function(df, var){
     df <- df %>% group_by(formatted_regime, beta) 
   } else if (var == "sigma") {
     df <- df %>% group_by(formatted_regime, sigma) 
-  } else if (var == "alpha") {
-    df <- df %>% group_by(formatted_regime, alpha)
+  } else if (var == "gamma") {
+    df <- df %>% group_by(formatted_regime, gamma)
   }
   df <- df %>%
     mutate(jaccard_mean = mean(pop_jaccard_avg),
@@ -141,9 +141,7 @@ graph_stats_diversity <- function(df, var, N, T_val, use_hrbrthemes){
                   position=position_dodge(.9)) + 
     labs(x=var, y="diversity",
          title=paste("N =", N, "T =", T_val, "Diversity",sep=" "))
-  if(use_hrbrthemes){
-    g <- g + theme_ipsum_rc()
-  }
[…]

### 2019-09-14 paper/filter_bubbles.tex
include new results and update analysis + sim code

@@
 \usepackage{float}
+\newtheorem{finding}{Finding}
+
+\usepackage{xspace}
+\newcommand{\OMIT}[1]{}
 %
@@
 
-\subsection{Local Search}
+\subsection{Local Consumption}
 
-We explore the degree to which users engage in \textit{local} consumption and how this varies with recommendation.
-We isolate the two main mechanisms in our model that can generate local search without recommendation.
-The first is by exploiting the correlation structure
+\begin{figure}
+\includegraphics[scale=0.4]{figures/consumption_dist_N_200T_20no_correlation}
+\caption{Consumption path with no correlation between utilities}
+\label{fig:no_correlation_consumption_path}
+\end{figure}
 
-\subsection{User Welfare}
+\begin{figure}
+\includegraphics[scale=0.1]{figures/consumption_dist_N_200T_20_overall}
+\caption{Consumption path between Recommendation Regimes}
+\label{fig:consumption_path_between_regimes}
+\end{figure}
 
-\subsection{Diversity}
+We characterize ``filter bubble" effects in our model as the degree to which consumers engage in \textit{local consumption}.
+We define local consumption in terms of the average consumption distance between the items consumed by the representative consumer at time $t-1$ and $t$.
+We compare the average movement within the product space across different recommendation regimes and associate lower average distance with more local consumption.
 
-\subsection{Homogenization}
+\begin{finding}\label{finding_local_consumption}
+The impact of recommendation on local consumption:
+\begin{enumerate}
+\item When $\rho = 0$, there is no difference in consumption distance between the three recommendation regimes.
+\item When $\rho > 0$, no recommendation induces more local consumption than both partial and omniscient recommendation.
+This effect is amplified as $\rho$ increases as well as when consumers are more risk averse ($\gamma$ increases)
+\end{enumerate}
+\end{finding}
+
+First, Figure \ref{fig:no_correlation_consumption_path} shows that, when $\rho = 0$, there is no difference in consumption distance between the three regimes.
+This is due to the fact that when $\rho = 0$, there is no reason that items that are close in the product space should have similar utilities and so the optimal consumption path does not depend on the similarity of the items.
+However this also means that consumers in the no recommendation regime do not learn anything about neighboring goods and so there is limited path-dependence in consumption.
+
+Figure \ref{fig:consumption_path_between_regimes} shows that, when $\rho > 0$, both partial and no recommendation lead to increasingly local consumption compared to the benchmark omniscient case.
+Further, the average consumption path between periods is \textit{decreasing} for the no recommendation case whereas it is \textit{increasing} for the omniscient case.
+Partial recommendation decreases the degree of local consumption but not as much as the omniscient benchmark.
+Due to the correlation of utilities, the omniscient consumption path exploits this and leads to the consumption of more similar products than in the case when $\rho = 0$.
+However, since these spillovers also impact consumer learning in the no recommendation case, consumers \textit{over-exploit} this and increasingly consume products similar to good products that they have consumed before.
+This is further illustrated by Figure \ref{fig:local_consumption_across_rho} which shows how the consumption paths between omniscient and no recommendation vary as $\rho$ increases.
+
+Finally, this effect is further amplified as the level of risk aversion, $\gamma$, increases.
+Figure \ref{fig:no_rec_risk_aversion} shows how drastically the degree of local consumption increases as $\gamma$ increases.
+This is due to the fact that the spillovers not only affect the mean expected belief about quality but also the degree of uncertainty.
+Local consumption therefore leads to consumers to have less uncertainty about certain areas of the product space and risk aversion may lead them to increasingly consume these products.
+
+\begin{figure*}
+\includegraphics[scale=0.1]{figures/consumption_dist_N_200T_20_no_rec_rho}
+\includegraphics[scale=0.1]{figures/consumption_dist_N_200T_20_omni_rho}
+\caption{Extent of Local Consumption for No Recommendation (Left) and Omniscient Recommendation (Right)}
+\label{fig:local_consumption_across_rho}
+\end{figure*}
+
+\begin{figure}
+\includegraphics[scale=0.1]{figures/consumption_dist_N_200T_20_no_rec_alpha}
+\caption{The Effect of Risk Aversion on Local Consumption}
+\label{fig:no_rec_risk_aversion}
+\end{figure}
+
+\subsection{User Welfare and Product Diversity}
+
+In this section we primarily focus on the impact of recommendation on the welfare on consumers as well as the overall diversity of the items that they consume.
+Consumer's \textit{ex-post} welfare is the average of realized utilities, to control for the effect of $T$, and is defined as follows:
+\begin{align*}
+W_i=\frac{1}{T}\sum_{n \in C_i^T} u_{in}
+\end{align*}
+While in the previous section we looked at the distance between consecutive goods, we also define a diversity measure that is defined over the entire consumed set.
+We utilize a diversity metric common in the literature (e.g.
+\cite{ziegler2005improving}) which is the average normalized pairwise distance between the consumed products:
+\begin{align*}
+D_i =\frac{1}{N} \frac{1}{T (T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)
+\end{align*}
+
+\begin{finding}\label{finding_diversity}
+The impact of recommendation on product diversity:
+\begin{enumerate}
+\item When $\rho = 0$, product diversity is the same across all three recommendation regimes.
+\item When $\rho > 0$, product diversity decreases across all recommendation regimes but decreases the most in the no recommendation regime.
+This effect is amplified as $\rho$ increases as well as when consumers become more risk-averse.
+\end{enumerate}
+\end{finding}
+
+As before, when there is no correlation between utilities product diversity is the same across different recommendation regimes.
+The over-exploitation of the learning spillovers when $\rho > 0$ leads to product diversity being lowest in the no recommendation regime.
+Figure \ref{fig:diversity_correlation} shows how diversity varies across recommendation regimes and the level of $\rho$ and Figure \ref{fig:risk_aversion_diversity} shows how diversity varies as risk aversion levels change.
+
+\begin{finding}
+The welfare gap between no recommendation and partial recommendation as well as no recommendation and omniscient recommendation is decreasing in $\rho$.
+\end{finding}
+However, interestingly, the decrease in diversity is \textit{not} associated with a decline in welfare.
+In fact, welfare stays roughly the same in the omniscient case and slightly increases for the partial recommendation case.
+The welfare gap between the no recommendation case and the other two cases \textit{decreases} as the diversity gap increases.
+
+\begin{finding}
+Without recommendation, diversity and welfare are:
+\begin{enumerate}
+\item Negatively correlated when consumers have no risk-aversion
+\item Uncorrelated when consumers have high levels of risk-aversion
+\end{enumerate}
+\end{finding}
+
+Figure \ref{fig:diversity_welfare_ra} shows how diversity and welfare correlate for the no recommendation case.
+When there is no risk-aversion then there is a negative correlation between welfare and diversity.
+This is since, with no risk-aversion, in a given period a consumer will select the good that they currently believe has the highest expected value.
+High product diversity in this case can arise from a consumer who consumes a bad item and updates her beliefs about nearby items negatively.
+As a result, in the next period she will pick an item far away in the product space from the item that she consumed previously.
+If the item that she had consumed was ``good", then she is more likely to pick a nearby item.
+The learning spillovers therefore lead to high product diversity being negatively correlated with welfare.
+
+\begin{figure}
+\includegraphics[scale=0.1]{figures/gamma_diversity_N_200_T_20}
+\caption{Risk Aversion and Diversity}
+\label{fig:risk_aversion_diversity}
+\end{figure}
+
+This only happens since $\gamma = 0$ leads to consumers only caring about the expected value of the good.
+However, as we saw in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, increasing $\gamma$ can lead to lower diversity and increasingly local consumption due to the fact that the degree of uncertainty now impacts consumers' choices.
+This weakens the negative relationship between diversity and welfare since both negative and positive experiences with a good reduce uncertainty about surrounding goods.
+This leads to the inverted-U shape found in Figure \ref{fig:diversity_welfare_ra} when $\gamma$ is relatively large (e.g $\gamma = 5$).
+
+\begin{figure*}
+\includegraphics[scale=0.1]{figures/rho_diversity_N_200_T_20}
+\includegraphics[scale=0.1]{figures/rho_welfare_N_200_T_20}
+\caption{Welfare, Diversity, and Correlation}
+\label{fig:diversity_correlation}
+\end{figure*}
+
+\begin{figure*}
+\includegraphics[scale=0.3]{figures/diversity_welfare_no_risk_aversion}
+\includegraphics[scale=0.3]{figures/diversity_welfare_high_ra}
+\caption{Diversity vs. Welfare, $\gamma = 0$ (Left), Diversity vs. Welfare, $\gamma = 5$ (Right)}
+\label{fig:diversity_welfare_ra}
+\end{figure*}
+
+\subsection{User Homogenization}
+
+In this section we focus on across user comparisons and investigate how the consumed set of items across consumers varies across different recommendation regimes and parameter values.
+In particular we look at the degree of \textit{homogenization} between consumers.
+Similar to \cite{chaney2018algorithmic} we use the Jaccard index between the consumption sets of consumers to measure homogeneity:
+\begin{align*}
+H:=\frac{1}{|I|(|I|-1)}\sum_{i,j \in I: i \ne j}d_J(C_i^T,C_j^T)
+\end{align*}
+where $d_J$ denotes the Jaccard index and $H \in [0,1]$.
+\begin{finding}
+Homogeneity is:
+\begin{enumerate}
+\item Highest under partial recommendation and lowest under no recommendation
+\item Increasing in $\beta$, or the weight of the common-value component
+\item Decreasing in $\rho$ for partial recommendation, but weakly increasing in $\rho$ for no recommendation
+\end{enumerate}
+\end{finding}
+
+Figure \ref{fig:beta_homo} shows that as the weight of the common-value component, $\beta$, increases consumers consume increasingly similar sets of items.
+The homogenization effect is strongest under partial recommendation since the revelation of the common-value component induces consumers to consume products in similar areas of the product space.
+However, as $\beta$ increases, some amount of homogenization is optimal as can be seen from the omniscient recommendation case.
+In the no recommendation case since consumers do not know the common-value component they engage in local consumption in different areas of the product space which leads to less than optimal homogeneity.
+
+Figure \ref{fig:cor_homo} shows that the degree of homogeneity in the partial recommendation case however is \textit{decreasing} as $\rho$ increases.
+As was discussed in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, the degree of local consumption increases with $\rho$.
+Even though the revelation of the common-value component induces them to search in similar parts of the product space, their idiosyncratic components induce them to consume products in a more localized area of the product space as $\rho$ increases which leads to a decline in homogeneity.
+
+\begin{figure}
+\includegraphics[scale=0.1]{figures/beta_homogeneity_N_200_T_20}
+\caption{Strength of Recommendation and Homogeneity}
+\label{fig:beta_homo}
+\end{figure}
+
+\begin{figure}
+\includegraphics[scale=0.1]{figures/rho_homogeneity_N_200_T_20}
+\caption{Correlation and Homogeneity}
+\label{fig:cor_homo}
+\end{figure}
 
 \subsection{Effectiveness of Recommendation}
-
-\begin{table}[bt]
-\centering
-\begin{tabular}{rlll}
-  \hline
-Rec Policy & Welfare & Diversity & Homogeneity \\ 
-  \hline
-No Rec & 0.42 $\pm$ 0.002 & 0.19 $\pm$ 0.0003 & 0.001 $\pm$ 0 \\ 
-Partial Rec & 1.25 $\pm$ 0.005 & 0.22 $\pm$ 0.0002 & 0.174 $\pm$ 0.004 \\ 
-Omniscient Rec & 2.24 $\pm$ 0.006 & 0.24 $\pm$ 0.0001 & 0.035 $\pm$ 0.002 \\ 
-   \hline
-\end{tabular}
-\caption{Mean Welfare, Diversity, and Homogeneity for $N = 1000, T = 25, I = 50, S = 50$. Reported intervals are 95\% confidence intervals.}
-\label{table:agg_results}
-\end{table}
-
-We are going to analyze three outcomes: how diverse consumer choices are, how similar are the choices that different individuals make and the resulting welfare.
-The first will be measured by $D_i =\frac{1}{N} \frac{1}{T (T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$, the average normalized pairwise distance between the consumed products, a natural measure of diversity in this environment that is utilized in the literature \cite{ziegler2005improving}.
-The second will be measured by a consumer homogeneity index $H:=\frac{1}{|I|(|I|-1)}\sum_{i,j \in I: i \ne j}d_J(C_i^T,C_j^T) $, where $d_J$ denotes the Jaccard index and $H \in [0,1]$ which is similar to the approach to measuring homogeneity utilized in \cite{chaney2018algorithmic}.
-Note that a higher $H$ indicates more homogeneity.
-Consumer $i$'s welfare will just be the average of realized utilities, to control for the effect of $T$, $W_i=\frac{1}{T}\sum_{n \in C_i^T} u_{in}$.
-\par
-\par
-\noindent \textbf{Results}.
-Table \ref{table:agg_results} shows aggregate results for welfare, diversity, and homogeneity.
-The omniscient recommendation case leads to the optimal consumption path for a consumer since she consumes the $T$ items with the highest utility.
-Compared to this case, both partial recommendation and no recommendation not only lead to lower welfare but also have lower diversity.
-In fact, the no recommendation case leads to the lowest levels of consumption diversity since users do not explore the space of products sufficiently and do excessive amounts of local search as can be seen in Figure \ref{fig:local_consumption}.
-Figure \ref{fig:local_consumption} shows how ``local" consumption choices are by looking at what fraction of consecutive consumption choices are close to each other with the definition of close being less than distance of 10.
-The partial recommendation case leads to higher consumption diversity and higher welfare, but also significantly higher homogeneity.
-
-In our model, filter bubble effects occur simply because of the correlation structure between the utility and beliefs of products.
-It highlights the fact that the correlation structure induces users to consume increasingly narrow content when they get no guidance from recommendation whereas the ``optimal" consumption path given by the omniscient policy does not exhibit this.
-The results of our model are consistent with the empirical results found in \cite{nguyen2014exploring} who show that filter bubble effects can arise even amongst users who do not utilize recommendations.
-\par
-Furthermore, \cite{nguyen2014exploring} also find that recommendation can lead to an increase in consumption diversity.
-As Table \ref{table:agg_results} shows, our model is consistent with this effect but it depends on the nature of recommendation.
-If recommendations only provides information on the common-value component (partially informed recommender) then welfare and consumption diversity increase when contrasted to the no recommendation case, but homogenization increases dramatically.
-Knowing the common-value component leads users to have more information on where to explore in the product space.
-However, the fact that this information is about the common-value component and the same across all individuals leads them to explore the same sections of the product space and leads to substantially higher homogeneity among users.
-If recommendation takes into account user beliefs on the idiosyncratic component of utility (omniscient recommender), then welfare and consumption diversity can further increase and homogeneity will decrease.
-\par
-Low consumption diversity is not necessarily bad by itself.
-Figure \ref{fig:diversity_welfare_no_rec} shows that, in the no recommendation case, welfare and consumption diversity are negatively correlated.
-This is primarily because content diversity in the no recommendation case can arise from the fact that the user consumes a bad product and is forced to jump around the product space looking for ``good" products.
-However, if a user finds a high utility product right away, then staying in this neighborhood may yield higher utility due to the correlation of utilities.
-The problem is that, even when the user finds a ``good" product, this may only be ``locally" good and the user will then excessively consume items around it due to having more information about these than more ``different'' ones, i.e.
-farther away.
-\par
-Finally, the value of recommendation in the partial recommendation case decreases over time as users learn more about the idiosyncratic component of their own preferences (i.e.
-as $t$ increases).
-Figure \ref{fig:rec_obey} shows that recommendation compliance decreases as $t$ increases.
-Additionally, Figure \ref{fig:beta_vary} shows how homogeneity and welfare change as $\beta$ increases.
-Most interestingly, when $\beta$ is high so that the common-value component is small, recommendation can be harmful to users.
-The information from recommendation still leads to large user homogeneity even though the optimum would be for there to be almost none and leads to lower welfare than under no recommendation.
 

### 2019-09-14 analysis/create_graphs.R

@@ -431,13 +431,13 @@ process_time_path <- function(N, T_val, use_hrbrthemes){
   no_rho_or_gamma <- filter(time_data, rho == 0.0 & gamma == 0.0)
   no_rho <- filter(time_data, rho == 0.0 & gamma > 0.0)
   no_gamma <- filter(time_data, gamma == 0.0 & rho > 0.0)
-  g <- ggplot(no_rho_or_gamma, aes(x=t, y=mean_consumption_dist)) +
-    geom_smooth(aes(colour=formatted_regime)) + theme_bw() +
-    ggtitle(paste("Consumption Distance, No Correlation or Risk Aversion")) + xlab("t") + ylab("Average Consumption Distance")
+  g <- ggplot(no_rho, aes(x=t, y=mean_consumption_dist)) +
+    geom_smooth(aes(colour=formatted_regime)) + theme_bw() + ylim(c(30, 60)) +
+    ggtitle(paste("Consumption Distance, No Correlation")) + xlab("t") + ylab("Average Consumption Distance")
   g <- g + theme(legend.position="bottom", legend.title = element_blank())  + 
     guides(color=guide_legend(override.aes=list(fill=NA)))
   ggsave(
-    filename= paste(WORKING_DIR, "all_figures/consumption_dist_N_", N,"T_", T_val,"no_correlation_risk_aversion.jpeg", sep=""), 
+    filename= paste(WORKING_DIR, "all_figures/consumption_dist_N_", N,"T_", T_val,"no_correlation", sep=""), 
     plot=g
   )
   

### 2019-09-14 simulation_code/simulations.jl

@@ -386,7 +386,7 @@ sigma_ibar = .1
 #
 rho_ibar = 0.0
 
-N_vals = [200]
+N_vals = [100, 500, 1000]
 
 T_vals = [20]
 

### 2019-09-14 simulation_code/simulations.jl
updates

@@ -1,6 +1,6 @@
 using Distributed
 using JSON
-addprocs(17)
+addprocs(30)
 @everywhere using IterTools;
 @everywhere using Dates;
 @everywhere using Distances;
@@ -380,7 +380,7 @@ end
 #
 nr_pop = 100
 #
-nr_ind = 100
+nr_ind = 25
 #
 sigma_ibar = .1
 #

### 2019-09-14 simulation_code/simulations.jl
reduce to 25

@@ -1,6 +1,6 @@
 using Distributed
 using JSON
-addprocs(30)
+addprocs(25)
 @everywhere using IterTools;
 @everywhere using Dates;
 @everywhere using Distances;

### 2019-09-14 simulation_code/simulations.jl
20

@@ -1,6 +1,6 @@
 using Distributed
 using JSON
-addprocs(25)
+addprocs(20)
 @everywhere using IterTools;
 @everywhere using Dates;
 @everywhere using Distances;

### 2019-09-14 simulation_code/simulations.jl
remove N = 1000

@@ -386,7 +386,7 @@ sigma_ibar = .1
 #
 rho_ibar = 0.0
 
-N_vals = [100, 500, 1000]
+N_vals = [100, 500]
 
 T_vals = [20]
 

### 2019-09-15 paper/filter_bubbles.tex
wrote the rec sys evaluation section while listening to deadmau5 so i'm sure this went great

@@
 
-\begin{finding}
+\begin{finding}\label{finding_welfare_gap}
 The welfare gap between no recommendation and partial recommendation as well as no recommendation and omniscient recommendation is decreasing in $\rho$.
@@
 
-\begin{finding}
+\begin{finding}\label{finding_diversity_welfare_corr}
 Without recommendation, diversity and welfare are:
@@
 where $d_J$ denotes the Jaccard index and $H \in [0,1]$.
-\begin{finding}
+\begin{finding}\label{finding_homogeneity}
 Homogeneity is:
@@
 
-\subsection{Effectiveness of Recommendation}
+\section{Recommender System Evaluation}
 
-\section{The Value of the Ex-Ante View}
+In this section we discuss how the insights from our model of user decision-making can be used to inform the evaluation and design of recommender systems.
 
-In this section we illustrate the differences between the ex-post and ex-ante view and show how it is useful for thinking about the design of RS.
-Our first point is trivial to state - since users do not know the true consumption values of the items, without any further information they may make \textit{ex-post} sub-optimal consumption decisions, but \textit{ex-ante} optimal given the information available to them at the moment of choice.
-Consider the following stylized example.
-The choice set of the individual is $\Omega= \{x_1, x_2, x_3 \}$ and the ex-post utility values are given by
-$u(x_1) = 1$, $u(x_2) = 2$, $u(x_3) = 3$. 
-\par
-However, when users are making decisions they have unobservable \textit{beliefs} about the ex-post utility value of the items and make decisions given these beliefs.
-As users obtain more information, their beliefs converge to the ex-post utility values but, generally, users do not have enough information and so have beliefs over the space of possible ex-post utility values.
-In this example we suppose these beliefs are simple so that users view each good as simple lotteries: for $n=1,2,3$, $u(x_n) = n + \varepsilon_n$, $\varepsilon_1 \sim \mathcal N (2, 1)$, $\varepsilon_2 \sim \mathcal N (0, 1)$, $\varepsilon_3 \sim \mathcal N (-5, 2)$.
-\par
-After uncertainty is unresolved, the user would consume $x_3$, but an expected utility maximizing consumer (without further information) would choose $x_1$.\footnote{We ignore an important component of decision-making under uncertainty, risk attitudes, by focusing on risk-neutral consumers.
-A \textit{risk-averse} user may choose to consume a good with lower expected \textit{ex-post} payoff but lower ex-ante variance compared to another good.
-While risk-aversion is important for decision-making under uncertainty, especially in the context of RS, it is not our main focus here.}
-\par
-This observation, while seemingly trivial, has important implications.
-The fact that the user would choose $C$ over $A$ reveals that $C$ prefers $A$ given her current beliefs\footnote{This is a statement of \textit{revealed preference} which has been studied heavily in economic decision theory (see e.g.
-\cite[]{mas1995microeconomic}).} and allows the system designer to get some information about a user's beliefs.
-Why are user beliefs important objects for designers of RS to understand and utilize?
-\par
-Beliefs are useful for understanding how users interact and get value from RS.
-In particular, users use RS to get \textit{information} about the set of items and update their beliefs about the value of the items.
-For instance, in the example, a RS could recommend to a user $x_3$ over $x_2$ and $x_1$ and that user will use this to update her beliefs about $x_3$.
-It is an empirical question that we leave for future work precisely how and to what extent users update their beliefs from recommendation and what factors of RS deisgn influence this.\footnote{There has been some work in this direction in understanding when recommendations are persuasive \cite{cremonesi2012investigating, gretzel2006persuasion} as well as the effect of their delivery on effectiveness \cite{murphy2014recommendation}.}
-\par
-Better understanding user beliefs is important for improving the performance of RS that optimize for metrics that move away from accuracy such as serendipity \cite{kotkov2016survey}.
-Serendipity-based RS attempt to provide recommendations that ``have the quality of being both unexpected and useful" \cite{maksai2015predicting} but it is hard to know what items are both unexpected and useful \cite{kotkov2016survey}.
-Understanding \textit{why} a user has not consumed a good depends on the beliefs that the user has and this is important for designing serendipitious recommendations.
-For instance, in the example, the user does not consume $x_3$ because she expects the good to turn out worse than the other two.
-However, if given more information, she would prefer to consume it over $x_1$ and $x_2$.
-Moreover, when the user can use her known valuation of an item to make inferences on how good another item is, consumption is path-dependent, as beliefs guide consumption and consumption affects beliefs about other goods.
-For instance, learning the value of $x_1$ says something about how good $x_2$ is.
-%When considering risk-averse individuals, when the user has alternatives with a known value,The consumer may have also consumed Another reason why the user may not .
-The user may have an alternative with a known value -- say going for a walk around the neighborhood -- which she could always prefer to watching a given movie that has uncertain value but that the consumer thinks is always lower than the value of the walk.\footnote{An additional reason this could be useful is if one considers a risk-averse individual, then it may be that a user has not consumed an item because she has a high perceived variance for the item's quality.}
-As a result, understanding the beliefs of users helps us understand why users have not consumed certain items and, in particular, which items a user may find useful given more information about it.
-This poses an important problem: designing systems that elicit not only ex-post valuations but also ex-ante beliefs about valuations, as only by knowing these can RS be effective in steering users' choices.
-In particular, choice has to be perceived as guided by ex-ante utility, whereas ratings by the ex-post realized utility.
-\par
-Understanding the choices that individuals will make in addition to what they will end up liking are complementary problems but require different viewpoints.
+The classic approach to recommender system evaluation is to predict user ratings for items and compare how accurate this prediction is to recorded ratings data, either explicitly given by users or inferred from behavioral data.
+Further, given this predicted rating, the recommender system should recommend to consumers the items with the highest predicted rating \cite{adomavicius2005toward}.
+
+There has been a recent movement away from such evaluation measures due to the observation that accurate recommendations are not necessarily useful recommendations \cite{mcnee2006being}.
+Our model illustrates why this could be the case.
+Consider the domain of movie recommendation and suppose a user has just watched the movie \textit{John Wick} and rated it highly.
+A recommender system attempting to predict user ratings may then predict that this user is very likely to enjoy the sequel, \textit{John Wick: Chapter Two}, as well.
+However, the user themselves may also have made this inference since the two movies are very similar to each other and, after enjoying \textit{John Wick}, the user will update her beliefs about \textit{John Wick: Chapter Two}.
+Thus, if the recommender system recommends this movie to the user then this recommendation is not useful since the recommendation gives the user little information that they did not already know.
+As a result, the user may watch \textit{John Wick: Chapter Two} even without recommendation and the value of the recommendation was small.
+
+To counteract this problem, alternative recommendation evaluation metrics have been proposed such as serendipity
+\cite{kotkov2016survey, kotkov2016challenges}.
+Ideally, serendipitous recommendations ``have the quality of being both unexpected and useful" \cite{maksai2015predicting}.
+\cite{iaquinta2010can} defines serendipitous recommendations as follows:
+\begin{quote}
+\textit{A serendipitous recommendation helps the
+user to find a surprisingly interesting item that
+she might not have otherwise discovered (or it
+would have been really hard to discover). [...]
+Serendipity cannot happen if the user already
+knows what is recommended to her, because a
+serendipitous happening is by definition something new. Thus the lower is the probability that user knows an item, the higher is the
+probability that a specific item could result
+in a serendipitous recommendation}.
+\end{quote}
+
+All approaches to serendipity to our knowledge, however, do not take into account the \textit{beliefs} that users have over the valuations for the items.
+For instance, \cite{vargas2011rank, kaminskas2014measuring} propose a unexpectedness metric that looks at the dissimilarity of the recommended items to what the recommender already knows the user likes either via a content-based approach or a collaborative-based approach.
+\cite{adamopoulos2015unexpectedness} use an expected utility approach but measure unexpectedness in terms of deviations from their ``expected set of items."
+
+We propose a definition of serendipity based on \textit{unexpectedness} and \textit{usefulness} which have have clear definitions in the context of our model of user decision-making.
+Users have \textit{beliefs} over the valuations for the items and thus \textit{unexpected} recommendations are given by items that consumers have low expected utility for, either due to the fact that their mean belief is low or the degree of uncertainty is high.
+Recall from Findings \ref{finding_local_consumption} and \ref{finding_diversity_welfare_corr} that, particularly if users are risk-averse, they will consume content in increasingly narrow portions of the product space and this may be detrimental to their realized welfare.
+As a result, identifying the portions of the product space where users may have significant uncertainty and utilizing recommendation to reduce this uncertainty can be welfare-enhancing for consumers.
+This motivates a definition of \textit{useful} recommendations as recommendations that give consumers information that would \textit{change the item they consume relative to no recommendation}.
+
+Operationalizing this approach requires the collection of data that is not traditionally collected for recommender system design.
+In particular, it becomes important to understand what choices users would make \textit{without recommendation} and to collect data on user beliefs.
 \cite{jiang2014choice, saavedra2016choice} argue that choice-based approaches alone can help in designing better RS, though they argue for these approaches because they do a better job at providing more accurate recommendations.
-We argue that the two problems should be considered separately -- though they may interact in interesting and useful ways.\footnote{In particular, some results in \cite{jiang2014choice, saavedra2016choice} may be driven by the fact that user choice embeds ratings information that may not have been observed by the recommender but is observed by the user.
-Ratings, reviews, friends, there are many sources of information affecting a person's beliefs and choices that are unobservable by RS.
+However, we argue that choice-based approaches are useful not as recommendations but as a baseline for what recommendations would be useful.\footnote{Furthermore, user beliefs and user choices contain information that may not have been observed by the recommender but is observed by the user.
+Ratings, reviews, friends, there are many sources of information affecting a person's beliefs and choices that are unobservable by recommender systems.
 Other consumption choices, e.g.
-movies seen in the cinema, change beliefs about for instance how good a director is, which affect beliefs about other movies and guide choices on a streaming platform that is unable to observe these data.}
-\par
-RS are designed to give users \textit{information} that is useful to help them make decisions.
-On the one hand, predicting a user's ex-post preferences is useful to know what items the user would like to consume.
-On the other hand, predicting what items a user will actually consume (without recommendation) can help reason about what information may actually be useful to give her.
-Furthermore, in designing RS to avoid filter bubble and homogenization effects, having accurate choice predictions allows mitigating such adverse effects to become a first-order component of design.
-In particular, by predicting both choice and ratings, RS can provide information to users that leads them to useful items but prevents them from falling into a filter bubble.
-\par
-Solving this choice prediction problem requires data on the choices that individuals make and the sets that they choose them from in addition to the traditional ratings or behavioral data that is collected.
-In the example given previously the fact that the consumer chooses $x_1$ from the set $\{x_1, x_2, x_3 \}$ would be the data that is useful for the choice prediction problem, as would the consumer's ex-ante beliefs about her valuation of the goods.
+movies seen in the cinema, change beliefs about for instance how good a director is, which affect beliefs about other movies and guide choices on a streaming platform that is unable to observe these data.
+However, this information can be inferred by the platform by collecting both beliefs and consumption choices.}
+
+Furthermore, our model shows the importance of understanding the underlying user preferences and the nature of the product space in the setting in which a recommender system is deployed since that will change the consumption patterns of consumers.
+For instance, understanding the degree of correlation of utilities and the level of risk-aversion become crucial in understanding how users will make consumption choices without recommendation and what kind of recommendation would be useful for users.
+
+\section{Discussion}
 

### 2019-09-15 paper/filter_bubbles.tex
fixups to section 4

@@
 All approaches to serendipity to our knowledge, however, do not take into account the \textit{beliefs} that users have over the valuations for the items.
-For instance, \cite{vargas2011rank, kaminskas2014measuring} propose a unexpectedness metric that looks at the dissimilarity of the recommended items to what the recommender already knows the user likes either via a content-based approach or a collaborative-based approach.
-\cite{adamopoulos2015unexpectedness} use an expected utility approach but measure unexpectedness in terms of deviations from their ``expected set of items."
+For instance, \cite{vargas2011rank, kaminskas2014measuring} propose unexpectedness metrics that look at the dissimilarity of the recommended items to what the recommender already knows the user likes either via a content-based approach or a collaborative-based approach.
+\cite{adamopoulos2015unexpectedness} use an expected utility approach but measure unexpectedness in terms of deviations from the ``expected consideration set" of a user and do not explicitly incorporate user beliefs.
 
@@
 This motivates a definition of \textit{useful} recommendations as recommendations that give consumers information that would \textit{change the item they consume relative to no recommendation}.
+Combining these, serendipitious recommendations are recommendations that provide users with information about \textit{unexpected} items that end up being \textit{useful} by changing consumption choices to increase the realized utility of a user.
 
-Operationalizing this approach requires the collection of data that is not traditionally collected for recommender system design.
-In particular, it becomes important to understand what choices users would make \textit{without recommendation} and to collect data on user beliefs.
+Operationalizing this approach requires the collection of data that is not traditionally collected for recommender systems.
+In particular, it becomes important to understand what choices users would make \textit{without recommendation} and to collect data on user beliefs, which can partially be inferred by observed choices.
 \cite{jiang2014choice, saavedra2016choice} argue that choice-based approaches alone can help in designing better RS, though they argue for these approaches because they do a better job at providing more accurate recommendations.

### 2019-09-16 simulation_code/simulations.jl
bump up

@@ -380,7 +380,7 @@ end
 #
 nr_pop = 100
 #
-nr_ind = 25
+nr_ind = 60
 #
 sigma_ibar = .1
 #

### 2019-09-16 paper/filter_bubbles.tex
update model seciton

@@
 \setcopyright{acmlicensed}
-\acmConference[RecSys '19]{Recomme '18: ACM Symposium on Neural Gaze Detection}{September 16--20, 2019}{Copenhagen, Denmark}
-\acmBooktitle{RecSys '19: ACM Conference on Neural Gaze Detection, June 03--05, 2018, Woodstock, NY}
+\acmConference[WWW '20]{Recomme '18: ACM Symposium on Neural Gaze Detection}{April 20--24, 2020}{Taipei, Taiwan}
+\acmBooktitle{WWW '20: ACM Conference on Neural Gaze Detection, June 03--05, 2018, Woodstock, NY}
 \acmPrice{15.00}
@@
 
-\subsection{Preliminaries and Illustrative Example}
-[Duarte can do this] \\
-\noindent \textbf{Expected Utility Theory} First, we review basic economic decision theory.
+\subsection{Preliminary on Expected Utility Theory}
+[Duarte can do this]
 
-\noindent \textbf{Illustrative Example}
-
-\subsection{Model}
+\subsection{Model Setup}
 \par
@@
 We denote by $u_{in}$ individual $i$'s realized utility from consuming item $n$.
-By realized utility we In particular, we consider that the realized utility derived from a given item can be decomposed in the following manner: $u_{in}= v_{in} + \beta v_n$, where $v_{in}$ denotes an idiosyncratic component -- i.e.
+By realized utility we In particular, we consider that the realized utility derived from a given item can be decomposed in the following manner:
+\begin{align*}
+u_{in}= v_{in} + \beta v_n
+\end{align*}
+where $v_{in}$ denotes an idiosyncratic component -- i.e.
 consumer $i$'s idiosyncratic taste for good $n$ --  and $v_{n}$, a common-value component.
 One can interpret $v_n$ as a measure of how much good $n$ is valued in society in general and, in a sense, $v_{in}$ denotes how $i$ diverges from this overall ranking.
-The scalar $\beta \in [0,1]$ denotes the degree to which utilities are idiosyncratic to each individual or common across individuals.
+The scalar $\beta \in \mathbb{R}_{+}$ denotes the degree to which utilities are idiosyncratic to each individual or common across individuals.
 If $\beta=0$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta \rightarrow \infty$, every individual has the same preferences.
+[include some discussion of what this implies about the predictability of a user's preferences]
 
-Stacking utilities in vector-form, we get ${\left(u_{in}\right)}_{n \in \mathcal{N}}=:U_i=V_i+ \beta V $, the vector of utilities associated with each good, where $V_i ={\left(v_{in}\right)}_{n \in \mathcal{N}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{N}}$.
+Stacking utilities in vector-form, we get:
+\begin{align*}
+{\left(u_{in}\right)}_{n \in \mathcal{N}}=:U_i=V_i+ \beta V
+\end{align*}
+We stack the idiosyncratic and common-value components as $V_i ={\left(v_{in}\right)}_{n \in \mathcal{N}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{N}}$.
 \par
 \noindent \textbf{Consumer Beliefs} We assume that consumers do not know the realized utilities of the goods before consuming an item.
-We make this assumption as our model pertains to markets where the goods are experience goods and modeling consumers as being uncertain over their preferences in these environments  is common in the economics literature [add citations].
-Even though consumers have access to online reviews for goods in many of the markets where recommender systems are deployed, it is prohibitively costly for consumers to acquire information on all of the goods in their choice set.
-Even if they did so, this information would be helpful for reducing uncertainty on the common-value component of their utility and not the idiosyncratic component.
-While we do not explicitly incorporate this into our model, we use it to motivate that consumers make consumption decisions under some degree of uncertainty.
-
 Formally, consumer $i$ starts with some beliefs about $U_i$, namely that the idiosyncratic and common-value parts of the utilities are independent -- $V_i \perp \!\!\!
-\perp V$ -- and that each is multivariate normal
-\begin{definition} The idiosyncratic utility and common-value utility are distributed as follows:
-\begin{center}
+\perp V$ -- and that each is multivariate normal.
+The idiosyncratic utility and common-value utility are distributed as follows:
 \begin{enumerate}
@@
 \end{enumerate}
-\end{center}
-\end{definition}
 
@@
 In our model we incorporate the idea that learning the utility of good $n$ gives a consumer information about similar items.
-This is drawn from recent empirical evidence in \cite{schulz2019structured} that consumers learn how to navigate large choice sets using similarity-based generalization.
 We assume that learning about the utility of good $n$ reveals more about the utility associated to items that are closer to it, which captures the idea that trying a product provides more information about similar products than about dissimilar ones.
 
-In order to have a well-defined notion of similarity we need to define a distance function between goods, which we define as $d(n,m):=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}$ where $m$ and $n$ are the index of the items in $\mathcal{N}$.
+In order to have a well-defined notion of similarity we need to define a distance function between goods, which we define as:
+\begin{align*}
+d(n,m):=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}
+\end{align*}
+ $m$ and $n$ are the indices of the items in $\mathcal{N}$.
 We consider that the entry of $n$-th row and the ($m$)-th column of $\Sigma_i$ is given by $\sigma_i^2 \rho^{d(n,m)}$, and that of $\Sigma$ is given by $\sigma^2 \rho^{d(n,m)}$.
@@
 The particular distance function that we utilize leads to this covariance structure being simple, where the $(n,n+1)$-th entry in the covariance matrix is $\rho$ , $(n,n+2)$-th entry is $\rho^2$, etc.
-This spill-over process is represented in Figure [insert fig].
 \par
@@
 This captures the idea that users are faced with an immense choice set but that ultimately they end up experiencing (and learning) about just a small fraction of it.
-Since this is a sequential decision-making problem under uncertainty, in principle consumers face an exploration-exploitation trade-off.
-However, for tractability, we impose that consumers are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
+For tractability we impose that consumers are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
 Any ties are broken uniformly at random.
@@
 
-We will consider three cases.
+We will consider three recommendation regimes:
+\begin{enumerate}
+\item Partial Recommendation
+\item No Recommendation
+\item Omniscient Recommendation
+\end{enumerate}
 The case of primary interest is \textit{partial recommendation} where the recommender observes utilities accrued, but does not know consumer $i$'s starting beliefs $\overline U_i$.
-In this case, the recommender knows $V$ but does not know neither $V_i$ nor, crucially, users' beliefs $\overline V_i$.
-The recommendation will be that the consumer $i$ chooses $r_{it} \in \arg \max_{n \in \mathcal{N} \backslash C_i^T} u_n$, but we assume the recommender provides full information about $V$.
+In this case, the recommender knows $V$ but does not know $V_i$.\footnote{We do not consider the acquisition of information for the recommender to know $V$ and suppose that she has sufficient data to learn $V$ before recommendation.} However, the recommender does know the users' beliefs, $\overline V_i$.
+Thus, at any given period, the recommender provides a personalized recommendation that combines the knowledge of the common value component $V$ with the user beliefs $\overline V_i$.
+User's beliefs about her valuation of each item become crucial in this case: just providing the user with information about $V$ may change the user's original ranking, but, without considering the user's beliefs, she will not necessarily follow recommendation.
+
+The notion of partial recommendation that we consider is idealized where the recommendation does the Bayesian updating for users.
+An alternative that is equivalent would be the recommender simply providing $V$ without knowing $\overline V_i$.
+In this case, the recommendation will be that the consumer $i$ chooses $r_{it} \in \arg \max_{n \in \mathcal{N} \backslash C_i^T} u_n$, but we assume the recommender provides full information about $V$.
 For instance, the recommender could display the whole distribution of utilities reported by other users or even its average, which is a good proxy for the common value component.\footnote{
@@
 Therefore, we assume that the RS reports the whole $V$, which results in higher expected welfare in choices and is not costly to implement.
-}
-Note that with partial recommendation it is not necessarily optimal for the user to follow the recommendation, that is, to pick the item with the highest common-value component $v_n$.
-Consumer's beliefs about her valuation of each item become crucial in this case: knowing $V$ may change the original ranking, but given this new information the consumer may find it best to pick an item other than the one recommended.
+} The results would be equivalent, but this would shift the burden of Bayesian updating to users.
 
@@
 This gives us a benchmark as to how consumers in our model would behave \textit{without} recommendation so that we can analyze what changes with the introduction of recommendation.
-The second is \textit{omniscient recommendation} where the recommender knows the true utility of each good for each consumer and can therefore recommend the best remaining good in every period.
+The second is \textit{omniscient recommendation} where the recommender knows the true realized utility of each good for each consumer and can therefore recommend the best remaining good in every period.
 This gives us a full information benchmark, which is the optimal consumption path for a consumer if all uncertainty about their preferences was resolved.
-We compare this benchmark to the partial recommendation case and it allows us to understand how much this form of recommendation affects behavior relative to the ideal.
 \par
@@
 We analyze our model using numerical simulation since the sequential decision-making component of our model paired with the rich covariance structure between the items make it difficult to characterize optimal consumer behavior analytically.
+The Gaussian assumption allows for closed form Bayesian-updating which enables us to simulate our model but does not provide much help in analytical characterizations.
+
 We explore how consumption patterns differ as we consider different recommendation regimes and report representative results from our simulations.
-$P$ populations of $I$ users and compare their consumption choices under different recommendation system regimes.
-We simulate over $100$ populations with $100$ consumers.
-We average over the consumers in a population to get a representative consumer and then report results where a single data point is over the set of representative consumers.
+We simulate our model over $100$ populations of users with $100$ users per population.
+For every population and every statistic we calculate, we average over the users in the population to get a representative statistic for this population.
+A given set of parameters and a population are a single data point in our dataset.
 
-\section{User Behavior}
+\subsection{Discussion of Assumptions}
+
+In this section we provide further discussion of several of the assumptions of our model.
+
+\noindent \textbf{Uncertainty}: We assume that users have underlying uncertainty about the realized utility of the items in their choice sets and as a result face a decision problem under uncertainty.
+The first justification for this is that environments where recommender systems are commonly deployed, such as on media streaming or e-commerce platforms, contain experience goods.
+A common assumption in the economics literature is that for experience goods, where users can only observe their realized utility after consumption, users face a decision problem under uncertainty when making consumption choices.
+The second justification is that, even if users are in an environment where they could observe utilities based on product reviews or introspection, it is prohibitively costly for them to do so in large choice sets.\footnote{Recent empirical work [cite the cameras paper and hodgson-lewis] has shown that, for product search, users engage in a ``zooming in" search strategy in terms of the information they acquire about products before purchasing and only end up acquiring information about a small set of products in a large product space.} Even if they acquired all the information from product reviews, this would only fully resolve uncertainty if individuals did not have idiosyncratic tastes.
+
+This motivates modeling recommendation as providing users \textit{information} in order to reduce their uncertainty.
+
+\noindent \textbf{Large Choice Sets}: We are interested in environments where the number of items in the choice set is very large and users only consume a small number of these items.
+This is realistic for platforms such as Netflix, Spotify, Amazon, and YouTube where there are sometimes billions of items and users can realistically only consume a handful of these items in their lifetime.
+
+\noindent \textbf{Learning Spillovers}: Users in our environment face a sequential decision-making problem.
+However, we suppose that individuals are myopic (i.e.
+consume the item with the highest expected utility, given their beliefs in a single period) in order to maintain tractability.
+The predominant difficulty in the learning problem they face is not that they need to repeatedly sample a single item in order to learn its value (as in standard multi-armed bandit formulations), but rather that they need to be able to generalize from their experience from consuming one item to other items.
+After consuming a single item they learn the realized utility of the good.
+
+\indent \cite{schulz2019structured} studies how individuals solve sequential decision-making problems under uncertainty in large choice sets in the context of mobile food delivery orders.
+They find that individuals engage in similarity-based generalization where learning about the realized utility of a particular good provides them with information about similar goods.
+Content-based recommender systems exploit a similar idea where, if a user liked one item, then it is also likely that they will like a similar, but different, item as well.
+
+Our model incorporates this observation in a stylized fashion where the degree of information spill-overs is controlled via parameter $\rho$ and we study how the strength of the spill-overs impacts behavior and the effect of recommendation.
+
+\section{Results}
 

### 2019-09-16 paper/filter_bubbles.tex
bleh

@@
 % These commands are for a PROCEEDINGS abstract or paper.
-\copyrightyear{2018}
-\acmYear{2018}
+\copyrightyear{2019}
+\acmYear{2019}
 \setcopyright{acmlicensed}
@@
 A given set of parameters and a population are a single data point in our dataset.
+
+We simulate over a grid of $\gamma, \sigma, \rho, \beta$.
+When we consider results varying a single parameter, we group the results over the other parameters and consider the differences in value only varying the parameter of interest.
+We report results for a relatively small $T$.
+Unless otherwise reported, our results are robust to varying $N$.
 
@@
 
-To counteract this problem, alternative recommendation evaluation metrics have been proposed such as serendipity
+As a result, there is a need to define alternative recommendation evaluation metrics that result in recommender systems that are more useful for users.
+One recent alternative has been proposed such as serendipity
 \cite{kotkov2016survey, kotkov2016challenges}.

### 2019-09-16 paper/filter_bubbles.tex
updated rec eval section

@@
 We denote by $u_{in}$ individual $i$'s realized utility from consuming item $n$.
-By realized utility we In particular, we consider that the realized utility derived from a given item can be decomposed in the following manner:
+In particular, we consider that the realized utility derived from a given item can be decomposed in the following manner:
 \begin{align*}
@@
 
-As a result, there is a need to define alternative recommendation evaluation metrics that result in recommender systems that are more useful for users.
-One recent alternative has been proposed such as serendipity
-\cite{kotkov2016survey, kotkov2016challenges}.
-Ideally, serendipitous recommendations ``have the quality of being both unexpected and useful" \cite{maksai2015predicting}.
+As a result, several alternative recomendation evaluation metrics, such as serendipity \cite{kotkov2016survey}, coverage \cite{ge2010beyond}, and novelty \cite{vargas2011rank}, have been proposed in order to develop recommender systems that produce recommendations that are more useful for users.
+We use our model to motivate an alternative recommendation evaluation metric that provides an alternative perspective on the serendipity metric proposed in the literature.
+Serendipitious recommendations ``have the quality of being both unexpected and useful" \cite{maksai2015predicting}.
 \cite{iaquinta2010can} defines serendipitous recommendations as follows:
@@
 
-All approaches to serendipity to our knowledge, however, do not take into account the \textit{beliefs} that users have over the valuations for the items.
+Our approach to serendipity comes from the intuition that a recommendation is only useful insofar as it leads a user to take a better action than she would without recommendation.
+This approach stems from the idea that the value of recommendation is the marginal gain in expected utility that a user receives from recommendation.\footnote{This notion is similar to the ``value of information" notion utilized in the literature in economics on the pricing of information goods \cite{bergemann2018design}.} This intuitively should come from providing information about products that, without additional information, users assign a low likelihood to being the best product currently available and so would not consume them without recommendation.
+
+This differs from other approaches in the literature since it depends on the beliefs of the users and what item the user would consume without recommendation.
 For instance, \cite{vargas2011rank, kaminskas2014measuring} propose unexpectedness metrics that look at the dissimilarity of the recommended items to what the recommender already knows the user likes either via a content-based approach or a collaborative-based approach.
-\cite{adamopoulos2015unexpectedness} use an expected utility approach but measure unexpectedness in terms of deviations from the ``expected consideration set" of a user and do not explicitly incorporate user beliefs.
+This depends only on the resulting item-set that the users get recommended and not necessarily user beliefs or the change in user action.
+\cite{adamopoulos2015unexpectedness} use an expected utility approach but use this to measure unexpectedness in terms of deviations from the ``expected consideration set" of a user and do not explicitly incorporate user beliefs.
 
-We propose a definition of serendipity based on \textit{unexpectedness} and \textit{usefulness} which have have clear definitions in the context of our model of user decision-making.
-Users have \textit{beliefs} over the valuations for the items and thus \textit{unexpected} recommendations are given by items that consumers have low expected utility for, either due to the fact that their mean belief is low or the degree of uncertainty is high.
+Our model allows us to give precise definitions for what it means for a recommendation to be \textit{unexpected} and \textit{useful}.
+Unexpected recommendations are given by recommended items that, given their beliefs, users have low expected utility from consuming, irregardless of what the true realized utility is.
 Recall from Findings \ref{finding_local_consumption} and \ref{finding_diversity_welfare_corr} that, particularly if users are risk-averse, they will consume content in increasingly narrow portions of the product space and this may be detrimental to their realized welfare.
 As a result, identifying the portions of the product space where users may have significant uncertainty and utilizing recommendation to reduce this uncertainty can be welfare-enhancing for consumers.
-This motivates a definition of \textit{useful} recommendations as recommendations that give consumers information that would \textit{change the item they consume relative to no recommendation}.
-Combining these, serendipitious recommendations are recommendations that provide users with information about \textit{unexpected} items that end up being \textit{useful} by changing consumption choices to increase the realized utility of a user.
+This is especially the case if this reduction of uncertainty leads users to explore under-explored portions of the product space where true realized utility is high, but expected utility, given beliefs, is low before recommendation.
+
+Useful recommendations are recommendations that give users information that would change the item they consume relative to no recommendation and lead to an increase in expected utility.
+In the previous example, recommending \textit{John Wick: Chapter Two} is not useful precisely because the expected utility before and after recommendation is the same since it does not change behavior.
+The most useful recommendations, or those that have the highest utility gain, are recommendations that induce users to consume unexpected goods.
 

### 2019-09-16 paper/filter_bubbles.tex
find and replace

@@
 
-Recommender Systems (RS) have become critical for assisting consumers in navigating the large choice sets that they face in many online markets.
-For instance, consumers have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
-Consumers in many cases are not aware of most items, let alone have full information about their preferences over them and, to make matters worse, the goods in these markets are usually experience goods whose true utility can only be learned after consumption.
-Furthermore, consumers interact with these systems routinely and watch more than one movie on Netflix, buy more than one product on Amazon, and listen to more than one video on YouTube.
+Recommender Systems (RS) have become critical for assisting users in navigating the large choice sets that they face in many online markets.
+For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
+users in many cases are not aware of most items, let alone have full information about their preferences over them and, to make matters worse, the goods in these markets are usually experience goods whose true utility can only be learned after consumption.
+Furthermore, users interact with these systems routinely and watch more than one movie on Netflix, buy more than one product on Amazon, and listen to more than one video on YouTube.
 
-Recommender systems have been influential in shaping consumer choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.
+Recommender systems have been influential in shaping user choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.
 While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems.
-There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals\footnote{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350/}, that personalized recommender systems lead consumers into \textit{filter bubbles} where they get effectively isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead consumers to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
+There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals\footnote{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350/}, that personalized recommender systems lead users into \textit{filter bubbles} where they get effectively isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
 
-\textbf{Our Model and Contributions} In this paper, drawing from recent work in psychology and decision theory in economics, we first develop a model of consumer decision-making in the context of large choice sets faced by consumers in online markets.
-We then ask how a stylized model of recommendation can shape the behavior of long-lived consumers and finally ask how the insights from our model can be used to improve recommender system design and to better understand the consequences of these systems.
+\textbf{Our Model and Contributions} In this paper, drawing from recent work in psychology and decision theory in economics, we first develop a model of user decision-making in the context of large choice sets faced by users in online markets.
+We then ask how a stylized model of recommendation can shape the behavior of long-lived users and finally ask how the insights from our model can be used to improve recommender system design and to better understand the consequences of these systems.
 
-We focus on understanding the degree to which recommender systems can lead consumers into \textit{filter bubbles}.
+We focus on understanding the degree to which recommender systems can lead users into \textit{filter bubbles}.
 Consistent with recent empirical work in the case of movie consumption \cite{nguyen2014exploring}, our model shows that individual consumption behavior over time naturally exhibits patterns consistent with a filter bubble effect \textit{without recommendation}.
 The key component of our model, drawn from the results of \cite{schulz2019structured}, is that the utilities of similar goods are correlated.
-This means that when consumers consume a good and learn its true utility this gives them information about similar goods.
+This means that when users consume a good and learn its true utility this gives them information about similar goods.
 Crucially, this not only impacts the underlying belief about the average utility of the good, but also the amount of uncertainty.
-This learning spill-over induces consumers to consume goods ``similar" to those they consumed before leading to an increasing narrowing of consumption patterns.
-This effect is further exemplified when consumers are \textit{risk-averse}, a concept from economic decision theory where, all else being equal, consumers have a preference for goods with lower uncertainty to those with higher uncertainty.
+This learning spill-over induces users to consume goods ``similar" to those they consumed before leading to an increasing narrowing of consumption patterns.
+This effect is further exemplified when users are \textit{risk-averse}, a concept from economic decision theory where, all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty.
 
@@
 We suppose each good is a sum of a common value component and an idiosyncratic component, where the idiosyncratic component is inherently unpredictable given other individual's preferences.
-We model existing recommendation systems as providing consumers with information on the common-value component and denote as this as partial recommendation.
-We consider how consumer behavior varies as we move from no recommendation to partial recommendation to omniscient recommendation, which is the optimal consumption path for consumers if they knew the true utility of all the goods in their choice set.
+We model existing recommendation systems as providing users with information on the common-value component and denote as this as partial recommendation.
+We consider how user behavior varies as we move from no recommendation to partial recommendation to omniscient recommendation, which is the optimal consumption path for users if they knew the true utility of all the goods in their choice set.
 
-Consistent with \cite{nguyen2014exploring}, we find that partial recommendation increases consumption diversity.
-However, we also find that partial recommendation increases homogeneity amongst consumers.
+Consistent with \cite{nguyen2014exploring}, we find that partial recommendation increases consumption diversity. However, we also find that partial recommendation increases homogeneity amongst users.
 
@@
 \par
-\noindent \textbf{Consumers} We consider a set of consumers $I$ where each individual $i \in I$ faces the same finite set of $N$ items $\mathcal{N} = \left\{0,1,...,N-1\right\}$.
+\noindent \textbf{users} We consider a set of users $I$ where each individual $i \in I$ faces the same finite set of $N$ items $\mathcal{N} = \left\{0,1,...,N-1\right\}$.
 For simplicity, we assume that individuals only derive pleasure from item $n \in \mathcal{N}$ the first time they consume it.
@@
 where $v_{in}$ denotes an idiosyncratic component -- i.e.
-consumer $i$'s idiosyncratic taste for good $n$ --  and $v_{n}$, a common-value component.
+user $i$'s idiosyncratic taste for good $n$ --  and $v_{n}$, a common-value component.
 One can interpret $v_n$ as a measure of how much good $n$ is valued in society in general and, in a sense, $v_{in}$ denotes how $i$ diverges from this overall ranking.
@@
 \par
-\noindent \textbf{Consumer Beliefs} We assume that consumers do not know the realized utilities of the goods before consuming an item.
-Formally, consumer $i$ starts with some beliefs about $U_i$, namely that the idiosyncratic and common-value parts of the utilities are independent -- $V_i \perp \!\!\!
+\noindent \textbf{user Beliefs} We assume that users do not know the realized utilities of the goods before consuming an item.
+Formally, user $i$ starts with some beliefs about $U_i$, namely that the idiosyncratic and common-value parts of the utilities are independent -- $V_i \perp \!\!\!
 \perp V$ -- and that each is multivariate normal.
@@
 We impose the normality assumption for two reasons.
-The first is that consumers update their beliefs using Bayesian updating and recall that the normal distribution forms a conjugate family, which allows for simple posterior updates.
+The first is that users update their beliefs using Bayesian updating and recall that the normal distribution forms a conjugate family, which allows for simple posterior updates.
 The second is that it allows us to incorporate an easily interpretable correlation structure between the items.
@@
 \par
-\noindent \textbf{Consumer Learning}
-When a consumer consumes a good $n$ they learn the realized utility for that good.
-In our model we incorporate the idea that learning the utility of good $n$ gives a consumer information about similar items.
+\noindent \textbf{user Learning}
+When a user consumes a good $n$ they learn the realized utility for that good.
+In our model we incorporate the idea that learning the utility of good $n$ gives a user information about similar items.
 We assume that learning about the utility of good $n$ reveals more about the utility associated to items that are closer to it, which captures the idea that trying a product provides more information about similar products than about dissimilar ones.
@@
 \par
-\textbf{Consumer Decision-Making}
+\textbf{user Decision-Making}
 We assume the user makes $T$ choices and therefore can only consume up to $T$ items, where $T$ is but a small fraction of $N$.
 This captures the idea that users are faced with an immense choice set but that ultimately they end up experiencing (and learning) about just a small fraction of it.
-For tractability we impose that consumers are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
+For tractability we impose that users are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
 Any ties are broken uniformly at random.
@@
 Our model of recommendation is stylized in order to provide qualitative insights into how recommendation shapes behavior, instead of looking at realistic implementations of recommender systems.
-We model recommendation as giving consumers information about the utility of the goods.
+We model recommendation as giving users information about the utility of the goods.
 
@@
 \end{enumerate}
-The case of primary interest is \textit{partial recommendation} where the recommender observes utilities accrued, but does not know consumer $i$'s starting beliefs $\overline U_i$.
+The case of primary interest is \textit{partial recommendation} where the recommender observes utilities accrued, but does not know user $i$'s starting beliefs $\overline U_i$.
 In this case, the recommender knows $V$ but does not know $V_i$.\footnote{We do not consider the acquisition of information for the recommender to know $V$ and suppose that she has sufficient data to learn $V$ before recommendation.} However, the recommender does know the users' beliefs, $\overline V_i$.
@@
 An alternative that is equivalent would be the recommender simply providing $V$ without knowing $\overline V_i$.
-In this case, the recommendation will be that the consumer $i$ chooses $r_{it} \in \arg \max_{n \in \mathcal{N} \backslash C_i^T} u_n$, but we assume the recommender provides full information about $V$.
+In this case, the recommendation will be that the user $i$ chooses $r_{it} \in \arg \max_{n \in \mathcal{N} \backslash C_i^T} u_n$, but we assume the recommender provides full information about $V$.
 For instance, the recommender could display the whole distribution of utilities reported by other users or even its average, which is a good proxy for the common value component.\footnote{
 The best item that could recommended with such information if the item with highest common value component.
-However, in that case, recommending only a single item generates costly Bayesian updating to the consumer and provides little guidance as to what she should indeed pick as the common value might be of little importance when compared to the idiosyncratic component.
+However, in that case, recommending only a single item generates costly Bayesian updating to the user and provides little guidance as to what she should indeed pick as the common value might be of little importance when compared to the idiosyncratic component.
 Therefore, we assume that the RS reports the whole $V$, which results in higher expected welfare in choices and is not costly to implement.
@@
 We further consider two cases that serve primarily as benchmarks.
-The first is \textit{no recommendation}, where consumers get no additional information about utilities and make consumption choices based on their beliefs and consumption history.
-This gives us a benchmark as to how consumers in our model would behave \textit{without} recommendation so that we can analyze what changes with the introduction of recommendation.
-The second is \textit{omniscient recommendation} where the recommender knows the true realized utility of each good for each consumer and can therefore recommend the best remaining good in every period.
-This gives us a full information benchmark, which is the optimal consumption path for a consumer if all uncertainty about their preferences was resolved.
+The first is \textit{no recommendation}, where users get no additional information about utilities and make consumption choices based on their beliefs and consumption history.
+This gives us a benchmark as to how users in our model would behave \textit{without} recommendation so that we can analyze what changes with the introduction of recommendation.
+The second is \textit{omniscient recommendation} where the recommender knows the true realized utility of each good for each user and can therefore recommend the best remaining good in every period.
+This gives us a full information benchmark, which is the optimal consumption path for a user if all uncertainty about their preferences was resolved.
 \par
 \noindent \textbf{Simulation Details}.
-We analyze our model using numerical simulation since the sequential decision-making component of our model paired with the rich covariance structure between the items make it difficult to characterize optimal consumer behavior analytically.
+We analyze our model using numerical simulation since the sequential decision-making component of our model paired with the rich covariance structure between the items make it difficult to characterize optimal user behavior analytically.
 The Gaussian assumption allows for closed form Bayesian-updating which enables us to simulate our model but does not provide much help in analytical characterizations.
@@
 
-We characterize ``filter bubble" effects in our model as the degree to which consumers engage in \textit{local consumption}.
-We define local consumption in terms of the average consumption distance between the items consumed by the representative consumer at time $t-1$ and $t$.
+We characterize ``filter bubble" effects in our model as the degree to which users engage in \textit{local consumption}.
+We define local consumption in terms of the average consumption distance between the items consumed by the representative user at time $t-1$ and $t$.
 We compare the average movement within the product space across different recommendation regimes and associate lower average distance with more local consumption.
@@
 \item When $\rho > 0$, no recommendation induces more local consumption than both partial and omniscient recommendation.
-This effect is amplified as $\rho$ increases as well as when consumers are more risk averse ($\gamma$ increases)
+This effect is amplified as $\rho$ increases as well as when users are more risk averse ($\gamma$ increases)
 \end{enumerate}
@@
 This is due to the fact that when $\rho = 0$, there is no reason that items that are close in the product space should have similar utilities and so the optimal consumption path does not depend on the similarity of the items.
-However this also means that consumers in the no recommendation regime do not learn anything about neighboring goods and so there is limited path-dependence in consumption.
+However this also means that users in the no recommendation regime do not learn anything about neighboring goods and so there is limited path-dependence in consumption.
 
@@
 Due to the correlation of utilities, the omniscient consumption path exploits this and leads to the consumption of more similar products than in the case when $\rho = 0$.
-However, since these spillovers also impact consumer learning in the no recommendation case, consumers \textit{over-exploit} this and increasingly consume products similar to good products that they have consumed before.
+However, since these spillovers also impact user learning in the no recommendation case, users \textit{over-exploit} this and increasingly consume products similar to good products that they have consumed before.
 This is further illustrated by Figure \ref{fig:local_consumption_across_rho} which shows how the consumption paths between omniscient and no recommendation vary as $\rho$ increases.
@@
 This is due to the fact that the spillovers not only affect the mean expected belief about quality but also the degree of uncertainty.
-Local consumption therefore leads to consumers to have less uncertainty about certain areas of the product space and risk aversion may lead them to increasingly consume these products.
+Local consumption therefore leads to users to have less uncertainty about certain areas of the product space and risk aversion may lead them to increasingly consume these products.
 
@@
 
-In this section we primarily focus on the impact of recommendation on the welfare on consumers as well as the overall diversity of the items that they consume.
-Consumer's \textit{ex-post} welfare is the average of realized utilities, to control for the effect of $T$, and is defined as follows:
+In this section we primarily focus on the impact of recommendation on the welfare on users as well as the overall diversity of the items that they consume.
+user's \textit{ex-post} welfare is the average of realized utilities, to control for the effect of $T$, and is defined as follows:
 \begin{align*}
@@
 \item When $\rho > 0$, product diversity decreases across all recommendation regimes but decreases the most in the no recommendation regime.
-This effect is amplified as $\rho$ increases as well as when consumers become more risk-averse.
+This effect is amplified as $\rho$ increases as well as when users become more risk-averse.
 \end{enumerate}
@@
 \begin{enumerate}
-\item Negatively correlated when consumers have no risk-aversion
-\item Uncorrelated when consumers have high levels of risk-aversion
+\item Negatively correlated when users have no risk-aversion
+\item Uncorrelated when users have high levels of risk-aversion
 \end{enumerate}
@@
 When there is no risk-aversion then there is a negative correlation between welfare and diversity.
-This is since, with no risk-aversion, in a given period a consumer will select the good that they currently believe has the highest expected value.
-High product diversity in this case can arise from a consumer who consumes a bad item and updates her beliefs about nearby items negatively.
+This is since, with no risk-aversion, in a given period a user will select the good that they currently believe has the highest expected value.
+High product diversity in this case can arise from a user who consumes a bad item and updates her beliefs about nearby items negatively.
 As a result, in the next period she will pick an item far away in the product space from the item that she consumed previously.
@@
 
-This only happens since $\gamma = 0$ leads to consumers only caring about the expected value of the good.
-However, as we saw in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, increasing $\gamma$ can lead to lower diversity and increasingly local consumption due to the fact that the degree of uncertainty now impacts consumers' choices.
+This only happens since $\gamma = 0$ leads to users only caring about the expected value of the good.
+However, as we saw in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, increasing $\gamma$ can lead to lower diversity and increasingly local consumption due to the fact that the degree of uncertainty now impacts users' choices.
 This weakens the negative relationship between diversity and welfare since both negative and positive experiences with a good reduce uncertainty about surrounding goods.
@@
 
-In this section we focus on across user comparisons and investigate how the consumed set of items across consumers varies across different recommendation regimes and parameter values.
-In particular we look at the degree of \textit{homogenization} between consumers.
-Similar to \cite{chaney2018algorithmic} we use the Jaccard index between the consumption sets of consumers to measure homogeneity:
+In this section we focus on across user comparisons and investigate how the consumed set of items across users varies across different recommendation regimes and parameter values.
+In particular we look at the degree of \textit{homogenization} between users.
+Similar to \cite{chaney2018algorithmic} we use the Jaccard index between the consumption sets of users to measure homogeneity:
 \begin{align*}
@@
 
-Figure \ref{fig:beta_homo} shows that as the weight of the common-value component, $\beta$, increases consumers consume increasingly similar sets of items.
-The homogenization effect is strongest under partial recommendation since the revelation of the common-value component induces consumers to consume products in similar areas of the product space.
+Figure \ref{fig:beta_homo} shows that as the weight of the common-value component, $\beta$, increases users consume increasingly similar sets of items.
+The homogenization effect is strongest under partial recommendation since the revelation of the common-value component induces users to consume products in similar areas of the product space.
 However, as $\beta$ increases, some amount of homogenization is optimal as can be seen from the omniscient recommendation case.
-In the no recommendation case since consumers do not know the common-value component they engage in local consumption in different areas of the product space which leads to less than optimal homogeneity.
+In the no recommendation case since users do not know the common-value component they engage in local consumption in different areas of the product space which leads to less than optimal homogeneity.
 
@@
 The classic approach to recommender system evaluation is to predict user ratings for items and compare how accurate this prediction is to recorded ratings data, either explicitly given by users or inferred from behavioral data.
-Further, given this predicted rating, the recommender system should recommend to consumers the items with the highest predicted rating \cite{adomavicius2005toward}.
+Further, given this predicted rating, the recommender system should recommend to users the items with the highest predicted rating \cite{adomavicius2005toward}.
 
@@
 Recall from Findings \ref{finding_local_consumption} and \ref{finding_diversity_welfare_corr} that, particularly if users are risk-averse, they will consume content in increasingly narrow portions of the product space and this may be detrimental to their realized welfare.
-As a result, identifying the portions of the product space where users may have significant uncertainty and utilizing recommendation to reduce this uncertainty can be welfare-enhancing for consumers.
+As a result, identifying the portions of the product space where users may have significant uncertainty and utilizing recommendation to reduce this uncertainty can be welfare-enhancing for users.
 This is especially the case if this reduction of uncertainty leads users to explore under-explored portions of the product space where true realized utility is high, but expected utility, given beliefs, is low before recommendation.
@@
 
-Furthermore, our model shows the importance of understanding the underlying user preferences and the nature of the product space in the setting in which a recommender system is deployed since that will change the consumption patterns of consumers.
+Furthermore, our model shows the importance of understanding the underlying user preferences and the nature of the product space in the setting in which a recommender system is deployed since that will change the consumption patterns of users.
 For instance, understanding the degree of correlation of utilities and the level of risk-aversion become crucial in understanding how users will make consumption choices without recommendation and what kind of recommendation would be useful for users.

### 2019-09-16 paper/filter_bubbles.tex
made some sentences red

@@
 
+\usepackage{color-edits}
 \usepackage{xspace}
@@
 \item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ 
-\item $V \sim \mathcal N(\overline V, \Sigma)$ with $\overline V =0$
+\item $V \sim \mathcal N(\overline V, \Sigma)$ with \ssedit{$\overline V =0}{red}$
 \end{enumerate}
@@
 We impose the normality assumption for two reasons.
-The first is that users update their beliefs using Bayesian updating and recall that the normal distribution forms a conjugate family, which allows for simple posterior updates.
-The second is that it allows us to incorporate an easily interpretable correlation structure between the items.
+\ssedit{The first is that users update their beliefs using Bayesian updating and recall that the normal distribution forms a conjugate family, which allows for simple posterior updates.}{red} The second is that it allows us to incorporate an easily interpretable correlation structure between the items.
 

### 2019-09-17 paper/filter_bubbles.tex
updates to paper

--- 2019-09-16 paper/filter_bubbles.tex
@@
 
+\usepackage{color-edits}
+\addauthor{ss}{red} 
 \usepackage{xspace}
@@
 
-Recommender systems have been influential in shaping user choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.
-While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems.
+Recommender systems have been influential in shaping user choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{https://www.mckinsey.com/industries/retail/our- insights/how-retailers-can-keep-up-with-consumers} While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems.
 There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals\footnote{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350/}, that personalized recommender systems lead users into \textit{filter bubbles} where they get effectively isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
 
-\textbf{Our Model and Contributions} In this paper, drawing from recent work in psychology and decision theory in economics, we first develop a model of user decision-making in the context of large choice sets faced by users in online markets.
-We then ask how a stylized model of recommendation can shape the behavior of long-lived users and finally ask how the insights from our model can be used to improve recommender system design and to better understand the consequences of these systems.
+\textbf{Our Model and Contributions} In this paper, drawing from recent work in psychology and decision theory in economics, we first develop a model of user decision-making of users in online markets \textit{without recommendation} and investigate whether the negative effects associated with recommender systems arise even in the absence of recommendation.
+We then ask how a stylized model of recommendation can shape the behavior of long-lived users and under what conditions it amplifies or deflates these effects.
+Finally, we use the insights from our model to improve the design and evaluation of recommender systems.
 
-We focus on understanding the degree to which recommender systems can lead users into \textit{filter bubbles}.
-Consistent with recent empirical work in the case of movie consumption \cite{nguyen2014exploring}, our model shows that individual consumption behavior over time naturally exhibits patterns consistent with a filter bubble effect \textit{without recommendation}.
-The key component of our model, drawn from the results of \cite{schulz2019structured}, is that the utilities of similar goods are correlated.
-This means that when users consume a good and learn its true utility this gives them information about similar goods.
+The key components of our model are:
+\begin{enumerate}
+\item Users have uncertainty about their preferences over goods and thus face a decision problem under uncertainty.
+\item Users face large choice sets and only consume a small portion of this choice set over their lifetime.
+\item Users are long-lived and sequentially consume items.
+Consumption of an item has learning spill-overs where consumption gives information about the utilities of similar items in the product space.
+\item Recommendation provides users with recommendation to reduce their uncertainty.
+\end{enumerate}
+Using this model we first investigate the degree to which recommender systems can lead users into filter bubbles.
+Our model rationalizes the empirical results in \cite{nguyen2014exploring} who show that, in the context of movie consumption, user behavior is consistent with filter bubble effects even without recommendation.
+The key component of our model that drives this result is that the utilities of similar goods are correlated and that, when users consume a good, it gives them information about similar goods.
 Crucially, this not only impacts the underlying belief about the average utility of the good, but also the amount of uncertainty.
-This learning spill-over induces users to consume goods ``similar" to those they consumed before leading to an increasing narrowing of consumption patterns.
+This learning spill-over induces users to consume goods ``similar" to those they consumed before that had high realized utility leading to an increasing narrowing of consumption towards these regions of the product space.
 This effect is further exemplified when users are \textit{risk-averse}, a concept from economic decision theory where, all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty.
+We consider recommendation as providing information to consumers in order to reduce their uncertainty and we find that this induces them to explore a more diverse set of products, which is consistent with the empirical results from \cite{nguyen2014exploring}.
 
-We then consider a stylized model of user recommendation and explore how recommendation can affect consumption patterns.
-We suppose each good is a sum of a common value component and an idiosyncratic component, where the idiosyncratic component is inherently unpredictable given other individual's preferences.
-We model existing recommendation systems as providing users with information on the common-value component and denote as this as partial recommendation.
-We consider how user behavior varies as we move from no recommendation to partial recommendation to omniscient recommendation, which is the optimal consumption path for users if they knew the true utility of all the goods in their choice set.
+We further explore how diverse the overall set of items that users consume is and find that, without recommendation, the diversity of items consumed is negatively correlated with welfare.
+This is due to the fact that, once users find good items, they consume items nearby in the product space due to the correlation of utilities.
+When they consume bad products, they move far away in the product space leading to a more diverse set of consumed products but lower welfare.
+However, when users are very risk-averse we find that they may stay in small portions of the product space.
+Thus, we find that consumption set diversity is not by a itself a useful measure for understanding the impact of recommender systems on users.
 
-Consistent with \cite{nguyen2014exploring}, we find that partial recommendation increases consumption diversity. However, we also find that partial recommendation increases homogeneity amongst users.
+We find that recommendation can lead to increases in homogeneity amongst users and that, without recommendation, users are almost fully inhomogenous.
+We model the realized utilities of users as being a weighted sum of a common-value and idiosyncratic component.
+The recommender can only learn the common-value component of utility since the idiosyncratic component is inherently unpredictable given other individuals' preferences.
+The recommender uses its knowledge of the common-value component and individuals' beliefs over the product space when designing personalized recommendation.
+This leads to an increase in welfare for users, but concentrates their consumption around items that have a high common-value component which leads to an increase in homogeneity.
+
+Finally, we apply our model and findings to inform the design and evaluation of recommender systems.
+Our evaluation metric relies on the intuition that a recommendation is only useful insofar as it leads a user to take a better action than she would without recommendation.
+Thus, in order to evaluate the value of recommendation it is important to be able to predict what a user would do without recommendation and how recommendation would change their behavior.
+This requires collecting data that is not traditionally collected by recommender systems - the beliefs that users have of the utilities of goods and understanding the degree of uncertainty users have over different portions of the product space.
 
@@
 \item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ 
-\item $V \sim \mathcal N(\overline V, \Sigma)$ with $\overline V =0$
+\item $V \sim \mathcal N(\overline V, \Sigma)$ with \ssedit{$\overline V =0$}
 \end{enumerate}
@@
 We impose the normality assumption for two reasons.
-The first is that users update their beliefs using Bayesian updating and recall that the normal distribution forms a conjugate family, which allows for simple posterior updates.
-The second is that it allows us to incorporate an easily interpretable correlation structure between the items.
+\ssedit{The first is that users update their beliefs using Bayesian updating and recall that the normal distribution forms a conjugate family, which allows for simple posterior updates.} The second is that it allows us to incorporate an easily interpretable correlation structure between the items.
 

### 2019-10-01 paper/filter_bubbles.tex
remove unused package
updated paper

--- 2019-09-16 paper/filter_bubbles.tex
@@
-\documentclass[sigconf]{acmart}
+\documentclass[sigconf, anonymous, review]{acmart}
 \settopmatter{printacmref=false}
@@
 %\usepackage{lmodern}
-\usepackage[font=small]{caption}
 \usepackage{float}
+\usepackage{enumitem}
 \newtheorem{finding}{Finding}
-
-\usepackage{xspace}
-\newcommand{\OMIT}[1]{}
 %
@@
 % These commands are for a PROCEEDINGS abstract or paper.
-\copyrightyear{2019}
-\acmYear{2019}
+\copyrightyear{2020}
+\acmYear{2020}
 \setcopyright{acmlicensed}
-\acmConference[WWW '20]{Recomme '18: ACM Symposium on Neural Gaze Detection}{April 20--24, 2020}{Taipei, Taiwan}
-\acmBooktitle{WWW '20: ACM Conference on Neural Gaze Detection, June 03--05, 2018, Woodstock, NY}
+\acmConference[WWW'20]{Recomme '18: ACM Symposium on Neural Gaze Detection}{April 20-24, 2020}{Taipai, Taiwan}
+\acmBooktitle{RecSys '19: ACM Conference on Neural Gaze Detection, June 03--05, 2018, Woodstock, NY}
 \acmPrice{15.00}
@@
 % The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
-\title{The Ex-Ante View of Recommender System Design}
+\title{A User-Centric Approach to the Design and Consequences of Recommender Systems}
 %
@@
 \begin{abstract}
+We study a model of user decision-making in the context of recommender systems.
+We explore the extent to which issues traditionally associated with recommender systems -- namely filter bubbles, user homogenization and consumption diversity -- can be attributed to recommendation systems.
+The key component of our model is that users face a decision problem under uncertainty and that there are informational spillovers where a user consuming a product and learning how she values it reduces uncertainty about similar products.
+We find that this generates behavior consistent with narrowing filter bubble effects and that recommendation can help alleviate these effects, as has been found in the empirical work of Nguyen et al.
+(2014).
+However, we find that recommendation, while alleviating filter bubble effects, leads to an increase in user homogeneity.
+\par
+
+We discuss the evaluation of recommender systems in light of these findings.
+Traditional design approaches to recommender systems only optimize for predicting user preferences after uncertainty has been resolved.
+We propose an alternative interpretation of the serendipity measure where good recommendations are those that provide useful information, that is, information without which choice would be different and lead to worse outcomes on average relative to the case where no recommendation is provided.
+As a result, we emphasize the importance of understanding user behavior without recommendation, both for designing recommendations and understanding their impacts and usefulness.
 \end{abstract}
@@
 % presented. Separate the keywords with commas.
-\keywords{Recommender Systems, Beliefs, Decision Theory, Filter Bubbles}
+%\keywords{Recommender Systems, Filter Bubbles, Serendipity}
 
@@
 
-Recommender Systems (RS) have become critical for assisting users in navigating the large choice sets that they face in many online markets.
-For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
-users in many cases are not aware of most items, let alone have full information about their preferences over them and, to make matters worse, the goods in these markets are usually experience goods whose true utility can only be learned after consumption.
-Furthermore, users interact with these systems routinely and watch more than one movie on Netflix, buy more than one product on Amazon, and listen to more than one video on YouTube.
+Recommender Systems (RS) have become critical for assisting consumers in navigating the large choice sets that they face in many online markets.
+For instance, consumers have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
+Consumers in many cases are not aware of most items, let alone have full information about their preferences over them and, to make matters worse, the goods in these markets are usually experience goods whose true utility can only be learned after consumption.
+Furthermore, consumers interact with these systems routinely and watch more than one movie on Netflix, buy more than one product on Amazon, and listen to more than one video on YouTube.
+\par
 
-Recommender systems have been influential in shaping user choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.
-While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems.
-There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals\footnote{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350/}, that personalized recommender systems lead users into \textit{filter bubbles} where they get effectively isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
+Recommender systems have been influential in shaping consumer choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{\url{https://www.mckinsey.com/industries/retail/our-insights/how-retailers-can-keep-up-with-consumers}} While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems.
+There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals,\footnote{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350/} that personalized recommender systems lead consumers into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead consumers to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
+\par
 
-\textbf{Our Model and Contributions} In this paper, drawing from recent work in psychology and decision theory in economics, we first develop a model of user decision-making in the context of large choice sets faced by users in online markets.
-We then ask how a stylized model of recommendation can shape the behavior of long-lived users and finally ask how the insights from our model can be used to improve recommender system design and to better understand the consequences of these systems.
+\textbf{Our Model and Contributions} In this paper, 
+%drawing from recent work in psychology and decision theory in economics, 
+we first propose a model of consumer decision-making in the context of large choice sets faced by consumers in online markets and show that the model is able to rationalize features of user behavior in such contexts by means of simulations.
+We then ask how a stylized model of recommendation can shape the behavior of long-lived consumers 
+and finally ask how the insights from our model can be used to improve recommender system design and to better understand the consequences of these systems.
+\par
 
-We focus on understanding the degree to which recommender systems can lead users into \textit{filter bubbles}.
-Consistent with recent empirical work in the case of movie consumption \cite{nguyen2014exploring}, our model shows that individual consumption behavior over time naturally exhibits patterns consistent with a filter bubble effect \textit{without recommendation}.
-The key component of our model, drawn from the results of \cite{schulz2019structured}, is that the utilities of similar goods are correlated.
-This means that when users consume a good and learn its true utility this gives them information about similar goods.
-Crucially, this not only impacts the underlying belief about the average utility of the good, but also the amount of uncertainty.
-This learning spill-over induces users to consume goods ``similar" to those they consumed before leading to an increasing narrowing of consumption patterns.
-This effect is further exemplified when users are \textit{risk-averse}, a concept from economic decision theory where, all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty.
+The key components of our model are:
+\begin{enumerate}[label=(\arabic*)]
+\item Users face uncertainty about their valuation of the different goods and thus face a decision problem under under uncertainty.
+\item Choice sets are large and users only consume a small fraction of this choice set over their lifetime.
+\item Consumption of these items is sequential and users live for multiple periods.
+\item Consuming an item provides information about the value of similar items in the product space.
+\item Recommendation provides users with information, reducing their uncertainty.
+\end{enumerate}
 
-We then consider a stylized model of user recommendation and explore how recommendation can affect consumption patterns.
-We suppose each good is a sum of a common value component and an idiosyncratic component, where the idiosyncratic component is inherently unpredictable given other individual's preferences.
-We model existing recommendation systems as providing users with information on the common-value component and denote as this as partial recommendation.
-We consider how user behavior varies as we move from no recommendation to partial recommendation to omniscient recommendation, which is the optimal consumption path for users if they knew the true utility of all the goods in their choice set.
+Using this model we first investigate the degree to which recommender systems lead users into filter bubbles.
+Our model rationalizes the empirical results in \cite{nguyen2014exploring} who show that in the context of movie consumption, user behavior is consistent with filter bubble effects \textit{even without recommendation}.
+The driving force of this in our model is that utilities of similar goods are correlated, which implies that when a good is consumed and the user learns its value, it provides information about similar goods.
+Crucially, this not only impacts the underlying belief about the average utility of the good, but also the level of uncertainty.
+Consequently, this learning spillover induces users to consume goods ``similar'' to those they consumed before that had high realized value, leading to an increasing narrowing of consumption patterns towards these regions of the product space.
+This effect is further amplified when users are \textit{risk-averse}, a concept from decision theory where all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty.
+We compare this benchmark to the case where recommendations are provided.
+We consider recommendations as giving information to consumers in order to reduce their uncertainty and we find that this reduces the narrowing effect, which is also consistent with the empirical results from \cite{nguyen2014exploring}.
+\par
 
-Consistent with \cite{nguyen2014exploring}, we find that partial recommendation increases consumption diversity. However, we also find that partial recommendation increases homogeneity amongst users.
+We then explore how diverse the overall set of items that users consume is and find that, without recommendation, higher diversity of items does not ensure higher welfare.
+This is due to the fact that once users find good items, they consume items nearby in the product space due to the correlation of utilities, leading to higher overall value from consumption.
+When they consume products they value little, they move to another part of the product space.
+A sequence of bad realizations will then lead to very intense search over very different parts of the product space and high consumption diversity, whereas discovering a sweet spot leads to highly rewarding local search and very low diversity.
+Thus we find that consumption set diversity is not by itself a useful measure for understanding the impact of recommender systems on users.
+\par
+
+We find that recommendation can lead to increases in homogeneity among users.
+We model the realized valuations as being a weighted sum of a common-value and an idiosyncratic component.
+This formulation gives a stylized notion of predictability of user preferences where the idiosyncratic component is inherently unpredictable given other individuals' preferences and the common-value component is what the recommonder can learn from previous users' data.
+The recommender uses its knowledge of the common-value component and the individuals' beliefs over the product space when designing personalized recommendation.
+This leads to an increase in welfare for users, but concentrates their consumption around items that have a high common-value component, resulting in an increase in homogeneity.
+\par
+
+Finally, we discuss how our model and findings can be used to inform the design and evaluation of recommender systems.
+Our evaluation metric relies on the intuition that a recommendation is only useful insofar as it leads a user to take a better action than she would without the recommendation.
+Thus, in order to evaluate the value of recommendation, it is important to be able to predict what a user would do without recommendation and how recommendation would change their behavior.
+If recommendations fail to affect behavior, they are of no use: the user would consume the same good and learn its value with and without the recommendation.
+Moreover, we argue that successful recommendation systems have to take into account the degree of uncertainty users have over different portions of the product space.
+Being able to predict valuations after consumption is not sufficient as uncertainty over the idiosyncratic component of one's own preferences may steer the user away from the recommended items in face of risk-aversion.
+\par
 
 \textbf{Related Work.} 
-Recommender Systems
+\cite{pariser2011filter} first informally described the ``filter bubble'' which is the idea that online personalization services would lead users down paths of increasingly narrower content so that they would effectively be isolated from a diversity of viewpoints or content.
+A number of studies, done in a wide range of disciplines, have since empirically studied whether or not this phenomenon exists in various contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}.
+The most relevant to our proposed study is \cite{nguyen2014exploring} who study whether this effect exists with recommender systems in the context of movie consumption.
+They find that even users whose consumption choices are not guided by recommendations fall into behavior consistent with ``filter bubbles'' and that recommender systems can actually increase the diversity of the content that users consume.
+To our knowledge there are no theoretical models that rationalize the empirical findings but predominantly empirical studies.
+Our paper provides a theoretical framework through which to view this problem.
+The second question is to what extent recommender systems can lead users to become increasingly homogenized or consume similar content as each other.
+\cite{celam2008hits,treviranus2009value} show that incorporating content popularity into recommender systems can lead to increased user homogenization.
+\cite{chaney2018algorithmic} more generally look at the issue of user homogenization in recommender systems.
+Our results are similar to previous work, where the limits of recommendation lead the recommender to over-recommend products from certain portions of the product space which leads users to consume goods in those regions.
+\cite{fleder2009blockbuster} also consider the effects of recommendation on multi-period user behavior, but predominantly focus on sales diversity and employ a different model of user behavior than we do here.
+\cite{hodgson2019horse} consider a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of utilities in the environment, but consider it in the context of search for a single item.
+Our discussion of recommender system evaluation contributes with insights closely tied to the value of information that is used in the economics literature (see e.g.
+\cite{bergemann2019information}) and builds off the large literature stemming from \cite{mcnee2006being} that makes the case that accurate predictions of true realized valuations may not lead to the best recommendations.
+This is related to the notion of serendipity in recommender systems that has been studied extensively (e.g.
+see \cite{kotkov2016survey} for a survey).
 
@@
 
-\subsection{Preliminary on Expected Utility Theory}
-[Duarte can do this]
+\subsection{Preliminaries on Expected Utility Theory}
+\noindent For every product $n$ in the product space $\mathcal J=\{0,1,....,N-1\}$, we assume that each consumer $i$ assigns a monetary equivalent $x_{i,n} \in \mathbb R$ to the experience of consuming it.
+Each consumer can value the same product in differently.
+However, we assume that consumers have the same utility over money, given by a utility function $u: \mathbb R \to \mathbb R$, strictly increasing and continuous.
+So, ex-post, the value of product $n$ for consumer $i$ is given by $u(x_{i,n})$ However, before consuming the product, the consumer does not know exactly how she will value it.
+In particular, even consumers that will end up having the same ex-post valuation of product $n$ may differ in their ex-ante valuation because they hold different beliefs about it.
+We denote by $p_{i}$ the beliefs consumer $i$ has about how she will value each of the products in the product space.
+Note that this implies that consuming product $n$ is the same as taking a gamble.
+Each individual evaluates the product according to the associated expected utility associated with the product, i.e.
+$U_i(n)=\mathbb E_{p_i}[u(x_n)]$.
+\par
 
-\subsection{Model Setup}
+Risk aversion captures the how different individuals react to the risk associated to a particular consumption opportunity.
+It is formalized as follows: Suppose for a given gamble $x$ taking real values and following distribution $p$, there is a certain amount of money that leaves the individual indifferent between taking the gamble and taking this sure amount, call it $\delta(x)$, termed the \textit{certainty equivalent} of $x$.
+An individual $i$ is more risk-averse than another individual $j$ when for every gamble $x$, individual $i$ requires a lower value to give up the gamble when compared to individual $j$, i.e.
+$\delta_i(x)$.
+Therefore, a more risk-averse individual is more willing to avoid the risk of taking the gamble.
+We assume that the utility function takes a flexible functional form $u(x)=1-\exp(-\gamma x)$ for $\gamma\ne0$ and $u(x)=x$ for $\gamma=0$ -- known as constant absolute risk-aversion preferences.
+Higher $\gamma$ implies higher risk-aversion with $\gamma=0$ corresponding to the risk-neutral case and $\gamma>0$ to the risk-averse one.
 \par
-\noindent \textbf{users} We consider a set of users $I$ where each individual $i \in I$ faces the same finite set of $N$ items $\mathcal{N} = \left\{0,1,...,N-1\right\}$.
-For simplicity, we assume that individuals only derive pleasure from item $n \in \mathcal{N}$ the first time they consume it.
 
-We denote by $u_{in}$ individual $i$'s realized utility from consuming item $n$.
-In particular, we consider that the realized utility derived from a given item can be decomposed in the following manner:
-\begin{align*}
-u_{in}= v_{in} + \beta v_n
-\end{align*}
-where $v_{in}$ denotes an idiosyncratic component -- i.e.
-user $i$'s idiosyncratic taste for good $n$ --  and $v_{n}$, a common-value component.
-One can interpret $v_n$ as a measure of how much good $n$ is valued in society in general and, in a sense, $v_{in}$ denotes how $i$ diverges from this overall ranking.
-The scalar $\beta \in \mathbb{R}_{+}$ denotes the degree to which utilities are idiosyncratic to each individual or common across individuals.
-If $\beta=0$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta \rightarrow \infty$, every individual has the same preferences.
-[include some discussion of what this implies about the predictability of a user's preferences]
+\subsection{Model}
+\par
+\noindent \textbf{Users} We consider a set of users $I$ where each individual $i \in I$ faces the same finite set of $N$ items $\mathcal J = \left\{0,1,...,N-1\right\}$.
+For simplicity, we assume that individuals only derive pleasure from item $n \in \mathcal{J}$ the first time they consume it.
+\par
 
-Stacking utilities in vector-form, we get:
-\begin{align*}
-{\left(u_{in}\right)}_{n \in \mathcal{N}}=:U_i=V_i+ \beta V
-\end{align*}
-We stack the idiosyncratic and common-value components as $V_i ={\left(v_{in}\right)}_{n \in \mathcal{N}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{N}}$.
+We denote by $x_{i,n}$ individual $i$'s realized value from consuming item $n$.
+In particular, we consider that the realized utility derived from a given item can be decomposed in the following manner: $x_{i,n}= v_{i,n} + \beta v_n$, where $v_{i,n}$ denotes an idiosyncratic component -- i.e.
+user $i$'s idiosyncratic taste for product $n$ --  and $v_{n}$, a common-value component.
+One can interpret $v_n$ as a measure of how much product $n$ is valued in society in general and, in a sense, $v_{i,n}$ denotes how $i$ diverges from this overall ranking.
+The scalar $\beta \in \mathbb{R}_{t}$ denotes the degree to which utilities are idiosyncratic to each individual or common across individuals.
+If $\beta=0$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta$ is large, every individual has similar preferences.
 \par
-\noindent \textbf{user Beliefs} We assume that users do not know the realized utilities of the goods before consuming an item.
-Formally, user $i$ starts with some beliefs about $U_i$, namely that the idiosyncratic and common-value parts of the utilities are independent -- $V_i \perp \!\!\!
-\perp V$ -- and that each is multivariate normal.
-The idiosyncratic utility and common-value utility are distributed as follows:
+Stacking utilities in vector-form, we get the vector of values associated with each item $${\left(x_{i,n}\right)}_{n \in \mathcal{N}}=:X_i =V_i+ \beta V $$ where $V_i ={\left(v_{i,n}\right)}_{n \in \mathcal{N}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{N}}$.
+\par
+\noindent\textbf{User Decision-Making}
+We assume the user makes $T$ choices and therefore can only consume up to $T$ items, where $T$ is but a small fraction of $N$.
+This captures the idea that users are faced with an immense choice set but that ultimately they end up experiencing (and learning) about just a small fraction of it.
+Since this is a sequential decision-making problem under uncertainty, in principle consumers face an exploration-exploitation trade-off.
+However, for tractability, we impose that consumers are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
+\par
+\noindent \textbf{User Beliefs} We assume that consumers do not know the realized utilities of the goods before consuming an item.
+Formally, consumer $i$ starts with some beliefs about $X_i$, namely that the idiosyncratic and common-value parts of the utilities are independent -- $V_i \perp \!\!\!
+\perp V$ -- and that each is multivariate normal
 \begin{enumerate}
@@
 We impose the normality assumption for two reasons.
-The first is that users update their beliefs using Bayesian updating and recall that the normal distribution forms a conjugate family, which allows for simple posterior updates.
+The first is that consumers update their beliefs using Bayesian updating and recall that the normal distribution forms a conjugate family, which allows for simple posterior updates.
 The second is that it allows us to incorporate an easily interpretable correlation structure between the items.
+\par
+Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{i,n}\right)_{n \in \mathcal{J}}$ are drawn independently from a jointly normal distribution, where $\overline v_{i,n} \sim \mathcal N (0, \overline \sigma^2)$ are independent and identically distributed.
+These $\overline v_{i,n}$ denote the prior belief that individual $i$ holds about the her valuation over good $n$.
+As people are exposed to different backgrounds, their beliefs about what is good for them also varies and $\overline v_{i,n}$ denotes this idiosyncrasy at the level of prior beliefs.
+\par
 
-Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{in}\right)_{n \in \mathcal{N}}$ are drawn independently from a jointly normal distribution, where $\overline v_{in} \sim \mathcal N (0, \overline \sigma^2)$ are i.i.d.
-These $\overline v_{in}$ denote the prior belief that $i$ holds about the her valuation over good $n$.
-As people are exposed to different backgrounds, their beliefs about what is good for them also varies and $\overline v_{in}$ denotes this idiosyncrasy at the level of prior beliefs.
+We assume users are expected utility maximizers.
+User $i$'s certainty equivalent for product $n$, the sure value that makes user $i$ indifferent between it and consuming the product $n$, conditional on the consumption history, is given by
+$\delta_{i}(n)\mid C_i^{t-1}=\mu_n-\frac{1}{2}\gamma \Sigma_{nn}$, where $\mu_n$ and $\Sigma_{nn}$ are the expected value and variance for item $n$ that the user has given their initial beliefs and consumption history up until time $t$.
+
+\noindent \textbf{User Learning}
+When a consumer consumes a product $n$ they learn the realized value for that product.
+In our model, we incorporate the idea that learning the value of product $n$ gives a consumer information about similar items.
+%This is drawn from recent empirical evidence in \cite{schulz2019structured} that consumers learn how to navigate large choice sets using similarity-based generalization. 
+We assume that learning about the value of good $n$ reveals more about the value associated to items that are closer to it, which captures the idea that trying a product provides more information about similar products than about dissimilar ones.
 \par
-\noindent \textbf{user Learning}
-When a user consumes a good $n$ they learn the realized utility for that good.
-In our model we incorporate the idea that learning the utility of good $n$ gives a user information about similar items.
-We assume that learning about the utility of good $n$ reveals more about the utility associated to items that are closer to it, which captures the idea that trying a product provides more information about similar products than about dissimilar ones.
 
-In order to have a well-defined notion of similarity we need to define a distance function between goods, which we define as:
-\begin{align*}
-d(n,m):=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}
-\end{align*}
- $m$ and $n$ are the indices of the items in $\mathcal{N}$.
+In order to have a well-defined notion of similarity we need to define a distance function between goods, which we define as $d(n,m):=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}$ where $m$ and $n$ are indices of items in $\mathcal{J}$.
+This distance function is not intended to model the intricacies of a realistic product space, but instead to provide a stylized product space to help us understand the effects of informational spillovers on user behavior.
 We consider that the entry of $n$-th row and the ($m$)-th column of $\Sigma_i$ is given by $\sigma_i^2 \rho^{d(n,m)}$, and that of $\Sigma$ is given by $\sigma^2 \rho^{d(n,m)}$.
@@
 Informativeness, for any $\rho \in (0,1)$, is decreasing in distance.
-The particular distance function that we utilize leads to this covariance structure being simple, where the $(n,n+1)$-th entry in the covariance matrix is $\rho$ , $(n,n+2)$-th entry is $\rho^2$, etc.
+The particular distance function that we utilize leads to this covariance structure being simple, where the $(n,n+1)$-th entry in the covariance matrix is $\rho$, the $(n,n+2)$-th entry is $\rho^2$, etc.
 \par
-\textbf{user Decision-Making}
-We assume the user makes $T$ choices and therefore can only consume up to $T$ items, where $T$ is but a small fraction of $N$.
-This captures the idea that users are faced with an immense choice set but that ultimately they end up experiencing (and learning) about just a small fraction of it.
-For tractability we impose that users are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
-Any ties are broken uniformly at random.
-This assumption is critical not only for tractability but also in order to have easily defined benchmarks.
+\begin{center}
+\includegraphics[width=.4\linewidth]{Example-Bubbles.pdf}
+\captionof{figure}{Illustrative Example}
+\label{fig:illustrative_example}
+\end{center}
+
+\noindent \textbf{An Illustrative Example} We illustrate our model with a simple example.
+Suppose that there are four products: 0, 1, 2, 3.
+The products are in different places of the product space, where 1 is close to 0 and 2 but more distant from 3, as shown in Figure \ref{fig:illustrative_example}.
+Suppose that the user actually believes that they all have expected value of 0 and consumes product 1.
+The underlying correlation structure implies that upon observing that $x_1=y$ the user will update beliefs about $\mathbb E x_0$, $\mathbb E x_2$ and $\mathbb E x_3$ to $\rho y$, $\rho y$ and $\rho^2 y$ respectively.
+This is because learning about product 1's value is more informative about similar products' value, 0 and 2, than products further away in the product space, 3.
+Therefore, the user will adjust her beliefs about their mean -- zero in the example -- more strongly for products which are similar.
+Moreover, now uncertainty about the value of products 0, 2 and 3 is also decreased but differentially so: learning product 1's value has contributed to a stronger reduction of uncertainty for nearby products that for product 3.
+If the actual value was positive, this induces the user to perform some local exploration, as 0 and 2 have higher mean and lower variance.
+If instead it was negative, then user's posterior beliefs penalize more strongly products 0 and 2 than 3, so if she is not too risk averse, she would go across the product space in search of some better element.
+However, if risk aversion is high enough, the user might be willing to trade-off ex-ante expected lower values for lower risk and stick to local exploration of products that have just because the user has less uncertainty about their values.
+As the saying goes, if you are risk-averse, better the devil you know.
 \par
+
 \noindent \textbf{Recommendation}
-Our model of recommendation is stylized in order to provide qualitative insights into how recommendation shapes behavior, instead of looking at realistic implementations of recommender systems.
-We model recommendation as giving users information about the utility of the goods.
+Our model of recommendation is stylized in order to provide qualitative insights into how recommendation shapes behavior, instead of focusing on the details of how recommender systems are implemented in practice.
+We model recommendation as giving consumers information about the valuation of the products.
+\par
 
-We will consider three recommendation regimes:
-\begin{enumerate}
-\item Partial Recommendation
-\item No Recommendation
-\item Omniscient Recommendation
-\end{enumerate}
-The case of primary interest is \textit{partial recommendation} where the recommender observes utilities accrued, but does not know user $i$'s starting beliefs $\overline U_i$.
-In this case, the recommender knows $V$ but does not know $V_i$.\footnote{We do not consider the acquisition of information for the recommender to know $V$ and suppose that she has sufficient data to learn $V$ before recommendation.} However, the recommender does know the users' beliefs, $\overline V_i$.
-Thus, at any given period, the recommender provides a personalized recommendation that combines the knowledge of the common value component $V$ with the user beliefs $\overline V_i$.
-User's beliefs about her valuation of each item become crucial in this case: just providing the user with information about $V$ may change the user's original ranking, but, without considering the user's beliefs, she will not necessarily follow recommendation.
+We will consider three cases.
+The case of primary interest is \textit{recommendation} where the recommender observes values accrued and knows $V$ but does not know neither $V_i$.\footnote{We do not consider the acquisition of information for the recommender to know $V$ and suppose that she has sufficient data to learn $V$ with arbitrary precision.} However, the recommender does know the users' beliefs $\bar V_i$.
+Thus, at any given period, the recommender provides a personalized recommendation that combines the knowledge of the common value component $V$ with the user beliefs $\bar V_i$.
+Knowing the user's beliefs about her valuation of each item become crucial in this case: just providing the user with information about $V$ may change the user's original ranking, but, without considering the user's beliefs, she will not necessarily follow the recommendation.
+\par
 
-The notion of partial recommendation that we consider is idealized where the recommendation does the Bayesian updating for users.
-An alternative that is equivalent would be the recommender simply providing $V$ without knowing $\overline V_i$.
-In this case, the recommendation will be that the user $i$ chooses $r_{it} \in \arg \max_{n \in \mathcal{N} \backslash C_i^T} u_n$, but we assume the recommender provides full information about $V$.
-For instance, the recommender could display the whole distribution of utilities reported by other users or even its average, which is a good proxy for the common value component.\footnote{
-The best item that could recommended with such information if the item with highest common value component.
+The notion of recommendation that we consider is idealized where the recommendation does the Bayesian updating for users.
+As users face a very difficult and costly Bayesian updating instance, a useful recommendation system provides recommendations taking into account this aspect and combines the history of past consumption and value realizations with the user's beliefs.
+An alternative that is equivalent would be the recommender simply providing $V$ without knowing $V_i$.
+In this case, the recommendation will be that the user $i$ chooses $r_{i,t} \in \arg \max_{n\in \mathcal J\backslash C^T_i} \mathbb E[u(x_n) \mid C_i^T]$
+but we assume the recommender provides full information about $V$.
+For instance, the recommender could display the whole distribution of utilities reported by other users or even its average, which is a good proxy for the common value component\footnote{It is possible for the recommender to only recommend the best item given such information, that is, the item with highest common value component.
 However, in that case, recommending only a single item generates costly Bayesian updating to the user and provides little guidance as to what she should indeed pick as the common value might be of little importance when compared to the idiosyncratic component.
-Therefore, we assume that the RS reports the whole $V$, which results in higher expected welfare in choices and is not costly to implement.
-} The results would be equivalent, but this would shift the burden of Bayesian updating to users.
+Therefore, we assume that the recommender reports the whole $V$, which results in higher expected welfare in choices and is not costly to implement.}.
+The results would be equivalent, but this would shift the burden of Bayesian updating to users.
+\par
 
@@
 This gives us a benchmark as to how users in our model would behave \textit{without} recommendation so that we can analyze what changes with the introduction of recommendation.
-The second is \textit{omniscient recommendation} where the recommender knows the true realized utility of each good for each user and can therefore recommend the best remaining good in every period.
+The second is the \textit{oracle recommendation} where the recommender knows the true realized utility of each good for each user and can therefore recommend the best remaining good in every period.
 This gives us a full information benchmark, which is the optimal consumption path for a user if all uncertainty about their preferences was resolved.
 \par
+
 \noindent \textbf{Simulation Details}. 
@@
 The Gaussian assumption allows for closed form Bayesian-updating which enables us to simulate our model but does not provide much help in analytical characterizations.
+\par
 
 We explore how consumption patterns differ as we consider different recommendation regimes and report representative results from our simulations.
-We simulate our model over $100$ populations of users with $100$ users per population.
+We simulate our model over 100 populations of users with 100 users per population.
 For every population and every statistic we calculate, we average over the users in the population to get a representative statistic for this population.
 A given set of parameters and a population are a single data point in our dataset.
+\par
 
-We simulate over a grid of $\gamma, \sigma, \rho, \beta$.
-When we consider results varying a single parameter, we group the results over the other parameters and consider the differences in value only varying the parameter of interest.
-We report results for a relatively small $T$.
+We simulate over risk-aversion parameters $\gamma \in \{ 0, 0.3, 0.6, 1, 5 \}$, standard-deviation $\sigma \in \{ 0.25, 0.5, 1.0, 2.0, 4.0 \}$, correlation parameters $\rho\in \{ 0, 0.1, 0.3, 0.5, 0.7, 0.9 \} $ and degree of idiosyncrasy of individual valuations $\beta \in \{ 0, 0.4, 0.8, 1, 2, 5\}$.
+When we consider results varying a single parameter, we group the results over the other parameters and provide reports varying only the parameter of interest.
+We report results for a relatively small consumption history $T=20$ with a produce space size $N=200$.
 Unless otherwise reported, our results are robust to varying $N$.
 
-\subsection{Discussion of Assumptions}
+\subsection{Discussion of the Assumptions}
+In this section we provide further discussion of several assumptions of our model.
+\par
 
-In this section we provide further discussion of several of the assumptions of our model.
+\noindent \textbf{Uncertainty}: We assume that users have underlying uncertainty about the realized value of the items in their choice sets and as a result face a decision problem under uncertainty.
+The first justification for this is that environments where recommender systems are commonly deployed, such as on media streaming and e-commerce platforms, are mostly dedicated to experience goods, where users only learn their value after consumption.
+Standard decision theory axiomatized the conditions under which behavior is described by (subjective) expected utility maximizers \cite{savage1954} who update beliefs according to Bayesian inference \cite{ghirardato2002revisiting}, to which we add the assumption that their prior is correct.
+The second justification is that, even if users are in an environment where they could observe utilities based on product reviews or introspection, it is prohibitively costly for them to do so in large choice sets\footnote{Recent empirical work \cite{bronnenberg2014zooming,hodgson2019horse} has shown that, for product search, users engage in a ``zooming in'' search strategy in terms of the information they acquire about products before purchasing and only end up acquiring information about a small set of products in a large product space}.
+Even if they acquired all the information from product reviews, this would not fully resolve uncertainty as to the extent that individuals have idiosyncratic tastes.
+Giving that users are facing choices problems uncertainty, this motivates modeling recommendation as providing users \textit{information} in order to reduce their uncertainty.
+\par
 
-\noindent \textbf{Uncertainty}: We assume that users have underlying uncertainty about the realized utility of the items in their choice sets and as a result face a decision problem under uncertainty.
-The first justification for this is that environments where recommender systems are commonly deployed, such as on media streaming or e-commerce platforms, contain experience goods.
-A common assumption in the economics literature is that for experience goods, where users can only observe their realized utility after consumption, users face a decision problem under uncertainty when making consumption choices.
-The second justification is that, even if users are in an environment where they could observe utilities based on product reviews or introspection, it is prohibitively costly for them to do so in large choice sets.\footnote{Recent empirical work [cite the cameras paper and hodgson-lewis] has shown that, for product search, users engage in a ``zooming in" search strategy in terms of the information they acquire about products before purchasing and only end up acquiring information about a small set of products in a large product space.} Even if they acquired all the information from product reviews, this would only fully resolve uncertainty if individuals did not have idiosyncratic tastes.
-
-This motivates modeling recommendation as providing users \textit{information} in order to reduce their uncertainty.
-
-\noindent \textbf{Large Choice Sets}: We are interested in environments where the number of items in the choice set is very large and users only consume a small number of these items.
-This is realistic for platforms such as Netflix, Spotify, Amazon, and YouTube where there are sometimes billions of items and users can realistically only consume a handful of these items in their lifetime.
+\noindent \textbf{Large Choice Sets}: We are interested in environments where the number of items in the choice set is large and users only consume a small number of these items.
+This is realistic for platforms such as Netflix, Spotify, Amazon and YouTube where there are sometimes billions of items and users can realistically only consume a handful of these items in their lifetime.\par
 
 \noindent \textbf{Learning Spillovers}: Users in our environment face a sequential decision-making problem.
-However, we suppose that individuals are myopic (i.e.
-consume the item with the highest expected utility, given their beliefs in a single period) in order to maintain tractability.
-The predominant difficulty in the learning problem they face is not that they need to repeatedly sample a single item in order to learn its value (as in standard multi-armed bandit formulations), but rather that they need to be able to generalize from their experience from consuming one item to other items.
-After consuming a single item they learn the realized utility of the good.
+However, we suppose that individuals are myopic, i.e.
+they consume the item with the highest expected utility given their beliefs at each period.
+This is assumed in order to preserve tractability.
+The predominant difficulty in the learning problem they face is not that they need to repeatedly sample a single item in order to learn its value as in the standard multi-armed bandit formulations, but rather that they need to be able to generalize from their experience from consuming one item to other items.
+After consuming a single item, they learn the realized value of the good, but have no desire to consume it again, i.e.
+we assume that there is no value to consuming the item again.
+\par
 
-\indent \cite{schulz2019structured} studies how individuals solve sequential decision-making problems under uncertainty in large choice sets in the context of mobile food delivery orders.
-They find that individuals engage in similarity-based generalization where learning about the realized utility of a particular good provides them with information about similar goods.
-Content-based recommender systems exploit a similar idea where, if a user liked one item, then it is also likely that they will like a similar, but different, item as well.
+\cite{schulz2019structured} studies how individuals solve sequential decision-making problems under uncertainty in large choice sets in the context of mobile food delivery orders.
+They find that individuals engage in similarity-based generalizations where learning about the realized value of a particular good provides them with information about similar goods.
+Content-based recommender systems exploit a similar idea where, if a user liked a given item, then she is also likely to like a similar but different item as well.
+\par
 
-Our model incorporates this observation in a stylized fashion where the degree of information spill-overs is controlled via parameter $\rho$ and we study how the strength of the spill-overs impacts behavior and the effect of recommendation.
+Our model incorporates this observation in a stylized fashion, where the degree of information spillovers is controlled via parameter $\rho$ and we study how the strength of the spillovers impacts behavior and the effect of recommendation.
+\par
 
 \section{Results}
-
 \subsection{Local Consumption}
-
 \begin{figure}
-\includegraphics[scale=0.4]{figures/consumption_dist_N_200T_20no_correlation}
-\caption{Consumption path with no correlation between utilities}
-\label{fig:no_correlation_consumption_path}
+\includegraphics[width=.9\linewidth]{figures/rho_pos_consumption_dist_N_200T_20_overall.pdf}\\
+\includegraphics[width=.9\linewidth]{figures/rho_zero_consumption_dist_N_200T_20_overall.pdf}\\
+\caption{Consumption path with (Top) and without (Bottom) correlation between values}
+\label{fig:correlation_consumption_path}
 \end{figure}
 
-\begin{figure}
-\includegraphics[scale=0.1]{figures/consumption_dist_N_200T_20_overall}
-\caption{Consumption path between Recommendation Regimes}
-\label{fig:consumption_path_between_regimes}
-\end{figure}
-
-We characterize ``filter bubble" effects in our model as the degree to which users engage in \textit{local consumption}.
-We define local consumption in terms of the average consumption distance between the items consumed by the representative user at time $t-1$ and $t$.
+We characterize ``filter bubble'' effects in our model as the degree to which users engage in \textit{local consumption}.
+We define local consumption in terms of the average consumption distance between the items consumed by the users at time $t-1$ and $t$.
 We compare the average movement within the product space across different recommendation regimes and associate lower average distance with more local consumption.
@@
 \item When $\rho = 0$, there is no difference in consumption distance between the three recommendation regimes.
-\item When $\rho > 0$, no recommendation induces more local consumption than both partial and omniscient recommendation.
+\item When $\rho > 0$, no recommendation induces more local consumption than both recommendation and oracle regimes.
 This effect is amplified as $\rho$ increases as well as when users are more risk averse ($\gamma$ increases)
@@
 
-First, Figure \ref{fig:no_correlation_consumption_path} shows that, when $\rho = 0$, there is no difference in consumption distance between the three regimes.
+First, the bottom panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho = 0$, there is no difference in consumption distance between the three regimes.
 This is due to the fact that when $\rho = 0$, there is no reason that items that are close in the product space should have similar utilities and so the optimal consumption path does not depend on the similarity of the items.
-However this also means that users in the no recommendation regime do not learn anything about neighboring goods and so there is limited path-dependence in consumption.
+However this also means that users do not learn anything about neighboring products and so there is limited path-dependence in consumption.
+\par
 
-Figure \ref{fig:consumption_path_between_regimes} shows that, when $\rho > 0$, both partial and no recommendation lead to increasingly local consumption compared to the benchmark omniscient case.
-Further, the average consumption path between periods is \textit{decreasing} for the no recommendation case whereas it is \textit{increasing} for the omniscient case.
-Partial recommendation decreases the degree of local consumption but not as much as the omniscient benchmark.
-Due to the correlation of utilities, the omniscient consumption path exploits this and leads to the consumption of more similar products than in the case when $\rho = 0$.
+The top panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho > 0$, both recommendation and no recommendation lead to increasingly local consumption compared to the oracle benchmark case.
+Further, the average consumption path between periods is \textit{decreasing} for the no recommendation case whereas it is \textit{increasing} for the oracle case.
+The recommendation regime decreases the degree of local consumption but not as much as the oracle benchmark.
+Due to the correlation of values, the oracle consumption path exploits this and leads to the consumption of more similar products than in the case when $\rho = 0$.
 However, since these spillovers also impact user learning in the no recommendation case, users \textit{over-exploit} this and increasingly consume products similar to good products that they have consumed before.
-This is further illustrated by Figure \ref{fig:local_consumption_across_rho} which shows how the consumption paths between omniscient and no recommendation vary as $\rho$ increases.
+This is further illustrated by Figure \ref{fig:local_consumption_across_rho} which shows how the consumption paths in the oracle and no recommendation regimes vary as $\rho$ increases.
+\par
 
@@
 This is due to the fact that the spillovers not only affect the mean expected belief about quality but also the degree of uncertainty.
-Local consumption therefore leads to users to have less uncertainty about certain areas of the product space and risk aversion may lead them to increasingly consume these products.
+Local consumption therefore leads to users having less uncertainty about certain areas of the product space and risk aversion may lead them to increasingly consume nearby products.
 
-\begin{figure*}
-\includegraphics[scale=0.1]{figures/consumption_dist_N_200T_20_no_rec_rho}
-\includegraphics[scale=0.1]{figures/consumption_dist_N_200T_20_omni_rho}
-\caption{Extent of Local Consumption for No Recommendation (Left) and Omniscient Recommendation (Right)}
+\begin{figure}[t]
+\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20.pdf}\\
+\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_partial.pdf}\\
+\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_omni.pdf}\\
+\caption{Extent of Local Consumption for No Recommendation (Top), Recommendation (Middle) and Oracle (Bottom) for different Correlation values}
 \label{fig:local_consumption_across_rho}
-\end{figure*}
+\end{figure}
 
-\begin{figure}
-\includegraphics[scale=0.1]{figures/consumption_dist_N_200T_20_no_rec_alpha}
-\caption{The Effect of Risk Aversion on Local Consumption}
+\begin{figure}[t]
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20.pdf}\\
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_partial.pdf}\\
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_omni.pdf}\\
+\caption{Effect of Risk Aversion on Local Consumption for No Recommendation (Top), Recommendation (Middle) and Oracle (Bottom)}
 \label{fig:no_rec_risk_aversion}
@@
 
+In sum, in our model filter bubbles are driven by learning-spillovers and risk aversion.
+It is the belief that there is an underlying correlation across different products' values, stronger for similar products, closer in the product space, that leads consumers to focus on local exploration upon consuming high valued items.
+For instance, upon discovering that the user likes movies by a given director, she will be more likely to explore other movies by the same director.
+Risk aversion may lead users into performing local exploration even when they have a low valuation of nearby products just because they know what to expect from the product.
+\par
+
 \subsection{User Welfare and Product Diversity}
+In this section we primarily focus on the impact of recommendation on the welfare on users as well as the overall diversity of the items that they consume.
+We define user's \textit{ex-post} welfare as the average of the realized values, controlling for the effect of $T$
+$$W_i:= \frac{1}{T}\sum_{n \in C_i^T} x_{i,n}$$
 
-In this section we primarily focus on the impact of recommendation on the welfare on users as well as the overall diversity of the items that they consume.
-user's \textit{ex-post} welfare is the average of realized utilities, to control for the effect of $T$, and is defined as follows:
-\begin{align*}
-W_i=\frac{1}{T}\sum_{n \in C_i^T} u_{in}
-\end{align*}
 While in the previous section we looked at the distance between consecutive goods, we also define a diversity measure that is defined over the entire consumed set.
@@
 \cite{ziegler2005improving}) which is the average normalized pairwise distance between the consumed products:
-\begin{align*}
-D_i =\frac{1}{N} \frac{1}{T (T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)
-\end{align*}
+$$D_i:=\frac{1}{N}\frac{1}{T(T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$$
 
@@
 \begin{enumerate}
-\item When $\rho = 0$, product diversity is the same across all three recommendation regimes.
+\item When $\rho = 0$, product diversity is the same across all three recommendation regimes;
 \item When $\rho > 0$, product diversity decreases across all recommendation regimes but decreases the most in the no recommendation regime.
@@
 The over-exploitation of the learning spillovers when $\rho > 0$ leads to product diversity being lowest in the no recommendation regime.
-Figure \ref{fig:diversity_correlation} shows how diversity varies across recommendation regimes and the level of $\rho$ and Figure \ref{fig:risk_aversion_diversity} shows how diversity varies as risk aversion levels change.
+Figure \ref{fig:risk_aversion_diversity} shows how diversity varies as risk aversion levels change and Figure \ref{fig:diversity_correlation} shows how diversity varies across recommendation regimes and the level of $\rho$.
+
+\begin{figure}[ht]
+\includegraphics[width=.9\linewidth]{figures/gamma_diversity_N_200_T_20}
+\captionof{figure}{Risk Aversion and Diversity}\label{fig:risk_aversion_diversity}
+\end{figure}
 
 \begin{finding}\label{finding_welfare_gap}
-The welfare gap between no recommendation and partial recommendation as well as no recommendation and omniscient recommendation is decreasing in $\rho$.
+The welfare gap between the no recommendation and the recommendation regimes and that between the no recommendation and the oracle regimes are decreasing in $\rho$.
 \end{finding}
 However, interestingly, the decrease in diversity is \textit{not} associated with a decline in welfare.
-In fact, welfare stays roughly the same in the omniscient case and slightly increases for the partial recommendation case.
+In fact, welfare stays roughly the same in the oracle and recommendation cases.
 The welfare gap between the no recommendation case and the other two cases \textit{decreases} as the diversity gap increases.
+\par
+\begin{figure}[t]
+\includegraphics[width=.9\linewidth]{figures/rho_diversity_N_200_T_20.pdf}\\
+\includegraphics[width=.9\linewidth]{figures/rho_welfare_N_200_T_20.pdf}\\
+\captionof{figure}{Welfare, Diversity, and Correlation}\label{fig:diversity_correlation}
+\end{figure}
 
 \begin{finding}\label{finding_diversity_welfare_corr}
-Without recommendation, diversity and welfare are:
+In the no recommendation regime, diversity and welfare are:
 \begin{enumerate}
-\item Negatively correlated when users have no risk-aversion
-\item Uncorrelated when users have high levels of risk-aversion
+\item Negatively correlated when users have no risk-aversion;
+\item Uncorrelated when users have high levels of risk-aversion.
 \end{enumerate}
+In the recommendation regime, diversity and welfare are:
+\begin{enumerate}
+\item Uncorrelated when users have no risk-aversion;
+\item Positively correlated when users have high levels of risk-aversion.
+\end{enumerate}
+In the oracle regime, diversity and welfare are always uncorrelated.
 \end{finding}
@@
 When there is no risk-aversion then there is a negative correlation between welfare and diversity.
-This is since, with no risk-aversion, in a given period a user will select the good that they currently believe has the highest expected value.
-High product diversity in this case can arise from a user who consumes a bad item and updates her beliefs about nearby items negatively.
-As a result, in the next period she will pick an item far away in the product space from the item that she consumed previously.
-If the item that she had consumed was ``good", then she is more likely to pick a nearby item.
+This is since, with no risk-aversion, in a given period a user will select the good that she currently believe has the highest expected value.
+High product diversity in this case can arise from a user who consumes an item she disliked and updates her beliefs about nearby items negatively.
+As a result, in the following period she will pick an item far away in the product space from the item that was previously consumed.
+If instead the user valued highly the item that she had consumed, then she is more likely to pick a nearby item.
 The learning spillovers therefore lead to high product diversity being negatively correlated with welfare.
+\par
 
-\begin{figure}
-\includegraphics[scale=0.1]{figures/gamma_diversity_N_200_T_20}
-\caption{Risk Aversion and Diversity}
-\label{fig:risk_aversion_diversity}
+%\begin{figure}
+
+%\end{figure}
+
+This only happens since $\gamma = 0$ leads to users only caring about the expected value of the product.
+However, as we saw in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, increasing $\gamma$ can lead to lower diversity and increasingly local consumption due to the fact that the degree of uncertainty now impacts users' choices.
+This weakens the negative relationship between diversity and welfare since both negative and positive experiences with a good reduce uncertainty about surrounding goods.
+This leads to the inverted-U shape found in Figure \ref{fig:diversity_welfare_ra} when $\gamma$ is relatively large (e.g $\gamma = 5$) though diversity and welfare are virtually uncorrelated in the data.
+In the recommendation and oracle regimes, under risk neutrality ($\gamma=0$), welfare and diversity and uncorrelated, while under risk aversion ($\gamma=5$), it is possible to observe in Figure \ref{fig:diversity_welfare_ra_partial} an actual positive relation between diversity and welfare as recommendations are able to reduce uncertainty and facilitate exploration of the product space.
+\par
+\begin{figure}[t]
+\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_rn.pdf}\\
+\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_ra.pdf}\\
+\captionof{figure}{Diversity vs. Welfare, No-Recommendation, $\gamma = 0$ (Top), Diversity vs. Welfare, $\gamma = 5$ (Bottom)}\label{fig:diversity_welfare_ra}
+\end{figure}
+\begin{figure}[t]
+\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_rn_partial.pdf}\\
+\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_ra_partial.pdf}\\
+\captionof{figure}{Diversity vs. Welfare, Recommendation, $\gamma = 0$ (Top), Diversity vs. Welfare, $\gamma = 5$ (Bottom)}\label{fig:diversity_welfare_ra_partial}
 \end{figure}
 
-This only happens since $\gamma = 0$ leads to users only caring about the expected value of the good.
-However, as we saw in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, increasing $\gamma$ can lead to lower diversity and increasingly local consumption due to the fact that the degree of uncertainty now impacts users' choices.
-This weakens the negative relationship between diversity and welfare since both negative and positive experiences with a good reduce uncertainty about surrounding goods.
-This leads to the inverted-U shape found in Figure \ref{fig:diversity_welfare_ra} when $\gamma$ is relatively large (e.g $\gamma = 5$).
-
-\begin{figure*}
-\includegraphics[scale=0.1]{figures/rho_diversity_N_200_T_20}
-\includegraphics[scale=0.1]{figures/rho_welfare_N_200_T_20}
-\caption{Welfare, Diversity, and Correlation}
-\label{fig:diversity_correlation}
-\end{figure*}
-
-\begin{figure*}
-\includegraphics[scale=0.3]{figures/diversity_welfare_no_risk_aversion}
-\includegraphics[scale=0.3]{figures/diversity_welfare_high_ra}
-\caption{Diversity vs. Welfare, $\gamma = 0$ (Left), Diversity vs. Welfare, $\gamma = 5$ (Right)}
-\label{fig:diversity_welfare_ra}
-\end{figure*}
-
 \subsection{User Homogenization}
-
 In this section we focus on across user comparisons and investigate how the consumed set of items across users varies across different recommendation regimes and parameter values.
@@
 \begin{enumerate}
-\item Highest under partial recommendation and lowest under no recommendation
-\item Increasing in $\beta$, or the weight of the common-value component
-\item Decreasing in $\rho$ for partial recommendation, but weakly increasing in $\rho$ for no recommendation
+\item Highest under recommendation and lowest under no recommendation;
+\item Increasing in $\beta$, or the weight of the common-value component;
+\item Decreasing in $\rho$ for partial recommendation, but weakly increasing in $\rho$ for no recommendation.
 \end{enumerate}
 \end{finding}
+Figure \ref{fig:beta_homo} shows that as the weight of the common-value component, $\beta$, increases users consume increasingly similar sets of items.
+The homogenization effect is strongest under the recommendation regime since the revelation of the common-value component induces users to consume products in similar areas of the product space.
+However, as $\beta$ increases, some amount of homogenization is optimal as can be seen from the oracle case.
+In the no recommendation regime since users do not know the common-value component they engage in local consumption in different areas of the product space which leads to less than optimal homogeneity.
+\par
 
-Figure \ref{fig:beta_homo} shows that as the weight of the common-value component, $\beta$, increases users consume increasingly similar sets of items.
-The homogenization effect is strongest under partial recommendation since the revelation of the common-value component induces users to consume products in similar areas of the product space.
-However, as $\beta$ increases, some amount of homogenization is optimal as can be seen from the omniscient recommendation case.
-In the no recommendation case since users do not know the common-value component they engage in local consumption in different areas of the product space which leads to less than optimal homogeneity.
-
-Figure \ref{fig:cor_homo} shows that the degree of homogeneity in the partial recommendation case however is \textit{decreasing} as $\rho$ increases.
-As was discussed in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, the degree of local consumption increases with $\rho$.
+Figure \ref{fig:cor_homo} shows that the degree of homogeneity in the recommendation case however is \textit{decreasing} as $\rho$ increases.
+As was highlighted in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, the degree of local consumption increases with $\rho$.
 Even though the revelation of the common-value component induces them to search in similar parts of the product space, their idiosyncratic components induce them to consume products in a more localized area of the product space as $\rho$ increases which leads to a decline in homogeneity.
 
-\begin{figure}
-\includegraphics[scale=0.1]{figures/beta_homogeneity_N_200_T_20}
-\caption{Strength of Recommendation and Homogeneity}
-\label{fig:beta_homo}
+\begin{figure}[t]
+\includegraphics[width=.8\linewidth]{figures/beta_homogeneity_N_200_T_20}
+\captionof{figure}{Strength of Recommendation and Homogeneity}\label{fig:beta_homo}
 \end{figure}
+~
 
-\begin{figure}
-\includegraphics[scale=0.1]{figures/rho_homogeneity_N_200_T_20}
-\caption{Correlation and Homogeneity}
-\label{fig:cor_homo}
+\begin{figure}[t]
+\includegraphics[width=.8\linewidth]{figures/rho_homogeneity_N_200_T_20}
+\captionof{figure}{Correlation and Homogeneity}\label{fig:cor_homo}
 \end{figure}
@@
 \section{Recommender System Evaluation}
-
 In this section we discuss how the insights from our model of user decision-making can be used to inform the evaluation and design of recommender systems.
+\par
 
@@
 Further, given this predicted rating, the recommender system should recommend to users the items with the highest predicted rating \cite{adomavicius2005toward}.
+\par
 
@@
 A recommender system attempting to predict user ratings may then predict that this user is very likely to enjoy the sequel, \textit{John Wick: Chapter Two}, as well.
-However, the user themselves may also have made this inference since the two movies are very similar to each other and, after enjoying \textit{John Wick}, the user will update her beliefs about \textit{John Wick: Chapter Two}.
+However, the user herself may also have made this inference since the two movies are very similar to each other and, after enjoying \textit{John Wick}, the user will update her beliefs about \textit{John Wick: Chapter Two}.
 Thus, if the recommender system recommends this movie to the user then this recommendation is not useful since the recommendation gives the user little information that they did not already know.
-As a result, the user may watch \textit{John Wick: Chapter Two} even without recommendation and the value of the recommendation was small.
+The key insight is that the recommendation is not useful since \textit{it ignores the inference the user themselves made and their updated beliefs}.
+The user may watch \textit{John Wick: Chapter Two}, then, even without recommendation and the value of the recommendation was small.
+\par
 
-As a result, several alternative recomendation evaluation metrics, such as serendipity \cite{kotkov2016survey}, coverage \cite{ge2010beyond}, and novelty \cite{vargas2011rank}, have been proposed in order to develop recommender systems that produce recommendations that are more useful for users.
+As a result, several alternative recommendation evaluation metrics, such as serendipity \cite{kotkov2016survey}, coverage \cite{ge2010beyond}, novelty \cite{vargas2011rank}, and many others have been proposed in order to develop recommender systems that produce recommendations that are more useful for users.
 We use our model to motivate an alternative recommendation evaluation metric that provides an alternative perspective on the serendipity metric proposed in the literature.
-Serendipitious recommendations ``have the quality of being both unexpected and useful" \cite{maksai2015predicting}.
+Serendipitous recommendations ``have the quality of being both unexpected and useful" \cite{maksai2015predicting}.
 \cite{iaquinta2010can} defines serendipitous recommendations as follows:
@@
 Our approach to serendipity comes from the intuition that a recommendation is only useful insofar as it leads a user to take a better action than she would without recommendation.
-This approach stems from the idea that the value of recommendation is the marginal gain in expected utility that a user receives from recommendation.\footnote{This notion is similar to the ``value of information" notion utilized in the literature in economics on the pricing of information goods \cite{bergemann2018design}.} This intuitively should come from providing information about products that, without additional information, users assign a low likelihood to being the best product currently available and so would not consume them without recommendation.
+This approach stems from the idea that the value of recommendation is the marginal gain that a user receives from recommendation.\footnote{This notion is similar to the ``value of information" notion utilized in the literature in economics on the pricing of information goods \cite{bergemann2018design}.} This intuitively should come from providing information about products that, without additional information, users assign a low likelihood to being the best products currently available for them to consume and so would not consume them without these being recommended.
+\par
 
@@
 This depends only on the resulting item-set that the users get recommended and not necessarily user beliefs or the change in user action.
-\cite{adamopoulos2015unexpectedness} use an expected utility approach but use this to measure unexpectedness in terms of deviations from the ``expected consideration set" of a user and do not explicitly incorporate user beliefs.
+\cite{adamopoulos2014unexpectedness} use an expected utility approach but use this to measure unexpectedness in terms of deviations from the ``expected consideration set" of a user and do not explicitly incorporate user beliefs.
+\par
 
 Our model allows us to give precise definitions for what it means for a recommendation to be \textit{unexpected} and \textit{useful}.
-Unexpected recommendations are given by recommended items that, given their beliefs, users have low expected utility from consuming, irregardless of what the true realized utility is.
+Unexpected recommendations are given by recommended items that, given their beliefs, users have low expected utility from consuming, irregardless of what the true realized value is.
 Recall from Findings \ref{finding_local_consumption} and \ref{finding_diversity_welfare_corr} that, particularly if users are risk-averse, they will consume content in increasingly narrow portions of the product space and this may be detrimental to their realized welfare.
@@
 This is especially the case if this reduction of uncertainty leads users to explore under-explored portions of the product space where true realized utility is high, but expected utility, given beliefs, is low before recommendation.
+\par
 
-Useful recommendations are recommendations that give users information that would change the item they consume relative to no recommendation and lead to an increase in expected utility.
+Useful recommendations are those that give users information that would change the item they consume relative to no recommendation and lead to an increase in expected utility given the updated beliefs when comparing the planned choice before and after the recommendation.
 In the previous example, recommending \textit{John Wick: Chapter Two} is not useful precisely because the expected utility before and after recommendation is the same since it does not change behavior.
-The most useful recommendations, or those that have the highest utility gain, are recommendations that induce users to consume unexpected goods.
+The most useful recommendations are those that (1) are informative and (2) induce users to consume unexpected goods.
+\par
 
@@
 \cite{jiang2014choice, saavedra2016choice} argue that choice-based approaches alone can help in designing better RS, though they argue for these approaches because they do a better job at providing more accurate recommendations.
-However, we argue that choice-based approaches are useful not as recommendations but as a baseline for what recommendations would be useful.\footnote{Furthermore, user beliefs and user choices contain information that may not have been observed by the recommender but is observed by the user.
+Instead, we stress that choice-based approaches are useful not as recommendations but as a baseline for what recommendations would be useful.\footnote{Additionally, user beliefs and user choices contain information that may not have been observed by the recommender but is observed by the user.
 Ratings, reviews, friends, there are many sources of information affecting a person's beliefs and choices that are unobservable by recommender systems.
@@
 However, this information can be inferred by the platform by collecting both beliefs and consumption choices.}
+\par
 
@@
 For instance, understanding the degree of correlation of utilities and the level of risk-aversion become crucial in understanding how users will make consumption choices without recommendation and what kind of recommendation would be useful for users.
-
-\section{Discussion}
+\par
 
 \section{Conclusion}
+We have studied a model of user decision-making in the context of markets where recommender systems are traditionally deployed in order to understand the consequences of recommender systems on consumption patterns of users.
+The key components of the model are that users make decisions under uncertainty and that consuming a good leads to informational spillovers about similar goods.We have shown that this alone can generate consumption behavior that is consistent with filter bubble effects that have traditionally been associated with personalized recommender systems.
+This effect is weakened once we allow for recommendation, but not as much as if users have full information.
+We have further explored the relationship between consumption diversity and user welfare and found that they are negatively correlated, implying that striving for consumption diversity is not necessarily welfare-improving for users.
+Finally, we have shown that recommendation can lead to increases in user homogenization both relative to no recommendation and full information since it concentrates users' consumption in portions of the product space with the highest utility according to the limits of what the recommender can predict about the utility of the good.
+\par
 
-We have argued that incorporating user choice under uncertainty should be a first-order component of RS design.
+We have used the insights from our model to guide the evaluation and design of recommender systems.
+The key insight is that a critical component of designing good recommendations is understanding what user behavior would be without recommendation and then understanding how recommendation would change their consumption.
+Operationalizing this requires collecting information on user beliefs about the goods in the product space, which is not data that is traditionally collected for the development of recommender systems.
+\par
+
+We leave for future work exploring to what extent user behavior changes if we drop the myopia assumption, so that users themselves are engaging in exploration.
+This likely interacts with recommendation in interesting ways as the recommendations may provide information that can complement or crowd-out user-led exploration.
+Our model also reveals a clear path-dependence in choice since consumption decisions today shape the information users have tomorrow about other products which affects their consumption choices tomorrow, etc.
+Even under the myopia assumption, understanding which recommendation system is optimal is an interesting and challenging problem.
+\par
+
+%We have argued that incorporating user choice under uncertainty should be a first-order component of RS design.
+By collecting appropriate data about user choices and user beliefs, RS can be built to better understand what choices users are likely to make and thus what information would be useful to give them as opposed to simply predicting what items a user will like.
+This approach can not only aid in designing more useful RS, but can also be utilized to better understand and prevent recently documented adverse effects of RS such as filter bubbles and homogenization.
+
+%\newpage
+
+%\subsection{User Homogenization}
+
+%
+%\begin{table}[bt]
+%\centering
+%\begin{tabular}{rlll}
+%  \hline
+%Rec Policy & Welfare & Diversity & Homogeneity \\ 
+%  \hline
+%No Rec & 0.42 $\pm$ 0.002 & 0.19 $\pm$ 0.0003 & 0.001 $\pm$ 0 \\ 
+%Partial Rec & 1.25 $\pm$ 0.005 & 0.22 $\pm$ 0.0002 & 0.174 $\pm$ 0.004 \\ 
+%Omniscient Rec & 2.24 $\pm$ 0.006 & 0.24 $\pm$ 0.0001 & 0.035 $\pm$ 0.002 \\ 
+%   \hline
+%\end{tabular}
+%\caption{Mean Welfare, Diversity, and Homogeneity for $N = 1000, T = 25, I = 50, S = 50$. Reported intervals are 95\% confidence intervals.}
+%\label{table:agg_results}
+%\end{table}
+%\begin{figure*}[!htbp]
+%    \centering
+%    \begin{minipage}[b]{0.5\textwidth}
+%        \centering
+%        \includegraphics[width=0.9\linewidth, height=1.5in]{local_moves_25}
+%        \caption{Extent of ``Local"  Consumption}
+%        \label{fig:local_consumption}
+%    \end{minipage}%
+%    ~ 
+%    \begin{minipage}[b]{0.5\textwidth}
+%        \centering
+%        \includegraphics[width=0.9\linewidth, height=1.5in]{welfare_homo_combo}
+%        \caption{Welfare and Homogeneity}
+%        \label{fig:beta_vary}
+%    \end{minipage}
+%    ~ \\
+%     \begin{minipage}{0.48\textwidth}
+%     \centering
+%     \includegraphics[width=0.9\linewidth, height=1.5in]{rec_obedience_25}
+%     \caption{Recommendation Effectiveness}\label{fig:rec_obey}
+%   \end{minipage}\hfill
+%   ~
+%   \begin{minipage}{0.48\textwidth}
+%     \centering
+%     \includegraphics[width=0.9\linewidth, height=1.5in]{Diversity_vs_Welfare_No_Recommendation}
+%     \caption{Diversity vs Welfare under No Recommendation}\label{fig:diversity_welfare_no_rec}
+%   \end{minipage}
+%\end{figure*}
+
+%We are going to analyze three outcomes: how diverse consumer choices are, how similar are the choices that different individuals make and the resulting welfare.
+The first will be measured by $D_i =\frac{1}{N} \frac{1}{T (T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$, the average normalized pairwise distance between the consumed products, a natural measure of diversity in this environment that is utilized in the literature \cite{ziegler2005improving}.
+The second will be measured by a consumer homogeneity index $H:=\frac{1}{|I|(|I|-1)}\sum_{i,j \in I: i \ne j}d_J(C_i^T,C_j^T) $, where $d_J$ denotes the Jaccard index and $H \in [0,1]$ which is similar to the approach to measuring homogeneity utilized in \cite{chaney2018algorithmic}.
+Note that a higher $H$ indicates more homogeneity.
+Consumer $i$'s welfare will just be the average of realized utilities, to control for the effect of $T$, $W_i=\frac{1}{T}\sum_{n \in C_i^T} u_{in}$.
+%\par
+%\par
+%\noindent \textbf{Results}.
+Table \ref{table:agg_results} shows aggregate results for welfare, diversity, and homogeneity.
+The omniscient recommendation case leads to the optimal consumption path for a consumer since she consumes the $T$ items with the highest utility.
+Compared to this case, both partial recommendation and no recommendation not only lead to lower welfare but also have lower diversity.
+In fact, the no recommendation case leads to the lowest levels of consumption diversity since users do not explore the space of products sufficiently and do excessive amounts of local search as can be seen in Figure \ref{fig:local_consumption}.
+Figure \ref{fig:local_consumption} shows how ``local" consumption choices are by looking at what fraction of consecutive consumption choices are close to each other with the definition of close being less than distance of 10.
+The partial recommendation case leads to higher consumption diversity and higher welfare, but also significantly higher homogeneity.
+%
+%In our model, filter bubble effects occur simply because of the correlation structure between the utility and beliefs of products.
+It highlights the fact that the correlation structure induces users to consume increasingly narrow content when they get no guidance from recommendation whereas the ``optimal" consumption path given by the omniscient policy does not exhibit this.
+The results of our model are consistent with the empirical results found in \cite{nguyen2014exploring} who show that filter bubble effects can arise even amongst users who do not utilize recommendations.
+%\par
+%Furthermore, \cite{nguyen2014exploring} also find that recommendation can lead to an increase in consumption diversity.
+As Table \ref{table:agg_results} shows, our model is consistent with this effect but it depends on the nature of recommendation.
+If recommendations only provides information on the common-value component (partially informed recommender) then welfare and consumption diversity increase when contrasted to the no recommendation case, but homogenization increases dramatically.
+Knowing the common-value component leads users to have more information on where to explore in the product space.
+However, the fact that this information is about the common-value component and the same across all individuals leads them to explore the same sections of the product space and leads to substantially higher homogeneity among users.
+If recommendation takes into account user beliefs on the idiosyncratic component of utility (omniscient recommender), then welfare and consumption diversity can further increase and homogeneity will decrease.
+%\par
+%Low consumption diversity is not necessarily bad by itself.
+Figure \ref{fig:diversity_welfare_no_rec} shows that, in the no recommendation case, welfare and consumption diversity are negatively correlated.
+This is primarily because content diversity in the no recommendation case can arise from the fact that the user consumes a bad product and is forced to jump around the product space looking for ``good" products.
+However, if a user finds a high utility product right away, then staying in this neighborhood may yield higher utility due to the correlation of utilities.
+The problem is that, even when the user finds a ``good" product, this may only be ``locally" good and the user will then excessively consume items around it due to having more information about these than more ``different'' ones, i.e.
+farther away.
+%\par
+%Finally, the value of recommendation in the partial recommendation case decreases over time as users learn more about the idiosyncratic component of their own preferences (i.e.
+as $t$ increases).
+Figure \ref{fig:rec_obey} shows that recommendation compliance decreases as $t$ increases.
+Additionally, Figure \ref{fig:beta_vary} shows how homogeneity and welfare change as $\beta$ increases.
+Most interestingly, when $\beta$ is high so that the common-value component is small, recommendation can be harmful to users.
+The information from recommendation still leads to large user homogeneity even though the optimum would be for there to be almost none and leads to lower welfare than under no recommendation.
+%
+%\section{The Value of the Ex-Ante View}
+%
+%In this section we illustrate the differences between the ex-post and ex-ante view and show how it is useful for thinking about the design of RS.
+Our first point is trivial to state - since users do not know the true consumption values of the items, without any further information they may make \textit{ex-post} sub-optimal consumption decisions, but \textit{ex-ante} optimal given the information available to them at the moment of choice.
+Consider the following stylized example.
+The choice set of the individual is $\Omega= \{x_1, x_2, x_3 \}$ and the ex-post utility values are given by
+%$u(x_1) = 1$, $u(x_2) = 2$, $u(x_3) = 3$. 
+%\par
+%However, when users are making decisions they have unobservable \textit{beliefs} about the ex-post utility value of the items and make decisions given these beliefs.
+As users obtain more information, their beliefs converge to the ex-post utility values but, generally, users do not have enough information and so have beliefs over the space of possible ex-post utility values.
+In this example we suppose these beliefs are simple so that users view each good as simple lotteries: for $n=1,2,3$, $u(x_n) = n + \varepsilon_n$, $\varepsilon_1 \sim \mathcal J (2, 1)$, $\varepsilon_2 \sim \mathcal J (0, 1)$, $\varepsilon_3 \sim \mathcal J (-5, 2)$.
+%\par
+%After uncertainty is unresolved, the user would consume $x_3$, but an expected utility maximizing consumer (without further information) would choose $x_1$.\footnote{We ignore an important component of decision-making under uncertainty, risk attitudes, by focusing on risk-neutral consumers.
+A \textit{risk-averse} user may choose to consume a good with lower expected \textit{ex-post} payoff but lower ex-ante variance compared to another good.
+While risk-aversion is important for decision-making under uncertainty, especially in the context of RS, it is not our main focus here.}
+%\par
+%This observation, while seemingly trivial, has important implications.
+The fact that the user would choose $C$ over $A$ reveals that $C$ prefers $A$ given her current beliefs\footnote{This is a statement of \textit{revealed preference} which has been studied heavily in economic decision theory (see e.g.
+\cite[]{mas1995microeconomic}).} and allows the system designer to get some information about a user's beliefs.
+Why are user beliefs important objects for designers of RS to understand and utilize?
+%\par
+%Beliefs are useful for understanding how users interact and get value from RS.
+In particular, users use RS to get \textit{information} about the set of items and update their beliefs about the value of the items.
+For instance, in the example, a RS could recommend to a user $x_3$ over $x_2$ and $x_1$ and that user will use this to update her beliefs about $x_3$.
+It is an empirical question that we leave for future work precisely how and to what extent users update their beliefs from recommendation and what factors of RS deisgn influence this.\footnote{There has been some work in this direction in understanding when recommendations are persuasive \cite{cremonesi2012investigating, gretzel2006persuasion} as well as the effect of their delivery on effectiveness \cite{murphy2014recommendation}.}
+%\par
+%Better understanding user beliefs is important for improving the performance of RS that optimize for metrics that move away from accuracy such as serendipity \cite{kotkov2016survey}.
+Serendipity-based RS attempt to provide recommendations that ``have the quality of being both unexpected and useful" \cite{maksai2015predicting} but it is hard to know what items are both unexpected and useful \cite{kotkov2016survey}.
+Understanding \textit{why} a user has not consumed a good depends on the beliefs that the user has and this is important for designing serendipitious recommendations.
+For instance, in the example, the user does not consume $x_3$ because she expects the good to turn out worse than the other two.
+However, if given more information, she would prefer to consume it over $x_1$ and $x_2$.
+Moreover, when the user can use her known valuation of an item to make inferences on how good another item is, consumption is path-dependent, as beliefs guide consumption and consumption affects beliefs about other goods.
+For instance, learning the value of $x_1$ says something about how good $x_2$ is.
+%When considering risk-averse individuals, when the user has alternatives with a known value,The consumer may have also consumed Another reason why the user may not .
+The user may have an alternative with a known value -- say going for a walk around the neighborhood -- which she could always prefer to watching a given movie that has uncertain value but that the consumer thinks is always lower than the value of the walk.\footnote{An additional reason this could be useful is if one considers a risk-averse individual, then it may be that a user has not consumed an item because she has a high perceived variance for the item's quality.}
+%As a result, understanding the beliefs of users helps us understand why users have not consumed certain items and, in particular, which items a user may find useful given more information about it.
+This poses an important problem: designing systems that elicit not only ex-post valuations but also ex-ante beliefs about valuations, as only by knowing these can RS be effective in steering users' choices.
+In particular, choice has to be perceived as guided by ex-ante utility, whereas ratings by the ex-post realized utility.
+%\par
+%Understanding the choices that individuals will make in addition to what they will end up liking are complementary problems but require different viewpoints.
+\cite{jiang2014choice, saavedra2016choice} argue that choice-based approaches alone can help in designing better RS, though they argue for these approaches because they do a better job at providing more accurate recommendations.
+We argue that the two problems should be considered separately -- though they may interact in interesting and useful ways.\footnote{In particular, some results in \cite{jiang2014choice, saavedra2016choice} may be driven by the fact that user choice embeds ratings information that may not have been observed by the recommender but is observed by the user.
+Ratings, reviews, friends, there are many sources of information affecting a person's beliefs and choices that are unobservable by RS.
+Other consumption choices, e.g.
+movies seen in the cinema, change beliefs about for instance how good a director is, which affect beliefs about other movies and guide choices on a streaming platform that is unable to observe these data.}
+%\par
+%RS are designed to give users \textit{information} that is useful to help them make decisions.
+On the one hand, predicting a user's ex-post preferences is useful to know what items the user would like to consume.
+On the other hand, predicting what items a user will actually consume (without recommendation) can help reason about what information may actually be useful to give her.
+Furthermore, in designing RS to avoid filter bubble and homogenization effects, having accurate choice predictions allows mitigating such adverse effects to become a first-order component of design.
+In particular, by predicting both choice and ratings, RS can provide information to users that leads them to useful items but prevents them from falling into a filter bubble.
+%\par
+%Solving this choice prediction problem requires data on the choices that individuals make and the sets that they choose them from in addition to the traditional ratings or behavioral data that is collected.
+In the example given previously the fact that the consumer chooses $x_1$ from the set $\{x_1, x_2, x_3 \}$ would be the data that is useful for the choice prediction problem, as would the consumer's ex-ante beliefs about her valuation of the goods.
+%
+%
+%\section{Conclusion}
+%
+%We have argued that incorporating user choice under uncertainty should be a first-order component of RS design.
 By collecting appropriate data about user choices and user beliefs, RS can be built to better understand what choices users are likely to make and thus what information would be useful to give them as opposed to simply predicting what items a user will like.

### 2019-10-03 paper/filter_bubbles.tex
updates to paper

@@
-\documentclass[sigconf, anonymous, review]{acmart}
+\documentclass[sigconf]{acmart}
 \settopmatter{printacmref=false}
@@
 
+\settopmatter{printacmref=false} % Removes citation information below abstract
+\makeatletter
+\def\-copyrightspace{\relax}
+\makeatother
+
 %
@@
 We study a model of user decision-making in the context of recommender systems.
-We explore the extent to which issues traditionally associated with recommender systems -- namely filter bubbles, user homogenization and consumption diversity -- can be attributed to recommendation systems.
+We explore the extent to which two issues traditionally associated with recommender systems -- \textit{filter bubbles} and \textit{user homogenization} -- can be attributed to recommendation systems.
 The key component of our model is that users face a decision problem under uncertainty and that there are informational spillovers where a user consuming a product and learning how she values it reduces uncertainty about similar products.
-We find that this generates behavior consistent with narrowing filter bubble effects and that recommendation can help alleviate these effects, as has been found in the empirical work of Nguyen et al.
+We find that this generates behavior consistent with filter bubble effects -- where users consume increasingly similar items over time -- and that recommendation can help alleviate these effects, as has been found in the empirical work of Nguyen et al.
 (2014).
@@
 \par
-
 We discuss the evaluation of recommender systems in light of these findings.
-Traditional design approaches to recommender systems only optimize for predicting user preferences after uncertainty has been resolved.
+Traditional design approaches to recommender systems only optimize for predicting user preferences after uncertainty has been resolved and ignore user beliefs and the inferences that users themselves make after consuming items.
 We propose an alternative interpretation of the serendipity measure where good recommendations are those that provide useful information, that is, information without which choice would be different and lead to worse outcomes on average relative to the case where no recommendation is provided.
@@
 %
+\author{Guy Aridor}
+\affiliation{
+  \institution{Columbia University \\ Department of Economics}}
 
+\author{Duarte Gon\c{c}alves}
+\affiliation{%
+  \institution{Columbia University, BRU-IL \\ Department of Economics}}
+
+  \author{Shan Sikdar}
+  \affiliation{Independent Researcher}
 %
@@
 
-Recommender Systems (RS) have become critical for assisting consumers in navigating the large choice sets that they face in many online markets.
-For instance, consumers have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
-Consumers in many cases are not aware of most items, let alone have full information about their preferences over them and, to make matters worse, the goods in these markets are usually experience goods whose true utility can only be learned after consumption.
-Furthermore, consumers interact with these systems routinely and watch more than one movie on Netflix, buy more than one product on Amazon, and listen to more than one video on YouTube.
+Recommender Systems (RS) have become critical for assisting users in navigating the large choice sets that they face in many online markets.
+For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
+Users in many cases are not aware of most items, let alone have full information about their preferences over them and, to make matters worse, the goods in these markets are usually experience goods whose true utility can only be learned after consumption.
+Furthermore, users interact with these systems routinely and watch more than one movie on Netflix, buy more than one product on Amazon, and listen to more than one video on YouTube.
 \par
@@
 Recommender systems have been influential in shaping consumer choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{\url{https://www.mckinsey.com/industries/retail/our-insights/how-retailers-can-keep-up-with-consumers}} While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems.
-There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals,\footnote{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350/} that personalized recommender systems lead consumers into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead consumers to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
+There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals,\footnote{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350/} that personalized recommender systems lead users into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
 \par
@@
 %drawing from recent work in psychology and decision theory in economics, 
-we first propose a model of consumer decision-making in the context of large choice sets faced by consumers in online markets and show that the model is able to rationalize features of user behavior in such contexts by means of simulations.
-We then ask how a stylized model of recommendation can shape the behavior of long-lived consumers 
+we first propose a model of consumer decision-making in the context of large choice sets faced by users in online markets and show that the model is able to rationalize features of user behavior in such contexts by means of simulations.
+We then ask how a stylized model of recommendation can shape the behavior of long-lived users
 and finally ask how the insights from our model can be used to improve recommender system design and to better understand the consequences of these systems.
@@
 We compare this benchmark to the case where recommendations are provided.
-We consider recommendations as giving information to consumers in order to reduce their uncertainty and we find that this reduces the narrowing effect, which is also consistent with the empirical results from \cite{nguyen2014exploring}.
+We consider recommendations as giving information to users in order to reduce their uncertainty and we find that this reduces the narrowing effect, which is also consistent with the empirical results from \cite{nguyen2014exploring}.
 \par
@@
 Moreover, we argue that successful recommendation systems have to take into account the degree of uncertainty users have over different portions of the product space.
-Being able to predict valuations after consumption is not sufficient as uncertainty over the idiosyncratic component of one's own preferences may steer the user away from the recommended items in face of risk-aversion.
+Being able to predict valuations after consumption is not sufficient as uncertainty over the idiosyncratic component of one's own preferences may steer the user away from the recommended items in the face of risk-aversion.
 \par
@@
 A number of studies, done in a wide range of disciplines, have since empirically studied whether or not this phenomenon exists in various contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}.
-The most relevant to our proposed study is \cite{nguyen2014exploring} who study whether this effect exists with recommender systems in the context of movie consumption.
+The most relevant to our study is \cite{nguyen2014exploring} who study whether this effect exists with recommender systems in the context of movie consumption.
 They find that even users whose consumption choices are not guided by recommendations fall into behavior consistent with ``filter bubbles'' and that recommender systems can actually increase the diversity of the content that users consume.
@@
 \cite{bergemann2019information}) and builds off the large literature stemming from \cite{mcnee2006being} that makes the case that accurate predictions of true realized valuations may not lead to the best recommendations.
-This is related to the notion of serendipity in recommender systems that has been studied extensively (e.g.
+Our evaluation measure is most closely related to the notion of serendipity in recommender systems that has been studied extensively (e.g.
 see \cite{kotkov2016survey} for a survey).
@@
 Each consumer can value the same product in differently.
-However, we assume that consumers have the same utility over money, given by a utility function $u: \mathbb R \to \mathbb R$, strictly increasing and continuous.
+However, we assume that users have the same utility over money, given by a utility function $u: \mathbb R \to \mathbb R$, strictly increasing and continuous.
 So, ex-post, the value of product $n$ for consumer $i$ is given by $u(x_{i,n})$ However, before consuming the product, the consumer does not know exactly how she will value it.
-In particular, even consumers that will end up having the same ex-post valuation of product $n$ may differ in their ex-ante valuation because they hold different beliefs about it.
+In particular, even users that will end up having the same ex-post valuation of product $n$ may differ in their ex-ante valuation because they hold different beliefs about it.
 We denote by $p_{i}$ the beliefs consumer $i$ has about how she will value each of the products in the product space.
@@
 This captures the idea that users are faced with an immense choice set but that ultimately they end up experiencing (and learning) about just a small fraction of it.
-Since this is a sequential decision-making problem under uncertainty, in principle consumers face an exploration-exploitation trade-off.
-However, for tractability, we impose that consumers are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
+Since this is a sequential decision-making problem under uncertainty, in principle users face an exploration-exploitation trade-off.
+However, for tractability, we impose that users are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
 \par
-\noindent \textbf{User Beliefs} We assume that consumers do not know the realized utilities of the goods before consuming an item.
+\noindent \textbf{User Beliefs} We assume that users do not know the realized utilities of the goods before consuming an item.
 Formally, consumer $i$ starts with some beliefs about $X_i$, namely that the idiosyncratic and common-value parts of the utilities are independent -- $V_i \perp \!\!\!
@@
 Our model of recommendation is stylized in order to provide qualitative insights into how recommendation shapes behavior, instead of focusing on the details of how recommender systems are implemented in practice.
-We model recommendation as giving consumers information about the valuation of the products.
+We model recommendation as giving users information about the valuation of the products.
 \par
@@
 but we assume the recommender provides full information about $V$.
-For instance, the recommender could display the whole distribution of utilities reported by other users or even its average, which is a good proxy for the common value component\footnote{It is possible for the recommender to only recommend the best item given such information, that is, the item with highest common value component.
+\footnote{For instance, the recommender could display the whole distribution of utilities reported by other users or even its average, which is a good proxy for the common value component.
+It is possible for the recommender to only recommend the best item given such information, that is, the item with highest common value component.
 However, in that case, recommending only a single item generates costly Bayesian updating to the user and provides little guidance as to what she should indeed pick as the common value might be of little importance when compared to the idiosyncratic component.
-Therefore, we assume that the recommender reports the whole $V$, which results in higher expected welfare in choices and is not costly to implement.}.
+Therefore, we assume that the recommender reports the whole $V$, which results in higher expected welfare in choices and is not costly to implement.}
 The results would be equivalent, but this would shift the burden of Bayesian updating to users.
@@
 Standard decision theory axiomatized the conditions under which behavior is described by (subjective) expected utility maximizers \cite{savage1954} who update beliefs according to Bayesian inference \cite{ghirardato2002revisiting}, to which we add the assumption that their prior is correct.
-The second justification is that, even if users are in an environment where they could observe utilities based on product reviews or introspection, it is prohibitively costly for them to do so in large choice sets\footnote{Recent empirical work \cite{bronnenberg2014zooming,hodgson2019horse} has shown that, for product search, users engage in a ``zooming in'' search strategy in terms of the information they acquire about products before purchasing and only end up acquiring information about a small set of products in a large product space}.
-Even if they acquired all the information from product reviews, this would not fully resolve uncertainty as to the extent that individuals have idiosyncratic tastes.
+The second justification is that, even if users are in an environment where they could observe utilities based on product reviews or introspection, it is prohibitively costly for them to do so in large choice sets.\footnote{Recent empirical work \cite{bronnenberg2014zooming,hodgson2019horse} has shown that, for product search, users engage in a ``zooming in'' search strategy in terms of the information they acquire about products before purchasing and only end up acquiring information about a small set of products in a large product space}  Even if they acquired all the information from product reviews, this would not fully resolve uncertainty as to the extent that individuals have idiosyncratic tastes.
 Giving that users are facing choices problems uncertainty, this motivates modeling recommendation as providing users \textit{information} in order to reduce their uncertainty.
@@
 In sum, in our model filter bubbles are driven by learning-spillovers and risk aversion.
-It is the belief that there is an underlying correlation across different products' values, stronger for similar products, closer in the product space, that leads consumers to focus on local exploration upon consuming high valued items.
+It is the belief that there is an underlying correlation across different products' values, stronger for similar products, closer in the product space, that leads users to focus on local consumption upon consuming high valued items.
 For instance, upon discovering that the user likes movies by a given director, she will be more likely to explore other movies by the same director.
-Risk aversion may lead users into performing local exploration even when they have a low valuation of nearby products just because they know what to expect from the product.
+Risk aversion may lead users into performing local consumption even when they have a low valuation of nearby products just because they know what to expect from the product.
 \par
@@
 We use our model to motivate an alternative recommendation evaluation metric that provides an alternative perspective on the serendipity metric proposed in the literature.
-Serendipitous recommendations ``have the quality of being both unexpected and useful" \cite{maksai2015predicting}.
-\cite{iaquinta2010can} defines serendipitous recommendations as follows:
-\begin{quote}
-\textit{A serendipitous recommendation helps the
-user to find a surprisingly interesting item that
-she might not have otherwise discovered (or it
-would have been really hard to discover). [...]
-Serendipity cannot happen if the user already
-knows what is recommended to her, because a
-serendipitous happening is by definition something new. Thus the lower is the probability that user knows an item, the higher is the
-probability that a specific item could result
-in a serendipitous recommendation}.
-\end{quote}
+Serendipitous recommendations are said to ``have the quality of being both unexpected and useful" \cite{maksai2015predicting} and we use this as a guide towards our definition.
 
 Our approach to serendipity comes from the intuition that a recommendation is only useful insofar as it leads a user to take a better action than she would without recommendation.
-This approach stems from the idea that the value of recommendation is the marginal gain that a user receives from recommendation.\footnote{This notion is similar to the ``value of information" notion utilized in the literature in economics on the pricing of information goods \cite{bergemann2018design}.} This intuitively should come from providing information about products that, without additional information, users assign a low likelihood to being the best products currently available for them to consume and so would not consume them without these being recommended.
+This approach stems from the idea that the value of recommendation is the marginal utility gain that a user receives from recommendation.\footnote{This notion is similar to the ``value of information" notion utilized in the literature in economics on the pricing of information goods \cite{bergemann2018design}.} This intuitively should come from providing information about products that, without additional information, users assign a low likelihood to being the best products currently available for them to consume and so would not consume them without these being recommended.
 \par
@@
 In the previous example, recommending \textit{John Wick: Chapter Two} is not useful precisely because the expected utility before and after recommendation is the same since it does not change behavior.
-The most useful recommendations are those that (1) are informative and (2) induce users to consume unexpected goods.
+The most useful recommendations are those that (1) are informative given user beliefs and (2) induce users to consume unexpected goods.
 \par
@@
 We have studied a model of user decision-making in the context of markets where recommender systems are traditionally deployed in order to understand the consequences of recommender systems on consumption patterns of users.
-The key components of the model are that users make decisions under uncertainty and that consuming a good leads to informational spillovers about similar goods.We have shown that this alone can generate consumption behavior that is consistent with filter bubble effects that have traditionally been associated with personalized recommender systems.
+The key components of the model are that users make decisions under uncertainty and that consuming a good leads to informational spillovers about similar goods.
+We have shown that this alone can generate consumption behavior that is consistent with filter bubble effects that have traditionally been associated with personalized recommender systems.
 This effect is weakened once we allow for recommendation, but not as much as if users have full information.

### 2019-10-31 paper/filter_bubbles.tex
in => $
commit finalized version of paper

@@
-\documentclass[sigconf]{acmart}
+\documentclass[sigconf, review, anonymous]{acmart}
 \settopmatter{printacmref=false}
@@
 \begin{document}
+\setlength{\abovedisplayskip}{0pt}
+\setlength{\belowdisplayskip}{0pt}
 % end of the preamble, start of the body of the document source.
@@
 \title{A User-Centric Approach to the Design and Consequences of Recommender Systems}
+
 %
@@
 For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
-Users in many cases are not aware of most items, let alone have full information about their preferences over them and, to make matters worse, the goods in these markets are usually experience goods whose true utility can only be learned after consumption.
+Users in many cases are not aware of most items, let alone have full information about their preferences over them and, to make matters worse, the goods in these markets are usually experience goods whose true value can only be learned after consumption.
 Furthermore, users interact with these systems routinely and watch more than one movie on Netflix, buy more than one product on Amazon, and listen to more than one video on YouTube.
@@
 
-Recommender systems have been influential in shaping consumer choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{\url{https://www.mckinsey.com/industries/retail/our-insights/how-retailers-can-keep-up-with-consumers}} While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems.
-There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals,\footnote{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350/} that personalized recommender systems lead users into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
+Recommender systems have been influential in shaping consumer choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{MacKenzie et al.
+(2013, Oct.),  How retailers can keep up with consumers.
+\url{https://www.mckinsey.com/industries/retail/our-insights/how-retailers-can-keep-up-with-consumers}.
+Retrieved on October 3, 2019.} While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems.
+There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals,\footnote{Friedersdorf (2018, Mar.
+12), YouTube Extremism and the Long Tail, \textit{The Atlantic}.
+\url{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350}.
+Retrieved on October 3, 2019.} that personalized recommender systems lead users into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
 \par
@@
 Our model rationalizes the empirical results in \cite{nguyen2014exploring} who show that in the context of movie consumption, user behavior is consistent with filter bubble effects \textit{even without recommendation}.
-The driving force of this in our model is that utilities of similar goods are correlated, which implies that when a good is consumed and the user learns its value, it provides information about similar goods.
+The driving force of this in our model is that preferences for similar goods are correlated, which implies that when a good is consumed and the user learns its value, it provides information about similar goods.
 Crucially, this not only impacts the underlying belief about the average utility of the good, but also the level of uncertainty.
@@
 We compare this benchmark to the case where recommendations are provided. 
-We consider recommendations as giving information to users in order to reduce their uncertainty and we find that this reduces the narrowing effect, which is also consistent with the empirical results from \cite{nguyen2014exploring}.
+We consider recommendations as giving information to users in order to reduce their uncertainty.
+Recommendations could potentially make matters worse: providing information about the value of the goods when this value is correlated could intensify local search and decrease consumption diversity.
+However, and in line with the empirical results from \cite{nguyen2014exploring}, we find that recommendations reduce the narrowing effect.
 \par
 
-We then explore how diverse the overall set of items that users consume is and find that, without recommendation, higher diversity of items does not ensure higher welfare.
-This is due to the fact that once users find good items, they consume items nearby in the product space due to the correlation of utilities, leading to higher overall value from consumption.
+We then explore how diverse the overall set of items that users consume is.
+It could be expected that higher diversity being associated with more intense search is associated with higher welfare.
+Much to the opposite, we find that, without recommendation, higher diversity of items does not ensure higher welfare.
+This is due to the fact that once users find good items, they consume items nearby in the product space due to the correlation of values, leading to higher overall value from consumption.
 When they consume products they value little, they move to another part of the product space.
 A sequence of bad realizations will then lead to very intense search over very different parts of the product space and high consumption diversity, whereas discovering a sweet spot leads to highly rewarding local search and very low diversity.
-Thus we find that consumption set diversity is not by itself a useful measure for understanding the impact of recommender systems on users.
+Thus, contrary to what a natural intuition could suggest, we find that consumption set diversity is not by itself a useful measure for understanding the impact of recommender systems on users.
 \par
@@
 \cite{fleder2009blockbuster} also consider the effects of recommendation on multi-period user behavior, but predominantly focus on sales diversity and employ a different model of user behavior than we do here.
-\cite{hodgson2019horse} consider a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of utilities in the environment, but consider it in the context of search for a single item.
+\cite{hodgson2019horse} consider a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of their preferences in the environment, but consider it in the context of search for a single item.
 Our discussion of recommender system evaluation contributes with insights closely tied to the value of information that is used in the economics literature (see e.g.
@@
 \par
-\noindent \textbf{Users} We consider a set of users $I$ where each individual $i \in I$ faces the same finite set of $N$ items $\mathcal J = \left\{0,1,...,N-1\right\}$.
+\noindent \textbf{Users.} We consider a set of users $I$ where each individual $i \in I$ faces the same finite set of $N$ items $\mathcal J = \left\{0,1,...,N-1\right\}$.
 For simplicity, we assume that individuals only derive pleasure from item $n \in \mathcal{J}$ the first time they consume it.
@@
 We denote by $x_{i,n}$ individual $i$'s realized value from consuming item $n$.
-In particular, we consider that the realized utility derived from a given item can be decomposed in the following manner: $x_{i,n}= v_{i,n} + \beta v_n$, where $v_{i,n}$ denotes an idiosyncratic component -- i.e.
+In particular, we consider that the realized value derived from a given item can be decomposed in the following manner: $x_{i,n}= v_{i,n} + \beta v_n$, where $v_{i,n}$ denotes an idiosyncratic component -- i.e.
 user $i$'s idiosyncratic taste for product $n$ --  and $v_{n}$, a common-value component.
 One can interpret $v_n$ as a measure of how much product $n$ is valued in society in general and, in a sense, $v_{i,n}$ denotes how $i$ diverges from this overall ranking.
-The scalar $\beta \in \mathbb{R}_{t}$ denotes the degree to which utilities are idiosyncratic to each individual or common across individuals.
+The scalar $\beta \in \mathbb{R}_{t}$ denotes the degree to which valuations are idiosyncratic to each individual or common across individuals.
 If $\beta=0$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta$ is large, every individual has similar preferences.
 \par
-Stacking utilities in vector-form, we get the vector of values associated with each item $${\left(x_{i,n}\right)}_{n \in \mathcal{N}}=:X_i =V_i+ \beta V $$ where $V_i ={\left(v_{i,n}\right)}_{n \in \mathcal{N}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{N}}$.
+Stacking values in vector-form, we get the vector of values associated with each item 
+$${\left(x_{i,n}\right)}_{n \in \mathcal{J}}=:X_i =V_i+ \beta V $$
+where $V_i ={\left(v_{i,n}\right)}_{n \in \mathcal{J}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{J}}$.
 \par
-\noindent\textbf{User Decision-Making}
+\noindent\textbf{User Decision-Making.}
 We assume the user makes $T$ choices and therefore can only consume up to $T$ items, where $T$ is but a small fraction of $N$.
@@
 \par
-\noindent \textbf{User Beliefs} We assume that users do not know the realized utilities of the goods before consuming an item.
-Formally, consumer $i$ starts with some beliefs about $X_i$, namely that the idiosyncratic and common-value parts of the utilities are independent -- $V_i \perp \!\!\!
+\noindent \textbf{User Beliefs.} We assume that users do not know the realized utilities of the goods before consuming an item.
+Formally, consumer $i$ starts with some beliefs about $X_i$, namely that the idiosyncratic and common-value parts of the valuations are independent -- $V_i \perp \!\!\!
 \perp V$ -- and that each is multivariate normal
-\begin{enumerate}
+\begin{enumerate}[topsep=0pt]
 \item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ 
@@
 \end{enumerate}
-
 We impose the normality assumption for two reasons.
@@
 \par
-\begin{center}
+\begin{figure}[t]
 \includegraphics[width=.4\linewidth]{Example-Bubbles.pdf}
@@
 \label{fig:illustrative_example}
-\end{center}
+\end{figure}
 
@@
 
-\noindent \textbf{Recommendation}
+\noindent \textbf{Recommendation.}
 Our model of recommendation is stylized in order to provide qualitative insights into how recommendation shapes behavior, instead of focusing on the details of how recommender systems are implemented in practice.
@@
 We further consider two cases that serve primarily as benchmarks.
-The first is \textit{no recommendation}, where users get no additional information about utilities and make consumption choices based on their beliefs and consumption history.
+The first is \textit{no recommendation}, where users get no additional information about values and make consumption choices based on their beliefs and consumption history.
 This gives us a benchmark as to how users in our model would behave \textit{without} recommendation so that we can analyze what changes with the introduction of recommendation.
@@
 We explore how consumption patterns differ as we consider different recommendation regimes and report representative results from our simulations.
-We simulate our model over 100 populations of users with 100 users per population.
-For every population and every statistic we calculate, we average over the users in the population to get a representative statistic for this population.
-A given set of parameters and a population are a single data point in our dataset.
+We simulate our model over 100 populations of users with 100 users per population for each combination of parameters.
+%For every population and every statistic we calculate, we average over the users in the population to get a representative statistic for this population. 
+A given set of parameters and a user are a single data point in our dataset.
 \par
@@
 When we consider results varying a single parameter, we group the results over the other parameters and provide reports varying only the parameter of interest.
-We report results for a relatively small consumption history $T=20$ with a produce space size $N=200$.
+We report results for a relatively small consumption history $T=20$ with a product space size $N=200$.
 Unless otherwise reported, our results are robust to varying $N$.
@@
 \par
+
+\begin{figure}[t]
+\includegraphics[width=.9\linewidth]{figures/rho_pos_consumption_dist_N_200T_20_overall.pdf}\\
+\includegraphics[width=.9\linewidth]{figures/rho_zero_consumption_dist_N_200T_20_overall.pdf}\\
+\caption{Consumption path with (Top) and without (Bottom) correlation between values}
+\label{fig:correlation_consumption_path}
+\end{figure}
 
@@
 \subsection{Local Consumption}
-\begin{figure}
-\includegraphics[width=.9\linewidth]{figures/rho_pos_consumption_dist_N_200T_20_overall.pdf}\\
-\includegraphics[width=.9\linewidth]{figures/rho_zero_consumption_dist_N_200T_20_overall.pdf}\\
-\caption{Consumption path with (Top) and without (Bottom) correlation between values}
-\label{fig:correlation_consumption_path}
-\end{figure}
-
 We characterize ``filter bubble'' effects in our model as the degree to which users engage in \textit{local consumption}.
@@
 First, the bottom panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho = 0$, there is no difference in consumption distance between the three regimes.
-This is due to the fact that when $\rho = 0$, there is no reason that items that are close in the product space should have similar utilities and so the optimal consumption path does not depend on the similarity of the items.
+This is due to the fact that when $\rho = 0$, there is no reason that items that are close in the product space should have similar va;ues and so the optimal consumption path does not depend on the similarity of the items.
 However this also means that users do not learn anything about neighboring products and so there is limited path-dependence in consumption.
@@
 \begin{figure}[t]
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20.pdf}\\
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_partial.pdf}\\
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_omni.pdf}\\
+\includegraphics[width=0.9\linewidth]{figures/gamma_consumption_dist_N_200T_20.pdf}\\
+\includegraphics[width=0.9\linewidth]{figures/gamma_consumption_dist_N_200T_20_partial.pdf}\\
+\includegraphics[width=0.9\linewidth]{figures/gamma_consumption_dist_N_200T_20_omni.pdf}\\
 \caption{Effect of Risk Aversion on Local Consumption for No Recommendation (Top), Recommendation (Middle) and Oracle (Bottom)}
@@
 
-As before, when there is no correlation between utilities product diversity is the same across different recommendation regimes.
+As before, when there is no correlation between valuations, product diversity is the same across different recommendation regimes.
 The over-exploitation of the learning spillovers when $\rho > 0$ leads to product diversity being lowest in the no recommendation regime.
@@
 \begin{figure}[ht]
-\includegraphics[width=.9\linewidth]{figures/gamma_diversity_N_200_T_20}
+\includegraphics[width=.85\linewidth]{figures/gamma_diversity_N_200_T_20}
 \captionof{figure}{Risk Aversion and Diversity}\label{fig:risk_aversion_diversity}
@@
 \begin{figure}[t]
-\includegraphics[width=.9\linewidth]{figures/rho_diversity_N_200_T_20.pdf}\\
-\includegraphics[width=.9\linewidth]{figures/rho_welfare_N_200_T_20.pdf}\\
+\includegraphics[width=.85\linewidth]{figures/rho_diversity_N_200_T_20.pdf}\\
+\includegraphics[width=.85\linewidth]{figures/rho_welfare_N_200_T_20.pdf}\\
 \captionof{figure}{Welfare, Diversity, and Correlation}\label{fig:diversity_correlation}
@@
 \begin{figure}[t]
-\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_rn_partial.pdf}\\
-\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_ra_partial.pdf}\\
+\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_partial.pdf}\\
+\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_partial.pdf}\\
 \captionof{figure}{Diversity vs. Welfare, Recommendation, $\gamma = 0$ (Top), Diversity vs. Welfare, $\gamma = 5$ (Bottom)}\label{fig:diversity_welfare_ra_partial}
@@
 \begin{figure}[t]
-\includegraphics[width=.8\linewidth]{figures/beta_homogeneity_N_200_T_20}
+\includegraphics[width=.7\linewidth]{figures/beta_homogeneity_N_200_T_20}
 \captionof{figure}{Strength of Recommendation and Homogeneity}\label{fig:beta_homo}
@@
 \begin{figure}[t]
-\includegraphics[width=.8\linewidth]{figures/rho_homogeneity_N_200_T_20}
+\includegraphics[width=.7\linewidth]{figures/rho_homogeneity_N_200_T_20}
 \captionof{figure}{Correlation and Homogeneity}\label{fig:cor_homo}
@@
 
-As a result, several alternative recommendation evaluation metrics, such as serendipity \cite{kotkov2016survey}, coverage \cite{ge2010beyond}, novelty \cite{vargas2011rank}, and many others have been proposed in order to develop recommender systems that produce recommendations that are more useful for users.
+As a result, several alternative recommendation evaluation metrics, such as serendipity \cite{kotkov2016survey}, coverage \cite{ge2010beyond}, novelty \cite{vargas2011rank}, and many others, have been proposed in order to develop recommender systems that produce recommendations that are more useful for users.
 We use our model to motivate an alternative recommendation evaluation metric that provides an alternative perspective on the serendipity metric proposed in the literature.
@@
 The key insight is that a critical component of designing good recommendations is understanding what user behavior would be without recommendation and then understanding how recommendation would change their consumption.
-Operationalizing this requires collecting information on user beliefs about the goods in the product space, which is not data that is traditionally collected for the development of recommender systems.
+Operationalizing this requires collecting information on user beliefs about the goods in the product space, which is not data that is traditionally collected for recommender systems.
 \par
@@
 
-%We have argued that incorporating user choice under uncertainty should be a first-order component of RS design.
-By collecting appropriate data about user choices and user beliefs, RS can be built to better understand what choices users are likely to make and thus what information would be useful to give them as opposed to simply predicting what items a user will like.
-This approach can not only aid in designing more useful RS, but can also be utilized to better understand and prevent recently documented adverse effects of RS such as filter bubbles and homogenization.
-
-%\newpage
-
-%\subsection{User Homogenization}
-
-%
-%\begin{table}[bt]
-%\centering
-%\begin{tabular}{rlll}
-%  \hline
-%Rec Policy & Welfare & Diversity & Homogeneity \\ 
-%  \hline
-%No Rec & 0.42 $\pm$ 0.002 & 0.19 $\pm$ 0.0003 & 0.001 $\pm$ 0 \\ 
-%Partial Rec & 1.25 $\pm$ 0.005 & 0.22 $\pm$ 0.0002 & 0.174 $\pm$ 0.004 \\ 
-%Omniscient Rec & 2.24 $\pm$ 0.006 & 0.24 $\pm$ 0.0001 & 0.035 $\pm$ 0.002 \\ 
-%   \hline
-%\end{tabular}
-%\caption{Mean Welfare, Diversity, and Homogeneity for $N = 1000, T = 25, I = 50, S = 50$. Reported intervals are 95\% confidence intervals.}
-%\label{table:agg_results}
-%\end{table}
-%\begin{figure*}[!htbp]
-%    \centering
-%    \begin{minipage}[b]{0.5\textwidth}
-%        \centering
-%        \includegraphics[width=0.9\linewidth, height=1.5in]{local_moves_25}
-%        \caption{Extent of ``Local"  Consumption}
-%        \label{fig:local_consumption}
-%    \end{minipage}%
-%    ~ 
-%    \begin{minipage}[b]{0.5\textwidth}
-%        \centering
-%        \includegraphics[width=0.9\linewidth, height=1.5in]{welfare_homo_combo}
-%        \caption{Welfare and Homogeneity}
-%        \label{fig:beta_vary}
-%    \end{minipage}
-%    ~ \\
-%     \begin{minipage}{0.48\textwidth}
-%     \centering
-%     \includegraphics[width=0.9\linewidth, height=1.5in]{rec_obedience_25}
-%     \caption{Recommendation Effectiveness}\label{fig:rec_obey}
-%   \end{minipage}\hfill
-%   ~
-%   \begin{minipage}{0.48\textwidth}
-%     \centering
-%     \includegraphics[width=0.9\linewidth, height=1.5in]{Diversity_vs_Welfare_No_Recommendation}
-%     \caption{Diversity vs Welfare under No Recommendation}\label{fig:diversity_welfare_no_rec}
-%   \end{minipage}
-%\end{figure*}
-
-%We are going to analyze three outcomes: how diverse consumer choices are, how similar are the choices that different individuals make and the resulting welfare.
-The first will be measured by $D_i =\frac{1}{N} \frac{1}{T (T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$, the average normalized pairwise distance between the consumed products, a natural measure of diversity in this environment that is utilized in the literature \cite{ziegler2005improving}.
-The second will be measured by a consumer homogeneity index $H:=\frac{1}{|I|(|I|-1)}\sum_{i,j \in I: i \ne j}d_J(C_i^T,C_j^T) $, where $d_J$ denotes the Jaccard index and $H \in [0,1]$ which is similar to the approach to measuring homogeneity utilized in \cite{chaney2018algorithmic}.
-Note that a higher $H$ indicates more homogeneity.
-Consumer $i$'s welfare will just be the average of realized utilities, to control for the effect of $T$, $W_i=\frac{1}{T}\sum_{n \in C_i^T} u_{in}$.
-%\par
-%\par
-%\noindent \textbf{Results}.
-Table \ref{table:agg_results} shows aggregate results for welfare, diversity, and homogeneity.
-The omniscient recommendation case leads to the optimal consumption path for a consumer since she consumes the $T$ items with the highest utility.
-Compared to this case, both partial recommendation and no recommendation not only lead to lower welfare but also have lower diversity.
-In fact, the no recommendation case leads to the lowest levels of consumption diversity since users do not explore the space of products sufficiently and do excessive amounts of local search as can be seen in Figure \ref{fig:local_consumption}.
-Figure \ref{fig:local_consumption} shows how ``local" consumption choices are by looking at what fraction of consecutive consumption choices are close to each other with the definition of close being less than distance of 10.
-The partial recommendation case leads to higher consumption diversity and higher welfare, but also significantly higher homogeneity.
-%
-%In our model, filter bubble effects occur simply because of the correlation structure between the utility and beliefs of products.
-It highlights the fact that the correlation structure induces users to consume increasingly narrow content when they get no guidance from recommendation whereas the ``optimal" consumption path given by the omniscient policy does not exhibit this.
-The results of our model are consistent with the empirical results found in \cite{nguyen2014exploring} who show that filter bubble effects can arise even amongst users who do not utilize recommendations.
-%\par
-%Furthermore, \cite{nguyen2014exploring} also find that recommendation can lead to an increase in consumption diversity.
-As Table \ref{table:agg_results} shows, our model is consistent with this effect but it depends on the nature of recommendation.
-If recommendations only provides information on the common-value component (partially informed recommender) then welfare and consumption diversity increase when contrasted to the no recommendation case, but homogenization increases dramatically.
-Knowing the common-value component leads users to have more information on where to explore in the product space.
-However, the fact that this information is about the common-value component and the same across all individuals leads them to explore the same sections of the product space and leads to substantially higher homogeneity among users.
-If recommendation takes into account user beliefs on the idiosyncratic component of utility (omniscient recommender), then welfare and consumption diversity can further increase and homogeneity will decrease.
-%\par
-%Low consumption diversity is not necessarily bad by itself.
-Figure \ref{fig:diversity_welfare_no_rec} shows that, in the no recommendation case, welfare and consumption diversity are negatively correlated.
-This is primarily because content diversity in the no recommendation case can arise from the fact that the user consumes a bad product and is forced to jump around the product space looking for ``good" products.
-However, if a user finds a high utility product right away, then staying in this neighborhood may yield higher utility due to the correlation of utilities.
-The problem is that, even when the user finds a ``good" product, this may only be ``locally" good and the user will then excessively consume items around it due to having more information about these than more ``different'' ones, i.e.
-farther away.
-%\par
-%Finally, the value of recommendation in the partial recommendation case decreases over time as users learn more about the idiosyncratic component of their own preferences (i.e.
-as $t$ increases).
-Figure \ref{fig:rec_obey} shows that recommendation compliance decreases as $t$ increases.
-Additionally, Figure \ref{fig:beta_vary} shows how homogeneity and welfare change as $\beta$ increases.
-Most interestingly, when $\beta$ is high so that the common-value component is small, recommendation can be harmful to users.
-The information from recommendation still leads to large user homogeneity even though the optimum would be for there to be almost none and leads to lower welfare than under no recommendation.
-%
-%\section{The Value of the Ex-Ante View}
-%
-%In this section we illustrate the differences between the ex-post and ex-ante view and show how it is useful for thinking about the design of RS.
-Our first point is trivial to state - since users do not know the true consumption values of the items, without any further information they may make \textit{ex-post} sub-optimal consumption decisions, but \textit{ex-ante} optimal given the information available to them at the moment of choice.
-Consider the following stylized example.
-The choice set of the individual is $\Omega= \{x_1, x_2, x_3 \}$ and the ex-post utility values are given by
-%$u(x_1) = 1$, $u(x_2) = 2$, $u(x_3) = 3$. 
-%\par
-%However, when users are making decisions they have unobservable \textit{beliefs} about the ex-post utility value of the items and make decisions given these beliefs.
-As users obtain more information, their beliefs converge to the ex-post utility values but, generally, users do not have enough information and so have beliefs over the space of possible ex-post utility values.
-In this example we suppose these beliefs are simple so that users view each good as simple lotteries: for $n=1,2,3$, $u(x_n) = n + \varepsilon_n$, $\varepsilon_1 \sim \mathcal J (2, 1)$, $\varepsilon_2 \sim \mathcal J (0, 1)$, $\varepsilon_3 \sim \mathcal J (-5, 2)$.
-%\par
-%After uncertainty is unresolved, the user would consume $x_3$, but an expected utility maximizing consumer (without further information) would choose $x_1$.\footnote{We ignore an important component of decision-making under uncertainty, risk attitudes, by focusing on risk-neutral consumers.
-A \textit{risk-averse} user may choose to consume a good with lower expected \textit{ex-post} payoff but lower ex-ante variance compared to another good.
-While risk-aversion is important for decision-making under uncertainty, especially in the context of RS, it is not our main focus here.}
-%\par
-%This observation, while seemingly trivial, has important implications.
-The fact that the user would choose $C$ over $A$ reveals that $C$ prefers $A$ given her current beliefs\footnote{This is a statement of \textit{revealed preference} which has been studied heavily in economic decision theory (see e.g.
-\cite[]{mas1995microeconomic}).} and allows the system designer to get some information about a user's beliefs.
-Why are user beliefs important objects for designers of RS to understand and utilize?
-%\par
-%Beliefs are useful for understanding how users interact and get value from RS.
-In particular, users use RS to get \textit{information} about the set of items and update their beliefs about the value of the items.
-For instance, in the example, a RS could recommend to a user $x_3$ over $x_2$ and $x_1$ and that user will use this to update her beliefs about $x_3$.
-It is an empirical question that we leave for future work precisely how and to what extent users update their beliefs from recommendation and what factors of RS deisgn influence this.\footnote{There has been some work in this direction in understanding when recommendations are persuasive \cite{cremonesi2012investigating, gretzel2006persuasion} as well as the effect of their delivery on effectiveness \cite{murphy2014recommendation}.}
-%\par
-%Better understanding user beliefs is important for improving the performance of RS that optimize for metrics that move away from accuracy such as serendipity \cite{kotkov2016survey}.
-Serendipity-based RS attempt to provide recommendations that ``have the quality of being both unexpected and useful" \cite{maksai2015predicting} but it is hard to know what items are both unexpected and useful \cite{kotkov2016survey}.
-Understanding \textit{why} a user has not consumed a good depends on the beliefs that the user has and this is important for designing serendipitious recommendations.
-For instance, in the example, the user does not consume $x_3$ because she expects the good to turn out worse than the other two.
-However, if given more information, she would prefer to consume it over $x_1$ and $x_2$.
-Moreover, when the user can use her known valuation of an item to make inferences on how good another item is, consumption is path-dependent, as beliefs guide consumption and consumption affects beliefs about other goods.
-For instance, learning the value of $x_1$ says something about how good $x_2$ is.
-%When considering risk-averse individuals, when the user has alternatives with a known value,The consumer may have also consumed Another reason why the user may not .
-The user may have an alternative with a known value -- say going for a walk around the neighborhood -- which she could always prefer to watching a given movie that has uncertain value but that the consumer thinks is always lower than the value of the walk.\footnote{An additional reason this could be useful is if one considers a risk-averse individual, then it may be that a user has not consumed an item because she has a high perceived variance for the item's quality.}
-%As a result, understanding the beliefs of users helps us understand why users have not consumed certain items and, in particular, which items a user may find useful given more information about it.
-This poses an important problem: designing systems that elicit not only ex-post valuations but also ex-ante beliefs about valuations, as only by knowing these can RS be effective in steering users' choices.
-In particular, choice has to be perceived as guided by ex-ante utility, whereas ratings by the ex-post realized utility.
-%\par
-%Understanding the choices that individuals will make in addition to what they will end up liking are complementary problems but require different viewpoints.
-\cite{jiang2014choice, saavedra2016choice} argue that choice-based approaches alone can help in designing better RS, though they argue for these approaches because they do a better job at providing more accurate recommendations.
-We argue that the two problems should be considered separately -- though they may interact in interesting and useful ways.\footnote{In particular, some results in \cite{jiang2014choice, saavedra2016choice} may be driven by the fact that user choice embeds ratings information that may not have been observed by the recommender but is observed by the user.
-Ratings, reviews, friends, there are many sources of information affecting a person's beliefs and choices that are unobservable by RS.
-Other consumption choices, e.g.
-movies seen in the cinema, change beliefs about for instance how good a director is, which affect beliefs about other movies and guide choices on a streaming platform that is unable to observe these data.}
-%\par
-%RS are designed to give users \textit{information} that is useful to help them make decisions.
-On the one hand, predicting a user's ex-post preferences is useful to know what items the user would like to consume.
-On the other hand, predicting what items a user will actually consume (without recommendation) can help reason about what information may actually be useful to give her.
-Furthermore, in designing RS to avoid filter bubble and homogenization effects, having accurate choice predictions allows mitigating such adverse effects to become a first-order component of design.
-In particular, by predicting both choice and ratings, RS can provide information to users that leads them to useful items but prevents them from falling into a filter bubble.
-%\par
-%Solving this choice prediction problem requires data on the choices that individuals make and the sets that they choose them from in addition to the traditional ratings or behavioral data that is collected.
-In the example given previously the fact that the consumer chooses $x_1$ from the set $\{x_1, x_2, x_3 \}$ would be the data that is useful for the choice prediction problem, as would the consumer's ex-ante beliefs about her valuation of the goods.
-%
-%
-%\section{Conclusion}
-%
-%We have argued that incorporating user choice under uncertainty should be a first-order component of RS design.
-By collecting appropriate data about user choices and user beliefs, RS can be built to better understand what choices users are likely to make and thus what information would be useful to give them as opposed to simply predicting what items a user will like.
-This approach can not only aid in designing more useful RS, but can also be utilized to better understand and prevent recently documented adverse effects of RS such as filter bubbles and homogenization.
-
 \bibliographystyle{plain}

### 2019-11-19 simulation_code/collab_filtering.jl
foo

@@ -3,6 +3,10 @@ User based Collaborative Filtering
 
 dumb example
 
+A Survey of Collaborative Filtering Techniques
+Xiaoyuan Su and Taghi M. Khoshgoftaar
+http://downloads.hindawi.com/archive/2009/421425.pdf
+
 https://realpython.com/build-recommendation-engine-collaborative-filtering/
 https://medium.com/@cfpinela/recommender-systems-user-based-and-item-based-collaborative-filtering-5d5f375a127f
 

### 2019-11-19 paper/filter_bubbles.tex
update figures'

@@
-\documentclass[sigconf, review, anonymous]{acmart}
+\documentclass[sigconf]{acmart}
 \settopmatter{printacmref=false}
@@
 \title{A User-Centric Approach to the Design and Consequences of Recommender Systems}
-
 %
@@
 When we consider results varying a single parameter, we group the results over the other parameters and provide reports varying only the parameter of interest.
-We report results for a relatively small consumption history $T=20$ with a product space size $N=200$.
+We report results for a relatively small consumption history $T=20$ with a produce space size $N=200$.
 Unless otherwise reported, our results are robust to varying $N$.
@@
 \begin{figure}[t]
-\includegraphics[width=0.9\linewidth]{figures/gamma_consumption_dist_N_200T_20.pdf}\\
-\includegraphics[width=0.9\linewidth]{figures/gamma_consumption_dist_N_200T_20_partial.pdf}\\
-\includegraphics[width=0.9\linewidth]{figures/gamma_consumption_dist_N_200T_20_omni.pdf}\\
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20.pdf}\\
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_partial.pdf}\\
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_omni.pdf}\\
 \caption{Effect of Risk Aversion on Local Consumption for No Recommendation (Top), Recommendation (Middle) and Oracle (Bottom)}
@@
 \begin{figure}[ht]
-\includegraphics[width=.85\linewidth]{figures/gamma_diversity_N_200_T_20}
+\includegraphics[width=.9\linewidth]{figures/gamma_diversity_N_200_T_20}
 \captionof{figure}{Risk Aversion and Diversity}\label{fig:risk_aversion_diversity}
@@
 \begin{figure}[t]
-\includegraphics[width=.85\linewidth]{figures/rho_diversity_N_200_T_20.pdf}\\
-\includegraphics[width=.85\linewidth]{figures/rho_welfare_N_200_T_20.pdf}\\
+\includegraphics[width=.9\linewidth]{figures/rho_diversity_N_200_T_20.pdf}\\
+\includegraphics[width=.9\linewidth]{figures/rho_welfare_N_200_T_20.pdf}\\
 \captionof{figure}{Welfare, Diversity, and Correlation}\label{fig:diversity_correlation}
@@
 \begin{figure}[t]
-\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_partial.pdf}\\
-\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_partial.pdf}\\
+\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_rn_partial.pdf}\\
+\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_ra_partial.pdf}\\
 \captionof{figure}{Diversity vs. Welfare, Recommendation, $\gamma = 0$ (Top), Diversity vs. Welfare, $\gamma = 5$ (Bottom)}\label{fig:diversity_welfare_ra_partial}
@@
 \begin{figure}[t]
-\includegraphics[width=.7\linewidth]{figures/beta_homogeneity_N_200_T_20}
+\includegraphics[width=.8\linewidth]{figures/beta_homogeneity_N_200_T_20}
 \captionof{figure}{Strength of Recommendation and Homogeneity}\label{fig:beta_homo}
@@
 \begin{figure}[t]
-\includegraphics[width=.7\linewidth]{figures/rho_homogeneity_N_200_T_20}
+\includegraphics[width=.8\linewidth]{figures/rho_homogeneity_N_200_T_20}
 \captionof{figure}{Correlation and Homogeneity}\label{fig:cor_homo}
@@
 
-As a result, several alternative recommendation evaluation metrics, such as serendipity \cite{kotkov2016survey}, coverage \cite{ge2010beyond}, novelty \cite{vargas2011rank}, and many others, have been proposed in order to develop recommender systems that produce recommendations that are more useful for users.
+As a result, several alternative recommendation evaluation metrics, such as serendipity \cite{kotkov2016survey}, coverage \cite{ge2010beyond}, novelty \cite{vargas2011rank}, and many others have been proposed in order to develop recommender systems that produce recommendations that are more useful for users.
 We use our model to motivate an alternative recommendation evaluation metric that provides an alternative perspective on the serendipity metric proposed in the literature.
@@
 The key insight is that a critical component of designing good recommendations is understanding what user behavior would be without recommendation and then understanding how recommendation would change their consumption.
-Operationalizing this requires collecting information on user beliefs about the goods in the product space, which is not data that is traditionally collected for recommender systems.
+Operationalizing this requires collecting information on user beliefs about the goods in the product space, which is not data that is traditionally collected for the development of recommender systems.
 \par

### 2019-12-11 paper/ec_submission/filter_bubble_ec.tex
Trying to work on ec formating

--- 2019-11-19 paper/filter_bubbles.tex
@@
-\documentclass[sigconf]{acmart}
-\settopmatter{printacmref=false}
-\usepackage[T1]{fontenc}
-%\usepackage{lmodern}
-\usepackage{float}
-\usepackage{enumitem}
+\documentclass[format=acmsmall, review=false]{acmart}
+\usepackage{acm-ec-20}
+\usepackage{booktabs} % For formal tables
+\usepackage[ruled]{algorithm2e} % For algorithms
+
+\usepackage{subcaption}
+
+\renewcommand{\algorithmcfname}{ALGORITHM}
+\SetAlFnt{\small}
+\SetAlCapFnt{\small}
+\SetAlCapNameFnt{\small}
+\SetAlCapHSkip{0pt}
+\IncMargin{-\parindent}
+
+% Choose a citation style by commenting/uncommenting the appropriate line:
+%\setcitestyle{authoryear}
+\setcitestyle{acmnumeric}
+
 \newtheorem{finding}{Finding}
-%
-% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
-\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
 
-% Rights management information. 
-% This information is sent to you when you complete the rights form.
-% These commands have SAMPLE values in them; it is your responsibility as an author to replace
-% the commands and values with those provided to you when you complete the rights form.
-%
-% These commands are for a PROCEEDINGS abstract or paper.
-\copyrightyear{2020}
-\acmYear{2020}
-\setcopyright{acmlicensed}
-\acmConference[WWW'20]{Recomme '18: ACM Symposium on Neural Gaze Detection}{April 20-24, 2020}{Taipai, Taiwan}
-\acmBooktitle{RecSys '19: ACM Conference on Neural Gaze Detection, June 03--05, 2018, Woodstock, NY}
-\acmPrice{15.00}
-\acmDOI{10.1145/1122445.1122456}
-\acmISBN{978-1-4503-9999-9/18/06}
+% Title. Note the optional short title for running heads. In the interest of anonymization, please do not include any acknowledgements.
+\newcommand\PaperTitle{A User-Centric Approach to the Design and Consequences of Recommender Systems}
+\title[\PaperTitle]{\PaperTitle}
 
-%
-% These commands are for a JOURNAL article.
-%\setcopyright{acmcopyright}
-%\acmJournal{TOG}
-%\acmYear{2018}\acmVolume{37}\acmNumber{4}\acmArticle{111}\acmMonth{8}
-%\acmDOI{10.1145/1122445.1122456}
+% Anonymized submission.
+\author{Submission 42?}
 
-%
-% Submission ID. 
-% Use this when submitting an article to a sponsored event. You'll receive a unique submission ID from the organizers
-% of the event, and this ID should be used as the parameter to this command.
-%\acmSubmissionID{123-A56-BU3}
-
-%
-% The majority of ACM publications use numbered citations and references. If you are preparing content for an event
-% sponsored by ACM SIGGRAPH, you must use the "author year" style of citations and references. Uncommenting
-% the next command will enable that style.
-%\citestyle{acmauthoryear}
-
-%
-% end of the preamble, start of the body of the document source.
-\begin{document}
-\setlength{\abovedisplayskip}{0pt}
-\setlength{\belowdisplayskip}{0pt}
-% end of the preamble, start of the body of the document source.
-
-%
-% The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
-\title{A User-Centric Approach to the Design and Consequences of Recommender Systems}
-%
-% By default, the full list of authors will be used in the page headers. Often, this list is too long, and will overlap
-% other information printed in the page headers. This command allows the author to define a more concise list
-% of authors' names for this purpose.
-
-\settopmatter{printacmref=false} % Removes citation information below abstract
-\makeatletter
-\def\-copyrightspace{\relax}
-\makeatother
-
-%
-% The abstract is a short summary of the work to be presented in the article.
+% Abstract. Note that this must come before \maketitle.
 \begin{abstract}
@@
 
-%
-% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-% Please copy and paste the code instead of the example below.
-%
-\author{Guy Aridor}
-\affiliation{
-  \institution{Columbia University \\ Department of Economics}}
-
-\author{Duarte Gon\c{c}alves}
-\affiliation{%
-  \institution{Columbia University, BRU-IL \\ Department of Economics}}
-
-  \author{Shan Sikdar}
-  \affiliation{Independent Researcher}
-%
-% Keywords. The author(s) should pick words that accurately describe the work being
-% presented. Separate the keywords with commas.
-%\keywords{Recommender Systems, Filter Bubbles, Serendipity}
-
-%
-% This command processes the author and affiliation and title information and builds
-% the first part of the formatted document.
+\begin{document}
 
@@
 
+% Paper body
 \section{Introduction}
@@
 
-Recommender systems have been influential in shaping consumer choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{MacKenzie et al.
-(2013, Oct.),  How retailers can keep up with consumers.
+Recommender systems have been influential in shaping consumer choice in these markets with 75\% of movies watched on Netflix and 35\% 
+of page-views on Amazon coming from recommendations.
+\footnote{MacKenzie et al. (2013, Oct.),  How retailers can keep up with consumers. 
 \url{https://www.mckinsey.com/industries/retail/our-insights/how-retailers-can-keep-up-with-consumers}. 
-Retrieved on October 3, 2019.} While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems.
-There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals,\footnote{Friedersdorf (2018, Mar.
-12), YouTube Extremism and the Long Tail, \textit{The Atlantic}.
-\url{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350}.
-Retrieved on October 3, 2019.} that personalized recommender systems lead users into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
+Retrieved on October 3, 2019.} 
+While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems. 
+There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals,
+\footnote{Friedersdorf (2018, Mar. 12), YouTube Extremism and the Long Tail, 
+\textit{The Atlantic}. 
+\url{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350}. Retrieved on October 3, 2019.} 
+that personalized recommender systems lead users into \textit{filter bubbles} 
+where they effectively get isolated from a diversity of viewpoints or content 
+\cite{pariser2011filter}, and that personalized recommender systems may also 
+lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
 \par
@@
 The key components of our model are:
-\begin{enumerate}[label=(\arabic*)]
+\begin{enumerate}
 \item Users face uncertainty about their valuation of the different goods and thus face a decision problem under under uncertainty.
@@
 \section{Our Model and Preliminaries}
-
 \subsection{Preliminaries on Expected Utility Theory}
@@
 \perp V$ -- and that each is multivariate normal
-\begin{enumerate}[topsep=0pt]
+\begin{enumerate}
 \item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ 
@@
 \begin{figure}[t]
-\includegraphics[width=.4\linewidth]{Example-Bubbles.pdf}
+\includegraphics[width=.4\linewidth]{../Example-Bubbles.pdf}
 \captionof{figure}{Illustrative Example}
@@
 
-We simulate over risk-aversion parameters $\gamma \in \{ 0, 0.3, 0.6, 1, 5 \}$, standard-deviation $\sigma \in \{ 0.25, 0.5, 1.0, 2.0, 4.0 \}$, correlation parameters $\rho\in \{ 0, 0.1, 0.3, 0.5, 0.7, 0.9 \} $ and degree of idiosyncrasy of individual valuations $\beta \in \{ 0, 0.4, 0.8, 1, 2, 5\}$.
+We simulate over risk-aversion parameters $\gamma \in \{ 0, 0.3, 0.6, 1, 5 \}$, \ standard-deviation $\sigma \in \{ 0.25, 0.5, 1.0, 2.0, 4.0 \}$, correlation parameters $\rho\in \{ 0, 0.1, 0.3, 0.5, 0.7, 0.9 \} $ and degree of idiosyncrasy of individual valuations $\beta \in \{ 0, 0.4, 0.8, 1, 2, 5\}$.
 When we consider results varying a single parameter, we group the results over the other parameters and provide reports varying only the parameter of interest.
@@
 
-\begin{figure}[t]
-\includegraphics[width=.9\linewidth]{figures/rho_pos_consumption_dist_N_200T_20_overall.pdf}\\
-\includegraphics[width=.9\linewidth]{figures/rho_zero_consumption_dist_N_200T_20_overall.pdf}\\
+% \begin{figure}[t]
+% \includegraphics[width=.9\linewidth]{../figures/rho_pos_consumption_dist_N_200T_20_overall.pdf}\\
+% \includegraphics[width=.9\linewidth]{../figures/rho_zero_consumption_dist_N_200T_20_overall.pdf}\\
+% \caption{Consumption path with (Top) and without (Bottom) correlation between values}
+% \label{fig:correlation_consumption_path}
+% \end{figure}
+
+\begin{figure}
+\begin{subfigure}{.5\linewidth}
+  \centering
+  \includegraphics[width=.8\linewidth]{../figures/rho_pos_consumption_dist_N_200T_20_overall.pdf}
+  \caption{1a}
+  \label{fig:sfig1}
+\end{subfigure}%
+\begin{subfigure}{.5\linewidth}
+  \centering
+  \includegraphics[width=.8\linewidth]{../figures/rho_zero_consumption_dist_N_200T_20_overall.pdf}
+  \caption{1b}
+  \label{fig:sfig2}
+\end{subfigure}
 \caption{Consumption path with (Top) and without (Bottom) correlation between values}
@@
 
+% \begin{figure}[t]
+% \includegraphics[width=\linewidth]{../figures/rho_consumption_dist_N_200T_20.pdf}\\
+% \includegraphics[width=\linewidth]{../figures/rho_consumption_dist_N_200T_20_partial.pdf}\\
+% \includegraphics[width=\linewidth]{../figures/rho_consumption_dist_N_200T_20_omni.pdf}\\
+% \caption{Extent of Local Consumption for No Recommendation (Top), Recommendation (Middle) and Oracle (Bottom) for different Correlation values}
+% \label{fig:local_consumption_across_rho}
+% \end{figure}
+
 \begin{figure}[t]
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20.pdf}\\
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_partial.pdf}\\
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_omni.pdf}\\
+\begin{subfigure}{.5\textwidth}
+\includegraphics[width=\linewidth]{../figures/rho_consumption_dist_N_200T_20.pdf}
+\end{subfigure}\\
+\begin{subfigure}{.5\textwidth}
+\includegraphics[width=\linewidth]{../figures/rho_consumption_dist_N_200T_20_partial.pdf}
+\end{subfigure}\\
+\begin{subfigure}{.5\textwidth}
+\includegraphics[width=\linewidth]{../figures/rho_consumption_dist_N_200T_20_omni.pdf}\\
 \caption{Extent of Local Consumption for No Recommendation (Top), Recommendation (Middle) and Oracle (Bottom) for different Correlation values}
+\end{subfigure}%
 \label{fig:local_consumption_across_rho}
@@
 \begin{figure}[t]
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20.pdf}\\
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_partial.pdf}\\
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_omni.pdf}\\
+\begin{subfigure}{.5\textwidth}
+\includegraphics[width=\linewidth]{../figures/gamma_consumption_dist_N_200T_20.pdf}
+\end{subfigure}\\
+\begin{subfigure}{.5\textwidth}
+\includegraphics[width=\linewidth]{../figures/gamma_consumption_dist_N_200T_20_partial.pdf}
+\end{subfigure}\\
+\begin{subfigure}{.5\textwidth}
+\includegraphics[width=\linewidth]{../figures/gamma_consumption_dist_N_200T_20_omni.pdf}\\
+\end{subfigure}%
 \caption{Effect of Risk Aversion on Local Consumption for No Recommendation (Top), Recommendation (Middle) and Oracle (Bottom)}
@@
 \begin{figure}[ht]
-\includegraphics[width=.9\linewidth]{figures/gamma_diversity_N_200_T_20}
+\includegraphics[width=.4\linewidth]{../figures/gamma_diversity_N_200_T_20}
 \captionof{figure}{Risk Aversion and Diversity}\label{fig:risk_aversion_diversity}
@@
 \par
-\begin{figure}[t]
-\includegraphics[width=.9\linewidth]{figures/rho_diversity_N_200_T_20.pdf}\\
-\includegraphics[width=.9\linewidth]{figures/rho_welfare_N_200_T_20.pdf}\\
+
+% \begin{figure}[t]
+% \includegraphics[width=.9\linewidth]{../figures/rho_diversity_N_200_T_20.pdf}\\
+% \includegraphics[width=.9\linewidth]{../figures/rho_welfare_N_200_T_20.pdf}\\
+% \captionof{figure}{Welfare, Diversity, and Correlation}\label{fig:diversity_correlation}
+% \end{figure}
+
+\begin{figure}
+\begin{subfigure}{.5\linewidth}
+  \includegraphics[width=.8\linewidth]{../figures/rho_diversity_N_200_T_20.pdf}
+\end{subfigure}\\
+\begin{subfigure}{.5\linewidth}
+  \includegraphics[width=.8\linewidth]{../figures/rho_welfare_N_200_T_20.pdf}
+\end{subfigure}
 \captionof{figure}{Welfare, Diversity, and Correlation}\label{fig:diversity_correlation}
@@
 
-%\begin{figure}
-
-%\end{figure}
-
 This only happens since $\gamma = 0$ leads to users only caring about the expected value of the product.
@@
 \begin{figure}[t]
-\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_rn.pdf}\\
-\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_ra.pdf}\\
+\begin{subfigure}{.5\textwidth}
+\includegraphics[width=1.05\linewidth]{../figures/diversity_welfare_rn.pdf}
+\end{subfigure}\\
+\begin{subfigure}{.5\textwidth}
+\includegraphics[width=1.05\linewidth]{../figures/diversity_welfare_ra.pdf}
+\end{subfigure}\\
 \captionof{figure}{Diversity vs. Welfare, No-Recommendation, $\gamma = 0$ (Top), Diversity vs. Welfare, $\gamma = 5$ (Bottom)}\label{fig:diversity_welfare_ra}
 \end{figure}
+
 \begin{figure}[t]
-\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_rn_partial.pdf}\\
-\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_ra_partial.pdf}\\
+\begin{subfigure}{.5\textwidth}
+\includegraphics[width=1.05\linewidth]{../figures/diversity_welfare_rn_partial.pdf}
+\end{subfigure}\\
+\begin{subfigure}{.5\textwidth}
+\includegraphics[width=1.05\linewidth]{../figures/diversity_welfare_ra_partial.pdf}
+\end{subfigure}\\
 \captionof{figure}{Diversity vs. Welfare, Recommendation, $\gamma = 0$ (Top), Diversity vs. Welfare, $\gamma = 5$ (Bottom)}\label{fig:diversity_welfare_ra_partial}
@@
 \begin{figure}[t]
-\includegraphics[width=.8\linewidth]{figures/beta_homogeneity_N_200_T_20}
+\includegraphics[width=.4\linewidth]{../figures/beta_homogeneity_N_200_T_20}
 \captionof{figure}{Strength of Recommendation and Homogeneity}\label{fig:beta_homo}
@@
 \begin{figure}[t]
-\includegraphics[width=.8\linewidth]{figures/rho_homogeneity_N_200_T_20}
+\includegraphics[width=.4\linewidth]{../figures/rho_homogeneity_N_200_T_20}
 \captionof{figure}{Correlation and Homogeneity}\label{fig:cor_homo}
@@
 
-\bibliographystyle{plain}
-\bibliography{refs}
+% Start of "Sample References" section
+% In the interest of anonymization, please do not include acknowledgements in your submission.
+%
+%\begin{acks}
+%
+%	The authors would like to thank Dr. Maura Turolla of Telecom
+%	Italia for providing specifications about the application scenario.
+%
+%	The work is supported by the \grantsponsor{GS501100001809}{National
+%		Natural Science Foundation of
+%		China}{http://dx.doi.org/10.13039/501100001809} under Grant
+%	No.:~\grantnum{GS501100001809}{61273304\_a}
+%	and~\grantnum[http://www.nnsf.cn/youngscientsts]{GS501100001809}{Young
+%		Scientsts' Support Program}.
+%
+%
+%\end{acks}
+
+% Bibliography
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{../refs}
+
 \end{document}

### 2019-12-14 paper/www20-rebuttal.txt
add rebuttal, WINE poster, and fix typos

[Review1and2]: We focused on building a model based on assumptions backed by work in psychology and economic decision theory. An online experiment with real users is a great idea and will be addressed in future work.
Clarity issues will be addressed in the revision.

[Review2Q1]: At the start V_i and V are drawn for each product and individual so that realized value (and utility) of each item is fixed throughout. Each user has correct initial beliefs about the distribution these are drawn from, a standard assumption in decision theory. As the user observes realized values of items they consume (x_{i,n}'s), they update beliefs on the distribution of values of the remaining items. The reason this updating is meaningful is that these values are correlated. The crucial point is that there is a difference between the perceived ("ex-ante") expected utility driving users’ decisions and the true/realized ("ex-post") utility. The user’s decisions depend on how much they know about the items, but the realized utility does not.

The realized X_i(=V_i+\beta*V) is fixed, only uncovered upon consumption. Without recommendation, user's beliefs are a combination of uncertainty over V and V_i. Recommendation reveals the realized V, but does not affect uncertainty over V_i.

Realized utility is fixed throughout so oracle recommendation does not need to account for user beliefs.

[Review2Q2]: We should have and will provide references from decision theory. We use a standard formulation of risk preferences, CARA preferences. The certainty equivalent of an item is homeomorphic to its expected utility, where risk attitudes play a role. The certainty equivalent of an item is the certain value x (line 248) that leads to the same expected utility: u(x)=E[u(item)]. As such, it is sufficient to know the certainty equivalents to which items have highest expected utility. In a Gaussian environment with CARA preferences, the expected utility of an item is increasing in mean and decreasing in variance (line 313), capturing users’ dislike of uncertainty. Higher risk aversion implies a higher penalty to uncertainty when ranking items. In our setup, having explored more nearby items decreases uncertainty, which may lead consumers to prefer safer items with lower expected values over riskier items with higher means.

[Review2Q3]: The illustrative example is under-specified as the arguments hold for any variance: information on item 0’s value impacts beliefs about both mean and variance and, under risk-aversion, both are relevant in determining user’s choices. Omitted algebraic details provided in the revision.

[Review2Q4]: We suppose the recommender knows V at t=0 and does not need to learn them. Having the recommender also learning V is an interesting extension, worth studying.

[Review3]: Existing approaches to serendipity and diversity focus on predicting realized utilities and "diversifying" the recommended itemset. Our approach emphasizes the role of beliefs and using as an evaluation metric the recommendation’s impact on user behavior.

### 2019-12-14 paper/filter_bubbles.tex
add rebuttal, WINE poster, and fix typos

@@
 \setcopyright{acmlicensed}
-\acmConference[WWW'20]{Recomme '18: ACM Symposium on Neural Gaze Detection}{April 20-24, 2020}{Taipai, Taiwan}
+\acmConference[WWW'20]{Recomme '18: ACM Symposium on Neural Gaze Detection}{April 20-24, 2020}{Taipei, Taiwan}
 \acmBooktitle{RecSys '19: ACM Conference on Neural Gaze Detection, June 03--05, 2018, Woodstock, NY}
@@
 
-\textbf{Our Model and Contributions} In this paper, 
+\noindent \textbf{Our Model and Contributions} In this paper, 
 %drawing from recent work in psychology and decision theory in economics, 
@@
 \begin{enumerate}[label=(\arabic*)]
-\item Users face uncertainty about their valuation of the different goods and thus face a decision problem under under uncertainty.
+\item Users face uncertainty about their valuation of the different goods and thus face a decision problem under uncertainty.
 \item Choice sets are large and users only consume a small fraction of this choice set over their lifetime.
@@
 First, the bottom panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho = 0$, there is no difference in consumption distance between the three regimes.
-This is due to the fact that when $\rho = 0$, there is no reason that items that are close in the product space should have similar va;ues and so the optimal consumption path does not depend on the similarity of the items.
+This is due to the fact that when $\rho = 0$, there is no reason that items that are close in the product space should have similar values and so the optimal consumption path does not depend on the similarity of the items.
 However this also means that users do not learn anything about neighboring products and so there is limited path-dependence in consumption.

### 2020-01-09 paper/filter_bubbles.tex
basic changes

--- 2019-11-19 paper/filter_bubbles.tex
@@
 We study a model of user decision-making in the context of recommender systems.
-We explore the extent to which two issues traditionally associated with recommender systems -- \textit{filter bubbles} and \textit{user homogenization} -- can be attributed to recommendation systems.
+We explore the extent to which two issues -- \textit{filter bubbles} and \textit{user homogenization} -- can be attributed to recommendation systems.
 The key component of our model is that users face a decision problem under uncertainty and that there are informational spillovers where a user consuming a product and learning how she values it reduces uncertainty about similar products.
@@
 Traditional design approaches to recommender systems only optimize for predicting user preferences after uncertainty has been resolved and ignore user beliefs and the inferences that users themselves make after consuming items.
-We propose an alternative interpretation of the serendipity measure where good recommendations are those that provide useful information, that is, information without which choice would be different and lead to worse outcomes on average relative to the case where no recommendation is provided.
+We propose an alternative interpretation of the serendipity measure where good recommendations are those that provide useful information, without which choice would lead to worse outcomes on average relative to when no recommendation is provided.
 As a result, we emphasize the importance of understanding user behavior without recommendation, both for designing recommendations and understanding their impacts and usefulness.
@@
 For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
-Users in many cases are not aware of most items, let alone have full information about their preferences over them and, to make matters worse, the goods in these markets are usually experience goods whose true value can only be learned after consumption.
+Users often are not aware of most items, let alone have full information about their preferences over them.
+To make matters worse, the goods in these markets are usually experience goods whose true value can only be learned after consumption.
 Furthermore, users interact with these systems routinely and watch more than one movie on Netflix, buy more than one product on Amazon, and listen to more than one video on YouTube.
@@
 %drawing from recent work in psychology and decision theory in economics, 
-we first propose a model of consumer decision-making in the context of large choice sets faced by users in online markets and show that the model is able to rationalize features of user behavior in such contexts by means of simulations.
+we propose a model of consumer decision-making in the context of large choice sets faced by users in online markets and show that the model is able to rationalize features of user behavior in such contexts by means of simulations.
 We then ask how a stylized model of recommendation can shape the behavior of long-lived users

### 2020-01-19 paper/filter_bubbles.tex
updates paper

--- 2019-12-14 paper/filter_bubbles.tex
@@
-\documentclass[sigconf]{acmart}
-\settopmatter{printacmref=false}
-\usepackage[T1]{fontenc}
-%\usepackage{lmodern}
-\usepackage{float}
-\usepackage{enumitem}
+\documentclass[format=acmsmall, review=false]{acmart}
+\usepackage{acm-ec-20}
+\usepackage{booktabs} % For formal tables
+\usepackage[ruled]{algorithm2e} % For algorithms
+
+\usepackage{subcaption}
+\newcommand{\xhdr}[1]{\vspace{1mm} \noindent{\bf #1}}
+
+\renewcommand{\algorithmcfname}{ALGORITHM}
+\SetAlFnt{\small}
+\SetAlCapFnt{\small}
+\SetAlCapNameFnt{\small}
+\SetAlCapHSkip{0pt}
+\IncMargin{-\parindent}
+
+% Choose a citation style by commenting/uncommenting the appropriate line:
+%\setcitestyle{authoryear}
+\setcitestyle{acmnumeric}
+
 \newtheorem{finding}{Finding}
-%
-% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
-\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
 
-% Rights management information. 
-% This information is sent to you when you complete the rights form.
-% These commands have SAMPLE values in them; it is your responsibility as an author to replace
-% the commands and values with those provided to you when you complete the rights form.
-%
-% These commands are for a PROCEEDINGS abstract or paper.
-\copyrightyear{2020}
-\acmYear{2020}
-\setcopyright{acmlicensed}
-\acmConference[WWW'20]{Recomme '18: ACM Symposium on Neural Gaze Detection}{April 20-24, 2020}{Taipei, Taiwan}
-\acmBooktitle{RecSys '19: ACM Conference on Neural Gaze Detection, June 03--05, 2018, Woodstock, NY}
-\acmPrice{15.00}
-\acmDOI{10.1145/1122445.1122456}
-\acmISBN{978-1-4503-9999-9/18/06}
+% Title. Note the optional short title for running heads. In the interest of anonymization, please do not include any acknowledgements.
+\newcommand\PaperTitle{Consumer Decision-Making and Recommender Systems}
+\title[\PaperTitle]{\PaperTitle}
 
-%
-% These commands are for a JOURNAL article.
-%\setcopyright{acmcopyright}
-%\acmJournal{TOG}
-%\acmYear{2018}\acmVolume{37}\acmNumber{4}\acmArticle{111}\acmMonth{8}
-%\acmDOI{10.1145/1122445.1122456}
+% Anonymized submission.
+\author{Submission 42?}
 
-%
-% Submission ID. 
-% Use this when submitting an article to a sponsored event. You'll receive a unique submission ID from the organizers
-% of the event, and this ID should be used as the parameter to this command.
-%\acmSubmissionID{123-A56-BU3}
-
-%
-% The majority of ACM publications use numbered citations and references. If you are preparing content for an event
-% sponsored by ACM SIGGRAPH, you must use the "author year" style of citations and references. Uncommenting
-% the next command will enable that style.
-%\citestyle{acmauthoryear}
-
-%
-% end of the preamble, start of the body of the document source.
-\begin{document}
-\setlength{\abovedisplayskip}{0pt}
-\setlength{\belowdisplayskip}{0pt}
-% end of the preamble, start of the body of the document source.
-
-%
-% The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
-\title{A User-Centric Approach to the Design and Consequences of Recommender Systems}
-%
-% By default, the full list of authors will be used in the page headers. Often, this list is too long, and will overlap
-% other information printed in the page headers. This command allows the author to define a more concise list
-% of authors' names for this purpose.
-
-\settopmatter{printacmref=false} % Removes citation information below abstract
-\makeatletter
-\def\-copyrightspace{\relax}
-\makeatother
-
-%
-% The abstract is a short summary of the work to be presented in the article.
+% Abstract. Note that this must come before \maketitle.
 \begin{abstract}
@@
 
+% Paper body
 \section{Introduction}
 
-Recommender Systems (RS) have become critical for assisting users in navigating the large choice sets that they face in many online markets.
+Users in many online markets face difficult choice problems as they have to sort through the vast choice sets present in these markets.
 For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
-Users in many cases are not aware of most items, let alone have full information about their preferences over them and, to make matters worse, the goods in these markets are usually experience goods whose true value can only be learned after consumption.
-Furthermore, users interact with these systems routinely and watch more than one movie on Netflix, buy more than one product on Amazon, and listen to more than one video on YouTube.
-\par
-
-Recommender systems have been influential in shaping consumer choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{MacKenzie et al.
+A common remedy by online platforms has been to deploy \textit{recommender systems}, which have become critical for assisting users in making consumption decisions by providing users with information and recommendations to allow them to make more informed consumption decisions.
+These systems have been shown to be influential in steering consumer choice with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{MacKenzie et al.
 (2013, Oct.),  How retailers can keep up with consumers.
 \url{https://www.mckinsey.com/industries/retail/our-insights/how-retailers-can-keep-up-with-consumers}. 
-Retrieved on October 3, 2019.} While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems.
-There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals,\footnote{Friedersdorf (2018, Mar.
-12), YouTube Extremism and the Long Tail, \textit{The Atlantic}.
-\url{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350}.
-Retrieved on October 3, 2019.} that personalized recommender systems lead users into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
+Retrieved on October 3, 2019.}
 \par 
+Despite evidence of the power of these systems in determining user choice, we have little theoretical understanding of how users make choices in these markets and how recommender systems shape these choices.
+Understanding this has become increasingly important as there have emerged a number of concerns regarding the consequences of such systems.
+For instance, there have been claims that YouTube's recommendation algorithm unintentionally led to the radicalization of many individuals,\footnote{Friedersdorf (2018, Mar.
+12), YouTube Extremism and the Long Tail,
+\textit{The Atlantic}. 
+\url{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350}. Retrieved on October 3, 2019.} 
+that personalized recommender systems lead users into \textit{filter bubbles} 
+where they effectively get isolated from a diversity of viewpoints or content 
+\cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
+Further, an emerging worry is that recommender systems can be utilized as a tool of market power by steering consumer choice towards products more profitable for online platforms to the detriment of consumer welfare [citations].
+Finally, there has emerged a burgeoning literature that focuses on understanding how recommender systems themselves can acquire information about product quality from the choices of self-interested consumers \cite{kremer2014implementing, che2017recommender,frazier2014incentivizing,mansour2015bayesian}.
+\par
+While these issues are important, it is hard to make progress on them without a model of user choice with foundations in economic decision theory and psychology.
+It is difficult to understand to what extent issues such as filter bubbles and homogenization are a consequence of recommender systems or a natural consequence of consumer decision-making in these contexts without such analysis.
+Further, it is challenging to understand how recommender systems can be used anti-competitively without understanding how such steering can influence user behavior.
+Finally, a major open question in the literature on incentivizing exploration in recommender systems is how to design recommendation policies in the context of realistic user behavior \cite{immorlica2018incentivizing}, which requires an understanding of what models of user decision-making are appropriate.
 
-\noindent \textbf{Our Model and Contributions} In this paper, 
-%drawing from recent work in psychology and decision theory in economics, 
-we first propose a model of consumer decision-making in the context of large choice sets faced by users in online markets and show that the model is able to rationalize features of user behavior in such contexts by means of simulations.
-We then ask how a stylized model of recommendation can shape the behavior of long-lived users
-and finally ask how the insights from our model can be used to improve recommender system design and to better understand the consequences of these systems.
+\xhdr{Our Model} In this paper we study a model of consumer decision-making in these markets via means of simulation and ask how a stylized model of recommendation can shape the behavior of users.
 \par
+The first component of our model is that users sequentially consume items and face large choice sets, though they only consume a small fraction of this choice set over their lifetime.
+\par
+The second component of our model is that users have \textit{uncertainty} about their valuation of the different items.
+This is motivated both by the fact that many markets where recommender systems are deployed contain experience goods, whose true value can only be learned after consumption, and the fact that such uncertainty is why recommender systems exist in the first place.
+Thus, users face a sequential decision-making problem under uncertainty.
+\par 
+The third, and key, component of our model is that users are Bayesian and employ similarity-based generalization to navigate the space of items.
+Unlike in standard sequential decision-making problems, in our model once an item is consumed all uncertainty about its valuation is resolved.
+This valuation provides information for consumers that enables them to update their beliefs about similar items.
+This exploits the fact that similar items have more correlated true valuations, which enables users to navigate the vast product space.
+\par 
+The fourth component of our model is that recommendation provides users with uncertainty about the true valuations.
+We model the realized valuations as being a weighted sum of a common-value and an idiosyncratic component.
+This formulation gives a stylized notion of predictability of user preferences where the idiosyncratic component is inherently unpredictable given other individuals' preferences and the common-value component is what the recommender can learn from previous users' data.
+The recommender uses its knowledge of the common-value component and the individuals' beliefs over the product space when designing personalized recommendation.
 
-The key components of our model are:
-\begin{enumerate}[label=(\arabic*)]
-\item Users face uncertainty about their valuation of the different goods and thus face a decision problem under uncertainty.
-\item Choice sets are large and users only consume a small fraction of this choice set over their lifetime.
-\item Consumption of these items is sequential and users live for multiple periods.
-\item Consuming an item provides information about the value of similar items in the product space.
-\item Recommendation provides users with information, reducing their uncertainty.
-\end{enumerate}
+\xhdr{Main Findings}
+Our first set of results focuses on understanding to what extent recommender systems lead users into filter bubbles and how such filter bubbles may naturally arise.
+The standard notion of filter bubbles is that, over time, user consumption becomes increasingly concentrated on narrower and narrower parts of the product space, leading to a decreasing distance between subsequent consumed items over time.
+However, empirical work, such as \cite{nguyen2014exploring}, finds that user consumption choices over time naturally exhibit such filter bubble effects without recommendation and that, in fact, recommender systems mitigate such effects.
+The results of our model are in line with these empirical results and give a clear intuition as to why such effects arise and why recommendation mitigates them to some degree.
+\par 
+We find that, even if there were no uncertainty about valuations, if there is no inherent correlation between the true valuations of the items then such filter bubble effects never arise, even in the optimal consumption path.
+However, once there is a correlation between the true valuations of the goods then such filter bubble effects arise even without recommendation.\footnote{The correlation between the true valuation of items in the product space is a natural assumption and, indeed, is the core idea behind standard content-based recommendation methods.
+These methods determine recommendations based on finding similar items to those that a user has consumed before and enjoyed \citep{adomavicius2005toward}.} The intuition is that, users initially have uncertainty about the set of items and when they begin consuming items this impacts the beliefs of the average utility of nearby items.
+When a user finds an item that has high realized utility, this induces them to update their beliefs positively about the realized utility of nearby items which leads to them consuming these items.
+Thus, without recommendation users increasingly consume items in this portion of the vast product space since they never get information about items in different parts of the product space.
+This learning spillover not only induces positive updates about the average utility of the item, but also reduces the level of uncertainty they have about these items.
+Thus, this effect is further amplified when users are \textit{risk-averse}, a concept from decision theory where all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty.
+\par 
+We then investigate whether the introduction of recommendation either mitigates or amplifies this effect.
+Recommendations could potentially make matters worse: providing information about the common value of the items when this value is correlated could intensify the extent of local consumption.
+However, we find the opposite - that recommendation partially mitigates this narrowing effect - since the revelation of the common value component allows for users to escape being stuck in portions of the product space due to lack of information about other portions of the product space.
+\par
+\xhdr{Additional Findings} We further explore the diversity of the overall set of items that users consume.
+Consumption diversity is increasingly a common goal of many recommender systems [references], which is reasonable since it could be expected that higher diversity is associated with more intense search and therefore higher welfare.
+However, we find that, without recommendation, higher diversity of items does not ensure higher welfare.
+This is due to the fact that high consumption set diversity in our model can arise from a sequence of bad realizations, which leads to very intense search over very different parts of the product space and high consumption diversity but very low welfare.
+However, discovering a sweet spot leads to highly rewarding local search and very low diversity.
+Thus, contrary to what a natural intuition could suggest, we find that consumption set diversity is not by itself a useful measure for understanding the impact of recommender systems on user welfare.
+\par
+We then turn to the extent to which recommendation leads to increasing user homogenization.
+We find that, without recommendation, there is very little homogenization as users are effectively all doing their own random exploration around the product space.
+In the ex-post utility maximizing consumption path (i.e.
+the consumption path the user would have taken if all uncertainty had been resolved) we find that some degree of homogenization is optimal.
+However, we find that while recommendation leads to an increase in consumer welfare, it concentrates user consumption choices in the same portions of the product space where there is a high common-value component.
+As a result, it leads to a substantially higher homogenization when compared both to the no recommendation and ex-post utility maximizing consumption path.
+\par 
+One implication from our model is that the user consumption paths have a clear path-dependence since the consumption of an item today influences the information a user has tomorrow which further influences consumption choices tomorrow.
+This is important for the debate regarding the extent to which recommender systems can be utilized as tools of market power since most of the literature analyzes this in a static context.
+However, this path-dependence implies that such platform steering can have dynamic consequences as steering towards certain items today can shape the long-run consumption choices of users.
 
-Using this model we first investigate the degree to which recommender systems lead users into filter bubbles.
-Our model rationalizes the empirical results in \cite{nguyen2014exploring} who show that in the context of movie consumption, user behavior is consistent with filter bubble effects \textit{even without recommendation}.
-The driving force of this in our model is that preferences for similar goods are correlated, which implies that when a good is consumed and the user learns its value, it provides information about similar goods.
-Crucially, this not only impacts the underlying belief about the average utility of the good, but also the level of uncertainty.
-Consequently, this learning spillover induces users to consume goods ``similar'' to those they consumed before that had high realized value, leading to an increasing narrowing of consumption patterns towards these regions of the product space.
-This effect is further amplified when users are \textit{risk-averse}, a concept from decision theory where all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty.
-We compare this benchmark to the case where recommendations are provided. 
-We consider recommendations as giving information to users in order to reduce their uncertainty.
-Recommendations could potentially make matters worse: providing information about the value of the goods when this value is correlated could intensify local search and decrease consumption diversity.
-However, and in line with the empirical results from \cite{nguyen2014exploring}, we find that recommendations reduce the narrowing effect.
-\par
-
-We then explore how diverse the overall set of items that users consume is.
-It could be expected that higher diversity being associated with more intense search is associated with higher welfare.
-Much to the opposite, we find that, without recommendation, higher diversity of items does not ensure higher welfare.
-This is due to the fact that once users find good items, they consume items nearby in the product space due to the correlation of values, leading to higher overall value from consumption.
-When they consume products they value little, they move to another part of the product space.
-A sequence of bad realizations will then lead to very intense search over very different parts of the product space and high consumption diversity, whereas discovering a sweet spot leads to highly rewarding local search and very low diversity.
-Thus, contrary to what a natural intuition could suggest, we find that consumption set diversity is not by itself a useful measure for understanding the impact of recommender systems on users.
-\par
-
-We find that recommendation can lead to increases in homogeneity among users.
-We model the realized valuations as being a weighted sum of a common-value and an idiosyncratic component.
-This formulation gives a stylized notion of predictability of user preferences where the idiosyncratic component is inherently unpredictable given other individuals' preferences and the common-value component is what the recommonder can learn from previous users' data.
-The recommender uses its knowledge of the common-value component and the individuals' beliefs over the product space when designing personalized recommendation.
-This leads to an increase in welfare for users, but concentrates their consumption around items that have a high common-value component, resulting in an increase in homogeneity.
-\par
-
-Finally, we discuss how our model and findings can be used to inform the design and evaluation of recommender systems.
-Our evaluation metric relies on the intuition that a recommendation is only useful insofar as it leads a user to take a better action than she would without the recommendation.
-Thus, in order to evaluate the value of recommendation, it is important to be able to predict what a user would do without recommendation and how recommendation would change their behavior.
-If recommendations fail to affect behavior, they are of no use: the user would consume the same good and learn its value with and without the recommendation.
-Moreover, we argue that successful recommendation systems have to take into account the degree of uncertainty users have over different portions of the product space.
-Being able to predict valuations after consumption is not sufficient as uncertainty over the idiosyncratic component of one's own preferences may steer the user away from the recommended items in the face of risk-aversion.
-\par
-
-\textbf{Related Work.} 
+\xhdr{Related Work.} 
 \cite{pariser2011filter} first informally described the ``filter bubble'' which is the idea that online personalization services would lead users down paths of increasingly narrower content so that they would effectively be isolated from a diversity of viewpoints or content.
@@
 The second question is to what extent recommender systems can lead users to become increasingly homogenized or consume similar content as each other.
-\cite{celam2008hits,treviranus2009value} show that incorporating content popularity into recommender systems can lead to increased user homogenization.
+\cite{celma2008hits, treviranus2009value} show that incorporating content popularity into recommender systems can lead to increased user homogenization.
 \cite{chaney2018algorithmic} more generally look at the issue of user homogenization in recommender systems.
@@
 \cite{hodgson2019horse} consider a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of their preferences in the environment, but consider it in the context of search for a single item.
-Our discussion of recommender system evaluation contributes with insights closely tied to the value of information that is used in the economics literature (see e.g.
-\cite{bergemann2019information}) and builds off the large literature stemming from \cite{mcnee2006being} that makes the case that accurate predictions of true realized valuations may not lead to the best recommendations.
-Our evaluation measure is most closely related to the notion of serendipity in recommender systems that has been studied extensively (e.g.
-see \cite{kotkov2016survey} for a survey).
 
 \section{Our Model and Preliminaries}
-
 \subsection{Preliminaries on Expected Utility Theory}
@@
 Higher $\gamma$ implies higher risk-aversion with $\gamma=0$ corresponding to the risk-neutral case and $\gamma>0$ to the risk-averse one.
+Our formulations here follow standard economic decision-theory (see \cite{mas1995microeconomic} for a textbook treatment of these topics).
 \par
@@
 \par
-\noindent\textbf{User Decision-Making.}
+\xhdr{User Decision-Making.}
 We assume the user makes $T$ choices and therefore can only consume up to $T$ items, where $T$ is but a small fraction of $N$.
@@
 \par
-\noindent \textbf{User Beliefs.} We assume that users do not know the realized utilities of the goods before consuming an item.
+\xhdr{User Beliefs.} We assume that all the true realized utilities are drawn at $t = 0$.
+However, users do not know the realized utilities before consuming an item, but rather have beliefs over the realized utilities.
 Formally, consumer $i$ starts with some beliefs about $X_i$, namely that the idiosyncratic and common-value parts of the valuations are independent -- $V_i \perp \!\!\!
 \perp V$ -- and that each is multivariate normal
-\begin{enumerate}[topsep=0pt]
+\begin{enumerate}
 \item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ 
@@
 $\delta_{i}(n)\mid C_i^{t-1}=\mu_n-\frac{1}{2}\gamma \Sigma_{nn}$, where $\mu_n$ and $\Sigma_{nn}$ are the expected value and variance for item $n$ that the user has given their initial beliefs and consumption history up until time $t$.
+Note that this expression is known to be the certainty equivalent for CARA preferences in a Gaussian environment \cite{mas1995microeconomic}.
 
-\noindent \textbf{User Learning}
-When a consumer consumes a product $n$ they learn the realized value for that product.
+\xhdr{User Learning}
+When a user consumes a product $n$ they learn the realized value for that product.
 In our model, we incorporate the idea that learning the value of product $n$ gives a consumer information about similar items.
@@
 Informativeness, for any $\rho \in (0,1)$, is decreasing in distance.
-The particular distance function that we utilize leads to this covariance structure being simple, where the $(n,n+1)$-th entry in the covariance matrix is $\rho$, the $(n,n+2)$-th entry is $\rho^2$, etc.
+The particular distance function that we utilize leads to this covariance structure being simple, where the $(n,n+1)$-th entry in the covariance matrix is $\sigma^{2} \rho$, the $(n,n+2)$-th entry is $\sigma^{2} \rho^2$, etc.
 \par
 \begin{figure}[t]
-\includegraphics[width=.4\linewidth]{Example-Bubbles.pdf}
+\includegraphics[width=.3\linewidth]{figures/Example-Bubbles.pdf}
 \captionof{figure}{Illustrative Example}
@@
 
-\noindent \textbf{An Illustrative Example} We illustrate our model with a simple example.
+\xhdr{An Illustrative Example} We illustrate our model with a simple example.
 Suppose that there are four products: 0, 1, 2, 3.
@@
 
-\noindent \textbf{Recommendation.}
+\xhdr{Recommendation.}
 Our model of recommendation is stylized in order to provide qualitative insights into how recommendation shapes behavior, instead of focusing on the details of how recommender systems are implemented in practice.
@@
 Thus, at any given period, the recommender provides a personalized recommendation that combines the knowledge of the common value component $V$ with the user beliefs $\bar V_i$.
-Knowing the user's beliefs about her valuation of each item become crucial in this case: just providing the user with information about $V$ may change the user's original ranking, but, without considering the user's beliefs, she will not necessarily follow the recommendation.
-\par
-
-The notion of recommendation that we consider is idealized where the recommendation does the Bayesian updating for users.
+Knowing the user's beliefs about her valuation of each item become crucial in this case: just providing the user with information about $V$ may change the user's original ranking, but, without considering the user's beliefs, she will not necessarily follow the recommendation.\footnote{The notion of recommendation that we consider is idealized where the recommendation does the Bayesian updating for users.
 As users face a very difficult and costly Bayesian updating instance, a useful recommendation system provides recommendations taking into account this aspect and combines the history of past consumption and value realizations with the user's beliefs.
@@
 but we assume the recommender provides full information about $V$.
-\footnote{For instance, the recommender could display the whole distribution of utilities reported by other users or even its average, which is a good proxy for the common value component.
+For instance, the recommender could display the whole distribution of utilities reported by other users or even its average, which is a good proxy for the common value component.
 It is possible for the recommender to only recommend the best item given such information, that is, the item with highest common value component.
 However, in that case, recommending only a single item generates costly Bayesian updating to the user and provides little guidance as to what she should indeed pick as the common value might be of little importance when compared to the idiosyncratic component.
-Therefore, we assume that the recommender reports the whole $V$, which results in higher expected welfare in choices and is not costly to implement.}
-The results would be equivalent, but this would shift the burden of Bayesian updating to users.
+Therefore, we assume that the recommender reports the whole $V$, which results in higher expected welfare in choices and is not costly to implement.
+The results would be equivalent, but this would shift the burden of Bayesian updating to users.}
 \par
@@
 This gives us a full information benchmark, which is the optimal consumption path for a user if all uncertainty about their preferences was resolved.
+Comparison to the oracle regime benchmark provides an analog to the standard regret measures utilized in the multi-armed bandit literature, which look at the difference in the expected value of the ex-post optimal action and the expected value of actions that were taken.
 \par
 
-\noindent \textbf{Simulation Details}. 
+\xhdr{Simulation Details}. 
 We analyze our model using numerical simulation since the sequential decision-making component of our model paired with the rich covariance structure between the items make it difficult to characterize optimal user behavior analytically.
@@
 
-We simulate over risk-aversion parameters $\gamma \in \{ 0, 0.3, 0.6, 1, 5 \}$, standard-deviation $\sigma \in \{ 0.25, 0.5, 1.0, 2.0, 4.0 \}$, correlation parameters $\rho\in \{ 0, 0.1, 0.3, 0.5, 0.7, 0.9 \} $ and degree of idiosyncrasy of individual valuations $\beta \in \{ 0, 0.4, 0.8, 1, 2, 5\}$.
+We simulate over risk-aversion parameters $\gamma \in \{ 0, 0.3, 0.6, 1, 5 \}$, \ standard-deviation $\sigma \in \{ 0.25, 0.5, 1.0, 2.0, 4.0 \}$, correlation parameters $\rho\in \{ 0, 0.1, 0.3, 0.5, 0.7, 0.9 \} $ and degree of idiosyncrasy of individual valuations $\beta \in \{ 0, 0.4, 0.8, 1, 2, 5\}$.
+We argue that these parameters are representative.
+First, for $\gamma$ we correspond parameter values from no risk aversion ($\gamma = 0$) to, for CARA preferences, a very high risk aversion level $(\gamma = 5)$.
+The correlation parameter, $\rho$, ranges from no correlation until almost full correlation $(\rho = 0.9)$.
+$\beta$ ranges from the common value component playing no role $(\beta = 0)$ to the idiosyncratic component dominating the true valuations $(\beta = 5)$.
+Finally, the standard deviation ranges from little uncertainty $(\sigma = 0.25)$ to a large amount of uncertainty $(\sigma = 4)$.
+Thus, the range of considered parameter values cover the relevant portions of the parameter space in order to provide full insight into the behavior of the model.
+\par
 When we consider results varying a single parameter, we group the results over the other parameters and provide reports varying only the parameter of interest.
@@
 
-\begin{figure}[t]
-\includegraphics[width=.9\linewidth]{figures/rho_pos_consumption_dist_N_200T_20_overall.pdf}\\
-\includegraphics[width=.9\linewidth]{figures/rho_zero_consumption_dist_N_200T_20_overall.pdf}\\
+\begin{figure}
+\begin{subfigure}{.5\linewidth}
+  \centering
+  \includegraphics[width=.8\linewidth]{figures/rho_pos_consumption_dist_N_200T_20_overall.pdf}
+  \caption{1a}
+  \label{fig:sfig1}
+\end{subfigure}%
+\begin{subfigure}{.5\linewidth}
+  \centering
+  \includegraphics[width=.8\linewidth]{figures/rho_zero_consumption_dist_N_200T_20_overall.pdf}
+  \caption{1b}
+  \label{fig:sfig2}
+\end{subfigure}
 \caption{Consumption path with (Top) and without (Bottom) correlation between values}
@@
 
-\noindent \textbf{Uncertainty}: We assume that users have underlying uncertainty about the realized value of the items in their choice sets and as a result face a decision problem under uncertainty.
+\xhdr{Uncertainty}: We assume that users have underlying uncertainty about the realized value of the items in their choice sets and as a result face a decision problem under uncertainty.
 The first justification for this is that environments where recommender systems are commonly deployed, such as on media streaming and e-commerce platforms, are mostly dedicated to experience goods, where users only learn their value after consumption.
 Standard decision theory axiomatized the conditions under which behavior is described by (subjective) expected utility maximizers \cite{savage1954} who update beliefs according to Bayesian inference \cite{ghirardato2002revisiting}, to which we add the assumption that their prior is correct.
-The second justification is that, even if users are in an environment where they could observe utilities based on product reviews or introspection, it is prohibitively costly for them to do so in large choice sets.\footnote{Recent empirical work \cite{bronnenberg2014zooming,hodgson2019horse} has shown that, for product search, users engage in a ``zooming in'' search strategy in terms of the information they acquire about products before purchasing and only end up acquiring information about a small set of products in a large product space}  Even if they acquired all the information from product reviews, this would not fully resolve uncertainty as to the extent that individuals have idiosyncratic tastes.
+The second justification is that, even if users are in an environment where they could observe utilities based on product reviews or introspection, it is prohibitively costly for them to do so in large choice sets.\footnote{Recent empirical work \cite{bronnenberg2016zooming,hodgson2019horse} has shown that, for product search, users engage in a ``zooming in'' search strategy in terms of the information they acquire about products before purchasing and only end up acquiring information about a small set of products in a large product space}  Even if they acquired all the information from product reviews, this would not fully resolve uncertainty as to the extent that individuals have idiosyncratic tastes.
 Giving that users are facing choices problems uncertainty, this motivates modeling recommendation as providing users \textit{information} in order to reduce their uncertainty.
@@
 
-\noindent \textbf{Large Choice Sets}: We are interested in environments where the number of items in the choice set is large and users only consume a small number of these items.
+\xhdr{Large Choice Sets}: We are interested in environments where the number of items in the choice set is large and users only consume a small number of these items.
 This is realistic for platforms such as Netflix, Spotify, Amazon and YouTube where there are sometimes billions of items and users can realistically only consume a handful of these items in their lifetime.\par
 
-\noindent \textbf{Learning Spillovers}: Users in our environment face a sequential decision-making problem.
+\xhdr{Learning Spillovers}: Users in our environment face a sequential decision-making problem.
 However, we suppose that individuals are myopic, i.e.
@@
 \section{Results}
-\subsection{Local Consumption}
+\subsection{Local Consumption and Filter Bubbles}
 We characterize ``filter bubble'' effects in our model as the degree to which users engage in \textit{local consumption}.
 We define local consumption in terms of the average consumption distance between the items consumed by the users at time $t-1$ and $t$.
-We compare the average movement within the product space across different recommendation regimes and associate lower average distance with more local consumption.
+Thus, in the context of our model, filter bubble effects arise when the average consumption distance decreases over time and, across regimes, when the overall levels are lower for a given recommendation regime compared to another.
 
@@
 However this also means that users do not learn anything about neighboring products and so there is limited path-dependence in consumption.
+Not only is there no difference in the levels between the three regimes, but they all have the same, flat, average consumption distance path.
+This underscores the fact that such concerns about filter bubbles \textit{only} arise in a world where there is correlation between the realized utility of items.
+If there were no correlation between the realized utilities then there would be no reason for users to consume similar goods and thus no narrowing effect, regardless of the information on the true utility of the items that users had.
 \par
-
 The top panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho > 0$, both recommendation and no recommendation lead to increasingly local consumption compared to the oracle benchmark case.
@@
 \begin{figure}[t]
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20.pdf}\\
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_partial.pdf}\\
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20.pdf}
+\end{subfigure}
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_partial.pdf}
+\end{subfigure}\\
+\begin{subfigure}{.45\textwidth}
 \includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_omni.pdf}\\
 \caption{Extent of Local Consumption for No Recommendation (Top), Recommendation (Middle) and Oracle (Bottom) for different Correlation values}
+\end{subfigure}%
 \label{fig:local_consumption_across_rho}
@@
 \begin{figure}[t]
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20.pdf}\\
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_partial.pdf}\\
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20.pdf}
+\end{subfigure}
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_partial.pdf}
+\end{subfigure}\\
+\begin{subfigure}{.45\textwidth}
 \includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_omni.pdf}\\
+\end{subfigure}%
 \caption{Effect of Risk Aversion on Local Consumption for No Recommendation (Top), Recommendation (Middle) and Oracle (Bottom)}
@@
 
-In sum, in our model filter bubbles are driven by learning-spillovers and risk aversion.
-It is the belief that there is an underlying correlation across different products' values, stronger for similar products, closer in the product space, that leads users to focus on local consumption upon consuming high valued items.
-For instance, upon discovering that the user likes movies by a given director, she will be more likely to explore other movies by the same director.
-Risk aversion may lead users into performing local consumption even when they have a low valuation of nearby products just because they know what to expect from the product.
-\par
+In sum, filter bubble effects can only arise when there is an inherent correlation between the realized utilities of the items in the product space.
+However, this is often the case in most environments where recommender systems are deployed.
+When there is a correlation between the realized utilities, then filter bubbles can naturally arise as a result of the information that users possess on the utilities of the items.
+We have shown that, absent any additional information, learning spillovers and risk aversion can lead users into filter bubbles.
+When users consume high valued items, they exploit the underlying correlation across different products' values, stronger for similar products, which leads them to increasingly consume products in narrower and narrower portions of the product space.
+Risk aversion may lead users into performing local consumption even when they have a low valuation of nearby items just because they know what to expect from the item.
+Recommendation leads to these effects being mitigated by providing users with additional information on products outside the already explored portions of the product space.
 
@@
 In this section we primarily focus on the impact of recommendation on the welfare on users as well as the overall diversity of the items that they consume.
-We define user's \textit{ex-post} welfare as the average of the realized values, controlling for the effect of $T$
-$$W_i:= \frac{1}{T}\sum_{n \in C_i^T} x_{i,n}$$
 
-While in the previous section we looked at the distance between consecutive goods, we also define a diversity measure that is defined over the entire consumed set.
-We utilize a diversity metric common in the literature (e.g.
+While in the previous section we looked at the distance between consecutive goods, in this section we focus on a diversity measure that considers the entire consumed set of items.
+The diversity measure we utilize is common in recommender system literature (e.g.
 \cite{ziegler2005improving}) which is the average normalized pairwise distance between the consumed products:
@@
 \end{finding}
-
+\par 
+Finding~\ref{finding_diversity} summarizes the main results on product diversity.
 As before, when there is no correlation between valuations, product diversity is the same across different recommendation regimes.
-The over-exploitation of the learning spillovers when $\rho > 0$ leads to product diversity being lowest in the no recommendation regime.
-Figure \ref{fig:risk_aversion_diversity} shows how diversity varies as risk aversion levels change and Figure \ref{fig:diversity_correlation} shows how diversity varies across recommendation regimes and the level of $\rho$.
+The over-exploitation of learning spillovers when $\rho > 0$ leads to product diversity being lowest in the no recommendation regime.
+As a result, this effect gets amplified as $\rho$ increases, which leads to the gap in diversity between the regimes to increase as $\rho$ increases.
+Figure \ref{fig:diversity_correlation} shows how diversity varies as $\rho$ increases across the three regimes that we consider.
+There is a similar increasing diversity gap as $\gamma$, or the level of risk-aversion, increases as can be seen in Figure \ref{fig:risk_aversion_diversity}.
+The mechanisms behind these effects have direct parallels to those discussed in the previous section since low average sequential consumption distance directly leads to low diversity.
+\par 
+We now study how recommendation impacts consumer welfare in our model.
+In our model users make consumption decisions that maximize their current period ex-ante utility that depends on their beliefs in that period.
+Thus, from an ex-ante perspective, they make optimal decisions, but our primary interest is in understanding how the ex-post, or realized, utility varies across regimes and parameter values.
+We define user's \textit{ex-post} welfare as the average of the realized values, controlling for the effect of $T$:
+$$W_i:= \frac{1}{T}\sum_{n \in C_i^T} x_{i,n}$$
 
 \begin{figure}[ht]
-\includegraphics[width=.9\linewidth]{figures/gamma_diversity_N_200_T_20}
+\includegraphics[width=.4\linewidth]{figures/gamma_diversity_N_200_T_20}
 \captionof{figure}{Risk Aversion and Diversity}\label{fig:risk_aversion_diversity}
@@
 \begin{finding}\label{finding_welfare_gap}
-The welfare gap between the no recommendation and the recommendation regimes and that between the no recommendation and the oracle regimes are decreasing in $\rho$.
+The impact of recommendation on consumer welfare is as follows:
+\begin{enumerate}
+\item Under oracle recommendation, welfare is constant as $\rho$ increases.
+\item Under no recommendation, welfare is increasing as $\rho$ increases.
+\item Recommendation introduces welfare gains, but these gains are decreasing as $\rho$ increases. 
+\end{enumerate}
 \end{finding}
-However, interestingly, the decrease in diversity is \textit{not} associated with a decline in welfare.
-In fact, welfare stays roughly the same in the oracle and recommendation cases.
-The welfare gap between the no recommendation case and the other two cases \textit{decreases} as the diversity gap increases.
 \par 
-\begin{figure}[t]
-\includegraphics[width=.9\linewidth]{figures/rho_diversity_N_200_T_20.pdf}\\
-\includegraphics[width=.9\linewidth]{figures/rho_welfare_N_200_T_20.pdf}\\
+Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which are drawn from Figure~\ref{fig:diversity_correlation}.
+The most interesting observation is that the value of recommendation decreases as $\rho$ decreases.
+The intuition for this is clear as recommendation provides users with information that allows them to better guide their decisions and increase welfare.
+However, as $\rho$ increases users get increasingly more information simply from consuming items since the realized utility is now more informative about the utility of nearby goods.
+Thus, since consumption decisions themselves reveal information, the information that recommendation reveals provides less  value.
+\par 
+One striking observation is that the decrease in diversity does not appear to be associated with a decline in welfare.
+Indeed, it appears that the opposite is the case - that low diversity is associated with higher welfare and vice versa.
+Thus, we next explore the relationship between welfare and diversity.
+\begin{figure}
+\begin{subfigure}{.45\linewidth}
+  \includegraphics[width=.8\linewidth]{figures/rho_diversity_N_200_T_20.pdf}
+\end{subfigure}
+\begin{subfigure}{.45\linewidth}
+  \includegraphics[width=.8\linewidth]{figures/rho_welfare_N_200_T_20.pdf}
+\end{subfigure}
 \captionof{figure}{Welfare, Diversity, and Correlation}\label{fig:diversity_correlation}
@@
 
-Figure \ref{fig:diversity_welfare_ra} shows how diversity and welfare correlate for the no recommendation case.
+Finding~\ref{finding_diversity_welfare_corr} summarizes our findings on the relationship between diversity and welfare.
+Figure \ref{fig:diversity_welfare_ra} shows how diversity and welfare correlate for the no recommendation case as we vary the degree of risk aversion.
 When there is no risk-aversion then there is a negative correlation between welfare and diversity.
@@
 \par
-
-%\begin{figure}
-
-%\end{figure}
-
 This only happens since $\gamma = 0$ leads to users only caring about the expected value of the product.
@@
 In the recommendation and oracle regimes, under risk neutrality ($\gamma=0$), welfare and diversity and uncorrelated, while under risk aversion ($\gamma=5$), it is possible to observe in Figure \ref{fig:diversity_welfare_ra_partial} an actual positive relation between diversity and welfare as recommendations are able to reduce uncertainty and facilitate exploration of the product space.
-\par
 \begin{figure}[t]
-\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_rn.pdf}\\
-\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_ra.pdf}\\
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_rn.pdf}
+\end{subfigure}
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_ra.pdf}
+\end{subfigure}\\
 \captionof{figure}{Diversity vs. Welfare, No-Recommendation, $\gamma = 0$ (Top), Diversity vs. Welfare, $\gamma = 5$ (Bottom)}\label{fig:diversity_welfare_ra}
 \end{figure}
+
 \begin{figure}[t]
-\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_rn_partial.pdf}\\
-\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_ra_partial.pdf}\\
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_rn_partial.pdf}
+\end{subfigure}
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_ra_partial.pdf}
+\end{subfigure}\\
 \captionof{figure}{Diversity vs. Welfare, Recommendation, $\gamma = 0$ (Top), Diversity vs. Welfare, $\gamma = 5$ (Bottom)}\label{fig:diversity_welfare_ra_partial}
@@
 In particular we look at the degree of \textit{homogenization} between users.
-Similar to \cite{chaney2018algorithmic} we use the Jaccard index between the consumption sets of users to measure homogeneity:
+Similar to other papers that study the degree of homogenization in recommender systems (e.g.
+\cite{chaney2018algorithmic}) we measure homogeneity via the Jaccard index between the consumption sets of users:
 \begin{align*}
@@
 \begin{finding}\label{finding_homogeneity}
-Homogeneity is:
+The impact of recommendation on homogeneity is as follows:
 \begin{enumerate}
@@
 \end{finding}
-Figure \ref{fig:beta_homo} shows that as the weight of the common-value component, $\beta$, increases users consume increasingly similar sets of items.
+Finding~\ref{finding_homogeneity} summarizes our findings on the impact of recommendation on user homogeneity.
+First, we study how the degree of homogenization varies as we increase $\beta$, the weight of the common value component.
+As $\beta$ increases we expect that users should become increasingly homogeneous as the realized utilities of the items are now increasingly similar.
+Figure \ref{fig:beta_homo} confirms that as the weight of the common-value component, $\beta$, increases users consume increasingly similar sets of items.
 The homogenization effect is strongest under the recommendation regime since the revelation of the common-value component induces users to consume products in similar areas of the product space.
-However, as $\beta$ increases, some amount of homogenization is optimal as can be seen from the oracle case.
-In the no recommendation regime since users do not know the common-value component they engage in local consumption in different areas of the product space which leads to less than optimal homogeneity.
+As $\beta$ increases, some amount of homogenization is optimal as can be seen from the oracle case.
+However, since users in the no recommendation regime do not know the common-value component they engage in local consumption in different areas of the product space which leads to less than optimal homogeneity.
 \par
-
-Figure \ref{fig:cor_homo} shows that the degree of homogeneity in the recommendation case however is \textit{decreasing} as $\rho$ increases.
+We next study how the degree of homogeneity varies as $\rho$ increases.
+Figure \ref{fig:cor_homo} shows how homogeneity varies as $\rho$ increases across the different recommendation regimes.
+The most striking finding is that homogeneity decreases as $\rho$ increases in the recommendation regime.
 As was highlighted in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, the degree of local consumption increases with $\rho$.
@@
 \begin{figure}[t]
-\includegraphics[width=.8\linewidth]{figures/beta_homogeneity_N_200_T_20}
+\includegraphics[width=.45\linewidth]{figures/beta_homogeneity_N_200_T_20}
 \captionof{figure}{Strength of Recommendation and Homogeneity}\label{fig:beta_homo}
@@
 \begin{figure}[t]
-\includegraphics[width=.8\linewidth]{figures/rho_homogeneity_N_200_T_20}
+\includegraphics[width=.45\linewidth]{figures/rho_homogeneity_N_200_T_20}
 \captionof{figure}{Correlation and Homogeneity}\label{fig:cor_homo}
 \end{figure}
-
-\section{Recommender System Evaluation}
-In this section we discuss how the insights from our model of user decision-making can be used to inform the evaluation and design of recommender systems.
-\par
-
-The classic approach to recommender system evaluation is to predict user ratings for items and compare how accurate this prediction is to recorded ratings data, either explicitly given by users or inferred from behavioral data.
-Further, given this predicted rating, the recommender system should recommend to users the items with the highest predicted rating \cite{adomavicius2005toward}.
-\par
-
-There has been a recent movement away from such evaluation measures due to the observation that accurate recommendations are not necessarily useful recommendations \cite{mcnee2006being}.
-Our model illustrates why this could be the case.
-Consider the domain of movie recommendation and suppose a user has just watched the movie \textit{John Wick} and rated it highly.
-A recommender system attempting to predict user ratings may then predict that this user is very likely to enjoy the sequel, \textit{John Wick: Chapter Two}, as well.
-However, the user herself may also have made this inference since the two movies are very similar to each other and, after enjoying \textit{John Wick}, the user will update her beliefs about \textit{John Wick: Chapter Two}.
-Thus, if the recommender system recommends this movie to the user then this recommendation is not useful since the recommendation gives the user little information that they did not already know.
-The key insight is that the recommendation is not useful since \textit{it ignores the inference the user themselves made and their updated beliefs}.
-The user may watch \textit{John Wick: Chapter Two}, then, even without recommendation and the value of the recommendation was small.
-\par
-
-As a result, several alternative recommendation evaluation metrics, such as serendipity \cite{kotkov2016survey}, coverage \cite{ge2010beyond}, novelty \cite{vargas2011rank}, and many others have been proposed in order to develop recommender systems that produce recommendations that are more useful for users.
-We use our model to motivate an alternative recommendation evaluation metric that provides an alternative perspective on the serendipity metric proposed in the literature.
-Serendipitous recommendations are said to ``have the quality of being both unexpected and useful" \cite{maksai2015predicting} and we use this as a guide towards our definition.
-
-Our approach to serendipity comes from the intuition that a recommendation is only useful insofar as it leads a user to take a better action than she would without recommendation.
-This approach stems from the idea that the value of recommendation is the marginal utility gain that a user receives from recommendation.\footnote{This notion is similar to the ``value of information" notion utilized in the literature in economics on the pricing of information goods \cite{bergemann2018design}.} This intuitively should come from providing information about products that, without additional information, users assign a low likelihood to being the best products currently available for them to consume and so would not consume them without these being recommended.
-\par
-
-This differs from other approaches in the literature since it depends on the beliefs of the users and what item the user would consume without recommendation.
-For instance, \cite{vargas2011rank, kaminskas2014measuring} propose unexpectedness metrics that look at the dissimilarity of the recommended items to what the recommender already knows the user likes either via a content-based approach or a collaborative-based approach.
-This depends only on the resulting item-set that the users get recommended and not necessarily user beliefs or the change in user action.
-\cite{adamopoulos2014unexpectedness} use an expected utility approach but use this to measure unexpectedness in terms of deviations from the ``expected consideration set" of a user and do not explicitly incorporate user beliefs.
-\par
-
-Our model allows us to give precise definitions for what it means for a recommendation to be \textit{unexpected} and \textit{useful}.
-Unexpected recommendations are given by recommended items that, given their beliefs, users have low expected utility from consuming, irregardless of what the true realized value is.
-Recall from Findings \ref{finding_local_consumption} and \ref{finding_diversity_welfare_corr} that, particularly if users are risk-averse, they will consume content in increasingly narrow portions of the product space and this may be detrimental to their realized welfare.
-As a result, identifying the portions of the product space where users may have significant uncertainty and utilizing recommendation to reduce this uncertainty can be welfare-enhancing for users.
-This is especially the case if this reduction of uncertainty leads users to explore under-explored portions of the product space where true realized utility is high, but expected utility, given beliefs, is low before recommendation.
-\par
-
-Useful recommendations are those that give users information that would change the item they consume relative to no recommendation and lead to an increase in expected utility given the updated beliefs when comparing the planned choice before and after the recommendation.
-In the previous example, recommending \textit{John Wick: Chapter Two} is not useful precisely because the expected utility before and after recommendation is the same since it does not change behavior.
-The most useful recommendations are those that (1) are informative given user beliefs and (2) induce users to consume unexpected goods.
-\par
-
-Operationalizing this approach requires the collection of data that is not traditionally collected for recommender systems.
-In particular, it becomes important to understand what choices users would make \textit{without recommendation} and to collect data on user beliefs, which can partially be inferred by observed choices.
-\cite{jiang2014choice, saavedra2016choice} argue that choice-based approaches alone can help in designing better RS, though they argue for these approaches because they do a better job at providing more accurate recommendations.
-Instead, we stress that choice-based approaches are useful not as recommendations but as a baseline for what recommendations would be useful.\footnote{Additionally, user beliefs and user choices contain information that may not have been observed by the recommender but is observed by the user.
-Ratings, reviews, friends, there are many sources of information affecting a person's beliefs and choices that are unobservable by recommender systems.
-Other consumption choices, e.g.
-movies seen in the cinema, change beliefs about for instance how good a director is, which affect beliefs about other movies and guide choices on a streaming platform that is unable to observe these data.
-However, this information can be inferred by the platform by collecting both beliefs and consumption choices.}
-\par
-
-Furthermore, our model shows the importance of understanding the underlying user preferences and the nature of the product space in the setting in which a recommender system is deployed since that will change the consumption patterns of users.
-For instance, understanding the degree of correlation of utilities and the level of risk-aversion become crucial in understanding how users will make consumption choices without recommendation and what kind of recommendation would be useful for users.
-\par
 
@@
 
-\bibliographystyle{plain}
+% Start of "Sample References" section
+% In the interest of anonymization, please do not include acknowledgements in your submission.
+%
+%\begin{acks}
+%
+%	The authors would like to thank Dr. Maura Turolla of Telecom
+%	Italia for providing specifications about the application scenario.
+%
+%	The work is supported by the \grantsponsor{GS501100001809}{National
+%		Natural Science Foundation of
+%		China}{http://dx.doi.org/10.13039/501100001809} under Grant
+%	No.:~\grantnum{GS501100001809}{61273304\_a}
+%	and~\grantnum[http://www.nnsf.cn/youngscientsts]{GS501100001809}{Young
+%		Scientsts' Support Program}.
+%
+%
+%\end{acks}
+
+% Bibliography
+\bibliographystyle{ACM-Reference-Format}
 \bibliography{refs}
+
 \end{document}

### 2020-01-19 paper/filter_bubbles.tex
udpates to paper

@@
 % Title. Note the optional short title for running heads. In the interest of anonymization, please do not include any acknowledgements.
-\newcommand\PaperTitle{Consumer Decision-Making and Recommender Systems}
+\newcommand\PaperTitle{Deconstructing the Filter Bubble: \\ Consumer Decision-Making and Recommender Systems}
 \title[\PaperTitle]{\PaperTitle}
@@
 We explore the extent to which two issues traditionally associated with recommender systems -- \textit{filter bubbles} and \textit{user homogenization} -- can be attributed to recommendation systems.
-The key component of our model is that users face a decision problem under uncertainty and that there are informational spillovers where a user consuming a product and learning how she values it reduces uncertainty about similar products.
+The key component of our model is that users face a sequential decision problem under uncertainty and that there are informational spillovers where a user consuming a product and learning how she values it reduces uncertainty and provides information about similar products.
 We find that this generates behavior consistent with filter bubble effects -- where users consume increasingly similar items over time -- and that recommendation can help alleviate these effects, as has been found in the empirical work of Nguyen et al.
@@
 However, we find that recommendation, while alleviating filter bubble effects, leads to an increase in user homogeneity.
-\par
-We discuss the evaluation of recommender systems in light of these findings.
-Traditional design approaches to recommender systems only optimize for predicting user preferences after uncertainty has been resolved and ignore user beliefs and the inferences that users themselves make after consuming items.
-We propose an alternative interpretation of the serendipity measure where good recommendations are those that provide useful information, that is, information without which choice would be different and lead to worse outcomes on average relative to the case where no recommendation is provided.
-As a result, we emphasize the importance of understanding user behavior without recommendation, both for designing recommendations and understanding their impacts and usefulness.
+Our results highlight the path-dependence of user choice in environments where recommender systems are deployed.
 \end{abstract}
@@
 
-Users in many online markets face difficult choice problems as they have to sort through the vast choice sets present in these markets.
+Users in many online markets face difficult decision problems as they have to sort through the vast choice sets present in these markets.
 For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
@@
 Despite evidence of the power of these systems in determining user choice, we have little theoretical understanding of how users make choices in these markets and how recommender systems shape these choices.
-Understanding this has become increasingly important as there have emerged a number of concerns regarding the consequences of such systems.
+Understanding this has become increasingly important as there have emerged a number of concerns regarding the broader social and economic consequences of such systems.
 For instance, there have been claims that YouTube's recommendation algorithm unintentionally led to the radicalization of many individuals,\footnote{Friedersdorf (2018, Mar.
@@
 \par 
-The third, and key, component of our model is that users are Bayesian and employ similarity-based generalization to navigate the space of items.
+The third, and key, component of our model is that consumption of an item reveals information that changes user beliefs about the valuation of similar items.
 Unlike in standard sequential decision-making problems, in our model once an item is consumed all uncertainty about its valuation is resolved.
 This valuation provides information for consumers that enables them to update their beliefs about similar items.
-This exploits the fact that similar items have more correlated true valuations, which enables users to navigate the vast product space.
+This exploits the fact that similar items have more correlated true valuations, which assists users in making decisions in a vast product space.
 \par 
@@
 The standard notion of filter bubbles is that, over time, user consumption becomes increasingly concentrated on narrower and narrower parts of the product space, leading to a decreasing distance between subsequent consumed items over time.
-However, empirical work, such as \cite{nguyen2014exploring}, finds that user consumption choices over time naturally exhibit such filter bubble effects without recommendation and that, in fact, recommender systems mitigate such effects.
+However, empirical work, such as \cite{nguyen2014exploring}, finds that user consumption choices over time naturally exhibit such filter bubble effects even without recommendation and that, in fact, recommender systems mitigate such effects.
 The results of our model are in line with these empirical results and give a clear intuition as to why such effects arise and why recommendation mitigates them to some degree.
@@
 We find that, even if there were no uncertainty about valuations, if there is no inherent correlation between the true valuations of the items then such filter bubble effects never arise, even in the optimal consumption path.
-However, once there is a correlation between the true valuations of the goods then such filter bubble effects arise even without recommendation.\footnote{The correlation between the true valuation of items in the product space is a natural assumption and, indeed, is the core idea behind standard content-based recommendation methods.
-These methods determine recommendations based on finding similar items to those that a user has consumed before and enjoyed \citep{adomavicius2005toward}.} The intuition is that, users initially have uncertainty about the set of items and when they begin consuming items this impacts the beliefs of the average utility of nearby items.
+However, once there is a correlation between the true valuations of the items then such filter bubble effects arise even without recommendation.\footnote{The correlation between the true valuation of items in the product space is a natural assumption and, indeed, is the core idea behind standard content-based recommendation methods.
+These methods determine recommendations based on finding similar items to those that a user has consumed before and enjoyed \citep{adomavicius2005toward}.} The intuition is that, users initially have uncertainty about the true valuations of the set of items and when they begin consuming items this impacts the beliefs of the average utility of nearby items.
 When a user finds an item that has high realized utility, this induces them to update their beliefs positively about the realized utility of nearby items which leads to them consuming these items.
@@
 This learning spillover not only induces positive updates about the average utility of the item, but also reduces the level of uncertainty they have about these items.
-Thus, this effect is further amplified when users are \textit{risk-averse}, a concept from decision theory where all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty.
+Thus, this effect is further amplified when users are \textit{risk-averse}, a concept from decision theory where all else being equal, users have a preference for items with lower uncertainty to those with higher uncertainty.
 \par 
@@
 \cite{chaney2018algorithmic} more generally look at the issue of user homogenization in recommender systems.
-Our results are similar to previous work, where the limits of recommendation lead the recommender to over-recommend products from certain portions of the product space which leads users to consume goods in those regions.
+Our results are similar to previous work, where the limits of recommendation lead the recommender to over-recommend products from certain portions of the product space which leads users to consume items in those regions.
 \cite{fleder2009blockbuster} also consider the effects of recommendation on multi-period user behavior, but predominantly focus on sales diversity and employ a different model of user behavior than we do here.
@@
 \subsection{Preliminaries on Expected Utility Theory}
-\noindent For every product $n$ in the product space $\mathcal J=\{0,1,....,N-1\}$, we assume that each consumer $i$ assigns a monetary equivalent $x_{i,n} \in \mathbb R$ to the experience of consuming it.
-Each consumer can value the same product in differently.
+\noindent For every item $n$ in the product space $\mathcal J=\{0,1,....,N-1\}$, we assume that each user $i$ assigns a monetary equivalent $x_{i,n} \in \mathbb R$ to the experience of consuming it.
+Each consumer can value the same item differently.
 However, we assume that users have the same utility over money, given by a utility function $u: \mathbb R \to \mathbb R$, strictly increasing and continuous.
-So, ex-post, the value of product $n$ for consumer $i$ is given by $u(x_{i,n})$ However, before consuming the product, the consumer does not know exactly how she will value it.
-In particular, even users that will end up having the same ex-post valuation of product $n$ may differ in their ex-ante valuation because they hold different beliefs about it.
-We denote by $p_{i}$ the beliefs consumer $i$ has about how she will value each of the products in the product space.
-Note that this implies that consuming product $n$ is the same as taking a gamble.
-Each individual evaluates the product according to the associated expected utility associated with the product, i.e.
+So, ex-post, the value of item $n$ for user $i$ is given by $u(x_{i,n})$ However, before consuming the item, the user does not know exactly how she will value it.
+In particular, even users that will end up having the same ex-post valuation of item $n$ may differ in their ex-ante valuation because they hold different beliefs about it.
+We denote by $p_{i}$ the beliefs item $i$ has about how she will value each of the items in the product space.
+Note that this implies that consuming item $n$ is the same as taking a gamble.
+Each user evaluates the item according to the associated expected utility associated with the item, i.e.
 $U_i(n)=\mathbb E_{p_i}[u(x_n)]$.
@@
 
-Risk aversion captures the how different individuals react to the risk associated to a particular consumption opportunity.
-It is formalized as follows: Suppose for a given gamble $x$ taking real values and following distribution $p$, there is a certain amount of money that leaves the individual indifferent between taking the gamble and taking this sure amount, call it $\delta(x)$, termed the \textit{certainty equivalent} of $x$.
-An individual $i$ is more risk-averse than another individual $j$ when for every gamble $x$, individual $i$ requires a lower value to give up the gamble when compared to individual $j$, i.e.
+Risk aversion captures the how different users react to the risk associated to a particular consumption opportunity.
+It is formalized as follows: Suppose for a given gamble $x$ taking real values and following distribution $p$, there is a certain amount of money that leaves the user indifferent between taking the gamble and taking this sure amount, call it $\delta(x)$, termed the \textit{certainty equivalent} of $x$.
+A user $i$ is more risk-averse than another user$j$ when for every gamble $x$, user $i$ requires a lower value to give up the gamble when compared to user $j$, i.e.
 $\delta_i(x)$.
-Therefore, a more risk-averse individual is more willing to avoid the risk of taking the gamble.
+Therefore, a more risk-averse user is more willing to avoid the risk of taking the gamble.
 We assume that the utility function takes a flexible functional form $u(x)=1-\exp(-\gamma x)$ for $\gamma\ne0$ and $u(x)=x$ for $\gamma=0$ -- known as constant absolute risk-aversion preferences.
@@
 \par
-\noindent \textbf{Users.} We consider a set of users $I$ where each individual $i \in I$ faces the same finite set of $N$ items $\mathcal J = \left\{0,1,...,N-1\right\}$.
-For simplicity, we assume that individuals only derive pleasure from item $n \in \mathcal{J}$ the first time they consume it.
+\noindent \textbf{Users.} We consider a set of users $I$ where each user $i \in I$ faces the same finite set of $N$ items $\mathcal J = \left\{0,1,...,N-1\right\}$.
+For simplicity, we assume that users only derive pleasure from item $n \in \mathcal{J}$ the first time they consume it.
 \par
 
-We denote by $x_{i,n}$ individual $i$'s realized value from consuming item $n$.
+We denote by $x_{i,n}$ user $i$'s realized value from consuming item $n$.
 In particular, we consider that the realized value derived from a given item can be decomposed in the following manner: $x_{i,n}= v_{i,n} + \beta v_n$, where $v_{i,n}$ denotes an idiosyncratic component -- i.e.
 user $i$'s idiosyncratic taste for product $n$ --  and $v_{n}$, a common-value component.
-One can interpret $v_n$ as a measure of how much product $n$ is valued in society in general and, in a sense, $v_{i,n}$ denotes how $i$ diverges from this overall ranking.
-The scalar $\beta \in \mathbb{R}_{t}$ denotes the degree to which valuations are idiosyncratic to each individual or common across individuals.
+One can interpret $v_n$ as a measure of how much item $n$ is valued in society in general and, in a sense, $v_{i,n}$ denotes how $i$ diverges from this overall ranking.
+The scalar $\beta \in \mathbb{R}_{t}$ denotes the degree to which valuations are idiosyncratic to each user or common across users.
 If $\beta=0$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta$ is large, every individual has similar preferences.
@@
 However, users do not know the realized utilities before consuming an item, but rather have beliefs over the realized utilities.
-Formally, consumer $i$ starts with some beliefs about $X_i$, namely that the idiosyncratic and common-value parts of the valuations are independent -- $V_i \perp \!\!\!
+Formally, user $i$ starts with some beliefs about $X_i$, namely that the idiosyncratic and common-value parts of the valuations are independent -- $V_i \perp \!\!\!
 \perp V$ -- and that each is multivariate normal
@@
 We impose the normality assumption for two reasons.
-The first is that users update their beliefs using Bayesian updating and recall that the normal distribution forms a conjugate family, which allows for simple posterior updates.
+The first is that this allows for simple and tractable belief updating.
 The second is that it allows us to incorporate an easily interpretable correlation structure between the items.
@@
 Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{i,n}\right)_{n \in \mathcal{J}}$ are drawn independently from a jointly normal distribution, where $\overline v_{i,n} \sim \mathcal N (0, \overline \sigma^2)$ are independent and identically distributed.
-These $\overline v_{i,n}$ denote the prior belief that individual $i$ holds about the her valuation over good $n$.
+These $\overline v_{i,n}$ denote the prior belief that individual $i$ holds about the her valuation over item $n$.
 As people are exposed to different backgrounds, their beliefs about what is good for them also varies and $\overline v_{i,n}$ denotes this idiosyncrasy at the level of prior beliefs.
@@
 We assume users are expected utility maximizers.
-User $i$'s certainty equivalent for product $n$, the sure value that makes user $i$ indifferent between it and consuming the product $n$, conditional on the consumption history, is given by
+User $i$'s certainty equivalent for item $n$, the sure value that makes user $i$ indifferent between it and consuming the item $n$, conditional on the consumption history, is given by
 $\delta_{i}(n)\mid C_i^{t-1}=\mu_n-\frac{1}{2}\gamma \Sigma_{nn}$, where $\mu_n$ and $\Sigma_{nn}$ are the expected value and variance for item $n$ that the user has given their initial beliefs and consumption history up until time $t$.
@@
 \xhdr{User Learning}
-When a user consumes a product $n$ they learn the realized value for that product.
-In our model, we incorporate the idea that learning the value of product $n$ gives a consumer information about similar items.
+When a user consumes a item $n$ they learn the realized value for that item. In our model, we incorporate the idea that learning the value of item $n$ gives a user information about similar items. 
 %This is drawn from recent empirical evidence in \cite{schulz2019structured} that consumers learn how to navigate large choice sets using similarity-based generalization. 
-We assume that learning about the value of good $n$ reveals more about the value associated to items that are closer to it, which captures the idea that trying a product provides more information about similar products than about dissimilar ones.
+We assume that learning about the value of item $n$ reveals more about the value associated to items that are closer to it, which captures the idea that trying a item provides more information about similar items than about dissimilar ones.
 \par
-
-In order to have a well-defined notion of similarity we need to define a distance function between goods, which we define as $d(n,m):=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}$ where $m$ and $n$ are indices of items in $\mathcal{J}$.
+In order to have a well-defined notion of similarity we need to define a distance function between items, which we define as $d(n,m):=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}$ where $m$ and $n$ are indices of items in $\mathcal{J}$.
 This distance function is not intended to model the intricacies of a realistic product space, but instead to provide a stylized product space to help us understand the effects of informational spillovers on user behavior.
@@
 \par
+The precise updating rule is as follows.
+Recall that at time $t$ the user's consumption history is given by $C_{i}^{t}$ and we denote the utility realizations of these items as $c_t$.
+We denote $\mu_t$ as the initial mean beliefs the user has over the items in $C_{i}^{t}$ and $\mu_{N-t}$ as the initial mean beliefs the user has over the remaining $N-t$ items, $\mathcal{J} \setminus C_{i}^{t}$.
+We partition the covariance matrix as follows:
+\[ \Sigma =  \left( \begin{array}{cc}
+\Sigma_{(N-t,N-t)} & \Sigma_{(N-t,t)} \\
+\Sigma_{(t,N-t)} & \Sigma_{(t,t)}
+\end{array} \right)
+\]
+Thus, after consuming the items in $C_{i}^{t}$ the resulting beliefs over the remaining items are given by $\mathcal{N}(\bar{\mu}, \bar{\Sigma})$ where $\bar{\mu}$ and $\bar{\Sigma}$ are as follows:
+\begin{align*}
+\bar{\mu} \mid c_t = \mu_{N-t} + \Sigma_{(N-t,t)} \Sigma_{(t,t)}^{-1}(c_t - \mu_t) \\
+\bar{\Sigma} \mid c_t = \Sigma_{(N-t,N-t)} - \Sigma_{(N-t,t)} \Sigma_{(t,t)}^{-1} \Sigma_{(t,N-t)}
+\end{align*}
 \begin{figure}[t]
@@
 The products are in different places of the product space, where 1 is close to 0 and 2 but more distant from 3, as shown in Figure \ref{fig:illustrative_example}.
-Suppose that the user actually believes that they all have expected value of 0 and consumes product 1.
+Suppose that the user's initial beliefs are that $\mu = 0$ and $\sigma^2 = 1$ for each product and, breaking the tie arbitrarily, consumes product 1.
 The underlying correlation structure implies that upon observing that $x_1=y$ the user will update beliefs about $\mathbb E x_0$, $\mathbb E x_2$ and $\mathbb E x_3$ to $\rho y$, $\rho y$ and $\rho^2 y$ respectively.
@@
 However, if risk aversion is high enough, the user might be willing to trade-off ex-ante expected lower values for lower risk and stick to local exploration of products that have just because the user has less uncertainty about their values.
-As the saying goes, if you are risk-averse, better the devil you know.
 \par
@@
 We analyze our model using numerical simulation since the sequential decision-making component of our model paired with the rich covariance structure between the items make it difficult to characterize optimal user behavior analytically.
-The Gaussian assumption allows for closed form Bayesian-updating which enables us to simulate our model but does not provide much help in analytical characterizations.
+The Gaussian assumption allows for closed form belief updating which enables us to simulate our model but does not provide much help in analytical characterizations.
 \par
@@
 We argue that these parameters are representative.
-First, for $\gamma$ we correspond parameter values from no risk aversion ($\gamma = 0$) to, for CARA preferences, a very high risk aversion level $(\gamma = 5)$.
-The correlation parameter, $\rho$, ranges from no correlation until almost full correlation $(\rho = 0.9)$.
+First, the values of $\gamma$ that we consider range from no risk aversion ($\gamma = 0$) to a very high risk aversion level under CARA preferences $(\gamma = 5)$.
+The correlation parameter, $\rho$, ranges from no correlation ($\rho = 0$) to nearly full correlation $(\rho = 0.9)$.
 $\beta$ ranges from the common value component playing no role $(\beta = 0)$ to the idiosyncratic component dominating the true valuations $(\beta = 5)$.
@@
 \cite{schulz2019structured} studies how individuals solve sequential decision-making problems under uncertainty in large choice sets in the context of mobile food delivery orders.
-They find that individuals engage in similarity-based generalizations where learning about the realized value of a particular good provides them with information about similar goods.
+They find that individuals engage in similarity-based generalizations where learning about the realized value of a particular good provides them with information about similar items.
 Content-based recommender systems exploit a similar idea where, if a user liked a given item, then she is also likely to like a similar but different item as well.
@@
 This underscores the fact that such concerns about filter bubbles \textit{only} arise in a world where there is correlation between the realized utility of items.
-If there were no correlation between the realized utilities then there would be no reason for users to consume similar goods and thus no narrowing effect, regardless of the information on the true utility of the items that users had.
+If there were no correlation between the realized utilities then there would be no reason for users to consume similar items and thus no narrowing effect, regardless of the information on the true utility of the items that users had.
 \par
@@
 Further, the average consumption path between periods is \textit{decreasing} for the no recommendation case whereas it is \textit{increasing} for the oracle case.
-The recommendation regime decreases the degree of local consumption but not as much as the oracle benchmark.
+The recommendation regime decreases the degree of local consumption, but not as much as the oracle benchmark.
 Due to the correlation of values, the oracle consumption path exploits this and leads to the consumption of more similar products than in the case when $\rho = 0$.
@@
 \includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_omni.pdf}\\
-\caption{Extent of Local Consumption for No Recommendation (Top), Recommendation (Middle) and Oracle (Bottom) for different Correlation values}
-\end{subfigure}%
+\end{subfigure}
+\caption{Extent of Local Consumption for No Recommendation (Top Left), Recommendation (Top Right) and Oracle (Bottom) for different Correlation values}
 \label{fig:local_consumption_across_rho}
@@
 \end{subfigure}%
-\caption{Effect of Risk Aversion on Local Consumption for No Recommendation (Top), Recommendation (Middle) and Oracle (Bottom)}
+\caption{Effect of Risk Aversion on Local Consumption for No Recommendation (Top Left), Recommendation (Top Right) and Oracle (Bottom)}
 \label{fig:no_rec_risk_aversion}
@@
 However, this is often the case in most environments where recommender systems are deployed.
-When there is a correlation between the realized utilities, then filter bubbles can naturally arise as a result of the information that users possess on the utilities of the items.
-We have shown that, absent any additional information, learning spillovers and risk aversion can lead users into filter bubbles.
+When there is a correlation between the realized utilities then filter bubbles can naturally arise due to the nature of how individuals acquire additional information about the remaining items.
+We have shown that, unless users are provided with additional information to guide their consumption choices, then these learning spillovers and user risk-aversion can lead users into filter bubbles.
 When users consume high valued items, they exploit the underlying correlation across different products' values, stronger for similar products, which leads them to increasingly consume products in narrower and narrower portions of the product space.
@@
 Recommendation leads to these effects being mitigated by providing users with additional information on products outside the already explored portions of the product space.
+Further, if all uncertainty was resolved as in the oracle regime, then such behavior is not present.
 
@@
 In this section we primarily focus on the impact of recommendation on the welfare on users as well as the overall diversity of the items that they consume.
-
-While in the previous section we looked at the distance between consecutive goods, in this section we focus on a diversity measure that considers the entire consumed set of items.
+\par 
+While in the previous section we looked at the distance between consecutive items, in this section we focus on a diversity measure that considers the entire consumed set of items.
 The diversity measure we utilize is common in recommender system literature (e.g.
@@
 $$D_i:=\frac{1}{N}\frac{1}{T(T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$$
+
+\begin{figure}[ht]
+\includegraphics[width=.45\linewidth]{figures/gamma_diversity_N_200_T_20}
+\captionof{figure}{Risk Aversion and Diversity}\label{fig:risk_aversion_diversity}
+\end{figure}
 
@@
 
-\begin{figure}[ht]
-\includegraphics[width=.4\linewidth]{figures/gamma_diversity_N_200_T_20}
-\captionof{figure}{Risk Aversion and Diversity}\label{fig:risk_aversion_diversity}
-\end{figure}
-
 \begin{finding}\label{finding_welfare_gap}
@@
 The intuition for this is clear as recommendation provides users with information that allows them to better guide their decisions and increase welfare.
-However, as $\rho$ increases users get increasingly more information simply from consuming items since the realized utility is now more informative about the utility of nearby goods.
+However, as $\rho$ increases users get increasingly more information simply from consuming items since the realized utility is now more informative about the utility of nearby items.
 Thus, since consumption decisions themselves reveal information, the information that recommendation reveals provides less  value.
@@
 \end{subfigure}
-\captionof{figure}{Welfare, Diversity, and Correlation}\label{fig:diversity_correlation}
+\captionof{figure}{User Welfare and Diversity as the strength of correlation, $\rho$, varies}\label{fig:diversity_correlation}
+\end{figure}
+
+\begin{figure}
+\begin{subfigure}{.45\linewidth}
+  \includegraphics[width=.8\linewidth]{figures/beta_diversity_N_200_T_20.pdf}
+\end{subfigure}
+\begin{subfigure}{.45\linewidth}
+  \includegraphics[width=.8\linewidth]{figures/beta_welfare_N_200_T_20.pdf}
+\end{subfigure}
+\captionof{figure}{User Welfare and Diversity as the strength of the common value component, $\beta$, varies}\label{fig:diversity_correlation}
 \end{figure}
@@
 However, as we saw in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, increasing $\gamma$ can lead to lower diversity and increasingly local consumption due to the fact that the degree of uncertainty now impacts users' choices.
-This weakens the negative relationship between diversity and welfare since both negative and positive experiences with a good reduce uncertainty about surrounding goods.
+This weakens the negative relationship between diversity and welfare since both negative and positive experiences with a good reduce uncertainty about surrounding items.
 This leads to the inverted-U shape found in Figure \ref{fig:diversity_welfare_ra} when $\gamma$ is relatively large (e.g $\gamma = 5$) though diversity and welfare are virtually uncorrelated in the data.
@@
 \end{subfigure}\\
-\captionof{figure}{Diversity vs. Welfare, No-Recommendation, $\gamma = 0$ (Top), Diversity vs. Welfare, $\gamma = 5$ (Bottom)}\label{fig:diversity_welfare_ra}
+\captionof{figure}{Diversity vs. Welfare, No-Recommendation, $\gamma = 0$ (Left), Diversity vs. Welfare, $\gamma = 5$ (Right)}\label{fig:diversity_welfare_ra}
 \end{figure}
@@
 \end{subfigure}\\
-\captionof{figure}{Diversity vs. Welfare, Recommendation, $\gamma = 0$ (Top), Diversity vs. Welfare, $\gamma = 5$ (Bottom)}\label{fig:diversity_welfare_ra_partial}
+\captionof{figure}{Diversity vs. Welfare, Recommendation, $\gamma = 0$ (Left), Diversity vs. Welfare, $\gamma = 5$ (Right)}\label{fig:diversity_welfare_ra_partial}
 \end{figure}
@@
 We have studied a model of user decision-making in the context of markets where recommender systems are traditionally deployed in order to understand the consequences of recommender systems on consumption patterns of users.
-The key components of the model are that users make decisions under uncertainty and that consuming a good leads to informational spillovers about similar goods.
+The key components of the model are that users make decisions under uncertainty and that consuming a good leads to informational spillovers about similar items.
 We have shown that this alone can generate consumption behavior that is consistent with filter bubble effects that have traditionally been associated with personalized recommender systems.
@@
 \par
-
-We have used the insights from our model to guide the evaluation and design of recommender systems.
-The key insight is that a critical component of designing good recommendations is understanding what user behavior would be without recommendation and then understanding how recommendation would change their consumption.
-Operationalizing this requires collecting information on user beliefs about the goods in the product space, which is not data that is traditionally collected for the development of recommender systems.
-\par
-
 We leave for future work exploring to what extent user behavior changes if we drop the myopia assumption, so that users themselves are engaging in exploration.
@@
 Even under the myopia assumption, understanding which recommendation system is optimal is an interesting and challenging problem.
-\par
-
-% Start of "Sample References" section
-% In the interest of anonymization, please do not include acknowledgements in your submission.
-%
-%\begin{acks}
-%
-%	The authors would like to thank Dr. Maura Turolla of Telecom
-%	Italia for providing specifications about the application scenario.
-%
-%	The work is supported by the \grantsponsor{GS501100001809}{National
-%		Natural Science Foundation of
-%		China}{http://dx.doi.org/10.13039/501100001809} under Grant
-%	No.:~\grantnum{GS501100001809}{61273304\_a}
-%	and~\grantnum[http://www.nnsf.cn/youngscientsts]{GS501100001809}{Young
-%		Scientsts' Support Program}.
-%
-%
-%\end{acks}
 

### 2020-01-19 paper/filter_bubbles.tex
udpated paper

@@
 
-Users in many online markets face difficult decision problems as they have to sort through the vast choice sets present in these markets.
+Users in many online platforms face difficult decision problems as they have to sort through the vast choice sets present in these markets.
 For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
@@
 \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
-Further, an emerging worry is that recommender systems can be utilized as a tool of market power by steering consumer choice towards products more profitable for online platforms to the detriment of consumer welfare [citations].
+Further, an emerging worry is that recommender systems can be utilized as a tool of market power by steering consumer choice towards products more profitable for online platforms to the detriment of consumer welfare \citep{aridor2020recommender, bourreau2018streaming, DeCorniereTaylor2019RAND}.
 Finally, there has emerged a burgeoning literature that focuses on understanding how recommender systems themselves can acquire information about product quality from the choices of self-interested consumers \cite{kremer2014implementing, che2017recommender,frazier2014incentivizing,mansour2015bayesian}.
@@
 \xhdr{Additional Findings} We further explore the diversity of the overall set of items that users consume.
-Consumption diversity is increasingly a common goal of many recommender systems [references], which is reasonable since it could be expected that higher diversity is associated with more intense search and therefore higher welfare.
+Consumption diversity is increasingly a common goal of many recommender systems \cite{castells2015novelty, kunaver2017diversity}, which is reasonable since it could be expected that higher diversity is associated with more intense search and therefore higher welfare.
 However, we find that, without recommendation, higher diversity of items does not ensure higher welfare.
@@
 \end{subfigure}
-\caption{Extent of Local Consumption for No Recommendation (Top Left), Recommendation (Top Right) and Oracle (Bottom) for different Correlation values}
+\caption{Extent of Local Consumption for No Recommendation (Top Left), Recommendation (Top Right) and Oracle (Bottom) as strength of correlation, $\rho$, varies}
 \label{fig:local_consumption_across_rho}
@@
 \subsection{User Welfare and Product Diversity}
-In this section we primarily focus on the impact of recommendation on the welfare on users as well as the overall diversity of the items that they consume.
+In this section we primarily focus on the impact of recommendation on user welfare and the overall diversity of the items that they consume.
 \par 
@@
 As a result, this effect gets amplified as $\rho$ increases, which leads to the gap in diversity between the regimes to increase as $\rho$ increases.
-Figure \ref{fig:diversity_correlation} shows how diversity varies as $\rho$ increases across the three regimes that we consider.
+Figure \ref{fig:diversity_welfare_correlation} shows how diversity varies as $\rho$ increases across the three regimes that we consider.
 There is a similar increasing diversity gap as $\gamma$, or the level of risk-aversion, increases as can be seen in Figure \ref{fig:risk_aversion_diversity}.
@@
 \par 
-We now study how recommendation impacts consumer welfare in our model.
+We now study how recommendation impacts user welfare in our model.
 In our model users make consumption decisions that maximize their current period ex-ante utility that depends on their beliefs in that period.
@@
 \item Under no recommendation, welfare is increasing as $\rho$ increases.
-\item Recommendation introduces welfare gains, but these gains are decreasing as $\rho$ increases. 
+\item Recommendation introduces welfare gains relative to no recommendation, but these gains are decreasing as $\rho$ increases. 
 \end{enumerate}
@@
 \par 
-Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which are drawn from Figure~\ref{fig:diversity_correlation}.
+Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which are drawn from Figure~\ref{fig:diversity_welfare_correlation}.
 The most interesting observation is that the value of recommendation decreases as $\rho$ decreases.
@@
 Thus, since consumption decisions themselves reveal information, the information that recommendation reveals provides less value.
+\par 
+Figure~\ref{fig:diversity_welfare_common_value} shows how diversity and welfare vary across recommendation regimes as the the strength of the common-value component, $\beta$, varies.
+As expected, when $\beta = 0$, these values are the same.
+However, as $\beta$ grows the idiosyncratic component of user valuations becomes increasingly less important and the recommendation regime begins to converge to the oracle regime.
+Indeed, we see that both product diversity and welfare increase under recommendation, whereas under no recommendation we see that there is a decrease in diversity with little changes to consumer welfare.
+This is intuitive as $\beta$ controls the informativeness of recommendation about user preferences.
 \par 
@@
 \end{subfigure}
-\captionof{figure}{User Welfare and Diversity as the strength of correlation, $\rho$, varies}\label{fig:diversity_correlation}
+\caption{User Welfare and Diversity as the strength of correlation, $\rho$, varies}\label{fig:diversity_welfare_correlation}
 \end{figure}
@@
 \end{subfigure}
-\captionof{figure}{User Welfare and Diversity as the strength of the common value component, $\beta$, varies}\label{fig:diversity_correlation}
+\caption{User Welfare and Diversity as the strength of the common value component, $\beta$, varies}\label{fig:diversity_welfare_common_value}
 \end{figure}
@@
 \end{subfigure}\\
-\captionof{figure}{Diversity vs. Welfare, No-Recommendation, $\gamma = 0$ (Left), Diversity vs. Welfare, $\gamma = 5$ (Right)}\label{fig:diversity_welfare_ra}
+\caption{Diversity vs. Welfare, No-Recommendation, $\gamma = 0$ (Left), Diversity vs. Welfare, $\gamma = 5$ (Right)}\label{fig:diversity_welfare_ra}
 \end{figure}
@@
 \end{subfigure}\\
-\captionof{figure}{Diversity vs. Welfare, Recommendation, $\gamma = 0$ (Left), Diversity vs. Welfare, $\gamma = 5$ (Right)}\label{fig:diversity_welfare_ra_partial}
+\caption{Diversity vs. Welfare, Recommendation, $\gamma = 0$ (Left), Diversity vs. Welfare, $\gamma = 5$ (Right)}\label{fig:diversity_welfare_ra_partial}
 \end{figure}
@@
 \par 
-We leave for future work exploring to what extent user behavior changes if we drop the myopia assumption, so that users themselves are engaging in exploration.
+Our results highlight that user choice in these markets is path dependent as a result of the nature of user preferences and how these preferences evolve over time.
+This has important ramifications for understanding the extent to which recommender systems can steer user behavior, potentially in anti-competitive ways.
+Steering users towards certain items today provides them with information that they utilize to make decisions tomorrow, which leads them to consume similar items to those they were originally steered towards.
+Thus, online platforms can potentially shape the long-term preferences of users and not only impact their current period choices.
+This has broader ramifications about the equilibrium effects of such behavior on competition and consumer welfare in online platform which we leave to future work.
+\par
+We leave for future work exploring to what extent user behavior changes if we drop the myopia assumption, so that users themselves are engaging in explicit exploration.
 This likely interacts with recommendation in interesting ways as the recommendations may provide information that can complement or crowd-out user-led exploration.
-Our model also reveals a clear path-dependence in choice since consumption decisions today shape the information users have tomorrow about other products which affects their consumption choices tomorrow, etc.
 Even under the myopia assumption, understanding which recommendation system is optimal is an interesting and challenging problem.
-
+Further, our results point to an important consideration in the literature on incentivizing exploration which is that the exploration faced by the recommender itself likely interacts in interesting and important ways with the choice problem faced by the users.
+\par 
+Overall, understanding how users make choices in online platforms and how recommender systems impact these choices is important to understand.
+Understanding these issues allows us to better understand the broader social and economic impact of such systems as well as can provide a guide to the improved design of such systems.
 % Bibliography

### 2020-01-19 paper/filter_bubbles.tex
new illustrative example

@@
 \[ \Sigma =  \left( \begin{array}{cc}
-\Sigma_{(N-t,N-t)} & \Sigma_{(N-t,t)} \\
+\Sigma_{(N-t, N-t)} & \Sigma_{(N-t,t)} \\
 \Sigma_{(t,N-t)} & \Sigma_{(t,t)}
@@
 \xhdr{An Illustrative Example} We illustrate our model with a simple example.
-Suppose that there are four products: 0, 1, 2, 3.
-The products are in different places of the product space, where 1 is close to 0 and 2 but more distant from 3, as shown in Figure \ref{fig:illustrative_example}.
-Suppose that the user's initial beliefs are that $\mu = 0$ and $\sigma^2 = 1$ for each product and, breaking the tie arbitrarily, consumes product 1.
-The underlying correlation structure implies that upon observing that $x_1=y$ the user will update beliefs about $\mathbb E x_0$, $\mathbb E x_2$ and $\mathbb E x_3$ to $\rho y$, $\rho y$ and $\rho^2 y$ respectively.
-This is because learning about product 1's value is more informative about similar products' value, 0 and 2, than products further away in the product space, 3.
-Therefore, the user will adjust her beliefs about their mean -- zero in the example -- more strongly for products which are similar.
-Moreover, now uncertainty about the value of products 0, 2 and 3 is also decreased but differentially so: learning product 1's value has contributed to a stronger reduction of uncertainty for nearby products that for product 3.
-If the actual value was positive, this induces the user to perform some local exploration, as 0 and 2 have higher mean and lower variance.
-If instead it was negative, then user's posterior beliefs penalize more strongly products 0 and 2 than 3, so if she is not too risk averse, she would go across the product space in search of some better element.
-However, if risk aversion is high enough, the user might be willing to trade-off ex-ante expected lower values for lower risk and stick to local exploration of products that have just because the user has less uncertainty about their values.
+Suppose that there are four items: 0, 1, 2, 3.
+The items are in different places of the product space, where 0 is close to 1 and 3 but more distant from 2, as shown in Figure \ref{fig:illustrative_example}.
+The initial beliefs are given as follows:
+\[ \mu = \left (\begin{array}{c}
+\mathbb{E}[x_0] \\
+\mathbb{E}[x_1] \\
+\mathbb{E}[x_2] \\
+\mathbb{E}[x_3]
+\end{array}  \right) = \left (\begin{array}{c}
+0 \\
+0\\
+0 \\
+0
+\end{array}  \right), \Sigma =  \sigma^{2} \left( \begin{array}{cccc}
+\rho^{0} & \rho^{1} & \rho^{2} & \rho^{0} \\
+\rho^{1} & \rho^{0} & \rho^{1} & \rho^{2} \\
+\rho^{2} & \rho^{1} & \rho^{0} & \rho^{1} \\
+\rho^{1} & \rho^{2} & \rho^{1} & \rho^{0} \\
+\end{array} \right)
+\]
+Suppose that, in period 1, the user breaks ties arbitrarily breaking the tie arbitrarily, consumes product 0.
+The underlying correlation structure implies that upon observing that $x_1=y$ the user will update beliefs about the remaining 3 products according to the previously specified updating rule.
+For concreteness, we suppose that $\sigma = 1$, $y = 0.5$, and $\rho = 0.5$.
+The resulting beliefs after observing $y$ are then as follows:
+\[ \bar{\mu} =   \left (\begin{array}{c}
+\mathbb{E}[x_1] \vspace{0.15cm} \\
+\mathbb{E}[x_2] \vspace{0.15cm} \\
+\mathbb{E}[x_3]
+\end{array}  \right) =\left (\begin{array}{c}
+\rho y = \frac{1}{4} \vspace{0.15cm} \\
+\rho^{2} y = \frac{1}{8} \vspace{0.15cm} \\
+ \rho y = \frac{1}{4} \\
+\end{array}  \right), \bar{\Sigma} =  \left( \begin{array}{ccc}
+\frac{3}{4} & \frac{3}{8} & 0.0 \vspace{0.15cm} \\
+\frac{3}{8} & \frac{15}{16} & \frac{3}{8} \vspace{0.15cm}  \\
+0.0 &\frac{3}{8} & \frac{3}{4}  \\
+\end{array} \right)
+\]
+Thus, upon learning $y$, the user updates beliefs about the remaining items.
+However, note that $\mathbb{E}[x_1] = \mathbb{E}[x_3] > \mathbb{E}[x_2]$ since item 0's value is more informative about similar items' values, items 1 and 3, than items further way in the product space such as item 2.
+Moreover, $\bar{\Sigma}_{11} = \bar{\Sigma}_{33} < \bar{\Sigma}_{22}$ as the uncertainty about items 1 and 3 is further reduced compared to item 2.
+Thus, since $y > 0$, the user in the next period will consume nearby items to item 0 since, even though initially she believed that all items had the same mean, the spillover from consuming item 0 leads her to believe that items 1 and 3 have higher expected valuations.
+\par 
+Suppose instead that $y = -0.5 < 0$.
+Now we have that $\mathbb{E}[x_1] = \mathbb{E}[x_3] = \frac{-1}{4}$ and $\mathbb{E}[x_2] = \frac{-1}{8}$ with $\bar{\Sigma}$ remaining the same.
+Now, if the user is risk-neutral ($\gamma = 0$), then she will go across the product space to consume item $2$ in the next period.
+However, if she is risk-averse she still may consume nearby items since her uncertainty about items $1$ and $3$ is still lower than item $2$.
+In particular, this will happen when
+\begin{align*}
+\delta(3) = \delta(1) = \rho y - \frac{1}{2} \gamma \bar{\Sigma}_{11} > \rho^{2} y - \frac{1}{2} \gamma \bar{\Sigma}_{22} = \delta(2)
+\end{align*}
+Supposing $y = -0.5$ and the aforementioned parametrization, the user will still consume items $1$ or $3$ when $\gamma > \frac{4}{3}$.
+Thus if risk aversion is high enough, the user might be willing to trade-off ex-ante expected lower values for lower risk and stick to consuming nearby items just because these items have lower uncertainty.
 \par

### 2020-01-19 paper/filter_bubbles.tex
first pass at EC paper

@@
 
-\xhdr{An Illustrative Example} We illustrate our model with a simple example.
+\xhdr{An Illustrative Example} We illustrate the main intuitions of our model with a simple example.
 Suppose that there are four items: 0, 1, 2, 3.
@@
 \]
-Suppose that, in period 1, the user breaks ties arbitrarily breaking the tie arbitrarily, consumes product 0.
-The underlying correlation structure implies that upon observing that $x_1=y$ the user will update beliefs about the remaining 3 products according to the previously specified updating rule.
+In period 1, every item is ex-ante identical since they have the same mean and variance and so suppose that the user breaks the tie arbitrarily and consumes item 0.
+The underlying correlation structure implies that upon observing that $x_0=y$ the user will update beliefs about the remaining 3 items according to the previously specified updating rule.
 For concreteness, we suppose that $\sigma = 1$, $y = 0.5$, and $\rho = 0.5$.
@@
 Thus, upon learning $y$, the user updates beliefs about the remaining items.
-However, note that $\mathbb{E}[x_1] = \mathbb{E}[x_3] > \mathbb{E}[x_2]$ since item 0's value is more informative about similar items' values, items 1 and 3, than items further way in the product space such as item 2.
+Note that $\mathbb{E}[x_1] = \mathbb{E}[x_3] > \mathbb{E}[x_2]$ since item 0's value is more informative about similar items' values, items 1 and 3, than items further way in the product space such as item 2.
 Moreover, $\bar{\Sigma}_{11} = \bar{\Sigma}_{33} < \bar{\Sigma}_{22}$ as the uncertainty about items 1 and 3 is further reduced compared to item 2.
-Thus, since $y > 0$, the user in the next period will consume nearby items to item 0 since, even though initially she believed that all items had the same mean, the spillover from consuming item 0 leads her to believe that items 1 and 3 have higher expected valuations.
+Thus, since $y > 0$, the user in the next period will consume items nearby to item 0 since, even though initially she believed that all items had the same mean, the spillover from consuming item 0 leads her to believe that items 1 and 3 have higher expected valuations.
+Since both the mean is higher for these items and the variance is lower, the user will consume items 1 and 3 regardless of her risk aversion levels.
 \par 
-Suppose instead that $y = -0.5 < 0$.
-Now we have that $\mathbb{E}[x_1] = \mathbb{E}[x_3] = \frac{-1}{4}$ and $\mathbb{E}[x_2] = \frac{-1}{8}$ with $\bar{\Sigma}$ remaining the same.
-Now, if the user is risk-neutral ($\gamma = 0$), then she will go across the product space to consume item $2$ in the next period.
-However, if she is risk-averse she still may consume nearby items since her uncertainty about items $1$ and $3$ is still lower than item $2$.
+Suppose instead that item 0 ends up having a negative valuation so that $y = -0.5 < 0$.
+Now we have that $\mathbb{E}[x_1] = \mathbb{E}[x_3] = -\frac{1}{4} <  -\frac{1}{8} = \mathbb{E}[x_2] = $ with $\bar{\Sigma}$ remaining the same as when $y = 0.5$.
+Now, the risk-aversion levels of the user determine the choice in the next period.
+If the user is risk-neutral ($\gamma = 0$), then she will go across the product space to consume item $2$ in the next period since it has a higher expected value.
+However, if she is sufficiently risk-averse then she still may consume item $1$ or $3$ since her uncertainty about these items is lower than item $2$.
 In particular, this will happen when
@@
 \end{align*}
-Supposing $y = -0.5$ and the aforementioned parametrization, the user will still consume items $1$ or $3$ when $\gamma > \frac{4}{3}$.
-Thus if risk aversion is high enough, the user might be willing to trade-off ex-ante expected lower values for lower risk and stick to consuming nearby items just because these items have lower uncertainty.
+Given the aforementioned parametrization and $y = -0.5$, the user will consume item $1$ or $3$ when $\gamma > \frac{4}{3}$ and will consume item $2$ when $\gamma < \frac{4}{3}$.
+Thus if the user is risk averse enough, then she might be willing to trade-off ex-ante lower expected values for lower risk and stick to consuming nearby items just because these items have lower uncertainty.
+\par 
+This example illustrates the main mechanisms in our model that can lead to excessive consumption of similar items.
+Once the user finds items in the product space with high valuations she will update her beliefs positively about items in this portion of the product space and continue consuming these items regardless of her level of risk aversion.
+However, this same updating leads to a reduction in uncertainty of these items and so, if she is sufficiently risk-averse, she still may continue consuming items in this portion of the product space, even if she has bad experiences with them, since they are perceived to be less risky.
 \par
@@
   \centering
-  \includegraphics[width=.8\linewidth]{figures/rho_pos_consumption_dist_N_200T_20_overall.pdf}
-  \caption{1a}
+  \includegraphics[width=.9\linewidth]{figures/rho_pos_consumption_dist_N_200T_20_overall.pdf}
   \label{fig:sfig1}
@@
   \centering
-  \includegraphics[width=.8\linewidth]{figures/rho_zero_consumption_dist_N_200T_20_overall.pdf}
-  \caption{1b}
+  \includegraphics[width=.9\linewidth]{figures/rho_zero_consumption_dist_N_200T_20_overall.pdf}
   \label{fig:sfig2}
 \end{subfigure}
-\caption{Consumption path with (Top) and without (Bottom) correlation between values}
+\caption{Consumption path with (Left) and without (Right) correlation between values}
 \label{fig:correlation_consumption_path}
@@
 \end{subfigure}
-\caption{Extent of Local Consumption for No Recommendation (Top Left), Recommendation (Top Right) and Oracle (Bottom) as strength of correlation, $\rho$, varies}
+\caption{Extent of Local Consumption as strength of correlation, $\rho$, varies. \\ No Recommendation (Top Left), Recommendation (Top Right) and Oracle (Bottom)}
 \label{fig:local_consumption_across_rho}
@@
 \end{subfigure}%
-\caption{Effect of Risk Aversion on Local Consumption for No Recommendation (Top Left), Recommendation (Top Right) and Oracle (Bottom)}
+\caption{Effect of Risk Aversion on Local Consumption \\ No Recommendation (Top Left), Recommendation (Top Right) and Oracle (Bottom)}
 \label{fig:no_rec_risk_aversion}

### 2020-01-19 simulation_code/toy_example.jl

@@ -1,0 +1,168 @@
+
+using JSON
+using IterTools;
+using Dates;
+using Distances;
+using Random;
+using Distributions;
+using LinearAlgebra;
+
+""" the minimum # of hops from node i to node j
+"""
+
+function hop_distance(
+        i::Int64,
+        j::Int64,
+        N::Int64
+    )::Int64
+    return min(abs(i-j), abs(j-i), abs(j-i-N), abs(i-j-N))
+end
+
+"""
+    cov_mat_fun(sigma::Float64, rho::Float64, N::Int64)
+Create covariance matrix
+# Arguments
+- `sigma::Float64`: 
+- `rho::Float64`: Covariance coefficient
+- `N::Int`: Number of goods
+# Returns
+- Covariance matrix 
+"""
+
+function cov_mat_fun(sigma::Float64, rho::Float64, N::Int64)::Array{Float64,2}
+    cov_mat = zeros(Float64, N, N)
+    for i in 1:N 
+        for j in 1:N
+            dist = hop_distance( i, j , N)
+            #dist = Distances.euclidean( iota(i,N), iota(j,N) )
+            cov_mat[i,j] = sigma * rho^dist
+        end
+    end
+    return cov_mat
+end
+
+""" 
+CARA
+# Arguments:
+- alpha (int) :
+- mu (numpy.ndarray): shape (N,)
+- sigma (numpy.ndarray) : shape (N,N)
+# Returns:
+    (numpy.ndarray) of shape (N,)
+# Notes:
+    alpha: the coefficient of absolute risk aversion
+    μ and σ2 are the mean and the variance of the distribution F
+    https://ocw.mit.edu/courses/economics/14-123-microeconomic-theory-iii-spring-2015/lecture-notes-and-slides/MIT14_123S15_Chap3.pdf 
+    pg 21
+"""
+
+function certainty_equivalent(
+        alpha::Float64, 
+        mu::Array{Float64,1},
+        sig::Array{Float64,2}
+    )
+    new_mu = mu - (.5 * alpha * diag(sig)) 
+    return new_mu
+end
+
+### Welfare Functions - Statistic Calculation Functions
+
+
+function init_sigma(Cit::Array{Int64,1},
+            Nit::Array{Int64,1},
[…]

### 2020-01-19 paper/filter_bubbles.tex
aesthetic
anotehr typo

@@
 \par
-\noindent \textbf{Users.} We consider a set of users $I$ where each user $i \in I$ faces the same finite set of $N$ items $\mathcal J = \left\{0,1,...,N-1\right\}$.
+\xhdr{Users.} We consider a set of users $I$ where each user $i \in I$ faces the same finite set of $N$ items $\mathcal J = \left\{0,1,...,N-1\right\}$.
 For simplicity, we assume that users only derive pleasure from item $n \in \mathcal{J}$ the first time they consume it.
@@
 \end{array}  \right), \bar{\Sigma} =  \left( \begin{array}{ccc}
-\frac{3}{4} & \frac{3}{8} & 0.0 \vspace{0.15cm} \\
+\frac{3}{4} & \frac{3}{8} & 0 \vspace{0.15cm} \\
 \frac{3}{8} & \frac{15}{16} & \frac{3}{8} \vspace{0.15cm}  \\
-0.0 &\frac{3}{8} & \frac{3}{4}  \\
+0 &\frac{3}{8} & \frac{3}{4}  \\
 \end{array} \right)
@@
 Suppose instead that item 0 ends up having a negative valuation so that $y = -0.5 < 0$.
-Now we have that $\mathbb{E}[x_1] = \mathbb{E}[x_3] = -\frac{1}{4} <  -\frac{1}{8} = \mathbb{E}[x_2] = $ with $\bar{\Sigma}$ remaining the same as when $y = 0.5$.
+Now we have that $\mathbb{E}[x_1] = \mathbb{E}[x_3] = -\frac{1}{4} <  -\frac{1}{8} = \mathbb{E}[x_2]$ with $\bar{\Sigma}$ remaining the same as when $y = 0.5$.
 Now, the risk-aversion levels of the user determine the choice in the next period.

### 2020-01-21 simulation_code/simulations.jl
add thompson

@@ -199,6 +199,26 @@ end
     return cit
 end 
 
+function thompson_sampling(
+    mu_V::Array{Float64,1},
+    Sigma_V::Array{Float64,2},
+    N
+)
+
+    
+    draws = [ 
+            rand(
+                MvNormal(
+                    [mu_V[ii]], 
+                    [Sigma_V[ii,ii]]
+                )
+            )  for ii in 1:N ]
+    c_it_index = argmax(draws)
+    c_it = draws[c_it_index][1]  
+
+    return c_it
+end
+
 @everywhere function choice_part(
     V_i::Array{Float64,1}, 
     mu_V_i::Array{Float64,1},
@@ -231,7 +251,7 @@ end
 
         c_it = nothing
         if use_thompson
-            c_it = argmax(rand(V))
+            c_it = thompson_sampling(mu_V_i, Sigma_V_i, N)
         elseif rand() < epsilon
             c_it = rand(choice_set)
         else
@@ -283,6 +303,8 @@ end
         c_it = 0
         if use_thompson
             c_it = argmax(rand(U_i)) 
+            c_it = thompson_sampling(mu_U_i, Sigma_U_i, N)
+            @show c_it
         elseif rand() < epsilon
             c_it = rand(choice_set)
         else
@@ -328,7 +350,8 @@ end
     Sigma_V_ibar::Array{Float64,2},
     alpha::Float64,
     epsilon::Float64,
-    seed::Int64
+    seed::Int64,
+    use_thompson::Bool
     )
 
     
@@ -349,7 +372,6 @@ end
     mu_V = zeros(Float64, N)
     V = rand(MvNormal(mu_V, Sigma_V))
 
-    use_thompson::Bool = true
     for it_ind=1:nr_ind
         # V_i = (v_in) n in I aka: consumer i’s idiosyncratic taste for good n in vector form
 
@@ -385,27 +407,28 @@ end
     return Dict( "Consumption" => C_pop, "Welfare" => W_pop, "Rec" => R_pop )
 end
 
+use_thompson::Bool = true
 #
-nr_pop = 100
+nr_pop = 10
 #
-nr_ind = 100
+nr_ind = 10
 #
 sigma_ibar = .1
 #
 rho_ibar = 0.0
 
[…]

### 2020-01-21 simulation_code/test.jl

@@ -3,6 +3,7 @@ using Distributions;
 using Distances;
 using Random;
 
+
 function hop_distance(
         i::Int64,
         j::Int64,
@@ -23,6 +24,35 @@ function cov_mat_fun(sigma::Float64, rho::Float64, N::Int64)::Array{Float64,2}
     return cov_mat
 end
 
+function thompson_sampling(
+    mu_V::Array{Float64,1},
+    Sigma_V::Array{Float64,2},
+    N
+)
+
+
+    draws = [
+            rand(
+                MvNormal(
+                    [mu_V[ii]],
+                    [Sigma_V[ii,ii]]
+                )
+            )  for ii in 1:N ]
+    c_it_index = argmax(draws)
+    c_it = draws[c_it_index][1]
+
+    @show draws
+    @show c_it_index
+    @show c_it
+
+    return c_it
+end
+
+
+
+seed = 1
+Random.seed!(seed);
+
 N = 20
 sigma = 0.25
 rho = 0.1
@@ -32,13 +62,11 @@ Nset = [ n for n=1:N]   # set of N items I = {1, ..., N}
 mu_V = zeros(Float64, N)
 V = rand(MvNormal(mu_V, Sigma_V))
 
-@show V
-x = rand(V)
-@show x
-@show typeof(x)
-y = argmax(x)
-@show y
+@show typeof(mu_V)
+@show mu_V[1]
+@show Sigma_V[1,1]
+#MvNormal(Int(0.0), 0.25)
 
-x = zeros(Float64, N)
-@show Random.rand!(V, x)
-@show x
+@show rand(MvNormal( [mu_V[1]] , [Sigma_V[1,1]] ))[1]
+
+thompson_sampling(mu_V, Sigma_V, N)

### 2020-01-21 simulation_code/simulations.jl
remove bad c_it

@@ -302,7 +302,6 @@ end
         
         c_it = 0
         if use_thompson
-            c_it = argmax(rand(U_i)) 
             c_it = thompson_sampling(mu_U_i, Sigma_U_i, N)
             @show c_it
         elseif rand() < epsilon

### 2020-01-21 simulation_code/simulations.jl
restore original params

@@ -408,17 +408,17 @@ end
 
 use_thompson::Bool = true
 #
-nr_pop = 10
+nr_pop = 100
 #
-nr_ind = 10
+nr_ind = 100
 #
 sigma_ibar = .1
 #
 rho_ibar = 0.0
 
-N_vals = [20]
+N_vals = [200]
 
-T_vals = [2]
+T_vals = [20]
 
 # Covariance structure
 rho_vals = [0.1, 0.3, 0.5, 0.7, 0.9]

### 2020-01-21 simulation_code/simulations.jl
update thompson_sampling use univariate normal

@@ -201,20 +201,18 @@ end
 
 function thompson_sampling(
     mu_V::Array{Float64,1},
-    Sigma_V::Array{Float64,2},
-    N
+    Sigma_V::Array{Float64,2}
 )
 
     
     draws = [ 
             rand(
-                MvNormal(
-                    [mu_V[ii]], 
-                    [Sigma_V[ii,ii]]
+                Normal(
+                    mu_V[ii],
+                    Sigma_V[ii,ii]
                 )
-            )  for ii in 1:N ]
-    c_it_index = argmax(draws)
-    c_it = draws[c_it_index][1]  
+            )  for ii in 1:length(mu_V) ]
+    c_it = argmax(draws)
 
     return c_it
 end
@@ -251,7 +249,7 @@ end
 
         c_it = nothing
         if use_thompson
-            c_it = thompson_sampling(mu_V_i, Sigma_V_i, N)
+            c_it = thompson_sampling(mu_V_i, Sigma_V_i)
         elseif rand() < epsilon
             c_it = rand(choice_set)
         else
@@ -302,7 +300,7 @@ end
         
         c_it = 0
         if use_thompson
-            c_it = thompson_sampling(mu_U_i, Sigma_U_i, N)
+            c_it = thompson_sampling(mu_U_i, Sigma_U_i)
             @show c_it
         elseif rand() < epsilon
             c_it = rand(choice_set)

### 2020-01-21 simulation_code/test.jl

@@ -26,23 +26,20 @@ end
 
 function thompson_sampling(
     mu_V::Array{Float64,1},
-    Sigma_V::Array{Float64,2},
-    N
+    Sigma_V::Array{Float64,2}
 )
 
 
     draws = [
             rand(
-                MvNormal(
-                    [mu_V[ii]],
-                    [Sigma_V[ii,ii]]
+               Normal(
+                    mu_V[ii],
+                    Sigma_V[ii,ii]
                 )
-            )  for ii in 1:N ]
-    c_it_index = argmax(draws)
-    c_it = draws[c_it_index][1]
+            )  for ii in 1:length(mu_V) ]
+    c_it = argmax(draws)
 
     @show draws
-    @show c_it_index
     @show c_it
 
     return c_it
@@ -65,8 +62,5 @@ V = rand(MvNormal(mu_V, Sigma_V))
 @show typeof(mu_V)
 @show mu_V[1]
 @show Sigma_V[1,1]
-#MvNormal(Int(0.0), 0.25)
 
-@show rand(MvNormal( [mu_V[1]] , [Sigma_V[1,1]] ))[1]
-
-thompson_sampling(mu_V, Sigma_V, N)
+thompson_sampling(mu_V, Sigma_V)

### 2020-01-21 simulation_code/simulations.jl
remove debug

@@ -301,7 +301,6 @@ end
         c_it = 0
         if use_thompson
             c_it = thompson_sampling(mu_U_i, Sigma_U_i)
-            @show c_it
         elseif rand() < epsilon
             c_it = rand(choice_set)
         else

### 2020-01-21 simulation_code/simulations.jl
thompson first try (#17)
* thompson first try

* some fixes

* Add a dummy test to run things faster

* add thompson

* remove bad c_it

* restore original params

* update thompson_sampling use univariate normal

* remove debug

@@ -199,6 +199,23 @@ end
     return cit
 end 
 
+function thompson_sampling(
+    mu_V::Array{Float64,1},
+    Sigma_V::Array{Float64,2}
+)
+
+    
+    draws = [ 
+            rand(
+                Normal(
+                    mu_V[ii],
+                    Sigma_V[ii,ii]
+                )
+            )  for ii in 1:length(mu_V) ]
+    c_it = argmax(draws)
+
+    return c_it
+end
 
 @everywhere function choice_part(
     V_i::Array{Float64,1}, 
@@ -210,7 +227,8 @@ end
     Nset::Array{Int64,1},
     alpha::Float64,
     epsilon::Float64,
-    beta::Float64
+    beta::Float64,
+    use_thompson::Bool
 )
     C_iT::Array{Int64,1} = []
     R_iT::Array{Int64,1} = []
@@ -228,8 +246,11 @@ end
         mu_Uit = mu_Vit + beta * cur_V
         # make choice
         ce_Uit = certainty_equivalent(alpha, mu_Uit, Sigma_Vit) # γ: uncertainty aversion
+
         c_it = nothing
-        if rand() < epsilon
+        if use_thompson
+            c_it = thompson_sampling(mu_V_i, Sigma_V_i)
+        elseif rand() < epsilon
             c_it = rand(choice_set)
         else
             c_it = choice_helper(ce_Uit, choice_set)
@@ -260,7 +281,9 @@ end
 			N::Int64, 
 			Nset::Array{Int64,1}, 
 			alpha::Float64, 
-			epsilon::Float64)
+			epsilon::Float64,
+                        use_thompson::Bool
+                        )
 
     C_iT::Array{Int64,1} = []
     for t=1:T
@@ -274,8 +297,11 @@ end
         # make choice
         ce_Uit = certainty_equivalent(alpha, mu_Uit, Sigma_Uit)
         choice_set = [n for n in Nset if n ∉ C_iT]
+        
         c_it = 0
-        if rand() < epsilon
+        if use_thompson
+            c_it = thompson_sampling(mu_U_i, Sigma_U_i)
+        elseif rand() < epsilon
[…]

### 2020-01-21 simulation_code/test.jl

@@ -1,0 +1,66 @@
+
+using Distributions;
+using Distances;
+using Random;
+
+
+function hop_distance(
+        i::Int64,
+        j::Int64,
+        N::Int64
+    )::Int64
+    return min(abs(i-j), abs(j-i), abs(j-i-N), abs(i-j-N))
+end
+
+function cov_mat_fun(sigma::Float64, rho::Float64, N::Int64)::Array{Float64,2}
+    cov_mat = zeros(Float64, N, N)
+    for i in 1:N 
+        for j in 1:N
+            dist = hop_distance( i, j , N)
+            #dist = Distances.euclidean( iota(i,N), iota(j,N) )
+            cov_mat[i,j] = sigma * rho^dist
+        end
+    end
+    return cov_mat
+end
+
+function thompson_sampling(
+    mu_V::Array{Float64,1},
+    Sigma_V::Array{Float64,2}
+)
+
+
+    draws = [
+            rand(
+               Normal(
+                    mu_V[ii],
+                    Sigma_V[ii,ii]
+                )
+            )  for ii in 1:length(mu_V) ]
+    c_it = argmax(draws)
+
+    @show draws
+    @show c_it
+
+    return c_it
+end
+
+
+
+seed = 1
+Random.seed!(seed);
+
+N = 20
+sigma = 0.25
+rho = 0.1
+
+Sigma_V = cov_mat_fun(sigma,rho,N)
+Nset = [ n for n=1:N]   # set of N items I = {1, ..., N}
+mu_V = zeros(Float64, N)
+V = rand(MvNormal(mu_V, Sigma_V))
+
+@show typeof(mu_V)
+@show mu_V[1]
+@show Sigma_V[1,1]
+
+thompson_sampling(mu_V, Sigma_V)

### 2020-01-21 simulation_code/simulations.jl
dumb errors

@@ -199,7 +199,7 @@ end
     return cit
 end 
 
-function thompson_sampling(
+@everywhere function thompson_sampling(
     mu_V::Array{Float64,1},
     Sigma_V::Array{Float64,2}
 )
@@ -403,7 +403,7 @@ end
     return Dict( "Consumption" => C_pop, "Welfare" => W_pop, "Rec" => R_pop )
 end
 
-use_thompson::Bool = true
+use_thompson = true
 #
 nr_pop = 100
 #

### 2020-01-22 simulation_code/simulations.jl
is this right?

@@ -249,7 +249,10 @@ end
 
         c_it = nothing
         if use_thompson
-            c_it = thompson_sampling(mu_V_i, Sigma_V_i)
+            c_it = thompson_sampling(copy(mu_Vit), copy(Sigma_Vit))
+            @show c_it
+            @show mu_V_i
+            @show choice_set
         elseif rand() < epsilon
             c_it = rand(choice_set)
         else
@@ -300,7 +303,7 @@ end
         
         c_it = 0
         if use_thompson
-            c_it = thompson_sampling(mu_U_i, Sigma_U_i)
+            c_it = thompson_sampling(copy(mu_Uit), copy(Sigma_Uit))
         elseif rand() < epsilon
             c_it = rand(choice_set)
         else
@@ -424,7 +427,7 @@ rho_vals = [0.1, 0.3, 0.5, 0.7, 0.9]
 beta_vals = [0.4, 0.8, 1., 2., 5.]
 
 # absolute risk aversion
-alpha_vals = [0.3, 0.6, 1., 5.]
+alpha_vals = [0.0]
 
 # action of the time for random exploration
 epsilon_vals = [0.0]

### 2020-01-22 simulation_code/simulations.jl
remove debug sludge

@@ -250,9 +250,6 @@ end
         c_it = nothing
         if use_thompson
             c_it = thompson_sampling(copy(mu_Vit), copy(Sigma_Vit))
-            @show c_it
-            @show mu_V_i
-            @show choice_set
         elseif rand() < epsilon
             c_it = rand(choice_set)
         else

### 2020-01-22 simulation_code/simulations.jl
attempt to fix errors with thompson (#19)
* is this right?

* remove debug sludge

@@ -249,7 +249,7 @@ end
 
         c_it = nothing
         if use_thompson
-            c_it = thompson_sampling(mu_V_i, Sigma_V_i)
+            c_it = thompson_sampling(copy(mu_Vit), copy(Sigma_Vit))
         elseif rand() < epsilon
             c_it = rand(choice_set)
         else
@@ -300,7 +300,7 @@ end
         
         c_it = 0
         if use_thompson
-            c_it = thompson_sampling(mu_U_i, Sigma_U_i)
+            c_it = thompson_sampling(copy(mu_Uit), copy(Sigma_Uit))
         elseif rand() < epsilon
             c_it = rand(choice_set)
         else
@@ -424,7 +424,7 @@ rho_vals = [0.1, 0.3, 0.5, 0.7, 0.9]
 beta_vals = [0.4, 0.8, 1., 2., 5.]
 
 # absolute risk aversion
-alpha_vals = [0.3, 0.6, 1., 5.]
+alpha_vals = [0.0]
 
 # action of the time for random exploration
 epsilon_vals = [0.0]

### 2020-01-22 simulation_code/simulations.jl
final fix

@@ -249,7 +249,7 @@ end
 
         c_it = nothing
         if use_thompson
-            c_it = thompson_sampling(mu_V_i, Sigma_V_i)
+            c_it = choice_set[thompson_sampling(mu_V_i, Sigma_V_i)]
         elseif rand() < epsilon
             c_it = rand(choice_set)
         else
@@ -300,7 +300,7 @@ end
         
         c_it = 0
         if use_thompson
-            c_it = thompson_sampling(mu_U_i, Sigma_U_i)
+            c_it = choice_set[thompson_sampling(mu_U_i, Sigma_U_i)]
         elseif rand() < epsilon
             c_it = rand(choice_set)
         else

### 2020-01-22 simulation_code/simulations.jl
final fix (#20)

@@ -249,7 +249,7 @@ end
 
         c_it = nothing
         if use_thompson
-            c_it = thompson_sampling(copy(mu_Vit), copy(Sigma_Vit))
+            c_it = choice_set[thompson_sampling(mu_V_i, Sigma_V_i)]
         elseif rand() < epsilon
             c_it = rand(choice_set)
         else
@@ -300,7 +300,7 @@ end
         
         c_it = 0
         if use_thompson
-            c_it = thompson_sampling(copy(mu_Uit), copy(Sigma_Uit))
+            c_it = choice_set[thompson_sampling(mu_U_i, Sigma_U_i)]
         elseif rand() < epsilon
             c_it = rand(choice_set)
         else

### 2020-01-22 simulation_code/dump_to_dataset.py
remove dependncey

@@ -7,7 +7,6 @@ import numpy as np
 
 import datetime
 from scipy.spatial.distance import jaccard, euclidean 
-from simulations import iota
 
 def get_files(root_path):
   dat_files = set()

### 2020-01-23 simulation_code/simulations.jl
fix parameters

@@ -418,10 +418,10 @@ N_vals = [200]
 T_vals = [20]
 
 # Covariance structure
-rho_vals = [0.1, 0.3, 0.5, 0.7, 0.9]
+rho_vals = [0.0, 0.1, 0.3, 0.5, 0.7, 0.9]
 
 # utility idiosyncratic degree 
-beta_vals = [0.4, 0.8, 1., 2., 5.]
+beta_vals = [0.0, 0.4, 0.8, 1., 2., 5.]
 
 # absolute risk aversion
 alpha_vals = [0.0]

### 2020-01-25 paper/filter_bubbles.tex
updates to paper

@@
 We explore the extent to which two issues traditionally associated with recommender systems -- \textit{filter bubbles} and \textit{user homogenization} -- can be attributed to recommendation systems.
-The key component of our model is that users face a sequential decision problem under uncertainty and that there are informational spillovers where a user consuming a product and learning how she values it reduces uncertainty and provides information about similar products.
+In our model users face a sequential decision problem under uncertainty and there are informational spillovers where a user consuming a product and learning how she values it reduces uncertainty and provides information about similar products.
 We find that this generates behavior consistent with filter bubble effects -- where users consume increasingly similar items over time -- and that recommendation can help alleviate these effects, as has been found in the empirical work of Nguyen et al.
@@
 For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
-A common remedy by online platforms has been to deploy \textit{recommender systems}, which have become critical for assisting users in making consumption decisions by providing users with information and recommendations to allow them to make more informed consumption decisions.
+A common remedy introduced by online platforms has been to deploy \textit{recommender systems}, which have become critical for assisting users in making consumption decisions.
+These systems provide users with information and recommendations to allow them to make more informed consumption decisions.
 These systems have been shown to be influential in steering consumer choice with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{MacKenzie et al.
@@
 Despite evidence of the power of these systems in determining user choice, we have little theoretical understanding of how users make choices in these markets and how recommender systems shape these choices.
-Understanding this has become increasingly important as there have emerged a number of concerns regarding the broader social and economic consequences of such systems.
-For instance, there have been claims that YouTube's recommendation algorithm unintentionally led to the radicalization of many individuals,\footnote{Friedersdorf (2018, Mar.
+Understanding this has become increasingly important as a number of concerns have emerged regarding the broader social and economic consequences of such systems.
+There have been claims that YouTube's recommendation algorithm unintentionally led to the radicalization of many individuals,\footnote{Friedersdorf (2018, Mar.
 12), YouTube Extremism and the Long Tail,
@@
 \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
-Further, an emerging worry is that recommender systems can be utilized as a tool of market power by steering consumer choice towards products more profitable for online platforms to the detriment of consumer welfare \citep{aridor2020recommender, bourreau2018streaming, DeCorniereTaylor2019RAND}.
+Furthermore, an emerging worry is that recommender systems can be utilized to steer consumer choice towards products more profitable for online platforms to the detriment of consumer welfare and the overall competitiveness of such markets \citep{aridor2020recommender, bourreau2018streaming, DeCorniereTaylor2019RAND}.
 Finally, there has emerged a burgeoning literature that focuses on understanding how recommender systems themselves can acquire information about product quality from the choices of self-interested consumers \cite{kremer2014implementing, che2017recommender,frazier2014incentivizing,mansour2015bayesian}.
 \par
-While these issues are important, it is hard to make progress on them without a model of user choice with foundations in economic decision theory and psychology.
-It is difficult to understand to what extent issues such as filter bubbles and homogenization are a consequence of recommender systems or a natural consequence of consumer decision-making in these contexts without such analysis.
-Further, it is challenging to understand how recommender systems can be used anti-competitively without understanding how such steering can influence user behavior.
-Finally, a major open question in the literature on incentivizing exploration in recommender systems is how to design recommendation policies in the context of realistic user behavior \cite{immorlica2018incentivizing}, which requires an understanding of what models of user decision-making are appropriate.
+While these issues are important and wide-reaching, it is hard to make progress on them without a model of user choice with foundations in economic decision theory and psychology.
+In this paper we provide such a model of user preferences and choice in these markets and study how the introduction of recommendation affects user choice.
+Our primary focus is on utilizing our model to understand the extent to which recommender systems lead individual users into filter bubbles and homogenize across user consumption choices.
+We rationalize existing empirical work on these topics and provide an argument for the underlying mechanisms behind such effects.
+We also discuss how our model can shed light on how recommender systems can be utilized to steer choice and affect the long-term consumption patterns of users.
+Finally, our model provides a direction for making progress on an open question in the literature on incentivizing exploration in recommender systems on how to incorporate realistic models of user choice into such models (e.g.
+\citep{immorlica2018incentivizing}).
 
-\xhdr{Our Model} In this paper we study a model of consumer decision-making in these markets via means of simulation and ask how a stylized model of recommendation can shape the behavior of users.
+\xhdr{Our Model} We consider a model of user choice that has the following components.
 \par
-The first component of our model is that users sequentially consume items and face large choice sets, though they only consume a small fraction of this choice set over their lifetime.
+The first component of our model is that users sequentially consume items and face large choice sets.
+In our setting of interest users are long-lived, but they only consume a small fraction of this choice set over their lifetime.
 \par
@@
 This valuation provides information for consumers that enables them to update their beliefs about similar items.
-This exploits the fact that similar items have more correlated true valuations, which assists users in making decisions in a vast product space.
+This exploits the fact that the valuations of similar items are correlated which assists users in navigating the vast product space.
+The idea that users make similarity-based assessments to guide their choice has grounding in theoretical work in decision theory \cite{gilboa1995case} and empirical evidence of how users navigate large choice sets \cite{schulz2019structured}.
 \par 
-The fourth component of our model is that recommendation provides users with uncertainty about the true valuations.
+The fourth component of our model is that recommendation provides users with information about the true valuations.
 We model the realized valuations as being a weighted sum of a common-value and an idiosyncratic component.
 This formulation gives a stylized notion of predictability of user preferences where the idiosyncratic component is inherently unpredictable given other individuals' preferences and the common-value component is what the recommender can learn from previous users' data.
-The recommender uses its knowledge of the common-value component and the individuals' beliefs over the product space when designing personalized recommendation.
+We suppose that the recommender knows the common-value component for each item and combines it with individuals' beliefs over the product space when designing personalized recommendation.
 
@@
 The standard notion of filter bubbles is that, over time, user consumption becomes increasingly concentrated on narrower and narrower parts of the product space, leading to a decreasing distance between subsequent consumed items over time.
-However, empirical work, such as \cite{nguyen2014exploring}, finds that user consumption choices over time naturally exhibit such filter bubble effects even without recommendation and that, in fact, recommender systems mitigate such effects.
+However, empirical work such as \cite{nguyen2014exploring}, finds that user consumption choices over time naturally exhibit such filter bubble effects even without recommendation and that, in fact, recommender systems mitigate such effects.
 The results of our model are in line with these empirical results and give a clear intuition as to why such effects arise and why recommendation mitigates them to some degree.
 \par
-We find that, even if there were no uncertainty about valuations, if there is no inherent correlation between the true valuations of the items then such filter bubble effects never arise, even in the optimal consumption path.
-However, once there is a correlation between the true valuations of the items then such filter bubble effects arise even without recommendation.\footnote{The correlation between the true valuation of items in the product space is a natural assumption and, indeed, is the core idea behind standard content-based recommendation methods.
-These methods determine recommendations based on finding similar items to those that a user has consumed before and enjoyed \citep{adomavicius2005toward}.} The intuition is that, users initially have uncertainty about the true valuations of the set of items and when they begin consuming items this impacts the beliefs of the average utility of nearby items.
-When a user finds an item that has high realized utility, this induces them to update their beliefs positively about the realized utility of nearby items which leads to them consuming these items.
+We first find that when there is no correlation between the valuations of goods there is no reason for filter bubble effects to arise, regardless of the information that users have on the valuations of the products.
+However, once there is a correlation between the true valuations of the items then such filter bubble effects arise even without recommendation.\footnote{The fact that in many environments where recommender systems are deployed the true valuations of items are correlated makes this the most relevant case to consider.
+Indeed, the core idea behind standard content-based recommendation methods relies on this fact as these methods determine recommendations based on finding similar items to those that a user has consumed before and enjoyed \citep{adomavicius2005toward}.} The intuition is that, users initially have uncertainty about their true valuations and initial consumption of items leads them to update their beliefs strongly about the average utility of nearby items.
+When a user finds an item that has high realized utility, this induces her to update her beliefs positively about the realized utility of nearby items which leads to  the consumption of these items.
 Thus, without recommendation users increasingly consume items in this portion of the vast product space since they never get information about items in different parts of the product space.
 This learning spillover not only induces positive updates about the average utility of the item, but also reduces the level of uncertainty they have about these items.
-Thus, this effect is further amplified when users are \textit{risk-averse}, a concept from decision theory where all else being equal, users have a preference for items with lower uncertainty to those with higher uncertainty.
+Thus, this effect is further amplified when users are \textit{risk-averse}.
+This is a standard concept from decision theory by which, all else being equal, users have a preference for items with lower uncertainty to those with higher uncertainty.
 \par 
@@
 Recommendations could potentially make matters worse: providing information about the common value of the items when this value is correlated could intensify the extent of local consumption.
-However, we find the opposite - that recommendation partially mitigates this narrowing effect - since the revelation of the common value component allows for users to escape being stuck in portions of the product space due to lack of information about other portions of the product space.
+However, in line with existing empirical work, we find the opposite - that recommendation partially mitigates this narrowing effect - since the revelation of the common value component allows for users to escape being stuck in portions of the product space due to lack of information about other portions of the product space.
 \par
 \xhdr{Additional Findings} We further explore the diversity of the overall set of items that users consume.
-Consumption diversity is increasingly a common goal of many recommender systems \cite{castells2015novelty, kunaver2017diversity}, which is reasonable since it could be expected that higher diversity is associated with more intense search and therefore higher welfare.
-However, we find that, without recommendation, higher diversity of items does not ensure higher welfare.
+Encouraging consumption diversity is increasingly a common goal of many recommender systems \cite{castells2015novelty, kunaver2017diversity}, which is reasonable since it could be expected that higher diversity is associated with more intense search and therefore higher welfare.
+However, we find that, without recommendation, higher diversity of items consumed does not ensure higher welfare.
 This is due to the fact that high consumption set diversity in our model can arise from a sequence of bad realizations, which leads to very intense search over very different parts of the product space and high consumption diversity but very low welfare.
@@
 We then turn to the extent to which recommendation leads to increasing user homogenization.
+In our context we define homogenization as the degree of similarity between the set of consumed items across individuals.
 We find that, without recommendation, there is very little homogenization as users are effectively all doing their own random exploration around the product space.
@@
 One implication from our model is that the user consumption paths have a clear path-dependence since the consumption of an item today influences the information a user has tomorrow which further influences consumption choices tomorrow.
-This is important for the debate regarding the extent to which recommender systems can be utilized as tools of market power since most of the literature analyzes this in a static context.
+This is important for the debate regarding the extent to which recommender systems can be utilized to steer consume choice as most of the literature analyzes this in a static context.
 However, this path-dependence implies that such platform steering can have dynamic consequences as steering towards certain items today can shape the long-run consumption choices of users.
@@
 \xhdr{Related Work.} 
-\cite{pariser2011filter} first informally described the ``filter bubble'' which is the idea that online personalization services would lead users down paths of increasingly narrower content so that they would effectively be isolated from a diversity of viewpoints or content.
-A number of studies, done in a wide range of disciplines, have since empirically studied whether or not this phenomenon exists in various contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}.
+The first set of related works studies the extent and implications of filter bubbles.
+\cite{pariser2011filter} first informally described the idea of the ``filter bubble'' which is that online personalization services would lead users down paths of increasingly narrower content so that they would effectively be isolated from a diversity of viewpoints or content.
+Following this a number of empirical studies, done in various disciplines, have since studied the extent to which this phenomena exists in a wide range of contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}.
 The most relevant to our study is \cite{nguyen2014exploring} who study whether this effect exists with recommender systems in the context of movie consumption.
 They find that even users whose consumption choices are not guided by recommendations fall into behavior consistent with ``filter bubbles'' and that recommender systems can actually increase the diversity of the content that users consume.
-To our knowledge there are no theoretical models that rationalize the empirical findings but predominantly empirical studies.
-Our paper provides a theoretical framework through which to view this problem.
-The second question is to what extent recommender systems can lead users to become increasingly homogenized or consume similar content as each other.
+To our knowledge there are no theoretical models that rationalize these empirical findings and our paper provides a theoretical framework through which to view this problem.
+Further, our paper provides a clear mechanism at play that drives such effects and how recommendation interacts with them.
+\par 
+Another set of papers have examined whether recommender systems can lead users to become increasingly homogenized or consume similar content as each other.
 \cite{celma2008hits, treviranus2009value} show that incorporating content popularity into recommender systems can lead to increased user homogenization.
-\cite{chaney2018algorithmic} more generally look at the issue of user homogenization in recommender systems.
-Our results are similar to previous work, where the limits of recommendation lead the recommender to over-recommend products from certain portions of the product space which leads users to consume items in those regions.
-\cite{fleder2009blockbuster} also consider the effects of recommendation on multi-period user behavior, but predominantly focus on sales diversity and employ a different model of user behavior than we do here.
-\cite{hodgson2019horse} consider a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of their preferences in the environment, but consider it in the context of search for a single item.
+\cite{chaney2018algorithmic} show how user homogenization may arise from training recommender systems on data from users exposed to algorithmic recommendations.
+\cite{fleder2009blockbuster} show that homogenization can increase due to a recommendation bias towards popularity that arises from lack of information about items with limited consumption histories.
+Our model leads to similar results as previous theoretical and empirical work where recommender systems lead to increased user homogenization.
+However, the mechanisms behind this from our model differ from existing work as homogenization arises due to the fact that recommendation leads users to coordinate their consumption decisions in certain portions of the product space.
+\par 
+Our model of decision-making draws from work in decision theory and psychology.
+\cite{schulz2019structured} empirically studies how individuals solve sequential decision-making problems under uncertainty in large choice sets in the context of mobile food delivery orders.
+They find that individuals engage in similarity-based generalizations where learning about the realized value of a particular good provides them with information about similar items.
+We incorporate this finding into our model in a stylized manner where a single parameter controls the degree of information spillovers.
+\cite{gilboa1995case} considers a model of case-based decision making where choices are made based on information from previously experienced choice problems that are similar to the one faced by the decision-maker.
+However, while conceptually related, our model incorporates this idea in a different manner by utilizing Bayesian reasoning as opposed to a weighted sum of previous similar choice problems.
+\cite{hodgson2019horse} consider a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of their preferences in the environment, but consider it in the context of search for a single item as opposed to consumption of multiple items.
+\cite{bronnenberg2016zooming} empirically show that users engage in a ``zooming in” search strategy in terms of the information they acquire about products before purchasing and only end up acquiring information about a small set of products in a large product space.
+Finally, \cite{chen2013human} surveys work that has studied decision-making models in the context of recommender systems, though to our knowledge none of these papers study user decision making over time or study filter bubble and homogenization issues.
 
@@
 \subsection{Preliminaries on Expected Utility Theory}
-\noindent For every item $n$ in the product space $\mathcal J=\{0,1,....,N-1\}$, we assume that each user $i$ assigns a monetary equivalent $x_{i,n} \in \mathbb R$ to the experience of consuming it.
-Each consumer can value the same item differently.
+\noindent For every item $n$ in the product space $\mathcal J$, we assume that each user $i$ assigns a monetary equivalent $x_{i,n} \in \mathbb R$ to the experience of consuming it.
+Each user can value the same item differently.
 However, we assume that users have the same utility over money, given by a utility function $u: \mathbb R \to \mathbb R$, strictly increasing and continuous.
@@
 Therefore, a more risk-averse user is more willing to avoid the risk of taking the gamble.
-We assume that the utility function takes a flexible functional form $u(x)=1-\exp(-\gamma x)$ for $\gamma\ne0$ and $u(x)=x$ for $\gamma=0$ -- known as constant absolute risk-aversion preferences.
+We assume that the utility function takes a flexible functional form $u(x)=1-\exp(-\gamma x)$ for $\gamma\ne0$ and $u(x)=x$ for $\gamma=0$ -- known as constant absolute risk-aversion preferences (from hereon CARA).
 Higher $\gamma$ implies higher risk-aversion with $\gamma=0$ corresponding to the risk-neutral case and $\gamma>0$ to the risk-averse one.
@@
 \xhdr{User Learning}
-When a user consumes a item $n$ they learn the realized value for that item. In our model, we incorporate the idea that learning the value of item $n$ gives a user information about similar items. 
+When a user consumes an item $n$ she learns the realized value for that item. In our model, we incorporate the idea that learning the value of item $n$ gives a user information about similar items. 
 %This is drawn from recent empirical evidence in \cite{schulz2019structured} that consumers learn how to navigate large choice sets using similarity-based generalization. 
-We assume that learning about the value of item $n$ reveals more about the value associated to items that are closer to it, which captures the idea that trying a item provides more information about similar items than about dissimilar ones.
+We assume that learning about the value of item $n$ reveals more about the value associated to items that are closer to it, which captures the idea that trying an item provides more information about similar items than about dissimilar ones.
 \par
@@
 The underlying correlation structure implies that upon observing that $x_0=y$ the user will update beliefs about the remaining 3 items according to the previously specified updating rule.
-For concreteness, we suppose that $\sigma = 1$, $y = 0.5$, and $\rho = 0.5$.
+For concreteness, we suppose that $\sigma = 1$ and $\rho = 0.5$, but the intuitions hold for any value of $\sigma$ and $\rho > 0$.
+First, we consider the case when $y = 0.5 > 0$.
 The resulting beliefs after observing $y$ are then as follows:
@@
 \end{array}  \right) =\left (\begin{array}{c}
-\rho y = \frac{1}{4} \vspace{0.15cm} \\
-\rho^{2} y = \frac{1}{8} \vspace{0.15cm} \\
- \rho y = \frac{1}{4} \\
+\rho y  \vspace{0.15cm} \\
+\rho^{2} y  \vspace{0.15cm} \\
+ \rho y \\
+\end{array} \right) =
+\left (\begin{array}{c}
+\frac{1}{4} \vspace{0.15cm} \\
+\frac{1}{8}  \vspace{0.15cm} \\
+\frac{1}{4}
 \end{array}  \right), \bar{\Sigma} =  \left( \begin{array}{ccc}
@@
 We will consider three cases.
-The case of primary interest is \textit{recommendation} where the recommender observes values accrued and knows $V$ but does not know neither $V_i$.\footnote{We do not consider the acquisition of information for the recommender to know $V$ and suppose that she has sufficient data to learn $V$ with arbitrary precision.} However, the recommender does know the users' beliefs $\bar V_i$.
+The case of primary interest is \textit{recommendation} where the recommender observes values accrued and knows $V$ but does not know neither $V_i$.\footnote{We do not consider the acquisition of information for the recommender to know $V$ and suppose that she has sufficient data to learn $V$ with arbitrary precision at $t = 0$.} However, the recommender does know the users' beliefs $\bar V_i$.
 Thus, at any given period, the recommender provides a personalized recommendation that combines the knowledge of the common value component $V$ with the user beliefs $\bar V_i$.
@@
 When we consider results varying a single parameter, we group the results over the other parameters and provide reports varying only the parameter of interest.
-We report results for a relatively small consumption history $T=20$ with a produce space size $N=200$.
+We report results for a relatively small consumption history $T=20$ with a product space size $N=200$.
 Unless otherwise reported, our results are robust to varying $N$.
-
-\subsection{Discussion of the Assumptions}
-In this section we provide further discussion of several assumptions of our model.
 \par
@@
 \end{figure}
-
-\xhdr{Uncertainty}: We assume that users have underlying uncertainty about the realized value of the items in their choice sets and as a result face a decision problem under uncertainty.
-The first justification for this is that environments where recommender systems are commonly deployed, such as on media streaming and e-commerce platforms, are mostly dedicated to experience goods, where users only learn their value after consumption.
-Standard decision theory axiomatized the conditions under which behavior is described by (subjective) expected utility maximizers \cite{savage1954} who update beliefs according to Bayesian inference \cite{ghirardato2002revisiting}, to which we add the assumption that their prior is correct.
-The second justification is that, even if users are in an environment where they could observe utilities based on product reviews or introspection, it is prohibitively costly for them to do so in large choice sets.\footnote{Recent empirical work \cite{bronnenberg2016zooming,hodgson2019horse} has shown that, for product search, users engage in a ``zooming in'' search strategy in terms of the information they acquire about products before purchasing and only end up acquiring information about a small set of products in a large product space}  Even if they acquired all the information from product reviews, this would not fully resolve uncertainty as to the extent that individuals have idiosyncratic tastes.
-Giving that users are facing choices problems uncertainty, this motivates modeling recommendation as providing users \textit{information} in order to reduce their uncertainty.
-\par
-
-\xhdr{Large Choice Sets}: We are interested in environments where the number of items in the choice set is large and users only consume a small number of these items.
-This is realistic for platforms such as Netflix, Spotify, Amazon and YouTube where there are sometimes billions of items and users can realistically only consume a handful of these items in their lifetime.\par
-
-\xhdr{Learning Spillovers}: Users in our environment face a sequential decision-making problem.
-However, we suppose that individuals are myopic, i.e.
-they consume the item with the highest expected utility given their beliefs at each period.
-This is assumed in order to preserve tractability.
-The predominant difficulty in the learning problem they face is not that they need to repeatedly sample a single item in order to learn its value as in the standard multi-armed bandit formulations, but rather that they need to be able to generalize from their experience from consuming one item to other items.
-After consuming a single item, they learn the realized value of the good, but have no desire to consume it again, i.e.
-we assume that there is no value to consuming the item again.
-\par
-
-\cite{schulz2019structured} studies how individuals solve sequential decision-making problems under uncertainty in large choice sets in the context of mobile food delivery orders.
-They find that individuals engage in similarity-based generalizations where learning about the realized value of a particular good provides them with information about similar items.
-Content-based recommender systems exploit a similar idea where, if a user liked a given item, then she is also likely to like a similar but different item as well.
-\par
-
-Our model incorporates this observation in a stylized fashion, where the degree of information spillovers is controlled via parameter $\rho$ and we study how the strength of the spillovers impacts behavior and the effect of recommendation.
-\par
 
@@
 When there is a correlation between the realized utilities then filter bubbles can naturally arise due to the nature of how individuals acquire additional information about the remaining items.
-We have shown that, unless users are provided with additional information to guide their consumption choices, then these learning spillovers and user risk-aversion can lead users into filter bubbles.
+We have shown that, unless users are provided with additional information to guide their consumption choices, then these information spillovers and user risk-aversion can lead users into filter bubbles.
 When users consume high valued items, they exploit the underlying correlation across different products' values, stronger for similar products, which leads them to increasingly consume products in narrower and narrower portions of the product space.
@@
 As before, when there is no correlation between valuations, product diversity is the same across different recommendation regimes.
-The over-exploitation of learning spillovers when $\rho > 0$ leads to product diversity being lowest in the no recommendation regime.
+The over-exploitation of information spillovers when $\rho > 0$ leads to product diversity being lowest in the no recommendation regime.
 As a result, this effect gets amplified as $\rho$ increases, which leads to the gap in diversity between the regimes to increase as $\rho$ increases.
@@
 If instead the user valued highly the item that she had consumed, then she is more likely to pick a nearby item.
-The learning spillovers therefore lead to high product diversity being negatively correlated with welfare.
+The information spillovers therefore lead to high product diversity being negatively correlated with welfare.
 \par

### 2020-01-25 paper/filter_bubbles.tex
derp

@@
 We argue that these parameters are representative.
-First, the values of $\gamma$ that we consider range from no risk aversion ($\gamma = 0$) to a very high risk aversion level under CARA preferences $(\gamma = 5)$.
-The correlation parameter, $\rho$, ranges from no correlation ($\rho = 0$) to nearly full correlation $(\rho = 0.9)$.
+First, the values of $\gamma$ that we consider range from no risk aversion ($\gamma = 0$) to reasonably high risk aversion levels $(\gamma = 5)$.
+The correlation parameter, $\rho$, ranges from uncorrelated valuations ($\rho = 0$) to nearly fully correlated valuations $(\rho = 0.9)$.
 $\beta$ ranges from the common value component playing no role $(\beta = 0)$ to the idiosyncratic component dominating the true valuations $(\beta = 5)$.

### 2020-01-25 paper/filter_bubbles.tex
typos

@@
 Risk aversion captures the how different users react to the risk associated to a particular consumption opportunity.
-It is formalized as follows: Suppose for a given gamble $x$ taking real values and following distribution $p$, there is a certain amount of money that leaves the user indifferent between taking the gamble and taking this sure amount, call it $\delta(x)$, termed the \textit{certainty equivalent} of $x$.
-A user $i$ is more risk-averse than another user$j$ when for every gamble $x$, user $i$ requires a lower value to give up the gamble when compared to user $j$, i.e.
+It is formalized as follows: a given gamble $x$ takes real values and follows distribution $p$.
+Then, for every gamble $x$, there is a certain amount of money that makes the user indifferent between taking the gamble or taking the sure amount of money.
+This sure amount of money is called the \textit{certainty equivalent} of gamble $x$ and is denoted as $\delta(x)$.
+A user $i$ is more risk-averse than another user $j$ when for every gamble $x$, user $i$ requires a lower value to give up the gamble when compared to user $j$, i.e.
 $\delta_i(x)$.
@@
 \xhdr{User Decision-Making.}
-We assume the user makes $T$ choices and therefore can only consume up to $T$ items, where $T$ is but a small fraction of $N$.
+We assume the user makes $T$ choices and therefore can only consume up to $T$ items, where $T$ is a small fraction of $N$.
 This captures the idea that users are faced with an immense choice set but that ultimately they end up experiencing (and learning) about just a small fraction of it.
@@
 The second is that it allows us to incorporate an easily interpretable correlation structure between the items.
+The precise formulation of $\Sigma$ and $\Sigma_i$ that we consider is defined below when we discuss user learning.
 \par
@@
 \subsection{User Homogenization}
-In this section we focus on across user comparisons and investigate how the consumed set of items across users varies across different recommendation regimes and parameter values.
+In this section, we focus on comparisons across users and investigate how the consumed set of items across users varies across different recommendation regimes and parameter values.
 In particular we look at the degree of \textit{homogenization} between users.

### 2020-01-26 paper/filter_bubbles.tex
fixes

@@
 One implication from our model is that the user consumption paths have a clear path-dependence since the consumption of an item today influences the information a user has tomorrow which further influences consumption choices tomorrow.
-This is important for the debate regarding the extent to which recommender systems can be utilized to steer consume choice as most of the literature analyzes this in a static context.
+This is important for the debate regarding the extent to which recommender systems can be utilized to steer consumer choice as most of the literature analyzes this in a static context.
 However, this path-dependence implies that such platform steering can have dynamic consequences as steering towards certain items today can shape the long-run consumption choices of users.
@@
 Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{i,n}\right)_{n \in \mathcal{J}}$ are drawn independently from a jointly normal distribution, where $\overline v_{i,n} \sim \mathcal N (0, \overline \sigma^2)$ are independent and identically distributed.
-These $\overline v_{i,n}$ denote the prior belief that individual $i$ holds about the her valuation over item $n$.
+These $\overline v_{i,n}$ denote the prior belief that individual $i$ holds about her valuation over item $n$.
 As people are exposed to different backgrounds, their beliefs about what is good for them also varies and $\overline v_{i,n}$ denotes this idiosyncrasy at the level of prior beliefs.
@@
 When there is no risk-aversion then there is a negative correlation between welfare and diversity.
-This is since, with no risk-aversion, in a given period a user will select the good that she currently believe has the highest expected value.
+This is since, with no risk-aversion, in a given period a user will select the good that she currently believes has the highest expected value.
 High product diversity in this case can arise from a user who consumes an item she disliked and updates her beliefs about nearby items negatively.
@@
 Thus, online platforms can potentially shape the long-term preferences of users and not only impact their current period choices.
-This has broader ramifications about the equilibrium effects of such behavior on competition and consumer welfare in online platform which we leave to future work.
+This has broader ramifications about the equilibrium effects of such behavior on competition and consumer welfare in online platforms which we leave to future work.
 \par

### 2020-01-26 paper/filter_bubbles.tex
updates

@@
 % Anonymized submission.
-\author{Submission 42?}
+\author{Submission 20}
 
@@
 \begin{abstract}
-We study a model of user decision-making in the context of recommender systems.
+We study a model of user decision-making in the context of recommender systems via numerical simulation.
 We explore the extent to which two issues traditionally associated with recommender systems -- \textit{filter bubbles} and \textit{user homogenization} -- can be attributed to recommendation systems.
@@
 Suppose instead that item 0 ends up having a negative valuation so that $y = -0.5 < 0$.
-Now we have that $\mathbb{E}[x_1] = \mathbb{E}[x_3] = -\frac{1}{4} <  -\frac{1}{8} = \mathbb{E}[x_2]$ with $\bar{\Sigma}$ remaining the same as when $y = 0.5$.
-Now, the risk-aversion levels of the user determine the choice in the next period.
+This results in $\mathbb{E}[x_1] = \mathbb{E}[x_3] = -\frac{1}{4} <  -\frac{1}{8} = \mathbb{E}[x_2]$ with $\bar{\Sigma}$ remaining the same as when $y = 0.5$.
+In this case the risk-aversion levels of the user determine the choice in the next period.
 If the user is risk-neutral ($\gamma = 0$), then she will go across the product space to consume item $2$ in the next period since it has a higher expected value.

### 2020-01-26 simulation_code/simulations.jl

@@ -226,9 +226,7 @@ end
     N::Int64,
     Nset::Array{Int64,1},
     alpha::Float64,
-    epsilon::Float64,
-    beta::Float64,
-    use_thompson::Bool
+    beta::Float64
 )
     C_iT::Array{Int64,1} = []
     R_iT::Array{Int64,1} = []
@@ -247,14 +245,7 @@ end
         # make choice
         ce_Uit = certainty_equivalent(alpha, mu_Uit, Sigma_Vit) # γ: uncertainty aversion
 
-        c_it = nothing
-        if use_thompson
-            c_it = choice_set[thompson_sampling(mu_V_i, Sigma_V_i)]
-        elseif rand() < epsilon
-            c_it = rand(choice_set)
-        else
-            c_it = choice_helper(ce_Uit, choice_set)
-        end
+        c_it = choice_helper(ce_Uit, choice_set)
         r_it = choice_set[argmax([V[i] for i in choice_set])]
         append!(R_iT,r_it)
         append!(C_iT,c_it)
@@ -280,10 +271,8 @@ end
 			T::Int64, 
 			N::Int64, 
 			Nset::Array{Int64,1}, 
-			alpha::Float64, 
-			epsilon::Float64,
-                        use_thompson::Bool
-                        )
+			alpha::Float64
+)
 
     C_iT::Array{Int64,1} = []
     for t=1:T
@@ -297,15 +286,7 @@ end
         # make choice
         ce_Uit = certainty_equivalent(alpha, mu_Uit, Sigma_Uit)
         choice_set = [n for n in Nset if n ∉ C_iT]
-        
-        c_it = 0
-        if use_thompson
-            c_it = choice_set[thompson_sampling(mu_U_i, Sigma_U_i)]
-        elseif rand() < epsilon
-            c_it = rand(choice_set)
-        else
-            c_it = choice_helper(ce_Uit, choice_set)
-        end
+        c_it = choice_helper(ce_Uit, choice_set)
         append!(C_iT, c_it)
     end
 		
@@ -326,7 +307,6 @@ end
         Sigma_V::Array{Float64,2},
         Sigma_V_ibar::Array{Float64,2},
         alpha::Float64,
[…]

### 2020-01-27 paper/filter_bubbles.tex
typo corrections

@@
 \cite{chaney2018algorithmic} show how user homogenization may arise from training recommender systems on data from users exposed to algorithmic recommendations.
-\cite{fleder2009blockbuster} show that homogenization can increase due to a recommendation bias towards popularity that arises from lack of information about items with limited consumption histories.
+\cite{fleder2009blockbuster} argues that homogenization can increase due to a recommendation bias towards popularity that arises from lack of information about items with limited consumption histories.
 Our model leads to similar results as previous theoretical and empirical work where recommender systems lead to increased user homogenization.
@@
 \par
-\xhdr{User Beliefs.} We assume that all the true realized utilities are drawn at $t = 0$.
-However, users do not know the realized utilities before consuming an item, but rather have beliefs over the realized utilities.
+\xhdr{User Beliefs.} We assume that all the true realized values are drawn at $t = 0$.
+However, users do not know the realized values before consuming an item, but rather have beliefs over the realized utilities.
 Formally, user $i$ starts with some beliefs about $X_i$, namely that the idiosyncratic and common-value parts of the valuations are independent -- $V_i \perp \!\!\!
@@
 These $\overline v_{i,n}$ denote the prior belief that individual $i$ holds about her valuation over item $n$.
-As people are exposed to different backgrounds, their beliefs about what is good for them also varies and $\overline v_{i,n}$ denotes this idiosyncrasy at the level of prior beliefs.
+As people are exposed to different backgrounds, their beliefs about how much they value a given item also varies and $\overline v_{i,n}$ denotes this idiosyncrasy at the level of prior beliefs.
 \par
@@
 The particular distance function that we utilize leads to this covariance structure being simple, where the $(n,n+1)$-th entry in the covariance matrix is $\sigma^{2} \rho$, the $(n,n+2)$-th entry is $\sigma^{2} \rho^2$, etc.
+\footnote{This exponential decay correlation structure can be related to the tenet of case-based similiartiy of Gilboa and Schmeidler (2008, MSS) for an axiomatization of exponential similarity.}
 \par

### 2020-05-06 paper/appendix.tex
udpate

--- 2020-01-26 paper/filter_bubbles.tex
@@
 % Title. Note the optional short title for running heads. In the interest of anonymization, please do not include any acknowledgements.
-\newcommand\PaperTitle{Deconstructing the Filter Bubble: \\ Consumer Decision-Making and Recommender Systems}
+\newcommand\PaperTitle{Appendix For Deconstructing the Filter Bubble: \\ Consumer Decision-Making and Recommender Systems}
 \title[\PaperTitle]{\PaperTitle}
-
-% Anonymized submission.
-\author{Submission 42?}
-
-% Abstract. Note that this must come before \maketitle.
-\begin{abstract}
-We study a model of user decision-making in the context of recommender systems.
-We explore the extent to which two issues traditionally associated with recommender systems -- \textit{filter bubbles} and \textit{user homogenization} -- can be attributed to recommendation systems.
-In our model users face a sequential decision problem under uncertainty and there are informational spillovers where a user consuming a product and learning how she values it reduces uncertainty and provides information about similar products.
-We find that this generates behavior consistent with filter bubble effects -- where users consume increasingly similar items over time -- and that recommendation can help alleviate these effects, as has been found in the empirical work of Nguyen et al.
-(2014).
-However, we find that recommendation, while alleviating filter bubble effects, leads to an increase in user homogeneity.
-Our results highlight the path-dependence of user choice in environments where recommender systems are deployed.
-\end{abstract}
 
@@
 % Paper body
-\section{Introduction}
-
-Users in many online platforms face difficult decision problems as they have to sort through the vast choice sets present in these markets.
-For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
-A common remedy introduced by online platforms has been to deploy \textit{recommender systems}, which have become critical for assisting users in making consumption decisions.
-These systems provide users with information and recommendations to allow them to make more informed consumption decisions.
-These systems have been shown to be influential in steering consumer choice with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{MacKenzie et al.
-(2013, Oct.),  How retailers can keep up with consumers.
-\url{https://www.mckinsey.com/industries/retail/our-insights/how-retailers-can-keep-up-with-consumers}. 
-Retrieved on October 3, 2019.}
-\par 
-Despite evidence of the power of these systems in determining user choice, we have little theoretical understanding of how users make choices in these markets and how recommender systems shape these choices.
-Understanding this has become increasingly important as a number of concerns have emerged regarding the broader social and economic consequences of such systems.
-There have been claims that YouTube's recommendation algorithm unintentionally led to the radicalization of many individuals,\footnote{Friedersdorf (2018, Mar.
-12), YouTube Extremism and the Long Tail,
-\textit{The Atlantic}. 
-\url{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350}. Retrieved on October 3, 2019.} 
-that personalized recommender systems lead users into \textit{filter bubbles} 
-where they effectively get isolated from a diversity of viewpoints or content 
-\cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
-Furthermore, an emerging worry is that recommender systems can be utilized to steer consumer choice towards products more profitable for online platforms to the detriment of consumer welfare and the overall competitiveness of such markets \citep{aridor2020recommender, bourreau2018streaming, DeCorniereTaylor2019RAND}.
-Finally, there has emerged a burgeoning literature that focuses on understanding how recommender systems themselves can acquire information about product quality from the choices of self-interested consumers \cite{kremer2014implementing, che2017recommender,frazier2014incentivizing,mansour2015bayesian}.
-\par
-While these issues are important and wide-reaching, it is hard to make progress on them without a model of user choice with foundations in economic decision theory and psychology.
-In this paper we provide such a model of user preferences and choice in these markets and study how the introduction of recommendation affects user choice.
-Our primary focus is on utilizing our model to understand the extent to which recommender systems lead individual users into filter bubbles and homogenize across user consumption choices.
-We rationalize existing empirical work on these topics and provide an argument for the underlying mechanisms behind such effects.
-We also discuss how our model can shed light on how recommender systems can be utilized to steer choice and affect the long-term consumption patterns of users.
-Finally, our model provides a direction for making progress on an open question in the literature on incentivizing exploration in recommender systems on how to incorporate realistic models of user choice into such models (e.g.
-\citep{immorlica2018incentivizing}).
-
-\xhdr{Our Model} We consider a model of user choice that has the following components.
-\par
-The first component of our model is that users sequentially consume items and face large choice sets.
-In our setting of interest users are long-lived, but they only consume a small fraction of this choice set over their lifetime.
-\par
-The second component of our model is that users have \textit{uncertainty} about their valuation of the different items.
-This is motivated both by the fact that many markets where recommender systems are deployed contain experience goods, whose true value can only be learned after consumption, and the fact that such uncertainty is why recommender systems exist in the first place.
-Thus, users face a sequential decision-making problem under uncertainty.
-\par 
-The third, and key, component of our model is that consumption of an item reveals information that changes user beliefs about the valuation of similar items.
-Unlike in standard sequential decision-making problems, in our model once an item is consumed all uncertainty about its valuation is resolved.
-This valuation provides information for consumers that enables them to update their beliefs about similar items.
-This exploits the fact that the valuations of similar items are correlated which assists users in navigating the vast product space.
-The idea that users make similarity-based assessments to guide their choice has grounding in theoretical work in decision theory \cite{gilboa1995case} and empirical evidence of how users navigate large choice sets \cite{schulz2019structured}.
-\par 
-The fourth component of our model is that recommendation provides users with information about the true valuations.
-We model the realized valuations as being a weighted sum of a common-value and an idiosyncratic component.
-This formulation gives a stylized notion of predictability of user preferences where the idiosyncratic component is inherently unpredictable given other individuals' preferences and the common-value component is what the recommender can learn from previous users' data.
-We suppose that the recommender knows the common-value component for each item and combines it with individuals' beliefs over the product space when designing personalized recommendation.
-
-\xhdr{Main Findings}
-Our first set of results focuses on understanding to what extent recommender systems lead users into filter bubbles and how such filter bubbles may naturally arise.
-The standard notion of filter bubbles is that, over time, user consumption becomes increasingly concentrated on narrower and narrower parts of the product space, leading to a decreasing distance between subsequent consumed items over time.
-However, empirical work such as \cite{nguyen2014exploring}, finds that user consumption choices over time naturally exhibit such filter bubble effects even without recommendation and that, in fact, recommender systems mitigate such effects.
-The results of our model are in line with these empirical results and give a clear intuition as to why such effects arise and why recommendation mitigates them to some degree.
-\par
-We first find that when there is no correlation between the valuations of goods there is no reason for filter bubble effects to arise, regardless of the information that users have on the valuations of the products.
-However, once there is a correlation between the true valuations of the items then such filter bubble effects arise even without recommendation.\footnote{The fact that in many environments where recommender systems are deployed the true valuations of items are correlated makes this the most relevant case to consider.
-Indeed, the core idea behind standard content-based recommendation methods relies on this fact as these methods determine recommendations based on finding similar items to those that a user has consumed before and enjoyed \citep{adomavicius2005toward}.} The intuition is that, users initially have uncertainty about their true valuations and initial consumption of items leads them to update their beliefs strongly about the average utility of nearby items.
-When a user finds an item that has high realized utility, this induces her to update her beliefs positively about the realized utility of nearby items which leads to  the consumption of these items.
-Thus, without recommendation users increasingly consume items in this portion of the vast product space since they never get information about items in different parts of the product space.
-This learning spillover not only induces positive updates about the average utility of the item, but also reduces the level of uncertainty they have about these items.
-Thus, this effect is further amplified when users are \textit{risk-averse}.
-This is a standard concept from decision theory by which, all else being equal, users have a preference for items with lower uncertainty to those with higher uncertainty.
-\par 
-We then investigate whether the introduction of recommendation either mitigates or amplifies this effect.
-Recommendations could potentially make matters worse: providing information about the common value of the items when this value is correlated could intensify the extent of local consumption.
-However, in line with existing empirical work, we find the opposite - that recommendation partially mitigates this narrowing effect - since the revelation of the common value component allows for users to escape being stuck in portions of the product space due to lack of information about other portions of the product space.
-\par
-\xhdr{Additional Findings} We further explore the diversity of the overall set of items that users consume.
-Encouraging consumption diversity is increasingly a common goal of many recommender systems \cite{castells2015novelty, kunaver2017diversity}, which is reasonable since it could be expected that higher diversity is associated with more intense search and therefore higher welfare.
-However, we find that, without recommendation, higher diversity of items consumed does not ensure higher welfare.
-This is due to the fact that high consumption set diversity in our model can arise from a sequence of bad realizations, which leads to very intense search over very different parts of the product space and high consumption diversity but very low welfare.
-However, discovering a sweet spot leads to highly rewarding local search and very low diversity.
-Thus, contrary to what a natural intuition could suggest, we find that consumption set diversity is not by itself a useful measure for understanding the impact of recommender systems on user welfare.
-\par
-We then turn to the extent to which recommendation leads to increasing user homogenization.
-In our context we define homogenization as the degree of similarity between the set of consumed items across individuals.
-We find that, without recommendation, there is very little homogenization as users are effectively all doing their own random exploration around the product space.
-In the ex-post utility maximizing consumption path (i.e.
-the consumption path the user would have taken if all uncertainty had been resolved) we find that some degree of homogenization is optimal.
-However, we find that while recommendation leads to an increase in consumer welfare, it concentrates user consumption choices in the same portions of the product space where there is a high common-value component.
-As a result, it leads to a substantially higher homogenization when compared both to the no recommendation and ex-post utility maximizing consumption path.
-\par 
-One implication from our model is that the user consumption paths have a clear path-dependence since the consumption of an item today influences the information a user has tomorrow which further influences consumption choices tomorrow.
-This is important for the debate regarding the extent to which recommender systems can be utilized to steer consumer choice as most of the literature analyzes this in a static context.
-However, this path-dependence implies that such platform steering can have dynamic consequences as steering towards certain items today can shape the long-run consumption choices of users.
-
-\xhdr{Related Work.} 
-The first set of related works studies the extent and implications of filter bubbles.
-\cite{pariser2011filter} first informally described the idea of the ``filter bubble'' which is that online personalization services would lead users down paths of increasingly narrower content so that they would effectively be isolated from a diversity of viewpoints or content.
-Following this a number of empirical studies, done in various disciplines, have since studied the extent to which this phenomena exists in a wide range of contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}.
-The most relevant to our study is \cite{nguyen2014exploring} who study whether this effect exists with recommender systems in the context of movie consumption.
-They find that even users whose consumption choices are not guided by recommendations fall into behavior consistent with ``filter bubbles'' and that recommender systems can actually increase the diversity of the content that users consume.
-To our knowledge there are no theoretical models that rationalize these empirical findings and our paper provides a theoretical framework through which to view this problem.
-Further, our paper provides a clear mechanism at play that drives such effects and how recommendation interacts with them.
-\par 
-Another set of papers have examined whether recommender systems can lead users to become increasingly homogenized or consume similar content as each other.
-\cite{celma2008hits, treviranus2009value} show that incorporating content popularity into recommender systems can lead to increased user homogenization.
-\cite{chaney2018algorithmic} show how user homogenization may arise from training recommender systems on data from users exposed to algorithmic recommendations.
-\cite{fleder2009blockbuster} show that homogenization can increase due to a recommendation bias towards popularity that arises from lack of information about items with limited consumption histories.
-Our model leads to similar results as previous theoretical and empirical work where recommender systems lead to increased user homogenization.
-However, the mechanisms behind this from our model differ from existing work as homogenization arises due to the fact that recommendation leads users to coordinate their consumption decisions in certain portions of the product space.
-\par 
-Our model of decision-making draws from work in decision theory and psychology.
-\cite{schulz2019structured} empirically studies how individuals solve sequential decision-making problems under uncertainty in large choice sets in the context of mobile food delivery orders.
-They find that individuals engage in similarity-based generalizations where learning about the realized value of a particular good provides them with information about similar items.
-We incorporate this finding into our model in a stylized manner where a single parameter controls the degree of information spillovers.
-\cite{gilboa1995case} considers a model of case-based decision making where choices are made based on information from previously experienced choice problems that are similar to the one faced by the decision-maker.
-However, while conceptually related, our model incorporates this idea in a different manner by utilizing Bayesian reasoning as opposed to a weighted sum of previous similar choice problems.
-\cite{hodgson2019horse} consider a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of their preferences in the environment, but consider it in the context of search for a single item as opposed to consumption of multiple items.
-\cite{bronnenberg2016zooming} empirically show that users engage in a ``zooming in” search strategy in terms of the information they acquire about products before purchasing and only end up acquiring information about a small set of products in a large product space.
-Finally, \cite{chen2013human} surveys work that has studied decision-making models in the context of recommender systems, though to our knowledge none of these papers study user decision making over time or study filter bubble and homogenization issues.
-
-\section{Our Model and Preliminaries}
-\subsection{Preliminaries on Expected Utility Theory}
-\noindent For every item $n$ in the product space $\mathcal J$, we assume that each user $i$ assigns a monetary equivalent $x_{i,n} \in \mathbb R$ to the experience of consuming it.
-Each user can value the same item differently.
-However, we assume that users have the same utility over money, given by a utility function $u: \mathbb R \to \mathbb R$, strictly increasing and continuous.
-So, ex-post, the value of item $n$ for user $i$ is given by $u(x_{i,n})$ However, before consuming the item, the user does not know exactly how she will value it.
-In particular, even users that will end up having the same ex-post valuation of item $n$ may differ in their ex-ante valuation because they hold different beliefs about it.
-We denote by $p_{i}$ the beliefs item $i$ has about how she will value each of the items in the product space.
-Note that this implies that consuming item $n$ is the same as taking a gamble.
-Each user evaluates the item according to the associated expected utility associated with the item, i.e.
-$U_i(n)=\mathbb E_{p_i}[u(x_n)]$.
-\par
-
-Risk aversion captures the how different users react to the risk associated to a particular consumption opportunity.
-It is formalized as follows: a given gamble $x$ takes real values and follows distribution $p$.
-Then, for every gamble $x$, there is a certain amount of money that makes the user indifferent between taking the gamble or taking the sure amount of money.
-This sure amount of money is called the \textit{certainty equivalent} of gamble $x$ and is denoted as $\delta(x)$.
-A user $i$ is more risk-averse than another user $j$ when for every gamble $x$, user $i$ requires a lower value to give up the gamble when compared to user $j$, i.e.
-$\delta_i(x)$.
-Therefore, a more risk-averse user is more willing to avoid the risk of taking the gamble.
-We assume that the utility function takes a flexible functional form $u(x)=1-\exp(-\gamma x)$ for $\gamma\ne0$ and $u(x)=x$ for $\gamma=0$ -- known as constant absolute risk-aversion preferences (from hereon CARA).
-Higher $\gamma$ implies higher risk-aversion with $\gamma=0$ corresponding to the risk-neutral case and $\gamma>0$ to the risk-averse one.
-Our formulations here follow standard economic decision-theory (see \cite{mas1995microeconomic} for a textbook treatment of these topics).
-\par
-
-\subsection{Model}
-\par
-\xhdr{Users.} We consider a set of users $I$ where each user $i \in I$ faces the same finite set of $N$ items $\mathcal J = \left\{0,1,...,N-1\right\}$.
-For simplicity, we assume that users only derive pleasure from item $n \in \mathcal{J}$ the first time they consume it.
-\par
-
-We denote by $x_{i,n}$ user $i$'s realized value from consuming item $n$.
-In particular, we consider that the realized value derived from a given item can be decomposed in the following manner: $x_{i,n}= v_{i,n} + \beta v_n$, where $v_{i,n}$ denotes an idiosyncratic component -- i.e.
-user $i$'s idiosyncratic taste for product $n$ --  and $v_{n}$, a common-value component.
-One can interpret $v_n$ as a measure of how much item $n$ is valued in society in general and, in a sense, $v_{i,n}$ denotes how $i$ diverges from this overall ranking.
-The scalar $\beta \in \mathbb{R}_{t}$ denotes the degree to which valuations are idiosyncratic to each user or common across users.
-If $\beta=0$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta$ is large, every individual has similar preferences.
-\par
-Stacking values in vector-form, we get the vector of values associated with each item 
-$${\left(x_{i,n}\right)}_{n \in \mathcal{J}}=:X_i =V_i+ \beta V $$
-where $V_i ={\left(v_{i,n}\right)}_{n \in \mathcal{J}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{J}}$.
-\par
-\xhdr{User Decision-Making.}
-We assume the user makes $T$ choices and therefore can only consume up to $T$ items, where $T$ is a small fraction of $N$.
-This captures the idea that users are faced with an immense choice set but that ultimately they end up experiencing (and learning) about just a small fraction of it.
-Since this is a sequential decision-making problem under uncertainty, in principle users face an exploration-exploitation trade-off.
-However, for tractability, we impose that users are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
-\par
-\xhdr{User Beliefs.} We assume that all the true realized utilities are drawn at $t = 0$.
-However, users do not know the realized utilities before consuming an item, but rather have beliefs over the realized utilities.
-Formally, user $i$ starts with some beliefs about $X_i$, namely that the idiosyncratic and common-value parts of the valuations are independent -- $V_i \perp \!\!\!
-\perp V$ -- and that each is multivariate normal
-\begin{enumerate}
-\item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ 
-\item $V \sim \mathcal N(\overline V, \Sigma)$ with $\overline V =0$
-\end{enumerate}
-We impose the normality assumption for two reasons.
-The first is that this allows for simple and tractable belief updating.
-The second is that it allows us to incorporate an easily interpretable correlation structure between the items.
-The precise formulation of $\Sigma$ and $\Sigma_i$ that we consider is defined below when we discuss user learning.
-\par
-Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{i,n}\right)_{n \in \mathcal{J}}$ are drawn independently from a jointly normal distribution, where $\overline v_{i,n} \sim \mathcal N (0, \overline \sigma^2)$ are independent and identically distributed.
-These $\overline v_{i,n}$ denote the prior belief that individual $i$ holds about her valuation over item $n$.
-As people are exposed to different backgrounds, their beliefs about what is good for them also varies and $\overline v_{i,n}$ denotes this idiosyncrasy at the level of prior beliefs.
-\par
-
-We assume users are expected utility maximizers.
-User $i$'s certainty equivalent for item $n$, the sure value that makes user $i$ indifferent between it and consuming the item $n$, conditional on the consumption history, is given by
-$\delta_{i}(n)\mid C_i^{t-1}=\mu_n-\frac{1}{2}\gamma \Sigma_{nn}$, where $\mu_n$ and $\Sigma_{nn}$ are the expected value and variance for item $n$ that the user has given their initial beliefs and consumption history up until time $t$.
-Note that this expression is known to be the certainty equivalent for CARA preferences in a Gaussian environment \cite{mas1995microeconomic}.
-
-\xhdr{User Learning}
-When a user consumes an item $n$ she learns the realized value for that item. In our model, we incorporate the idea that learning the value of item $n$ gives a user information about similar items. 
-%This is drawn from recent empirical evidence in \cite{schulz2019structured} that consumers learn how to navigate large choice sets using similarity-based generalization. 
-We assume that learning about the value of item $n$ reveals more about the value associated to items that are closer to it, which captures the idea that trying an item provides more information about similar items than about dissimilar ones.
-\par
-In order to have a well-defined notion of similarity we need to define a distance function between items, which we define as $d(n,m):=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}$ where $m$ and $n$ are indices of items in $\mathcal{J}$.
-This distance function is not intended to model the intricacies of a realistic product space, but instead to provide a stylized product space to help us understand the effects of informational spillovers on user behavior.
-We consider that the entry of $n$-th row and the ($m$)-th column of $\Sigma_i$ is given by $\sigma_i^2 \rho^{d(n,m)}$, and that of $\Sigma$ is given by $\sigma^2 \rho^{d(n,m)}$.
-The scalar $\rho \in [0,1]$ therefore impacts the covariance structure: a higher $\rho$ implies that learning the utility of $n$ is more informative about products nearby.
-Informativeness, for any $\rho \in (0,1)$, is decreasing in distance.
-The particular distance function that we utilize leads to this covariance structure being simple, where the $(n,n+1)$-th entry in the covariance matrix is $\sigma^{2} \rho$, the $(n,n+2)$-th entry is $\sigma^{2} \rho^2$, etc.
-\par
-The precise updating rule is as follows.
-Recall that at time $t$ the user's consumption history is given by $C_{i}^{t}$ and we denote the utility realizations of these items as $c_t$.
-We denote $\mu_t$ as the initial mean beliefs the user has over the items in $C_{i}^{t}$ and $\mu_{N-t}$ as the initial mean beliefs the user has over the remaining $N-t$ items, $\mathcal{J} \setminus C_{i}^{t}$.
-We partition the covariance matrix as follows:
-\[ \Sigma =  \left( \begin{array}{cc}
-\Sigma_{(N-t, N-t)} & \Sigma_{(N-t,t)} \\
-\Sigma_{(t,N-t)} & \Sigma_{(t,t)}
-\end{array} \right)
-\]
-Thus, after consuming the items in $C_{i}^{t}$ the resulting beliefs over the remaining items are given by $\mathcal{N}(\bar{\mu}, \bar{\Sigma})$ where $\bar{\mu}$ and $\bar{\Sigma}$ are as follows:
-\begin{align*}
-\bar{\mu} \mid c_t = \mu_{N-t} + \Sigma_{(N-t,t)} \Sigma_{(t,t)}^{-1}(c_t - \mu_t) \\
-\bar{\Sigma} \mid c_t = \Sigma_{(N-t,N-t)} - \Sigma_{(N-t,t)} \Sigma_{(t,t)}^{-1} \Sigma_{(t,N-t)}
-\end{align*}
-\begin{figure}[t]
-\includegraphics[width=.3\linewidth]{figures/Example-Bubbles.pdf}
-\captionof{figure}{Illustrative Example}
-\label{fig:illustrative_example}
-\end{figure}
-
-\xhdr{An Illustrative Example} We illustrate the main intuitions of our model with a simple example.
-Suppose that there are four items: 0, 1, 2, 3.
-The items are in different places of the product space, where 0 is close to 1 and 3 but more distant from 2, as shown in Figure \ref{fig:illustrative_example}.
-The initial beliefs are given as follows:
-\[ \mu = \left (\begin{array}{c}
-\mathbb{E}[x_0] \\
-\mathbb{E}[x_1] \\
-\mathbb{E}[x_2] \\
-\mathbb{E}[x_3]
-\end{array}  \right) = \left (\begin{array}{c}
-0 \\
-0\\
-0 \\
-0
-\end{array}  \right), \Sigma =  \sigma^{2} \left( \begin{array}{cccc}
-\rho^{0} & \rho^{1} & \rho^{2} & \rho^{0} \\
-\rho^{1} & \rho^{0} & \rho^{1} & \rho^{2} \\
-\rho^{2} & \rho^{1} & \rho^{0} & \rho^{1} \\
-\rho^{1} & \rho^{2} & \rho^{1} & \rho^{0} \\
-\end{array} \right)
-\]
-In period 1, every item is ex-ante identical since they have the same mean and variance and so suppose that the user breaks the tie arbitrarily and consumes item 0.
-The underlying correlation structure implies that upon observing that $x_0=y$ the user will update beliefs about the remaining 3 items according to the previously specified updating rule.
-For concreteness, we suppose that $\sigma = 1$ and $\rho = 0.5$, but the intuitions hold for any value of $\sigma$ and $\rho > 0$.
-First, we consider the case when $y = 0.5 > 0$.
-The resulting beliefs after observing $y$ are then as follows:
-\[ \bar{\mu} =   \left (\begin{array}{c}
-\mathbb{E}[x_1] \vspace{0.15cm} \\
-\mathbb{E}[x_2] \vspace{0.15cm} \\
-\mathbb{E}[x_3]
-\end{array}  \right) =\left (\begin{array}{c}
-\rho y  \vspace{0.15cm} \\
-\rho^{2} y  \vspace{0.15cm} \\
- \rho y \\
-\end{array} \right) =
-\left (\begin{array}{c}
-\frac{1}{4} \vspace{0.15cm} \\
-\frac{1}{8}  \vspace{0.15cm} \\
-\frac{1}{4}
-\end{array}  \right), \bar{\Sigma} =  \left( \begin{array}{ccc}
-\frac{3}{4} & \frac{3}{8} & 0 \vspace{0.15cm} \\
-\frac{3}{8} & \frac{15}{16} & \frac{3}{8} \vspace{0.15cm}  \\
-0 &\frac{3}{8} & \frac{3}{4}  \\
-\end{array} \right)
-\]
-Thus, upon learning $y$, the user updates beliefs about the remaining items.
-Note that $\mathbb{E}[x_1] = \mathbb{E}[x_3] > \mathbb{E}[x_2]$ since item 0's value is more informative about similar items' values, items 1 and 3, than items further way in the product space such as item 2.
-Moreover, $\bar{\Sigma}_{11} = \bar{\Sigma}_{33} < \bar{\Sigma}_{22}$ as the uncertainty about items 1 and 3 is further reduced compared to item 2.
-Thus, since $y > 0$, the user in the next period will consume items nearby to item 0 since, even though initially she believed that all items had the same mean, the spillover from consuming item 0 leads her to believe that items 1 and 3 have higher expected valuations.
-Since both the mean is higher for these items and the variance is lower, the user will consume items 1 and 3 regardless of her risk aversion levels.
-\par 
-Suppose instead that item 0 ends up having a negative valuation so that $y = -0.5 < 0$.
-Now we have that $\mathbb{E}[x_1] = \mathbb{E}[x_3] = -\frac{1}{4} <  -\frac{1}{8} = \mathbb{E}[x_2]$ with $\bar{\Sigma}$ remaining the same as when $y = 0.5$.
-Now, the risk-aversion levels of the user determine the choice in the next period.
-If the user is risk-neutral ($\gamma = 0$), then she will go across the product space to consume item $2$ in the next period since it has a higher expected value.
-However, if she is sufficiently risk-averse then she still may consume item $1$ or $3$ since her uncertainty about these items is lower than item $2$.
-In particular, this will happen when
-\begin{align*}
-\delta(3) = \delta(1) = \rho y - \frac{1}{2} \gamma \bar{\Sigma}_{11} > \rho^{2} y - \frac{1}{2} \gamma \bar{\Sigma}_{22} = \delta(2)
-\end{align*}
-Given the aforementioned parametrization and $y = -0.5$, the user will consume item $1$ or $3$ when $\gamma > \frac{4}{3}$ and will consume item $2$ when $\gamma < \frac{4}{3}$.
-Thus if the user is risk averse enough, then she might be willing to trade-off ex-ante lower expected values for lower risk and stick to consuming nearby items just because these items have lower uncertainty.
-\par 
-This example illustrates the main mechanisms in our model that can lead to excessive consumption of similar items.
-Once the user finds items in the product space with high valuations she will update her beliefs positively about items in this portion of the product space and continue consuming these items regardless of her level of risk aversion.
-However, this same updating leads to a reduction in uncertainty of these items and so, if she is sufficiently risk-averse, she still may continue consuming items in this portion of the product space, even if she has bad experiences with them, since they are perceived to be less risky.
-\par
-
-\xhdr{Recommendation.}
-Our model of recommendation is stylized in order to provide qualitative insights into how recommendation shapes behavior, instead of focusing on the details of how recommender systems are implemented in practice.
-We model recommendation as giving users information about the valuation of the products.
-\par
-
-We will consider three cases.
-The case of primary interest is \textit{recommendation} where the recommender observes values accrued and knows $V$ but does not know neither $V_i$.\footnote{We do not consider the acquisition of information for the recommender to know $V$ and suppose that she has sufficient data to learn $V$ with arbitrary precision at $t = 0$.} However, the recommender does know the users' beliefs $\bar V_i$.
-Thus, at any given period, the recommender provides a personalized recommendation that combines the knowledge of the common value component $V$ with the user beliefs $\bar V_i$.
-Knowing the user's beliefs about her valuation of each item become crucial in this case: just providing the user with information about $V$ may change the user's original ranking, but, without considering the user's beliefs, she will not necessarily follow the recommendation.\footnote{The notion of recommendation that we consider is idealized where the recommendation does the Bayesian updating for users.
-As users face a very difficult and costly Bayesian updating instance, a useful recommendation system provides recommendations taking into account this aspect and combines the history of past consumption and value realizations with the user's beliefs.
-An alternative that is equivalent would be the recommender simply providing $V$ without knowing $V_i$.
-In this case, the recommendation will be that the user $i$ chooses $r_{i,t} \in \arg \max_{n\in \mathcal J\backslash C^T_i} \mathbb E[u(x_n) \mid C_i^T]$
-but we assume the recommender provides full information about $V$.
-For instance, the recommender could display the whole distribution of utilities reported by other users or even its average, which is a good proxy for the common value component.
-It is possible for the recommender to only recommend the best item given such information, that is, the item with highest common value component.
-However, in that case, recommending only a single item generates costly Bayesian updating to the user and provides little guidance as to what she should indeed pick as the common value might be of little importance when compared to the idiosyncratic component.
-Therefore, we assume that the recommender reports the whole $V$, which results in higher expected welfare in choices and is not costly to implement.
-The results would be equivalent, but this would shift the burden of Bayesian updating to users.}
-\par
-
-We further consider two cases that serve primarily as benchmarks.
-The first is \textit{no recommendation}, where users get no additional information about values and make consumption choices based on their beliefs and consumption history.
-This gives us a benchmark as to how users in our model would behave \textit{without} recommendation so that we can analyze what changes with the introduction of recommendation.
-The second is the \textit{oracle recommendation} where the recommender knows the true realized utility of each good for each user and can therefore recommend the best remaining good in every period.
-This gives us a full information benchmark, which is the optimal consumption path for a user if all uncertainty about their preferences was resolved.
-Comparison to the oracle regime benchmark provides an analog to the standard regret measures utilized in the multi-armed bandit literature, which look at the difference in the expected value of the ex-post optimal action and the expected value of actions that were taken.
-\par
-
-\xhdr{Simulation Details}. 
-We analyze our model using numerical simulation since the sequential decision-making component of our model paired with the rich covariance structure between the items make it difficult to characterize optimal user behavior analytically.
-The Gaussian assumption allows for closed form belief updating which enables us to simulate our model but does not provide much help in analytical characterizations.
-\par
-
-We explore how consumption patterns differ as we consider different recommendation regimes and report representative results from our simulations.
-We simulate our model over 100 populations of users with 100 users per population for each combination of parameters.
-%For every population and every statistic we calculate, we average over the users in the population to get a representative statistic for this population. 
-A given set of parameters and a user are a single data point in our dataset.
-\par
-
-We simulate over risk-aversion parameters $\gamma \in \{ 0, 0.3, 0.6, 1, 5 \}$, \ standard-deviation $\sigma \in \{ 0.25, 0.5, 1.0, 2.0, 4.0 \}$, correlation parameters $\rho\in \{ 0, 0.1, 0.3, 0.5, 0.7, 0.9 \} $ and degree of idiosyncrasy of individual valuations $\beta \in \{ 0, 0.4, 0.8, 1, 2, 5\}$.
-We argue that these parameters are representative.
-First, the values of $\gamma$ that we consider range from no risk aversion ($\gamma = 0$) to reasonably high risk aversion levels $(\gamma = 5)$.
-The correlation parameter, $\rho$, ranges from uncorrelated valuations ($\rho = 0$) to nearly fully correlated valuations $(\rho = 0.9)$.
-$\beta$ ranges from the common value component playing no role $(\beta = 0)$ to the idiosyncratic component dominating the true valuations $(\beta = 5)$.
-Finally, the standard deviation ranges from little uncertainty $(\sigma = 0.25)$ to a large amount of uncertainty $(\sigma = 4)$.
-Thus, the range of considered parameter values cover the relevant portions of the parameter space in order to provide full insight into the behavior of the model.
-\par
-When we consider results varying a single parameter, we group the results over the other parameters and provide reports varying only the parameter of interest.
-We report results for a relatively small consumption history $T=20$ with a product space size $N=200$.
-Unless otherwise reported, our results are robust to varying $N$.
-\par
-
-\begin{figure}
-\begin{subfigure}{.5\linewidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{figures/rho_pos_consumption_dist_N_200T_20_overall.pdf}
-  \label{fig:sfig1}
-\end{subfigure}%
-\begin{subfigure}{.5\linewidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{figures/rho_zero_consumption_dist_N_200T_20_overall.pdf}
-  \label{fig:sfig2}
-\end{subfigure}
-\caption{Consumption path with (Left) and without (Right) correlation between values}
-\label{fig:correlation_consumption_path}
-\end{figure}
-
-\section{Results}
-\subsection{Local Consumption and Filter Bubbles}
-We characterize ``filter bubble'' effects in our model as the degree to which users engage in \textit{local consumption}.
-We define local consumption in terms of the average consumption distance between the items consumed by the users at time $t-1$ and $t$.
-Thus, in the context of our model, filter bubble effects arise when the average consumption distance decreases over time and, across regimes, when the overall levels are lower for a given recommendation regime compared to another.
-
-\begin{finding}\label{finding_local_consumption}
-The impact of recommendation on local consumption:
-\begin{enumerate}
-\item When $\rho = 0$, there is no difference in consumption distance between the three recommendation regimes.
-\item When $\rho > 0$, no recommendation induces more local consumption than both recommendation and oracle regimes.
-This effect is amplified as $\rho$ increases as well as when users are more risk averse ($\gamma$ increases)
-\end{enumerate}
-\end{finding}
-
-First, the bottom panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho = 0$, there is no difference in consumption distance between the three regimes.
-This is due to the fact that when $\rho = 0$, there is no reason that items that are close in the product space should have similar values and so the optimal consumption path does not depend on the similarity of the items.
-However this also means that users do not learn anything about neighboring products and so there is limited path-dependence in consumption.
-Not only is there no difference in the levels between the three regimes, but they all have the same, flat, average consumption distance path.
-This underscores the fact that such concerns about filter bubbles \textit{only} arise in a world where there is correlation between the realized utility of items.
-If there were no correlation between the realized utilities then there would be no reason for users to consume similar items and thus no narrowing effect, regardless of the information on the true utility of the items that users had.
-\par
-The top panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho > 0$, both recommendation and no recommendation lead to increasingly local consumption compared to the oracle benchmark case.
-Further, the average consumption path between periods is \textit{decreasing} for the no recommendation case whereas it is \textit{increasing} for the oracle case.
-The recommendation regime decreases the degree of local consumption, but not as much as the oracle benchmark.
-Due to the correlation of values, the oracle consumption path exploits this and leads to the consumption of more similar products than in the case when $\rho = 0$.
-However, since these spillovers also impact user learning in the no recommendation case, users \textit{over-exploit} this and increasingly consume products similar to good products that they have consumed before.
-This is further illustrated by Figure \ref{fig:local_consumption_across_rho} which shows how the consumption paths in the oracle and no recommendation regimes vary as $\rho$ increases.
-\par
-
-Finally, this effect is further amplified as the level of risk aversion, $\gamma$, increases.
-Figure \ref{fig:no_rec_risk_aversion} shows how drastically the degree of local consumption increases as $\gamma$ increases.
-This is due to the fact that the spillovers not only affect the mean expected belief about quality but also the degree of uncertainty.
-Local consumption therefore leads to users having less uncertainty about certain areas of the product space and risk aversion may lead them to increasingly consume nearby products.
-
-\begin{figure}[t]
+\begin{figure}[H]
 \begin{subfigure}{.45\textwidth}
@@
 
-\begin{figure}[t]
+\begin{figure}[H]
 \begin{subfigure}{.45\textwidth}
@@
 \end{subfigure}%
-\caption{Effect of Risk Aversion on Local Consumption \\ No Recommendation (Top Left), Recommendation (Top Right) and Oracle (Bottom)}
+\caption{Effect of Risk Aversion ($\gamma$) on Local Consumption \\ No Recommendation (Top Left), Recommendation (Top Right) and Oracle (Bottom)}
 \label{fig:no_rec_risk_aversion}
-\end{figure}
-
-In sum, filter bubble effects can only arise when there is an inherent correlation between the realized utilities of the items in the product space.
-However, this is often the case in most environments where recommender systems are deployed.
-When there is a correlation between the realized utilities then filter bubbles can naturally arise due to the nature of how individuals acquire additional information about the remaining items.
-We have shown that, unless users are provided with additional information to guide their consumption choices, then these information spillovers and user risk-aversion can lead users into filter bubbles.
-When users consume high valued items, they exploit the underlying correlation across different products' values, stronger for similar products, which leads them to increasingly consume products in narrower and narrower portions of the product space.
-Risk aversion may lead users into performing local consumption even when they have a low valuation of nearby items just because they know what to expect from the item.
-Recommendation leads to these effects being mitigated by providing users with additional information on products outside the already explored portions of the product space.
-Further, if all uncertainty was resolved as in the oracle regime, then such behavior is not present.
-
-\subsection{User Welfare and Product Diversity}
-In this section we primarily focus on the impact of recommendation on user welfare and the overall diversity of the items that they consume.
-\par 
-While in the previous section we looked at the distance between consecutive items, in this section we focus on a diversity measure that considers the entire consumed set of items.
-The diversity measure we utilize is common in recommender system literature (e.g.
-\cite{ziegler2005improving}) which is the average normalized pairwise distance between the consumed products:
-$$D_i:=\frac{1}{N}\frac{1}{T(T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$$
-
-\begin{figure}[ht]
-\includegraphics[width=.45\linewidth]{figures/gamma_diversity_N_200_T_20}
-\captionof{figure}{Risk Aversion and Diversity}\label{fig:risk_aversion_diversity}
-\end{figure}
-
-\begin{finding}\label{finding_diversity}
-The impact of recommendation on product diversity:
-\begin{enumerate}
-\item When $\rho = 0$, product diversity is the same across all three recommendation regimes;
-\item When $\rho > 0$, product diversity decreases across all recommendation regimes but decreases the most in the no recommendation regime.
-This effect is amplified as $\rho$ increases as well as when users become more risk-averse.
-\end{enumerate}
-\end{finding}
-\par 
-Finding~\ref{finding_diversity} summarizes the main results on product diversity.
-As before, when there is no correlation between valuations, product diversity is the same across different recommendation regimes.
-The over-exploitation of information spillovers when $\rho > 0$ leads to product diversity being lowest in the no recommendation regime.
-As a result, this effect gets amplified as $\rho$ increases, which leads to the gap in diversity between the regimes to increase as $\rho$ increases.
-Figure \ref{fig:diversity_welfare_correlation} shows how diversity varies as $\rho$ increases across the three regimes that we consider.
-There is a similar increasing diversity gap as $\gamma$, or the level of risk-aversion, increases as can be seen in Figure \ref{fig:risk_aversion_diversity}.
-The mechanisms behind these effects have direct parallels to those discussed in the previous section since low average sequential consumption distance directly leads to low diversity.
-\par 
-We now study how recommendation impacts user welfare in our model.
-In our model users make consumption decisions that maximize their current period ex-ante utility that depends on their beliefs in that period.
-Thus, from an ex-ante perspective, they make optimal decisions, but our primary interest is in understanding how the ex-post, or realized, utility varies across regimes and parameter values.
-We define user's \textit{ex-post} welfare as the average of the realized values, controlling for the effect of $T$:
-$$W_i:= \frac{1}{T}\sum_{n \in C_i^T} x_{i,n}$$
-
-\begin{finding}\label{finding_welfare_gap}
-The impact of recommendation on consumer welfare is as follows:
-\begin{enumerate}
-\item Under oracle recommendation, welfare is constant as $\rho$ increases.
-\item Under no recommendation, welfare is increasing as $\rho$ increases.
-\item Recommendation introduces welfare gains relative to no recommendation, but these gains are decreasing as $\rho$ increases. 
-\end{enumerate}
-\end{finding}
-\par 
-Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which are drawn from Figure~\ref{fig:diversity_welfare_correlation}.
-The most interesting observation is that the value of recommendation decreases as $\rho$ decreases.
-The intuition for this is clear as recommendation provides users with information that allows them to better guide their decisions and increase welfare.
-However, as $\rho$ increases users get increasingly more information simply from consuming items since the realized utility is now more informative about the utility of nearby items.
-Thus, since consumption decisions themselves reveal information, the information that recommendation reveals provides less value.
-\par 
-Figure~\ref{fig:diversity_welfare_common_value} shows how diversity and welfare vary across recommendation regimes as the the strength of the common-value component, $\beta$, varies.
-As expected, when $\beta = 0$, these values are the same.
-However, as $\beta$ grows the idiosyncratic component of user valuations becomes increasingly less important and the recommendation regime begins to converge to the oracle regime.
-Indeed, we see that both product diversity and welfare increase under recommendation, whereas under no recommendation we see that there is a decrease in diversity with little changes to consumer welfare.
-This is intuitive as $\beta$ controls the informativeness of recommendation about user preferences.
-\par 
-One striking observation is that the decrease in diversity does not appear to be associated with a decline in welfare.
-Indeed, it appears that the opposite is the case - that low diversity is associated with higher welfare and vice versa.
-Thus, we next explore the relationship between welfare and diversity.
-\begin{figure}
-\begin{subfigure}{.45\linewidth}
-  \includegraphics[width=.8\linewidth]{figures/rho_diversity_N_200_T_20.pdf}
-\end{subfigure}
-\begin{subfigure}{.45\linewidth}
-  \includegraphics[width=.8\linewidth]{figures/rho_welfare_N_200_T_20.pdf}
-\end{subfigure}
-\caption{User Welfare and Diversity as the strength of correlation, $\rho$, varies}\label{fig:diversity_welfare_correlation}
 \end{figure}
@@
 
-\begin{finding}\label{finding_diversity_welfare_corr}
-In the no recommendation regime, diversity and welfare are:
-\begin{enumerate}
-\item Negatively correlated when users have no risk-aversion;
-\item Uncorrelated when users have high levels of risk-aversion.
-\end{enumerate}
-In the recommendation regime, diversity and welfare are:
-\begin{enumerate}
-\item Uncorrelated when users have no risk-aversion;
-\item Positively correlated when users have high levels of risk-aversion.
-\end{enumerate}
-In the oracle regime, diversity and welfare are always uncorrelated.
-\end{finding}
-
-Finding~\ref{finding_diversity_welfare_corr} summarizes our findings on the relationship between diversity and welfare.
-Figure \ref{fig:diversity_welfare_ra} shows how diversity and welfare correlate for the no recommendation case as we vary the degree of risk aversion.
-When there is no risk-aversion then there is a negative correlation between welfare and diversity.
-This is since, with no risk-aversion, in a given period a user will select the good that she currently believes has the highest expected value.
-High product diversity in this case can arise from a user who consumes an item she disliked and updates her beliefs about nearby items negatively.
-As a result, in the following period she will pick an item far away in the product space from the item that was previously consumed.
-If instead the user valued highly the item that she had consumed, then she is more likely to pick a nearby item.
-The information spillovers therefore lead to high product diversity being negatively correlated with welfare.
-\par
-This only happens since $\gamma = 0$ leads to users only caring about the expected value of the product.
-However, as we saw in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, increasing $\gamma$ can lead to lower diversity and increasingly local consumption due to the fact that the degree of uncertainty now impacts users' choices.
-This weakens the negative relationship between diversity and welfare since both negative and positive experiences with a good reduce uncertainty about surrounding items.
-This leads to the inverted-U shape found in Figure \ref{fig:diversity_welfare_ra} when $\gamma$ is relatively large (e.g $\gamma = 5$) though diversity and welfare are virtually uncorrelated in the data.
-In the recommendation and oracle regimes, under risk neutrality ($\gamma=0$), welfare and diversity and uncorrelated, while under risk aversion ($\gamma=5$), it is possible to observe in Figure \ref{fig:diversity_welfare_ra_partial} an actual positive relation between diversity and welfare as recommendations are able to reduce uncertainty and facilitate exploration of the product space.
 \begin{figure}[t]
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_rn.pdf}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_ra.pdf}
-\end{subfigure}\\
-\caption{Diversity vs. Welfare, No-Recommendation, $\gamma = 0$ (Left), Diversity vs. Welfare, $\gamma = 5$ (Right)}\label{fig:diversity_welfare_ra}
+\includegraphics[width=.45\linewidth]{figures/beta_homogeneity_N_200_T_20}
+\captionof{figure}{Strength of Recommendation ($\beta$) and Homogeneity}\label{fig:beta_homo}
+\end{figure}
+\begin{figure}[t]
+\includegraphics[width=.45\linewidth]{figures/rho_homogeneity_N_200_T_20}
+\captionof{figure}{Correlation ($\rho$) and Homogeneity}\label{fig:cor_homo}
 \end{figure}
 
-\begin{figure}[t]
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_rn_partial.pdf}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_ra_partial.pdf}
-\end{subfigure}\\
-\caption{Diversity vs. Welfare, Recommendation, $\gamma = 0$ (Left), Diversity vs. Welfare, $\gamma = 5$ (Right)}\label{fig:diversity_welfare_ra_partial}
-\end{figure}
-
-\subsection{User Homogenization}
-In this section, we focus on comparisons across users and investigate how the consumed set of items across users varies across different recommendation regimes and parameter values.
-In particular we look at the degree of \textit{homogenization} between users.
-Similar to other papers that study the degree of homogenization in recommender systems (e.g.
-\cite{chaney2018algorithmic}) we measure homogeneity via the Jaccard index between the consumption sets of users:
-\begin{align*}
-H:=\frac{1}{|I|(|I|-1)}\sum_{i,j \in I: i \ne j}d_J(C_i^T,C_j^T)
-\end{align*}
-where $d_J$ denotes the Jaccard index and $H \in [0,1]$.
-\begin{finding}\label{finding_homogeneity}
-The impact of recommendation on homogeneity is as follows:
-\begin{enumerate}
-\item Highest under recommendation and lowest under no recommendation;
-\item Increasing in $\beta$, or the weight of the common-value component;
-\item Decreasing in $\rho$ for partial recommendation, but weakly increasing in $\rho$ for no recommendation.
-\end{enumerate}
-\end{finding}
-Finding~\ref{finding_homogeneity} summarizes our findings on the impact of recommendation on user homogeneity.
-First, we study how the degree of homogenization varies as we increase $\beta$, the weight of the common value component.
-As $\beta$ increases we expect that users should become increasingly homogeneous as the realized utilities of the items are now increasingly similar.
-Figure \ref{fig:beta_homo} confirms that as the weight of the common-value component, $\beta$, increases users consume increasingly similar sets of items.
-The homogenization effect is strongest under the recommendation regime since the revelation of the common-value component induces users to consume products in similar areas of the product space.
-As $\beta$ increases, some amount of homogenization is optimal as can be seen from the oracle case.
-However, since users in the no recommendation regime do not know the common-value component they engage in local consumption in different areas of the product space which leads to less than optimal homogeneity.
-\par
-We next study how the degree of homogeneity varies as $\rho$ increases.
-Figure \ref{fig:cor_homo} shows how homogeneity varies as $\rho$ increases across the different recommendation regimes.
-The most striking finding is that homogeneity decreases as $\rho$ increases in the recommendation regime.
-As was highlighted in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, the degree of local consumption increases with $\rho$.
-Even though the revelation of the common-value component induces them to search in similar parts of the product space, their idiosyncratic components induce them to consume products in a more localized area of the product space as $\rho$ increases which leads to a decline in homogeneity.
-
-\begin{figure}[t]
-\includegraphics[width=.45\linewidth]{figures/beta_homogeneity_N_200_T_20}
-\captionof{figure}{Strength of Recommendation and Homogeneity}\label{fig:beta_homo}
-\end{figure}
-~
-
-\begin{figure}[t]
-\includegraphics[width=.45\linewidth]{figures/rho_homogeneity_N_200_T_20}
-\captionof{figure}{Correlation and Homogeneity}\label{fig:cor_homo}
-\end{figure}
-
-\section{Conclusion}
-We have studied a model of user decision-making in the context of markets where recommender systems are traditionally deployed in order to understand the consequences of recommender systems on consumption patterns of users.
-The key components of the model are that users make decisions under uncertainty and that consuming a good leads to informational spillovers about similar items.
-We have shown that this alone can generate consumption behavior that is consistent with filter bubble effects that have traditionally been associated with personalized recommender systems.
-This effect is weakened once we allow for recommendation, but not as much as if users have full information.
-We have further explored the relationship between consumption diversity and user welfare and found that they are negatively correlated, implying that striving for consumption diversity is not necessarily welfare-improving for users.
-Finally, we have shown that recommendation can lead to increases in user homogenization both relative to no recommendation and full information since it concentrates users' consumption in portions of the product space with the highest utility according to the limits of what the recommender can predict about the utility of the good.
-\par 
-Our results highlight that user choice in these markets is path dependent as a result of the nature of user preferences and how these preferences evolve over time.
-This has important ramifications for understanding the extent to which recommender systems can steer user behavior, potentially in anti-competitive ways.
-Steering users towards certain items today provides them with information that they utilize to make decisions tomorrow, which leads them to consume similar items to those they were originally steered towards.
-Thus, online platforms can potentially shape the long-term preferences of users and not only impact their current period choices.
-This has broader ramifications about the equilibrium effects of such behavior on competition and consumer welfare in online platforms which we leave to future work.
-\par
-We leave for future work exploring to what extent user behavior changes if we drop the myopia assumption, so that users themselves are engaging in explicit exploration.
-This likely interacts with recommendation in interesting ways as the recommendations may provide information that can complement or crowd-out user-led exploration.
-Even under the myopia assumption, understanding which recommendation system is optimal is an interesting and challenging problem.
-Further, our results point to an important consideration in the literature on incentivizing exploration which is that the exploration faced by the recommender itself likely interacts in interesting and important ways with the choice problem faced by the users.
-\par 
-Overall, understanding how users make choices in online platforms and how recommender systems impact these choices is important to understand.
-Understanding these issues allows us to better understand the broader social and economic impact of such systems as well as can provide a guide to the improved design of such systems.
 % Bibliography

### 2020-05-06 paper/filter_bubbles.tex
udpate

@@
 
-Users in many online platforms face difficult decision problems as they have to sort through the vast choice sets present in these markets.
+Recommender Systems (RS) have become critical for assisting users in navigating the large choice sets that they face in many online markets.
 For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
-A common remedy introduced by online platforms has been to deploy \textit{recommender systems}, which have become critical for assisting users in making consumption decisions.
-These systems provide users with information and recommendations to allow them to make more informed consumption decisions.
-These systems have been shown to be influential in steering consumer choice with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{MacKenzie et al.
+Users in many cases are not aware of most items, let alone have full information about their preferences over them and, to make matters worse, the goods in these markets are usually experience goods whose true value can only be learned after consumption.
+Furthermore, users interact with these systems routinely and watch more than one movie on Netflix, buy more than one product on Amazon, and listen to more than one video on YouTube.
+\par
+
+Recommender systems have been influential in shaping consumer choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{MacKenzie et al.
 (2013, Oct.),  How retailers can keep up with consumers.
 \url{https://www.mckinsey.com/industries/retail/our-insights/how-retailers-can-keep-up-with-consumers}.
-Retrieved on October 3, 2019.}
+Retrieved on October 3, 2019.} While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems.
+There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals,\footnote{Friedersdorf (2018, Mar.
+12), YouTube Extremism and the Long Tail, \textit{The Atlantic}.
+\url{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350}.
+Retrieved on October 3, 2019.} that personalized recommender systems lead users into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
 \par
-Despite evidence of the power of these systems in determining user choice, we have little theoretical understanding of how users make choices in these markets and how recommender systems shape these choices.
-Understanding this has become increasingly important as a number of concerns have emerged regarding the broader social and economic consequences of such systems.
-There have been claims that YouTube's recommendation algorithm unintentionally led to the radicalization of many individuals,\footnote{Friedersdorf (2018, Mar.
-12), YouTube Extremism and the Long Tail,
-\textit{The Atlantic}. 
-\url{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350}. Retrieved on October 3, 2019.} 
-that personalized recommender systems lead users into \textit{filter bubbles} 
-where they effectively get isolated from a diversity of viewpoints or content 
-\cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
-Furthermore, an emerging worry is that recommender systems can be utilized to steer consumer choice towards products more profitable for online platforms to the detriment of consumer welfare and the overall competitiveness of such markets \citep{aridor2020recommender, bourreau2018streaming, DeCorniereTaylor2019RAND}.
-Finally, there has emerged a burgeoning literature that focuses on understanding how recommender systems themselves can acquire information about product quality from the choices of self-interested consumers \cite{kremer2014implementing, che2017recommender,frazier2014incentivizing,mansour2015bayesian}.
-\par
-While these issues are important and wide-reaching, it is hard to make progress on them without a model of user choice with foundations in economic decision theory and psychology.
-In this paper we provide such a model of user preferences and choice in these markets and study how the introduction of recommendation affects user choice.
-Our primary focus is on utilizing our model to understand the extent to which recommender systems lead individual users into filter bubbles and homogenize across user consumption choices.
-We rationalize existing empirical work on these topics and provide an argument for the underlying mechanisms behind such effects.
-We also discuss how our model can shed light on how recommender systems can be utilized to steer choice and affect the long-term consumption patterns of users.
-Finally, our model provides a direction for making progress on an open question in the literature on incentivizing exploration in recommender systems on how to incorporate realistic models of user choice into such models (e.g.
-\citep{immorlica2018incentivizing}).
 
-\xhdr{Our Model} We consider a model of user choice that has the following components.
+\xhdr{Model} We consider a model of user choice that has the following components.
 \par
@@
 
-\xhdr{Main Findings}
-Our first set of results focuses on understanding to what extent recommender systems lead users into filter bubbles and how such filter bubbles may naturally arise.
-The standard notion of filter bubbles is that, over time, user consumption becomes increasingly concentrated on narrower and narrower parts of the product space, leading to a decreasing distance between subsequent consumed items over time.
-However, empirical work such as \cite{nguyen2014exploring}, finds that user consumption choices over time naturally exhibit such filter bubble effects even without recommendation and that, in fact, recommender systems mitigate such effects.
-The results of our model are in line with these empirical results and give a clear intuition as to why such effects arise and why recommendation mitigates them to some degree.
+\xhdr{Contributions}
+Using this model we first investigate the degree to which recommender systems lead users into filter bubbles.
+Our model rationalizes the empirical results in \cite{nguyen2014exploring} who show that in the context of movie consumption, user behavior is consistent with filter bubble effects \textit{even without recommendation}.
+The driving force of this in our model is that preferences for similar goods are correlated, which implies that when a good is consumed and the user learns its value, it provides information about similar goods.
+Crucially, this not only impacts the underlying belief about the average utility of the good, but also the level of uncertainty.
+Consequently, this learning spillover induces users to consume goods ``similar'' to those they consumed before that had high realized value, leading to an increasing narrowing of consumption patterns towards these regions of the product space.
+This effect is further amplified when users are \textit{risk-averse}, a concept from decision theory where all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty.
+We compare this benchmark to the case where recommendations are provided. 
+We consider recommendations as giving information to users in order to reduce their uncertainty.
+Recommendations could potentially make matters worse: providing information about the value of the goods when this value is correlated could intensify local search and decrease consumption diversity.
+However, and in line with the empirical results from \cite{nguyen2014exploring}, we find that recommendations reduce the narrowing effect.
 \par
-We first find that when there is no correlation between the valuations of goods there is no reason for filter bubble effects to arise, regardless of the information that users have on the valuations of the products.
-However, once there is a correlation between the true valuations of the items then such filter bubble effects arise even without recommendation.\footnote{The fact that in many environments where recommender systems are deployed the true valuations of items are correlated makes this the most relevant case to consider.
-Indeed, the core idea behind standard content-based recommendation methods relies on this fact as these methods determine recommendations based on finding similar items to those that a user has consumed before and enjoyed \citep{adomavicius2005toward}.} The intuition is that, users initially have uncertainty about their true valuations and initial consumption of items leads them to update their beliefs strongly about the average utility of nearby items.
-When a user finds an item that has high realized utility, this induces her to update her beliefs positively about the realized utility of nearby items which leads to  the consumption of these items.
-Thus, without recommendation users increasingly consume items in this portion of the vast product space since they never get information about items in different parts of the product space.
-This learning spillover not only induces positive updates about the average utility of the item, but also reduces the level of uncertainty they have about these items.
-Thus, this effect is further amplified when users are \textit{risk-averse}.
-This is a standard concept from decision theory by which, all else being equal, users have a preference for items with lower uncertainty to those with higher uncertainty.
+
+We then explore how diverse the overall set of items that users consume is.
+It could be expected that higher diversity being associated with more intense search is associated with higher welfare.
+Much to the opposite, we find that, without recommendation, higher diversity of items does not ensure higher welfare.
+This is due to the fact that once users find good items, they consume items nearby in the product space due to the correlation of values, leading to higher overall value from consumption.
+When they consume products they value little, they move to another part of the product space.
+A sequence of bad realizations will then lead to very intense search over very different parts of the product space and high consumption diversity, whereas discovering a sweet spot leads to highly rewarding local search and very low diversity.
+Thus, contrary to what a natural intuition could suggest, we find that consumption set diversity is not by itself a useful measure for understanding the impact of recommender systems on users.
 \par
-We then investigate whether the introduction of recommendation either mitigates or amplifies this effect.
-Recommendations could potentially make matters worse: providing information about the common value of the items when this value is correlated could intensify the extent of local consumption.
-However, in line with existing empirical work, we find the opposite - that recommendation partially mitigates this narrowing effect - since the revelation of the common value component allows for users to escape being stuck in portions of the product space due to lack of information about other portions of the product space.
+
+We find that recommendation can lead to increases in homogeneity among users.
+We model the realized valuations as being a weighted sum of a common-value and an idiosyncratic component.
+This formulation gives a stylized notion of predictability of user preferences where the idiosyncratic component is inherently unpredictable given other individuals' preferences and the common-value component is what the recommonder can learn from previous users' data.
+The recommender uses its knowledge of the common-value component and the individuals' beliefs over the product space when designing personalized recommendation.
+This leads to an increase in welfare for users, but concentrates their consumption around items that have a high common-value component, resulting in an increase in homogeneity.
 \par
-\xhdr{Additional Findings} We further explore the diversity of the overall set of items that users consume.
-Encouraging consumption diversity is increasingly a common goal of many recommender systems \cite{castells2015novelty, kunaver2017diversity}, which is reasonable since it could be expected that higher diversity is associated with more intense search and therefore higher welfare.
-However, we find that, without recommendation, higher diversity of items consumed does not ensure higher welfare.
-This is due to the fact that high consumption set diversity in our model can arise from a sequence of bad realizations, which leads to very intense search over very different parts of the product space and high consumption diversity but very low welfare.
-However, discovering a sweet spot leads to highly rewarding local search and very low diversity.
-Thus, contrary to what a natural intuition could suggest, we find that consumption set diversity is not by itself a useful measure for understanding the impact of recommender systems on user welfare.
+
+Finally, we discuss how our model and findings can be used to inform the design and evaluation of recommender systems.
+Our evaluation metric relies on the intuition that a recommendation is only useful insofar as it leads a user to take a better action than she would without the recommendation.
+Thus, in order to evaluate the value of recommendation, it is important to be able to predict what a user would do without recommendation and how recommendation would change their behavior.
+If recommendations fail to affect behavior, they are of no use: the user would consume the same good and learn its value with and without the recommendation.
+Moreover, we argue that successful recommendation systems have to take into account the degree of uncertainty users have over different portions of the product space.
+Being able to predict valuations after consumption is not sufficient as uncertainty over the idiosyncratic component of one's own preferences may steer the user away from the recommended items in the face of risk-aversion.
 \par
-We then turn to the extent to which recommendation leads to increasing user homogenization.
-In our context we define homogenization as the degree of similarity between the set of consumed items across individuals.
-We find that, without recommendation, there is very little homogenization as users are effectively all doing their own random exploration around the product space.
-In the ex-post utility maximizing consumption path (i.e.
-the consumption path the user would have taken if all uncertainty had been resolved) we find that some degree of homogenization is optimal.
-However, we find that while recommendation leads to an increase in consumer welfare, it concentrates user consumption choices in the same portions of the product space where there is a high common-value component.
-As a result, it leads to a substantially higher homogenization when compared both to the no recommendation and ex-post utility maximizing consumption path.
-\par 
-One implication from our model is that the user consumption paths have a clear path-dependence since the consumption of an item today influences the information a user has tomorrow which further influences consumption choices tomorrow.
-This is important for the debate regarding the extent to which recommender systems can be utilized to steer consumer choice as most of the literature analyzes this in a static context.
-However, this path-dependence implies that such platform steering can have dynamic consequences as steering towards certain items today can shape the long-run consumption choices of users.
 
-\xhdr{Related Work.} 
-The first set of related works studies the extent and implications of filter bubbles.
-\cite{pariser2011filter} first informally described the idea of the ``filter bubble'' which is that online personalization services would lead users down paths of increasingly narrower content so that they would effectively be isolated from a diversity of viewpoints or content.
-Following this a number of empirical studies, done in various disciplines, have since studied the extent to which this phenomena exists in a wide range of contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}.
+\textbf{Related Work.} 
+\cite{pariser2011filter} first informally described the ``filter bubble'' which is the idea that online personalization services would lead users down paths of increasingly narrower content so that they would effectively be isolated from a diversity of viewpoints or content.
+A number of studies, done in a wide range of disciplines, have since empirically studied whether or not this phenomenon exists in various contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}.
 The most relevant to our study is \cite{nguyen2014exploring} who study whether this effect exists with recommender systems in the context of movie consumption.
 They find that even users whose consumption choices are not guided by recommendations fall into behavior consistent with ``filter bubbles'' and that recommender systems can actually increase the diversity of the content that users consume.
-To our knowledge there are no theoretical models that rationalize these empirical findings and our paper provides a theoretical framework through which to view this problem.
-Further, our paper provides a clear mechanism at play that drives such effects and how recommendation interacts with them.
-\par 
-Another set of papers have examined whether recommender systems can lead users to become increasingly homogenized or consume similar content as each other.
-\cite{celma2008hits, treviranus2009value} show that incorporating content popularity into recommender systems can lead to increased user homogenization.
-\cite{chaney2018algorithmic} show how user homogenization may arise from training recommender systems on data from users exposed to algorithmic recommendations.
-\cite{fleder2009blockbuster} argues that homogenization can increase due to a recommendation bias towards popularity that arises from lack of information about items with limited consumption histories.
-Our model leads to similar results as previous theoretical and empirical work where recommender systems lead to increased user homogenization.
-However, the mechanisms behind this from our model differ from existing work as homogenization arises due to the fact that recommendation leads users to coordinate their consumption decisions in certain portions of the product space.
-\par 
-Our model of decision-making draws from work in decision theory and psychology.
-\cite{schulz2019structured} empirically studies how individuals solve sequential decision-making problems under uncertainty in large choice sets in the context of mobile food delivery orders.
-They find that individuals engage in similarity-based generalizations where learning about the realized value of a particular good provides them with information about similar items.
-We incorporate this finding into our model in a stylized manner where a single parameter controls the degree of information spillovers.
-\cite{gilboa1995case} considers a model of case-based decision making where choices are made based on information from previously experienced choice problems that are similar to the one faced by the decision-maker.
-However, while conceptually related, our model incorporates this idea in a different manner by utilizing Bayesian reasoning as opposed to a weighted sum of previous similar choice problems.
-\cite{hodgson2019horse} consider a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of their preferences in the environment, but consider it in the context of search for a single item as opposed to consumption of multiple items.
-\cite{bronnenberg2016zooming} empirically show that users engage in a ``zooming in” search strategy in terms of the information they acquire about products before purchasing and only end up acquiring information about a small set of products in a large product space.
-Finally, \cite{chen2013human} surveys work that has studied decision-making models in the context of recommender systems, though to our knowledge none of these papers study user decision making over time or study filter bubble and homogenization issues.
+To our knowledge there are no theoretical models that rationalize the empirical findings but predominantly empirical studies.
+Our paper provides a theoretical framework through which to view this problem.
+The second question is to what extent recommender systems can lead users to become increasingly homogenized or consume similar content as each other.
+\cite{celam2008hits,treviranus2009value} show that incorporating content popularity into recommender systems can lead to increased user homogenization.
+\cite{chaney2018algorithmic} more generally look at the issue of user homogenization in recommender systems.
+Our results are similar to previous work, where the limits of recommendation lead the recommender to over-recommend products from certain portions of the product space which leads users to consume goods in those regions.
+\cite{fleder2009blockbuster} also consider the effects of recommendation on multi-period user behavior, but predominantly focus on sales diversity and employ a different model of user behavior than we do here.
+\cite{hodgson2019horse} consider a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of their preferences in the environment, but consider it in the context of search for a single item.
+Our discussion of recommender system evaluation contributes with insights closely tied to the value of information that is used in the economics literature (see e.g.
+\cite{bergemann2019information}) and builds off the large literature stemming from \cite{mcnee2006being} that makes the case that accurate predictions of true realized valuations may not lead to the best recommendations.
+Our evaluation measure is most closely related to the notion of serendipity in recommender systems that has been studied extensively (e.g.
+see \cite{kotkov2016survey} for a survey).
 
@@
 The particular distance function that we utilize leads to this covariance structure being simple, where the $(n,n+1)$-th entry in the covariance matrix is $\sigma^{2} \rho$, the $(n,n+2)$-th entry is $\sigma^{2} \rho^2$, etc.
-\footnote{This exponential decay correlation structure can be related to the tenet of case-based similiartiy of Gilboa and Schmeidler (2008, MSS) for an axiomatization of exponential similarity.}
+\footnote{This exponential decay correlation structure can be related to the tenet of case-based similarity of \cite{gilboa1995case} -- see \cite{billot2008axiomatization} for an axiomatization of exponential similarity.}
 \par
@@
 For concreteness, we suppose that $\sigma = 1$ and $\rho = 0.5$, but the intuitions hold for any value of $\sigma$ and $\rho > 0$.
-First, we consider the case when $y = 0.5 > 0$.
+First, we consider the case when the realization of $y > 0$ and, for concreteness, specifically when $y = 0.5$ though the general intuitions hold for any $y > 0$.
 The resulting beliefs after observing $y$ are then as follows:
@@
 Thus, since $y > 0$, the user in the next period will consume items nearby to item 0 since, even though initially she believed that all items had the same mean, the spillover from consuming item 0 leads her to believe that items 1 and 3 have higher expected valuations.
-Since both the mean is higher for these items and the variance is lower, the user will consume items 1 and 3 regardless of her risk aversion levels.
+Since both the mean is higher for these items and the variance is lower, the user will consume items 1 and 3 regardless of her risk aversion level.
 \par 
-Suppose instead that item 0 ends up having a negative valuation so that $y = -0.5 < 0$.
+Now we consider the case when item 0 ends up having a negative valuation so that $y = -0.5 < 0$.
 This results in $\mathbb{E}[x_1] = \mathbb{E}[x_3] = -\frac{1}{4} <  -\frac{1}{8} = \mathbb{E}[x_2]$ with $\bar{\Sigma}$ remaining the same as when $y = 0.5$.
@@
 We simulate over risk-aversion parameters $\gamma \in \{ 0, 0.3, 0.6, 1, 5 \}$, \ standard-deviation $\sigma \in \{ 0.25, 0.5, 1.0, 2.0, 4.0 \}$, correlation parameters $\rho\in \{ 0, 0.1, 0.3, 0.5, 0.7, 0.9 \} $ and degree of idiosyncrasy of individual valuations $\beta \in \{ 0, 0.4, 0.8, 1, 2, 5\}$.
-We argue that these parameters are representative.
-First, the values of $\gamma$ that we consider range from no risk aversion ($\gamma = 0$) to reasonably high risk aversion levels $(\gamma = 5)$.
-The correlation parameter, $\rho$, ranges from uncorrelated valuations ($\rho = 0$) to nearly fully correlated valuations $(\rho = 0.9)$.
-$\beta$ ranges from the common value component playing no role $(\beta = 0)$ to the idiosyncratic component dominating the true valuations $(\beta = 5)$.
-Finally, the standard deviation ranges from little uncertainty $(\sigma = 0.25)$ to a large amount of uncertainty $(\sigma = 4)$.
-Thus, the range of considered parameter values cover the relevant portions of the parameter space in order to provide full insight into the behavior of the model.
+The range of considered parameter values cover the relevant portions of the parameter space in order to provide full insight into the behavior of the model.
 \par
@@
 We report results for a relatively small consumption history $T=20$ with a product space size $N=200$.
-Unless otherwise reported, our results are robust to varying $N$.
+Unless otherwise reported, our results are robust to increasing $N$.
 \par
@@
 
-First, the bottom panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho = 0$, there is no difference in consumption distance between the three regimes.
+First, the right panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho = 0$, there is no difference in consumption distance between the three regimes.
 This is due to the fact that when $\rho = 0$, there is no reason that items that are close in the product space should have similar values and so the optimal consumption path does not depend on the similarity of the items.
@@
 
-\begin{figure}[t]
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20.pdf}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_partial.pdf}
-\end{subfigure}\\
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_omni.pdf}\\
-\end{subfigure}
-\caption{Extent of Local Consumption as strength of correlation, $\rho$, varies. \\ No Recommendation (Top Left), Recommendation (Top Right) and Oracle (Bottom)}
-\label{fig:local_consumption_across_rho}
-\end{figure}
-
-\begin{figure}[t]
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20.pdf}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_partial.pdf}
-\end{subfigure}\\
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_omni.pdf}\\
-\end{subfigure}%
-\caption{Effect of Risk Aversion on Local Consumption \\ No Recommendation (Top Left), Recommendation (Top Right) and Oracle (Bottom)}
-\label{fig:no_rec_risk_aversion}
-\end{figure}
-
 In sum, filter bubble effects can only arise when there is an inherent correlation between the realized utilities of the items in the product space.
@@
 $$D_i:=\frac{1}{N}\frac{1}{T(T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$$
-
-\begin{figure}[ht]
-\includegraphics[width=.45\linewidth]{figures/gamma_diversity_N_200_T_20}
-\captionof{figure}{Risk Aversion and Diversity}\label{fig:risk_aversion_diversity}
-\end{figure}
 
@@
 
-\begin{figure}
-\begin{subfigure}{.45\linewidth}
-  \includegraphics[width=.8\linewidth]{figures/beta_diversity_N_200_T_20.pdf}
-\end{subfigure}
-\begin{subfigure}{.45\linewidth}
-  \includegraphics[width=.8\linewidth]{figures/beta_welfare_N_200_T_20.pdf}
-\end{subfigure}
-\caption{User Welfare and Diversity as the strength of the common value component, $\beta$, varies}\label{fig:diversity_welfare_common_value}
-\end{figure}
-
 \begin{finding}\label{finding_diversity_welfare_corr}
@@
 \end{subfigure}\\
-\caption{Diversity vs. Welfare, No-Recommendation, $\gamma = 0$ (Left), Diversity vs. Welfare, $\gamma = 5$ (Right)}\label{fig:diversity_welfare_ra}
+\caption{Diversity vs. Welfare, \textbf{No-Recommendation}, $\gamma = 0$ (Left), Diversity vs. Welfare, $\gamma = 5$ (Right)}\label{fig:diversity_welfare_ra}
 \end{figure}
@@
 \end{subfigure}\\
-\caption{Diversity vs. Welfare, Recommendation, $\gamma = 0$ (Left), Diversity vs. Welfare, $\gamma = 5$ (Right)}\label{fig:diversity_welfare_ra_partial}
+\caption{Diversity vs. Welfare, \textbf{Recommendation}, $\gamma = 0$ (Left), Diversity vs. Welfare, $\gamma = 5$ (Right)}\label{fig:diversity_welfare_ra_partial}
 \end{figure}
@@
 
-\begin{figure}[t]
-\includegraphics[width=.45\linewidth]{figures/beta_homogeneity_N_200_T_20}
-\captionof{figure}{Strength of Recommendation and Homogeneity}\label{fig:beta_homo}
-\end{figure}
-~
+\section{Recommender System Evaluation}
+In this section we discuss how the insights from our model of user decision-making can be used to inform the evaluation and design of recommender systems.
+\par
 
-\begin{figure}[t]
-\includegraphics[width=.45\linewidth]{figures/rho_homogeneity_N_200_T_20}
-\captionof{figure}{Correlation and Homogeneity}\label{fig:cor_homo}
-\end{figure}
+The classic approach to recommender system evaluation is to predict user ratings for items and compare how accurate this prediction is to recorded ratings data, either explicitly given by users or inferred from behavioral data.
+Further, given this predicted rating, the recommender system should recommend to users the items with the highest predicted rating \cite{adomavicius2005toward}.
+\par
+
+There has been a recent movement away from such evaluation measures due to the observation that accurate recommendations are not necessarily useful recommendations \cite{mcnee2006being}.
+Our model illustrates why this could be the case.
+Consider the domain of movie recommendation and suppose a user has just watched the movie \textit{John Wick} and rated it highly.
+A recommender system attempting to predict user ratings may then predict that this user is very likely to enjoy the sequel, \textit{John Wick: Chapter Two}, as well.
+However, the user herself may also have made this inference since the two movies are very similar to each other and, after enjoying \textit{John Wick}, the user will update her beliefs about \textit{John Wick: Chapter Two}.
+Thus, if the recommender system recommends this movie to the user then this recommendation is not useful since the recommendation gives the user little information that they did not already know.
+The key insight is that the recommendation is not useful since \textit{it ignores the inference the user themselves made and their updated beliefs}.
+The user may watch \textit{John Wick: Chapter Two}, then, even without recommendation and the value of the recommendation was small.
+\par
+
+As a result, several alternative recommendation evaluation metrics, such as serendipity \cite{kotkov2016survey}, coverage \cite{ge2010beyond}, novelty \cite{vargas2011rank}, and many others have been proposed in order to develop recommender systems that produce recommendations that are more useful for users.
+We use our model to motivate an alternative recommendation evaluation metric that provides an alternative perspective on the serendipity metric proposed in the literature.
+Serendipitous recommendations are said to ``have the quality of being both unexpected and useful" \cite{maksai2015predicting} and we use this as a guide towards our definition.
+
+Our approach to serendipity comes from the intuition that a recommendation is only useful insofar as it leads a user to take a better action than she would without recommendation.
+This approach stems from the idea that the value of recommendation is the marginal utility gain that a user receives from recommendation.\footnote{This notion is similar to the ``value of information" notion utilized in the literature in economics on the pricing of information goods \cite{bergemann2018design}.} This intuitively should come from providing information about products that, without additional information, users assign a low likelihood to being the best products currently available for them to consume and so would not consume them without these being recommended.
+\par
+
+This differs from other approaches in the literature since it depends on the beliefs of the users and what item the user would consume without recommendation.
+For instance, \cite{vargas2011rank, kaminskas2014measuring} propose unexpectedness metrics that look at the dissimilarity of the recommended items to what the recommender already knows the user likes either via a content-based approach or a collaborative-based approach.
+This depends only on the resulting item-set that the users get recommended and not necessarily user beliefs or the change in user action.
+\cite{adamopoulos2014unexpectedness} use an expected utility approach but use this to measure unexpectedness in terms of deviations from the ``expected consideration set" of a user and do not explicitly incorporate user beliefs.
+\par
+
+Our model allows us to give precise definitions for what it means for a recommendation to be \textit{unexpected} and \textit{useful}.
+Unexpected recommendations are given by recommended items that, given their beliefs, users have low expected utility from consuming, irregardless of what the true realized value is.
+Recall from Findings \ref{finding_local_consumption} and \ref{finding_diversity_welfare_corr} that, particularly if users are risk-averse, they will consume content in increasingly narrow portions of the product space and this may be detrimental to their realized welfare.
+As a result, identifying the portions of the product space where users may have significant uncertainty and utilizing recommendation to reduce this uncertainty can be welfare-enhancing for users.
+This is especially the case if this reduction of uncertainty leads users to explore under-explored portions of the product space where true realized utility is high, but expected utility, given beliefs, is low before recommendation.
+\par
+
+Useful recommendations are those that give users information that would change the item they consume relative to no recommendation and lead to an increase in expected utility given the updated beliefs when comparing the planned choice before and after the recommendation.
+In the previous example, recommending \textit{John Wick: Chapter Two} is not useful precisely because the expected utility before and after recommendation is the same since it does not change behavior.
+The most useful recommendations are those that (1) are informative given user beliefs and (2) induce users to consume unexpected goods.
+\par
+
+Operationalizing this approach requires the collection of data that is not traditionally collected for recommender systems.
+In particular, it becomes important to understand what choices users would make \textit{without recommendation} and to collect data on user beliefs, which can partially be inferred by observed choices.
+\cite{jiang2014choice, saavedra2016choice} argue that choice-based approaches alone can help in designing better RS, though they argue for these approaches because they do a better job at providing more accurate recommendations.
+Instead, we stress that choice-based approaches are useful not as recommendations but as a baseline for what recommendations would be useful.\footnote{Additionally, user beliefs and user choices contain information that may not have been observed by the recommender but is observed by the user.
+Ratings, reviews, friends, there are many sources of information affecting a person's beliefs and choices that are unobservable by recommender systems.
+Other consumption choices, e.g.
+movies seen in the cinema, change beliefs about for instance how good a director is, which affect beliefs about other movies and guide choices on a streaming platform that is unable to observe these data.
+However, this information can be inferred by the platform by collecting both beliefs and consumption choices.}
+\par
+
+Furthermore, our model shows the importance of understanding the underlying user preferences and the nature of the product space in the setting in which a recommender system is deployed since that will change the consumption patterns of users.
+For instance, understanding the degree of correlation of utilities and the level of risk-aversion become crucial in understanding how users will make consumption choices without recommendation and what kind of recommendation would be useful for users.
+\par
 
@@
 \par 
-Our results highlight that user choice in these markets is path dependent as a result of the nature of user preferences and how these preferences evolve over time.
-This has important ramifications for understanding the extent to which recommender systems can steer user behavior, potentially in anti-competitive ways.
-Steering users towards certain items today provides them with information that they utilize to make decisions tomorrow, which leads them to consume similar items to those they were originally steered towards.
-Thus, online platforms can potentially shape the long-term preferences of users and not only impact their current period choices.
-This has broader ramifications about the equilibrium effects of such behavior on competition and consumer welfare in online platforms which we leave to future work.
-\par
 We leave for future work exploring to what extent user behavior changes if we drop the myopia assumption, so that users themselves are engaging in explicit exploration.
@@
 Overall, understanding how users make choices in online platforms and how recommender systems impact these choices is important to understand.
-Understanding these issues allows us to better understand the broader social and economic impact of such systems as well as can provide a guide to the improved design of such systems.
+Understanding these issues allows us to better understand the broader social and economic impact of such systems as well as can provide a guide to improve the design of such systems.
 % Bibliography

### 2020-05-06 paper/filter_bubbles.tex
nix conclusion

@@
 \end{align*}
-\begin{figure}[t]
-\includegraphics[width=.3\linewidth]{figures/Example-Bubbles.pdf}
-\captionof{figure}{Illustrative Example}
-\label{fig:illustrative_example}
-\end{figure}
 
@@
 For instance, understanding the degree of correlation of utilities and the level of risk-aversion become crucial in understanding how users will make consumption choices without recommendation and what kind of recommendation would be useful for users.
-\par
 
-\section{Conclusion}
-We have studied a model of user decision-making in the context of markets where recommender systems are traditionally deployed in order to understand the consequences of recommender systems on consumption patterns of users.
-The key components of the model are that users make decisions under uncertainty and that consuming a good leads to informational spillovers about similar items.
-We have shown that this alone can generate consumption behavior that is consistent with filter bubble effects that have traditionally been associated with personalized recommender systems.
-This effect is weakened once we allow for recommendation, but not as much as if users have full information.
-We have further explored the relationship between consumption diversity and user welfare and found that they are negatively correlated, implying that striving for consumption diversity is not necessarily welfare-improving for users.
-Finally, we have shown that recommendation can lead to increases in user homogenization both relative to no recommendation and full information since it concentrates users' consumption in portions of the product space with the highest utility according to the limits of what the recommender can predict about the utility of the good.
-\par 
-We leave for future work exploring to what extent user behavior changes if we drop the myopia assumption, so that users themselves are engaging in explicit exploration.
-This likely interacts with recommendation in interesting ways as the recommendations may provide information that can complement or crowd-out user-led exploration.
-Even under the myopia assumption, understanding which recommendation system is optimal is an interesting and challenging problem.
-Further, our results point to an important consideration in the literature on incentivizing exploration which is that the exploration faced by the recommender itself likely interacts in interesting and important ways with the choice problem faced by the users.
-\par 
-Overall, understanding how users make choices in online platforms and how recommender systems impact these choices is important to understand.
-Understanding these issues allows us to better understand the broader social and economic impact of such systems as well as can provide a guide to improve the design of such systems.
 % Bibliography

### 2020-05-17 paper/appendix.tex
fix the figures in the paper, add appendix, and rewrite section 4

@@
-\documentclass[format=acmsmall, review=false]{acmart}
-\usepackage{acm-ec-20}
+\documentclass[format=acmsmall, review=true]{acmart}
+\usepackage{acm-tweaks}
 \usepackage{booktabs} % For formal tables
@@
 \IncMargin{-\parindent}
-
 % Choose a citation style by commenting/uncommenting the appropriate line:
@@
 % Title. Note the optional short title for running heads. In the interest of anonymization, please do not include any acknowledgements.
-\newcommand\PaperTitle{Appendix For Deconstructing the Filter Bubble: \\ Consumer Decision-Making and Recommender Systems}
+\newcommand\PaperTitle{Appendix For Deconstructing the Filter Bubble: Consumer Decision-Making and Recommender Systems}
 \title[\PaperTitle]{\PaperTitle}
@@
 
-% Paper body
+\section{Omitted Figures}
+
+This section contains figures that were omitted from the main text due to space constraints.
+Thus, the figures in this section are computed utilizing the same parameter values described in the main text for $N = 200$.
+
+\begin{figure}[ht]
+\caption{Relationship between $\beta$ and Homogeneity, $N = 200$}
+\includegraphics[width=.45\linewidth]{figures/beta_homogeneity_N_200_T_20}\label{fig:beta_homo}
+\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the weight of the common value component, $\beta$.
+Each line represents this plot for a single recommendation regime.}
+\end{figure}
+\begin{figure}[ht]
+\caption{Relationship between $\rho$ and Homogeneity, $N = 200$}
+\includegraphics[width=.45\linewidth]{figures/rho_homogeneity_N_200_T_20}
+\caption*{\scriptsize Notes: Notes: This figure displays the value of the homogeneity measure as we vary the inherent correlation between the valuation of the items, $\rho$.
+Each line represents this plot for a single recommendation regime.}\label{fig:cor_homo}
+\end{figure}
+
 \begin{figure}[H]
+\caption{Relationship between Local Consumption and $\rho$, $N = 200$}
 \begin{subfigure}{.45\textwidth}
@@
 \end{subfigure}
-\caption{Extent of Local Consumption as strength of correlation, $\rho$, varies. \\ No Recommendation (Top Left), Recommendation (Top Right) and Oracle (Bottom)}
+\caption*{\scriptsize Notes: Each figure plots the average consecutive consumption distance across time as the inherent correlation between the valuation of the items, $\rho$.
+The top left displays the no recommendation regime, the top right displays the recommendation regime, and the bottom displays the oracle regime.}
 \label{fig:local_consumption_across_rho}
@@
 \begin{figure}[H]
+\caption{Relationship between Local Consumption and $\gamma$, $N = 200$}
 \begin{subfigure}{.45\textwidth}
@@
 \end{subfigure}%
-\caption{Effect of Risk Aversion ($\gamma$) on Local Consumption \\ No Recommendation (Top Left), Recommendation (Top Right) and Oracle (Bottom)}
+\caption*{\scriptsize Notes: Each figure plots the average consecutive consumption distance across time as the risk aversion level of users, $\gamma$, varies.
+The top left displays the no recommendation regime, the top right displays the recommendation regime, and the bottom displays the oracle regime.}
 \label{fig:no_rec_risk_aversion}
@@
 
-\begin{figure}
-\begin{subfigure}{.45\linewidth}
-  \includegraphics[width=.8\linewidth]{figures/beta_diversity_N_200_T_20.pdf}
-\end{subfigure}
-\begin{subfigure}{.45\linewidth}
-  \includegraphics[width=.8\linewidth]{figures/beta_welfare_N_200_T_20.pdf}
-\end{subfigure}
-\caption{User Welfare and Diversity as the strength of the common value component, $\beta$, varies}\label{fig:diversity_welfare_common_value}
+\clearpage
+\section{N = 100}
+This section contains the same figures as reported in the main text and appendix, but for $N = 100$ instead of $N = 200$.
+
+\begin{figure}[ht]
+\caption{Relationship between $\beta$ and Homogeneity, $N = 100$}
+\includegraphics[width=.45\linewidth]{figures/beta_homogeneity_N_200_T_20}\label{fig:beta_homo}
+\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the weight of the common value component, $\beta$.
+Each line represents this plot for a single recommendation regime.}
+\end{figure}
+\begin{figure}[ht]
+\caption{Relationship between $\rho$ and Homogeneity, $N = 100$}
+\includegraphics[width=.45\linewidth]{figures/rho_homogeneity_N_100_T_20}
+\caption*{\scriptsize Notes: Notes: This figure displays the value of the homogeneity measure as we vary the inherent correlation between the valuation of the items, $\rho$.
+Each line represents this plot for a single recommendation regime.}\label{fig:cor_homo}
 \end{figure}
 
-\begin{figure}[t]
-\includegraphics[width=.45\linewidth]{figures/beta_homogeneity_N_200_T_20}
-\captionof{figure}{Strength of Recommendation ($\beta$) and Homogeneity}\label{fig:beta_homo}
+\begin{figure}[H]
+\caption{Relationship between Local Consumption and $\gamma$, $N = 100$}
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_100T_20.pdf}
+\end{subfigure}
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_100T_20_partial.pdf}
+\end{subfigure}\\
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_100T_20_omni.pdf}\\
+\end{subfigure}%
+\caption*{\scriptsize Notes: Each figure plots the average consecutive consumption distance across time as the risk aversion level of users, $\gamma$, varies.
+The top left displays the no recommendation regime, the top right displays the recommendation regime, and the bottom displays the oracle regime.}
+\label{fig:no_rec_risk_aversion}
 \end{figure}
-\begin{figure}[t]
-\includegraphics[width=.45\linewidth]{figures/rho_homogeneity_N_200_T_20}
-\captionof{figure}{Correlation ($\rho$) and Homogeneity}\label{fig:cor_homo}
+
+\begin{figure}[H]
+\caption{Local Consumption and Correlation, $N = 100$}
+\begin{subfigure}{.5\linewidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{figures/rho_pos_consumption_dist_N_100T_20_overall.pdf}
+  \label{fig:sfig1}
+\end{subfigure}%
+\begin{subfigure}{.5\linewidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{figures/rho_zero_consumption_dist_N_100T_20_overall.pdf}
+  \label{fig:sfig2}
+\end{subfigure}
+\caption*{\footnotesize Notes: The figure shows the consecutive consumption path difference between the no recommendation, recommendation, and oracle regime.
+The figure on the left displays the average consecutive consumption distance aggregating over simulations where $\rho = 0$ and the figure on the right displays the average consecutive consumption distance aggregating over simulations where $\rho > 0$}
+\label{fig:correlation_consumption_path_n_100}
+\end{figure}
+
+% Paper body
+\begin{figure}[H]
+\captionof{figure}{Relationship between Local Consumption and $\rho$, $N = 100$}
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_100T_20.pdf}
+\end{subfigure}
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_100T_20_partial.pdf}
+\end{subfigure}\\
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_100T_20_omni.pdf}\\
+\end{subfigure}
+\caption*{\scriptsize Notes: Each figure plots the average consecutive consumption distance across time as the inherent correlation between the valuation of the items, $\rho$.
+The top left displays the no recommendation regime, the top right displays the recommendation regime, and the bottom displays the oracle regime.}
+\label{fig:local_consumption_across_rho}
+\end{figure}
+
+\clearpage
+
+\section{N = 500}
+
+This section contains the same figures as reported in the main text and appendix, but for $N = 500$ instead of $N = 200$.
+
+\begin{figure}[ht]
+\caption{Relationship between $\beta$ and Homogeneity, $N = 500$}
+\includegraphics[width=.45\linewidth]{figures/beta_homogeneity_N_200_T_20}\label{fig:beta_homo}
+\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the weight of the common value component, $\beta$.
+Each line represents this plot for a single recommendation regime.}
+\end{figure}
+\begin{figure}[ht]
+\caption{Relationship between $\rho$ and Homogeneity, $N = 500$}
+\includegraphics[width=.45\linewidth]{figures/rho_homogeneity_N_100_T_20}
+\caption*{\scriptsize Notes: Notes: This figure displays the value of the homogeneity measure as we vary the inherent correlation between the valuation of the items, $\rho$.
+Each line represents this plot for a single recommendation regime.}\label{fig:cor_homo}
+\end{figure}
+
+\begin{figure}[H]
+\captionof{figure}{Relationship between Local Consumption and $\gamma$, $N = 500$}
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_100T_20.pdf}
+\end{subfigure}
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_100T_20_partial.pdf}
+\end{subfigure}\\
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_100T_20_omni.pdf}\\
+\end{subfigure}%
+\caption*{\scriptsize Notes: Each figure plots the average consecutive consumption distance across time as the risk aversion level of users, $\gamma$, varies.
+The top left displays the no recommendation regime, the top right displays the recommendation regime, and the bottom displays the oracle regime.}
+\label{fig:no_rec_risk_aversion}
+\end{figure}
+
+% Paper body
+\begin{figure}[H]
+\caption{Relationship between Local Consumption and $\rho$, $N = 500$}
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_100T_20.pdf}
+\end{subfigure}
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_100T_20_partial.pdf}
+\end{subfigure}\\
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_100T_20_omni.pdf}\\
+\end{subfigure}
+\caption*{\scriptsize Notes: Each figure plots the average consecutive consumption distance across time as the inherent correlation between the valuation of the items, $\rho$.
+The top left displays the no recommendation regime, the top right displays the recommendation regime, and the bottom displays the oracle regime.}
+\label{fig:local_consumption_across_rho}
+\end{figure}
+
+\begin{figure}[H]
+\caption{Local Consumption and Correlation, $N = 500$}
+\begin{subfigure}{.5\linewidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{figures/rho_pos_consumption_dist_N_100T_20_overall.pdf}
+  \label{fig:sfig1}
+\end{subfigure}%
+\begin{subfigure}{.5\linewidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{figures/rho_zero_consumption_dist_N_100T_20_overall.pdf}
+  \label{fig:sfig2}
+\end{subfigure}
+\caption*{\scriptsize Notes: The figure shows the consecutive consumption path difference between the no recommendation, recommendation, and oracle regime.
+The figure on the left displays the average consecutive consumption distance aggregating over simulations where $\rho = 0$ and the figure on the right displays the average consecutive consumption distance aggregating over simulations where $\rho > 0$}
+\label{fig:correlation_consumption_path_n_100}
 \end{figure}
@@
 % Bibliography
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{refs}
 

### 2020-05-17 paper/filter_bubbles.tex
fix the figures in the paper, add appendix, and rewrite section 4

@@
 \documentclass[format=acmsmall, review=false]{acmart}
-\usepackage{acm-ec-20}
+\settopmatter{printacmref=false, printccs=false}
 \usepackage{booktabs} % For formal tables
@@
 % Anonymized submission.
-\author{Submission 20}
 
@@
 
-\xhdr{Model} We consider a model of user choice that has the following components.
+\xhdr{Our Model} We consider a model of user choice that has the following components.
 \par
@@
 In particular, even users that will end up having the same ex-post valuation of item $n$ may differ in their ex-ante valuation because they hold different beliefs about it.
-We denote by $p_{i}$ the beliefs item $i$ has about how she will value each of the items in the product space.
+We denote by $p_{i}$ the beliefs user $i$ has about how she will value each of the items in the product space.
 Note that this implies that consuming item $n$ is the same as taking a gamble.
@@
 
-Risk aversion captures the how different users react to the risk associated to a particular consumption opportunity.
+Risk aversion captures how different users react to the risk associated to a particular consumption opportunity.
 It is formalized as follows: a given gamble $x$ takes real values and follows distribution $p$.
@@
 We will consider three cases.
-The case of primary interest is \textit{recommendation} where the recommender observes values accrued and knows $V$ but does not know neither $V_i$.\footnote{We do not consider the acquisition of information for the recommender to know $V$ and suppose that she has sufficient data to learn $V$ with arbitrary precision at $t = 0$.} However, the recommender does know the users' beliefs $\bar V_i$.
+The case of primary interest is \textit{recommendation} where the recommender observes values accrued and knows $V$ but does not know $V_i$.\footnote{We do not consider the acquisition of information for the recommender to know $V$ and suppose that she has sufficient data to learn $V$ with arbitrary precision at $t = 0$.} However, the recommender does know the users' beliefs $\bar V_i$.
 Thus, at any given period, the recommender provides a personalized recommendation that combines the knowledge of the common value component $V$ with the user beliefs $\bar V_i$.
@@
 We report results for a relatively small consumption history $T=20$ with a product space size $N=200$.
-Unless otherwise reported, our results are robust to increasing $N$.
 \par
@@
 \begin{figure}
+\caption{Local Consumption and Correlation, $N = 200$}
 \begin{subfigure}{.5\linewidth}
@@
 \end{subfigure}
-\caption{Consumption path with (Left) and without (Right) correlation between values}
+\caption*{\scriptsize Notes: The figure shows the consecutive consumption path difference between the no recommendation, recommendation, and oracle regime.
+The figure on the left displays the average consecutive consumption distance aggregating over simulations where $\rho = 0$ and the figure on the right displays the average consecutive consumption distance aggregating over simulations where $\rho > 0$}
 \label{fig:correlation_consumption_path}
@@
 
-In sum, filter bubble effects can only arise when there is an inherent correlation between the realized utilities of the items in the product space.
-However, this is often the case in most environments where recommender systems are deployed.
+In sum, filter bubble effects only arise when there is an inherent correlation between the realized utilities of the items in the product space.
 When there is a correlation between the realized utilities then filter bubbles can naturally arise due to the nature of how individuals acquire additional information about the remaining items.
@@
 \begin{figure}
+\caption{Relationship between User Welfare and Diversity, varying $\rho$}\label{fig:diversity_welfare_correlation}
 \begin{subfigure}{.45\linewidth}
@@
 \end{subfigure}
-\caption{User Welfare and Diversity as the strength of correlation, $\rho$, varies}\label{fig:diversity_welfare_correlation}
+\caption*{\scriptsize Notes: The figure on the left displays the relationship between $rho$, the correlation between valuations of items, and overall consumed product diversity.
+The figure on the right displays the relationship between $\rho$ and overall welfare.}
 \end{figure}
@@
 \begin{figure}[t]
+\caption{Diversity vs. Welfare, \textbf{No Recommendation}}
 \begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_rn.pdf}
+\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn.pdf}
 \end{subfigure}
 \begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_ra.pdf}
+\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra.pdf}
 \end{subfigure}\\
-\caption{Diversity vs. Welfare, \textbf{No-Recommendation}, $\gamma = 0$ (Left), Diversity vs. Welfare, $\gamma = 5$ (Right)}\label{fig:diversity_welfare_ra}
+\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under the no recommendation regime.
+The figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.}\label{fig:diversity_welfare_ra}
 \end{figure}
@@
 \begin{figure}[t]
+\caption{Diversity vs. Welfare, \textbf{Recommendation}}
 \begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_rn_partial.pdf}
+\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_partial.pdf}
 \end{subfigure}
 \begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.05\linewidth]{figures/diversity_welfare_ra_partial.pdf}
+\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_partial.pdf}
 \end{subfigure}\\
-\caption{Diversity vs. Welfare, \textbf{Recommendation}, $\gamma = 0$ (Left), Diversity vs. Welfare, $\gamma = 5$ (Right)}\label{fig:diversity_welfare_ra_partial}
+\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under the recommendation regime.
+The figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.}\label{fig:diversity_welfare_ra_partial}
 \end{figure}
@@
 \section{Recommender System Evaluation}
-In this section we discuss how the insights from our model of user decision-making can be used to inform the evaluation and design of recommender systems.
+In this section we discuss how the insights from our model of user decision-making can inform the evaluation and design of recommender systems.
+The classic approach to recommender system evaluation is predicting user ratings for items and comparing how accurate this prediction is to recorded ratings data, either explicitly given by users or inferred from behavioral data.
+The recommender system should then recommend to users the items with the highest predicted rating \cite{adomavicius2005toward}.
 \par
-
-The classic approach to recommender system evaluation is to predict user ratings for items and compare how accurate this prediction is to recorded ratings data, either explicitly given by users or inferred from behavioral data.
-Further, given this predicted rating, the recommender system should recommend to users the items with the highest predicted rating \cite{adomavicius2005toward}.
-\par
-
 There has been a recent movement away from such evaluation measures due to the observation that accurate recommendations are not necessarily useful recommendations \cite{mcnee2006being}.
-Our model illustrates why this could be the case.
+Our model illustrates a mechanism behind this observation.
 Consider the domain of movie recommendation and suppose a user has just watched the movie \textit{John Wick} and rated it highly.
 A recommender system attempting to predict user ratings may then predict that this user is very likely to enjoy the sequel, \textit{John Wick: Chapter Two}, as well.
-However, the user herself may also have made this inference since the two movies are very similar to each other and, after enjoying \textit{John Wick}, the user will update her beliefs about \textit{John Wick: Chapter Two}.
+However, the user herself may also have made this inference since the two movies are very similar to each other.
 Thus, if the recommender system recommends this movie to the user then this recommendation is not useful since the recommendation gives the user little information that they did not already know.
 The key insight is that the recommendation is not useful since \textit{it ignores the inference the user themselves made and their updated beliefs}.
-The user may watch \textit{John Wick: Chapter Two}, then, even without recommendation and the value of the recommendation was small.
+The user may watch \textit{John Wick: Chapter Two}, then, even without recommendation, and the value of the recommendation was small.
 \par
-
-As a result, several alternative recommendation evaluation metrics, such as serendipity \cite{kotkov2016survey}, coverage \cite{ge2010beyond}, novelty \cite{vargas2011rank}, and many others have been proposed in order to develop recommender systems that produce recommendations that are more useful for users.
-We use our model to motivate an alternative recommendation evaluation metric that provides an alternative perspective on the serendipity metric proposed in the literature.
-Serendipitous recommendations are said to ``have the quality of being both unexpected and useful" \cite{maksai2015predicting} and we use this as a guide towards our definition.
-
-Our approach to serendipity comes from the intuition that a recommendation is only useful insofar as it leads a user to take a better action than she would without recommendation.
-This approach stems from the idea that the value of recommendation is the marginal utility gain that a user receives from recommendation.\footnote{This notion is similar to the ``value of information" notion utilized in the literature in economics on the pricing of information goods \cite{bergemann2018design}.} This intuitively should come from providing information about products that, without additional information, users assign a low likelihood to being the best products currently available for them to consume and so would not consume them without these being recommended.
-\par
-
-This differs from other approaches in the literature since it depends on the beliefs of the users and what item the user would consume without recommendation.
-For instance, \cite{vargas2011rank, kaminskas2014measuring} propose unexpectedness metrics that look at the dissimilarity of the recommended items to what the recommender already knows the user likes either via a content-based approach or a collaborative-based approach.
-This depends only on the resulting item-set that the users get recommended and not necessarily user beliefs or the change in user action.
-\cite{adamopoulos2014unexpectedness} use an expected utility approach but use this to measure unexpectedness in terms of deviations from the ``expected consideration set" of a user and do not explicitly incorporate user beliefs.
-\par
-
-Our model allows us to give precise definitions for what it means for a recommendation to be \textit{unexpected} and \textit{useful}.
-Unexpected recommendations are given by recommended items that, given their beliefs, users have low expected utility from consuming, irregardless of what the true realized value is.
-Recall from Findings \ref{finding_local_consumption} and \ref{finding_diversity_welfare_corr} that, particularly if users are risk-averse, they will consume content in increasingly narrow portions of the product space and this may be detrimental to their realized welfare.
-As a result, identifying the portions of the product space where users may have significant uncertainty and utilizing recommendation to reduce this uncertainty can be welfare-enhancing for users.
-This is especially the case if this reduction of uncertainty leads users to explore under-explored portions of the product space where true realized utility is high, but expected utility, given beliefs, is low before recommendation.
-\par
-
-Useful recommendations are those that give users information that would change the item they consume relative to no recommendation and lead to an increase in expected utility given the updated beliefs when comparing the planned choice before and after the recommendation.
-In the previous example, recommending \textit{John Wick: Chapter Two} is not useful precisely because the expected utility before and after recommendation is the same since it does not change behavior.
-The most useful recommendations are those that (1) are informative given user beliefs and (2) induce users to consume unexpected goods.
-\par
-
-Operationalizing this approach requires the collection of data that is not traditionally collected for recommender systems.
-In particular, it becomes important to understand what choices users would make \textit{without recommendation} and to collect data on user beliefs, which can partially be inferred by observed choices.
-\cite{jiang2014choice, saavedra2016choice} argue that choice-based approaches alone can help in designing better RS, though they argue for these approaches because they do a better job at providing more accurate recommendations.
-Instead, we stress that choice-based approaches are useful not as recommendations but as a baseline for what recommendations would be useful.\footnote{Additionally, user beliefs and user choices contain information that may not have been observed by the recommender but is observed by the user.
-Ratings, reviews, friends, there are many sources of information affecting a person's beliefs and choices that are unobservable by recommender systems.
+This intuition implies that recommender systems should collect additional data beyond the type of data that is traditionally recorded.
+The first and most crucial type of data to collect is individual user \textit{beliefs} about items that they have not yet consumed.
+As illustrated by our model, these beliefs are what drive the future consumption decisions of users and understanding these beliefs is crucial for determining the value of recommending certain items.\footnote{Additionally, user beliefs contain information that may not have been observed by the recommender that only observes user choices on the platform.
+There are many sources of information that affect a user's beliefs and choices that are unobservable by recommender systems such as ratings, reviews, and friends.
 Other consumption choices, e.g.
 movies seen in the cinema, change beliefs about for instance how good a director is, which affect beliefs about other movies and guide choices on a streaming platform that is unable to observe these data.
-However, this information can be inferred by the platform by collecting both beliefs and consumption choices.}
+However, this information can be inferred by the platform by collecting both beliefs and consumption choices.} The second type of data that is critical for recommender system designers to collect is how user beliefs change over time and, in particular, not just how individuals value the item they just consumed, but also how it impacts their beliefs about the quality of similar items.
+The third type of data is the risk-aversion levels of users as our model illustrates that the risk preferences of users are important for understanding what information recommender systems can provide that materially leads users to alter their consumption patterns.
 \par 
-
-Furthermore, our model shows the importance of understanding the underlying user preferences and the nature of the product space in the setting in which a recommender system is deployed since that will change the consumption patterns of users.
-For instance, understanding the degree of correlation of utilities and the level of risk-aversion become crucial in understanding how users will make consumption choices without recommendation and what kind of recommendation would be useful for users.
-
+A natural follow-up question is how this additional data should be utilized in the design of good recommendations.
+Our model posits that recommendation provides value to users by providing them with information about the true valuation of an item if they were to consume it.
+Thus, the prediction problem for the recommender becomes predicting what item would the user choose with no recommendation and, correspondingly, what would be the most useful information to provide to the user that would lead her to consume a \textit{better} item than she would without recommendation.
+This links back to the intuition our model provided for the \textit{John Wick} example whereby collecting user beliefs and measuring how the user updated beliefs about similar goods would lead the recommender to understand that the user would consume \textit{John Wick: Chapter Two}.
+Our approach would therefore imply that, with this as a starting point, the recommender's problem would be to predict what is the most useful information to give the user leading them to change the item that they eventually consume.
+\par 
+There have been a number of alternative recommendation evaluation metrics proposed in the literature with the aim of providing more useful recommendations than those provided by accuracy metrics, such as serendipity \cite{kotkov2016survey}, coverage \cite{ge2010beyond}, novelty \cite{vargas2011rank}, and many others.
+Our approach most closely follows the set of proposed serendipity measures which are surveyed in \cite{kotkov2016survey}.
+As discussed by \cite{maksai2015predicting},  serendipitous recommendations are said to ``have the quality of being both unexpected and useful" which is in line with the primary intuition behind our approach.
+The primary difference between our proposed approach and those existing in the literature is that ours crucially hinges on understanding user beliefs and the risk-preferences of users.
+For instance, \cite{vargas2011rank, kaminskas2014measuring} propose unexpectedness metrics that look at the dissimilarity of the proposed recommended items compared to what the recommender already knows the user likes - either via a content-based approach or a collaborative-based approach.
+This metric depends only on the resulting item-set that the recommender proposes and not necessarily on the user's beliefs or how such a recommendation will change the item that the user consumes.
+\cite{adamopoulos2015unexpectedness} propose an expected utility approach but use this to measure unexpectedness in terms of deviations from the ``expected consideration set" of a user and do not explicitly incorporate user beliefs.
+\par 
+Indeed, our approach allows us to give precise definitions for what it means for a recommendation to be \textit{unexpected} and \textit{useful} in the spirit of serendipitous recommendations.
+Our evaluation measure leads to useful recommendations since it leads users towards better items than they would consume without recommendation.
+It further results in 	``unexpected" recommendations since it explicitly incorporates user beliefs and thus allows the recommender to understand how ``unexpected" a recommendation would be from the perspective of a user.
+Thus, a good recommendation provides users with information that leads them to consume an item that they would not have expected to consume without the information from the recommender and that they enjoyed consuming more than the item they would have without recommendation.
 % Bibliography

### 2020-05-17 paper/filter_bubbles.tex
updatred intro

@@
 \newcommand{\xhdr}[1]{\vspace{1mm} \noindent{\bf #1}}
-
+\renewcommand{\footnotetextcopyrightpermission}[1]{} % remove copyright block
+\renewcommand{\footnotetextauthorsaddresses}[1]{} % remove author address 
 \renewcommand{\algorithmcfname}{ALGORITHM}
@@
 
-Recommender Systems (RS) have become critical for assisting users in navigating the large choice sets that they face in many online markets.
+Recommender Systems (RS) have become critical for assisting users in navigating the large choice sets that they face in many online platforms.
 For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
-Users in many cases are not aware of most items, let alone have full information about their preferences over them and, to make matters worse, the goods in these markets are usually experience goods whose true value can only be learned after consumption.
-Furthermore, users interact with these systems routinely and watch more than one movie on Netflix, buy more than one product on Amazon, and listen to more than one video on YouTube.
+Users in many cases are not aware of most items, let alone have full information about their preferences over them.
+To make matters worse, the item in these contexts are usually experience goods whose true value can only be learned after consumption.
 \par
@@
 \par
+Understanding how recommender systems influence user behavior is important not only for characterizing the broader consequences of such systems but also for guiding their design.
+In this paper we develop a theoretical model of user decision-making in contexts where recommender systems are traditionally deployed.
+We utilize previous empirical studies that characterize how recommender systems influence user choice as a benchmark and our theoretical model provides an intuitive mechanism that provides an explanation for these empirical results.
+The key insight of our model is that individual \textit{beliefs} drive the consumption choices of individuals and that recommendations serve to provide users with information that leads them to update their beliefs.
+However, a crucial component of our model is that individuals' beliefs about items are driven not only by recommendations, but also from their previous experiences with similar items.
+We use these insights to provide guidance for recommender system designers.
+In order to design good recommendations it becomes crucial for designers to understand user beliefs of goods they haven't consumed thus far and how these beliefs evolve over time.
 
@@
 In our setting of interest users are long-lived, but they only consume a small fraction of this choice set over their lifetime.
+This is traditionally the case on online platforms that have thousands or millions of options for users.
 \par
-The second component of our model is that users have \textit{uncertainty} about their valuation of the different items.
-This is motivated both by the fact that many markets where recommender systems are deployed contain experience goods, whose true value can only be learned after consumption, and the fact that such uncertainty is why recommender systems exist in the first place.
+The second component of our model is that users have \textit{uncertainty} about their valuation of the different items before they consume them.
+This is motivated both by the fact that many contexts where recommender systems are deployed contain experience goods, whose true value can only be learned after consumption, and the fact that such uncertainty is why recommender systems exist in the first place.
 Thus, users face a sequential decision-making problem under uncertainty.
@@
 
-\xhdr{Contributions}
-Using this model we first investigate the degree to which recommender systems lead users into filter bubbles.
-Our model rationalizes the empirical results in \cite{nguyen2014exploring} who show that in the context of movie consumption, user behavior is consistent with filter bubble effects \textit{even without recommendation}.
-The driving force of this in our model is that preferences for similar goods are correlated, which implies that when a good is consumed and the user learns its value, it provides information about similar goods.
-Crucially, this not only impacts the underlying belief about the average utility of the good, but also the level of uncertainty.
-Consequently, this learning spillover induces users to consume goods ``similar'' to those they consumed before that had high realized value, leading to an increasing narrowing of consumption patterns towards these regions of the product space.
+\xhdr{Our Contributions}
+Our model provides a clear mechanism that explains the empirical results in \cite{nguyen2014exploring} who show that, in the context of movie consumption, user behavior is consistent with filter bubble effects even without recommendation and that recommendation leads to users being less likely to fall into such filter bubbles.
+In this context, filter bubble effects are defined as users consuming items in an increasingly narrow portion of the product space.
+The simple and intuitive driving force of this in our model is that preferences for similar items are correlated, which implies that when an item is consumed and the user learns its value, it provides information about similar item.
+Crucially, this not only impacts the underlying belief about the average utility of the item, but also the level of uncertainty.
+Consequently, this learning spillover induces users to consume items ``similar'' to those they consumed before that had high realized value, leading to an increasing narrowing of consumption patterns towards these regions of the product space.
 This effect is further amplified when users are \textit{risk-averse}, a concept from decision theory where all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty.
-We compare this benchmark to the case where recommendations are provided. 
-We consider recommendations as giving information to users in order to reduce their uncertainty.
-Recommendations could potentially make matters worse: providing information about the value of the goods when this value is correlated could intensify local search and decrease consumption diversity.
-However, and in line with the empirical results from \cite{nguyen2014exploring}, we find that recommendations reduce the narrowing effect.
+However, by providing information to users, recommendation leads users to be more likely to explore other portions of the product space, limiting the filter bubble effect.
 \par
-
-We then explore how diverse the overall set of items that users consume is.
-It could be expected that higher diversity being associated with more intense search is associated with higher welfare.
-Much to the opposite, we find that, without recommendation, higher diversity of items does not ensure higher welfare.
-This is due to the fact that once users find good items, they consume items nearby in the product space due to the correlation of values, leading to higher overall value from consumption.
-When they consume products they value little, they move to another part of the product space.
-A sequence of bad realizations will then lead to very intense search over very different parts of the product space and high consumption diversity, whereas discovering a sweet spot leads to highly rewarding local search and very low diversity.
-Thus, contrary to what a natural intuition could suggest, we find that consumption set diversity is not by itself a useful measure for understanding the impact of recommender systems on users.
+We find that, while recommendation leads a single user to be more likely to explore diverse portions of the product space, it also coordinates consumption choices across users.
+This leads to an increase in homogeneity across users, resulting in a tradeoff between homogenizing across user consumption and diversifying within user consumption.
+We finally explore the relationship between the overall diversity of consumed items and user welfare.
+We find that more diverse sets of consumed items do not always correspond to higher user welfare.
 \par
-
-We find that recommendation can lead to increases in homogeneity among users.
-We model the realized valuations as being a weighted sum of a common-value and an idiosyncratic component.
-This formulation gives a stylized notion of predictability of user preferences where the idiosyncratic component is inherently unpredictable given other individuals' preferences and the common-value component is what the recommonder can learn from previous users' data.
-The recommender uses its knowledge of the common-value component and the individuals' beliefs over the product space when designing personalized recommendation.
-This leads to an increase in welfare for users, but concentrates their consumption around items that have a high common-value component, resulting in an increase in homogeneity.
-\par
-
-Finally, we discuss how our model and findings can be used to inform the design and evaluation of recommender systems.
-Our evaluation metric relies on the intuition that a recommendation is only useful insofar as it leads a user to take a better action than she would without the recommendation.
-Thus, in order to evaluate the value of recommendation, it is important to be able to predict what a user would do without recommendation and how recommendation would change their behavior.
-If recommendations fail to affect behavior, they are of no use: the user would consume the same good and learn its value with and without the recommendation.
-Moreover, we argue that successful recommendation systems have to take into account the degree of uncertainty users have over different portions of the product space.
-Being able to predict valuations after consumption is not sufficient as uncertainty over the idiosyncratic component of one's own preferences may steer the user away from the recommended items in the face of risk-aversion.
+Finally, we discuss how our model and findings can be used to inform the design and evaluation of recommender systems as well as the data that is traditionally collected for them.
+Our model highlights the importance of user beliefs in determining the consumption choices of individuals and how both recommendation and informational spillovers determine how these beliefs change over time.
+By collecting information on user beliefs, recommender system designers can understand what items a user would consume \textit{without} recommendation and then predict how providing information to the user would change their beliefs and resulting consumption decisions.
+Thus, our evaluation measure determines the value of a recommendation based on the marginal welfare gain associated with providing a user with a recommendation over what the user would do without recommendation.
+We discuss how this provides an additional rationale as to why ``accurate" recommendations are not always good recommendations.
 \par
@@
 \textbf{Related Work.} 
-\cite{pariser2011filter} first informally described the ``filter bubble'' which is the idea that online personalization services would lead users down paths of increasingly narrower content so that they would effectively be isolated from a diversity of viewpoints or content.
-A number of studies, done in a wide range of disciplines, have since empirically studied whether or not this phenomenon exists in various contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}.
+The first set of related works studies the extent and implications of filter bubbles.
+\cite{pariser2011filter} first informally described the idea of the ``filter bubble'' which is that online personalization services would lead users down paths of increasingly narrower content so that they would effectively be isolated from a diversity of viewpoints or content.
+Following this a number of empirical studies, done in various disciplines, have since studied the extent to which this phenomenon exists in a wide range of contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}.
 The most relevant to our study is \cite{nguyen2014exploring} who study whether this effect exists with recommender systems in the context of movie consumption.
 They find that even users whose consumption choices are not guided by recommendations fall into behavior consistent with ``filter bubbles'' and that recommender systems can actually increase the diversity of the content that users consume.
-To our knowledge there are no theoretical models that rationalize the empirical findings but predominantly empirical studies.
-Our paper provides a theoretical framework through which to view this problem.
-The second question is to what extent recommender systems can lead users to become increasingly homogenized or consume similar content as each other.
-\cite{celam2008hits,treviranus2009value} show that incorporating content popularity into recommender systems can lead to increased user homogenization.
-\cite{chaney2018algorithmic} more generally look at the issue of user homogenization in recommender systems.
-Our results are similar to previous work, where the limits of recommendation lead the recommender to over-recommend products from certain portions of the product space which leads users to consume goods in those regions.
-\cite{fleder2009blockbuster} also consider the effects of recommendation on multi-period user behavior, but predominantly focus on sales diversity and employ a different model of user behavior than we do here.
-\cite{hodgson2019horse} consider a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of their preferences in the environment, but consider it in the context of search for a single item.
-Our discussion of recommender system evaluation contributes with insights closely tied to the value of information that is used in the economics literature (see e.g.
-\cite{bergemann2019information}) and builds off the large literature stemming from \cite{mcnee2006being} that makes the case that accurate predictions of true realized valuations may not lead to the best recommendations.
-Our evaluation measure is most closely related to the notion of serendipity in recommender systems that has been studied extensively (e.g.
-see \cite{kotkov2016survey} for a survey).
+To our knowledge there are no theoretical models that rationalize these empirical findings and our paper provides a theoretical framework through which to view this problem.
+Further, our paper provides a clear mechanism at play that drives such effects and how recommendation interacts with them.
+\par 
+Another set of papers has examined whether recommender systems can lead users to become increasingly homogenized or consume similar content to one another.
+\cite{celma2008hits, treviranus2009value} show that incorporating content popularity into recommender systems can lead to increased user homogenization.
+\cite{chaney2018algorithmic} show how user homogenization may arise from training recommender systems on data from users exposed to algorithmic recommendations.
+\cite{fleder2009blockbuster} show that homogenization can increase due to a recommendation bias towards popularity that arises from lack of information about items with limited consumption histories.
+Our model leads to similar results as previous theoretical and empirical work where recommender systems lead to increased user homogenization.
+However, the mechanisms behind this from our model differ from existing work as homogenization arises due to the fact that recommendation leads users to coordinate their consumption decisions in certain portions of the product space.
+\par
+Another set of papers in the recommender systems community that studies the impact of human decision-making on the design and evaluation of recommender systems.
+\cite{chen2013human} surveys the literature on the relationship between human decision making and recommender systems.
+The closest set of papers pointed out in this survey are those related to preference construction \cite{bettman1998constructive, lichtenstein2006construction} whereby users develop preferences over time through the context of a decision process.
+Our model points out that the true underlying preferences of users may be stable over time, but, due to the nature of items in contexts where recommender systems are deployed, they have incomplete information of them and both consumption and recommendation provide them with information to reduce their uncertainty.
+Thus, the primary insight of our paper is that individual beliefs and how individuals update their beliefs about similar items after consumption are important and previously unconsidered elements of human decision making that are critical for understanding the design and consequences of recommender systems.
+Within this literature, there is a set of papers that focus on ``user-centric" approaches to recommendation \cite{celma2008new, cremonesi2013user, pu2011user} whereby user evaluation of the usefulness of recommendation is a key evaluation measure.
+Our evaluation measure is similar, but, unlike previous approaches, emphasizes the importance of user beliefs.
+Finally, \cite{hodgson2019horse} consider a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of their preferences in the environment, but consider it in the context of search for a single item.
 
@@
 movies seen in the cinema, change beliefs about for instance how good a director is, which affect beliefs about other movies and guide choices on a streaming platform that is unable to observe these data.
-However, this information can be inferred by the platform by collecting both beliefs and consumption choices.} The second type of data that is critical for recommender system designers to collect is how user beliefs change over time and, in particular, not just how individuals value the item they just consumed, but also how it impacts their beliefs about the quality of similar items.
-The third type of data is the risk-aversion levels of users as our model illustrates that the risk preferences of users are important for understanding what information recommender systems can provide that materially leads users to alter their consumption patterns.
+However, this information can be inferred by the platform by collecting both beliefs and consumption choices.} The second type of data that is critical for recommender system designers to collect is how user beliefs change over time and, in particular, not just how individuals value the item they just consumed, but also how it impacts their beliefs about the quality of similar items.\footnote{Characterizing the similarity between items has been an important goal of designing content-based recommendations, though as noted by \cite{winecoff2019users}, how users perceive similarity between items is not always in line with how similarity is computed in content-based recommender systems.
+Further understanding how users perceive item similarity and, in particular, how this impacts which items users update their beliefs about is an important direction for future work.} The third type of data is the risk-aversion levels of users as our model illustrates that the risk preferences of users are important for understanding what information recommender systems can provide that materially leads users to alter their consumption patterns.
 \par 
@@
 \par 
-There have been a number of alternative recommendation evaluation metrics proposed in the literature with the aim of providing more useful recommendations than those provided by accuracy metrics, such as serendipity \cite{kotkov2016survey}, coverage \cite{ge2010beyond}, novelty \cite{vargas2011rank}, and many others.
+There have been a number of alternative recommendation evaluation metrics proposed in the literature with the aim of providing more useful recommendations than those provided by accuracy metrics, such as serendipity \cite{kotkov2016survey}, calibration \cite{steck2018calibrated}, coverage \cite{ge2010beyond}, novelty \cite{vargas2011rank}, and many others.
 Our approach most closely follows the set of proposed serendipity measures which are surveyed in \cite{kotkov2016survey}.

### 2020-05-17 paper/appendix.tex
finish revised except appendix

@@
 \usepackage[ruled]{algorithm2e} % For algorithms
-
 \usepackage{subcaption}
@@
 \begin{figure}[ht]
-\caption{Relationship between $\beta$ and Homogeneity, $N = 200$}
-\includegraphics[width=.45\linewidth]{figures/beta_homogeneity_N_200_T_20}\label{fig:beta_homo}
-\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the weight of the common value component, $\beta$.
-Each line represents this plot for a single recommendation regime.}
+\caption{Relationship between User Welfare and Diversity, varying $\rho$}\label{fig:diversity_welfare_correlation}
+\begin{subfigure}{.45\linewidth}
+  \includegraphics[width=.8\linewidth]{figures/rho_diversity_N_200_T_20.pdf}
+\end{subfigure}
+\ContinuedFloat
+\begin{subfigure}{.45\linewidth}
+  \includegraphics[width=.8\linewidth]{figures/rho_welfare_N_200_T_20.pdf}
+\end{subfigure}
+\caption*{\scriptsize Notes: The figure on the left displays the relationship between $rho$, the correlation between valuations of items, and overall consumed product diversity.
+The figure on the right displays the relationship between $\rho$ and overall welfare.}
 \end{figure}
+
+\addtocounter{figure}{-1}
+
 \begin{figure}[ht]
@@
 \end{figure}
+\addtocounter{figure}{-1}
 
@@
 
+\addtocounter{figure}{-1}
+
 % Paper body
@@
 \end{figure}
+
+\addtocounter{figure}{-1}
 

### 2020-05-17 paper/filter_bubbles.tex
finish revised except appendix

@@
 \usepackage[ruled]{algorithm2e} % For algorithms
-
+\usepackage{enumitem}
 \usepackage{subcaption}
@@
 \perp V$ -- and that each is multivariate normal
-\begin{enumerate}
+\begin{enumerate}[topsep=0pt]
 \item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ 
@@
 Suppose that there are four items: 0, 1, 2, 3.
-The items are in different places of the product space, where 0 is close to 1 and 3 but more distant from 2, as shown in Figure \ref{fig:illustrative_example}.
+The items are in different places of the product space, where 0 is close to 1 and 3 but more distant from 2.
 The initial beliefs are given as follows:
@@
 Thus, at any given period, the recommender provides a personalized recommendation that combines the knowledge of the common value component $V$ with the user beliefs $\bar V_i$.
-Knowing the user's beliefs about her valuation of each item become crucial in this case: just providing the user with information about $V$ may change the user's original ranking, but, without considering the user's beliefs, she will not necessarily follow the recommendation.\footnote{The notion of recommendation that we consider is idealized where the recommendation does the Bayesian updating for users.
-As users face a very difficult and costly Bayesian updating instance, a useful recommendation system provides recommendations taking into account this aspect and combines the history of past consumption and value realizations with the user's beliefs.
-An alternative that is equivalent would be the recommender simply providing $V$ without knowing $V_i$.
-In this case, the recommendation will be that the user $i$ chooses $r_{i,t} \in \arg \max_{n\in \mathcal J\backslash C^T_i} \mathbb E[u(x_n) \mid C_i^T]$
-but we assume the recommender provides full information about $V$.
-For instance, the recommender could display the whole distribution of utilities reported by other users or even its average, which is a good proxy for the common value component.
-It is possible for the recommender to only recommend the best item given such information, that is, the item with highest common value component.
-However, in that case, recommending only a single item generates costly Bayesian updating to the user and provides little guidance as to what she should indeed pick as the common value might be of little importance when compared to the idiosyncratic component.
-Therefore, we assume that the recommender reports the whole $V$, which results in higher expected welfare in choices and is not costly to implement.
-The results would be equivalent, but this would shift the burden of Bayesian updating to users.}
+Knowing the user's beliefs about her valuation of each item become crucial in this case: just providing the user with information about $V$ may change the user's original ranking, but, without considering the user's beliefs, she will not necessarily follow the recommendation.
 \par
@@
 When we consider results varying a single parameter, we group the results over the other parameters and provide reports varying only the parameter of interest.
-We report results for a relatively small consumption history $T=20$ with a product space size $N=200$.
+We report results for a relatively small consumption history $T=20$ with a product space size $N=200$.\footnote{In the supplementary material we include the same exercises presented here for $N = 100$ and $N = 500$ and show that the results are qualitatively similar to those reported here.}
 \par
@@
 
-First, the right panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho = 0$, there is no difference in consumption distance between the three regimes.
+First, the left panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho = 0$, there is no difference in consumption distance between the three regimes.
 This is due to the fact that when $\rho = 0$, there is no reason that items that are close in the product space should have similar values and so the optimal consumption path does not depend on the similarity of the items.
@@
 \par
-The top panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho > 0$, both recommendation and no recommendation lead to increasingly local consumption compared to the oracle benchmark case.
+The right panel of Figure~\ref{fig:correlation_consumption_path} shows that, when $\rho > 0$, both recommendation and no recommendation lead to increasingly local consumption compared to the oracle benchmark case.
 Further, the average consumption path between periods is \textit{decreasing} for the no recommendation case whereas it is \textit{increasing} for the oracle case.
@@
 However, since these spillovers also impact user learning in the no recommendation case, users \textit{over-exploit} this and increasingly consume products similar to good products that they have consumed before.
-This is further illustrated by Figure \ref{fig:local_consumption_across_rho} which shows how the consumption paths in the oracle and no recommendation regimes vary as $\rho$ increases.
+This is further illustrated by Figure 3 in the supplementary material  which shows how the consumption paths in the oracle and no recommendation regimes vary as $\rho$ increases.
 \par
@@
 Finally, this effect is further amplified as the level of risk aversion, $\gamma$, increases.
-Figure \ref{fig:no_rec_risk_aversion} shows how drastically the degree of local consumption increases as $\gamma$ increases.
+Figure 4 in the supplementary material shows how drastically the degree of local consumption increases as $\gamma$ increases.
 This is due to the fact that the spillovers not only affect the mean expected belief about quality but also the degree of uncertainty.
@@
 As a result, this effect gets amplified as $\rho$ increases, which leads to the gap in diversity between the regimes to increase as $\rho$ increases.
-Figure \ref{fig:diversity_welfare_correlation} shows how diversity varies as $\rho$ increases across the three regimes that we consider.
-There is a similar increasing diversity gap as $\gamma$, or the level of risk-aversion, increases as can be seen in Figure \ref{fig:risk_aversion_diversity}.
+Figure \ref{fig:diversity_welfare_correlation} shows how diversity varies as $\rho$ increases across the three regimes that we consider.There is a similar increasing diversity gap as $\gamma$, or the level of risk-aversion, increases as can be seen in Figure 4 in the supplementary material.
 The mechanisms behind these effects have direct parallels to those discussed in the previous section since low average sequential consumption distance directly leads to low diversity.
@@
 \par 
-Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which are drawn from Figure~\ref{fig:diversity_welfare_correlation}.
+Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which are drawn from Figure 1 in the supplementary material.
 The most interesting observation is that the value of recommendation decreases as $\rho$ decreases.
-The intuition for this is clear as recommendation provides users with information that allows them to better guide their decisions and increase welfare.
-However, as $\rho$ increases users get increasingly more information simply from consuming items since the realized utility is now more informative about the utility of nearby items.
+The intuition is clear as recommendation provides users with information that allows them to better guide their decisions and increase welfare.
+However, as $\rho$ increases users get increasingly more information from consuming items since the realized utility is now more informative about the utility of nearby items.
 Thus, since consumption decisions themselves reveal information, the information that recommendation reveals provides less value.
 \par 
-Figure~\ref{fig:diversity_welfare_common_value} shows how diversity and welfare vary across recommendation regimes as the the strength of the common-value component, $\beta$, varies.
+Figure 1 in the appendix shows how diversity and welfare vary across recommendation regimes as the the strength of the common-value component, $\beta$, varies.
 As expected, when $\beta = 0$, these values are the same.
@@
 Thus, we next explore the relationship between welfare and diversity.
-\begin{figure}
-\caption{Relationship between User Welfare and Diversity, varying $\rho$}\label{fig:diversity_welfare_correlation}
-\begin{subfigure}{.45\linewidth}
-  \includegraphics[width=.8\linewidth]{figures/rho_diversity_N_200_T_20.pdf}
+
+\addtocounter{figure}{-1}
+\begin{figure}[ht]
+\caption{Diversity vs. Welfare, \textbf{No Recommendation}}
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn.pdf}
 \end{subfigure}
-\begin{subfigure}{.45\linewidth}
-  \includegraphics[width=.8\linewidth]{figures/rho_welfare_N_200_T_20.pdf}
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra.pdf}
+\end{subfigure}\\
+\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under the no recommendation regime.
+The figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.}\label{fig:diversity_welfare_ra}
+\end{figure}
+
+\addtocounter{figure}{-1}
+\begin{figure}[ht]
+\caption{Diversity vs. Welfare, \textbf{Recommendation}}
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_partial.pdf}
 \end{subfigure}
-\caption*{\scriptsize Notes: The figure on the left displays the relationship between $rho$, the correlation between valuations of items, and overall consumed product diversity.
-The figure on the right displays the relationship between $\rho$ and overall welfare.}
+\begin{subfigure}{.45\textwidth}
+\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_partial.pdf}
+\end{subfigure}\\
+\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under the recommendation regime.
+The figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.}\label{fig:diversity_welfare_ra_partial}
 \end{figure}
+\addtocounter{figure}{-1}
 
@@
 In the recommendation and oracle regimes, under risk neutrality ($\gamma=0$), welfare and diversity and uncorrelated, while under risk aversion ($\gamma=5$), it is possible to observe in Figure \ref{fig:diversity_welfare_ra_partial} an actual positive relation between diversity and welfare as recommendations are able to reduce uncertainty and facilitate exploration of the product space.
-\begin{figure}[t]
-\caption{Diversity vs. Welfare, \textbf{No Recommendation}}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn.pdf}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra.pdf}
-\end{subfigure}\\
-\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under the no recommendation regime.
-The figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.}\label{fig:diversity_welfare_ra}
+
+\subsection{User Homogenization}
+
+\begin{figure}[ht]
+\caption{Relationship between $\beta$ and Homogeneity, $N = 200$}
+\includegraphics[width=.45\linewidth]{figures/beta_homogeneity_N_200_T_20}\label{fig:beta_homo}
+\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the weight of the common value component, $\beta$.
+Each line represents this plot for a single recommendation regime.}
 \end{figure}
 
-\begin{figure}[t]
-\caption{Diversity vs. Welfare, \textbf{Recommendation}}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_partial.pdf}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_partial.pdf}
-\end{subfigure}\\
-\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under the recommendation regime.
-The figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.}\label{fig:diversity_welfare_ra_partial}
-\end{figure}
-
-\subsection{User Homogenization}
 In this section, we focus on comparisons across users and investigate how the consumed set of items across users varies across different recommendation regimes and parameter values.
@@
 We next study how the degree of homogeneity varies as $\rho$ increases.
-Figure \ref{fig:cor_homo} shows how homogeneity varies as $\rho$ increases across the different recommendation regimes.
+Figure 2 in the supplementary material shows how homogeneity varies as $\rho$ increases across the different recommendation regimes.
 The most striking finding is that homogeneity decreases as $\rho$ increases in the recommendation regime.
@@
 In this section we discuss how the insights from our model of user decision-making can inform the evaluation and design of recommender systems.
-The classic approach to recommender system evaluation is predicting user ratings for items and comparing how accurate this prediction is to recorded ratings data, either explicitly given by users or inferred from behavioral data.
-The recommender system should then recommend to users the items with the highest predicted rating \cite{adomavicius2005toward}.
+The classic approach to recommender system evaluation is to predict user ratings for items and to compare how accurate this prediction is to recorded ratings data, either explicitly given by users or inferred from behavioral data.
+The recommender system should then recommend the items with the highest predicted ratings \cite{adomavicius2005toward}.
 \par
@@
 However, the user herself may also have made this inference since the two movies are very similar to each other.
-Thus, if the recommender system recommends this movie to the user then this recommendation is not useful since the recommendation gives the user little information that they did not already know.
+Thus, recommending this movie would not be not useful since the recommendation gives the user little information that she did not already know.
 The key insight is that the recommendation is not useful since \textit{it ignores the inference the user themselves made and their updated beliefs}.
@@
 There are many sources of information that affect a user's beliefs and choices that are unobservable by recommender systems such as ratings, reviews, and friends.
-Other consumption choices, e.g.
-movies seen in the cinema, change beliefs about for instance how good a director is, which affect beliefs about other movies and guide choices on a streaming platform that is unable to observe these data.
 However, this information can be inferred by the platform by collecting both beliefs and consumption choices.} The second type of data that is critical for recommender system designers to collect is how user beliefs change over time and, in particular, not just how individuals value the item they just consumed, but also how it impacts their beliefs about the quality of similar items.\footnote{Characterizing the similarity between items has been an important goal of designing content-based recommendations, though as noted by \cite{winecoff2019users}, how users perceive similarity between items is not always in line with how similarity is computed in content-based recommender systems.

### 2020-05-18 paper/appendix.tex
more updates to paper, fixed figures and new abstract

@@
 \caption{Relationship between $\beta$ and Homogeneity, $N = 100$}
-\includegraphics[width=.45\linewidth]{figures/beta_homogeneity_N_200_T_20}\label{fig:beta_homo}
+\includegraphics[width=.45\linewidth]{figures/beta_homogeneity_N_100_T_20}\label{fig:beta_homo}
 \caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the weight of the common value component, $\beta$.
@@
 \caption{Relationship between $\beta$ and Homogeneity, $N = 500$}
-\includegraphics[width=.45\linewidth]{figures/beta_homogeneity_N_200_T_20}\label{fig:beta_homo}
+\includegraphics[width=.45\linewidth]{figures/beta_homogeneity_N_500_T_20}\label{fig:beta_homo}
 \caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the weight of the common value component, $\beta$.
@@
 \caption{Relationship between $\rho$ and Homogeneity, $N = 500$}
-\includegraphics[width=.45\linewidth]{figures/rho_homogeneity_N_100_T_20}
+\includegraphics[width=.45\linewidth]{figures/rho_homogeneity_N_500_T_20}
 \caption*{\scriptsize Notes: Notes: This figure displays the value of the homogeneity measure as we vary the inherent correlation between the valuation of the items, $\rho$.
@@
 \begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_100T_20.pdf}
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_500T_20.pdf}
 \end{subfigure}
 \begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_100T_20_partial.pdf}
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_500T_20_partial.pdf}
 \end{subfigure}\\
 \begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_100T_20_omni.pdf}\\
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_500T_20_omni.pdf}\\
 \end{subfigure}%
@@
 \begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_100T_20.pdf}
+\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_500T_20.pdf}
 \end{subfigure}
 \begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_100T_20_partial.pdf}
+\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_500T_20_partial.pdf}
 \end{subfigure}\\
 \begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_100T_20_omni.pdf}\\
+\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_500T_20_omni.pdf}\\
 \end{subfigure}
@@
   \centering
-  \includegraphics[width=.9\linewidth]{figures/rho_pos_consumption_dist_N_100T_20_overall.pdf}
+  \includegraphics[width=.9\linewidth]{figures/rho_pos_consumption_dist_N_500T_20_overall.pdf}
   \label{fig:sfig1}
@@
   \centering
-  \includegraphics[width=.9\linewidth]{figures/rho_zero_consumption_dist_N_100T_20_overall.pdf}
+  \includegraphics[width=.9\linewidth]{figures/rho_zero_consumption_dist_N_500T_20_overall.pdf}
   \label{fig:sfig2}

### 2020-05-18 paper/filter_bubbles.tex
more updates to paper, fixed figures and new abstract

@@
 We study a model of user decision-making in the context of recommender systems via numerical simulation.
-We explore the extent to which two issues traditionally associated with recommender systems -- \textit{filter bubbles} and \textit{user homogenization} -- can be attributed to recommendation systems.
-In our model users face a sequential decision problem under uncertainty and there are informational spillovers where a user consuming a product and learning how she values it reduces uncertainty and provides information about similar products.
-We find that this generates behavior consistent with filter bubble effects -- where users consume increasingly similar items over time -- and that recommendation can help alleviate these effects, as has been found in the empirical work of Nguyen et al.
-(2014).
-However, we find that recommendation, while alleviating filter bubble effects, leads to an increase in user homogeneity.
-Our results highlight the path-dependence of user choice in environments where recommender systems are deployed.
+Our model provides an explanation for the findings of Nguyen, et.
+al (2014), where users consume increasingly similar items over time even without recommendation.
+Our model naturally generates such filter bubble effects since users face a sequential decision making problem under uncertainty and there are informational spillovers where a user consuming a item and learning how she values it reduces uncertainty and provides information about similar items.
+As shown in Nguyen, et.
+al (2014), recommendation alleviates these filter bubble effects and we further find that, by providing users with information, recommendation leads users to consume increasingly similar items.
+Finally, our model highlights the importance of collecting data on user beliefs and their evolution over time both to design better recommendations and to further understand their impact.
 \end{abstract}
@@
 \par
-The right panel of Figure~\ref{fig:correlation_consumption_path} shows that, when $\rho > 0$, both recommendation and no recommendation lead to increasingly local consumption compared to the oracle benchmark case.
+The left panel of Figure~\ref{fig:correlation_consumption_path} shows that, when $\rho > 0$, both recommendation and no recommendation lead to increasingly local consumption compared to the oracle benchmark case.
 Further, the average consumption path between periods is \textit{decreasing} for the no recommendation case whereas it is \textit{increasing} for the oracle case.
@@
 \begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn.pdf}
+\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_n_200.pdf}
 \end{subfigure}
 \begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra.pdf}
+\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_n_200.pdf}
 \end{subfigure}\\
@@
 \begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_partial.pdf}
+\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_partial_n_200.pdf}
 \end{subfigure}
 \begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_partial.pdf}
+\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_partial_n_200.pdf}
 \end{subfigure}\\

### 2020-05-18 paper/filter_bubbles.tex
updates to the paper

@@
 Retrieved on October 3, 2019.} While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems.
-There have been claims that YouTube's recommendation algorithm unintentionally lead to the radicalization of many individuals,\footnote{Friedersdorf (2018, Mar.
-12), YouTube Extremism and the Long Tail, \textit{The Atlantic}.
-\url{https://www.theatlantic.com/politics/archive/2018/03/youtube-extremism-and-the-long-tail/555350}.
-Retrieved on October 3, 2019.} that personalized recommender systems lead users into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
+There have been claims that personalized recommender systems lead users into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
 \par
@@
 In this context, filter bubble effects are defined as users consuming items in an increasingly narrow portion of the product space.
-The simple and intuitive driving force of this in our model is that preferences for similar items are correlated, which implies that when an item is consumed and the user learns its value, it provides information about similar item.
+The simple and intuitive driving force of this in our model is that preferences for similar items are correlated, which implies that when an item is consumed and the user learns its value, it provides information about similar items.
 Crucially, this not only impacts the underlying belief about the average utility of the item, but also the level of uncertainty.
@@
 By collecting information on user beliefs, recommender system designers can understand what items a user would consume \textit{without} recommendation and then predict how providing information to the user would change their beliefs and resulting consumption decisions.
-Thus, our evaluation measure determines the value of a recommendation based on the marginal welfare gain associated with providing a user with a recommendation over what the user would do without recommendation.
+Thus,  our evaluation measure determines the value of a recommendation based on the marginal welfare gain associated with providing a user with a recommendation over what the user would do without it.
 We discuss how this provides an additional rationale as to why ``accurate" recommendations are not always good recommendations.
@@
 \cite{celma2008hits, treviranus2009value} show that incorporating content popularity into recommender systems can lead to increased user homogenization.
-\cite{chaney2018algorithmic} show how user homogenization may arise from training recommender systems on data from users exposed to algorithmic recommendations.
+\cite{chaney2018algorithmic} shows how user homogenization may arise from training recommender systems on data from users exposed to algorithmic recommendations.
 \cite{fleder2009blockbuster} show that homogenization can increase due to a recommendation bias towards popularity that arises from lack of information about items with limited consumption histories.
@@
 \par
-Another set of papers in the recommender systems community that studies the impact of human decision-making on the design and evaluation of recommender systems.
+Another set of papers studies the impact of human decision-making on the design and evaluation of recommender systems.
 \cite{chen2013human} surveys the literature on the relationship between human decision making and recommender systems.
@@
 Our evaluation measure is similar, but, unlike previous approaches, emphasizes the importance of user beliefs.
-Finally, \cite{hodgson2019horse} consider a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of their preferences in the environment, but consider it in the context of search for a single item.
+Finally, \cite{hodgson2019horse} considers a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of their preferences in the environment, but consider it in the context of search for a single item.
 

### 2020-05-20 paper/appendix.tex
more edits

@@
 \maketitle
-
-\section{Omitted Figures}
+\addtocounter{figure}{+4}
+\section{Additional Figures}
 
@@
 \begin{figure}[ht]
-\caption{Relationship between User Welfare and Diversity, varying $\rho$}\label{fig:diversity_welfare_correlation}
+\caption{Relationship between $\rho$ and User Welfare, Diversity, $N = 200$}\label{fig:diversity_welfare_correlation}
 \begin{subfigure}{.45\linewidth}
@@
 \end{subfigure}
-\caption*{\scriptsize Notes: The figure on the left displays the relationship between $rho$, the correlation between valuations of items, and overall consumed product diversity.
+\caption*{\scriptsize Notes: The figure on the left displays the relationship between $\rho$, the correlation between valuations of items, and overall consumed product diversity.
 The figure on the right displays the relationship between $\rho$ and overall welfare.}
@@
 \includegraphics[width=.45\linewidth]{figures/rho_homogeneity_N_200_T_20}
-\caption*{\scriptsize Notes: Notes: This figure displays the value of the homogeneity measure as we vary the inherent correlation between the valuation of the items, $\rho$.
+\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the inherent correlation between the valuation of the items, $\rho$.
 Each line represents this plot for a single recommendation regime.}\label{fig:cor_homo}
@@
 
+\begin{figure}[ht]
+\caption{Relationship between $\gamma$ and Diversity, $N = 200$}
+\includegraphics[width=.45\linewidth]{figures/gamma_diversity_N_200_T_20}
+\caption*{\scriptsize Notes: This figure displays the value of the diversity measure as we vary the inherent correlation between the risk aversion level, $\gamma$.}\label{fig:cor_homo}
+\end{figure}
+
+\begin{figure}[ht]
+\caption{Relationship between $\beta$ and User Welfare, Diversity, $N = 200$}\label{fig:diversity_welfare_correlation}
+\begin{subfigure}{.45\linewidth}
+  \includegraphics[width=.8\linewidth]{figures/beta_diversity_N_200_T_20.pdf}
+\end{subfigure}
+\ContinuedFloat
+\begin{subfigure}{.45\linewidth}
+  \includegraphics[width=.8\linewidth]{figures/beta_welfare_N_200_T_20.pdf}
+\end{subfigure}
+\caption*{\scriptsize Notes: The figure on the left displays the relationship between $\beta$, the strength of the common value component, and overall consumed product diversity.
+The figure on the right displays the relationship between $\beta$ and overall welfare.}
+\end{figure}
+
+\addtocounter{figure}{-1}
+
 \clearpage
@@
 \includegraphics[width=.45\linewidth]{figures/rho_homogeneity_N_100_T_20}
-\caption*{\scriptsize Notes: Notes: This figure displays the value of the homogeneity measure as we vary the inherent correlation between the valuation of the items, $\rho$.
+\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the inherent correlation between the valuation of the items, $\rho$.
 Each line represents this plot for a single recommendation regime.}\label{fig:cor_homo}
@@
 \end{subfigure}
-\caption*{\footnotesize Notes: The figure shows the consecutive consumption path difference between the no recommendation, recommendation, and oracle regime.
-The figure on the left displays the average consecutive consumption distance aggregating over simulations where $\rho = 0$ and the figure on the right displays the average consecutive consumption distance aggregating over simulations where $\rho > 0$}
+\caption*{\scriptsize Notes: The figure shows the consecutive consumption path difference between the no recommendation, recommendation, and oracle regime.
+The figure on the left displays the average consecutive consumption distance aggregating over simulations where $\rho > 0$ and the figure on the right displays the average consecutive consumption distance aggregating over simulations where $\rho = 0$.}
 \label{fig:correlation_consumption_path_n_100}
@@
 
+\begin{figure}[ht]
+\caption{Relationship between $\gamma$ and Diversity, $N = 100$}
+\includegraphics[width=.45\linewidth]{figures/gamma_diversity_N_100_T_20}
+\caption*{\scriptsize Notes: This figure displays the value of the diversity measure as we vary the inherent correlation between the risk aversion level, $\gamma$.}\label{fig:cor_homo}
+\end{figure}
+
 \clearpage
@@
 \includegraphics[width=.45\linewidth]{figures/rho_homogeneity_N_500_T_20}
-\caption*{\scriptsize Notes: Notes: This figure displays the value of the homogeneity measure as we vary the inherent correlation between the valuation of the items, $\rho$.
+\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the inherent correlation between the valuation of the items, $\rho$.
 Each line represents this plot for a single recommendation regime.}\label{fig:cor_homo}
@@
 \caption*{\scriptsize Notes: The figure shows the consecutive consumption path difference between the no recommendation, recommendation, and oracle regime.
-The figure on the left displays the average consecutive consumption distance aggregating over simulations where $\rho = 0$ and the figure on the right displays the average consecutive consumption distance aggregating over simulations where $\rho > 0$}
+The figure on the left displays the average consecutive consumption distance aggregating over simulations where $\rho > 0$ and the figure on the right displays the average consecutive consumption distance aggregating over simulations where $\rho = 0$.}
 \label{fig:correlation_consumption_path_n_100}
+\end{figure}
+
+\begin{figure}[ht]
+\caption{Relationship between $\gamma$ and Diversity, $N = 500$}
+\includegraphics[width=.45\linewidth]{figures/gamma_diversity_N_500_T_20}
+\caption*{\scriptsize Notes: This figure displays the value of the diversity measure as we vary the inherent correlation between the risk aversion level, $\gamma$.}\label{fig:cor_homo}
 \end{figure}

### 2020-05-20 paper/filter_bubbles.tex
more edits

@@
 
-Recommender Systems (RS) have become critical for assisting users in navigating the large choice sets that they face in many online platforms.
+Recommender systems (RS) have become critical for assisting users in navigating the large choice sets that they face on many online platforms.
 For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
@@
 
-Recommender systems have been influential in shaping consumer choice in these markets with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{MacKenzie et al.
+Recommender systems have driven a significant fraction of consumer choice on these platforms with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{MacKenzie et al.
 (2013, Oct.),  How retailers can keep up with consumers.
@@
 In this paper we develop a theoretical model of user decision-making in contexts where recommender systems are traditionally deployed.
-We utilize previous empirical studies that characterize how recommender systems influence user choice as a benchmark and our theoretical model provides an intuitive mechanism that provides an explanation for these empirical results.
-The key insight of our model is that individual \textit{beliefs} drive the consumption choices of individuals and that recommendations serve to provide users with information that leads them to update their beliefs.
+We utilize previous empirical studies that characterize how recommender systems influence user choice as a benchmark and our theoretical model provides an intuitive mechanism that can explain these empirical results.
+The key insight of our model is that individual \textit{beliefs} drive the consumption choices of individuals and that recommendations provide users with information that leads them to update their beliefs and alter their choices.
 However, a crucial component of our model is that individuals' beliefs about items are driven not only by recommendations, but also from their previous experiences with similar items.
-We use these insights to provide guidance for recommender system designers.
-In order to design good recommendations it becomes crucial for designers to understand user beliefs of goods they haven't consumed thus far and how these beliefs evolve over time.
+We use these insights to provide guidance for recommender system design.
+In order to design good recommendations it becomes crucial for designers to understand user beliefs of the goods they haven't consumed thus far and track how these beliefs evolve over time.
 
@@
 The first component of our model is that users sequentially consume items and face large choice sets.
-In our setting of interest users are long-lived, but they only consume a small fraction of this choice set over their lifetime.
+In our setting of interest, users are long-lived, but they only consume a small fraction of this choice set over their lifetime.
 This is traditionally the case on online platforms that have thousands or millions of options for users.
 \par
-The second component of our model is that users have \textit{uncertainty} about their valuation of the different items before they consume them.
+The second component is that users have \textit{uncertainty} about their valuation of the different items before they consume them.
 This is motivated both by the fact that many contexts where recommender systems are deployed contain experience goods, whose true value can only be learned after consumption, and the fact that such uncertainty is why recommender systems exist in the first place.
@@
 \par 
-The third, and key, component of our model is that consumption of an item reveals information that changes user beliefs about the valuation of similar items.
+The third, and most crucial, element is that consumption of an item reveals information that changes user beliefs about the valuation of similar items.
 Unlike in standard sequential decision-making problems, in our model once an item is consumed all uncertainty about its valuation is resolved.
@@
 \par 
-The fourth component of our model is that recommendation provides users with information about the true valuations.
+Finally, in our model recommendation provides users with information about the true valuations.
 We model the realized valuations as being a weighted sum of a common-value and an idiosyncratic component.
@@
 Our model provides a clear mechanism that explains the empirical results in \cite{nguyen2014exploring} who show that, in the context of movie consumption, user behavior is consistent with filter bubble effects even without recommendation and that recommendation leads to users being less likely to fall into such filter bubbles.
-In this context, filter bubble effects are defined as users consuming items in an increasingly narrow portion of the product space.
+In this context, filter bubble effects are defined as users consuming items in an increasingly narrow portion of the product space over time.
 The simple and intuitive driving force of this in our model is that preferences for similar items are correlated, which implies that when an item is consumed and the user learns its value, it provides information about similar items.
 Crucially, this not only impacts the underlying belief about the average utility of the item, but also the level of uncertainty.
-Consequently, this learning spillover induces users to consume items ``similar'' to those they consumed before that had high realized value, leading to an increasing narrowing of consumption patterns towards these regions of the product space.
+Consequently, this learning spillover induces users to consume items similar to those they consumed before that had high realized value, leading to an increasing narrowing of consumption patterns towards these regions of the product space.
 This effect is further amplified when users are \textit{risk-averse}, a concept from decision theory where all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty.
@@
 This leads to an increase in homogeneity across users, resulting in a tradeoff between homogenizing across user consumption and diversifying within user consumption.
-We finally explore the relationship between the overall diversity of consumed items and user welfare.
-We find that more diverse sets of consumed items do not always correspond to higher user welfare.
+We explore the relationship between the overall diversity of consumed items and user welfare and find that more diverse sets of consumed items do not always correspond to higher user welfare.
 \par
-Finally, we discuss how our model and findings can be used to inform the design and evaluation of recommender systems as well as the data that is traditionally collected for them.
-Our model highlights the importance of user beliefs in determining the consumption choices of individuals and how both recommendation and informational spillovers determine how these beliefs change over time.
-By collecting information on user beliefs, recommender system designers can understand what items a user would consume \textit{without} recommendation and then predict how providing information to the user would change their beliefs and resulting consumption decisions.
+Lastly, we discuss how our model and findings can be used to inform the design and evaluation of recommender systems as well as the data that is traditionally collected for them.
+This highlights the importance of user beliefs in determining the consumption choices of individuals and how both recommendation and informational spillovers determine how these beliefs change over time.
+By collecting information on user beliefs, recommender system designers can understand what items a user would consume \textit{without} recommendation and then predict how providing information to the user would change her beliefs and resulting consumption decisions.
 Thus,  our evaluation measure determines the value of a recommendation based on the marginal welfare gain associated with providing a user with a recommendation over what the user would do without it.
@@
 The most relevant to our study is \cite{nguyen2014exploring} who study whether this effect exists with recommender systems in the context of movie consumption.
-They find that even users whose consumption choices are not guided by recommendations fall into behavior consistent with ``filter bubbles'' and that recommender systems can actually increase the diversity of the content that users consume.
-To our knowledge there are no theoretical models that rationalize these empirical findings and our paper provides a theoretical framework through which to view this problem.
-Further, our paper provides a clear mechanism at play that drives such effects and how recommendation interacts with them.
+They find that even users whose consumption choices are not guided by recommendations exhibit behavior consistent with ``filter bubbles'' and that recommender systems can actually increase the diversity of the content that users consume.
+To our knowledge there are no theoretical models that rationalize these empirical findings and we provide a theoretical framework through which to view this problem.
+Furthermore, our paper provides a clear mechanism at play that drives such effects and how recommendation interacts with them.
 \par 
@@
 \cite{fleder2009blockbuster} show that homogenization can increase due to a recommendation bias towards popularity that arises from lack of information about items with limited consumption histories.
-Our model leads to similar results as previous theoretical and empirical work where recommender systems lead to increased user homogenization.
-However, the mechanisms behind this from our model differ from existing work as homogenization arises due to the fact that recommendation leads users to coordinate their consumption decisions in certain portions of the product space.
+Our model leads to similar results as previous work where recommender systems lead to increased user homogenization.
+However, the mechanisms behind this differ from existing work as homogenization arises due to the fact that recommendation leads users to coordinate their consumption decisions in certain portions of the product space.
 \par
@@
 The closest set of papers pointed out in this survey are those related to preference construction \cite{bettman1998constructive, lichtenstein2006construction} whereby users develop preferences over time through the context of a decision process.
-Our model points out that the true underlying preferences of users may be stable over time, but, due to the nature of items in contexts where recommender systems are deployed, they have incomplete information of them and both consumption and recommendation provide them with information to reduce their uncertainty.
+Our model points out that the true underlying preferences of users may be stable over time, but, due to the nature of items in contexts where recommender systems are deployed, they have incomplete information of their valuations and both consumption and recommendation provide them with information to reduce their uncertainty.
 Thus, the primary insight of our paper is that individual beliefs and how individuals update their beliefs about similar items after consumption are important and previously unconsidered elements of human decision making that are critical for understanding the design and consequences of recommender systems.
-Within this literature, there is a set of papers that focus on ``user-centric" approaches to recommendation \cite{celma2008new, cremonesi2013user, pu2011user} whereby user evaluation of the usefulness of recommendation is a key evaluation measure.
+Within this literature, \cite{celma2008new, cremonesi2013user, pu2011user} focus on ``user-centric" approaches to recommendation whereby user evaluation of the usefulness of recommendation is a key evaluation measure.
 Our evaluation measure is similar, but, unlike previous approaches, emphasizes the importance of user beliefs.
@@
 Note that this implies that consuming item $n$ is the same as taking a gamble.
-Each user evaluates the item according to the associated expected utility associated with the item, i.e.
+Each user evaluates the item according to the expected utility associated with the item, i.e.
 $U_i(n)=\mathbb E_{p_i}[u(x_n)]$.
 \par
-
 Risk aversion captures how different users react to the risk associated to a particular consumption opportunity.
@@
 One can interpret $v_n$ as a measure of how much item $n$ is valued in society in general and, in a sense, $v_{i,n}$ denotes how $i$ diverges from this overall ranking.
-The scalar $\beta \in \mathbb{R}_{t}$ denotes the degree to which valuations are idiosyncratic to each user or common across users.
+The scalar $\beta \in \mathbb{R}_{+}$ denotes the degree to which valuations are idiosyncratic to each user or common across users.
 If $\beta=0$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta$ is large, every individual has similar preferences.
@@
 This captures the idea that users are faced with an immense choice set but that ultimately they end up experiencing (and learning) about just a small fraction of it.
-Since this is a sequential decision-making problem under uncertainty, in principle users face an exploration-exploitation trade-off.
-However, for tractability, we impose that users are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
+For tractability, we impose that users are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
 \par
@@
 \xhdr{User Learning}
-When a user consumes an item $n$ she learns the realized value for that item. In our model, we incorporate the idea that learning the value of item $n$ gives a user information about similar items. 
-%This is drawn from recent empirical evidence in \cite{schulz2019structured} that consumers learn how to navigate large choice sets using similarity-based generalization. 
-We assume that learning about the value of item $n$ reveals more about the value associated to items that are closer to it, which captures the idea that trying an item provides more information about similar items than about dissimilar ones.
+When a user consumes an item $n$ she learns the realized value for that item.
+We assume that learning about the value of item $n$ reveals more about the value associated to items that are closer to it, which captures the idea that trying an item provides more information about similar items than about dissimilar ones.\footnote{\cite{schulz2019structured} empirically studies how individuals solve sequential decision-making problems under uncertainty in large choice sets in the context of mobile food delivery orders.
+They find that individuals engage in similarity-based generalizations where learning about the realized value of a particular good provides them with information about similar items.
+We incorporate this finding into our model in a stylized manner.} In order to have a well-defined notion of similarity we need to define a distance function between items, which we define as $d(n,m):=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}$ where $m$ and $n$ are indices of items in $\mathcal{J}$.
+This distance function is not intended to model the intricacies of a realistic product space, but instead to provide a stylized product space to help us understand the effects of informational spillovers on user behavior.
+The basic intuition, in the context of movie consumption, is that a user watching \textit{John Wick} (item $n$) provides more information about how much she will like \textit{John Wick: Chapter Two} (item $m$) than \textit{Titanic} (item $q$) since $d(n, m) < d(n, q)$.
 \par
-In order to have a well-defined notion of similarity we need to define a distance function between items, which we define as $d(n,m):=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}$ where $m$ and $n$ are indices of items in $\mathcal{J}$.
-This distance function is not intended to model the intricacies of a realistic product space, but instead to provide a stylized product space to help us understand the effects of informational spillovers on user behavior.
 We consider that the entry of $n$-th row and the ($m$)-th column of $\Sigma_i$ is given by $\sigma_i^2 \rho^{d(n,m)}$, and that of $\Sigma$ is given by $\sigma^2 \rho^{d(n,m)}$.
@@
 If the user is risk-neutral ($\gamma = 0$), then she will go across the product space to consume item $2$ in the next period since it has a higher expected value.
-However, if she is sufficiently risk-averse then she still may consume item $1$ or $3$ since her uncertainty about these items is lower than item $2$.
+However, if she is sufficiently risk-averse then she may still consume item $1$ or $3$ since her uncertainty about these items is lower than item $2$.
 In particular, this will happen when
@@
 \par 
-This example illustrates the main mechanisms in our model that can lead to excessive consumption of similar items.
+This example illustrates the main mechanisms that can lead to excessive consumption of similar items.
 Once the user finds items in the product space with high valuations she will update her beliefs positively about items in this portion of the product space and continue consuming these items regardless of her level of risk aversion.
@@
 Thus, at any given period, the recommender provides a personalized recommendation that combines the knowledge of the common value component $V$ with the user beliefs $\bar V_i$.
-Knowing the user's beliefs about her valuation of each item become crucial in this case: just providing the user with information about $V$ may change the user's original ranking, but, without considering the user's beliefs, she will not necessarily follow the recommendation.
+Knowing the user's beliefs about her valuation of each item become crucial in this case: just providing the user with information about $V$ may change the user's original ranking, but, without considering the user's beliefs, she will not necessarily follow the recommendation.\footnote{The notion of recommendation that we consider is idealized where the recommendation does the Bayesian updating for users, but the results are equivalent to if the users did the updating themselves.}
 \par
@@
 We further consider two cases that serve primarily as benchmarks.
-The first is \textit{no recommendation}, where users get no additional information about values and make consumption choices based on their beliefs and consumption history.
+The first is \textit{no recommendation}, where users get no additional information and make consumption choices based on their beliefs and consumption history.
 This gives us a benchmark as to how users in our model would behave \textit{without} recommendation so that we can analyze what changes with the introduction of recommendation.
@@
 \caption*{\scriptsize Notes: The figure shows the consecutive consumption path difference between the no recommendation, recommendation, and oracle regime.
-The figure on the left displays the average consecutive consumption distance aggregating over simulations where $\rho = 0$ and the figure on the right displays the average consecutive consumption distance aggregating over simulations where $\rho > 0$}
+The figure on the left displays the average consecutive consumption distance aggregating over simulations where $\rho > 0$ and the figure on the right displays the average consecutive consumption distance aggregating over simulations where $\rho = 0$.}
 \label{fig:correlation_consumption_path}
@@
 
-First, the left panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho = 0$, there is no difference in consumption distance between the three regimes.
+First, the right panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho = 0$, there is no difference in consumption distance between the three regimes.
 This is due to the fact that when $\rho = 0$, there is no reason that items that are close in the product space should have similar values and so the optimal consumption path does not depend on the similarity of the items.
@@
 Not only is there no difference in the levels between the three regimes, but they all have the same, flat, average consumption distance path.
-This underscores the fact that such concerns about filter bubbles \textit{only} arise in a world where there is correlation between the realized utility of items.
-If there were no correlation between the realized utilities then there would be no reason for users to consume similar items and thus no narrowing effect, regardless of the information on the true utility of the items that users had.
+This underscores the fact that if there were no correlation between the realized utilities then there would be no reason for users to consume similar items and thus no narrowing effect, regardless of the information on the true utility of the items that users had.
 \par
@@
 However, since these spillovers also impact user learning in the no recommendation case, users \textit{over-exploit} this and increasingly consume products similar to good products that they have consumed before.
-This is further illustrated by Figure 3 in the supplementary material  which shows how the consumption paths in the oracle and no recommendation regimes vary as $\rho$ increases.
+This is further illustrated in the supplementary material (Figure 7), which shows how the consumption paths in the oracle and no recommendation regimes vary as $\rho$ increases.
 \par
@@
 Finally, this effect is further amplified as the level of risk aversion, $\gamma$, increases.
-Figure 4 in the supplementary material shows how drastically the degree of local consumption increases as $\gamma$ increases.
+We show in the supplementary material (Figure 8) that the degree of local consumption drastically increases as $\gamma$ increases.
 This is due to the fact that the spillovers not only affect the mean expected belief about quality but also the degree of uncertainty.
@@
 In this section we primarily focus on the impact of recommendation on user welfare and the overall diversity of the items that they consume.
-\par 
 While in the previous section we looked at the distance between consecutive items, in this section we focus on a diversity measure that considers the entire consumed set of items.
@@
 As a result, this effect gets amplified as $\rho$ increases, which leads to the gap in diversity between the regimes to increase as $\rho$ increases.
-Figure \ref{fig:diversity_welfare_correlation} shows how diversity varies as $\rho$ increases across the three regimes that we consider.There is a similar increasing diversity gap as $\gamma$, or the level of risk-aversion, increases as can be seen in Figure 4 in the supplementary material.
+In the supplementary material (Figure 5), we show how diversity varies as $\rho$ increases across the three regimes that we consider.
+There is a similar increasing diversity gap as $\gamma$, or the level of risk-aversion, increases as can be seen in the supplementary material (Figure 9).
 The mechanisms behind these effects have direct parallels to those discussed in the previous section since low average sequential consumption distance directly leads to low diversity.
@@
 \par 
-Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which are drawn from Figure 1 in the supplementary material.
+Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which are drawn from the supplementary material (Figure 5).
 The most interesting observation is that the value of recommendation decreases as $\rho$ decreases.
@@
 \par 
-Figure 1 in the appendix shows how diversity and welfare vary across recommendation regimes as the the strength of the common-value component, $\beta$, varies.
+In the supplementary material (Figure 10), we show how diversity and welfare vary across recommendation regimes as the the strength of the common-value component, $\beta$, varies.
 As expected, when $\beta = 0$, these values are the same.
 However, as $\beta$ grows the idiosyncratic component of user valuations becomes increasingly less important and the recommendation regime begins to converge to the oracle regime.
-Indeed, we see that both product diversity and welfare increase under recommendation, whereas under no recommendation we see that there is a decrease in diversity with little changes to consumer welfare.
+Indeed, we see that both product diversity and welfare increase under recommendation, whereas under no recommendation we see that there is a decrease in diversity with little changes to user welfare.
 This is intuitive as $\beta$ controls the informativeness of recommendation about user preferences.
@@
 Indeed, it appears that the opposite is the case - that low diversity is associated with higher welfare and vice versa.
-Thus, we next explore the relationship between welfare and diversity.
+We next explore the relationship between welfare and diversity.
 
@@
 \begin{figure}[ht]
-\caption{Diversity vs. Welfare, \textbf{No Recommendation}}
+\caption{Diversity vs. Welfare, \textbf{No Recommendation}, N = 200}
 \begin{subfigure}{.45\textwidth}
@@
 \begin{figure}[ht]
-\caption{Diversity vs. Welfare, \textbf{Recommendation}}
+\caption{Diversity vs. Welfare, \textbf{Recommendation}, N  = 200}
 \begin{subfigure}{.45\textwidth}
@@
 When there is no risk-aversion then there is a negative correlation between welfare and diversity.
-This is since, with no risk-aversion, in a given period a user will select the good that she currently believes has the highest expected value.
+This is since, with no risk-aversion, in a given period a user will select the item that she currently believes has the highest expected value.
 High product diversity in this case can arise from a user who consumes an item she disliked and updates her beliefs about nearby items negatively.
@@
 \par
-This only happens since $\gamma = 0$ leads to users only caring about the expected value of the product.
+This only happens since $\gamma = 0$ leads to users only caring about the expected value of the item.
 However, as we saw in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, increasing $\gamma$ can lead to lower diversity and increasingly local consumption due to the fact that the degree of uncertainty now impacts users' choices.
 This weakens the negative relationship between diversity and welfare since both negative and positive experiences with a good reduce uncertainty about surrounding items.
-This leads to the inverted-U shape found in Figure \ref{fig:diversity_welfare_ra} when $\gamma$ is relatively large (e.g $\gamma = 5$) though diversity and welfare are virtually uncorrelated in the data.
+This leads to the inverted-U shape found in Figure \ref{fig:diversity_welfare_ra} when $\gamma$ is relatively large (e.g.
+$\gamma = 5$) though diversity and welfare are virtually uncorrelated in the data.
 In the recommendation and oracle regimes, under risk neutrality ($\gamma=0$), welfare and diversity and uncorrelated, while under risk aversion ($\gamma=5$), it is possible to observe in Figure \ref{fig:diversity_welfare_ra_partial} an actual positive relation between diversity and welfare as recommendations are able to reduce uncertainty and facilitate exploration of the product space.
@@
 We next study how the degree of homogeneity varies as $\rho$ increases.
-Figure 2 in the supplementary material shows how homogeneity varies as $\rho$ increases across the different recommendation regimes.
+In the supplementary material (Figure 6) we show how homogeneity varies as $\rho$ increases.
 The most striking finding is that homogeneity decreases as $\rho$ increases in the recommendation regime.

### 2020-05-20 paper/appendix.tex
updates
more updates

@@
 \includegraphics[width=.45\linewidth]{figures/gamma_diversity_N_200_T_20}
-\caption*{\scriptsize Notes: This figure displays the value of the diversity measure as we vary the inherent correlation between the risk aversion level, $\gamma$.}\label{fig:cor_homo}
+\caption*{\scriptsize Notes: This figure displays the value of the diversity measure as we vary the risk aversion level, $\gamma$.}\label{fig:cor_homo}
 \end{figure}
@@
 \end{subfigure}
-\caption*{\scriptsize Notes: Each figure plots the average consecutive consumption distance across time as the inherent correlation between the valuation of the items, $\rho$.
+\caption*{\scriptsize Notes: Each figure plots the average consecutive consumption distance across time as the inherent correlation between the valuation of the items, $\rho$, varies.
 The top left displays the no recommendation regime, the top right displays the recommendation regime, and the bottom displays the oracle regime.}
@@
 
+\begin{figure}[ht]
+\caption{Relationship between $\beta$ and User Welfare, Diversity, $N = 100$}\label{fig:diversity_welfare_correlation}
+\begin{subfigure}{.45\linewidth}
+  \includegraphics[width=.8\linewidth]{figures/beta_diversity_N_100_T_20.pdf}
+\end{subfigure}
+\ContinuedFloat
+\begin{subfigure}{.45\linewidth}
+  \includegraphics[width=.8\linewidth]{figures/beta_welfare_N_100_T_20.pdf}
+\end{subfigure}
+\caption*{\scriptsize Notes: The figure on the left displays the relationship between $\beta$, the strength of the common value component, and overall consumed product diversity.
+The figure on the right displays the relationship between $\beta$ and overall welfare.}
+\end{figure}
+
+\addtocounter{figure}{-1}
 \clearpage
@@
 
+\begin{figure}[ht]
+\caption{Relationship between $\beta$ and User Welfare, Diversity, $N = 500$}\label{fig:diversity_welfare_correlation}
+\begin{subfigure}{.45\linewidth}
+  \includegraphics[width=.8\linewidth]{figures/beta_diversity_N_500_T_20.pdf}
+\end{subfigure}
+\ContinuedFloat
+\begin{subfigure}{.45\linewidth}
+  \includegraphics[width=.8\linewidth]{figures/beta_welfare_N_500_T_20.pdf}
+\end{subfigure}
+\caption*{\scriptsize Notes: The figure on the left displays the relationship between $\beta$, the strength of the common value component, and overall consumed product diversity.
+The figure on the right displays the relationship between $\beta$ and overall welfare.}
+\end{figure}
+
+\addtocounter{figure}{-1}
+
 % Bibliography

### 2020-05-20 paper/filter_bubbles.tex
more updates

@@
 \par
-
-\textbf{Related Work.} 
+\xhdr{Related Work.} 
 The first set of related works studies the extent and implications of filter bubbles.

### 2020-05-20 paper/filter_bubbles.tex
more edits

@@
 We utilize previous empirical studies that characterize how recommender systems influence user choice as a benchmark and our theoretical model provides an intuitive mechanism that can explain these empirical results.
-The key insight of our model is that individual \textit{beliefs} drive the consumption choices of individuals and that recommendations provide users with information that leads them to update their beliefs and alter their choices.
-However, a crucial component of our model is that individuals' beliefs about items are driven not only by recommendations, but also from their previous experiences with similar items.
+The key insight of our model is that user \textit{beliefs} drive the consumption choices of users and that recommendations provide them with information that leads them to update their beliefs and alter their choices.
+A crucial component of our model is that users' beliefs about items are driven not only by recommendations, but also from their previous experiences with similar items.
 We use these insights to provide guidance for recommender system design.
-In order to design good recommendations it becomes crucial for designers to understand user beliefs of the goods they haven't consumed thus far and track how these beliefs evolve over time.
+In order to design good recommendations it becomes important for designers to understand user beliefs of the goods they haven't consumed thus far and track how these beliefs evolve over time.
 
@@
 The third, and most crucial, element is that consumption of an item reveals information that changes user beliefs about the valuation of similar items.
-Unlike in standard sequential decision-making problems, in our model once an item is consumed all uncertainty about its valuation is resolved.
-This valuation provides information for consumers that enables them to update their beliefs about similar items.
+Unlike in standard sequential decision-making problems, once an item is consumed all uncertainty about its valuation is resolved and provides information that enables users to update their beliefs about similar items.
 This exploits the fact that the valuations of similar items are correlated which assists users in navigating the vast product space.
@@
 We model the realized valuations as being a weighted sum of a common-value and an idiosyncratic component.
-This formulation gives a stylized notion of predictability of user preferences where the idiosyncratic component is inherently unpredictable given other individuals' preferences and the common-value component is what the recommender can learn from previous users' data.
-We suppose that the recommender knows the common-value component for each item and combines it with individuals' beliefs over the product space when designing personalized recommendation.
+This formulation gives a stylized notion of predictability of user preferences where the idiosyncratic component is inherently unpredictable given other users' preferences and the common-value component is what the recommender can learn from previous users' data.
+We suppose that the recommender knows the common-value component for each item and combines it with users' beliefs over the product space when designing personalized recommendation.
 
@@
 Crucially, this not only impacts the underlying belief about the average utility of the item, but also the level of uncertainty.
-Consequently, this learning spillover induces users to consume items similar to those they consumed before that had high realized value, leading to an increasing narrowing of consumption patterns towards these regions of the product space.
+Consequently, this learning spillover induces users to consume items similar to those they consumed before that had high realized value, leading to an increasing narrowing of consumption towards these regions of the product space.
 This effect is further amplified when users are \textit{risk-averse}, a concept from decision theory where all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty.
@@
 Lastly, we discuss how our model and findings can be used to inform the design and evaluation of recommender systems as well as the data that is traditionally collected for them.
-This highlights the importance of user beliefs in determining the consumption choices of individuals and how both recommendation and informational spillovers determine how these beliefs change over time.
+This highlights the importance of user beliefs in determining user consumption choices and how both recommendation and informational spillovers determine how these beliefs change over time.
 By collecting information on user beliefs, recommender system designers can understand what items a user would consume \textit{without} recommendation and then predict how providing information to the user would change her beliefs and resulting consumption decisions.
@@
 Following this a number of empirical studies, done in various disciplines, have since studied the extent to which this phenomenon exists in a wide range of contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}.
-The most relevant to our study is \cite{nguyen2014exploring} who study whether this effect exists with recommender systems in the context of movie consumption.
+The most relevant to our study is \cite{nguyen2014exploring} who study whether this effect exists in the context of movie consumption.
 They find that even users whose consumption choices are not guided by recommendations exhibit behavior consistent with ``filter bubbles'' and that recommender systems can actually increase the diversity of the content that users consume.
@@
 The closest set of papers pointed out in this survey are those related to preference construction \cite{bettman1998constructive, lichtenstein2006construction} whereby users develop preferences over time through the context of a decision process.
-Our model points out that the true underlying preferences of users may be stable over time, but, due to the nature of items in contexts where recommender systems are deployed, they have incomplete information of their valuations and both consumption and recommendation provide them with information to reduce their uncertainty.
-Thus, the primary insight of our paper is that individual beliefs and how individuals update their beliefs about similar items after consumption are important and previously unconsidered elements of human decision making that are critical for understanding the design and consequences of recommender systems.
+We point out that the true underlying preferences of users may be stable over time, but, due to the nature of items in contexts where recommender systems are deployed, they have incomplete information of their valuations and both consumption and recommendation provide them with information to reduce their uncertainty.
+Thus, the primary insight of our paper is that user beliefs and how users update their beliefs about similar items after consumption are important and previously unconsidered elements of human decision making that are critical for understanding the design and consequences of recommender systems.
 Within this literature, \cite{celma2008new, cremonesi2013user, pu2011user} focus on ``user-centric" approaches to recommendation whereby user evaluation of the usefulness of recommendation is a key evaluation measure.
@@
 We assume the user makes $T$ choices and therefore can only consume up to $T$ items, where $T$ is a small fraction of $N$.
-This captures the idea that users are faced with an immense choice set but that ultimately they end up experiencing (and learning) about just a small fraction of it.
-For tractability, we impose that users are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
+This captures the idea that users are faced with an immense choice set, but that ultimately they end up experiencing (and learning) about just a small fraction of it.
+For tractability, we impose that users are myopic and every period consume the item that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
 \par
-\xhdr{User Beliefs.} We assume that all the true realized values are drawn at $t = 0$.
-However, users do not know the realized values before consuming an item, but rather have beliefs over the realized utilities.
+\xhdr{User Beliefs.} We assume that all the true realized values are drawn at $t = 0$. However, users do not know the realized values before consuming an item, but rather have beliefs over them.
 Formally, user $i$ starts with some beliefs about $X_i$, namely that the idiosyncratic and common-value parts of the valuations are independent -- $V_i \perp \!\!\!
@@
 The recommendation regime decreases the degree of local consumption, but not as much as the oracle benchmark.
-Due to the correlation of values, the oracle consumption path exploits this and leads to the consumption of more similar products than in the case when $\rho = 0$.
-However, since these spillovers also impact user learning in the no recommendation case, users \textit{over-exploit} this and increasingly consume products similar to good products that they have consumed before.
-This is further illustrated in the supplementary material (Figure 7), which shows how the consumption paths in the oracle and no recommendation regimes vary as $\rho$ increases.
+Due to the correlation of values, the oracle consumption path exploits this and leads to the consumption of more similar items than in the case when $\rho = 0$.
+However, since these spillovers also impact user learning in the no recommendation case, users \textit{over-exploit} this and increasingly consume items similar to high value items that they have consumed before.
+This is further illustrated in the supplementary material (Figure 7), which shows how the consumption paths in the oracle and no recommendation regimes vary as $\rho$ increases and is in line with this intuition.
 \par
-
-Finally, this effect is further amplified as the level of risk aversion, $\gamma$, increases.
+This effect is further amplified as the level of risk aversion, $\gamma$, increases.
 We show in the supplementary material (Figure 8) that the degree of local consumption drastically increases as $\gamma$ increases.
@@
 We have shown that, unless users are provided with additional information to guide their consumption choices, then these information spillovers and user risk-aversion can lead users into filter bubbles.
-When users consume high valued items, they exploit the underlying correlation across different products' values, stronger for similar products, which leads them to increasingly consume products in narrower and narrower portions of the product space.
+When users consume high valued items, they exploit the underlying correlation across different items' values, stronger for similar items, which leads them to increasingly consume products in narrower and narrower portions of the product space.
 Risk aversion may lead users into performing local consumption even when they have a low valuation of nearby items just because they know what to expect from the item.
@@
 The diversity measure we utilize is common in recommender system literature (e.g.
-\cite{ziegler2005improving}) which is the average normalized pairwise distance between the consumed products:
+\cite{ziegler2005improving}) which is the average normalized pairwise distance between the consumed items:
 $$D_i:=\frac{1}{N}\frac{1}{T(T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$$

### 2020-05-22 paper/appendix.tex
shorthand

--- 2020-05-20 paper/appendix.tex
@@
 \caption{Relationship between $\rho$ and User Welfare, Diversity, $N = 200$}\label{fig:diversity_welfare_correlation}
-\begin{subfigure}{.45\linewidth}
-  \includegraphics[width=.8\linewidth]{figures/rho_diversity_N_200_T_20.pdf}
+\begin{subfigure}{0.3\linewidth}
+  \includegraphics[width=1.0\linewidth]{figures/rho_diversity_N_200_T_20.pdf}
 \end{subfigure}
 \ContinuedFloat
-\begin{subfigure}{.45\linewidth}
-  \includegraphics[width=.8\linewidth]{figures/rho_welfare_N_200_T_20.pdf}
+\begin{subfigure}{0.3\linewidth}
+  \includegraphics[width=1.0\linewidth]{figures/rho_welfare_N_200_T_20.pdf}
 \end{subfigure}
@@
 
-\begin{figure}[ht]
-\caption{Relationship between $\rho$ and Homogeneity, $N = 200$}
-\includegraphics[width=.45\linewidth]{figures/rho_homogeneity_N_200_T_20}
-\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the inherent correlation between the valuation of the items, $\rho$.
-Each line represents this plot for a single recommendation regime.}\label{fig:cor_homo}
-\end{figure}
-
 \begin{figure}[H]
 \caption{Relationship between Local Consumption and $\rho$, $N = 200$}
-\begin{subfigure}{.45\textwidth}
+\begin{subfigure}{.3\textwidth}
 \includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20.pdf}
 \end{subfigure}
-\begin{subfigure}{.45\textwidth}
+\begin{subfigure}{.3\textwidth}
 \includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_partial.pdf}
 \end{subfigure}\\
-\begin{subfigure}{.45\textwidth}
+\begin{subfigure}{.3\textwidth}
 \includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_omni.pdf}\\
@@
 \caption{Relationship between Local Consumption and $\gamma$, $N = 200$}
-\begin{subfigure}{.45\textwidth}
+\begin{subfigure}{.3\textwidth}
 \includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20.pdf}
 \end{subfigure}
-\begin{subfigure}{.45\textwidth}
+\begin{subfigure}{.3\textwidth}
 \includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_partial.pdf}
-\end{subfigure}\\
-\begin{subfigure}{.45\textwidth}
+\end{subfigure}
+\begin{subfigure}{.3\textwidth}
 \includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_omni.pdf}\\
@@
 
-\clearpage
-\section{N = 100}
-This section contains the same figures as reported in the main text and appendix, but for $N = 100$ instead of $N = 200$.
-
-\begin{figure}[ht]
-\caption{Relationship between $\beta$ and Homogeneity, $N = 100$}
-\includegraphics[width=.45\linewidth]{figures/beta_homogeneity_N_100_T_20}\label{fig:beta_homo}
-\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the weight of the common value component, $\beta$.
-Each line represents this plot for a single recommendation regime.}
-\end{figure}
-\begin{figure}[ht]
-\caption{Relationship between $\rho$ and Homogeneity, $N = 100$}
-\includegraphics[width=.45\linewidth]{figures/rho_homogeneity_N_100_T_20}
-\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the inherent correlation between the valuation of the items, $\rho$.
-Each line represents this plot for a single recommendation regime.}\label{fig:cor_homo}
-\end{figure}
-
-\begin{figure}[H]
-\caption{Relationship between Local Consumption and $\gamma$, $N = 100$}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_100T_20.pdf}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_100T_20_partial.pdf}
-\end{subfigure}\\
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_100T_20_omni.pdf}\\
-\end{subfigure}%
-\caption*{\scriptsize Notes: Each figure plots the average consecutive consumption distance across time as the risk aversion level of users, $\gamma$, varies.
-The top left displays the no recommendation regime, the top right displays the recommendation regime, and the bottom displays the oracle regime.}
-\label{fig:no_rec_risk_aversion}
-\end{figure}
-\addtocounter{figure}{-1}
-
-\begin{figure}[H]
-\caption{Local Consumption and Correlation, $N = 100$}
-\begin{subfigure}{.5\linewidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{figures/rho_pos_consumption_dist_N_100T_20_overall.pdf}
-  \label{fig:sfig1}
-\end{subfigure}%
-\begin{subfigure}{.5\linewidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{figures/rho_zero_consumption_dist_N_100T_20_overall.pdf}
-  \label{fig:sfig2}
-\end{subfigure}
-\caption*{\scriptsize Notes: The figure shows the consecutive consumption path difference between the no recommendation, recommendation, and oracle regime.
-The figure on the left displays the average consecutive consumption distance aggregating over simulations where $\rho > 0$ and the figure on the right displays the average consecutive consumption distance aggregating over simulations where $\rho = 0$.}
-\label{fig:correlation_consumption_path_n_100}
-\end{figure}
-\addtocounter{figure}{-1}
-
-% Paper body
-\begin{figure}[H]
-\captionof{figure}{Relationship between Local Consumption and $\rho$, $N = 100$}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_100T_20.pdf}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_100T_20_partial.pdf}
-\end{subfigure}\\
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_100T_20_omni.pdf}\\
-\end{subfigure}
-\caption*{\scriptsize Notes: Each figure plots the average consecutive consumption distance across time as the inherent correlation between the valuation of the items, $\rho$.
-The top left displays the no recommendation regime, the top right displays the recommendation regime, and the bottom displays the oracle regime.}
-\label{fig:local_consumption_across_rho}
-\end{figure}
-
-\begin{figure}[ht]
-\caption{Relationship between $\gamma$ and Diversity, $N = 100$}
-\includegraphics[width=.45\linewidth]{figures/gamma_diversity_N_100_T_20}
-\caption*{\scriptsize Notes: This figure displays the value of the diversity measure as we vary the inherent correlation between the risk aversion level, $\gamma$.}\label{fig:cor_homo}
-\end{figure}
-
-\clearpage
-
-\section{N = 500}
-
-This section contains the same figures as reported in the main text and appendix, but for $N = 500$ instead of $N = 200$.
-
-\begin{figure}[ht]
-\caption{Relationship between $\beta$ and Homogeneity, $N = 500$}
-\includegraphics[width=.45\linewidth]{figures/beta_homogeneity_N_500_T_20}\label{fig:beta_homo}
-\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the weight of the common value component, $\beta$.
-Each line represents this plot for a single recommendation regime.}
-\end{figure}
-\begin{figure}[ht]
-\caption{Relationship between $\rho$ and Homogeneity, $N = 500$}
-\includegraphics[width=.45\linewidth]{figures/rho_homogeneity_N_500_T_20}
-\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the inherent correlation between the valuation of the items, $\rho$.
-Each line represents this plot for a single recommendation regime.}\label{fig:cor_homo}
-\end{figure}
-
-\begin{figure}[H]
-\captionof{figure}{Relationship between Local Consumption and $\gamma$, $N = 500$}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_500T_20.pdf}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_500T_20_partial.pdf}
-\end{subfigure}\\
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_500T_20_omni.pdf}\\
-\end{subfigure}%
-\caption*{\scriptsize Notes: Each figure plots the average consecutive consumption distance across time as the risk aversion level of users, $\gamma$, varies.
-The top left displays the no recommendation regime, the top right displays the recommendation regime, and the bottom displays the oracle regime.}
-\label{fig:no_rec_risk_aversion}
-\end{figure}
-
-\addtocounter{figure}{-1}
-
-% Paper body
-\begin{figure}[H]
-\caption{Relationship between Local Consumption and $\rho$, $N = 500$}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_500T_20.pdf}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_500T_20_partial.pdf}
-\end{subfigure}\\
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_500T_20_omni.pdf}\\
-\end{subfigure}
-\caption*{\scriptsize Notes: Each figure plots the average consecutive consumption distance across time as the inherent correlation between the valuation of the items, $\rho$.
-The top left displays the no recommendation regime, the top right displays the recommendation regime, and the bottom displays the oracle regime.}
-\label{fig:local_consumption_across_rho}
-\end{figure}
-
-\addtocounter{figure}{-1}
-
-\begin{figure}[H]
-\caption{Local Consumption and Correlation, $N = 500$}
-\begin{subfigure}{.5\linewidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{figures/rho_pos_consumption_dist_N_500T_20_overall.pdf}
-  \label{fig:sfig1}
-\end{subfigure}%
-\begin{subfigure}{.5\linewidth}
-  \centering
-  \includegraphics[width=.9\linewidth]{figures/rho_zero_consumption_dist_N_500T_20_overall.pdf}
-  \label{fig:sfig2}
-\end{subfigure}
-\caption*{\scriptsize Notes: The figure shows the consecutive consumption path difference between the no recommendation, recommendation, and oracle regime.
-The figure on the left displays the average consecutive consumption distance aggregating over simulations where $\rho > 0$ and the figure on the right displays the average consecutive consumption distance aggregating over simulations where $\rho = 0$.}
-\label{fig:correlation_consumption_path_n_100}
-\end{figure}
-
-\begin{figure}[ht]
-\caption{Relationship between $\gamma$ and Diversity, $N = 500$}
-\includegraphics[width=.45\linewidth]{figures/gamma_diversity_N_500_T_20}
-\caption*{\scriptsize Notes: This figure displays the value of the diversity measure as we vary the inherent correlation between the risk aversion level, $\gamma$.}\label{fig:cor_homo}
-\end{figure}
-
 % Bibliography

### 2020-05-22 paper/filter_bubbles.tex
shorthand

@@
-\documentclass[format=acmsmall, review=false]{acmart}
+\documentclass[manuscript]{acmart}
 \settopmatter{printacmref=false, printccs=false}
-\usepackage{booktabs} % For formal tables
-\usepackage[ruled]{algorithm2e} % For algorithms
 \usepackage{enumitem}
@@
 \newcommand{\xhdr}[1]{\vspace{1mm} \noindent{\bf #1}}
-\renewcommand{\footnotetextcopyrightpermission}[1]{} % remove copyright block
-\renewcommand{\footnotetextauthorsaddresses}[1]{} % remove author address 
-\renewcommand{\algorithmcfname}{ALGORITHM}
-\SetAlFnt{\small}
-\SetAlCapFnt{\small}
-\SetAlCapNameFnt{\small}
-\SetAlCapHSkip{0pt}
-\IncMargin{-\parindent}
 
@@
 % Title. Note the optional short title for running heads. In the interest of anonymization, please do not include any acknowledgements.
-\newcommand\PaperTitle{Deconstructing the Filter Bubble: \\ Consumer Decision-Making and Recommender Systems}
+\newcommand\PaperTitle{Deconstructing the Filter Bubble: Consumer Decision-Making and RSs}
 \title[\PaperTitle]{\PaperTitle}
@@
 \begin{abstract}
-We study a model of user decision-making in the context of recommender systems via numerical simulation.
+We study a model of user decision-making in the context of RSs via numerical simulation.
 Our model provides an explanation for the findings of Nguyen, et.
 al (2014), where users consume increasingly similar items over time even without recommendation.
-Our model naturally generates such filter bubble effects since users face a sequential decision making problem under uncertainty and there are informational spillovers where a user consuming a item and learning how she values it reduces uncertainty and provides information about similar items.
-As shown in Nguyen, et.
-al (2014), recommendation alleviates these filter bubble effects and we further find that, by providing users with information, recommendation leads users to consume increasingly similar items.
+Our model naturally generates such filter bubble effects since users face a sequential decision making problem under uncertainty and there are informational spillovers where a user consuming an item and learning how she values it reduces uncertainty and provides information about similar items.
+\par
+In line with Nguyen, et.
+al (2014), recommendation alleviates these filter bubble effects.
+However, we find that recommendation leads to an increase in homogeneity across users, resulting in a trade-off between homogenizing across user consumption and diversifying within user consumption.
 Finally, our model highlights the importance of collecting data on user beliefs and their evolution over time both to design better recommendations and to further understand their impact.
@@
 Users in many cases are not aware of most items, let alone have full information about their preferences over them.
-To make matters worse, the item in these contexts are usually experience goods whose true value can only be learned after consumption.
+To make matters worse, the items in these contexts are usually experience goods whose true value for users can only be learned after consumption.
 \par
 
-Recommender systems have driven a significant fraction of consumer choice on these platforms with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{MacKenzie et al.
+RS have driven a significant fraction of consumer choice on these platforms with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{MacKenzie et al.
 (2013, Oct.),  How retailers can keep up with consumers.
@@
 Retrieved on October 3, 2019.} While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems.
-There have been claims that personalized recommender systems lead users into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
+There have been claims that personalized RSs lead users into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized RSs may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
 \par
-Understanding how recommender systems influence user behavior is important not only for characterizing the broader consequences of such systems but also for guiding their design.
-In this paper we develop a theoretical model of user decision-making in contexts where recommender systems are traditionally deployed.
-We utilize previous empirical studies that characterize how recommender systems influence user choice as a benchmark and our theoretical model provides an intuitive mechanism that can explain these empirical results.
+Understanding how RS influence user behavior is important not only for characterizing the broader consequences of such systems but also for guiding their design.
+In this paper we develop a theoretical model of user decision-making in contexts where RS are traditionally deployed.
+We utilize previous empirical studies that characterize how RS influence user choice as a benchmark and our theoretical model provides an intuitive mechanism that can explain these empirical results.
 The key insight of our model is that user \textit{beliefs} drive the consumption choices of users and that recommendations provide them with information that leads them to update their beliefs and alter their choices.
 A crucial component of our model is that users' beliefs about items are driven not only by recommendations, but also from their previous experiences with similar items.
-We use these insights to provide guidance for recommender system design.
+We use these insights to provide guidance for RS design.
 In order to design good recommendations it becomes important for designers to understand user beliefs of the goods they haven't consumed thus far and track how these beliefs evolve over time.
@@
 The second component is that users have \textit{uncertainty} about their valuation of the different items before they consume them.
-This is motivated both by the fact that many contexts where recommender systems are deployed contain experience goods, whose true value can only be learned after consumption, and the fact that such uncertainty is why recommender systems exist in the first place.
+This is motivated both by the fact that many contexts where RSs are deployed contain experience goods, whose true value can only be learned after consumption, and the fact that such uncertainty is why RSs exist in the first place.
 Thus, users face a sequential decision-making problem under uncertainty.
@@
 \par
-Lastly, we discuss how our model and findings can be used to inform the design and evaluation of recommender systems as well as the data that is traditionally collected for them.
+Lastly, we discuss how our model and findings can be used to inform the design and evaluation of RSs as well as the data that is traditionally collected for them.
 This highlights the importance of user beliefs in determining user consumption choices and how both recommendation and informational spillovers determine how these beliefs change over time.
-By collecting information on user beliefs, recommender system designers can understand what items a user would consume \textit{without} recommendation and then predict how providing information to the user would change her beliefs and resulting consumption decisions.
+By collecting information on user beliefs, RS designers can understand what items a user would consume \textit{without} recommendation and then predict how providing information to the user would change her beliefs and resulting consumption decisions.
 Thus,  our evaluation measure determines the value of a recommendation based on the marginal welfare gain associated with providing a user with a recommendation over what the user would do without it.
@@
 The most relevant to our study is \cite{nguyen2014exploring} who study whether this effect exists in the context of movie consumption.
-They find that even users whose consumption choices are not guided by recommendations exhibit behavior consistent with ``filter bubbles'' and that recommender systems can actually increase the diversity of the content that users consume.
+They find that even users whose consumption choices are not guided by recommendations exhibit behavior consistent with ``filter bubbles'' and that RSs can actually increase the diversity of the content that users consume.
 To our knowledge there are no theoretical models that rationalize these empirical findings and we provide a theoretical framework through which to view this problem.
-Furthermore, our paper provides a clear mechanism at play that drives such effects and how recommendation interacts with them.
+Furthermore, we provide a clear mechanism that drives such effects and how recommendation interacts with them.
 \par 
-Another set of papers has examined whether recommender systems can lead users to become increasingly homogenized or consume similar content to one another.
-\cite{celma2008hits, treviranus2009value} show that incorporating content popularity into recommender systems can lead to increased user homogenization.
-\cite{chaney2018algorithmic} shows how user homogenization may arise from training recommender systems on data from users exposed to algorithmic recommendations.
-\cite{fleder2009blockbuster} show that homogenization can increase due to a recommendation bias towards popularity that arises from lack of information about items with limited consumption histories.
-Our model leads to similar results as previous work where recommender systems lead to increased user homogenization.
+Another set of papers has examined whether RSs can lead users to become increasingly homogenized.
+\cite{celma2008hits, treviranus2009value} show that incorporating content popularity into RSs can lead to increased user homogenization.
+\cite{chaney2018algorithmic} shows how user homogenization may arise from training RSs on data from users exposed to algorithmic recommendations.
+\cite{fleder2009blockbuster} show that homogenization can increase due to a popularity recommendation bias that arises from lack of information about items with limited consumption histories.
+We show similar results as previous work where RSs lead to increased user homogenization.
 However, the mechanisms behind this differ from existing work as homogenization arises due to the fact that recommendation leads users to coordinate their consumption decisions in certain portions of the product space.
 \par
-Another set of papers studies the impact of human decision-making on the design and evaluation of recommender systems.
-\cite{chen2013human} surveys the literature on the relationship between human decision making and recommender systems.
+Another set of papers studies the impact of human decision-making on the design and evaluation of RSs.
+\cite{chen2013human} surveys the literature on the relationship between human decision making and RSs.
 The closest set of papers pointed out in this survey are those related to preference construction \cite{bettman1998constructive, lichtenstein2006construction} whereby users develop preferences over time through the context of a decision process.
-We point out that the true underlying preferences of users may be stable over time, but, due to the nature of items in contexts where recommender systems are deployed, they have incomplete information of their valuations and both consumption and recommendation provide them with information to reduce their uncertainty.
-Thus, the primary insight of our paper is that user beliefs and how users update their beliefs about similar items after consumption are important and previously unconsidered elements of human decision making that are critical for understanding the design and consequences of recommender systems.
+We point out that the true underlying preferences of users may be stable over time, but, due to the nature of items in contexts where RSs are deployed, they have incomplete information of their valuations and both consumption and recommendation provide them with information to reduce their uncertainty.
+Thus, the primary insight of our paper is that user beliefs and how users update their beliefs about similar items after consumption are important and previously unconsidered elements of human decision making that are critical for understanding the design and consequences of RSs.
 Within this literature, \cite{celma2008new, cremonesi2013user, pu2011user} focus on ``user-centric" approaches to recommendation whereby user evaluation of the usefulness of recommendation is a key evaluation measure.
@@
 Informativeness, for any $\rho \in (0,1)$, is decreasing in distance.
-The particular distance function that we utilize leads to this covariance structure being simple, where the $(n,n+1)$-th entry in the covariance matrix is $\sigma^{2} \rho$, the $(n,n+2)$-th entry is $\sigma^{2} \rho^2$, etc.
-\footnote{This exponential decay correlation structure can be related to the tenet of case-based similarity of \cite{gilboa1995case} -- see \cite{billot2008axiomatization} for an axiomatization of exponential similarity.}
+The particular distance function that we utilize leads to this covariance structure being simple, where the $(n,n+1)$-th entry in the covariance matrix is $\sigma^{2} \rho$, the $(n,n+2)$-th entry is $\sigma^{2} \rho^2$, etc.\footnote{This exponential decay correlation structure can be related to the tenet of case-based similarity of \cite{gilboa1995case} -- see \cite{billot2008axiomatization} for an axiomatization of exponential similarity.}
 \par
@@
 \xhdr{Recommendation.}
-Our model of recommendation is stylized in order to provide qualitative insights into how recommendation shapes behavior, instead of focusing on the details of how recommender systems are implemented in practice.
+Our model of recommendation is stylized in order to provide qualitative insights into how recommendation shapes behavior, instead of focusing on the details of how RSs are implemented in practice.
 We model recommendation as giving users information about the valuation of the products.
@@
 The range of considered parameter values cover the relevant portions of the parameter space in order to provide full insight into the behavior of the model.
-\par
 When we consider results varying a single parameter, we group the results over the other parameters and provide reports varying only the parameter of interest.
@@
 \caption{Local Consumption and Correlation, $N = 200$}
-\begin{subfigure}{.5\linewidth}
+\begin{subfigure}{.3\linewidth}
   \centering
-  \includegraphics[width=.9\linewidth]{figures/rho_pos_consumption_dist_N_200T_20_overall.pdf}
+  \includegraphics[width=1.0\linewidth]{figures/rho_pos_consumption_dist_N_200T_20_overall.pdf}
   \label{fig:sfig1}
 \end{subfigure}%
-\begin{subfigure}{.5\linewidth}
+\begin{subfigure}{.3\linewidth}
   \centering
-  \includegraphics[width=.9\linewidth]{figures/rho_zero_consumption_dist_N_200T_20_overall.pdf}
+  \includegraphics[width=1.0\linewidth]{figures/rho_zero_consumption_dist_N_200T_20_overall.pdf}
   \label{fig:sfig2}
@@
 
+\begin{figure}[ht]
+\caption{Relationship between Local Consumption and $\rho$, $N = 200$}
+\begin{subfigure}{.3\textwidth}
+\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20.pdf}
+\end{subfigure}
+\begin{subfigure}{.3\textwidth}
+\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_partial.pdf}
+\end{subfigure}
+\begin{subfigure}{.3\textwidth}
+\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_omni.pdf}\\
+\end{subfigure}
+\begin{subfigure}{.3\textwidth}
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20.pdf}
+\end{subfigure}
+\begin{subfigure}{.3\textwidth}
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_partial.pdf}
+\end{subfigure}
+\begin{subfigure}{.3\textwidth}
+\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_omni.pdf}
+\end{subfigure}%
+\caption*{\scriptsize Notes: Each figure plots the average consecutive consumption distance across time as the inherent correlation between the valuation of the items, $\rho$.
+The top left displays the no recommendation regime, the top right displays the recommendation regime, and the bottom displays the oracle regime.
+Notes: Each figure plots the average consecutive consumption distance across time as the risk aversion level of users, $\gamma$, varies.
+The top left displays the no recommendation regime, the top right displays the recommendation regime, and the bottom displays the oracle regime.}
+\label{fig:local_consumption_across_rho}
+\end{figure}
+
 \section{Results}
@@
 We define local consumption in terms of the average consumption distance between the items consumed by the users at time $t-1$ and $t$.
-Thus, in the context of our model, filter bubble effects arise when the average consumption distance decreases over time and, across regimes, when the overall levels are lower for a given recommendation regime compared to another.
+Thus, in the context of our model, filter bubble effects arise when the average consumption distance decreases over time and, across regimes, when the levels are lower for a given recommendation regime compared to another.
 
@@
 While in the previous section we looked at the distance between consecutive items, in this section we focus on a diversity measure that considers the entire consumed set of items.
-The diversity measure we utilize is common in recommender system literature (e.g.
+The diversity measure we utilize is common in RS literature (e.g.
 \cite{ziegler2005improving}) which is the average normalized pairwise distance between the consumed items:
@@
 \par
+
+\begin{figure}[ht]
+\caption{Diversity vs. Welfare, \textbf{No Recommendation}, N = 200}
+\begin{subfigure}{.3\textwidth}
+\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_n_200.pdf}
+\end{subfigure}
+\begin{subfigure}{.3\textwidth}
+\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_n_200.pdf}
+\end{subfigure}\\
+\begin{subfigure}{.3\textwidth}
+\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_partial_n_200.pdf}
+\end{subfigure}
+\begin{subfigure}{.3\textwidth}
+\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_partial_n_200.pdf}
+\end{subfigure}\\
+\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under the no recommendation regime.
+The figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.
+Notes: The figure plots the relationship between diversity and welfare under the recommendation regime.
+The figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.}\label{fig:diversity_welfare_ra}
+\end{figure}
+\addtocounter{figure}{-1}
 We now study how recommendation impacts user welfare in our model.
@@
 \par 
-In the supplementary material (Figure 10), we show how diversity and welfare vary across recommendation regimes as the the strength of the common-value component, $\beta$, varies.
-As expected, when $\beta = 0$, these values are the same.
-However, as $\beta$ grows the idiosyncratic component of user valuations becomes increasingly less important and the recommendation regime begins to converge to the oracle regime.
-Indeed, we see that both product diversity and welfare increase under recommendation, whereas under no recommendation we see that there is a decrease in diversity with little changes to user welfare.
-This is intuitive as $\beta$ controls the informativeness of recommendation about user preferences.
-\par 
 One striking observation is that the decrease in diversity does not appear to be associated with a decline in welfare.
@@
 
-\addtocounter{figure}{-1}
 \begin{figure}[ht]
-\caption{Diversity vs. Welfare, \textbf{No Recommendation}, N = 200}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_n_200.pdf}
+\caption{Relationship between $\rho$ and User Welfare, Diversity, $N = 200$}\label{fig:diversity_welfare_correlation}
+\begin{subfigure}{0.3\linewidth}
+  \includegraphics[width=1.0\linewidth]{figures/rho_diversity_N_200_T_20.pdf}
 \end{subfigure}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_n_200.pdf}
-\end{subfigure}\\
-\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under the no recommendation regime.
-The figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.}\label{fig:diversity_welfare_ra}
+\ContinuedFloat
+\begin{subfigure}{0.3\linewidth}
+  \includegraphics[width=1.0\linewidth]{figures/rho_welfare_N_200_T_20.pdf}
+\end{subfigure}
+\begin{subfigure}{0.3\linewidth}
+\includegraphics[width=1.0\linewidth]{figures/gamma_diversity_N_200_T_20}
+\end{subfigure}
+\caption*{\scriptsize Notes: The figure on the left displays the relationship between $\rho$, the correlation between valuations of items, and overall consumed product diversity.
+The figure on the right displays the relationship between $\rho$ and overall welfare.}
 \end{figure}
 
-\addtocounter{figure}{-1}
-\begin{figure}[ht]
-\caption{Diversity vs. Welfare, \textbf{Recommendation}, N  = 200}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_partial_n_200.pdf}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_partial_n_200.pdf}
-\end{subfigure}\\
-\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under the recommendation regime.
-The figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.}\label{fig:diversity_welfare_ra_partial}
-\end{figure}
 \addtocounter{figure}{-1}
@@
 When there is no risk-aversion then there is a negative correlation between welfare and diversity.
-This is since, with no risk-aversion, in a given period a user will select the item that she currently believes has the highest expected value.
+This is since, with no risk-aversion, a user will select the item that she currently believes has the highest expected value.
 High product diversity in this case can arise from a user who consumes an item she disliked and updates her beliefs about nearby items negatively.
@@
 $\gamma = 5$) though diversity and welfare are virtually uncorrelated in the data.
-In the recommendation and oracle regimes, under risk neutrality ($\gamma=0$), welfare and diversity and uncorrelated, while under risk aversion ($\gamma=5$), it is possible to observe in Figure \ref{fig:diversity_welfare_ra_partial} an actual positive relation between diversity and welfare as recommendations are able to reduce uncertainty and facilitate exploration of the product space.
+In the recommendation and oracle regimes, under risk neutrality ($\gamma=0$), welfare and diversity are uncorrelated, while under risk aversion ($\gamma=5$), it is possible to observe an actual positive relation between diversity and welfare as recommendations are able to reduce uncertainty and facilitate exploration of the product space.
 
@@
 \caption{Relationship between $\beta$ and Homogeneity, $N = 200$}
-\includegraphics[width=.45\linewidth]{figures/beta_homogeneity_N_200_T_20}\label{fig:beta_homo}
-\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the weight of the common value component, $\beta$.
+\includegraphics[width=.3\linewidth]{figures/beta_homogeneity_N_200_T_20}
+\includegraphics[width=.3\linewidth]{figures/rho_homogeneity_N_200_T_20}\label{fig:beta_homo}
+\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the weight of the common value component, $\beta$ (left) and correlation between valuations, $\rho$ (right).
 Each line represents this plot for a single recommendation regime.}
@@
 In particular we look at the degree of \textit{homogenization} between users.
-Similar to other papers that study the degree of homogenization in recommender systems (e.g.
+Similar to other papers that study the degree of homogenization in RSs (e.g.
 \cite{chaney2018algorithmic}) we measure homogeneity via the Jaccard index between the consumption sets of users:
@@
 
-\section{Recommender System Evaluation}
-In this section we discuss how the insights from our model of user decision-making can inform the evaluation and design of recommender systems.
-The classic approach to recommender system evaluation is to predict user ratings for items and to compare how accurate this prediction is to recorded ratings data, either explicitly given by users or inferred from behavioral data.
-The recommender system should then recommend the items with the highest predicted ratings \cite{adomavicius2005toward}.
+\section{RS Evaluation}
+In this section we discuss how the insights from our model of user decision-making can inform the evaluation and design of RSs.
+The classic approach to evaluation is to predict user ratings for items and to compare how accurate this prediction is to recorded ratings data, either explicitly given by users or inferred from behavioral data.
+The RS should then recommend the items with the highest predicted ratings \cite{adomavicius2005toward}.
 \par
@@
 Consider the domain of movie recommendation and suppose a user has just watched the movie \textit{John Wick} and rated it highly.
-A recommender system attempting to predict user ratings may then predict that this user is very likely to enjoy the sequel, \textit{John Wick: Chapter Two}, as well.
+A RS attempting to predict user ratings may then predict that this user is very likely to enjoy the sequel, \textit{John Wick: Chapter Two}, as well.
 However, the user herself may also have made this inference since the two movies are very similar to each other.
 Thus, recommending this movie would not be not useful since the recommendation gives the user little information that she did not already know.
-The key insight is that the recommendation is not useful since \textit{it ignores the inference the user themselves made and their updated beliefs}.
+The key insight is it is not useful since \textit{it ignores the inference the user themselves made and their updated beliefs}.
 The user may watch \textit{John Wick: Chapter Two}, then, even without recommendation, and the value of the recommendation was small.
 \par
-This intuition implies that recommender systems should collect additional data beyond the type of data that is traditionally recorded.
+This intuition implies that RSs should collect additional data beyond the type of data that is traditionally recorded.
 The first and most crucial type of data to collect is individual user \textit{beliefs} about items that they have not yet consumed.
-As illustrated by our model, these beliefs are what drive the future consumption decisions of users and understanding these beliefs is crucial for determining the value of recommending certain items.\footnote{Additionally, user beliefs contain information that may not have been observed by the recommender that only observes user choices on the platform.
-There are many sources of information that affect a user's beliefs and choices that are unobservable by recommender systems such as ratings, reviews, and friends.
-However, this information can be inferred by the platform by collecting both beliefs and consumption choices.} The second type of data that is critical for recommender system designers to collect is how user beliefs change over time and, in particular, not just how individuals value the item they just consumed, but also how it impacts their beliefs about the quality of similar items.\footnote{Characterizing the similarity between items has been an important goal of designing content-based recommendations, though as noted by \cite{winecoff2019users}, how users perceive similarity between items is not always in line with how similarity is computed in content-based recommender systems.
-Further understanding how users perceive item similarity and, in particular, how this impacts which items users update their beliefs about is an important direction for future work.} The third type of data is the risk-aversion levels of users as our model illustrates that the risk preferences of users are important for understanding what information recommender systems can provide that materially leads users to alter their consumption patterns.
+As illustrated by our model, these beliefs are what drive the future consumption decisions of users and understanding these beliefs is crucial for determining the value of recommending certain items.\footnote{Additionally, user beliefs contain information that may not have been observed by the recommender that only observes user choices on the platform.} The second type of data that is critical for RS designers to collect is how user beliefs change over time and, in particular, not just how individuals value the item they just consumed, but also how it impacts their beliefs about the quality of similar items.\footnote{Characterizing the similarity between items has been an important goal of designing content-based recommendations, though as noted by \cite{winecoff2019users}, how users perceive similarity between items is not always in line with how similarity is computed in content-based RSs.
+Understanding how this impacts which items users update their beliefs about is an important direction for future work.} The third type of data is the risk-aversion levels of users as our model illustrates that the risk preferences of users are important for understanding what information RSs can provide that materially leads users to alter their consumption patterns.
 \par 
@@
 For instance, \cite{vargas2011rank, kaminskas2014measuring} propose unexpectedness metrics that look at the dissimilarity of the proposed recommended items compared to what the recommender already knows the user likes - either via a content-based approach or a collaborative-based approach.
-This metric depends only on the resulting item-set that the recommender proposes and not necessarily on the user's beliefs or how such a recommendation will change the item that the user consumes.
+This metric depends only on the proposed item-set and not necessarily on the user's beliefs or how such a recommendation will change the item that the user consumes.
 \cite{adamopoulos2015unexpectedness} propose an expected utility approach but use this to measure unexpectedness in terms of deviations from the ``expected consideration set" of a user and do not explicitly incorporate user beliefs.
@@
 It further results in ``unexpected" recommendations since it explicitly incorporates user beliefs and thus allows the recommender to understand how ``unexpected" a recommendation would be from the perspective of a user.
-Thus, a good recommendation provides users with information that leads them to consume an item that they would not have expected to consume without the information from the recommender and that they enjoyed consuming more than the item they would have without recommendation.
 % Bibliography

### 2020-05-27 paper/filter_bubbles.tex
updated TeX

@@
 % Title. Note the optional short title for running heads. In the interest of anonymization, please do not include any acknowledgements.
-\newcommand\PaperTitle{Deconstructing the Filter Bubble: Consumer Decision-Making and RSs}
+\newcommand\PaperTitle{Deconstructing the Filter Bubble: \\User Decision-Making and Recommender Systems}
 \title[\PaperTitle]{\PaperTitle}
@@
 \begin{abstract}
-We study a model of user decision-making in the context of RSs via numerical simulation.
+We study a model of user decision-making in the context of recommender systems via numerical simulation.
 Our model provides an explanation for the findings of Nguyen, et.
@@
 al (2014), recommendation alleviates these filter bubble effects.
-However, we find that recommendation leads to an increase in homogeneity across users, resulting in a trade-off between homogenizing across user consumption and diversifying within user consumption.
+However, we find that it also leads to an increase in homogeneity across users, resulting in a trade-off between homogenizing across user consumption and diversifying
+within user consumption.
 Finally, our model highlights the importance of collecting data on user beliefs and their evolution over time both to design better recommendations and to further understand their impact.
@@
 \url{https://www.mckinsey.com/industries/retail/our-insights/how-retailers-can-keep-up-with-consumers}.
-Retrieved on October 3, 2019.} While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems.
-There have been claims that personalized RSs lead users into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized RSs may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
+Retrieved on October 3, 2019.} While there are many positive effects from these systems, there is an increasing worry about their unintended side-effects.
+There have been claims that personalized RS lead users into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized RS may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
 \par
@@
 The second component is that users have \textit{uncertainty} about their valuation of the different items before they consume them.
-This is motivated both by the fact that many contexts where RSs are deployed contain experience goods, whose true value can only be learned after consumption, and the fact that such uncertainty is why RSs exist in the first place.
+This is motivated both by the fact that many contexts where RS are deployed contain experience goods, whose true value can only be learned after consumption, and the fact that such uncertainty is why RS exist in the first place.
 Thus, users face a sequential decision-making problem under uncertainty.
@@
 \par
-Lastly, we discuss how our model and findings can be used to inform the design and evaluation of RSs as well as the data that is traditionally collected for them.
+Lastly, we discuss how our model and findings can be used to inform the design and evaluation of RS as well as the data that is traditionally collected for them.
 This highlights the importance of user beliefs in determining user consumption choices and how both recommendation and informational spillovers determine how these beliefs change over time.
@@
 The most relevant to our study is \cite{nguyen2014exploring} who study whether this effect exists in the context of movie consumption.
-They find that even users whose consumption choices are not guided by recommendations exhibit behavior consistent with ``filter bubbles'' and that RSs can actually increase the diversity of the content that users consume.
+They find that even users whose consumption choices are not guided by recommendations exhibit behavior consistent with ``filter bubbles'' and that RS can actually increase the diversity of the content that users consume.
 To our knowledge there are no theoretical models that rationalize these empirical findings and we provide a theoretical framework through which to view this problem.
@@
 \par 
-Another set of papers has examined whether RSs can lead users to become increasingly homogenized.
-\cite{celma2008hits, treviranus2009value} show that incorporating content popularity into RSs can lead to increased user homogenization.
-\cite{chaney2018algorithmic} shows how user homogenization may arise from training RSs on data from users exposed to algorithmic recommendations.
+Another set of papers has examined whether RS can lead users to become increasingly homogenized.
+\cite{celma2008hits, treviranus2009value} show that incorporating content popularity into RS can lead to increased user homogenization.
+\cite{chaney2018algorithmic} shows how user homogenization may arise from training RS on data from users exposed to algorithmic recommendations.
 \cite{fleder2009blockbuster} show that homogenization can increase due to a popularity recommendation bias that arises from lack of information about items with limited consumption histories.
-We show similar results as previous work where RSs lead to increased user homogenization.
+We show similar results as previous work where RS lead to increased user homogenization.
 However, the mechanisms behind this differ from existing work as homogenization arises due to the fact that recommendation leads users to coordinate their consumption decisions in certain portions of the product space.
 \par
-Another set of papers studies the impact of human decision-making on the design and evaluation of RSs.
-\cite{chen2013human} surveys the literature on the relationship between human decision making and RSs.
+Another set of papers studies the impact of human decision-making on the design and evaluation of RS.
+\cite{chen2013human} surveys the literature on the relationship between human decision making and RS.
 The closest set of papers pointed out in this survey are those related to preference construction \cite{bettman1998constructive, lichtenstein2006construction} whereby users develop preferences over time through the context of a decision process.
-We point out that the true underlying preferences of users may be stable over time, but, due to the nature of items in contexts where RSs are deployed, they have incomplete information of their valuations and both consumption and recommendation provide them with information to reduce their uncertainty.
-Thus, the primary insight of our paper is that user beliefs and how users update their beliefs about similar items after consumption are important and previously unconsidered elements of human decision making that are critical for understanding the design and consequences of RSs.
+We point out that the true underlying preferences of users may be stable over time, but, due to the nature of items in contexts where RS are deployed, they have incomplete information of their valuations and both consumption and recommendation provide them with information to reduce their uncertainty.
+Thus, the primary insight of our paper is that user beliefs and how users update their beliefs about similar items after consumption are important and previously unconsidered elements of human decision making that are critical for understanding the design and consequences of RS.
 Within this literature, \cite{celma2008new, cremonesi2013user, pu2011user} focus on ``user-centric" approaches to recommendation whereby user evaluation of the usefulness of recommendation is a key evaluation measure.
@@
 \xhdr{Recommendation.}
-Our model of recommendation is stylized in order to provide qualitative insights into how recommendation shapes behavior, instead of focusing on the details of how RSs are implemented in practice.
+Our model of recommendation is stylized in order to provide qualitative insights into how recommendation shapes behavior, instead of focusing on the details of how RS are implemented in practice.
 We model recommendation as giving users information about the valuation of the products.
@@
 When we consider results varying a single parameter, we group the results over the other parameters and provide reports varying only the parameter of interest.
-We report results for a relatively small consumption history $T=20$ with a product space size $N=200$.\footnote{In the supplementary material we include the same exercises presented here for $N = 100$ and $N = 500$ and show that the results are qualitatively similar to those reported here.}
-\par
+We report results for a relatively small consumption history $T=20$ with a product space size $N=200$.\footnote{Due to space constraints, we only report the values for $N = 200$.
+However, we further conducted the same exercises for $N = 100$ and $N = 500$ and the results are qualitatively similar to those reported here.}
+\section{Results}
+\subsection{Local Consumption and Filter Bubbles}
+We characterize ``filter bubble'' effects in our model as the degree to which users engage in \textit{local consumption}.
+We define local consumption in terms of the average consumption distance between the items consumed by the users at time $t-1$ and $t$.
+Thus, in the context of our model, filter bubble effects arise when the average consumption distance decreases over time and, across regimes, when the levels are lower for a given recommendation regime compared to another.
 
-\begin{figure}
+\begin{figure}[ht]
 \caption{Local Consumption and Correlation, $N = 200$}
@@
 
+\addtocounter{figure}{-1}
+
 \begin{figure}[ht]
-\caption{Relationship between Local Consumption and $\rho$, $N = 200$}
+\caption{Relationship between Local Consumption and $\rho$, $\gamma$}
 \begin{subfigure}{.3\textwidth}
@@
 \end{subfigure}%
-\caption*{\scriptsize Notes: Each figure plots the average consecutive consumption distance across time as the inherent correlation between the valuation of the items, $\rho$.
-The top left displays the no recommendation regime, the top right displays the recommendation regime, and the bottom displays the oracle regime.
-Notes: Each figure plots the average consecutive consumption distance across time as the risk aversion level of users, $\gamma$, varies.
-The top left displays the no recommendation regime, the top right displays the recommendation regime, and the bottom displays the oracle regime.}
-\label{fig:local_consumption_across_rho}
+\caption*{\scriptsize Notes: Each figure plots the average consecutive consumption distance across time as the inherent correlation between the valuation of the items, $\rho$, varies (top row) and the level of risk-aversion, $\gamma$, varies (bottom row).
+In both rows, the left displays the no recommendation regime, the center displays the recommendation regime, and the right displays the oracle regime.}
+\label{fig:local_consumption_vary_rho_gamma}
 \end{figure}
-
-\section{Results}
-\subsection{Local Consumption and Filter Bubbles}
-We characterize ``filter bubble'' effects in our model as the degree to which users engage in \textit{local consumption}.
-We define local consumption in terms of the average consumption distance between the items consumed by the users at time $t-1$ and $t$.
-Thus, in the context of our model, filter bubble effects arise when the average consumption distance decreases over time and, across regimes, when the levels are lower for a given recommendation regime compared to another.
+\addtocounter{figure}{-1}
 
@@
 However, since these spillovers also impact user learning in the no recommendation case, users \textit{over-exploit} this and increasingly consume items similar to high value items that they have consumed before.
-This is further illustrated in the supplementary material (Figure 7), which shows how the consumption paths in the oracle and no recommendation regimes vary as $\rho$ increases and is in line with this intuition.
+This is further illustrated in the top row of~\autoref{fig:local_consumption_vary_rho_gamma}, which shows how the consumption paths in the oracle and no recommendation regimes vary as $\rho$ increases and is in line with this intuition.
 \par
 This effect is further amplified as the level of risk aversion, $\gamma$, increases.
-We show in the supplementary material (Figure 8) that the degree of local consumption drastically increases as $\gamma$ increases.
+The bottom row of~\autoref{fig:local_consumption_vary_rho_gamma} validates that the degree of local consumption drastically increases as $\gamma$ increases.
 This is due to the fact that the spillovers not only affect the mean expected belief about quality but also the degree of uncertainty.
@@
 
+\begin{figure}[ht]
+\caption{Relationship between $\rho$, $\gamma$ and User Welfare, Diversity}\label{fig:diversity_welfare_correlation}
+\begin{subfigure}{0.3\linewidth}
+  \includegraphics[width=1.0\linewidth]{figures/rho_diversity_N_200_T_20.pdf}
+\end{subfigure}
+\begin{subfigure}{0.3\linewidth}
+\includegraphics[width=1.0\linewidth]{figures/gamma_diversity_N_200_T_20}
+\end{subfigure}
+\begin{subfigure}{0.3\linewidth}
+  \includegraphics[width=1.0\linewidth]{figures/rho_welfare_N_200_T_20.pdf}
+\end{subfigure}
+\caption*{\scriptsize Notes: The figures on the left and center display the relationship between $\rho$ and overall consumed product diversity (left) as well as $\gamma$ and overall consumed product diversity (center).
+The figure on the right displays the relationship between $\rho$ and overall welfare.}
+\end{figure}
+
+\addtocounter{figure}{-1}
 \subsection{User Welfare and Product Diversity}
@@
 As a result, this effect gets amplified as $\rho$ increases, which leads to the gap in diversity between the regimes to increase as $\rho$ increases.
-In the supplementary material (Figure 5), we show how diversity varies as $\rho$ increases across the three regimes that we consider.
-There is a similar increasing diversity gap as $\gamma$, or the level of risk-aversion, increases as can be seen in the supplementary material (Figure 9).
+The top row of~\autoref{fig:local_consumption_vary_rho_gamma} shows how diversity varies as $\rho$ increases across the three regimes that we consider.
+There is a similar increasing diversity gap as $\gamma$, or the level of risk-aversion, increases as can be seen in the bottom row of~\autoref{fig:local_consumption_vary_rho_gamma}.
 The mechanisms behind these effects have direct parallels to those discussed in the previous section since low average sequential consumption distance directly leads to low diversity.
@@
 \begin{figure}[ht]
-\caption{Diversity vs. Welfare, \textbf{No Recommendation}, N = 200}
+\caption{Diversity vs. Welfare}
 \begin{subfigure}{.3\textwidth}
@@
 \end{subfigure}\\
-\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under the no recommendation regime.
-The figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.
-Notes: The figure plots the relationship between diversity and welfare under the recommendation regime.
-The figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.}\label{fig:diversity_welfare_ra}
+\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under the recommendation (top row)  and no recommendation (bottom row) regime.
+In both rows, the figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.}\label{fig:diversity_welfare_ra}
 \end{figure}
@@
 \par 
-Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which are drawn from the supplementary material (Figure 5).
+Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which can be seen in the rightmost plot of~\autoref{fig:diversity_welfare_correlation}.
 The most interesting observation is that the value of recommendation decreases as $\rho$ decreases.
@@
 We next explore the relationship between welfare and diversity.
-
-\begin{figure}[ht]
-\caption{Relationship between $\rho$ and User Welfare, Diversity, $N = 200$}\label{fig:diversity_welfare_correlation}
-\begin{subfigure}{0.3\linewidth}
-  \includegraphics[width=1.0\linewidth]{figures/rho_diversity_N_200_T_20.pdf}
-\end{subfigure}
-\ContinuedFloat
-\begin{subfigure}{0.3\linewidth}
-  \includegraphics[width=1.0\linewidth]{figures/rho_welfare_N_200_T_20.pdf}
-\end{subfigure}
-\begin{subfigure}{0.3\linewidth}
-\includegraphics[width=1.0\linewidth]{figures/gamma_diversity_N_200_T_20}
-\end{subfigure}
-\caption*{\scriptsize Notes: The figure on the left displays the relationship between $\rho$, the correlation between valuations of items, and overall consumed product diversity.
-The figure on the right displays the relationship between $\rho$ and overall welfare.}
-\end{figure}
-
-\addtocounter{figure}{-1}
 
@@
 In particular we look at the degree of \textit{homogenization} between users.
-Similar to other papers that study the degree of homogenization in RSs (e.g.
+Similar to other papers that study the degree of homogenization in RS (e.g.
 \cite{chaney2018algorithmic}) we measure homogeneity via the Jaccard index between the consumption sets of users:
@@
 First, we study how the degree of homogenization varies as we increase $\beta$, the weight of the common value component.
-As $\beta$ increases we expect that users should become increasingly homogeneous as the realized utilities of the items are now increasingly similar.
-Figure \ref{fig:beta_homo} confirms that as the weight of the common-value component, $\beta$, increases users consume increasingly similar sets of items.
+As $\beta$ increases we expect that users should become increasingly homogeneous as the realized utilities of the items are now increasingly similar.~\autoref{fig:beta_homo} confirms that as the weight of the common-value component $\beta$ increases, users consume increasingly similar sets of items.
 The homogenization effect is strongest under the recommendation regime since the revelation of the common-value component induces users to consume products in similar areas of the product space.
@@
 We next study how the degree of homogeneity varies as $\rho$ increases.
-In the supplementary material (Figure 6) we show how homogeneity varies as $\rho$ increases.
+~\autoref{fig:beta_homo} shows how homogeneity varies as $\rho$ increases.
 The most striking finding is that homogeneity decreases as $\rho$ increases in the recommendation regime.
@@
 
-\section{RS Evaluation}
-In this section we discuss how the insights from our model of user decision-making can inform the evaluation and design of RSs.
+\section{Recommender System Evaluation}
+In this section we discuss how the insights from our model of user decision-making can inform the evaluation and design of recommender systems.
 The classic approach to evaluation is to predict user ratings for items and to compare how accurate this prediction is to recorded ratings data, either explicitly given by users or inferred from behavioral data.
@@
 \par
-This intuition implies that RSs should collect additional data beyond the type of data that is traditionally recorded.
+This intuition implies that RS should collect additional data beyond the type of data that is traditionally recorded.
 The first and most crucial type of data to collect is individual user \textit{beliefs} about items that they have not yet consumed.
-As illustrated by our model, these beliefs are what drive the future consumption decisions of users and understanding these beliefs is crucial for determining the value of recommending certain items.\footnote{Additionally, user beliefs contain information that may not have been observed by the recommender that only observes user choices on the platform.} The second type of data that is critical for RS designers to collect is how user beliefs change over time and, in particular, not just how individuals value the item they just consumed, but also how it impacts their beliefs about the quality of similar items.\footnote{Characterizing the similarity between items has been an important goal of designing content-based recommendations, though as noted by \cite{winecoff2019users}, how users perceive similarity between items is not always in line with how similarity is computed in content-based RSs.
-Understanding how this impacts which items users update their beliefs about is an important direction for future work.} The third type of data is the risk-aversion levels of users as our model illustrates that the risk preferences of users are important for understanding what information RSs can provide that materially leads users to alter their consumption patterns.
+As illustrated by our model, these beliefs are what drive the future consumption decisions of users and understanding these beliefs is crucial for determining the value of recommending certain items.\footnote{Additionally, user beliefs contain information that may not have been observed by the recommender that only observes user choices on the platform.} The second type of data that is critical for RS designers to collect is how user beliefs change over time and, in particular, not just how individuals value the item they just consumed, but also how it impacts their beliefs about the quality of similar items.\footnote{Characterizing the similarity between items has been an important goal of designing content-based recommendations, though as noted by \cite{winecoff2019users}, how users perceive similarity between items is not always in line with how similarity is computed in content-based RS.
+Understanding how this impacts which items users update their beliefs about is an important direction for future work.} The third type of data is the risk-aversion levels of users as our model illustrates that the risk preferences of users are important for understanding what information RS can provide that materially leads users to alter their consumption patterns.
 \par 

### 2020-06-22 paper/filter_bubbles.tex
update

@@
 
+\usepackage{afterpage}
+\usepackage[suppress]{color-edits}
+\addauthor{dg}{red}    % dg for Duarte
+%\addauthor{ga}{blue}  % ga for Guy
+
 % Choose a citation style by commenting/uncommenting the appropriate line:
@@
 \newtheorem{finding}{Finding}
+\setlength{\topsep}{0pt}
 
@@
 al (2014), where users consume increasingly similar items over time even without recommendation.
-Our model naturally generates such filter bubble effects since users face a sequential decision making problem under uncertainty and there are informational spillovers where a user consuming an item and learning how she values it reduces uncertainty and provides information about similar items.
+Our model naturally generates such filter\dgedit{-}bubble effects since users face a sequential decision making problem under uncertainty and there are informational spillovers where a user consuming an item and learning how she values it reduces uncertainty and provides information about similar items.
 \par
 In line with Nguyen, et.
-al (2014), recommendation alleviates these filter bubble effects.
+al (2014), recommendation alleviates these filter\dgedit{-}bubble effects.
 However, we find that it also leads to an increase in homogeneity across users, resulting in a trade-off between homogenizing across user consumption and diversifying
@@
 A crucial component of our model is that users' beliefs about items are driven not only by recommendations, but also from their previous experiences with similar items.
-We use these insights to provide guidance for RS design.
-In order to design good recommendations it becomes important for designers to understand user beliefs of the goods they haven't consumed thus far and track how these beliefs evolve over time.
+\dgdelete{We use these insights to provide guidance for RS design.
+In order to design good recommendations it becomes important for designers to understand user beliefs of the items they haven't consumed thus far and track how these beliefs evolve over time.}\dgedit{We use these insights to provide guidance for RS design, highlighting that understanding users' beliefs about the quality of the available items is essential in order to design recommendations and evaluate their impact.}
 
-\xhdr{Our Model} We consider a model of user choice that has the following components.
+\xhdr{Our Model\dgedit{.}} We \dgedit{analyze} a model of user choice \dgedit{with four central components}.
 \par
@@
 \par
-The second component is that users have \textit{uncertainty} about their valuation of the different items before they consume them.
-This is motivated both by the fact that many contexts where RS are deployed contain experience goods, whose true value can only be learned after consumption, and the fact that such uncertainty is why RS exist in the first place.
+The second component is\dgedit{, prior to consuming them, users are \textit{uncertain} about how much they value the different items.}
+This is motivated both by the fact \dgedit{that, in many contexts, recommendations regard experience goods,} whose true value can only be learned after consumption, and the fact that such uncertainty is why RS exist in the first place.
 Thus, users face a sequential decision-making problem under uncertainty.
 \par 
-The third, and most crucial, element is that consumption of an item reveals information that changes user beliefs about the valuation of similar items.
+The third, and most crucial, element is that consumption of an item reveals information that changes user beliefs about the\dgedit{ir} valuation of similar items.
 Unlike in standard sequential decision-making problems, once an item is consumed all uncertainty about its valuation is resolved and provides information that enables users to update their beliefs about similar items.
 This exploits the fact that the valuations of similar items are correlated which assists users in navigating the vast product space.
-The idea that users make similarity-based assessments to guide their choice has grounding in theoretical work in decision theory \cite{gilboa1995case} and empirical evidence of how users navigate large choice sets \cite{schulz2019structured}.
+The idea that users make similarity-based assessments to guide their choice has grounding in \dgdelete{theoretical work in decision theory \cite{gilboa1995case} and } empirical evidence on how users navigate large choice sets \cite{schulz2019structured}.
 \par 
@@
 
-\xhdr{Our Contributions}
-Our model provides a clear mechanism that explains the empirical results in \cite{nguyen2014exploring} who show that, in the context of movie consumption, user behavior is consistent with filter bubble effects even without recommendation and that recommendation leads to users being less likely to fall into such filter bubbles.
-In this context, filter bubble effects are defined as users consuming items in an increasingly narrow portion of the product space over time.
-The simple and intuitive driving force of this in our model is that preferences for similar items are correlated, which implies that when an item is consumed and the user learns its value, it provides information about similar items.
-Crucially, this not only impacts the underlying belief about the average utility of the item, but also the level of uncertainty.
+\xhdr{Our Contributions\dgedit{.}}
+We provide a clear mechanism that explains the empirical results in \cite{nguyen2014exploring} who show that, in the context of movie consumption, user behavior is consistent with filter\dgedit{-}bubble effects even without recommendation and that recommendation leads to users being less likely to fall into such filter bubbles.
+In this context, filter\dgedit{-}bubble effects are defined as users consuming items in an increasingly narrow portion of the product space over time.
+The simple and intuitive driving force of this\dgdelete{ in our model} is that preferences for similar items are correlated, which implies that when an item is consumed and the user learns its value, it provides information about similar items.
+Crucially, this not only impacts the underlying belief about the \dgedit{expected value of similar items, but also how uncertain the user is about their valuation of them}.
 Consequently, this learning spillover induces users to consume items similar to those they consumed before that had high realized value, leading to an increasing narrowing of consumption towards these regions of the product space.
-This effect is further amplified when users are \textit{risk-averse}, a concept from decision theory where all else being equal, users have a preference for goods with lower uncertainty to those with higher uncertainty.
+This effect is further amplified when users are \textit{risk-averse}, a concept from decision theory where all else being equal, users have a preference for items with lower uncertainty to those with higher uncertainty.
 However, by providing information to users, recommendation leads users to be more likely to explore other portions of the product space, limiting the filter bubble effect.
@@
 \cite{pariser2011filter} first informally described the idea of the ``filter bubble'' which is that online personalization services would lead users down paths of increasingly narrower content so that they would effectively be isolated from a diversity of viewpoints or content.
-Following this a number of empirical studies, done in various disciplines, have since studied the extent to which this phenomenon exists in a wide range of contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}.
+Following this\dgedit{,} a number of empirical studies\dgdelete{, done} in various disciplines, have since studied the extent to which this phenomenon exists in a wide range of contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}.
 The most relevant to our study is \cite{nguyen2014exploring} who study whether this effect exists in the context of movie consumption.
@@
 However, we assume that users have the same utility over money, given by a utility function $u: \mathbb R \to \mathbb R$, strictly increasing and continuous.
-So, ex-post, the value of item $n$ for user $i$ is given by $u(x_{i,n})$ However, before consuming the item, the user does not know exactly how she will value it.
+So, ex-post, the value of item $n$ for user $i$ is given by $u(x_{i,n})$\dgedit{.
+B}efore consuming the item, the user does not know exactly how she will value it.
 In particular, even users that will end up having the same ex-post valuation of item $n$ may differ in their ex-ante valuation because they hold different beliefs about it.
@@
 This sure amount of money is called the \textit{certainty equivalent} of gamble $x$ and is denoted as $\delta(x)$.
-A user $i$ is more risk-averse than another user $j$ when for every gamble $x$, user $i$ requires a lower value to give up the gamble when compared to user $j$, i.e.
-$\delta_i(x)$.
+A user $i$ is more risk-averse than another user $j$ \dgedit{if whenever user $j$ prefers a sure thing to the gamble, then user $i$ does too}\dgdelete{when for every gamble $x$, user $i$ requires a lower value to give up the gamble when compared to user $j$, i.e.
+$\delta_i(x)$}.
 Therefore, a more risk-averse user is more willing to avoid the risk of taking the gamble.
-We assume that the utility function takes a flexible functional form $u(x)=1-\exp(-\gamma x)$ for $\gamma\ne0$ and $u(x)=x$ for $\gamma=0$ -- known as constant absolute risk-aversion preferences (from hereon CARA).
-Higher $\gamma$ implies higher risk-aversion with $\gamma=0$ corresponding to the risk-neutral case and $\gamma>0$ to the risk-averse one.
+We assume that the utility function takes a flexible functional form $u(x)=1-\exp(-\gamma x)$ for $\gamma\ne0$ and $u(x)=x$ for $\gamma\dgedit{\to}0$ -- known as constant absolute risk-aversion preferences (from hereon CARA).
+Higher $\gamma$ implies higher risk-aversion with $\gamma\dgedit{\to}0$ corresponding to the risk-neutral case and $\gamma>0$ to the risk-averse one.
 Our formulations here follow standard economic decision-theory (see \cite{mas1995microeconomic} for a textbook treatment of these topics).
@@
 \par
-Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{i,n}\right)_{n \in \mathcal{J}}$ are drawn independently from a jointly normal distribution, where $\overline v_{i,n} \sim \mathcal N (0, \overline \sigma^2)$ are independent and identically distributed.
+\dgedit{Recalling} that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{i,n}\right)_{n \in \mathcal{J}}$ are drawn independently from a jointly normal distribution, where $\overline v_{i,n} \sim \mathcal N (0, \overline \sigma^2)$ are independent and identically distributed.
 These $\overline v_{i,n}$ denote the prior belief that individual $i$ holds about her valuation over item $n$.
@@
 
-\xhdr{User Learning}
+\xhdr{User Learning\dgedit{.}}
 When a user consumes an item $n$ she learns the realized value for that item.
-We assume that learning about the value of item $n$ reveals more about the value associated to items that are closer to it, which captures the idea that trying an item provides more information about similar items than about dissimilar ones.\footnote{\cite{schulz2019structured} empirically studies how individuals solve sequential decision-making problems under uncertainty in large choice sets in the context of mobile food delivery orders.
-They find that individuals engage in similarity-based generalizations where learning about the realized value of a particular good provides them with information about similar items.
+We \dgedit{consider the case where}  learning about the value of item $n$ reveals more about the value associated to items that are closer to it, which captures the idea that trying an item provides more information about similar items than about dissimilar ones.\footnote{\cite{schulz2019structured} empirically studies how individuals solve sequential decision-making problems under uncertainty in large choice sets in the context of mobile food delivery orders.
+They find that individuals engage in similarity-based generalizations where learning about the realized value of a particular item provides them with information about similar items.
 We incorporate this finding into our model in a stylized manner.} In order to have a well-defined notion of similarity we need to define a distance function between items, which we define as $d(n,m):=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}$ where $m$ and $n$ are indices of items in $\mathcal{J}$.
 This distance function is not intended to model the intricacies of a realistic product space, but instead to provide a stylized product space to help us understand the effects of informational spillovers on user behavior.
-The basic intuition, in the context of movie consumption, is that a user watching \textit{John Wick} (item $n$) provides more information about how much she will like \textit{John Wick: Chapter Two} (item $m$) than \textit{Titanic} (item $q$) since $d(n, m) < d(n, q)$.
+The basic intuition, in the context of movie consumption, is that a \dgedit{user's valuation of} \textit{John Wick} (item $n$) provides more information about how much she will like \textit{John Wick: Chapter Two} (item $m$) than \textit{Titanic} (item $q$) since $d(n, m) < d(n, q)$.
 \par
@@
 
-\xhdr{An Illustrative Example} We illustrate the main intuitions of our model with a simple example.
+\xhdr{An Illustrative Example\dgedit{.}} We illustrate the main intuitions of our model with a simple example.
 Suppose that there are four items: 0, 1, 2, 3.
 The items are in different places of the product space, where 0 is close to 1 and 3 but more distant from 2.
-The initial beliefs are given as follows:
+\dgedit{For the sake of expositional clarity, suppose that t}he initial beliefs are given as follows:
 \[ \mu = \left (\begin{array}{c}
@@
 0
-\end{array}  \right), \Sigma =  \sigma^{2} \left( \begin{array}{cccc}
+\end{array}  \right), \dgedit{\qquad}\Sigma =  \sigma^{2} \left( \begin{array}{cccc}
 \rho^{0} & \rho^{1} & \rho^{2} & \rho^{0} \\
@@
 In period 1, every item is ex-ante identical since they have the same mean and variance and so suppose that the user breaks the tie arbitrarily and consumes item 0.
-The underlying correlation structure implies that upon observing that $x_0=y$ the user will update beliefs about the remaining 3 items according to the previously specified updating rule.
+The underlying correlation structure implies that upon observing that $x_0 = y$ the user will update beliefs about the remaining 3 items according to the previously specified updating rule.
 For concreteness, we suppose that $\sigma = 1$ and $\rho = 0.5$, but the intuitions hold for any value of $\sigma$ and $\rho > 0$.
-First, we consider the case when the realization of $y > 0$ and, for concreteness, specifically when $y = 0.5$ though the general intuitions hold for any $y > 0$.
+First, we consider the case when the realization of $y > 0$ and\dgdelete{, for concreteness}, specifically\dgedit{,} \dgdelete{when} $y = 0.5$\dgedit{ --} though the general intuitions hold for any $y > 0$.
 The resulting beliefs after observing $y$ are then as follows:
@@
 \]
-Thus, upon learning $y$, the user updates beliefs about the remaining items.
+Thus, upon learning $\dgedit{x_0=}y$, the user updates beliefs about the remaining items.
 Note that $\mathbb{E}[x_1] = \mathbb{E}[x_3] > \mathbb{E}[x_2]$ since item 0's value is more informative about similar items' values, items 1 and 3, than items further way in the product space such as item 2.
@@
 
-We further consider two cases that serve primarily as benchmarks.
+We further consider two cases that serve \dgedit{mainly} as benchmarks.
 The first is \textit{no recommendation}, where users get no additional information and make consumption choices based on their beliefs and consumption history.
-This gives us a benchmark as to how users in our model would behave \textit{without} recommendation so that we can analyze what changes with the introduction of recommendation.
-The second is the \textit{oracle recommendation} where the recommender knows the true realized utility of each good for each user and can therefore recommend the best remaining good in every period.
+This gives us a benchmark as to how users would behave \textit{without} recommendation so that we can analyze what changes with the introduction of recommendation.
+The second is the \textit{oracle recommendation} where the recommender knows the true realized utility of each item for each user and can therefore recommend the best remaining item in every period.
 This gives us a full information benchmark, which is the optimal consumption path for a user if all uncertainty about their preferences was resolved.
@@
 
-\xhdr{Simulation Details}. 
-We analyze our model using numerical simulation since the sequential decision-making component of our model paired with the rich covariance structure between the items make it difficult to characterize optimal user behavior analytically.
+\xhdr{Simulation Details.}
+We analyze our model using numerical simulation since the sequential decision-making component paired with the rich covariance structure between the items make it difficult to characterize optimal user behavior analytically.
 The Gaussian assumption allows for closed form belief updating which enables us to simulate our model but does not provide much help in analytical characterizations.
@@
 We explore how consumption patterns differ as we consider different recommendation regimes and report representative results from our simulations.
-We simulate our model over 100 populations of users with 100 users per population for each combination of parameters.
+We run this simulation over 100 populations of users with 100 users per population for each combination of parameters.
 %For every population and every statistic we calculate, we average over the users in the population to get a representative statistic for this population. 
@@
 \subsection{Local Consumption and Filter Bubbles}
-We characterize ``filter bubble'' effects in our model as the degree to which users engage in \textit{local consumption}.
+We characterize ``filter bubble'' effects as the degree to which users engage in \textit{local consumption}.
 We define local consumption in terms of the average consumption distance between the items consumed by the users at time $t-1$ and $t$.
-Thus, in the context of our model, filter bubble effects arise when the average consumption distance decreases over time and, across regimes, when the levels are lower for a given recommendation regime compared to another.
+Thus, in the context of our model, filter\dgedit{-}bubble effects arise when the average consumption distance decreases over time and, across regimes, when the levels are lower for a given recommendation regime compared to another.
 
-\begin{figure}[ht]
+\dgedit{Our first finding can be summarized as follows:}
+
+\begin{finding}\label{finding_local_consumption}
+The impact of recommendation on local consumption:
+\begin{enumerate}
+\item When $\rho = 0$, there is no difference in consumption distance between the three recommendation regimes.
+\item When $\rho > 0$, no recommendation induces more local consumption than both recommendation and oracle regimes.
+This effect is amplified as $\rho$ increases as well as when users are more risk averse ($\gamma$ increases).
+\end{enumerate}
+\end{finding}
+
+\afterpage{
+\begin{figure}[t]
 \caption{Local Consumption and Correlation, $N = 200$}
@@
 \end{figure}
+\addtocounter{figure}{-1}}
 
-\addtocounter{figure}{-1}
+First, the right panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho = 0$, there is no difference in consumption distance between the three regimes.
+This is due to the fact that when $\rho = 0$, there is no reason that items that are close in the product space should have similar values and so the optimal consumption path does not depend on the similarity of the items.
+However this also means that users do not learn anything about neighboring products and so there is limited path-dependence in consumption.
+Not only is there no difference in the levels between the three regimes, but they all have the same, flat, average consumption distance path.
+This underscores the fact that if there were no correlation between the realized utilities then there would be no reason for users to consume similar items and thus no narrowing effect, regardless of the information on the true utility of the items that users had.
+\par
+The left panel of Figure~\ref{fig:correlation_consumption_path} shows that, when $\rho > 0$, both recommendation and no recommendation lead to increasingly local consumption compared to the oracle benchmark case.
+Further, the average consumption path between periods is \textit{decreasing} for the no recommendation case whereas it is \textit{increasing} for the oracle case.
+The recommendation regime decreases the degree of local consumption, but not as much as the oracle benchmark.
+Due to the correlation of values, the oracle consumption path exploits this and leads to the consumption of more similar items than in the case when $\rho = 0$.
+However, since these spillovers also impact user learning in the no recommendation case, users \textit{over-exploit} this and increasingly consume items similar to high value items that they have consumed before.
+This is further illustrated in the top row of~\autoref{fig:local_consumption_vary_rho_gamma}, which shows how the consumption paths in the oracle and no recommendation regimes vary as $\rho$ increases and is in line with this intuition.
+\par
+This effect is further amplified as the level of risk aversion, $\gamma$, increases.
+The bottom row of~\autoref{fig:local_consumption_vary_rho_gamma} vaFurtherlidates that the degree of local consumption drastically increases as $\gamma$ increases.
+This is due to the fact that the spillovers not only impact the mean expected belief about quality but also the degree of uncertainty.
+Local consumption therefore leads to users having less uncertainty about certain areas of the product space and risk aversion may lead them to increasingly consume nearby products.
 
-\begin{figure}[ht]
+In sum, filter\dgedit{-}bubble effects only arise when there is an inherent correlation between the realized utilities of the items in the product space.
+When there is a correlation between the realized utilities then filter bubbles can naturally arise due to the nature of how individuals acquire additional information about the remaining items.
+We have shown that, unless users are provided with additional information to guide their consumption choices, then these information spillovers and user risk-aversion can lead users into filter bubbles.
+When users consume high valued items, they exploit the underlying correlation across different items' values, stronger for similar items, which leads them to increasingly consume products in narrower and narrower portions of the product space.
+Risk aversion may lead users into performing local consumption even when they have a low valuation of nearby items just because they know what to expect from the item.
+Recommendation leads to these effects being mitigated by providing users with additional information on products outside the already explored portions of the product space.
+Furthermore, if all uncertainty was resolved as in the oracle regime, then such behavior is not present.
+
+\subsection{User Welfare and Product Diversity}
+
+In this section we primarily focus on the impact of recommendation on user welfare and the overall diversity of the items that they consume.
+While in the previous section we looked at the distance between consecutive items, in this section we focus on a diversity measure that considers the entire consumed set of items.
+The diversity measure we utilize is common in RS literature (e.g.
+\cite{ziegler2005improving}) which is the average normalized pairwise distance between the consumed items:
+$$D_i:=\frac{1}{N}\frac{1}{T(T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$$
+\dgedit{\noindent 
+Finding~\ref{finding_diversity} summarizes the main results on product diversity:}
+\begin{finding}\label{finding_diversity}
+The impact of recommendation on product diversity:
+\begin{enumerate}
+\item When $\rho = 0$, product diversity is the same across all three recommendation regimes;
+\item When $\rho > 0$, product diversity decreases across all recommendation regimes but decreases the most in the no recommendation regime.
+This effect is amplified as $\rho$ increases as well as when users become more risk-averse.
+\end{enumerate}
+\end{finding}
+\par
+
+\begin{figure}[t]
 \caption{Relationship between Local Consumption and $\rho$, $\gamma$}
@@
 \addtocounter{figure}{-1}
+\dgdelete{Finding~\ref{finding_diversity} summarizes the main results on product diversity.
+}As before, when there is no correlation between valuations, product diversity is the same across different recommendation regimes.
+The over-exploitation of information spillovers when $\rho > 0$ leads to product diversity being lowest in the no recommendation regime.
+As a result, this effect gets amplified as $\rho$ increases, which leads to the gap in diversity between the regimes to increase as $\rho$ increases.
+The top row of~\autoref{fig:local_consumption_vary_rho_gamma} shows how diversity varies as $\rho$ increases across the three regimes that we consider.
+There is a similar increasing diversity gap as $\gamma$, or the level of risk-aversion, increases as can be seen in the bottom row of~\autoref{fig:local_consumption_vary_rho_gamma}.
+The mechanisms behind these effects have direct parallels to those discussed in the previous section since low average sequential consumption distance directly leads to low diversity.
+\par
 
-\begin{finding}\label{finding_local_consumption}
-The impact of recommendation on local consumption:
+We now study how recommendation impacts user welfare.
+In our model users make consumption decisions that maximize their current period ex-ante utility that depends on their beliefs in that period.
+Thus, from an ex-ante perspective, they make optimal decisions, but our primary interest is in understanding how the ex-post, or realized, utility varies across regimes and parameter values.
+We define user's \textit{ex-post} welfare as the average of the realized values, controlling for the effect of $T$:
+$$W_i:= \frac{1}{T}\sum_{n \in C_i^T} x_{i,n}$$
+
+\dgedit{\noindent Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which can be seen in the rightmost plot of~\autoref{fig:diversity_welfare_correlation}:}
+\begin{finding}\label{finding_welfare_gap}
+The impact of recommendation on consumer welfare is as follows:
 \begin{enumerate}
-\item When $\rho = 0$, there is no difference in consumption distance between the three recommendation regimes.
-\item When $\rho > 0$, no recommendation induces more local consumption than both recommendation and oracle regimes.
-This effect is amplified as $\rho$ increases as well as when users are more risk averse ($\gamma$ increases)
+\item Under oracle recommendation, welfare is \dgedit{invariant with respect to }$\rho$.
+\item Under no recommendation, welfare is increasing \dgedit{in} $\rho$.
+\item Recommendation introduces welfare gains relative to no recommendation, but these gains are decreasing as $\rho$ increases.
 \end{enumerate}
 \end{finding}
+\par
 
-First, the right panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho = 0$, there is no difference in consumption distance between the three regimes.
-This is due to the fact that when $\rho = 0$, there is no reason that items that are close in the product space should have similar values and so the optimal consumption path does not depend on the similarity of the items.
-However this also means that users do not learn anything about neighboring products and so there is limited path-dependence in consumption.
-Not only is there no difference in the levels between the three regimes, but they all have the same, flat, average consumption distance path.
-This underscores the fact that if there were no correlation between the realized utilities then there would be no reason for users to consume similar items and thus no narrowing effect, regardless of the information on the true utility of the items that users had.
-\par
-The left panel of Figure~\ref{fig:correlation_consumption_path} shows that, when $\rho > 0$, both recommendation and no recommendation lead to increasingly local consumption compared to the oracle benchmark case.
-Further, the average consumption path between periods is \textit{decreasing} for the no recommendation case whereas it is \textit{increasing} for the oracle case.
-The recommendation regime decreases the degree of local consumption, but not as much as the oracle benchmark.
-Due to the correlation of values, the oracle consumption path exploits this and leads to the consumption of more similar items than in the case when $\rho = 0$.
-However, since these spillovers also impact user learning in the no recommendation case, users \textit{over-exploit} this and increasingly consume items similar to high value items that they have consumed before.
-This is further illustrated in the top row of~\autoref{fig:local_consumption_vary_rho_gamma}, which shows how the consumption paths in the oracle and no recommendation regimes vary as $\rho$ increases and is in line with this intuition.
-\par
-This effect is further amplified as the level of risk aversion, $\gamma$, increases.
-The bottom row of~\autoref{fig:local_consumption_vary_rho_gamma} validates that the degree of local consumption drastically increases as $\gamma$ increases.
-This is due to the fact that the spillovers not only affect the mean expected belief about quality but also the degree of uncertainty.
-Local consumption therefore leads to users having less uncertainty about certain areas of the product space and risk aversion may lead them to increasingly consume nearby products.
-
-In sum, filter bubble effects only arise when there is an inherent correlation between the realized utilities of the items in the product space.
-When there is a correlation between the realized utilities then filter bubbles can naturally arise due to the nature of how individuals acquire additional information about the remaining items.
-We have shown that, unless users are provided with additional information to guide their consumption choices, then these information spillovers and user risk-aversion can lead users into filter bubbles.
-When users consume high valued items, they exploit the underlying correlation across different items' values, stronger for similar items, which leads them to increasingly consume products in narrower and narrower portions of the product space.
-Risk aversion may lead users into performing local consumption even when they have a low valuation of nearby items just because they know what to expect from the item.
-Recommendation leads to these effects being mitigated by providing users with additional information on products outside the already explored portions of the product space.
-Further, if all uncertainty was resolved as in the oracle regime, then such behavior is not present.
-
-\begin{figure}[ht]
+\begin{figure}[t]
 \caption{Relationship between $\rho$, $\gamma$ and User Welfare, Diversity}\label{fig:diversity_welfare_correlation}
@@
 \end{figure}
+\addtocounter{figure}{-1}
 
-\addtocounter{figure}{-1}
-\subsection{User Welfare and Product Diversity}
-In this section we primarily focus on the impact of recommendation on user welfare and the overall diversity of the items that they consume.
-While in the previous section we looked at the distance between consecutive items, in this section we focus on a diversity measure that considers the entire consumed set of items.
-The diversity measure we utilize is common in RS literature (e.g.
-\cite{ziegler2005improving}) which is the average normalized pairwise distance between the consumed items:
-$$D_i:=\frac{1}{N}\frac{1}{T(T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$$
+\dgdelete{Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which can be seen in the rightmost plot of~\autoref{fig:diversity_welfare_correlation}.
+}
+The most interesting observation is that the value of recommendation decreases as $\rho$ decreases.
+While welfare in the recommendation regime is flat as we increase $\rho$, it is increasing in the no recommendation regime and thus the welfare gap between the two shrinks as $\rho$ increases.
+The intuition is clear as recommendation provides users with information that allows them to better guide their decisions and increase welfare.
+However, as $\rho$ increases users get increasingly more information from consuming items since the realized utility is now more informative about the utility of nearby items.
+Thus, since consumption decisions themselves reveal information, the information that recommendation reveals provides less value.
+\par 
+One striking observation is that the decrease in diversity does not appear to be associated with a decline in welfare.
+Indeed, it appears that the opposite is the case - that low diversity is associated with higher welfare and vice versa.
+We next explore the relationship between welfare and diversity.
+\dgedit{
+\par
+\noindent Finding~\ref{finding_diversity_welfare_corr} summarizes our findings on the relationship between diversity and welfare:}
 
-\begin{finding}\label{finding_diversity}
-The impact of recommendation on product diversity:
+\begin{finding}\label{finding_diversity_welfare_corr}
+In the no recommendation regime, diversity and welfare are:
 \begin{enumerate}
-\item When $\rho = 0$, product diversity is the same across all three recommendation regimes;
-\item When $\rho > 0$, product diversity decreases across all recommendation regimes but decreases the most in the no recommendation regime.
-This effect is amplified as $\rho$ increases as well as when users become more risk-averse.
+\item Negatively correlated when users have no risk-aversion;
+\item Uncorrelated when users have high levels of risk-aversion.
 \end{enumerate}
+In the recommendation regime, diversity and welfare are:
+\begin{enumerate}
+\item Uncorrelated when users have no risk-aversion;
+\item Positively correlated when users have high levels of risk-aversion.
+\end{enumerate}
+In the oracle regime, diversity and welfare are always uncorrelated.
 \end{finding}
-\par 
-Finding~\ref{finding_diversity} summarizes the main results on product diversity.
-As before, when there is no correlation between valuations, product diversity is the same across different recommendation regimes.
-The over-exploitation of information spillovers when $\rho > 0$ leads to product diversity being lowest in the no recommendation regime.
-As a result, this effect gets amplified as $\rho$ increases, which leads to the gap in diversity between the regimes to increase as $\rho$ increases.
-The top row of~\autoref{fig:local_consumption_vary_rho_gamma} shows how diversity varies as $\rho$ increases across the three regimes that we consider.
-There is a similar increasing diversity gap as $\gamma$, or the level of risk-aversion, increases as can be seen in the bottom row of~\autoref{fig:local_consumption_vary_rho_gamma}.
-The mechanisms behind these effects have direct parallels to those discussed in the previous section since low average sequential consumption distance directly leads to low diversity.
-\par
 
-\begin{figure}[ht]
+\begin{figure}[t]
 \caption{Diversity vs. Welfare}
@@
 \addtocounter{figure}{-1}
-We now study how recommendation impacts user welfare in our model.
-In our model users make consumption decisions that maximize their current period ex-ante utility that depends on their beliefs in that period.
-Thus, from an ex-ante perspective, they make optimal decisions, but our primary interest is in understanding how the ex-post, or realized, utility varies across regimes and parameter values.
-We define user's \textit{ex-post} welfare as the average of the realized values, controlling for the effect of $T$:
-$$W_i:= \frac{1}{T}\sum_{n \in C_i^T} x_{i,n}$$
 
-\begin{finding}\label{finding_welfare_gap}
-The impact of recommendation on consumer welfare is as follows:
-\begin{enumerate}
-\item Under oracle recommendation, welfare is constant as $\rho$ increases.
-\item Under no recommendation, welfare is increasing as $\rho$ increases.
-\item Recommendation introduces welfare gains relative to no recommendation, but these gains are decreasing as $\rho$ increases. 
-\end{enumerate}
-\end{finding}
-\par 
-Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which can be seen in the rightmost plot of~\autoref{fig:diversity_welfare_correlation}.
-The most interesting observation is that the value of recommendation decreases as $\rho$ decreases.
-The intuition is clear as recommendation provides users with information that allows them to better guide their decisions and increase welfare.
-However, as $\rho$ increases users get increasingly more information from consuming items since the realized utility is now more informative about the utility of nearby items.
-Thus, since consumption decisions themselves reveal information, the information that recommendation reveals provides less value.
-\par 
-One striking observation is that the decrease in diversity does not appear to be associated with a decline in welfare.
-Indeed, it appears that the opposite is the case - that low diversity is associated with higher welfare and vice versa.
-We next explore the relationship between welfare and diversity.
-
-\begin{finding}\label{finding_diversity_welfare_corr}
-In the no recommendation regime, diversity and welfare are:
-\begin{enumerate}
-\item Negatively correlated when users have no risk-aversion;
-\item Uncorrelated when users have high levels of risk-aversion.
-\end{enumerate}
-In the recommendation regime, diversity and welfare are:
-\begin{enumerate}
-\item Uncorrelated when users have no risk-aversion;
-\item Positively correlated when users have high levels of risk-aversion.
-\end{enumerate}
-In the oracle regime, diversity and welfare are always uncorrelated.
-\end{finding}
-
-Finding~\ref{finding_diversity_welfare_corr} summarizes our findings on the relationship between diversity and welfare.
-Figure \ref{fig:diversity_welfare_ra} shows how diversity and welfare correlate for the no recommendation case as we vary the degree of risk aversion.
+\dgdelete{Finding~\ref{finding_diversity_welfare_corr} summarizes our findings on the relationship between diversity and welfare.
+}Figure \ref{fig:diversity_welfare_ra} shows how diversity and welfare correlate for the no recommendation case as we vary the degree of risk aversion.
 When there is no risk-aversion then there is a negative correlation between welfare and diversity.
@@
 However, as we saw in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, increasing $\gamma$ can lead to lower diversity and increasingly local consumption due to the fact that the degree of uncertainty now impacts users' choices.
-This weakens the negative relationship between diversity and welfare since both negative and positive experiences with a good reduce uncertainty about surrounding items.
+This weakens the negative relationship between diversity and welfare since both negative and positive experiences with an item reduce uncertainty about surrounding items.
 This leads to the inverted-U shape found in Figure \ref{fig:diversity_welfare_ra} when $\gamma$ is relatively large (e.g.
@@
 \subsection{User Homogenization}
-
-\begin{figure}[ht]
-\caption{Relationship between $\beta$ and Homogeneity, $N = 200$}
-\includegraphics[width=.3\linewidth]{figures/beta_homogeneity_N_200_T_20}
-\includegraphics[width=.3\linewidth]{figures/rho_homogeneity_N_200_T_20}\label{fig:beta_homo}
-\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the weight of the common value component, $\beta$ (left) and correlation between valuations, $\rho$ (right).
-Each line represents this plot for a single recommendation regime.}
-\end{figure}
 
@@
 where $d_J$ denotes the Jaccard index and $H \in [0,1]$.
+\dgedit{\par
+\noindent Finding~\ref{finding_homogeneity} summarizes our findings on the impact of recommendation on user homogeneity:}
+
 \begin{finding}\label{finding_homogeneity}
@@
 \end{finding}
-Finding~\ref{finding_homogeneity} summarizes our findings on the impact of recommendation on user homogeneity.
-First, we study how the degree of homogenization varies as we increase $\beta$, the weight of the common value component.
+
+\begin{figure}[t]
+\caption{Relationship between $\beta$ and Homogeneity, $N = 200$}
+\includegraphics[width=.3\linewidth]{figures/beta_homogeneity_N_200_T_20}
+\includegraphics[width=.3\linewidth]{figures/rho_homogeneity_N_200_T_20}\label{fig:beta_homo}
+\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the weight of the common value component, $\beta$ (left) and correlation between valuations, $\rho$ (right).
+Each line represents this plot for a single recommendation regime.}
+\end{figure}
+
+\dgdelete{Finding~\ref{finding_homogeneity} summarizes our findings on the impact of recommendation on user homogeneity.
+}First, we study how the degree of homogenization varies as we increase $\beta$, the weight of the common value component.
 As $\beta$ increases we expect that users should become increasingly homogeneous as the realized utilities of the items are now increasingly similar.~\autoref{fig:beta_homo} confirms that as the weight of the common-value component $\beta$ increases, users consume increasingly similar sets of items.
@@
 \par
-We next study how the degree of homogeneity varies as $\rho$ increases.
-~\autoref{fig:beta_homo} shows how homogeneity varies as $\rho$ increases.
-The most striking finding is that homogeneity decreases as $\rho$ increases in the recommendation regime.
+
+We next study how the degree of homogeneity varies \dgedit{with} $\rho$\dgdelete{ increases}.
+~\autoref{fig:beta_homo} shows how \dgdelete{homogeneity varies as $\rho$ increases.
+The most striking finding is that }homogeneity decreases as $\rho$ increases in the recommendation regime.
 As was highlighted in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, the degree of local consumption increases with $\rho$.
@@
 The first and most crucial type of data to collect is individual user \textit{beliefs} about items that they have not yet consumed.
-As illustrated by our model, these beliefs are what drive the future consumption decisions of users and understanding these beliefs is crucial for determining the value of recommending certain items.\footnote{Additionally, user beliefs contain information that may not have been observed by the recommender that only observes user choices on the platform.} The second type of data that is critical for RS designers to collect is how user beliefs change over time and, in particular, not just how individuals value the item they just consumed, but also how it impacts their beliefs about the quality of similar items.\footnote{Characterizing the similarity between items has been an important goal of designing content-based recommendations, though as noted by \cite{winecoff2019users}, how users perceive similarity between items is not always in line with how similarity is computed in content-based RS.
+As illustrated by our model, these beliefs are what drive the future consumption decisions of users and understanding these beliefs is crucial for determining the value of recommending certain items.\footnote{Additionally, user beliefs contain information that may not have been observed by the recommender that only observes user choices on the platform.} The second type of data that is \dgedit{relevant} for RS designers to collect is how user beliefs change over time and, in particular, not just how individuals value the item they just consumed, but also how it impacts their beliefs about the quality of similar items.\footnote{Characterizing the similarity between items has been an important goal of designing content-based recommendations, though as noted by \cite{winecoff2019users}, how users perceive similarity between items is not always in line with how similarity is computed in content-based RS.
 Understanding how this impacts which items users update their beliefs about is an important direction for future work.} The third type of data is the risk-aversion levels of users as our model illustrates that the risk preferences of users are important for understanding what information RS can provide that materially leads users to alter their consumption patterns.
@@
 Thus, the prediction problem for the recommender becomes predicting what item would the user choose with no recommendation and, correspondingly, what would be the most useful information to provide to the user that would lead her to consume a \textit{better} item than she would without recommendation.
-This links back to the intuition our model provided for the \textit{John Wick} example whereby collecting user beliefs and measuring how the user updated beliefs about similar goods would lead the recommender to understand that the user would consume \textit{John Wick: Chapter Two}.
+This links back to the intuition our model provided for the \textit{John Wick} example whereby collecting user beliefs and measuring how the user updated beliefs about similar items would lead the recommender to understand that the user would consume \textit{John Wick: Chapter Two}.
 Our approach would therefore imply that, with this as a starting point, the recommender's problem would be to predict what is the most useful information to give the user leading them to change the item that they eventually consume.

### 2020-06-22 paper/filter_bubbles_extended_abstract.tex
add extended abdstract

--- 2020-06-22 paper/filter_bubbles.tex
@@
-\documentclass[manuscript]{acmart}
-\settopmatter{printacmref=false, printccs=false}
+\documentclass{article}
+
+\usepackage{icml2020}
+%\usepackage{lmodern}
+\usepackage{float}
+\usepackage{amssymb}
 \usepackage{enumitem}
-\usepackage{subcaption}
-\newcommand{\xhdr}[1]{\vspace{1mm} \noindent{\bf #1}}
-
 \usepackage{afterpage}
@@
 \addauthor{dg}{red}    % dg for Duarte
-%\addauthor{ga}{blue}  % ga for Guy
-
-% Choose a citation style by commenting/uncommenting the appropriate line:
-%\setcitestyle{authoryear}
-\setcitestyle{acmnumeric}
+%
+\usepackage{amsmath}
+\usepackage{amstext}
+\newcommand{\xhdr}[1]{\vspace{1mm} \noindent{\bf #1}}
+\usepackage{microtype}
+\usepackage{graphicx}
+\usepackage{subfigure}
+\usepackage{booktabs} % for professional tables
 
@@
 
-% Title. Note the optional short title for running heads. In the interest of anonymization, please do not include any acknowledgements.
-\newcommand\PaperTitle{Deconstructing the Filter Bubble: \\User Decision-Making and Recommender Systems}
-\title[\PaperTitle]{\PaperTitle}
+% hyperref makes hyperlinks in the resulting PDF.
+% If your build breaks (sometimes temporarily if a hyperlink spans a page)
+% please comment out the following usepackage line and replace
+% \usepackage{icml2020} with \usepackage[nohyperref]{icml2020} above.
+\usepackage{hyperref}
+\begin{document}
+
+% Rights management information. 
+% This information is sent to you when you complete the rights form.
+% These commands have SAMPLE values in them; it is your responsibility as an author to replace
+% the commands and values with those provided to you when you complete the rights form.
+%
+% These commands are for a PROCEEDINGS abstract or paper.
+
+%
+% These commands are for a JOURNAL article.
+%\setcopyright{acmcopyright}
+%\acmJournal{TOG}
+%\acmYear{2018}\acmVolume{37}\acmNumber{4}\acmArticle{111}\acmMonth{8}
+%\acmDOI{10.1145/1122445.1122456}
+
+%
+% Submission ID. 
+% Use this when submitting an article to a sponsored event. You'll receive a unique submission ID from the organizers
+% of the event, and this ID should be used as the parameter to this command.
+%\acmSubmissionID{123-A56-BU3}
+
+%
+% The majority of ACM publications use numbered citations and references. If you are preparing content for an event
+% sponsored by ACM SIGGRAPH, you must use the "author year" style of citations and references. Uncommenting
+% the next command will enable that style.
+%\citestyle{acmauthoryear}
+
+%
+% end of the preamble, start of the body of the document source.
+\icmltitlerunning{Deconstructing the Filter Bubble}
+
+\twocolumn[
+\icmltitle{Deconstructing the Filter Bubble:  \\User Decision-Making and Recommender Systems}
+
+% It is OKAY to include author information, even for blind
+% submissions: the style file will automatically remove it for you
+% unless you've provided the [accepted] option to the icml2020
+% package.
+
+% List of affiliations: The first argument should be a (short)
+% identifier you will use later to specify author affiliations
+% Academic affiliations should list Department, University, City, Region, Country
+% Industry affiliations should list Company, City, Region, Country
+
+% You can specify symbols, otherwise they are numbered in order.
+% Ideally, you should not use this facility. Affiliations will be numbered
+% in order of appearance and this is the preferred way.
+]
 
@@
 We study a model of user decision-making in the context of recommender systems via numerical simulation.
-Our model provides an explanation for the findings of Nguyen, et.
-al (2014), where users consume increasingly similar items over time even without recommendation.
-Our model naturally generates such filter\dgedit{-}bubble effects since users face a sequential decision making problem under uncertainty and there are informational spillovers where a user consuming an item and learning how she values it reduces uncertainty and provides information about similar items.
-\par
-In line with Nguyen, et.
-al (2014), recommendation alleviates these filter\dgedit{-}bubble effects.
-However, we find that it also leads to an increase in homogeneity across users, resulting in a trade-off between homogenizing across user consumption and diversifying
-within user consumption.
-Finally, our model highlights the importance of collecting data on user beliefs and their evolution over time both to design better recommendations and to further understand their impact.
+We show that this model provides an explanation for the findings of Nguyen, et.
+al (2014), where, in environments where recommender systems are typically deployed, users consume increasingly similar items over time even without recommendation.
+We find that recommendation alleviates these natural filter-bubble effects, but that it also leads to an increase in homogeneity across users, resulting in a trade-off between homogenizing across user consumption and diversifying within user consumption.
+Finally, we discuss how our model highlights the importance of collecting data on user beliefs and their evolution over time both to design better recommendations and to further understand their impact.
 \end{abstract}
 
-\begin{document}
-
-\maketitle
-
+\textit{Note: This is an extended abstract of the paper - the full version is attached as an appendix and contains the omitted technical details and full description of the model and the exercises.}
 % Paper body
@@
 The closest set of papers pointed out in this survey are those related to preference construction \cite{bettman1998constructive, lichtenstein2006construction} whereby users develop preferences over time through the context of a decision process.
-We point out that the true underlying preferences of users may be stable over time, but, due to the nature of items in contexts where RS are deployed, they have incomplete information of their valuations and both consumption and recommendation provide them with information to reduce their uncertainty.
-Thus, the primary insight of our paper is that user beliefs and how users update their beliefs about similar items after consumption are important and previously unconsidered elements of human decision making that are critical for understanding the design and consequences of RS.
+The primary insight of our paper is that user beliefs and how users update their beliefs about similar items after consumption are important and previously unconsidered elements of human decision making that are critical for understanding the design and consequences of RS.
 Within this literature, \cite{celma2008new, cremonesi2013user, pu2011user} focus on ``user-centric" approaches to recommendation whereby user evaluation of the usefulness of recommendation is a key evaluation measure.
 Our evaluation measure is similar, but, unlike previous approaches, emphasizes the importance of user beliefs.
-Finally, \cite{hodgson2019horse} considers a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of their preferences in the environment, but consider it in the context of search for a single item.
 
-\section{Our Model and Preliminaries}
-\subsection{Preliminaries on Expected Utility Theory}
-\noindent For every item $n$ in the product space $\mathcal J$, we assume that each user $i$ assigns a monetary equivalent $x_{i,n} \in \mathbb R$ to the experience of consuming it.
-Each user can value the same item differently.
-However, we assume that users have the same utility over money, given by a utility function $u: \mathbb R \to \mathbb R$, strictly increasing and continuous.
-So, ex-post, the value of item $n$ for user $i$ is given by $u(x_{i,n})$\dgedit{.
-B}efore consuming the item, the user does not know exactly how she will value it.
-In particular, even users that will end up having the same ex-post valuation of item $n$ may differ in their ex-ante valuation because they hold different beliefs about it.
-We denote by $p_{i}$ the beliefs user $i$ has about how she will value each of the items in the product space.
-Note that this implies that consuming item $n$ is the same as taking a gamble.
-Each user evaluates the item according to the expected utility associated with the item, i.e.
-$U_i(n)=\mathbb E_{p_i}[u(x_n)]$.
-\par
-Risk aversion captures how different users react to the risk associated to a particular consumption opportunity.
-It is formalized as follows: a given gamble $x$ takes real values and follows distribution $p$.
-Then, for every gamble $x$, there is a certain amount of money that makes the user indifferent between taking the gamble or taking the sure amount of money.
-This sure amount of money is called the \textit{certainty equivalent} of gamble $x$ and is denoted as $\delta(x)$.
-A user $i$ is more risk-averse than another user $j$ \dgedit{if whenever user $j$ prefers a sure thing to the gamble, then user $i$ does too}\dgdelete{when for every gamble $x$, user $i$ requires a lower value to give up the gamble when compared to user $j$, i.e.
-$\delta_i(x)$}.
-Therefore, a more risk-averse user is more willing to avoid the risk of taking the gamble.
-We assume that the utility function takes a flexible functional form $u(x)=1-\exp(-\gamma x)$ for $\gamma\ne0$ and $u(x)=x$ for $\gamma\dgedit{\to}0$ -- known as constant absolute risk-aversion preferences (from hereon CARA).
-Higher $\gamma$ implies higher risk-aversion with $\gamma\dgedit{\to}0$ corresponding to the risk-neutral case and $\gamma>0$ to the risk-averse one.
-Our formulations here follow standard economic decision-theory (see \cite{mas1995microeconomic} for a textbook treatment of these topics).
-\par
-
-\subsection{Model}
-\par
-\xhdr{Users.} We consider a set of users $I$ where each user $i \in I$ faces the same finite set of $N$ items $\mathcal J = \left\{0,1,...,N-1\right\}$.
-For simplicity, we assume that users only derive pleasure from item $n \in \mathcal{J}$ the first time they consume it.
-\par
-
-We denote by $x_{i,n}$ user $i$'s realized value from consuming item $n$.
-In particular, we consider that the realized value derived from a given item can be decomposed in the following manner: $x_{i,n}= v_{i,n} + \beta v_n$, where $v_{i,n}$ denotes an idiosyncratic component -- i.e.
-user $i$'s idiosyncratic taste for product $n$ --  and $v_{n}$, a common-value component.
-One can interpret $v_n$ as a measure of how much item $n$ is valued in society in general and, in a sense, $v_{i,n}$ denotes how $i$ diverges from this overall ranking.
-The scalar $\beta \in \mathbb{R}_{+}$ denotes the degree to which valuations are idiosyncratic to each user or common across users.
-If $\beta=0$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta$ is large, every individual has similar preferences.
-\par
-Stacking values in vector-form, we get the vector of values associated with each item 
-$${\left(x_{i,n}\right)}_{n \in \mathcal{J}}=:X_i =V_i+ \beta V $$
-where $V_i ={\left(v_{i,n}\right)}_{n \in \mathcal{J}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{J}}$.
-\par
-\xhdr{User Decision-Making.}
-We assume the user makes $T$ choices and therefore can only consume up to $T$ items, where $T$ is a small fraction of $N$.
-This captures the idea that users are faced with an immense choice set, but that ultimately they end up experiencing (and learning) about just a small fraction of it.
-For tractability, we impose that users are myopic and every period consume the item that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
-\par
-\xhdr{User Beliefs.} We assume that all the true realized values are drawn at $t = 0$. However, users do not know the realized values before consuming an item, but rather have beliefs over them.
-Formally, user $i$ starts with some beliefs about $X_i$, namely that the idiosyncratic and common-value parts of the valuations are independent -- $V_i \perp \!\!\!
-\perp V$ -- and that each is multivariate normal
-\begin{enumerate}[topsep=0pt]
-\item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ 
-\item $V \sim \mathcal N(\overline V, \Sigma)$ with $\overline V =0$
-\end{enumerate}
-We impose the normality assumption for two reasons.
-The first is that this allows for simple and tractable belief updating.
-The second is that it allows us to incorporate an easily interpretable correlation structure between the items.
-The precise formulation of $\Sigma$ and $\Sigma_i$ that we consider is defined below when we discuss user learning.
-\par
-\dgedit{Recalling} that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{i,n}\right)_{n \in \mathcal{J}}$ are drawn independently from a jointly normal distribution, where $\overline v_{i,n} \sim \mathcal N (0, \overline \sigma^2)$ are independent and identically distributed.
-These $\overline v_{i,n}$ denote the prior belief that individual $i$ holds about her valuation over item $n$.
-As people are exposed to different backgrounds, their beliefs about how much they value a given item also varies and $\overline v_{i,n}$ denotes this idiosyncrasy at the level of prior beliefs.
-\par
-
-We assume users are expected utility maximizers.
-User $i$'s certainty equivalent for item $n$, the sure value that makes user $i$ indifferent between it and consuming the item $n$, conditional on the consumption history, is given by
-$\delta_{i}(n)\mid C_i^{t-1}=\mu_n-\frac{1}{2}\gamma \Sigma_{nn}$, where $\mu_n$ and $\Sigma_{nn}$ are the expected value and variance for item $n$ that the user has given their initial beliefs and consumption history up until time $t$.
-Note that this expression is known to be the certainty equivalent for CARA preferences in a Gaussian environment \cite{mas1995microeconomic}.
-
-\xhdr{User Learning\dgedit{.}}
-When a user consumes an item $n$ she learns the realized value for that item.
-We \dgedit{consider the case where}  learning about the value of item $n$ reveals more about the value associated to items that are closer to it, which captures the idea that trying an item provides more information about similar items than about dissimilar ones.\footnote{\cite{schulz2019structured} empirically studies how individuals solve sequential decision-making problems under uncertainty in large choice sets in the context of mobile food delivery orders.
-They find that individuals engage in similarity-based generalizations where learning about the realized value of a particular item provides them with information about similar items.
-We incorporate this finding into our model in a stylized manner.} In order to have a well-defined notion of similarity we need to define a distance function between items, which we define as $d(n,m):=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}$ where $m$ and $n$ are indices of items in $\mathcal{J}$.
-This distance function is not intended to model the intricacies of a realistic product space, but instead to provide a stylized product space to help us understand the effects of informational spillovers on user behavior.
-The basic intuition, in the context of movie consumption, is that a \dgedit{user's valuation of} \textit{John Wick} (item $n$) provides more information about how much she will like \textit{John Wick: Chapter Two} (item $m$) than \textit{Titanic} (item $q$) since $d(n, m) < d(n, q)$.
-\par
-We consider that the entry of $n$-th row and the ($m$)-th column of $\Sigma_i$ is given by $\sigma_i^2 \rho^{d(n,m)}$, and that of $\Sigma$ is given by $\sigma^2 \rho^{d(n,m)}$.
-The scalar $\rho \in [0,1]$ therefore impacts the covariance structure: a higher $\rho$ implies that learning the utility of $n$ is more informative about products nearby.
-Informativeness, for any $\rho \in (0,1)$, is decreasing in distance.
-The particular distance function that we utilize leads to this covariance structure being simple, where the $(n,n+1)$-th entry in the covariance matrix is $\sigma^{2} \rho$, the $(n,n+2)$-th entry is $\sigma^{2} \rho^2$, etc.\footnote{This exponential decay correlation structure can be related to the tenet of case-based similarity of \cite{gilboa1995case} -- see \cite{billot2008axiomatization} for an axiomatization of exponential similarity.}
-\par
-The precise updating rule is as follows.
-Recall that at time $t$ the user's consumption history is given by $C_{i}^{t}$ and we denote the utility realizations of these items as $c_t$.
-We denote $\mu_t$ as the initial mean beliefs the user has over the items in $C_{i}^{t}$ and $\mu_{N-t}$ as the initial mean beliefs the user has over the remaining $N-t$ items, $\mathcal{J} \setminus C_{i}^{t}$.
-We partition the covariance matrix as follows:
-\[ \Sigma =  \left( \begin{array}{cc}
-\Sigma_{(N-t, N-t)} & \Sigma_{(N-t,t)} \\
-\Sigma_{(t,N-t)} & \Sigma_{(t,t)}
-\end{array} \right)
-\]
-Thus, after consuming the items in $C_{i}^{t}$ the resulting beliefs over the remaining items are given by $\mathcal{N}(\bar{\mu}, \bar{\Sigma})$ where $\bar{\mu}$ and $\bar{\Sigma}$ are as follows:
-\begin{align*}
-\bar{\mu} \mid c_t = \mu_{N-t} + \Sigma_{(N-t,t)} \Sigma_{(t,t)}^{-1}(c_t - \mu_t) \\
-\bar{\Sigma} \mid c_t = \Sigma_{(N-t,N-t)} - \Sigma_{(N-t,t)} \Sigma_{(t,t)}^{-1} \Sigma_{(t,N-t)}
-\end{align*}
-
-\xhdr{An Illustrative Example\dgedit{.}} We illustrate the main intuitions of our model with a simple example.
+\section{Model Illustration}
+We illustrate the main intuitions of our model with a simple example.
 Suppose that there are four items: 0, 1, 2, 3.
 The items are in different places of the product space, where 0 is close to 1 and 3 but more distant from 2.
-\dgedit{For the sake of expositional clarity, suppose that t}he initial beliefs are given as follows:
+\dgedit{For the sake of expositional clarity, suppose that t}he initial beliefs are given as follows where $\rho$ denotes the correlation between the utilities of the items, $\mu$ is the mean belief, and $\Sigma$ is the covariance matrix:\footnote{See section 2.2 of the full paper in the appendix for more details about the model}
 \[ \mu = \left (\begin{array}{c}
-\mathbb{E}[x_0] \\
-\mathbb{E}[x_1] \\
-\mathbb{E}[x_2] \\
-\mathbb{E}[x_3]
-\end{array}  \right) = \left (\begin{array}{c}
 0 \\
@@
 0
-\end{array}  \right), \dgedit{\qquad}\Sigma =  \sigma^{2} \left( \begin{array}{cccc}
+\end{array}  \right), \\ 
+\dgedit{\qquad}\Sigma =  \sigma^{2} \left( \begin{array}{cccc}
 \rho^{0} & \rho^{1} & \rho^{2} & \rho^{0} \\
@@
 In period 1, every item is ex-ante identical since they have the same mean and variance and so suppose that the user breaks the tie arbitrarily and consumes item 0.
-The underlying correlation structure implies that upon observing that $x_0 = y$ the user will update beliefs about the remaining 3 items according to the previously specified updating rule.
+The underlying correlation structure implies that upon observing that $x_0 = y$ the user will update beliefs about the remaining 3 items.
 For concreteness, we suppose that $\sigma = 1$ and $\rho = 0.5$, but the intuitions hold for any value of $\sigma$ and $\rho > 0$.
@@
 
-\xhdr{Recommendation.}
-Our model of recommendation is stylized in order to provide qualitative insights into how recommendation shapes behavior, instead of focusing on the details of how RS are implemented in practice.
-We model recommendation as giving users information about the valuation of the products.
-\par
-
-We will consider three cases.
-The case of primary interest is \textit{recommendation} where the recommender observes values accrued and knows $V$ but does not know $V_i$.\footnote{We do not consider the acquisition of information for the recommender to know $V$ and suppose that she has sufficient data to learn $V$ with arbitrary precision at $t = 0$.} However, the recommender does know the users' beliefs $\bar V_i$.
-Thus, at any given period, the recommender provides a personalized recommendation that combines the knowledge of the common value component $V$ with the user beliefs $\bar V_i$.
-Knowing the user's beliefs about her valuation of each item become crucial in this case: just providing the user with information about $V$ may change the user's original ranking, but, without considering the user's beliefs, she will not necessarily follow the recommendation.\footnote{The notion of recommendation that we consider is idealized where the recommendation does the Bayesian updating for users, but the results are equivalent to if the users did the updating themselves.}
-\par
-
-We further consider two cases that serve \dgedit{mainly} as benchmarks.
-The first is \textit{no recommendation}, where users get no additional information and make consumption choices based on their beliefs and consumption history.
-This gives us a benchmark as to how users would behave \textit{without} recommendation so that we can analyze what changes with the introduction of recommendation.
-The second is the \textit{oracle recommendation} where the recommender knows the true realized utility of each item for each user and can therefore recommend the best remaining item in every period.
-This gives us a full information benchmark, which is the optimal consumption path for a user if all uncertainty about their preferences was resolved.
-Comparison to the oracle regime benchmark provides an analog to the standard regret measures utilized in the multi-armed bandit literature, which look at the difference in the expected value of the ex-post optimal action and the expected value of actions that were taken.
-\par
-
-\xhdr{Simulation Details.}
-We analyze our model using numerical simulation since the sequential decision-making component paired with the rich covariance structure between the items make it difficult to characterize optimal user behavior analytically.
-The Gaussian assumption allows for closed form belief updating which enables us to simulate our model but does not provide much help in analytical characterizations.
-\par
-
-We explore how consumption patterns differ as we consider different recommendation regimes and report representative results from our simulations.
-We run this simulation over 100 populations of users with 100 users per population for each combination of parameters.
-%For every population and every statistic we calculate, we average over the users in the population to get a representative statistic for this population. 
-A given set of parameters and a user are a single data point in our dataset.
-\par
-
-We simulate over risk-aversion parameters $\gamma \in \{ 0, 0.3, 0.6, 1, 5 \}$, \ standard-deviation $\sigma \in \{ 0.25, 0.5, 1.0, 2.0, 4.0 \}$, correlation parameters $\rho\in \{ 0, 0.1, 0.3, 0.5, 0.7, 0.9 \} $ and degree of idiosyncrasy of individual valuations $\beta \in \{ 0, 0.4, 0.8, 1, 2, 5\}$.
-The range of considered parameter values cover the relevant portions of the parameter space in order to provide full insight into the behavior of the model.
-When we consider results varying a single parameter, we group the results over the other parameters and provide reports varying only the parameter of interest.
-We report results for a relatively small consumption history $T=20$ with a product space size $N=200$.\footnote{Due to space constraints, we only report the values for $N = 200$.
-However, we further conducted the same exercises for $N = 100$ and $N = 500$ and the results are qualitatively similar to those reported here.}
-\section{Results}
-\subsection{Local Consumption and Filter Bubbles}
-We characterize ``filter bubble'' effects as the degree to which users engage in \textit{local consumption}.
-We define local consumption in terms of the average consumption distance between the items consumed by the users at time $t-1$ and $t$.
-Thus, in the context of our model, filter\dgedit{-}bubble effects arise when the average consumption distance decreases over time and, across regimes, when the levels are lower for a given recommendation regime compared to another.
-
-\dgedit{Our first finding can be summarized as follows:}
-
-\begin{finding}\label{finding_local_consumption}
-The impact of recommendation on local consumption:
-\begin{enumerate}
-\item When $\rho = 0$, there is no difference in consumption distance between the three recommendation regimes.
-\item When $\rho > 0$, no recommendation induces more local consumption than both recommendation and oracle regimes.
-This effect is amplified as $\rho$ increases as well as when users are more risk averse ($\gamma$ increases).
-\end{enumerate}
-\end{finding}
-
-\afterpage{
-\begin{figure}[t]
-\caption{Local Consumption and Correlation, $N = 200$}
-\begin{subfigure}{.3\linewidth}
-  \centering
-  \includegraphics[width=1.0\linewidth]{figures/rho_pos_consumption_dist_N_200T_20_overall.pdf}
-  \label{fig:sfig1}
-\end{subfigure}%
-\begin{subfigure}{.3\linewidth}
-  \centering
-  \includegraphics[width=1.0\linewidth]{figures/rho_zero_consumption_dist_N_200T_20_overall.pdf}
-  \label{fig:sfig2}
-\end{subfigure}
-\caption*{\scriptsize Notes: The figure shows the consecutive consumption path difference between the no recommendation, recommendation, and oracle regime.
-The figure on the left displays the average consecutive consumption distance aggregating over simulations where $\rho > 0$ and the figure on the right displays the average consecutive consumption distance aggregating over simulations where $\rho = 0$.}
-\label{fig:correlation_consumption_path}
-\end{figure}
-\addtocounter{figure}{-1}}
-
-First, the right panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho = 0$, there is no difference in consumption distance between the three regimes.
-This is due to the fact that when $\rho = 0$, there is no reason that items that are close in the product space should have similar values and so the optimal consumption path does not depend on the similarity of the items.
-However this also means that users do not learn anything about neighboring products and so there is limited path-dependence in consumption.
-Not only is there no difference in the levels between the three regimes, but they all have the same, flat, average consumption distance path.
-This underscores the fact that if there were no correlation between the realized utilities then there would be no reason for users to consume similar items and thus no narrowing effect, regardless of the information on the true utility of the items that users had.
-\par
-The left panel of Figure~\ref{fig:correlation_consumption_path} shows that, when $\rho > 0$, both recommendation and no recommendation lead to increasingly local consumption compared to the oracle benchmark case.
-Further, the average consumption path between periods is \textit{decreasing} for the no recommendation case whereas it is \textit{increasing} for the oracle case.
-The recommendation regime decreases the degree of local consumption, but not as much as the oracle benchmark.
-Due to the correlation of values, the oracle consumption path exploits this and leads to the consumption of more similar items than in the case when $\rho = 0$.
-However, since these spillovers also impact user learning in the no recommendation case, users \textit{over-exploit} this and increasingly consume items similar to high value items that they have consumed before.
-This is further illustrated in the top row of~\autoref{fig:local_consumption_vary_rho_gamma}, which shows how the consumption paths in the oracle and no recommendation regimes vary as $\rho$ increases and is in line with this intuition.
-\par
-This effect is further amplified as the level of risk aversion, $\gamma$, increases.
-The bottom row of~\autoref{fig:local_consumption_vary_rho_gamma} vaFurtherlidates that the degree of local consumption drastically increases as $\gamma$ increases.
-This is due to the fact that the spillovers not only impact the mean expected belief about quality but also the degree of uncertainty.
-Local consumption therefore leads to users having less uncertainty about certain areas of the product space and risk aversion may lead them to increasingly consume nearby products.
-
-In sum, filter\dgedit{-}bubble effects only arise when there is an inherent correlation between the realized utilities of the items in the product space.
-When there is a correlation between the realized utilities then filter bubbles can naturally arise due to the nature of how individuals acquire additional information about the remaining items.
-We have shown that, unless users are provided with additional information to guide their consumption choices, then these information spillovers and user risk-aversion can lead users into filter bubbles.
-When users consume high valued items, they exploit the underlying correlation across different items' values, stronger for similar items, which leads them to increasingly consume products in narrower and narrower portions of the product space.
-Risk aversion may lead users into performing local consumption even when they have a low valuation of nearby items just because they know what to expect from the item.
-Recommendation leads to these effects being mitigated by providing users with additional information on products outside the already explored portions of the product space.
-Furthermore, if all uncertainty was resolved as in the oracle regime, then such behavior is not present.
-
-\subsection{User Welfare and Product Diversity}
-
-In this section we primarily focus on the impact of recommendation on user welfare and the overall diversity of the items that they consume.
-While in the previous section we looked at the distance between consecutive items, in this section we focus on a diversity measure that considers the entire consumed set of items.
-The diversity measure we utilize is common in RS literature (e.g.
-\cite{ziegler2005improving}) which is the average normalized pairwise distance between the consumed items:
-$$D_i:=\frac{1}{N}\frac{1}{T(T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$$
-\dgedit{\noindent 
-Finding~\ref{finding_diversity} summarizes the main results on product diversity:}
-\begin{finding}\label{finding_diversity}
-The impact of recommendation on product diversity:
-\begin{enumerate}
-\item When $\rho = 0$, product diversity is the same across all three recommendation regimes;
-\item When $\rho > 0$, product diversity decreases across all recommendation regimes but decreases the most in the no recommendation regime.
-This effect is amplified as $\rho$ increases as well as when users become more risk-averse.
-\end{enumerate}
-\end{finding}
-\par
-
-\begin{figure}[t]
-\caption{Relationship between Local Consumption and $\rho$, $\gamma$}
-\begin{subfigure}{.3\textwidth}
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20.pdf}
-\end{subfigure}
-\begin{subfigure}{.3\textwidth}
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_partial.pdf}
-\end{subfigure}
-\begin{subfigure}{.3\textwidth}
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_omni.pdf}\\
-\end{subfigure}
-\begin{subfigure}{.3\textwidth}
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20.pdf}
-\end{subfigure}
-\begin{subfigure}{.3\textwidth}
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_partial.pdf}
-\end{subfigure}
-\begin{subfigure}{.3\textwidth}
-\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_omni.pdf}
-\end{subfigure}%
-\caption*{\scriptsize Notes: Each figure plots the average consecutive consumption distance across time as the inherent correlation between the valuation of the items, $\rho$, varies (top row) and the level of risk-aversion, $\gamma$, varies (bottom row).
-In both rows, the left displays the no recommendation regime, the center displays the recommendation regime, and the right displays the oracle regime.}
-\label{fig:local_consumption_vary_rho_gamma}
-\end{figure}
-\addtocounter{figure}{-1}
-\dgdelete{Finding~\ref{finding_diversity} summarizes the main results on product diversity.
-}As before, when there is no correlation between valuations, product diversity is the same across different recommendation regimes.
-The over-exploitation of information spillovers when $\rho > 0$ leads to product diversity being lowest in the no recommendation regime.
-As a result, this effect gets amplified as $\rho$ increases, which leads to the gap in diversity between the regimes to increase as $\rho$ increases.
-The top row of~\autoref{fig:local_consumption_vary_rho_gamma} shows how diversity varies as $\rho$ increases across the three regimes that we consider.
-There is a similar increasing diversity gap as $\gamma$, or the level of risk-aversion, increases as can be seen in the bottom row of~\autoref{fig:local_consumption_vary_rho_gamma}.
-The mechanisms behind these effects have direct parallels to those discussed in the previous section since low average sequential consumption distance directly leads to low diversity.
-\par
-
-We now study how recommendation impacts user welfare.
-In our model users make consumption decisions that maximize their current period ex-ante utility that depends on their beliefs in that period.
-Thus, from an ex-ante perspective, they make optimal decisions, but our primary interest is in understanding how the ex-post, or realized, utility varies across regimes and parameter values.
-We define user's \textit{ex-post} welfare as the average of the realized values, controlling for the effect of $T$:
-$$W_i:= \frac{1}{T}\sum_{n \in C_i^T} x_{i,n}$$
-
-\dgedit{\noindent Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which can be seen in the rightmost plot of~\autoref{fig:diversity_welfare_correlation}:}
-\begin{finding}\label{finding_welfare_gap}
-The impact of recommendation on consumer welfare is as follows:
-\begin{enumerate}
-\item Under oracle recommendation, welfare is \dgedit{invariant with respect to }$\rho$.
-\item Under no recommendation, welfare is increasing \dgedit{in} $\rho$.
-\item Recommendation introduces welfare gains relative to no recommendation, but these gains are decreasing as $\rho$ increases.
-\end{enumerate}
-\end{finding}
-\par
-
-\begin{figure}[t]
-\caption{Relationship between $\rho$, $\gamma$ and User Welfare, Diversity}\label{fig:diversity_welfare_correlation}
-\begin{subfigure}{0.3\linewidth}
-  \includegraphics[width=1.0\linewidth]{figures/rho_diversity_N_200_T_20.pdf}
-\end{subfigure}
-\begin{subfigure}{0.3\linewidth}
-\includegraphics[width=1.0\linewidth]{figures/gamma_diversity_N_200_T_20}
-\end{subfigure}
-\begin{subfigure}{0.3\linewidth}
-  \includegraphics[width=1.0\linewidth]{figures/rho_welfare_N_200_T_20.pdf}
-\end{subfigure}
-\caption*{\scriptsize Notes: The figures on the left and center display the relationship between $\rho$ and overall consumed product diversity (left) as well as $\gamma$ and overall consumed product diversity (center).
-The figure on the right displays the relationship between $\rho$ and overall welfare.}
-\end{figure}
-\addtocounter{figure}{-1}
-
-\dgdelete{Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which can be seen in the rightmost plot of~\autoref{fig:diversity_welfare_correlation}.
-}
-The most interesting observation is that the value of recommendation decreases as $\rho$ decreases.
-While welfare in the recommendation regime is flat as we increase $\rho$, it is increasing in the no recommendation regime and thus the welfare gap between the two shrinks as $\rho$ increases.
-The intuition is clear as recommendation provides users with information that allows them to better guide their decisions and increase welfare.
-However, as $\rho$ increases users get increasingly more information from consuming items since the realized utility is now more informative about the utility of nearby items.
-Thus, since consumption decisions themselves reveal information, the information that recommendation reveals provides less value.
-\par 
-One striking observation is that the decrease in diversity does not appear to be associated with a decline in welfare.
-Indeed, it appears that the opposite is the case - that low diversity is associated with higher welfare and vice versa.
-We next explore the relationship between welfare and diversity.
-\dgedit{
-\par
-\noindent Finding~\ref{finding_diversity_welfare_corr} summarizes our findings on the relationship between diversity and welfare:}
-
-\begin{finding}\label{finding_diversity_welfare_corr}
-In the no recommendation regime, diversity and welfare are:
-\begin{enumerate}
-\item Negatively correlated when users have no risk-aversion;
-\item Uncorrelated when users have high levels of risk-aversion.
-\end{enumerate}
-In the recommendation regime, diversity and welfare are:
-\begin{enumerate}
-\item Uncorrelated when users have no risk-aversion;
-\item Positively correlated when users have high levels of risk-aversion.
-\end{enumerate}
-In the oracle regime, diversity and welfare are always uncorrelated.
-\end{finding}
-
-\begin{figure}[t]
-\caption{Diversity vs. Welfare}
-\begin{subfigure}{.3\textwidth}
-\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_n_200.pdf}
-\end{subfigure}
-\begin{subfigure}{.3\textwidth}
-\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_n_200.pdf}
-\end{subfigure}\\
-\begin{subfigure}{.3\textwidth}
-\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_partial_n_200.pdf}
-\end{subfigure}
-\begin{subfigure}{.3\textwidth}
-\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_partial_n_200.pdf}
-\end{subfigure}\\
-\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under the recommendation (top row)  and no recommendation (bottom row) regime.
-In both rows, the figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.}\label{fig:diversity_welfare_ra}
-\end{figure}
-\addtocounter{figure}{-1}
-
-\dgdelete{Finding~\ref{finding_diversity_welfare_corr} summarizes our findings on the relationship between diversity and welfare.
-}Figure \ref{fig:diversity_welfare_ra} shows how diversity and welfare correlate for the no recommendation case as we vary the degree of risk aversion.
-When there is no risk-aversion then there is a negative correlation between welfare and diversity.
-This is since, with no risk-aversion, a user will select the item that she currently believes has the highest expected value.
-High product diversity in this case can arise from a user who consumes an item she disliked and updates her beliefs about nearby items negatively.
-As a result, in the following period she will pick an item far away in the product space from the item that was previously consumed.
-If instead the user valued highly the item that she had consumed, then she is more likely to pick a nearby item.
-The information spillovers therefore lead to high product diversity being negatively correlated with welfare.
-\par
-This only happens since $\gamma = 0$ leads to users only caring about the expected value of the item.
-However, as we saw in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, increasing $\gamma$ can lead to lower diversity and increasingly local consumption due to the fact that the degree of uncertainty now impacts users' choices.
-This weakens the negative relationship between diversity and welfare since both negative and positive experiences with an item reduce uncertainty about surrounding items.
-This leads to the inverted-U shape found in Figure \ref{fig:diversity_welfare_ra} when $\gamma$ is relatively large (e.g.
-$\gamma = 5$) though diversity and welfare are virtually uncorrelated in the data.
-In the recommendation and oracle regimes, under risk neutrality ($\gamma=0$), welfare and diversity are uncorrelated, while under risk aversion ($\gamma=5$), it is possible to observe an actual positive relation between diversity and welfare as recommendations are able to reduce uncertainty and facilitate exploration of the product space.
-
-\subsection{User Homogenization}
-
-In this section, we focus on comparisons across users and investigate how the consumed set of items across users varies across different recommendation regimes and parameter values.
-In particular we look at the degree of \textit{homogenization} between users.
-Similar to other papers that study the degree of homogenization in RS (e.g.
-\cite{chaney2018algorithmic}) we measure homogeneity via the Jaccard index between the consumption sets of users:
-\begin{align*}
-H:=\frac{1}{|I|(|I|-1)}\sum_{i,j \in I: i \ne j}d_J(C_i^T,C_j^T)
-\end{align*}
-where $d_J$ denotes the Jaccard index and $H \in [0,1]$.
-\dgedit{\par
-\noindent Finding~\ref{finding_homogeneity} summarizes our findings on the impact of recommendation on user homogeneity:}
-
-\begin{finding}\label{finding_homogeneity}
-The impact of recommendation on homogeneity is as follows:
-\begin{enumerate}
-\item Highest under recommendation and lowest under no recommendation;
-\item Increasing in $\beta$, or the weight of the common-value component;
-\item Decreasing in $\rho$ for partial recommendation, but weakly increasing in $\rho$ for no recommendation.
-\end{enumerate}
-\end{finding}
-
-\begin{figure}[t]
-\caption{Relationship between $\beta$ and Homogeneity, $N = 200$}
-\includegraphics[width=.3\linewidth]{figures/beta_homogeneity_N_200_T_20}
-\includegraphics[width=.3\linewidth]{figures/rho_homogeneity_N_200_T_20}\label{fig:beta_homo}
-\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the weight of the common value component, $\beta$ (left) and correlation between valuations, $\rho$ (right).
-Each line represents this plot for a single recommendation regime.}
-\end{figure}
-
-\dgdelete{Finding~\ref{finding_homogeneity} summarizes our findings on the impact of recommendation on user homogeneity.
-}First, we study how the degree of homogenization varies as we increase $\beta$, the weight of the common value component.
-As $\beta$ increases we expect that users should become increasingly homogeneous as the realized utilities of the items are now increasingly similar.~\autoref{fig:beta_homo} confirms that as the weight of the common-value component $\beta$ increases, users consume increasingly similar sets of items.
-The homogenization effect is strongest under the recommendation regime since the revelation of the common-value component induces users to consume products in similar areas of the product space.
-As $\beta$ increases, some amount of homogenization is optimal as can be seen from the oracle case.
-However, since users in the no recommendation regime do not know the common-value component they engage in local consumption in different areas of the product space which leads to less than optimal homogeneity.
-\par
-
-We next study how the degree of homogeneity varies \dgedit{with} $\rho$\dgdelete{ increases}.
-~\autoref{fig:beta_homo} shows how \dgdelete{homogeneity varies as $\rho$ increases.
-The most striking finding is that }homogeneity decreases as $\rho$ increases in the recommendation regime.
-As was highlighted in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, the degree of local consumption increases with $\rho$.
-Even though the revelation of the common-value component induces them to search in similar parts of the product space, their idiosyncratic components induce them to consume products in a more localized area of the product space as $\rho$ increases which leads to a decline in homogeneity.
-
-\section{Recommender System Evaluation}
+\section{Recommender System Design}
 In this section we discuss how the insights from our model of user decision-making can inform the evaluation and design of recommender systems.
@@
 The first and most crucial type of data to collect is individual user \textit{beliefs} about items that they have not yet consumed.
-As illustrated by our model, these beliefs are what drive the future consumption decisions of users and understanding these beliefs is crucial for determining the value of recommending certain items.\footnote{Additionally, user beliefs contain information that may not have been observed by the recommender that only observes user choices on the platform.} The second type of data that is \dgedit{relevant} for RS designers to collect is how user beliefs change over time and, in particular, not just how individuals value the item they just consumed, but also how it impacts their beliefs about the quality of similar items.\footnote{Characterizing the similarity between items has been an important goal of designing content-based recommendations, though as noted by \cite{winecoff2019users}, how users perceive similarity between items is not always in line with how similarity is computed in content-based RS.
-Understanding how this impacts which items users update their beliefs about is an important direction for future work.} The third type of data is the risk-aversion levels of users as our model illustrates that the risk preferences of users are important for understanding what information RS can provide that materially leads users to alter their consumption patterns.
+As illustrated by our model, these beliefs are what drive the future consumption decisions of users and understanding these beliefs is crucial for determining the value of recommending certain items.\footnote{Additionally, user beliefs contain information that may not have been observed by the recommender that only observes user choices on the platform.} The second type of data that is \dgedit{relevant} for RS designers to collect is how user beliefs change over time and, in particular, not just how individuals value the item they just consumed, but also how it impacts their beliefs about the quality of similar items.
+The third type of data is the risk-aversion levels of users as our model illustrates that the risk preferences of users are important for understanding what information RS can provide that materially leads users to alter their consumption patterns.
 \par 
@@
 This metric depends only on the proposed item-set and not necessarily on the user's beliefs or how such a recommendation will change the item that the user consumes.
-\cite{adamopoulos2015unexpectedness} propose an expected utility approach but use this to measure unexpectedness in terms of deviations from the ``expected consideration set" of a user and do not explicitly incorporate user beliefs.
 \par 
@@
 It further results in ``unexpected" recommendations since it explicitly incorporates user beliefs and thus allows the recommender to understand how ``unexpected" a recommendation would be from the perspective of a user.
+
 % Bibliography
-\bibliographystyle{ACM-Reference-Format}
+\bibliographystyle{icml2020}
 \bibliography{refs}

### 2020-08-01 paper/filter_bubbles.tex
check in materials

@@
 \documentclass[manuscript]{acmart}
-\settopmatter{printacmref=false, printccs=false}
+\settopmatter{printacmref=true, printccs=true}
 \usepackage{enumitem}
@@
 
+  \keywords{Filter Bubbles, Recommender Systems, Similarity-based Generalization}
+
 \usepackage{afterpage}
@@
 \addauthor{dg}{red}    % dg for Duarte
-%\addauthor{ga}{blue}  % ga for Guy
+\addauthor{ga}{blue}  % ga for Guy
 
@@
 \newcommand\PaperTitle{Deconstructing the Filter Bubble: \\User Decision-Making and Recommender Systems}
-\title[\PaperTitle]{\PaperTitle}
 
-% Anonymized submission.
+\title[Deconstructing the Filter Bubble]{\PaperTitle}
+
+\author{Guy Aridor}
+\affiliation{%
+  \institution{Columbia University}
+  \city{New York}
+  \state{NY}
+  \country{USA}
+}
+\email{g.aridor@columbia.edu}
+
+\author{Duarte Gon\c{c}alves}
+\affiliation{%
+  \institution{Columbia University}
+  \city{New York}
+  \state{NY}
+    \country{USA}
+}
+\email{duarte.goncalves@columbia.edu}
+
+\author{Shan Sikdar}
+\affiliation{%
+  \institution{Everquote}
+  \city{Cambridge}
+  \state{MA}
+    \country{USA}
+}
+\email{shan.sikdar@gmail.com}
+
+\copyrightyear{2020}
+\acmYear{2020}
+\setcopyright{acmlicensed}\acmConference[RecSys '20]{Fourteenth ACM Conference on Recommender Systems}{September 21--26, 2020}{Virtual Event, Brazil}
+\acmBooktitle{Fourteenth ACM Conference on Recommender Systems (RecSys '20), September 21--26, 2020, Virtual Event, Brazil}
+\acmPrice{15.00}
+\acmDOI{10.1145/3383313.3412246}
+\acmISBN{978-1-4503-7583-2/20/09}
 
 % Abstract. Note that this must come before \maketitle.
+
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10010405.10010455.10010460</concept_id>
+<concept_desc>Applied computing~Economics</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10002951.10003317.10003347.10003350</concept_id>
+<concept_desc>Information systems~Recommender systems</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Applied computing~Economics}
+\ccsdesc[500]{Information systems~Recommender systems}
+
 \begin{abstract}
@@
 Our model provides an explanation for the findings of Nguyen, et.
-al (2014), where users consume increasingly similar items over time even without recommendation.
-Our model naturally generates such filter\dgedit{-}bubble effects since users face a sequential decision making problem under uncertainty and there are informational spillovers where a user consuming an item and learning how she values it reduces uncertainty and provides information about similar items.
-\par
-In line with Nguyen, et.
-al (2014), recommendation alleviates these filter\dgedit{-}bubble effects.
-However, we find that it also leads to an increase in homogeneity across users, resulting in a trade-off between homogenizing across user consumption and diversifying
-within user consumption.
-Finally, our model highlights the importance of collecting data on user beliefs and their evolution over time both to design better recommendations and to further understand their impact.
+al (2014), where, in environments where recommender systems are typically deployed, users consume increasingly similar items over time even without recommendation.
+We find that recommendation alleviates these natural filter-bubble effects, but that it also leads to an increase in homogeneity across users, resulting in a trade-off between homogenizing across-user consumption and diversifying within-user consumption.
+Finally, we discuss how our model highlights the importance of collecting data on user beliefs and their evolution over time both to design better recommendations and to further understand their impact.
 \end{abstract}
@@
 Recommender systems (RS) have become critical for assisting users in navigating the large choice sets that they face on many online platforms.
-For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
+For instance, users have to choose from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube.
 Users in many cases are not aware of most items, let alone have full information about their preferences over them.
@@
 Understanding how RS influence user behavior is important not only for characterizing the broader consequences of such systems but also for guiding their design.
-In this paper we develop a theoretical model of user decision-making in contexts where RS are traditionally deployed.
+In this paper, we develop a theoretical model of user decision-making in contexts where RS are traditionally deployed.
 We utilize previous empirical studies that characterize how RS influence user choice as a benchmark and our theoretical model provides an intuitive mechanism that can explain these empirical results.
@@
 A crucial component of our model is that users' beliefs about items are driven not only by recommendations, but also from their previous experiences with similar items.
-\dgdelete{We use these insights to provide guidance for RS design.
-In order to design good recommendations it becomes important for designers to understand user beliefs of the items they haven't consumed thus far and track how these beliefs evolve over time.}\dgedit{We use these insights to provide guidance for RS design, highlighting that understanding users' beliefs about the quality of the available items is essential in order to design recommendations and evaluate their impact.}
+We use these insights to provide guidance for RS design, highlighting that understanding users' beliefs about the quality of the available items is essential in order to design recommendations and evaluate their impact.
 
-\xhdr{Our Model\dgedit{.}} We \dgedit{analyze} a model of user choice \dgedit{with four central components}.
+\xhdr{Our Model.} We analyze a model of user choice with four central components.
 \par
@@
 \par
-The second component is\dgedit{, prior to consuming them, users are \textit{uncertain} about how much they value the different items.}
-This is motivated both by the fact \dgedit{that, in many contexts, recommendations regard experience goods,} whose true value can only be learned after consumption, and the fact that such uncertainty is why RS exist in the first place.
+The second component is that, prior to consuming them, users are \textit{uncertain} about how much they value the different items.
+This is motivated both by the fact that, in many contexts, recommendations regard experience goods, whose true value can only be learned after consumption, and the fact that such uncertainty is why RS exist in the first place.
 Thus, users face a sequential decision-making problem under uncertainty.
 \par 
-The third, and most crucial, element is that consumption of an item reveals information that changes user beliefs about the\dgedit{ir} valuation of similar items.
+The third, and most crucial, element is that consumption of an item reveals information that changes user beliefs about their valuation of similar items.
 Unlike in standard sequential decision-making problems, once an item is consumed all uncertainty about its valuation is resolved and provides information that enables users to update their beliefs about similar items.
 This exploits the fact that the valuations of similar items are correlated which assists users in navigating the vast product space.
-The idea that users make similarity-based assessments to guide their choice has grounding in \dgdelete{theoretical work in decision theory \cite{gilboa1995case} and } empirical evidence on how users navigate large choice sets \cite{schulz2019structured}.
+The idea that users make similarity-based assessments to guide their choice has grounding in empirical evidence on how users navigate large choice sets \cite{schulz2019structured}.
 \par 
@@
 
-\xhdr{Our Contributions\dgedit{.}}
-We provide a clear mechanism that explains the empirical results in \cite{nguyen2014exploring} who show that, in the context of movie consumption, user behavior is consistent with filter\dgedit{-}bubble effects even without recommendation and that recommendation leads to users being less likely to fall into such filter bubbles.
-In this context, filter\dgedit{-}bubble effects are defined as users consuming items in an increasingly narrow portion of the product space over time.
-The simple and intuitive driving force of this\dgdelete{ in our model} is that preferences for similar items are correlated, which implies that when an item is consumed and the user learns its value, it provides information about similar items.
-Crucially, this not only impacts the underlying belief about the \dgedit{expected value of similar items, but also how uncertain the user is about their valuation of them}.
+\xhdr{Our Contributions.}
+We provide a clear mechanism that explains the empirical results in \cite{nguyen2014exploring} who show that, in the context of movie consumption, user behavior is consistent with filter-bubble effects even without recommendation and that recommendation leads to users being less likely to fall into such filter bubbles.
+In this context, filter-bubble effects are defined as users consuming items in an increasingly narrow portion of the product space over time.
+The simple and intuitive driving force of this is that preferences for similar items are correlated, which implies that when an item is consumed and the user learns its value, it provides information about similar items.
+Crucially, this not only impacts the underlying belief about the expected value of similar items, but also how uncertain the user is about their valuation of them.
 Consequently, this learning spillover induces users to consume items similar to those they consumed before that had high realized value, leading to an increasing narrowing of consumption towards these regions of the product space.
@@
 We find that, while recommendation leads a single user to be more likely to explore diverse portions of the product space, it also coordinates consumption choices across users.
-This leads to an increase in homogeneity across users, resulting in a tradeoff between homogenizing across user consumption and diversifying within user consumption.
+This leads to an increase in homogeneity across users, resulting in a trade-off between homogenizing across-user consumption and diversifying within-user consumption.
 We explore the relationship between the overall diversity of consumed items and user welfare and find that more diverse sets of consumed items do not always correspond to higher user welfare.
@@
 Thus,  our evaluation measure determines the value of a recommendation based on the marginal welfare gain associated with providing a user with a recommendation over what the user would do without it.
-We discuss how this provides an additional rationale as to why ``accurate" recommendations are not always good recommendations.
+We discuss how this provides an additional rationale as to why ``accurate'' recommendations are not always good recommendations.
 \par
@@
 \cite{pariser2011filter} first informally described the idea of the ``filter bubble'' which is that online personalization services would lead users down paths of increasingly narrower content so that they would effectively be isolated from a diversity of viewpoints or content.
-Following this\dgedit{,} a number of empirical studies\dgdelete{, done} in various disciplines, have since studied the extent to which this phenomenon exists in a wide range of contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}.
+Following this, a number of empirical studies in various disciplines, have since studied the extent to which this phenomenon exists in a wide range of contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}.
 The most relevant to our study is \cite{nguyen2014exploring} who study whether this effect exists in the context of movie consumption.
@@
 To our knowledge there are no theoretical models that rationalize these empirical findings and we provide a theoretical framework through which to view this problem.
-Furthermore, we provide a clear mechanism that drives such effects and how recommendation interacts with them.
+Moreover, we provide a clear mechanism that drives such effects and how recommendation interacts with them.
 \par 
@@
 \par
-Another set of papers studies the impact of human decision-making on the design and evaluation of RS.
+A third strand in the literature studies the impact of human decision-making on the design and evaluation of RS.
 \cite{chen2013human} surveys the literature on the relationship between human decision making and RS.
@@
 Thus, the primary insight of our paper is that user beliefs and how users update their beliefs about similar items after consumption are important and previously unconsidered elements of human decision making that are critical for understanding the design and consequences of RS.
-Within this literature, \cite{celma2008new, cremonesi2013user, pu2011user} focus on ``user-centric" approaches to recommendation whereby user evaluation of the usefulness of recommendation is a key evaluation measure.
+Within this literature, \cite{celma2008new, cremonesi2013user, pu2011user} focus on ``user-centric'' approaches to recommendation whereby user evaluation of the usefulness of recommendation is a key evaluation measure.
 Our evaluation measure is similar, but, unlike previous approaches, emphasizes the importance of user beliefs.
@@
 However, we assume that users have the same utility over money, given by a utility function $u: \mathbb R \to \mathbb R$, strictly increasing and continuous.
-So, ex-post, the value of item $n$ for user $i$ is given by $u(x_{i,n})$\dgedit{.
-B}efore consuming the item, the user does not know exactly how she will value it.
+So, ex-post, the value of item $n$ for user $i$ is given by $u(x_{i,n})$.
+Before consuming the item, the user does not know exactly how she will value it.
 In particular, even users that will end up having the same ex-post valuation of item $n$ may differ in their ex-ante valuation because they hold different beliefs about it.
@@
 This sure amount of money is called the \textit{certainty equivalent} of gamble $x$ and is denoted as $\delta(x)$.
-A user $i$ is more risk-averse than another user $j$ \dgedit{if whenever user $j$ prefers a sure thing to the gamble, then user $i$ does too}\dgdelete{when for every gamble $x$, user $i$ requires a lower value to give up the gamble when compared to user $j$, i.e.
-$\delta_i(x)$}.
+A user $i$ is more risk-averse than another user $j$ if whenever user $j$ prefers a sure thing to the gamble, then user $i$ does too.
 Therefore, a more risk-averse user is more willing to avoid the risk of taking the gamble.
-We assume that the utility function takes a flexible functional form $u(x)=1-\exp(-\gamma x)$ for $\gamma\ne0$ and $u(x)=x$ for $\gamma\dgedit{\to}0$ -- known as constant absolute risk-aversion preferences (from hereon CARA).
-Higher $\gamma$ implies higher risk-aversion with $\gamma\dgedit{\to}0$ corresponding to the risk-neutral case and $\gamma>0$ to the risk-averse one.
-Our formulations here follow standard economic decision-theory (see \cite{mas1995microeconomic} for a textbook treatment of these topics).
+We assume that the utility function takes a flexible functional form $u(x)=1-\exp(-\gamma x)$ for $\gamma\ne0$ and $u(x)=x$ for $\gamma\to 0$ -- known as constant absolute risk-aversion preferences (from hereon CARA).
+Higher $\gamma$ implies higher risk-aversion, with $\gamma \to 0$ corresponding to the risk-neutral case and $\gamma>0$ to the risk-averse one.
+Our formulations here follow standard economic consumer theory (see \cite{mas1995microeconomic} for a textbook treatment of these topics).
 \par
@@
 Stacking values in vector-form, we get the vector of values associated with each item 
-$${\left(x_{i,n}\right)}_{n \in \mathcal{J}}=:X_i =V_i+ \beta V $$
+$${\left(x_{i,n}\right)}_{n \in \mathcal{J}}=:X_i =V_i+ \beta V, $$
 where $V_i ={\left(v_{i,n}\right)}_{n \in \mathcal{J}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{J}}$.
@@
 Formally, user $i$ starts with some beliefs about $X_i$, namely that the idiosyncratic and common-value parts of the valuations are independent -- $V_i \perp \!\!\!
-\perp V$ -- and that each is multivariate normal
+\perp V$ -- and that each is multivariate normal:
 \begin{enumerate}[topsep=0pt]
-\item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ 
-\item $V \sim \mathcal N(\overline V, \Sigma)$ with $\overline V =0$
+\item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$; and 
+\item $V \sim \mathcal N(\overline V, \Sigma)$ with $\overline V =0$.
 \end{enumerate}
@@
 \par
-\dgedit{Recalling} that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{i,n}\right)_{n \in \mathcal{J}}$ are drawn independently from a jointly normal distribution, where $\overline v_{i,n} \sim \mathcal N (0, \overline \sigma^2)$ are independent and identically distributed.
+Recalling that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{i,n}\right)_{n \in \mathcal{J}}$ are drawn independently from a jointly normal distribution, where $\overline v_{i,n} \sim \mathcal N (0, \overline \sigma^2)$ are independent and identically distributed.
 These $\overline v_{i,n}$ denote the prior belief that individual $i$ holds about her valuation over item $n$.
@@
 Note that this expression is known to be the certainty equivalent for CARA preferences in a Gaussian environment \cite{mas1995microeconomic}.
+As it is immediate from this expression, the user assigns greater value to items for which the expected monetary equivalent, $\mu_n$, is higher, but penalizes those about which there is greater uncertainty $\Sigma_{nn}$, the more so the greater the user's degree of risk aversion $\gamma$.
 
-\xhdr{User Learning\dgedit{.}}
+\xhdr{User Learning.}
 When a user consumes an item $n$ she learns the realized value for that item.
-We \dgedit{consider the case where}  learning about the value of item $n$ reveals more about the value associated to items that are closer to it, which captures the idea that trying an item provides more information about similar items than about dissimilar ones.\footnote{\cite{schulz2019structured} empirically studies how individuals solve sequential decision-making problems under uncertainty in large choice sets in the context of mobile food delivery orders.
+We consider the case where  learning about the value of item $n$ reveals more about the value associated to items that are closer to it, which captures the idea that trying an item provides more information about similar items than about dissimilar ones.\footnote{\cite{schulz2019structured} empirically studies how individuals solve sequential decision-making problems under uncertainty in large choice sets in the context of mobile food delivery orders.
 They find that individuals engage in similarity-based generalizations where learning about the realized value of a particular item provides them with information about similar items.
@@
 This distance function is not intended to model the intricacies of a realistic product space, but instead to provide a stylized product space to help us understand the effects of informational spillovers on user behavior.
-The basic intuition, in the context of movie consumption, is that a \dgedit{user's valuation of} \textit{John Wick} (item $n$) provides more information about how much she will like \textit{John Wick: Chapter Two} (item $m$) than \textit{Titanic} (item $q$) since $d(n, m) < d(n, q)$.
+The basic intuition, in the context of movie consumption, is that a user's valuation of \textit{John Wick} (item $n$) provides more information about how much she will like \textit{John Wick: Chapter Two} (item $m$) than \textit{Titanic} (item $q$) since $d(n, m) < d(n, q)$.
 \par
 We consider that the entry of $n$-th row and the ($m$)-th column of $\Sigma_i$ is given by $\sigma_i^2 \rho^{d(n,m)}$, and that of $\Sigma$ is given by $\sigma^2 \rho^{d(n,m)}$.
-The scalar $\rho \in [0,1]$ therefore impacts the covariance structure: a higher $\rho$ implies that learning the utility of $n$ is more informative about products nearby.
-Informativeness, for any $\rho \in (0,1)$, is decreasing in distance.
-The particular distance function that we utilize leads to this covariance structure being simple, where the $(n,n+1)$-th entry in the covariance matrix is $\sigma^{2} \rho$, the $(n,n+2)$-th entry is $\sigma^{2} \rho^2$, etc.\footnote{This exponential decay correlation structure can be related to the tenet of case-based similarity of \cite{gilboa1995case} -- see \cite{billot2008axiomatization} for an axiomatization of exponential similarity.}
+The scalar $\rho \in [0,1]$ therefore impacts the covariance structure: a higher $\rho$ implies that learning the utility of $n$ is more informative about products nearby and, for $\rho \in (0,1)$, this effect is decreasing in distance.
+The particular distance function that we rely on leads to a simple covariance structure, where the $(n,n+1)$-th entry in the covariance matrix is $\sigma^{2} \rho$, the $(n,n+2)$-th entry is $\sigma^{2} \rho^2$, etc.\footnote{This exponential decay correlation structure can be related to the tenet of case-based similarity of \cite{gilboa1995case} -- see \cite{billot2008axiomatization} for an axiomatization of exponential similarity.}
 \par
@@
 \Sigma_{(t,N-t)} & \Sigma_{(t,t)}
-\end{array} \right)
+\end{array} \right).
 \]
-Thus, after consuming the items in $C_{i}^{t}$ the resulting beliefs over the remaining items are given by $\mathcal{N}(\bar{\mu}, \bar{\Sigma})$ where $\bar{\mu}$ and $\bar{\Sigma}$ are as follows:
+After consuming the items in $C_{i}^{t}$, the resulting beliefs over the remaining items are given by $\mathcal{N}(\bar{\mu}, \bar{\Sigma})$ where $\bar{\mu}$ and $\bar{\Sigma}$ are as follows:
 \begin{align*}
-\bar{\mu} \mid c_t = \mu_{N-t} + \Sigma_{(N-t,t)} \Sigma_{(t,t)}^{-1}(c_t - \mu_t) \\
-\bar{\Sigma} \mid c_t = \Sigma_{(N-t,N-t)} - \Sigma_{(N-t,t)} \Sigma_{(t,t)}^{-1} \Sigma_{(t,N-t)}
+\bar{\mu} \mid c_t &= \mu_{N-t} + \Sigma_{(N-t,t)} \Sigma_{(t,t)}^{-1}(c_t - \mu_t); \\
+\bar{\Sigma} \mid c_t &= \Sigma_{(N-t,N-t)} - \Sigma_{(N-t,t)} \Sigma_{(t,t)}^{-1} \Sigma_{(t,N-t)}.
 \end{align*}
 
-\xhdr{An Illustrative Example\dgedit{.}} We illustrate the main intuitions of our model with a simple example.
+\xhdr{An Illustrative Example.} We illustrate the main intuitions of our model with a simple example.
 Suppose that there are four items: 0, 1, 2, 3.
 The items are in different places of the product space, where 0 is close to 1 and 3 but more distant from 2.
-\dgedit{For the sake of expositional clarity, suppose that t}he initial beliefs are given as follows:
-\[ \mu = \left (\begin{array}{c}
-\mathbb{E}[x_0] \\
-\mathbb{E}[x_1] \\
-\mathbb{E}[x_2] \\
-\mathbb{E}[x_3]
-\end{array}  \right) = \left (\begin{array}{c}
-0 \\
-0\\
-0 \\
-0
-\end{array}  \right), \dgedit{\qquad}\Sigma =  \sigma^{2} \left( \begin{array}{cccc}
-\rho^{0} & \rho^{1} & \rho^{2} & \rho^{0} \\
-\rho^{1} & \rho^{0} & \rho^{1} & \rho^{2} \\
-\rho^{2} & \rho^{1} & \rho^{0} & \rho^{1} \\
-\rho^{1} & \rho^{2} & \rho^{1} & \rho^{0} \\
-\end{array} \right)
-\]
+For the sake of expositional clarity, suppose that the initial mean beliefs are given by $\mu=(\mathbb E[x_n])_{n=0}^3=(0)_{n=0}^3$.
+\par
 In period 1, every item is ex-ante identical since they have the same mean and variance and so suppose that the user breaks the tie arbitrarily and consumes item 0.
@@
 For concreteness, we suppose that $\sigma = 1$ and $\rho = 0.5$, but the intuitions hold for any value of $\sigma$ and $\rho > 0$.
-First, we consider the case when the realization of $y > 0$ and\dgdelete{, for concreteness}, specifically\dgedit{,} \dgdelete{when} $y = 0.5$\dgedit{ --} though the general intuitions hold for any $y > 0$.
+First, we consider the case when the realization of $y > 0$ and, specifically, $y = 0.5$ -- though the general intuitions hold for any $y > 0$.
 The resulting beliefs after observing $y$ are then as follows:
-\[ \bar{\mu} =   \left (\begin{array}{c}
-\mathbb{E}[x_1] \vspace{0.15cm} \\
-\mathbb{E}[x_2] \vspace{0.15cm} \\
-\mathbb{E}[x_3]
+\[ \bar{\mu} = (\mu \mid x_0=y) =  \left (\begin{array}{c}
+\mathbb{E}[x_1\mid x_0=y] \vspace{0.15cm} \\
+\mathbb{E}[x_2\mid x_0=y] \vspace{0.15cm} \\
+\mathbb{E}[x_3\mid x_0=y]
 \end{array}  \right) =\left (\begin{array}{c}
@@
 \frac{1}{4}
-\end{array}  \right), \bar{\Sigma} =  \left( \begin{array}{ccc}
+\end{array}  \right),\qquad 
+\bar{\Sigma} =(\Sigma\mid x_0=y)=  \left( \begin{array}{ccc}
 \frac{3}{4} & \frac{3}{8} & 0 \vspace{0.15cm} \\
@@
 0 &\frac{3}{8} & \frac{3}{4}  \\
-\end{array} \right)
+\end{array} \right).
 \]
-Thus, upon learning $\dgedit{x_0=}y$, the user updates beliefs about the remaining items.
-Note that $\mathbb{E}[x_1] = \mathbb{E}[x_3] > \mathbb{E}[x_2]$ since item 0's value is more informative about similar items' values, items 1 and 3, than items further way in the product space such as item 2.
+Thus, upon learning $x_0=y$, the user updates beliefs about the remaining items.
+Note that $\mathbb{E}[x_1\mid x_0=y] = \mathbb{E}[x_3\mid x_0=y] > \mathbb{E}[x_2\mid x_0=y]$ since item 0's value is more informative about similar items' values, items 1 and 3, than items further away in the product space such as item 2.
 Moreover, $\bar{\Sigma}_{11} = \bar{\Sigma}_{33} < \bar{\Sigma}_{22}$ as the uncertainty about items 1 and 3 is further reduced compared to item 2.
@@
 Now we consider the case when item 0 ends up having a negative valuation so that $y = -0.5 < 0$.
-This results in $\mathbb{E}[x_1] = \mathbb{E}[x_3] = -\frac{1}{4} <  -\frac{1}{8} = \mathbb{E}[x_2]$ with $\bar{\Sigma}$ remaining the same as when $y = 0.5$.
+This results in $\mathbb{E}[x_1\mid x_0=y] = \mathbb{E}[x_3\mid x_0=y] = -\frac{1}{4} <  -\frac{1}{8} = \mathbb{E}[x_2\mid x_0=y]$ with $\bar{\Sigma}$ remaining the same as when $y = 0.5$.
 In this case the risk-aversion levels of the user determine the choice in the next period.
@@
 
-We further consider two cases that serve \dgedit{mainly} as benchmarks.
+We further consider two cases that serve mainly as benchmarks.
 The first is \textit{no recommendation}, where users get no additional information and make consumption choices based on their beliefs and consumption history.
@@
 \xhdr{Simulation Details.}
-We analyze our model using numerical simulation since the sequential decision-making component paired with the rich covariance structure between the items make it difficult to characterize optimal user behavior analytically.
-The Gaussian assumption allows for closed form belief updating which enables us to simulate our model but does not provide much help in analytical characterizations.
+We analyze our model using numerical simulation since the sequential decision-making component paired with the rich covariance structure between the items make it difficult to characterize optimal user behavior analytically.\footnote{The Gaussian assumption allows for closed form belief updating which enables us to simulate our model but does not provide much help in analytical characterizations.}
 \par
@@
 We define local consumption in terms of the average consumption distance between the items consumed by the users at time $t-1$ and $t$.
-Thus, in the context of our model, filter\dgedit{-}bubble effects arise when the average consumption distance decreases over time and, across regimes, when the levels are lower for a given recommendation regime compared to another.
+Thus, in the context of our model, filter-bubble effects arise when the average consumption distance decreases over time and, across regimes, when the levels are lower for a given recommendation regime compared to another.
 
-\dgedit{Our first finding can be summarized as follows:}
+Our first finding can be summarized as follows:
 
@@
 \begin{figure}[t]
-\caption{Local Consumption and Correlation, $N = 200$}
+\caption{Local Consumption and Correlation}
 \begin{subfigure}{.3\linewidth}
@@
 \end{subfigure}%
+\hspace{1cm}
 \begin{subfigure}{.3\linewidth}
@@
 \caption*{\scriptsize Notes: The figure shows the consecutive consumption path difference between the no recommendation, recommendation, and oracle regime.
-The figure on the left displays the average consecutive consumption distance aggregating over simulations where $\rho > 0$ and the figure on the right displays the average consecutive consumption distance aggregating over simulations where $\rho = 0$.}
+The figure on the left displays the average consecutive consumption distance aggregating over simulations where $\rho \in (0,1)$ and the figure on the right displays the average consecutive consumption distance aggregating over simulations where $\rho = 0$.}
 \label{fig:correlation_consumption_path}
@@
 \par
-The left panel of Figure~\ref{fig:correlation_consumption_path} shows that, when $\rho > 0$, both recommendation and no recommendation lead to increasingly local consumption compared to the oracle benchmark case.
-Further, the average consumption path between periods is \textit{decreasing} for the no recommendation case whereas it is \textit{increasing} for the oracle case.
+The left panel of Figure~\ref{fig:correlation_consumption_path} shows that, when $\rho \in (0,1)$, both recommendation and no recommendation lead to increasingly local consumption compared to the oracle benchmark case.
+Moreover, the average consumption path between periods is \textit{decreasing} for the no recommendation case whereas it is \textit{increasing} for the oracle case.
 The recommendation regime decreases the degree of local consumption, but not as much as the oracle benchmark.
 Due to the correlation of values, the oracle consumption path exploits this and leads to the consumption of more similar items than in the case when $\rho = 0$.
-However, since these spillovers also impact user learning in the no recommendation case, users \textit{over-exploit} this and increasingly consume items similar to high value items that they have consumed before.
-This is further illustrated in the top row of~\autoref{fig:local_consumption_vary_rho_gamma}, which shows how the consumption paths in the oracle and no recommendation regimes vary as $\rho$ increases and is in line with this intuition.
+However, since these spillovers also impact user learning in the no recommendation case, users \textit{over-exploit} these and increasingly consume items similar to high value items that they have consumed before.
+This is also illustrated in the top row of~\autoref{fig:local_consumption_vary_rho_gamma}, which shows how the consumption paths in the oracle and no-recommendation regimes vary as $\rho$ increases and is in line with this intuition.
 \par
-This effect is further amplified as the level of risk aversion, $\gamma$, increases.
-The bottom row of~\autoref{fig:local_consumption_vary_rho_gamma} vaFurtherlidates that the degree of local consumption drastically increases as $\gamma$ increases.
+As shown in the bottom row of~\autoref{fig:local_consumption_vary_rho_gamma}, this effect is further amplified as the level of risk aversion increases: the degree of local consumption drastically increases as $\gamma$ increases.
 This is due to the fact that the spillovers not only impact the mean expected belief about quality but also the degree of uncertainty.
@@
 
-In sum, filter\dgedit{-}bubble effects only arise when there is an inherent correlation between the realized utilities of the items in the product space.
+In sum, filter-bubble effects only arise when there is an inherent correlation between the realized utilities of the items in the product space.
 When there is a correlation between the realized utilities then filter bubbles can naturally arise due to the nature of how individuals acquire additional information about the remaining items.
@@
 Recommendation leads to these effects being mitigated by providing users with additional information on products outside the already explored portions of the product space.
-Furthermore, if all uncertainty was resolved as in the oracle regime, then such behavior is not present.
+Logically, if all uncertainty were resolved as in the oracle regime, then such behavior is not present.
 
@@
 $$D_i:=\frac{1}{N}\frac{1}{T(T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$$
-\dgedit{\noindent 
-Finding~\ref{finding_diversity} summarizes the main results on product diversity:}
+\noindent 
+Finding~\ref{finding_diversity} summarizes the main results on product diversity:
 \begin{finding}\label{finding_diversity}
@@
 \item When $\rho = 0$, product diversity is the same across all three recommendation regimes;
-\item When $\rho > 0$, product diversity decreases across all recommendation regimes but decreases the most in the no recommendation regime.
+\item When $\rho \in (0,1)$, product diversity decreases across all recommendation regimes but decreases the most in the no-recommendation regime.
 This effect is amplified as $\rho$ increases as well as when users become more risk-averse.
@@
 \begin{figure}[t]
-\caption{Relationship between Local Consumption and $\rho$, $\gamma$}
-\begin{subfigure}{.3\textwidth}
+\centering
+\caption{Relationship between Local Consumption and Correlation ($\rho$) and Risk Aversion ($\gamma$)}
+\begin{subfigure}{.3\linewidth}
 \includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20.pdf}
 \end{subfigure}
-\begin{subfigure}{.3\textwidth}
+\begin{subfigure}{.3\linewidth}
 \includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_partial.pdf}
 \end{subfigure}
-\begin{subfigure}{.3\textwidth}
-\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_omni.pdf}\\
-\end{subfigure}
-\begin{subfigure}{.3\textwidth}
+\begin{subfigure}{.3\linewidth}
+\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_omni.pdf}
+\end{subfigure}\\
+\begin{subfigure}{.3\linewidth}
 \includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20.pdf}
 \end{subfigure}
-\begin{subfigure}{.3\textwidth}
+\begin{subfigure}{.3\linewidth}
 \includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_partial.pdf}
 \end{subfigure}
-\begin{subfigure}{.3\textwidth}
+\begin{subfigure}{.3\linewidth}
 \includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_omni.pdf}
@@
 \caption*{\scriptsize Notes: Each figure plots the average consecutive consumption distance across time as the inherent correlation between the valuation of the items, $\rho$, varies (top row) and the level of risk-aversion, $\gamma$, varies (bottom row).
-In both rows, the left displays the no recommendation regime, the center displays the recommendation regime, and the right displays the oracle regime.}
+In both rows, the left displays the no-recommendation regime, the center displays the recommendation regime, and the right displays the oracle regime.}
 \label{fig:local_consumption_vary_rho_gamma}
@@
 \addtocounter{figure}{-1}
-\dgdelete{Finding~\ref{finding_diversity} summarizes the main results on product diversity.
-}As before, when there is no correlation between valuations, product diversity is the same across different recommendation regimes.
-The over-exploitation of information spillovers when $\rho > 0$ leads to product diversity being lowest in the no recommendation regime.
+As before, when there is no correlation between valuations, product diversity is the same across different recommendation regimes.
+The over-exploitation of information spillovers when $\rho \in (0,1)$ leads to product diversity being lowest in the no-recommendation regime.
 As a result, this effect gets amplified as $\rho$ increases, which leads to the gap in diversity between the regimes to increase as $\rho$ increases.
@@
 There is a similar increasing diversity gap as $\gamma$, or the level of risk-aversion, increases as can be seen in the bottom row of~\autoref{fig:local_consumption_vary_rho_gamma}.
-The mechanisms behind these effects have direct parallels to those discussed in the previous section since low average sequential consumption distance directly leads to low diversity.
+The mechanisms behind these effects directly parallels those discussed in the previous section since low average sequential consumption distance directly leads to low diversity.
 \par
@@
 
-\dgedit{\noindent Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which can be seen in the rightmost plot of~\autoref{fig:diversity_welfare_correlation}:}
+\noindent Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which can be seen in the rightmost plot of~\autoref{fig:diversity_welfare_correlation}:
+
+~\par
 \begin{finding}\label{finding_welfare_gap}
@@
 \begin{enumerate}
-\item Under oracle recommendation, welfare is \dgedit{invariant with respect to }$\rho$.
-\item Under no recommendation, welfare is increasing \dgedit{in} $\rho$.
+\item Under oracle recommendation, welfare is invariant with respect to $\rho$.
+\item Under no recommendation, welfare is increasing in $\rho$.
 \item Recommendation introduces welfare gains relative to no recommendation, but these gains are decreasing as $\rho$ increases.
@@
 \begin{figure}[t]
-\caption{Relationship between $\rho$, $\gamma$ and User Welfare, Diversity}\label{fig:diversity_welfare_correlation}
+\caption{Relationship between Correlation ($\rho$) and Risk Aversion ($\gamma$), and User Welfare and Diversity}\label{fig:diversity_welfare_correlation}
 \begin{subfigure}{0.3\linewidth}
@@
 \begin{subfigure}{0.3\linewidth}
-\includegraphics[width=1.0\linewidth]{figures/gamma_diversity_N_200_T_20}
+\includegraphics[width=1.0\linewidth]{figures/gamma_diversity_N_200_T_20.pdf}
 \end{subfigure}
@@
 
-\dgdelete{Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which can be seen in the rightmost plot of~\autoref{fig:diversity_welfare_correlation}.
-}
 The most interesting observation is that the value of recommendation decreases as $\rho$ decreases.
-While welfare in the recommendation regime is flat as we increase $\rho$, it is increasing in the no recommendation regime and thus the welfare gap between the two shrinks as $\rho$ increases.
+While welfare in the recommendation regime is flat as we increase $\rho$, it is increasing in the no-recommendation regime and thus the welfare gap between the two shrinks as $\rho$ increases.
 The intuition is clear as recommendation provides users with information that allows them to better guide their decisions and increase welfare.
 However, as $\rho$ increases users get increasingly more information from consuming items since the realized utility is now more informative about the utility of nearby items.
-Thus, since consumption decisions themselves reveal information, the information that recommendation reveals provides less value.
+Thus, since consumption decisions themselves yield valuable information, the information provided by recommendation is less valuable to the user.
 \par 
@@
 We next explore the relationship between welfare and diversity.
-\dgedit{
 \par
-\noindent Finding~\ref{finding_diversity_welfare_corr} summarizes our findings on the relationship between diversity and welfare:}
+\noindent Finding~\ref{finding_diversity_welfare_corr} summarizes our findings on the relationship between diversity and welfare:
 
 \begin{finding}\label{finding_diversity_welfare_corr}
-In the no recommendation regime, diversity and welfare are:
+In the no-recommendation regime, diversity and welfare are:
 \begin{enumerate}
@@
 \caption{Diversity vs. Welfare}
-\begin{subfigure}{.3\textwidth}
+\begin{minipage}{1.2\textwidth}
+\hspace*{-1cm}\begin{subfigure}{.23\textwidth}
 \includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_n_200.pdf}
 \end{subfigure}
-\begin{subfigure}{.3\textwidth}
+\begin{subfigure}{.23\textwidth}
 \includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_n_200.pdf}
-\end{subfigure}\\
-\begin{subfigure}{.3\textwidth}
+\end{subfigure}
+\begin{subfigure}{.23\textwidth}
 \includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_partial_n_200.pdf}
 \end{subfigure}
-\begin{subfigure}{.3\textwidth}
+\begin{subfigure}{.23\textwidth}
 \includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_partial_n_200.pdf}
-\end{subfigure}\\
-\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under the recommendation (top row)  and no recommendation (bottom row) regime.
-In both rows, the figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.}\label{fig:diversity_welfare_ra}
+\end{subfigure}
+\end{minipage}\\
+\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under recommendation for$\gamma = 0$ (first), $\gamma = 5$ (second) and no recommendation for $\gamma = 0$ (third), $\gamma = 5$ (fourth).}\label{fig:diversity_welfare_ra}
 \end{figure}
@@
 
-\dgdelete{Finding~\ref{finding_diversity_welfare_corr} summarizes our findings on the relationship between diversity and welfare.
-}Figure \ref{fig:diversity_welfare_ra} shows how diversity and welfare correlate for the no recommendation case as we vary the degree of risk aversion.
+Figure \ref{fig:diversity_welfare_ra} shows how diversity and welfare correlate for the no recommendation case as we vary the degree of risk aversion.
 When there is no risk-aversion then there is a negative correlation between welfare and diversity.
@@
 where $d_J$ denotes the Jaccard index and $H \in [0,1]$.
-\dgedit{\par
-\noindent Finding~\ref{finding_homogeneity} summarizes our findings on the impact of recommendation on user homogeneity:}
+\par
+\noindent Finding~\ref{finding_homogeneity} summarizes our findings on the impact of recommendation on user homogeneity:
 
@@
 \begin{figure}[t]
-\caption{Relationship between $\beta$ and Homogeneity, $N = 200$}
-\includegraphics[width=.3\linewidth]{figures/beta_homogeneity_N_200_T_20}
+\caption{Relationship between $\beta$ and Homogeneity}
+\includegraphics[width=.3\linewidth]{figures/beta_homogeneity_N_200_T_20} \hspace{1.0cm}
 \includegraphics[width=.3\linewidth]{figures/rho_homogeneity_N_200_T_20}\label{fig:beta_homo}
@@
 
-\dgdelete{Finding~\ref{finding_homogeneity} summarizes our findings on the impact of recommendation on user homogeneity.
-}First, we study how the degree of homogenization varies as we increase $\beta$, the weight of the common value component.
+First, we study how the degree of homogenization varies as we increase $\beta$, the weight of the common value component.
 As $\beta$ increases we expect that users should become increasingly homogeneous as the realized utilities of the items are now increasingly similar.~\autoref{fig:beta_homo} confirms that as the weight of the common-value component $\beta$ increases, users consume increasingly similar sets of items.
@@
 As $\beta$ increases, some amount of homogenization is optimal as can be seen from the oracle case.
-However, since users in the no recommendation regime do not know the common-value component they engage in local consumption in different areas of the product space which leads to less than optimal homogeneity.
+However, since users in the no-recommendation regime do not know the common-value component they engage in local consumption in different areas of the product space which leads to less than optimal homogeneity.
 \par
 
-We next study how the degree of homogeneity varies \dgedit{with} $\rho$\dgdelete{ increases}.
-~\autoref{fig:beta_homo} shows how \dgdelete{homogeneity varies as $\rho$ increases.
-The most striking finding is that }homogeneity decreases as $\rho$ increases in the recommendation regime.
+We next study how the degree of homogeneity varies with $\rho$.
+~\autoref{fig:beta_homo} shows how homogeneity decreases as $\rho$ increases in the recommendation regime.
 As was highlighted in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, the degree of local consumption increases with $\rho$.
@@
 The first and most crucial type of data to collect is individual user \textit{beliefs} about items that they have not yet consumed.
-As illustrated by our model, these beliefs are what drive the future consumption decisions of users and understanding these beliefs is crucial for determining the value of recommending certain items.\footnote{Additionally, user beliefs contain information that may not have been observed by the recommender that only observes user choices on the platform.} The second type of data that is \dgedit{relevant} for RS designers to collect is how user beliefs change over time and, in particular, not just how individuals value the item they just consumed, but also how it impacts their beliefs about the quality of similar items.\footnote{Characterizing the similarity between items has been an important goal of designing content-based recommendations, though as noted by \cite{winecoff2019users}, how users perceive similarity between items is not always in line with how similarity is computed in content-based RS.
+As illustrated by our model, these beliefs are what drive the future consumption decisions of users and understanding these beliefs is crucial for determining the value of recommending certain items.\footnote{Additionally, user beliefs contain information that may not have been observed by the recommender that only observes user choices on the platform.} The second type of data that is relevant for RS designers to collect is how user beliefs change over time and, in particular, not just how individuals value the item they just consumed, but also how it impacts their beliefs about the quality of similar items.\footnote{Characterizing the similarity between items has been an important goal of designing content-based recommendations, though as noted by \cite{winecoff2019users}, how users perceive similarity between items is not always in line with how similarity is computed in content-based RS.
 Understanding how this impacts which items users update their beliefs about is an important direction for future work.} The third type of data is the risk-aversion levels of users as our model illustrates that the risk preferences of users are important for understanding what information RS can provide that materially leads users to alter their consumption patterns.
@@
 Our approach most closely follows the set of proposed serendipity measures which are surveyed in \cite{kotkov2016survey}.
-As discussed by \cite{maksai2015predicting},  serendipitous recommendations are said to ``have the quality of being both unexpected and useful" which is in line with the primary intuition behind our approach.
+As discussed by \cite{maksai2015predicting},  serendipitous recommendations are said to ``have the quality of being both unexpected and useful'' which is in line with the primary intuition behind our approach.
 The primary difference between our proposed approach and those existing in the literature is that ours crucially hinges on understanding user beliefs and the risk-preferences of users.
@@
 This metric depends only on the proposed item-set and not necessarily on the user's beliefs or how such a recommendation will change the item that the user consumes.
-\cite{adamopoulos2015unexpectedness} propose an expected utility approach but use this to measure unexpectedness in terms of deviations from the ``expected consideration set" of a user and do not explicitly incorporate user beliefs.
+\cite{adamopoulos2015unexpectedness} propose an expected utility approach but use this to measure unexpectedness in terms of deviations from the ``expected consideration set'' of a user and do not explicitly incorporate user beliefs.
 \par 
@@
 Our evaluation measure leads to useful recommendations since it leads users towards better items than they would consume without recommendation.
-It further results in ``unexpected" recommendations since it explicitly incorporates user beliefs and thus allows the recommender to understand how ``unexpected" a recommendation would be from the perspective of a user.
+It further results in ``unexpected'' recommendations since it explicitly incorporates user beliefs and thus allows the recommender to understand how ``unexpected'' a recommendation would be from the perspective of a user.
 % Bibliography

### 2020-08-01 paper/filter_bubbles_arxiv.tex
check in materials

--- 2020-08-01 paper/filter_bubbles.tex
@@
-\documentclass[manuscript]{acmart}
-\settopmatter{printacmref=true, printccs=true}
+\documentclass[manuscript, nonacm]{acmart}
+\settopmatter{printacmref=false, printccs=true}
+\renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference information in first column
 \usepackage{enumitem}
@@
 \usepackage{afterpage}
-\usepackage[suppress]{color-edits}
-\addauthor{dg}{red}    % dg for Duarte
-\addauthor{ga}{blue}  % ga for Guy
 
@@
 
-\author{Guy Aridor}
-\affiliation{%
-  \institution{Columbia University}
-  \city{New York}
-  \state{NY}
-  \country{USA}
-}
-\email{g.aridor@columbia.edu}
+% Abstract. Note that this must come before \maketitle.
 
-\author{Duarte Gon\c{c}alves}
-\affiliation{%
-  \institution{Columbia University}
-  \city{New York}
-  \state{NY}
-    \country{USA}
-}
-\email{duarte.goncalves@columbia.edu}
-
-\author{Shan Sikdar}
-\affiliation{%
-  \institution{Everquote}
-  \city{Cambridge}
-  \state{MA}
-    \country{USA}
-}
-\email{shan.sikdar@gmail.com}
-
-\copyrightyear{2020}
-\acmYear{2020}
-\setcopyright{acmlicensed}\acmConference[RecSys '20]{Fourteenth ACM Conference on Recommender Systems}{September 21--26, 2020}{Virtual Event, Brazil}
-\acmBooktitle{Fourteenth ACM Conference on Recommender Systems (RecSys '20), September 21--26, 2020, Virtual Event, Brazil}
-\acmPrice{15.00}
-\acmDOI{10.1145/3383313.3412246}
-\acmISBN{978-1-4503-7583-2/20/09}
-
-% Abstract. Note that this must come before \maketitle.
+\begin{document}
 
@@
 
-\begin{document}
+\author{Guy Aridor}
+\affiliation{%
+  \institution{Columbia University}
+  \city{New York}
+  \state{NY}
+}
+\email{g.aridor@columbia.edu}
+
+\author{Duarte Gon\c{c}alves}
+\affiliation{%
+  \institution{Columbia University}
+  \city{New York}
+  \state{NY}
+}
+\email{duarte.goncalves@columbia.edu}
+
+\author{Shan Sikdar}
+\affiliation{%
+  \institution{Everquote}
+  \city{Cambridge}
+  \state{MA}
+}
+\email{shan.sikdar@gmail.com}
 
@@
 A crucial component of our model is that users' beliefs about items are driven not only by recommendations, but also from their previous experiences with similar items.
-We use these insights to provide guidance for RS design, highlighting that understanding users' beliefs about the quality of the available items is essential in order to design recommendations and evaluate their impact.
+We use these insights to provide guidance for RS design, highlighting that understanding users' beliefs about the quality of the available items is essential to design recommendations and evaluate their impact.
 
@@
 The second component is that, prior to consuming them, users are \textit{uncertain} about how much they value the different items.
-This is motivated both by the fact that, in many contexts, recommendations regard experience goods, whose true value can only be learned after consumption, and the fact that such uncertainty is why RS exist in the first place.
+This is motivated both by the fact that recommender systems are traditionally deployed in contexts with experience goods, whose true value can only be learned after consumption, and the fact that such uncertainty is why RS exist in the first place.
 Thus, users face a sequential decision-making problem under uncertainty.
@@
 We discuss how this provides an additional rationale as to why ``accurate'' recommendations are not always good recommendations.
+\section{Related Work.} 
 \par
-\xhdr{Related Work.} 
 The first set of related works studies the extent and implications of filter bubbles.
@@
 In particular, we consider that the realized value derived from a given item can be decomposed in the following manner: $x_{i,n}= v_{i,n} + \beta v_n$, where $v_{i,n}$ denotes an idiosyncratic component -- i.e.
-user $i$'s idiosyncratic taste for product $n$ --  and $v_{n}$, a common-value component.
+user $i$'s idiosyncratic taste for item $n$ --  and $v_{n}$, a common-value component.
 One can interpret $v_n$ as a measure of how much item $n$ is valued in society in general and, in a sense, $v_{i,n}$ denotes how $i$ diverges from this overall ranking.
@@
 We consider that the entry of $n$-th row and the ($m$)-th column of $\Sigma_i$ is given by $\sigma_i^2 \rho^{d(n,m)}$, and that of $\Sigma$ is given by $\sigma^2 \rho^{d(n,m)}$.
-The scalar $\rho \in [0,1]$ therefore impacts the covariance structure: a higher $\rho$ implies that learning the utility of $n$ is more informative about products nearby and, for $\rho \in (0,1)$, this effect is decreasing in distance.
+The scalar $\rho \in [0,1]$ therefore impacts the covariance structure: a higher $\rho$ implies that learning the utility of $n$ is more informative about items nearby and, for $\rho \in (0,1)$, this effect is decreasing in distance.
 The particular distance function that we rely on leads to a simple covariance structure, where the $(n,n+1)$-th entry in the covariance matrix is $\sigma^{2} \rho$, the $(n,n+2)$-th entry is $\sigma^{2} \rho^2$, etc.\footnote{This exponential decay correlation structure can be related to the tenet of case-based similarity of \cite{gilboa1995case} -- see \cite{billot2008axiomatization} for an axiomatization of exponential similarity.}
@@
 In period 1, every item is ex-ante identical since they have the same mean and variance and so suppose that the user breaks the tie arbitrarily and consumes item 0.
-The underlying correlation structure implies that upon observing that $x_0 = y$ the user will update beliefs about the remaining 3 items according to the previously specified updating rule.
+The underlying correlation structure implies that upon observing that $x_0 = y$ the user will update beliefs about the remaining three items according to the previously specified updating rule.
 For concreteness, we suppose that $\sigma = 1$ and $\rho = 0.5$, but the intuitions hold for any value of $\sigma$ and $\rho > 0$.
@@
 Our model of recommendation is stylized in order to provide qualitative insights into how recommendation shapes behavior, instead of focusing on the details of how RS are implemented in practice.
-We model recommendation as giving users information about the valuation of the products.
+We model recommendation as giving users information about the valuation of the items.
 \par
@@
 \end{finding}
-
 \afterpage{
@@
 \caption*{\scriptsize Notes: The figure shows the consecutive consumption path difference between the no recommendation, recommendation, and oracle regime.
-The figure on the left displays the average consecutive consumption distance aggregating over simulations where $\rho \in (0,1)$ and the figure on the right displays the average consecutive consumption distance aggregating over simulations where $\rho = 0$.}
+The figure on the left displays the average consecutive consumption distance aggregating over simulations where $\rho \in (0,1)$ and the figure on the right displays the average consecutive consumption distance aggregating over simulations where $\rho = 0$.
+The shaded area represents the 95\% confidence interval.}
 \label{fig:correlation_consumption_path}
@@
 This is due to the fact that when $\rho = 0$, there is no reason that items that are close in the product space should have similar values and so the optimal consumption path does not depend on the similarity of the items.
-However this also means that users do not learn anything about neighboring products and so there is limited path-dependence in consumption.
+However this also means that users do not learn anything about neighboring items and so there is limited path-dependence in consumption.
 Not only is there no difference in the levels between the three regimes, but they all have the same, flat, average consumption distance path.
@@
 This is due to the fact that the spillovers not only impact the mean expected belief about quality but also the degree of uncertainty.
-Local consumption therefore leads to users having less uncertainty about certain areas of the product space and risk aversion may lead them to increasingly consume nearby products.
+Local consumption therefore leads to users having less uncertainty about certain areas of the product space and risk aversion may lead them to increasingly consume nearby items.
 
@@
 We have shown that, unless users are provided with additional information to guide their consumption choices, then these information spillovers and user risk-aversion can lead users into filter bubbles.
-When users consume high valued items, they exploit the underlying correlation across different items' values, stronger for similar items, which leads them to increasingly consume products in narrower and narrower portions of the product space.
+When users consume high valued items, they exploit the underlying correlation across different items' values, stronger for similar items, which leads them to increasingly consume items in narrower and narrower portions of the product space.
 Risk aversion may lead users into performing local consumption even when they have a low valuation of nearby items just because they know what to expect from the item.
-Recommendation leads to these effects being mitigated by providing users with additional information on products outside the already explored portions of the product space.
+Recommendation leads to these effects being mitigated by providing users with additional information on items outside the already explored portions of the product space.
 Logically, if all uncertainty were resolved as in the oracle regime, then such behavior is not present.
 
-\subsection{User Welfare and Product Diversity}
+\subsection{User Welfare and Item Diversity}
 
@@
 \noindent 
-Finding~\ref{finding_diversity} summarizes the main results on product diversity:
+Finding~\ref{finding_diversity} summarizes the main results on item diversity:
 \begin{finding}\label{finding_diversity}
-The impact of recommendation on product diversity:
+The impact of recommendation on item diversity:
 \begin{enumerate}
-\item When $\rho = 0$, product diversity is the same across all three recommendation regimes;
-\item When $\rho \in (0,1)$, product diversity decreases across all recommendation regimes but decreases the most in the no-recommendation regime.
+\item When $\rho = 0$, item diversity is the same across all three recommendation regimes;
+\item When $\rho \in (0,1)$, item diversity decreases across all recommendation regimes but decreases the most in the no-recommendation regime.
 This effect is amplified as $\rho$ increases as well as when users become more risk-averse.
@@
 \centering
-\caption{Relationship between Local Consumption and Correlation ($\rho$) and Risk Aversion ($\gamma$)}
+\caption{Relationship between Local Consumption and Correlation ($\rho$), Risk Aversion ($\gamma$)}
 \begin{subfigure}{.3\linewidth}
@@
 \addtocounter{figure}{-1}
-As before, when there is no correlation between valuations, product diversity is the same across different recommendation regimes.
-The over-exploitation of information spillovers when $\rho \in (0,1)$ leads to product diversity being lowest in the no-recommendation regime.
+
+As before, when there is no correlation between valuations, item diversity is the same across different recommendation regimes.
+The over-exploitation of information spillovers when $\rho \in (0,1)$ leads to item diversity being lowest in the no-recommendation regime.
 As a result, this effect gets amplified as $\rho$ increases, which leads to the gap in diversity between the regimes to increase as $\rho$ increases.
@@
 \begin{figure}[t]
-\caption{Relationship between Correlation ($\rho$) and Risk Aversion ($\gamma$), and User Welfare and Diversity}\label{fig:diversity_welfare_correlation}
+\caption{Relationship between User Welfare, Diversity and Correlation ($\rho$), Risk Aversion ($\gamma$)}\label{fig:diversity_welfare_correlation}
 \begin{subfigure}{0.3\linewidth}
@@
 \begin{subfigure}{0.3\linewidth}
-\includegraphics[width=1.0\linewidth]{figures/gamma_diversity_N_200_T_20.pdf}
+\includegraphics[width=1.0\linewidth]{figures/gamma_diversity_N_200_T_20}
 \end{subfigure}
@@
 \end{subfigure}
-\caption*{\scriptsize Notes: The figures on the left and center display the relationship between $\rho$ and overall consumed product diversity (left) as well as $\gamma$ and overall consumed product diversity (center).
-The figure on the right displays the relationship between $\rho$ and overall welfare.}
+\caption*{\scriptsize Notes: The figures on the left and center display the relationship between $\rho$ and overall consumed item diversity (left) as well as $\gamma$ and overall consumed item diversity (center).
+The figure on the right displays the relationship between $\rho$ and overall welfare.
+The shaded area represents the 95\% confidence interval.}
 \end{figure}
@@
 \end{minipage}\\
-\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under recommendation for$\gamma = 0$ (first), $\gamma = 5$ (second) and no recommendation for $\gamma = 0$ (third), $\gamma = 5$ (fourth).}\label{fig:diversity_welfare_ra}
+\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under no recommendation, with $\gamma = 0$ (first) and $\gamma = 5$ (second), and under recommendation, with $\gamma = 0$ (third) and $\gamma = 5$ (fourth).}\label{fig:diversity_welfare_ra}
 \end{figure}
@@
 This is since, with no risk-aversion, a user will select the item that she currently believes has the highest expected value.
-High product diversity in this case can arise from a user who consumes an item she disliked and updates her beliefs about nearby items negatively.
+High item diversity in this case can arise from a user who consumes an item she disliked and updates her beliefs about nearby items negatively.
 As a result, in the following period she will pick an item far away in the product space from the item that was previously consumed.
 If instead the user valued highly the item that she had consumed, then she is more likely to pick a nearby item.
-The information spillovers therefore lead to high product diversity being negatively correlated with welfare.
+The information spillovers therefore lead to high item diversity being negatively correlated with welfare.
 \par
@@
 \begin{figure}[t]
-\caption{Relationship between $\beta$ and Homogeneity}
+\caption{Relationship between Homogeneity and Common-Value Strength ($\beta$), Correlation ($\rho$)}
 \includegraphics[width=.3\linewidth]{figures/beta_homogeneity_N_200_T_20} \hspace{1.0cm}
@@
 \caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the weight of the common value component, $\beta$ (left) and correlation between valuations, $\rho$ (right).
-Each line represents this plot for a single recommendation regime.}
+Each line represents this plot for a single recommendation regime.
+The shaded area represents the 95\% confidence interval.}
 \end{figure}
@@
 As $\beta$ increases we expect that users should become increasingly homogeneous as the realized utilities of the items are now increasingly similar.~\autoref{fig:beta_homo} confirms that as the weight of the common-value component $\beta$ increases, users consume increasingly similar sets of items.
-The homogenization effect is strongest under the recommendation regime since the revelation of the common-value component induces users to consume products in similar areas of the product space.
+The homogenization effect is strongest under the recommendation regime since the revelation of the common-value component induces users to consume items in similar areas of the product space.
 As $\beta$ increases, some amount of homogenization is optimal as can be seen from the oracle case.
@@
 As was highlighted in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, the degree of local consumption increases with $\rho$.
-Even though the revelation of the common-value component induces them to search in similar parts of the product space, their idiosyncratic components induce them to consume products in a more localized area of the product space as $\rho$ increases which leads to a decline in homogeneity.
+Even though the revelation of the common-value component induces them to search in similar parts of the product space, their idiosyncratic components induce them to consume items in a more localized area of the product space as $\rho$ increases which leads to a decline in homogeneity.
 
@@
 \par
-This intuition implies that RS should collect additional data beyond the type of data that is traditionally recorded.
+This intuition implies that RS should collect additional data beyond that which is traditionally recorded.
 The first and most crucial type of data to collect is individual user \textit{beliefs} about items that they have not yet consumed.
@@
 Our model posits that recommendation provides value to users by providing them with information about the true valuation of an item if they were to consume it.
-Thus, the prediction problem for the recommender becomes predicting what item would the user choose with no recommendation and, correspondingly, what would be the most useful information to provide to the user that would lead her to consume a \textit{better} item than she would without recommendation.
+Thus, the prediction problem for the recommender becomes predicting what item the user would choose with no recommendation and, correspondingly, what would be the most useful information to provide to the user that would lead her to consume a \textit{better} item than she would without recommendation.
 This links back to the intuition our model provided for the \textit{John Wick} example whereby collecting user beliefs and measuring how the user updated beliefs about similar items would lead the recommender to understand that the user would consume \textit{John Wick: Chapter Two}.
@@
 The primary difference between our proposed approach and those existing in the literature is that ours crucially hinges on understanding user beliefs and the risk-preferences of users.
-For instance, \cite{vargas2011rank, kaminskas2014measuring} propose unexpectedness metrics that look at the dissimilarity of the proposed recommended items compared to what the recommender already knows the user likes - either via a content-based approach or a collaborative-based approach.
+For instance, \cite{vargas2011rank, kaminskas2014measuring} propose unexpectedness metrics that look at the dissimilarity of the proposed recommended items compared to what the recommender already knows the user likes.
 This metric depends only on the proposed item-set and not necessarily on the user's beliefs or how such a recommendation will change the item that the user consumes.
-\cite{adamopoulos2015unexpectedness} propose an expected utility approach but use this to measure unexpectedness in terms of deviations from the ``expected consideration set'' of a user and do not explicitly incorporate user beliefs.
+\cite{kotkov2018investigating} provide a comprehensive overview of possible definitions of serendipity and ours is closest to their ``motivational novelty" definition, which is that the user was persuaded to consume an item as a result of recommendation.
 \par 
@@
 Our evaluation measure leads to useful recommendations since it leads users towards better items than they would consume without recommendation.
-It further results in ``unexpected'' recommendations since it explicitly incorporates user beliefs and thus allows the recommender to understand how ``unexpected'' a recommendation would be from the perspective of a user.
+It further results in ``unexpected'' recommendations since it explicitly incorporates user beliefs and thus allows the RS to understand how ``unexpected'' a recommendation would be from the perspective of a user.
+Finally, such a measure may lead to a perceived broadening of user preferences as has been discussed in \cite{herlocker2004evaluating, zhang2012auralist}.
+However, under our interpretation, it may be that their underlying preferences are unchanged and, instead, that recommendation and consumption themselves provides information that encourages users to explore different portions of the product space.
+\newpage
 % Bibliography
